### 2019-05-20 boek.tex
initial commit

\documentclass[11pt,a4paper]{book}

%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORT PACKAGES %%%
%%%%%%%%%%%%%%%%%%%%%%%

% adjust margins (should be done before importing fancyhdr!)
\usepackage[margin=2.5cm,a4paper]{geometry} %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm

% text packages
\usepackage[british]{babel}
\usepackage{setspace}

% scripting packages
\usepackage{ifthen}
\usepackage{blindtext}
\usepackage{todonotes}

% style packages
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage[]{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion

% used for defining abbreviations, see nomenclature.tex
\usepackage{nomencl}

% figure packages
\usepackage{tikz,pgfplots}
\usetikzlibrary{calc,shapes,arrows,intersections,shadows}
\tikzset{>=latex}
\usepackage{graphicx}
\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{import}
\graphicspath{{fig/}}
\usepackage{xcolor}

% pseudocode packages
\usepackage{algorithm}
\usepackage{algpseudocode}

% symbols packages
\usepackage{amssymb}
\usepackage{amsmath}

% table packages
\usepackage{tabularx}
\usepackage{tabulary}

% references packages
\usepackage[sorting=none]{biblatex}
\bibliography{library}

% link packages
\usepackage{url}
\usepackage{doi}

%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE HEADER %%%
%%%%%%%%%%%%%%%%%%%%%
\usepackage{fancyhdr}  
\pagestyle{fancy} 
\renewcommand{\chaptermark}[1]{\markboth{#1}{}}
\fancypagestyle{frontmatter}{%    
	\fancyhf{} 	
	\fancyfoot[RO,LE]{\textsf{Page \thepage}} 
	\fancypagestyle{plain}{
		\renewcommand{\headrulewidth}{0pt}
		\fancyhead{}
		\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	} 
}
\fancypagestyle{mainmatter}{%    
	\fancyhf{} 	
	\fancyhead[RO,LE]{\nouppercase{\small \textsf{\leftmark}}}
	\fancyfoot[RO,LE]{\textsf{Page \thepage}} 
}
\setlength{\headheight}{15pt}

\newcounter{letternum}
\newcounter{lettersum}
\setcounter{lettersum}{13}
\newlength{\thumbtopmargin}
\setlength{\thumbtopmargin}{1cm}
\newlength{\thumbbottommargin}
\setlength{\thumbbottommargin}{3cm}
\newlength{\thumbheight}
\pgfmathsetlength{\thumbheight}{%
	(\paperheight-\thumbtopmargin-\thumbbottommargin)/\value{lettersum}}

\newlength{\thumbwidth}
\setlength{\thumbwidth}{1.5cm}

\tikzset{
	thumb/.style={
		%   draw=black,
		fill=light-gray,
		text=black,
		minimum height=\thumbheight, %\thumbheight,
		text width=\thumbwidth,
		outer sep=0pt,%   outer sep=10pt,
		font=\sffamily\Huge,
	}
}
\newcommand{\oddthumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north east,] at ($%
	(current page.north east)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
\newcommand{\eventhumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north west,] at ($%
	(current page.north west)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
% create a new command to set a new lettergroup
\newcommand{\lettergroup}[1]{%
	\fancyhead[LO]{\oddthumb{#1}}%
	\fancyhead[RE]{\eventhumb{#1}}%
	\fancypagestyle{chapterstart}{%
		\renewcommand{\headrulewidth}{0pt}
		\renewcommand{\footrulewidth}{0pt}
		\fancyhf{}
		\chead{\oddthumb{#1}}% chapters start only on odd pages
		\fancyfoot[RO,LE]{\textsf{Page \thepage}} %\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	}
	\thispagestyle{chapterstart}
	\stepcounter{letternum}%
}

%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE STYLES %%%
%%%%%%%%%%%%%%%%%%%%%

% use Sonny chapter style
\usepackage[Sonny]{fncychap} %Sonny

\definecolor{light-gray}{gray}{0.90} %not always visible on a dell screen!!!

% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}
\hypersetup{%
	pdftitle = {A title},
	pdfsubject = {A subject},
	pdfkeywords = {import keywords},
	pdfauthor = {Me of course},
	pdfcreator = {\LaTeX},
}

% Set style variables
\setlength{\parindent}{0cm} 
\renewcommand{\contentsname}{Table of Contents}
\renewcommand{\listfigurename}{List of Figures}
\renewcommand{\listtablename}{List of Tables}
\renewcommand{\appendixname}{}

\begin{document}

%\onehalfspace

% thesis obligatory content: 
% * an abstract in Dutch
% * a list of abbreviations
% * one uniform (compiling) reference list
% * a resume or Curriculum Vitae of the candidate
% * an introductory chapter and additional linking texts between chapters (papers) if the dissertation consists of a compilation of papers
% * acknowledgements of financing institutions
% * apply for an ISBN number for your thesis. This takes 1-2 days.

% geen paginanummering tot we aan de inhoudsopgave komen
\pagenumbering{alph}

\frontmatter
\pagestyle{empty}

\include{ch0_frontmatter}

% tabel van afkortingen en symbolen

% hoofdstukken
\mainmatter
\pagestyle{mainmatter}

\include{ch1_singlecellbiology}

\include{ch2_dynbenchmark}

\include{ch3_scorpius}

\include{ch4_dyno}

\include{ch5_incgraph}

\include{ch6_brood}

\include{ch7_discussion}

% appendices

\appendix

\printbibliography

\backmatter

% eventueel: lijst van figuren en tabellen
%\listoffigures
%\listoftables

% lege pagina (!!)

\end{document}

### 2019-05-20 ch0_frontmatter.tex
initial commit

%%%%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE TITLEPAGE %%%
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{titlepage}
	\centering
	\includegraphics[width=\textwidth]{fig/DI-eps-converted-to.pdf}

	\vspace{1cm}

	\includegraphics[width=2.25cm]{fig/ITGlogo_new} 
	\vspace{2.25cm}

	{\Huge \textsf{Proteomic analysis of \textit{Taenia} spp. excretion/secretion proteins: the search for \textit{Taenia solium}-specific diagnostic antigens}}

	\vspace{2.25cm}

	{\LARGE \textsf{Bjorn Victor}}

	\vspace{2.25cm}

	{\textsf{Proefschrift voorgedragen tot het behalen van de graad van \\Doctor in de Diergeneeskundige Wetenschappen, 2014}}

	\vfill

	{\textsf{Promotoren: \\
			Prof. Dr. Pierre Dorny\\Dr. Sarah Gabri\"el\\ Dr. Magnus Palmblad}}

	\vfill

	{\textsf{Vakgroep Virologie, Parasitologie en Immunologie\\
			Faculteit Diergeneeskunde, Universiteit Gent \\
			Salisburylaan 133, B-9820 Merelbeke}}
\end{titlepage}

%%%%%%%%%%%%%%%%%%%%%%%%
%%% ACKNOWLEDGEMENTS %%%
%%%%%%%%%%%%%%%%%%%%%%%%

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Acknowledgements}
\blindtext

\newpage{\thispagestyle{empty}\cleardoublepage}

%%%%%%%%%%%%%
%%% LISTS %%%
%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\tableofcontents

\newpage{\thispagestyle{empty}\cleardoublepage}
\listoffigures

\newpage{\thispagestyle{empty}\cleardoublepage}
\listoftables

\newpage{\thispagestyle{empty}\cleardoublepage}

%%%%%%%%%%%%%%%%%%%%
%%% NOMENCLATURE %%%
%%%%%%%%%%%%%%%%%%%%
\makenomenclature
\printnomenclature

\newcommand{\grn}{GRN}
\newcommand{\grns}{GRNs}
\nomenclature{GRN}{Gene Regulatory Network}

\newcommand{\im}{IM}
\newcommand{\ims}{IMs}
\nomenclature{IM}{Importance Measure}

\newcommand{\tf}{TF}
\newcommand{\tfs}{TFs}
\nomenclature{TF}{Transcription Factor}

\newcommand{\dna}{DNA}
\nomenclature{DNA}{Deoxyribonucleic Acid}

\newcommand{\rna}{RNA}
\nomenclature{RNA}{Ribonucleic Acid}

\newcommand{\mrna}{mRNA}
\nomenclature{mRNA}{Messenger RNA}

\newcommand{\nin}{NI}
\nomenclature{NI}{Network Inference}

\newcommand{\ml}{ML}
\nomenclature{ML}{Machine Learning}

\newcommand{\rf}{RF}
\nomenclature{RF}{Random Forests}

\newcommand{\cart}{CART}
\nomenclature{CART}{Classification And Regression Trees}

\newcommand{\vananh}{Huynh-Thu}%{V\^an Anh}

### 2019-05-20 ch1_singlecellbiology.tex
initial commit

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Single Cell Biology} 
\chaptermark{Single Cell Biology.} % short title
\label{chap:singlecellbiology}
\lettergroup{\thechapter}

\begin{quote}
	\textbf{Abstract:} \blindtext
\end{quote}

\vfill

Adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. Computational methods for trajectory inference from single-cell transcriptomics. \textit{European Journal of Immunology} 46, 11 (2016), 2496--2506. \doi{10.1002/eji.201646347}.\\
{\footnotesize $^*$ Equal contribution}
\newpage

\blindtext

### 2019-05-20 ch2_dynbenchmark.tex
initial commit

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Benchmarking trajectory inference methods} 
\chaptermark{Benchmarking trajectory inference methods.} % short title
\label{chap:dynbenchmark}

\begin{quote}
	\textbf{Abstract:} \blindtext
\end{quote}

\vfill

Adapted from:\\
Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y. A comparison of single-cell trajectory inference methods. \textit{Nature Biotechnology} 37, 5 (2019), 547--554. \doi{10.1038/s41587-019-0071-9}.\\
{\footnotesize $^*$ Equal contribution}
\newpage

\blindtext

### 2019-05-20 ch3_scorpius.tex
initial commit

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{SCORPIUS} 
\chaptermark{SCORPIUS.} % short title
\label{chap:scorpius}

\begin{quote}
	\textbf{Abstract:} \blindtext
\end{quote}

\vfill

Adapted from:\\
\textbf{Cannoodt, R.}, \ldots, De Preter, K., and Saeys, Y. SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development. \textit{Journal} vol, issue (2019), page--page. \doi{10.1101/079509v2}.
\newpage

\blindtext

### 2019-05-20 ch0_frontmatter.tex
update

@@
 
-	{\LARGE \textsf{Bjorn Victor}}
+	{\LARGE \textsf{Robrecht Cannoodt}}
 
@@
 
-	{\textsf{Proefschrift voorgedragen tot het behalen van de graad van \\Doctor in de Diergeneeskundige Wetenschappen, 2014}}
+	{\textsf{A thesis submitted in fulfillment of the requirements for the degree of \\Doctor in Computer Science, 2019}}
 
@@
 
-	{\textsf{Promotoren: \\
-			Prof. Dr. Pierre Dorny\\Dr. Sarah Gabri\"el\\ Dr. Magnus Palmblad}}
+	{\textsf{Supervisors: \\
+			Prof. Dr. Yvan Saeys\\Prof. Dr. Katleen De Preter}}
 
@@
 
-	{\textsf{Vakgroep Virologie, Parasitologie en Immunologie\\
-			Faculteit Diergeneeskunde, Universiteit Gent \\
-			Salisburylaan 133, B-9820 Merelbeke}}
+	{\textsf{Vakgroep Toegepaste Wiskunde, Informatica, en Statistiek\\
+			Faculteit Wetenschappen, Universiteit Gent \\
+			Krijgslaan 281 - S2, 9000 Gent}}
 \end{titlepage}

### 2019-05-22 ch0_frontmatter.tex
update document

@@
 
-	{\Huge \textsf{Proteomic analysis of \textit{Taenia} spp. excretion/secretion proteins: the search for \textit{Taenia solium}-specific diagnostic antigens}}
+	{\Huge \textsf{Studying dynamic processes\\with networks and trajectories}}
 
@@
 
-	{\textsf{A thesis submitted in fulfillment of the requirements for the degree of \\Doctor in Computer Science, 2019}}
+	{\textsf{Thesis submitted in fulfilment of the requirements for the degree of \\Doctor in Computer Science, 2019}}
 

### 2019-05-22 ch1_introduction.tex
update document

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Introduction} 
\chaptermark{Introduction.} % short title
\label{chap:introduction}
\lettergroup{\thechapter}

\begin{quote}
	\textbf{Abstract:} Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner. Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following. In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods. We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation. As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
\end{quote}

\vfill

Adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. Computational methods for trajectory inference from single-cell transcriptomics. \textit{European Journal of Immunology} 46, 11 (2016), 2496--2506. \doi{10.1002/eji.201646347}.\\
{\footnotesize $^*$ Equal contribution}
\newpage

\section{Single cell biology}

\subsection{Dynamic processes}

\subsection{Gene regulation}

\subsection{Single cell transcriptomics}

\section{Machine learning}

\subsection{Supervised learning}

\subsection{Unsupervised learning}

\subsection{Feature selection}

\subsection{Trajectory inference}

\subsection{Network inference}

\section{Research objectives}

\section{Outline}

### 2019-05-22 ch6_incgraph.tex
update document

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Optimising regulatory networks} 
\chaptermark{Optimising regulatory networks.} % short title
\label{chap:incgraph}

\begin{quote}
	\textbf{Abstract:} \blindtext
\end{quote}

\vfill

Adapted from:\\
\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y. IncGraph: Incremental graphlet counting for topology optimisation. \textit{PLOS ONE} 13, 4 (2018), e0195997. \doi{10.1371/journal.pone.0195997}.
\newpage

\blindtext \cite{Eren2013}

### 2019-05-22 chX_backmatter.tex
update document

\appendix

\backmatter

\chapter{Samenvatting}

\chapter{Summary}

\chapter{List of Publications}

\printbibliography

### 2019-05-22 ch6_incgraph.tex
update manuscript

@@
 \begin{quote}
-	\textbf{Abstract:} \blindtext
+	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
+	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
+	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
+\\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
+	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
+	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
+	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
+\\
+	IncGraph is freely available as an open-source R package on CRAN (incgraph). The development version is also available on GitHub (\href{http://www.github.com/rcannood/incgraph}{rcannood/incgraph}).
 \end{quote}
@@
 
-\blindtext \cite{Eren2013}
+\section*{Introduction}
+Even the simplest of living organisms already consist of complex biochemical networks which must be able to respond to a variety of stressful conditions in order to survive. 
+An organism can be characterised using numerous interaction networks, including gene regulation, metabolic, signalling, and protein-protein interaction. 
+The advent of high-throughput profiling methods (e.g.
+microarrays and RNA sequencing) have allowed to observe the molecular contents of a cell, which in turn have enabled computational network inference methods to reverse engineer the biochemical interaction networks~\cite{Albert2007}.
+Improving the accuracy of inferred networks has been a long-standing challenge, but the development of ever more sophisticated algorithms and community-wide benchmarking studies have resulted in significant process~\cite{Marbach2010, Narendra2011, Marbach2012, Aijo2017}.
+
+Several recent developments involve incorporating topological priors, either to guide the inference process~\cite{Lopes2014} or post-process the network~\cite{Ruyssinck2016}.
+A common prior is that biological networks are highly modular~\cite{Rives2003}, as they consist of multiple collections of functionally or physically linked molecules~\cite{Hartwell1999,Barabasi2004}.
+At the lowest level, each module is made up out of biochemical interactions arranged in small topological patterns, which act as fundamental building blocks~\cite{Milo2002}.
+
+Graphlets~\cite{Przulj2004} are one of the tools which could be used to add a topological prior to a biological network,
+Graphlets are small connected subnetworks which can be counted to identify which low-level topological patterns are present in a network.
+By comparing how topologically similar a predicted network is to what would be expected of a true biological network, a predicted network can be optimised in order to obtain a better topology.
+
+By counting the number of occurrences of each of the different graphlets (Fig~\ref{fig_graphlets}A) touching a specific node, one can characterise the topology surrounding it.
+The graphlet counts of a network can be represented as a matrix with one row for each of the nodes and one column for each of the graphets (Fig~\ref{fig_graphlets}B).
+An orbit represents a class of isomorphic (i.e. resulting in the same structure) positions of nodes within a graphlet (Fig \ref{fig_graphlets}A, coloured in red).
+Both graphlets and orbits have been used extensively for predicting the properties of nodes such as protein functionality~\cite{Milenkovic2008,Guerrero2008,Singh2014} and gene oncogenicity~\cite{Milenkovic2010a}, by performing network alignment~\cite{Kuchaiev2010,Milenkovic2013} or using them as a similarity measure in machine learning tasks~\cite{Shervashidze2009,Vacic2010}.
+
+%% FIGURE 1
+\picfig{fig/incgraph/figure1.pdf}{\LARGEfigure}{fig_graphlets}{{\bf Graphlet counting in a network characterises its local topologies.}
+	(A) In total, there are 30 different graphlets containing 2 to 5 nodes, ranging from $G_0$ to $G_{29}$.
+Orbits are an extension of graphlets which also take into account the position of a node within a graphlet.
+The 73 different orbits are coloured in red.
+(B) By counting the occurrences of these graphlets in the network, the local topology surrounding a node can be quantified.}
+
+In this work, we focus on optimising gene regulatory networks by incorporating a topological prior as part of a topology optimisation process.
+We seek to optimise a predicted network by iteratively modifying the network and accepting modifications that lead to topological properties that resemble better those of true biological networks.
+
+However, using graphlets to perform topology optimisation was hitherto not possible.
+Calculating the graphlet counts using the most state-of-the-art graphlet counting of a moderately sized gene regulatory network already has an execution time of about five seconds (\textit{E.
+coli}, $\sim 3000$ genes, $\sim 10000$ interactions, up to graphlets up to 5 nodes).
+While this computational time poses no issue for regular static networks, recalculating all graphlet counts for every network modification made as part of a topology optimisation is computationally intractable.
+For example, performing 100'000 iterations of topology optimisation on a similarly sized network and calculating the topological impact of 10 possible edge modification at each iteration, already results in a computational time of about 12 days.
+Graphlet-based topology optimisation thus quickly becomes infeasible for larger networks.
+
+When adding or removing an edge to a large network, the number of altered graphlets is much smaller than the total number of graphlets in the network.
+It could therefore be much more efficient to iterate over and count all the graphlets that have been added or removed as a result of the edge modification, than it is to recalculate the graphlet counts from scratch.
+
+Eppstein et al.
+introduced data structures and algorithms for updating the counts of size-three\cite{Eppstein2009} and size-four\cite{Eppstein2012} subgraphs in a dynamic setting.
+The data structures were determined such that the amortised time is $O(h)$ and $O(h^2)$, respectively, where $h$ is the h-index of the network\cite{Hirsch2005}.
+
+We developed IncGraph, an alternative algorithm and implementation for performing incremental counting of graphlets up to size five.
+We show empirically that IncGraph is several orders of magnitude faster at calculating the differences in graphlet counts in comparison to non-incremental counting approaches.
+In addition, we demonstrate the practical applicability by developing a graphlet-based optimisation criterion for biological networks.
+
+\section*{Materials and methods}
+Assume a network $G$ of which the graphlet counts $C_G$ are known.
+$C_G$ is an $n$-by-$m$ matrix, with $n$ the number of vertices in the network, $m = 73$ is the number of different orbits, and where $C_G[i,j]$ is the number of times node $i$ is part of a graphlet at orbit $O_j$.
+Further assume that one edge has either been added or removed from $G$, resulting in $G'$, at which point the counts $C_{G'}$ need to be observed.
+If multiple edges have been modified, the method described below can be repeated for each edge individually.
+
+\subsection*{Incremental graphlet counting}
+As stated earlier, recalculating the graphlet counts for each modification made to the network quickly becomes computationally intractable for larger network sizes.
+However, as the differences in topology between $G$ and $G'$ are small, it is instead possible to calculate the differences in graphlet counts $\Delta_{G, G'}$ instead.
+This is much more efficient to calculate, as only the neighbourhood of the modified edges needs to be explored.
+$C_{G'}$ can thus be calculated as $C_{G'} = C_G + \Delta_{G,G'}$.
+
+The differences in graphlet counts $\Delta_{G, G'}$ are calculated by iterating recursively over the neighbours surrounding each of the modified edges (See \nameref{code_incgraph}).
+Several strategies are used in order to calculate $\Delta_{G,G'}$ as efficiently as possible (Fig~\ref{fig_efficient}).
+(A) The delta matrix is calculated separately for each modified edge.
+Since the edge already contains two out of five nodes and any modified graphlet is a connected subgraph, the neighbourhood of this edge only needs to be explored up to depth 3 in order to iterate over all modified graphlets.
+(B) We propose a lookup table to look up the graphlet index of each node of a given subgraph.
+By weighting each possible edge in a 5-node graphlet, the sum of the edges of a subgraph can be used to easily look up the corresponding graphlet index.
+(C) During the recursive iteration of the neighbourhood, the same combination of nodes is never visited twice.
+(D) As the network can be relatively large, the adjacency matrix is binary compressed in order to save memory.
+One integer requires 4 bytes and contains the adjacency boolean values of 32 edges, whereas otherwise 32 booleans would require 32 bytes.
+For example, 1GB of memory is large enough to store a compressed adjacency matrix of 92681 nodes.
+If the network contains too many nodes to fit a compressed adjacency matrix into the memory, a list of sets containing each node's neighbours is used instead.
+
+%% FIGURE 2
+\picfig{fig/incgraph/figure2.pdf}{\LARGEfigure}{fig_efficient}{{\bf Several strategies are employed in order to reduce time and memory consumption.}
+(A) Only the depth 3 neighbourhood of each modified edge needs to be explored in order to have visited all modified five-node graphlets.
+(B) A lookup table can be used to easily look up the graphlet index of a subgraph, by weighing each edge in a 5-node subgraph by a power of~2.
+(C) The same combination of five nodes is never repeated, as to avoid counting the same graphlet multiple times.
+(D) The adjacency matrix of the network is compressed in order to reduce memory usage.}
+
+IncGraph supports counting graphlets and orbits of subgraphs up to five nodes in undirected networks.
+By modifying the lookup table, the method can be easily extended to directed graphlets or larger-node graphlets, or limited to only a selection of graphlets.
+This allows for variations of the typical graphlets to be studied in an incremental setting.
+
+\subsection*{Timing experiments}
+We compared the execution time needed to calculate the graphlet counts in iteratively modified networks between our method and a state-of-the-art non-incremental algorithm, Orca~\cite{Hocevar2014}.
+Orca is a heavily optimised algorithm for counting 5-node graphlets in static networks, and outperforms all other static graphlet counting algorithms by an order of magnitude~\cite{Hocevar2014}.
+
+The timing experiments were performed by generating networks from 3 different network models, making modifications to those networks while still adhering to the network model, and measuring the execution times taken for both approaches to calculate the new graphlet counts (Fig~\ref{fig_method_speedup}).
+
+%% FIGURE 3
+\picfig{fig/incgraph/figure3.pdf}{\LARGEfigure}{fig_method_speedup}{{\bf Static network model generators were modified to generate dynamic networks.}
+	Three network models were used: Barabási–Albert, Erdős–Rényi, and Geometric.
+Step 0: a network is generated according to the network model and the given parameters.
+Step 1: the network is modified such that the result is as likely to have been generated by the network model.
+Step 2: The time for IncGraph to calculate the differences in graphlet counts is measured ($T_{IG}$).
+Step 3: The time for the non-incremental approach to calculate the graphlet counts of the new network is measured ($T_{NI}$).
+Steps 1 to 3 are repeated until all modifications generated at step 0 are processed, or until an execution time threshold has been reached.}
+
+The network models were based on three static network models: Barab\'asi-Albert~\cite{Albert2002}, Erd\H{o}s--R\'enyi~\cite{Erdos1959}, and Geometric~\cite{Appel1997}.
+Pseudo code for these random evolving network models can be found in \nameref{code_ba}, \nameref{code_er}, and \nameref{code_geo} respectively.
+Each model generates an initial network according to the static network model it is based on, and a list of network modifications (removing an edge from or adding an edge to the network).
+These network modifications are made such that at any given time point in the evolving network, it is likely that the network at its current state could have been generated by the original static network model.
+
+Networks were generated with varying network models, between 1000 and 16000 nodes, node degrees between 2 and 64, and 10000 time points.
+We measured the time needed to calculate the delta matrix at random time points until 1 hour has passed.
+All timings experiments were carried out on Intel(R) Xeon(R) CPU E5-2665 $@$ 2.40GHz processors, with one thread per processor.
+The generation of networks with higher node counts or degrees was constrained by the execution time of the network generators, not by IncGraph.
+All data and scripts are made available at \href{https://github.com/rcannood/incgraph-scripts}{github.com/rcannood/incgraph-scripts}.
+
+\subsection*{Gene regulatory network optimisation experiments}
+We demonstrate the usefulness of IncGraph by using a simple graphlet-based metric to optimise gene regulatory networks. 
+One of the striking differences between real and predicted gene regulatory networks is that the predicted networks contain highly connected subnetworks, which contain high amounts of false positives.
+We determine a penalty score such that predicted networks containing graphlets with many redundant edges will be penalised in comparison to very sparse networks.
+
+The \textit{redundancy penalty} (Fig~\ref{fig_method_reranking}A) of a network is defined as the sum of occurrences of each graphlet multiplied by the redundancy associated with each individual graphlet.
+The redundancy of a graphlet is the number of edges that can be removed without disconnecting the nodes from one another.
+By using the redundancy penalty score, we aim to improve the gene regulatory network (Fig~\ref{fig_method_reranking}B).
+
+The topology optimisation procedure uses an empty network as initialisation and grows the network by selecting interactions iteratively.
+Each iteration, the top $k = 100$ highest ranked interactions that are not currently part of the network are evaluated, and the highest ranked interaction passing the redundancy criterion is selected (Fig~\ref{fig_method_reranking}C).
+This procedure is repeated until a predefined amount of time has passed.
+As the aim of this experiment is not to obtain the highest performing topology optimisation method, parameter optimisation of $k$  has not been performed and is considered to be outside the scope of this work.
+
+%% FIGURE 4
+\picfig{fig/incgraph/figure4.pdf}{\hugefigure}{fig_method_reranking}{{\bf Predicted gene regulatory networks of model organisms are optimised to reduce the false positive rate.}
+	A)~The number of redundant edges in each graphlet are counted.
+B) The network is optimised in order to obtain a lower redundancy over time.
+Two networks are shown, one before and one after the optimisation procedure.
+Edges coloured in red have been removed from the network after optimisation, green edges have been added.
+C) Starting from an empty network, the interactions are modified by iteratively evaluating the increase in redundancy of the next $k$ interactions, and adding the first edge for which its redundancy is less than the $90^\textrm{th}$ percentile redundancy.}
+
+We optimised gene regulatory networks of \textit{E.
+coli} and \textit{S.
+cerevisiae}.
+The predicted networks were generated using the network inference method GENIE3~\cite{Vananh2010} with default parameters.
+Gene expression data was obtained from COLOMBOS~\cite{Moretto2016} and GEO~\cite{Edgar2002}, respectively.
+The predicted networks and the optimised versions thereof were compared against respective lists of known gene regulatory interactions~\cite{Gama-Castro2016,Ma2014}.
+
+% Results and Discussion can be combined.
+\section*{Results and discussion}
+The contributions of this work are twofold.
+Firstly, we propose a new method for incrementally calculating the differences in graphlet counts in changing graphs, and show that it is orders of magnitude faster than non-incremental approaches.
+Secondly, we demonstrate its applicability by optimising a predicted gene regulatory network in order to reduce the false positive rate therein.
+
+\subsection*{Execution time is reduced by orders of magnitude}
+Timing experiments show that IncGraph is significantly faster in calculating the delta matrix in comparison to calculating the graphlet counts from scratch at each iteration (Fig~\ref{fig_speedup}).
+The observed speedup ratios between IncGraph and the non-incremental approach Orca ranges from about 50$\times$ to 10000$\times$.
+The speedup ratio increases as the network size increases.
+For larger networks, IncGraph can thus calculate the delta matrices of 10000 edge modifications while the non-incremental approach calculates one graphlet count matrix.
+
+Surprisingly, IncGraph obtains higher execution times for networks with 5657 nodes than for networks with 8000 nodes.
+One possible explanation is that the size of the data structures containing those networks are particularly favourable in avoiding cache misses.
+Confirmation of this explanation, however, would require further investigation.
+
+%% FIGURE 5
+\picfig{fig/incgraph/figure5.pdf}{\LARGEfigure}{fig_speedup}{{\bf IncGraph is significantly faster than non-incremental approaches.}
+	For small networks, the execution time of IncGraph $T_{IG}$ is already 50 times less than that of non-incremental approaches $T_{NI}$.
+This ratio increases even further for networks with higher numbers of nodes or higher average degrees.}
+
+Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (\nameref{sfig_hindex}).
+This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{Eppstein2012} for counting three-size and four-size subgraphs respectively.
+
+\subsection*{IncGraph allows for better regulatory network optimisation}
+We implemented a graphlet-based optimisation algorithm for improving the false positive rate of the predicted gene regulatory networks of \textit{E.
+coli} and \textit{S.
+cerevisiae}.
+After reranking the regulatory interactions, the F1 score of the first 1000~interactions had increased by 7.6\% and 2.2\% respectively (Fig~\ref{fig_reranking}A).
+The obtained speedup of about 15-30$\times$ (Fig~\ref{fig_reranking}B) is in line with the experiments on \textit{in silico} networks.
+Namely, for the \textit{E. coli} network at 9618~interactions and 889~nodes (average degree~=~10.8), a speedup of about 30$\times$ was obtained.
+Similarly, for the \textit{S. cerevisiae} network at 8013~interactions and 1254~nodes (average degree~=~6.4), a speedup of about 15$\times$ was obtained.
+These speedups are in the same order of magnitude of similarly sized networks (1000~nodes and 8000~interactions) generated with a Barab\'asi-Albert model, with a speedup of 65$\times$.
+This is to be expected, as such networks share the same scale-free property that gene regulatory networks have.
+
+%% FIGURE 6
+\picfig{fig/incgraph/figure6.pdf}{\Largefigure}{fig_reranking}{{\bf A simple graphlet-based scoring method improves predicted regulatory networks.}
+	(A) The F1 score was calculated by calculating the harmonic mean of the AUROC and AUPR scores of the first 1000 interactions.
+(B) IncGraph is significantly faster than the non-incremental approach.
+Note that for each interaction added to the network, the graphlet counts of 100 putative interactions were evaluated.}
+
+\section*{Conclusion}
+Many improvements over the past few years have resulted in efficient graphlet counting algorithms, even for large static networks.
+However, needing to perform the simplest of tasks tens of thousands of times quickly becomes computationally intractable.
+As such, recalculating the graphlet counts of a network after each of the many network modification is infeasible.
+
+This study introduces a method for calculating the differences in graphlet (and orbit) counts, which we call incremental graphlet counting or IncGraph for short.
+We show that IncGraph is at least 10-100 times faster than non-incremental methods for networks of moderate size, and that the speedup ratio increases even further for larger networks.
+To demonstrate the applicability of IncGraph, we optimised a predicted gene regulatory network by enumerating over the ranked edges and observing the graphlet counts of several candidate edges before deciding which edge to add to the network.
+
+IncGraph enables graphlet-based metrics to characterize online networks, e.g.
+in order to track certain network patterns over time, as a similarity measure in a machine learning task, or as a criterion in a topology optimisation.
+
+\section*{Supporting information}
+
+%% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Pseudocode.}
+\label{code_incgraph}
+{\bf IncGraph calculates $\Delta_{G,G'}$ using a strict branch-and-bound strategy.}
+
+\paragraph*{S2 Pseudocode.}
+\label{code_ba}
+{\bf Pseudo code for generating an evolving Barab\'asi-Albert (BA) network.} It first generates a BA network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a BA network.
+
+\paragraph*{S3 Pseudocode.}
+\label{code_er}
+{\bf Pseudo code for generating an evolving Erd\H{o}s--R\'enyi (ER) network.} It first generates an ER network, and then generates $o$ operations such that at any time point, the network is or very closely resembles an ER network.
+
+\paragraph*{S4 Pseudocode.}
+\label{code_geo}
+{\bf Pseudo code for generating an evolving geometric network.} It first generates a geometric network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a geometric network.
+
+\paragraph*{S1 Fig.}
+\label{sfig_hindex}
+{\bf Empirical measurements show a strong relation between the execution time of IncGraph and the h-index cubed of the network it was applied to.} This is in line with the findings by Eppstein et al., where counting 3-size subgraphs has an amortised time of $O(h)$ and counting 4-size subgraphs has an amortised time of $O(h^2)$.

### 2019-08-09 ch0_frontmatter.tex
update manuscript
update texs and files

@@
+\hypersetup{%
+	pdftitle = {Modelling single cell dynamics with trajectories and gene regulatory networks},
+	pdfsubject = {A subject},
+	pdfkeywords = {import keywords},
+	pdfauthor = {Robrecht Cannoodt},
+	pdfcreator = {\LaTeX},
+}
+
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-	{\Huge \textsf{Studying dynamic processes\\with networks and trajectories}}
+	{\Huge \textsf{Modelling single cell dynamics\\with trajectories and gene regulatory networks}}
 
@@
 
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\listoffigures
+%\newpage{\thispagestyle{empty}\cleardoublepage}
+%\listoffigures
 
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\listoftables
+%\newpage{\thispagestyle{empty}\cleardoublepage}
+%\listoftables
 
@@
 \nomenclature{CART}{Classification And Regression Trees}
-
-\newcommand{\vananh}{Huynh-Thu}%{V\^an Anh}

### 2019-08-09 ch2_dynbenchmark.tex
update document
update texs and files

@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Benchmarking trajectory inference methods} 
-\chaptermark{Benchmarking trajectory inference methods.} % short title
+\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
+\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
 \label{chap:dynbenchmark}
@@
 \begin{quote}
-	\textbf{Abstract:} \blindtext
+	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
+Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
+Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
+Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
+Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
+Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
 \end{quote}
@@
 
-\blindtext
+\section{Introduction}
+
+Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scaling_2017,Etzrodt2014-tr}.
+Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{Trapnell2015,cannoodt_computational_2016,moon_manifoldlearningbasedmethods_2018}.
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.
+TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process1, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{Aibar2017a}.
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_scienceforumhuman_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{Aibar2017a,angerer_singlecellsmake_2017} and user-friendly TI methods.
+
+A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1).
+Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecelllistsoftware_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
+While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect.
+Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{Haghverdi2016,Setty2016-og}) or through parameters provided by the user \cite{Trapnell2014,matsumoto_scoupprobabilisticmodel_2016}.
+These methods therefore mainly focus on correctly ordering the cells along the fixed topology.
+More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2017,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.
+
+Given the diversity in TI methods, it is important to quantitatively assess their performance, scalability, robustness and usability.
+Many attempts at tackling this issue have already been made \cite{Haghverdi2016,Ji2016-il,Welch2016-ue,matsumoto_scoupprobabilisticmodel_2016,duverle_celltreebioconductorpackage_2016,Cannoodt2016a,lonnberg_singlecellrnaseqcomputational_2017,campbell_probabilisticmodelingbifurcations_2017,wolf_graphabstractionreconciles_2017}, but a comprehensive comparison of TI methods across a large number of different datasets is still lacking.
+This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea of which would optimally solve their problem.
+Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
+
+In this study, we evaluated the accuracy, scalability, stability and usability of 45 TI methods (Figure \ref{fig:figure_1}a).
+We found substantial complementarity between current methods, with different sets of methods performing most optimally depending on the characteristics of the data.
+For method users, we created an interactive set of guidelines (available at \href{http://guidelines.dynverse.org}{guidelines.dynverse.org}), which gives context-specific recommendations for method usage.
+Our evaluation also highlights some challenges for current methods, and our evaluation strategy can be useful to spearhead the development of new tools that accurately infer trajectories on ever more complex use cases.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_1.pdf}
+	\caption{
+		\textbf{Overview of several key aspects of the evaluation.}
+		\textbf{a}, A schematic overview of our evaluation pipeline.
+\textbf{b}, To make the trajectories comparable to each other, a common trajectory model was used to represent reference trajectories from the real and synthetic datasets, as well as any predictions of TI methods.
+\textbf{c}, Trajectories are automatically classified into one of seven trajectory types, with increasing complexity.
+\textbf{d}, We defined four metrics, each assessing the quality of a different aspect of the trajectory.
+The HIM score assesses the similarity between the two topologies, taking into account differences in edge lengths and degree distributions.
+The F1branches assesses the similarity of the assignment of cells onto branches.
+The cordist quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances.
+Finally, wcorfeatures quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.
+	}
+	\label{fig:figure_1}
+\end{figure}
+
+\section{Results}
+
+\subsection{Trajectory inference methods}
+To make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (Figure \ref{fig:figure_1}b).
+In this model, the overall topology is represented by a network of ‘milestones’, and the cells are placed within the space formed by each set of connected milestones.
+Although almost every method returned a unique set of outputs, we were able to classify these outputs into seven distinct groups (Figure \ref{fig:supfigure_1}) and we wrote a common output converter for each of these groups (Figure \ref{fig:figure_2}a).
+When strictly required, we also provided prior information to the method.
+These different priors can range from weak priors that are relatively easy to acquire, such as a start cell, to strong priors, such as a known grouping of cells, that are much harder to know a priori, and which can potentially introduce a large bias into the analysis (Figure \ref{fig:figure_2}a).
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_1.pdf}
+	\caption{
+		\textbf{A common interface for TI methods.}
+		\textbf{a} The input and output of each TI method is standardized.
+As input, each TI method receives either raw or normalized counts, several parameters, and a selection of prior information.
+After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model.
+This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
+\textbf{b} Illustrations of the specific transformations performed by each of the wrapper functions.
+	}
+	\label{fig:supfigure_1}
+\end{figure}
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_2.pdf}
+	Figure caption on next page $\rightarrow$
+\end{figure}
+
+\clearpage % insert a page break
+\captionof{figure}{
+	\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
+	\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies.
+The methods are grouped vertically based on the most complex trajectory type they can infer.
+\textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation.
+Methods that errored on more than 50\% of the datasets are not included in this figure and are shown instead in Supplementary Fig.
+2.
+}
+\label{fig:figure_2}
+
+The largest difference between TI methods is whether a method fixes the topology and, if it does not, what kind of topology it can detect.
+We defined seven possible types of topology, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
+Most methods either focus on inferring linear trajectories or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (Figure \ref{fig:figure_2}a).
+
+We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript.
+Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b).
+We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig.
+2), after which we conclude with guidelines for method users and future perspectives for method developers.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_3.pdf}
+	Figure caption on next page $\rightarrow$
+\end{figure}
+\clearpage % insert a page break
+\captionof{figure}{	
+	\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
+	\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}.
+\textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types.
+The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
+\textbf{c}, Predicted execution times for varying numbers of cells and features (no.
+of cells $\times$ no.
+of features).
+Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features.
+k, thousands; m, millions; cor, correlation.
+\textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method.
+\textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category.
+Off-the-shelf methods were directly implemented in R and thus do not have a usability score.
+}
+\label{fig:figure_3}
+
+\subsection{Accuracy}
+
+We defined several metrics to compare a prediction to a reference trajectory (Supplementary Note 1).
+Based on an analysis of their robustness and conformity to a set of rules (Supplementary Note 1), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures).
+The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance.
+These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table 2).
+Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}.
+All other real datasets were classified as ‘silver standard’.
+For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter2011}.
+For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.
+
+We found that method performance was very variable across datasets, indicating that there is no ‘one-size-fits-all’ method that works well on every dataset (Figure \ref{fig:supfigure_3}a).
+Even methods that can detect most of the trajectory types, such as PAGA, RaceID/StemID and SLICER were not the best methods across all trajectory types (Figure \ref{fig:figure_3}b).
+The overall score between the different dataset sources was moderately to highly correlated (Spearman rank correlation between 0.5–0.9) with the scores on real datasets containing a gold standard (Figure \ref{fig:supfigure_3}b), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data.
+On the other hand, the different metrics frequently disagreed with each other, with Monocle and PAGA Tree scoring better on the topology scores, whereas other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (Figure \ref{fig:figure_3}b).
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_3.pdf}
+	Figure caption on next page $\rightarrow$
+\end{figure}
+
+\clearpage % insert a page break
+\captionof{figure}{
+	\textbf{Accuracy of trajectory inference methods.}
+	\textbf{a} Overall score for all methods across 339 datasets, colored by the source of the datasets.
+Black line indicates the mean.
+\textbf{b} Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard, across all methods (n = 46, after filtering out methods that errored too frequently).
+Shown in the top left is the Pearson correlation.
+\textbf{c} Bias in the overall score towards trajectory types for all methods across 339 datasets.
+Black line indicates the mean.
+\textbf{d} Distributions of the difference in size between predicted and reference topologies.
+A positive difference means that the topology predicted by the method is more complex than the one in the reference.
+}
+\label{fig:supfigure_3}
+
+The performance of a method was strongly dependent on the type of trajectory present in the data (Figure \ref{fig:figure_3}b).
+Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories (Figure \ref{fig:supfigure_3}c).
+This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, whereas those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (Figure \ref{fig:supfigure_3}d).
+This analysis therefore indicates that detecting the right topology is still a difficult task for most of these methods, because methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.
+
+The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods.
+To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods.
+A top model in this case was defined as a model with an overall score of at least 95\% as the best model.
+On all datasets, using one method resulted in getting a top model about 27\% of the time.
+This increased up to 74\% with the addition of six other methods (Figure \ref{fig:figure_4}a).
+The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA.
+We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (Figure \ref{fig:figure_4}b), although in these cases less methods were necessary to obtain at least one top model for a given dataset.
+Altogether, this shows that there is considerable complementarity between the different methods and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori.
+Moreover, it also opens up the possibilities for new ensemble methods that utilize this complementarity.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_4.pdf}
+	\caption{
+		\textbf{Complementarity between different trajectory inference methods.}
+		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95\% of the best model) when applied to all datasets.
+\textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types.
+For this figure, we did not include any methods requiring a cell grouping or a time course as prior information.
+	}
+	\label{fig:figure_4}
+\end{figure}
+
+\subsection{Scalability}
+
+While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million \cite{svensson_exponentialscalingsinglecell_2018}.
+Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features \cite{cao_jointprofilingchromatin_2018}.
+To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets.
+We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a).
+As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Supplementary Fig.
+5), and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig.
+2a).
+
+\begin{figure}
+	\centering\includegraphics[width=0.9\linewidth]{fig/dynbenchmark/supfigure_4.pdf}
+	\caption{
+		\textbf{Scalability of trajectory inference methods.}
+		\textbf{a} Three examples of average observed running times across five datasets (left) and the predicted running time (right).
+\textbf{b} Overview of the scalability results of all methods, ordered by their average predicted running time from (a).
+We predicted execution times and memory usage for each method with increasing number of features or cells, and used these values to classify each method into sublinear, linear, quadratic and superquadratic based on the shape of the curve.
+	}
+	\label{fig:supfigure_4}
+\end{figure}
+
+We found that the scalability of most methods was overall very poor, with most graph and tree methods not finishing within an hour on a dataset with ten thousand cells and ten thousand features (Figure \ref{fig:figure_3}c), which is around the size of a typical droplet-based single-cell dataset \cite{svensson_exponentialscalingsinglecell_2018}.
+Running times increased further with increasing number of cells, with only a handful of graph/tree methods completing within a day on a million cells (PAGA, PAGA Tree, Monocle DDRTree, Stemnet and GrandPrix).
+Some methods, such as Monocle DDRTree and GrandPrix, also suffered from unsatisfactory running times when given a high number of features.
+
+Methods with a low running time typically had two defining aspects: they had a linear time complexity with respect to the features and/or cells, and adding new cells or features led to a relatively low increase in time (Figure \ref{fig:supfigure_4}b).
+We found that more than half of all methods had a quadratic or superquadratic complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (Figure \ref{fig:supfigure_4}b).
+
+We also assessed the memory requirements of each method (Supplementary Fig.
+2c).
+Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features.
+Notably, the memory requirements were very high for several methods on datasets with high numbers of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).
+
+Altogether, the scalability analysis indicated that the dimensions of the data are an important factor in the choice of method, and that method development should pay more attention to maintaining reasonable running times and memory usage.
+
+\subsection{Stability}
+
+It is not only important that a method is able to infer an accurate model in a reasonable time frame, but also that it produces a similar model when given very similar input data.
+To test the stability of each method, we executed each method on ten different subsamples of the datasets (95\% of the cells, 95\% of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (Figure \ref{fig:figure_3}d).
+
+Given that the trajectories of methods that fix the topology either algorithmically or through a parameter are already very constrained, it is to be expected that such methods tend to generate very stable results.
+Nonetheless, some fixed topology methods still produced slightly more stable results, such as SCORPIUS and MATCHER for linear methods and MFA for multifurcating methods.
+Stability was much more diverse among methods with a free topology.
+Slingshot produced more stable models than PAGA (Tree), which in turn produced more stable results than pCreode and Monocle DDRTree.
+
+\subsection{Usability}
+
+While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_ten_2017}.
+We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}).
+It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB).
+The reasons for not including certain tools are all discussed on our repository (https://github.com/dynverse/dynmethods/issues?q=label:unwrappable).
+Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.
+
+We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Supplementary Fig.
+6).
+While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Supplementary Fig.
+6 right) and we believe that these should receive extra attention from developers.
+Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_best_2014,artaza_top10metrics_2016}.
+Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods.
+We observed no clear relation between usability and method accuracy or usability (Figure \ref{fig:figure_2}b).
+
+\begin{table}
+	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}
+
+	\centering\begingroup\fontsize{7}{9}\selectfont
+
+	\begin{tabular}{p{2.5cm}p{7.5cm}p{3cm}}
+		\toprule
+		Aspect & Items & References \\ \midrule
+
+		\multicolumn{3}{l}{\textbf{Availability}}\\*
+
+		Open source & (1) Method's code is freely available (2) The code can be run on a freely available platform & \cite{l_rpackages_2017,wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
+
+		Version control & The code is available on a public version controlled repository, such as Github & \cite{l_rpackages_2017,wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
+
+		Packaging & (1) The code is provided as a "package", exposing functionality through functions or shell commands (2) The code can be easily installed through a repository such as CRAN, Bioconductor, PyPI, CPAN, debian packages, … & \cite{l_rpackages_2017,wickham_packagesorganizetest_2015,jimenez_foursimplerecommendations_2017,silva_generalguidelinesbiomedical_2017}\\*
+
+		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*
+
+		License & (1) The code is licensed (2) License allows academic use & \cite{l_rpackages_2017,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
+
+		Interface & (1) The tool can be run using a graphical user interface, either locally or on a web server (2) The tool can be run through the command line or through a programming language & \cite{silva_generalguidelinesbiomedical_2017}\\
+
+		\midrule
+		\multicolumn{3}{l}{\textbf{Code quality}} \\*
+
+		Function and object naming & (1) Functions/commands have well chosen names (2) Arguments/parameters have well chosen names & \cite{wilson_best_2014,wickham_packagesorganizetest_2015}\\*
+
+		Code style & (1) Code has a consistent style (2) Code follows (basic) good practices in the programming language of choice, for example PEP8 or the tidyverse style guide & \cite{wilson_best_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016}\\*
+
+		Code duplication & Duplicated code is minimal & \cite{wilson_best_2014,wickham_packagesorganizetest_2015}\\
+
+		Self-contained functions & The method is exposed to the user as self-contained functions or commands & \cite{anderson_writing_2016,taschuk_ten_2017,silva_generalguidelinesbiomedical_2017}\\*
+
+		Plotting & Plotting functions are provided for the final and/or intermediate results & \\*
+
+		Dummy proofing & Package contains dummy proofing, i.e.
+testing whether the parameters and data supplied by the user make sense and are useful & \cite{l_rpackages_2017,karimzadeh_topconsiderationscreating_}\\
+
+		\midrule
+		\multicolumn{3}{l}{\textbf{Code assurance}} \\*
+
+		Unit testing & Method is tested using unit tests & \cite{l_rpackages_2017,wilson_best_2014,anderson_writing_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
+
+		Unit testing & Tests are run automatically using functionality from the programming language & \cite{l_rpackages_2017,wilson_best_2014,anderson_writing_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
+
+		Continuous integration & The method uses continuous integration, for example on Travis CI & \cite{beaulieu-jones_reproducibility_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
+
+		Code coverage & (1) The code coverage of the repository is assessed. (2) What is the percentage of code coverage & \\
+
+		\midrule
+		\multicolumn{3}{l}{\textbf{Documentation}} \\*
+
+		Support & (1) There is a support ticket system, for example on Github (2) The authors respond to tickets and issues are resolved within a reasonable time frame & \cite{wilson_best_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
+
+		Development model & (1) The repository separates the development code from master code, for example using git master en developer branches (2) The repository has created releases, or several branches corresponding to major releases.
+(3) The repository has branches for the development of separate features.
+& \cite{driessen_successfulgitbranching_2010}\\*
+
+		Tutorial & (1) A tutorial or vignette is available (2) The tutorial has example results (3) The tutorial has real example data (4) The tutorial showcases the method on several datasets (1=0, 2=0.5, >2=1) & \cite{wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017,karimzadeh_topconsiderationscreating_,boulesteix_tensimplerules_2015}\\*
+
+		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
+
+		Inline documentation & Inline documentation is present in the code & \cite{wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
+
+		Parameter transparency & All important parameters are exposed to the user & \cite{taschuk_ten_2017}\\
+
+		\midrule
+		\multicolumn{3}{l}{\textbf{Behaviour}} \\*
+
+		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{jfp_greendiceare_2016}\\*
+
+		Unexpected output & (1) No unexpected output messages are generated by the method (2) No unexpected files, folders or plots are generated (3) No unexpected warnings during runtime or compilation are generated & \cite{artaza_top10metrics_2016}\\*
+
+		Trajectory format & The postprocessing necessary to extract the relevant output from the method is minimal (1), moderate (0.5) or extensive (0) & \\*
+
+		Prior information & Prior information is required (0), optional (1) or not required (1) & \\*
+
+		\midrule
+		\multicolumn{3}{l}{\textbf{Paper}} \\*
+
+		Publishing & The method is published & \\*
+
+		Peer review & The paper is published in a peer-reviewed journal & \cite{karimzadeh_topconsiderationscreating_,gannon_essentialrolepeer_2001,baldwin_refereeswetrust_2017}\\*
+
+		Evaluation on real data & (1) The paper shows the method's usefulness on several (1), one (0.25) or no real datasets.
+(2) The paper quantifies the accuracy of the method given a gold or silver standard trajectory & \cite{aniba_issuesbioinformaticsbenchmarking_2010,jelizarow_overoptimismbioinformaticsillustration_2010}\\*
+
+		Evaluation of robustness & The paper assessed method robustness (to eg.
+noise, subsampling, parameter changes, stability) in one (0.5) or several (1) ways & \cite{karimzadeh_topconsiderationscreating_,aniba_issuesbioinformaticsbenchmarking_2010,boulesteix_tensimplerules_2015,jelizarow_overoptimismbioinformaticsillustration_2010}\\
+
+		\bottomrule
+	\end{tabular}
+	\endgroup{}
+\end{table}
+
+\section{Discussion}
+
+In this study, we presented a large-scale evaluation of the performance of 45 TI methods.
+By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the accuracy of the methods on more than 200 datasets.
+We also assessed several other important quality measures, such as the quality of the method’s implementation, the scalability to hundreds of thousands of cells and the stability of the output on small variations of the datasets.
+
+Based on the results of our benchmark, we propose a set of practical guidelines for method users (Figure \ref{fig:figure_5} and guidelines.dynverse.org).
+We postulate that, as a method’s performance is heavily dependent on the trajectory type being studied, the choice of method should currently be primarily driven by the anticipated trajectory topology in the data.
+For most use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
+In each of these use cases, our evaluation suggests a different set of optimal methods, as shown in Figure \ref{fig:figure_5}.
+Several other factors will also impact the choice of methods, such as the dimensions of the dataset and the prior information that is available.
+These factors and several others can all be dynamically explored in our interactive app (guidelines.dynverse.org).
+This app can also be used to query the results of this evaluation, such as filtering the datasets or changing the importance of the evaluation metrics for the final ranking.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_5.pdf}
+	\caption{
+		\textbf{Practical guidelines for method users.}
+		As the performance of a method mostly depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user’s existing knowledge about the expected topology in the data.
+We therefore devised a set of practical guidelines, which combines the method’s performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
+Methods to the right are ranked according to their performance on a particular (set of) trajectory type.
+Further to the right are shown the accuracy (+: scaled performance $\geq$ 0.9, $\pm$: >0.6), usability scores (+:$\geq$0.9, $\pm$ $\geq$0.6), estimated running times and required prior information.
+k, thousands; m, millions.
+	}
+	\label{fig:figure_5}
+\end{figure}
+
+When inferring a trajectory on a dataset of interest, it is important to take two further points into account.
+First, it is critical that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
+This is to make sure that the prediction is not biased due to the given parameter setting or the particular algorithm underlying a TI method.
+The value of using different methods is further supported by our analysis indicating substantial complementarity between the different methods.
+Second, even if the expected topology is known, it can be beneficial to also try out methods that make less assumptions about the trajectory topology.
+When the expected topology is confirmed using such a method, it provides additional evidence to the user.
+When a more complex topology is produced, this could indicate that the underlying biology is much more complex than anticipated by the user.
+
+Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{Aibar2017a} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}.
+Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}).
+Using our framework, this figure can be recreated using only a couple of lines of R code (https://methods.dynverse.org).
+In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses.
+In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_6.pdf}
+	\caption{
+		\textbf{Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.}
+		Trajectories inferred by each method were projected to a common dimensionality reduction using multidimensional scaling.
+For each dataset, we also calculated a ‘consensus’ prediction, by calculating the cordist between each pair of models and picking the model with the highest score on average.
+\textbf{a}, The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
+\textbf{b}, The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts.
+\textbf{c}, A synthetic dataset generated by dyntoy, containing four disconnected trajectories.
+\textbf{d}, A synthetic dataset generated by dyngen, containing a cyclic trajectory.
+	}
+	\label{fig:figure_6}
+\end{figure}
+
+Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories (Figure \ref{fig:figure_6}a,b).
+However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analyzing single-cell omics datasets with complex trajectories.
+Foremost, new methods should focus on improving the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be identified using a dimensionality reduction method (Figure \ref{fig:figure_6}c,d).
+Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell omics field.
+Finally, new tools should be designed to scale well with the increasing number of cells and features.
+We found that only a handful of current methods can handle datasets with more than 10,000 cells within a reasonable time frame.
+To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method on the different measures proposed in this study at \href{https://benchmark.dynverse.org}{https://benchmark.dynverse.org}.
+
+We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
+However, ‘good-yet-not-the-best’ methods \cite{norel_selfassessmenttrap_2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases.
+This is also supported by our analysis of method complementarity.
+Some examples for the latter include PhenoPath, which can include additional covariates in its model, ouija, which returns a measure of uncertainty of each cell’s position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
+
+\section{Methods}
+
+\subsection{Trajectory inference methods}
+
+We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecelllistsoftware_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellpseudotimeoverviewalgorithms_2018}.
+We welcome any contributions by creating an issue at https://methods.dynverse.org.
+
+Methods were excluded from the evaluation based on several criteria: (1) not freely available; (2) no code available; (3) superseded by another method; (4) requires data types other than expression; (5) no programming interface; (6) unresolved errors during wrapping; (7) too slow (requires more than 1 h on a 100 $\times$ 100 dataset); (8) does not return an ordering; and (9) requires additional user input during the algorithm (other than prior information).
+The discussions on why these methods were excluded can be found at \href{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}.
+In the end, we included 45 methods in the evaluation.
+
+\subsection{Method wrappers}
+
+To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}).
+These containers are automatically built and tested using Travis continuous integration (\href{https://travis-ci.org/dynverse}{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity.
+For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts).
+This script reads in the input data, runs the method and outputs the files required to construct a trajectory.
+We also created a script to generate an example dataset, which is used for automated testing.
+
+We used the Github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores.
+About one-third of the authors responded and we improved the wrappers based on their feedback.
+These discussions can be viewed on Github: \href{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}{https://github.com/dynverse/dynmethods/issues?q=label:method\_discussion}
+
+\subsubsection{Method input}
+
+As input, we provided each method with either the raw count data (after cell and gene filtering) or normalized expression values, based on the description in the method documentation or from the study describing the method.
+A large portion of the methods requires some form of prior information (for example, a start cell) to be executable.
+Other methods optionally allow the exploitation of certain prior information.
+Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes or even a grouping of cells into cell states.
+Providing prior information to a TI method can be both a blessing and a curse.
+In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
+On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
+Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well-studied systems.
+
+The prior information was extracted from the reference trajectory as follows:
+
+\begin{itemize}
+	\item \textbf{Start cells}: the identity of one or more start cells.
+For both real and synthetic data, a cell was chosen that was the closest (in geodesic distance) to each milestone with only outgoing edges.
+For ties, one random cell was chosen.
+For cyclic datasets, a random cell was chosen.
+	\item \textbf{End cells}: the identity of one or more end cells. This is similar to the start cells, but now for every state with only incoming edges.
+	\item \textbf{No. of end states}: number of terminal states, i.e., the number of milestones with only incoming edges.
+	\item \textbf{Grouping}: for each cell a label showing which state/cluster/branch it belongs to.
+For real data, the states were from the gold/silver standard.
+For synthetic data, each milestone was seen as one group and cells were assigned to their closest milestone.
+	\item \textbf{No.
+of branches}: number of branches/intermediate states.
+For real data, this was the number of states in the gold/silver standard.
+For synthetic data, this was the number of milestones.
+	\item \textbf{Discrete time course}: for each cell a time point from which it was sampled.
+If available, this was directly extracted from the reference trajectory; otherwise the geodesic distance from the root milestone was used.
+For synthetic data, the simulation time was uniformily discretized into four timepoints.
+	\item \textbf{Continuous time course}: for each cell a time point from which it was sampled.
+For real data, this was equal to the discrete time course.
+For synthetic data, we used the internal simulation time of each simulator.
+\end{itemize}
+
+\subsubsection{Common trajectory model}
+
+Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps.
+We therefore post-processed the output of each method into a common probabilistic trajectory model (Figure \ref{fig:supfigure_1}a).
+This model consisted of three parts.
+(1) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
+(2) The milestone percentages contain, for each cell, its position between milestones and sums for each cell to one.
+(3) The regions of delayed commitment define connections between three or more milestones.
+These must be explicitly defined in the trajectory model and per region one milestone must be directly connected to all other milestones of the region.
+
+Depending on the output of a method, we used different strategies to convert the output to our model (Figure \ref{fig:supfigure_1}b).
+Special conversions are denoted by an asterisk and will be explained in more detail in the second list below.
+
+\begin{itemize}
+	\item \textbf{Type 1, direct}: CALISTA*, DPT*, ElPiGraph, ElPiGraph cycle, ElPiGraph linear, MERLoT, PAGA, SLICE*, Slingshot, URD* and Wishbone.
+The wrapped method directly returned a network of milestones, the regions of delayed commitment and for each cell it is given to what extent it belongs to a milestone.
+In some cases, this indicates that additional transformations were required for the method, not covered by any of the following output formats.
+Some methods returned a branch network instead of a milestone network and this network was converted by calculating the line graph of the branch network.
+	\item \textbf{Type 2, linear pseudotime}: Component 1, Embeddr, FORKS, MATCHER, ouija, ouijaflow, PhenoPath, pseudogp, SCIMITAR, SCORPIUS, topslam, TSCAN, Wanderlust and Waterfall.
+The method returned a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
+	\item \textbf{Type 3, cyclical pseudotime}: Angle and reCAT.
+The method returned a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones.
+These milestones were positioned at pseudotime 0, 1/3 and 2/3.
+	\item \textbf{Type 4, end state probability}: FateID, GPfates, GrandPrix, MFA*, SCOUP and STEMNET.
+The method returned a pseudotime and for each cell and end state a probability (Pr) for how likely a cell will end up in a certain end state.
+This was translated into a star-shaped milestone network, with one starting milestone (M0) and several outer milestones (Mi), with regions of delayed commitment between all milestones.
+The milestone percentage of a cell to one of the outer milestones was equal to pseudotime$\times$PrMi.
+The milestone percentage to the starting milestone was equal to 1 $-$ pseudotime.
+	\item \textbf{Type 5, cluster assignment}: Mpath and SCUBA.
+The method returned a milestone network and an assignment of each cell to a specific milestone.
+Cells were positioned onto the milestones they are assigned to, with milestone percentage equal to 1.
+	\item \textbf{Type 6, orthogonal projection}: MST, pCreode and RaceID/StemID.
+The method returned a milestone network, and a dimensionality reduction of the cells and milestones.
+The cells were projected to the closest nearest segment, thus determining the cells’ position along the milestone network.
+If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell.
+For these methods, we usually wrote two wrappers, one which included the projection and one without.
+	\item\textbf{ Type 7, cell graph}: CellRouter, CellTrails, cellTree Gibbs, cellTree maptpx, cellTree VEM, Monocle DDRTree, Monocle ICA, Sincell* and SLICER.
+The method returned a network of cells and which cell–cell transitions were part of the ‘backbone’ structure.
+Backbone cells with degree $\neq$ 2 were regarded as milestones and all other cells were placed on transitions between the milestones.
+If a method did not return a distance between pairs of cells, the cells were uniformly positioned between the two milestones.
+Otherwise, we first calculated the distance between two milestones as the sum of the distances between the cells and then divided the distance of each pair of cells with the total distance to get the milestone percentages.
+\end{itemize}
+
+Special conversions were necessary for certain methods:
+
+\begin{itemize}
+	\item \textbf{CALISTA}: We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest.
+The cluster probabilities of the two selected milestones were then used as milestone percentages.
+This was then processed as a type 1, direct, method.
+	\item \textbf{DPT}: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contained the cells that were assigned to the ‘unknown’ branch) and three terminal milestones, each corresponding to a tip point.
+This was then processed as a type 1, direct, method.
+	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25\%.
+This was then processed as a type 7, cell graph, method.
+	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\href{https://research.cchmc.org/pbge/slice.html}{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE.
+This was then processed as a type 1, direct, method.
+	\item \textbf{MFA}: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as a type 4, end state probabilities, method.
+	\item \textbf{URD}: We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.
+
+\end{itemize}
+
+More information on how each method was wrapped can be found within the comments of each wrapper script, listed at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.
+
+\subsubsection{Off-the-shelf methods}
+
+For baseline performance, we added several ‘off-the-shelf’ TI methods that can be run using a few lines of code in R.
+
+\begin{itemize}
+
+	\item \textbf{Component 1}: This method returns the first component of a principal component analysis (PCA) dimensionality reduction as a linear trajectory.
+This method is especially relevant as it has been used in a few studies already \cite{kouno_temporaldynamicstranscriptional_2013,zeng_pseudotemporalorderingsingle_2017}.
+
+	\item \textbf{Angle}: Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA and uses this angle as a pseudotime for generating a cyclical trajectory.
+
+	\item \textbf{MST}: This method performs PCA dimensionality reduction, followed by clustering using the R mclust package, after which the clusters are connected using a minimum spanning tree.
+The trees are orthogonally projected to the nearest segment of the tree.
+This baseline is highly relevant as many methods follow the same methodology: dimensionality reduction, clustering, topology inference and project cells to topology.
+
+\end{itemize}
+
+\subsection{Trajectory types}
+
+We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (Figure \ref{fig:figure_1}c).
+These trajectory types start from the most general trajectory type, a disconnected graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types are reached: linear and cyclical.
+A disconnected graph is a graph in which only one edge can exist between two nodes.
+A (connected) graph is a disconnected graph in which all nodes are connected.
+An acyclic graph is a graph containing no cycles.
+A tree is an acyclic graph containing no convergences (no nodes with in-degree higher than 1).
+A convergence is an acyclic graph in which only one node has a degree larger than 1 and this same node has an in-degree of 1.
+A multifurcation is a tree in which only one node has a degree larger than 1.
+A bifurcation is a multifurcation in which only one node has a degree equal to 3.
+A linear topology is a graph in which no node has a degree larger than 3.
+Finally, a cycle is a connected graph in which every node has a degree equal to 2.
+In most cases, a method that was able to detect a complex trajectory type was also able to detect less complex trajectory types, with some exceptions shown in Figure \ref{fig:figure_2}a.
+
+For simplicity, we merged the bifurcation and convergence trajectory type, and the acyclic graph and connected graph trajectory type in the main figures of the paper.
+
+\subsection{Real datasets}
+
+We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table 2).
+The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}).
+Whenever possible, we preferred to start from the raw counts data.
+These raw counts were all normalized and filtered using a common pipeline, as discussed later.
+Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
+In the end, we included 110 datasets for this evaluation.
+
+For each dataset, we extracted a reference trajectory, consisting of two parts: the cellular grouping (milestones) and the connections between these groups (milestone network).
+The cellular grouping was provided by the authors of the original study, and we classified it as a gold standard when it was created independently from the expression matrix (such as from cell sorting, the origin of the sample, the time it was sampled or cellular mixing) or as a silver standard otherwise (usually by clustering the expression values).
+To connect these cell groups, we used the original study to determine the network that the authors validated or otherwise found to be the most likely.
+In the end, each group of cells was placed on a milestone, having a percentage of 1 for that particular milestone.
+The known connections between these groups were used to construct the milestone network.
+If there was biological or experimental time data available, we used this as the length of the edge; otherwise we set all the lengths equal to one.
+
+\subsection{Synthetic datasets}
+
+To generate synthetic datasets, we used four different synthetic data simulators:
+
+\begin{itemize}
+	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
+	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
+	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulosPROSSTTProbabilisticSimulation2018}
+	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
+
+\end{itemize}
+
+For every simulator, we took great care to make the datasets as realistic as possible.
+To do this, we extracted several parameters from all real datasets.
+We calculated the number of differentially expressed features within a trajectory using a two-way Mann–Whitney U test between every pair of cell groups.
+These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5\% of cells, and had at least a fold-change of 2.
+We also calculated several other parameters, such as drop-out rates and library sizes using the Splatter package \cite{zappia_splattersimulationsinglecell_2017}.
+These parameters were then given to the simulators when applicable, as described for each simulator below.
+Not every real dataset was selected to serve as a reference for a synthetic dataset.
+Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (with $k = 10$, implemented in the R cluster package) to generate synthetic data.
+
+\subsubsection{dyngen}
+
+The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdom_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
+We simulated a total of 110 datasets, with 11 different topologies.
+
+\hfill\break
+\textit{Network generation}
+\hfill\break
+
+One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
+The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
+One driver of bifurcation seems to be mutual antagonism, where genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
+Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
+Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
+
+To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
+Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
+The structure of these module networks is available at \href{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}.
+
+From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added.
+For each dataset, we used the same number of genes as were differentially expressed in the real datasets.
+5\% of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module).
+We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module.
+To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits \cite{marbach_tissuespecificregulatorycircuits_2016}), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the page\_rank function of the \textit{igraph} package.
+
+\hfill\break
+\textit{Simulation of gene regulatory systems using thermodynamic models}
+\hfill\break
+
+To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdom_2012}.
+In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):
+
+$$
+\begin{aligned}
+\label{eq:mrna_ode}
+\frac{dx_i}{dt} &= \underbrace{m \times f(y_1, y_2, ...)}_\text{production} - \underbrace{\lambda \times x_i}_\text{degradation}
+\end{aligned}
+$$
+$$
+\begin{aligned}
+\label{eq:prot_ode}
+\frac{dy_i}{dt} &= \underbrace{r \times x_i}_\text{production} - \underbrace{\Lambda \times y_i}_\text{degradation}
+\end{aligned}
+$$
+
+where $m$, $\lambda$, $r$ and $\Lambda$ represent production and degradation rates, the ratio of which determines the maximal gene and protein expression.
+The two types of equations are coupled because the production of protein $y_i$ depends on the amount of gene expression $x_i$, which in turn depends on the amount of other proteins through the activation function $f(y_1, y_2, ...)$.
+
+The activation function is inspired by a thermodynamic model of gene regulation, in which the promoter of a gene can be bound or unbound by a set of transcription factors, each representing a certain state of the promoter.
+Each state is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state.
+The production rate of the gene is calculated by combining the probabilities of the promoter being in each state with the relative activation:
+
+$$
+\begin{aligned}
+\label{eq:input_function_probabilities}
+f(y_1, y_2, ..., y_n) = \sum_{j \in \{0, 1, ..., n^2\}} \alpha_j \times P_j
+\end{aligned}
+$$
+
+The probability of being in a state is based on the thermodynamics of transcription factor binding. When only one transcription factor is bound in a state:
+$$
+\begin{aligned}
+P_j \propto \nu = \left(\frac{y}{k}\right)^{n}
+\end{aligned}
+$$
+
+where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
+$$
+\begin{aligned}
+P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
+\end{aligned}
+$$
+
+where $\rho$ represents the cooperativity of binding between the different transcription factors.
+
+$P_i$ is only proportional to $\nu$ because $\nu$ is normalized such that $\sum_{i} P_i = 1$.
+
+To each differential equation, we added an additional stochastic term: 
+$$
+\begin{aligned}
+\frac{dx_i}{dt} & = m \times f(y_1, y_2, ...) - \lambda \times x_i + \eta \times \sqrt{x_i} \times \Delta W_t \\
+\frac{dy_i}{dt} & = r \times x_i - \Lambda \times y_i + \eta \times \sqrt{y_i} \times \Delta W_t
+\end{aligned}
+$$
+
+with $\Delta W_t \sim \mathcal{N}(0, h)$.
+
+Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows.
+$e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.
+
+$$
+\begin{aligned}
+r & = \mathcal{U}(10, 200) \\
+d & = \mathcal{U}(2, 8) \\
+p & = \mathcal{U}(2, 8) \\
+q & = \mathcal{U}(1, 5) \\
+a_0 & = \begin{cases}1 & \text{if } |e| = 0 \\ 1 & \text{if } \forall x \in e \text{, } x = -1 \\ 0 & \text{if } \forall x \in e \text{, } x = 1 \\ 0.5 & \text{otherwise}\end{cases} \\
+a_i & = \begin{cases}0 & \text{if } \exists x \in e_i \text{, } x = -1 \\ 1 & \text{otherwise}\end{cases} \\
+s & = \mathcal{U}(1, 20) \\
+k & = y_{max} / (2 * s) \text{,} \\
+\ & \quad \text{where } y_{max} = r / d \times p / q \\
+c & = \mathcal{U}(1, 4)
+\end{aligned}
+$$
+
+We converted each ODE to an SDE by adding a chemical Langevin equation, as described in \cite{schaffter_genenetweaversilicobenchmark_2011}.
+These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 8$.
+The total simulation time varied between 5 for linear and bifurcating datasets, 10 for consecutive bifurcating, trifurcating and converging datasets, 15 for bifurcating converging datasets and 30 for linear long, cycle and bifurcating loop datasets.
+The burn-in period was for each simulation 2.
+Each network was simulated 32 times.
+
+\hfill\break
+\textit{Simulation of the single-cell RNA-seq experiment}
+\hfill\break
+
+For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in.
+These cells were sampled uniformly across the different steps of the 32 simulations.
+Next, we used the Splatter package \cite{zappia_splattersimulationsinglecell_2017} to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
+We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$.
+These were combined with the expression levels of the genes simulated within a trajectory.
+Next, true counts were simulated using $Y'_{i,j} \sim \text{Poisson}(\lambda_{i,j})$.
+Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
+Both $x_0$ (the midpoint for the dropout logistic function) and $k$ (the shape of the dropout logistic function) were estimated by Splatter.
+
+This count matrix was then filtered and normalised using the pipeline described below.
+
+\hfill\break
+\textit{Gold standard extraction}
+\hfill\break
+
+Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward.
+Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable.
+We therefore first constructed a backbone expression profile for each branch within the trajectory.
+To do this, we first defined in which order the expression of the modules is expected to change, and then generated a backbone expression profile in which the expression of these modules increases and decreases smoothly between 0 and 1.
+We also smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
+We used dynamic time warping, implemented in the dtw R package \cite{giorgino_computingvisualizingdynamic_2009,tormene_matchingincompletetime_2009}, with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
+In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.
+
+\subsubsection{dyntoy}
+
+For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (\url{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}) .
+We created 12 topology generators (described below), and with 10 datasets per generator, this lead to a total of 120 datasets.
+
+We created a set of topology generators, were $B(n, p)$ denotes a binomial distribution, and $U(a, b)$ denotes a uniform distribution:
+
+* Linear and cyclic, with number of milestones $\sim B(10, 0.25)$
+* Bifurcating and converging, with four milestones
+* Binary tree, with number of branching points $\sim U(3, 6)$
+* Tree, with number of branching points $\sim U(3, 6)$ and maximal degree $\sim U(3, 6)$
+
+For more complex topologies we first calculated a random number of "modifications" $\sim U(3, 6)$ and a $\textit{deg}_{\textit{max}} \sim B(10, 0.25) + 1$.
+For each type of topology, we defined what kind of modifications are possible: divergences, loops, convergences and divergence-convergence.
+We then iteratively constructed the topology by uniformly sampling from the set of possible modifications, and adding this modification to the existing topology.
+For a divergence, we connected an existing milestone to a number of a new milestones.
+Conversely, for a convergence we connected a number of new nodes to an existing node.
+For a loop, we connected two existing milestones with a number of milestones in between.
+Finally for a divergence-convergence we connected an existing milestone to several new milestones which again converged on a new milestone.
+The number of nodes was sampled from $\sim B(\textit{deg}_{\textit{max}} - 3, 0.25) + 2$
+
+* Looping, allowed loop modifications
+* Diverging-converging, allowed divergence and converging modifications
+* Diverging with loops, allowed divergence and loop modifications
+* Multiple looping, allowed looping modifications
+* Connected, allowed looping, divergence and convergence modifications
+* Disconnected, number of components sampled from $\sim B(5, 0.25) + 2$, for each component we randomly chose a topology from the ones listed above
+
+After generating the topology, we sampled the length of each edge $\sim U(0.5, 1)$.
+We added regions of delayed commitment to a divergence in a random half of the cases.
+We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, based on the length of the edges in the milestone network.
+
+For each gene (same number as from the reference real dataset), we calculated the Kamada-Kawai layout in 2 dimensions, with edge weight equal to the length of the edge.
+For this gene, we then extracted for each cell a density value using a bivariate normal distribution with $\mu \sim U(x_{\textit{min}}, x_{\textit{min}})$ and $\sigma \sim U(x_{\textit{min}}/10, x_{\textit{min}}/8)$.
+We used this density as input for a zero-inflated negative binomial distribution with $\mu ~ U(100, 1000) \times \textit{density}$, $k ~ U(\mu / 10, \mu / 4)$ and $pi$ from the parameters of the reference real dataset, to get the final count values.
+
+This count matrix was then filtered and normalised using the pipeline described below.
+
+\subsubsection{PROSSTT}
+
+PROSSTT is a recent data simulator \cite{papadopoulosPROSSTTProbabilisticSimulation2018}, which simulates expression using linear mixtures of expression programs and random walks through the trajectory.
+We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets.
+However, due to frequent crashes of the tool, only 19 datasets created output and were thus used in the evaluation.
+
+Using the simulate\_lineage function, we simulated the lineage expression, with parameters $a \sim U(0.01, 0.1)$, $\textit{branch-tol}_{\textit{intra}} \sim U(0, 0.9)$ and $\textit{branch-tol}_{\textit{inter}} \sim U(0, 0.9)$.
+These parameter distributions were chosen very broad so as to make sure both easy and difficult datasets are simulated.
+After simulating base gene expression with simulate\_base\_gene\_exp, we used the sample\_density function to finally simulate expression values of a number of cells (the same as from the reference real dataset), with $\alpha \sim \textit{Lognormal}$ ($\mu = 0.3$ and $\sigma = 1.5$) and $\beta \sim \textit{Lognormal}$ ($\mu = 2$ and $\sigma = 1.5$).
+Each of these parameters were centered around the default values of PROSSTT, but with enough variability to ensure a varied set of datasets.
+
+This count matrix was then filtered and normalised using the pipeline described below.
+
+\subsubsection{Splatter}
+
+Splatter \cite{zappia_splattersimulationsinglecell_2017} simulates expression values by constructing non-linear paths between different states, each having a distinct expression profile.
+We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets, leading to a total of 50 datasets.
+
+We used the splatSimulatePaths function from Splatter to simulate datasets, with number of cells and genes equal to those in the reference real dataset, and with parameters  $\textit{nonlinearProb}$, $\textit{sigmaFac}$ and $\textit{skew}$ all sampled from $U(0, 1)$.
+
+\subsection{Dataset filtering and normalization}
+
+We used a standard single-cell RNA-seq preprocessing pipeline that applies parts of the scran and scater Bioconductor packages \cite{lun_stepbystepworkflowlowlevel_2016}.
+The advantages of this pipeline are that it works both with and without spike-ins, and it includes a harsh cell filtering that looks at abnormalities in library sizes, mitochondrial gene expression and the number of genes expressed using median absolute deviations (which we set to 3).
+We required that a gene was expressed in at least 5\% of the cells and that it should have an average expression higher than 0.02.
+Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5\% and a biological component higher than 0.5.
+As a final filter, we removed both all-zero genes and cells until convergence.
+
+\subsection{Benchmark metrics}
+
+The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{norel_selfassessmenttrap_2011}.
+Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer.
+We therefore defined several possible metrics for comparing trajectories, each investigating different layers.
+These are all discussed in Supplementary Note 1 along with examples and robustness analyses when appropriate.
+
+Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (Supplementary Note 1).
+Based on this analysis, we chose four metrics for the evaluation, each assessing a different aspect of the trajectory: (1) the HIM measures the topological similarity; (2) the F1branches compares the branch assignment; (3) the cordist assesses the similarity in pairwise cell–cell distances and thus the cellular positions; and (4) the wcorfeatures looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
+
+\subsubsection{The Hamming–Ipsen–Mikhailov metric}
+
+The HIM metric \cite{jurman_HIM_Glocal_Metric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length).
+It is a linear combination of the normalized Hamming distance, which gives an indication of the differences in edge lengths, and the normalized Ipsen–Mikhailov distance, which assesses the similarity in degree distributions.
+The latter has a parameter $\gamma$, which was fixed at 0.1 to make the scores comparable between datasets.
+We illustrate the metric and discuss alternatives in Supplementary Note 1.
+
+\subsubsection{The F1 between branch assignments}
+
+To compare branch assignment, we used an F1 score, also used used for comparing biclustering methods \cite{saelens_comprehensiveevaluationmodule_2018}.
+To calculate this metric, we first calculated the similarity of all pairs of branches between the two trajectories using the Jaccard similarity.
+Next, we defined the ‘Recovery' (respectively ‘Relevance') as the average maximal similarity of all branches in the reference dataset (respectively prediction).
+The F1branches was then defined as the harmonic mean between Recovery and Relevance.
+We illustrate this metric further in Supplementary Note 1.
+
+\subsubsection{Correlation between geodesic distances}
+
+When the position of a cell is the same in both the reference and the prediction, its relative distances to all other cells in the trajectory should also be the same.
+This observation is the basis for the cordist metric.
+To calculate the cordist, we first sampled 100 waypoint cells in both the prediction and the reference dataset, using stratified sampling between the different milestones, edges and regions of delayed commitment, weighted by the number of cells in each collection.
+We then calculated the geodesic distances between the union of waypoint cells from both datasets and all other cells.
+The calculation of the geodesic distance depended on the location of the two cells within the trajectory, further discussed in Supplementary Note 1, and was weighted by the length of the edge in the milestone network.
+Finally, the cordist was defined as the Spearman rank correlation between the distances of both datasets.
+We illustrate the metric and assess the effect of the number of waypoint cells in Supplementary Note 1.
+
+\subsubsection{The correlation between important features}
+
+The wcorfeatures assesses whether the same differentially expressed features are found using the predicted trajectory as in the known trajectory.
+To calculate this metric, we used Random Forest regression (implemented in the R ranger package \cite{wright_rangerfastimplementation_2017}), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone.
+We then extracted feature importance values for each feature and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset to give more weight to large differences.
+As hyperparameters we set the number of trees to 10,000 and the number of features on which to split to 1\% of all available features.
+We illustrate this metric and assess the effect of its hyperparameters in Supplementary Note 1.
+
+\subsubsection{Score aggregation}
+
+To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics. This aggregation strategy is explained in more detail in Supplementary Note 1.
+
+To ensure that easy and difficult datasets have equal influence on the final score, we first normalized the scores on each dataset across the different methods.
+We shifted and scaled the scores to $\sigma$ = 1 and $\mu$ = 0, and then applied the unit probability density function of a normal distribution on these values to get the scores back into the [0,1] range.
+
+Since there is a bias in dataset source and trajectory type (for example, there are many more linear datasets), we aggregated the scores per method and dataset in multiple steps.
+We first aggregated the datasets with the same dataset source and trajectory type using an arithmetic mean of their scores.
+Next, the scores were averaged over different dataset sources, using an arithmetic mean that was weighted based on how much the synthetic and silver scores correlated with the real gold scores.
+Finally, the scores were aggregated over the different trajectory types again using an arithmetic mean.
+
+Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.
+
+\subsection{Method execution}
+
+Each execution of a method on a dataset was performed in a separate task as part of a gridengine job.
+Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 at 2.40 GHz, and one R session was started for each task.
+During the execution of a method on a dataset, if the time limit (>1 h) or memory limit (16 GB) was exceeded, or an error was produced, a zero score was returned for that execution.
+
+\subsection{Complementarity}
+
+To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score was equal to or higher than 95\% of the best overall score for that particular dataset.
+We then calculated for every method the weighted percentage of datasets that fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method.
+We iteratively added new methods until all methods were selected.
+For this analysis, we did not include any methods that require any strong prior information and only included methods that could detect the trajectory types present in at least one of the datasets.
+
+\subsection{Scalability}
+
+To assess the scalability of each method, we started from five real datasets, selected using the centers from a k-medoids as discussed before.
+We up- and downscaled these datasets between 10 and 100,000 cells and 10 and 100,000 features, while never going higher than 1,000,000 values in total.
+To generate new cells or features, we first generated a 10-nearest-neighbor graph of both the cells and features from the expression space.
+For every new cell or feature, we used a linear combination of one to three existing cells or features, where each cell or feature was given a weight sampled from a uniform distribution between 0 and 1.
+
+We ran each method on each dataset for maximally 1 h and gave each process 10 GB of memory.
+To determine the running time of each method, we started the timer right after data loading and the loading of any packages, and stopped the clock before postprocessing and saving of the output.
+Pre- and postprocessing steps specific to a method, such as dimensionality reduction and gene filtering, were included in the time.
+To estimate the maximal memory usage, we used the max\_vmem value from the qacct command provided by a gridengine cluster.
+We acknowledge, however, that these memory estimates are very noisy and the averages provided in this study are therefore only rough estimates.
+
+The relationship between the dimensions of a dataset and the running time or maximal memory usage was modeled using shape constrained additive models \cite{pya_shapeconstrainedadditive_2015}, with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the scam function as implemented in the R scam package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
+
+To classify the time complexity of each method with respect to the number of cells, we predicted the running time at 10,000 features with increasing number of cells from 100 to 100,000, with steps of 100.
+We trained a generalized linear model with the following function: $y \approxeq \log{x} + \sqrt{x} + x + x^2 + x^3$ with y as running time and x as the number of cells or features.
+The time complexity of a method was then classified using the weights w from this model:
+
+$$
+\begin{cases}
+\text{superquadratic} & \text{if } w_{x^3} > 0.25,\\
+\text{quadratic} & \text{if } w_{x^2} > 0.25,\\
+\text{linear} & \text{if } w_x > 0.25,\\
+\text{sublinear} & \text{if } w_{\text{log}(x)} > 0.25 \text{ or } w_{\text{sqrt}(x)} > 0.25,\\
+\text{case with highest weight} & \text{else.}
+\end{cases}
+$$
+
+This process was repeated for classifying the time complexity with respect to the number of features, and the memory complexity both with respect to the number of cells and features.
+
+\subsection{Stability}
+
+In the ideal case, a method should produce a similar trajectory, even when the input data is slightly different.
+However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R or the `random.seed` function in numpy).
+To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25\% of the datasets accounting for 15\% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
+We subsampled each dataset 10 times with 95\% of the original cells and 95\% of the original features.
+We ran every method on each of the bootstraps, and assessed the stability by calculating the benchmarking scores between each pair of subsequent models (run $i$ is compared to run $i+1$).
+For the cordist and F1branches, we only used the intersection between the cells of two datasets, while the intersection of the features was used for the wcorfeatures.
+
+\subsection{Usability}
+
+We created a transparent scoring scheme to quantify the usability of each method based on several existing tool quality and programming guidelines in the literature and online (Table \ref{tab:scoresheet}).
+The main goal of this quality control is to stimulate the improvement of current methods, and the development of user- and developer-friendly new methods.
+The quality control assessed six categories, each looking at several aspects, which are further divided into individual items.
+The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
+The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
+The code assurance category is frequently overlooked, and checks for code testing, continuous integration \cite{beaulieu-jones_reproducibility_2017} and an active support system.
+The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
+The behavior category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
+Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of datasets in which the usefulness of the method was shown and the scope of method evaluation in the paper.
+
+Each quality aspect received a weight depending on how frequently it was found in several papers and online sources that discuss tool quality (Table \ref{tab:scoresheet}).
+This was to make sure that more important aspects, such as the open source availability of the method, outweighed other less important aspects, such as the availability of a graphical user interface.
+For each aspect, we also assigned a weight to the individual questions being investigated (Table \ref{tab:scoresheet}).
+For calculating the final score, we weighed each of the six categories equally.
+
+\subsection{Guidelines}
+
+For each set of outcomes in the guidelines figure, we selected one to four methods, by first filtering the methods on those that can detect all required trajectory types, and ordering the methods according to their average accuracy score on datasets containing these trajectory types (aggregated according to the scheme presented in the section Accuracy).
+
+We used the same approach for selecting the best set of methods in the guidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}), developed using the R shiny package.
+This app will also filter the methods, among other things, depending on the predicted running time and memory requirements, the prior information available and the preferred execution environment (using the dynmethods package or standalone).
+
+\subsection{Reporting Summary}
+
+Further information on research design is available in the Nature Research Reporting Summary, available at \href{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}
+
+\section{Supplementary Note 1: Metrics to compare two trajectories}
+
+A trajectory, as defined in our evaluation, is a model with multiple abstractions.
+The top abstraction is the topology which contains information about the paths each cell can take from their starting point.
+Deeper abstractions involve the mapping of each cell to a particular branch within this network, and the position (or ordering) of each cells within these branches.
+Internally, the topology is represented by the milestone network and regions of delayed commitment, the branch assignment and cellular positions are represented by the milestone percentages (Figure \ref{fig:snote1fig_1}).
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_1.pdf}
+	\caption{
+		\textbf{An example trajectory that will be used throughout this section.}
+		It contains contains four milestones (W to Z) and five cells (a to e).
+	}
+	\label{fig:snote1fig_1}
+\end{figure}
+
+Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models using only one metric.
+We therefore sought to use different comparison metrics, each serving a different purpose:
+
+\begin{itemize}
+	\item \textbf{Specific metrics} investigate one particular aspect of the trajectory.
+Such metrics make it possible to find particular weak points for methods, e.g.
+that a method is very good at ordering but does not frequently find the correct topology.
+Moreover, having multiple individual metrics allow personalised rankings of methods, for example for users which are primarily interested in using the method correct topology.
+	\item \textbf{Application metrics} focus on the quality of a downstream analysis using the trajectory.
+For example, it measures whether the trajectory can be used to find accurate differentially expressed genes.
+	\item \textbf{Overall metrics} should capture all the different abstractions, in other words such metrics measure whether the resulting trajectory has a good topology, that the cells belong to similar branches \textit{and} that they are ordered correctly.
+\end{itemize}
+
+Here, we first describe and illustrate several possible specific, application and overall metrics.
+Next, we test these metrics on several test cases, to make sure they robustly identify "wrong" trajectory predictions.
+
+All metrics described here were implemented within the \href{https://github.com/dynverse/dyneval}{dyneval R package (https://github.com/dynverse/dyneval)}.
+
+\subsection{Metric characterisation and testing}
+
+\subsubsection{\textit{isomorphic}, \textit{edgeflip} and \textit{HIM}: Edit distance between two trajectory topologies}
+
+We used three different scores to assess the similarity in the topology between two trajectories, irregardless of where the cells were positioned.
+
+For all three scores, we first simplified the topology of the trajectory to make both graph structures comparable:
+
+\begin{itemize}
+	\item As we are only interested in the main structure of the topology without start or end, the graph was made undirected.
+	\item All milestones with degree 2 were removed. For example in the topology A $\Rightarrow$ B $\Rightarrow$ C $\Rightarrow$ D, C $\Rightarrow$ D, the B milestone was removed
+	\item A linear topology was converted to A $\Rightarrow$ B $\Rightarrow$ C
+	\item A cyclical topology such as A $\Rightarrow$ B $\Rightarrow$ C $\Rightarrow$ D or A $\Rightarrow$ B $\Rightarrow$ A were all simplified to A $\Rightarrow$ B $\Rightarrow$ C $\Rightarrow$ A
+	\item Duplicated edges such as A $\Rightarrow$ B, A $\Rightarrow$ B were decoupled to A $\Rightarrow$ B, A $\Rightarrow$ C $\Rightarrow$ B
+\end{itemize}
+
+The \textit{isomorphic} score returns 1 if two graphs are isomorphic, and 0 if they were not.
+For this, we used the used the BLISS algorithm \cite{junttila_engineeringefficientcanonical_2007}, as implemented in the R *igraph* package.
+
+The \textit{edgeflip} score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
+This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution \cite{bahiense_maximumcommonedge_2012}.
+We implemented a branch and bound approach for this problem, using several heuristics to speed up the search:
+
+\begin{itemize}
+	\item First check all possible edge additions and removals corresponding to the number of different edges between the two graphs.
+	\item For each possible solution, first check whether: \begin{enumerate}
+		\item The maximal degree is the same
+		\item The minimal degree is the same
+		\item All degrees are the same after sorting
+	\end{enumerate}
+	\item Only then check if the two graphs are isomorphic as described earlier.
+	\item If no solution is found, check all possible solutions with two extra edge additions/removals.
+\end{itemize}
+
+The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_HIM_Glocal_Metric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}).
+It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network.
+Conceptually, \textit{HIM} is a linear combination of:
+
+\begin{itemize}
+	\item The normalised Hamming distance \cite{doughertyValidationGeneRegulatory2011}, which calculates the distance between two graphs by matching individual edges in the adjacency matrix, but disregards overall structural similarity.
+	\item The normalised Ipsen-Mikhailov distance \cite{ipsenEvolutionaryReconstructionNetworks2002}, which calculates the overall distance of two graphs based on matches between its degree and adjacency matrix, while disregarding local structural similarities.
+It requires a $\gamma$ parameter, which is usually estimated based on the number of nodes in the graph, but which we fixed at $0.1$ so as to make the score comparable across different graph sizes.
+\end{itemize}
+
+We compared the three scores on several common topologies (Figure \ref{fig:snote1fig_2}a).
+While conceptually very different, the \textit{edgeflip} and \textit{HIM} still produce similar scores (Figure \ref{fig:snote1fig_2}b).
+The \textit{HIM} tends to punish the detection of cycles, while the \textit{edgeflip} is more harsh for differences in the number of bifurcations (Figure \ref{fig:snote1fig_2}b).
+The main difference however is that the \textit{HIM} takes into account edge lengths when comparing two trajectories, as illustrated in (Figure \ref{fig:snote1fig_2}c).
+Short "extra" edges in the topology are less punished by the \textit{HIM} than by the \textit{edgeflip}.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_2.pdf}
+	\caption{
+		\textbf{Showcase of three metrics to evaluate topologies: \textit{isomorphic}, \textit{edgeflip} and \textit{HIM}}
+		\textbf{(a)} The used topologies.
+\textbf{(b)} The scores when comparing each pair of trajectory types.
+(c) Four datasets in which aan extra edge is added and made progressively longer.
+This shows how the HIM can take into account edge lengths.
+	}
+	\label{fig:snote1fig_2}
+\end{figure}
+
+To summarise, the different topology based scores are useful for different scenarios:
+
+\begin{itemize}
+	\item If the two trajectories should only be compared when the topology is exactly the same, the \textit{isomorphic} should be used.
+	\item If it is important that the topologies are similar, but not necessarily isomorphic, the \textit{edgeflip} is most appropriate.
+	\item If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the \textit{HIM} is most appropriate.
+\end{itemize}
+
+\subsubsection{$\textit{F1}_{\textit{branches}}$ and $\textit{F1}_{\textit{milestones}}$: Comparing how well the cells are clustered in the trajectory}
+
+Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone \textit{or} branch \ref{fig:snote1fig_3}.
+These clusters of cells can then be compared using one of the many external cluster evaluation measures \cite{saelens_comprehensiveevaluationmodule_2018}.
+When selecting a cluster evaluation metric, we had two main conditions:
+
+- Because we allow methods to filter cells in the trajectory, the metric should be able to handle "non-exhaustive assignment", where some cells are not assigned to any cluster.
+- The metric should give each cluster equal weight, so that rare cell stages are equally important as large stages.
+
+The $\textit{F1}$ score between the $\textit{Recovery}$ and $\textit{Relevance}$ is a metric which conforms to both these conditions.
+This metric will map two clustersets by using their shared members based on the $\textit{Jaccard}$ similarity.
+It then calculates the $\textit{Recovery}$ as the average maximal $\textit{Jaccard}$ for every cluster in the first set of clusters (in our case the reference trajectory).
+Conversely, the $\textit{Relevance}$ is calculated based on the average maximal similarity in the second set of clusters (in our case the prediction).
+Both the $\textit{Recovery}$ and $\textit{Relevance}$ are then given equal weight in a harmonic mean ($\textit{F1}$).
+Formally, if $C$ and $C'$ are two cell clusters:
+
+\begin{align*}
+\textit{Jaccard}(c, c') &= \frac{|c \cap c'|}{|c \cup c'|} \\
+\textit{Recovery} &= \frac{1}{|C|} \sum_{c \in C}{\max_{c' \in C'}{\textit{Jaccard(c, c')}}} \\
+\textit{Relevance} &= \frac{1}{|C'|} \sum_{c' \in C'}{\max_{c \in C}{\textit{Jaccard(c, c')}}} \\
+\textit{F1} &= \frac{2}{\frac{1}{\textit{Recovery}} + \frac{1}{\textit{Relevance}}}
+\end{align*}
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_3.pdf}
+	\caption{
+		\textbf{Mapping cells to their closest milestone or branch for the calculation of the {$\textit{F1}_{\textit{milestones}}$} and {$\textit{F1}_{\textit{branches}}$} .}
+		To calculate the {$\textit{F1}_{\textit{milestones}}$}, cells are mapped towards the nearest milestone, i.e.
+the milestone with the highest milestone percentage.
+For the {$\textit{F1}_{\textit{branches}}$}, the cells are mapped to the closest edge.
+	}
+	\label{fig:snote1fig_3}
+\end{figure}
+
+\subsubsection{$\textit{cor}_{\textit{dist}}$: Correlation between geodesic distances}
+
+When the position of a cell is the same in both the reference and the prediction, its \textit{relative} distances to all other cells in the trajectory should also be the same.
+This observation is the basis for the $\textit{cor}_{\textit{dist}}$ metric.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_4.pdf}
+	\caption{
+		\textbf{The calculation of geodesic distances on a small example trajectory.}
+		\textbf{a)} A toy example containing four milestones (W to Z) and five cells (a to e).
+\textbf{b)} The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
+\textbf{c)} The calculations made for calculating the pairwise geodesic distances.
+\textbf{d)} A heatmap representation of the pairwise geodesic distances.
+	}
+	\label{fig:snote1fig_4}
+\end{figure}
+
+The geodesic distance is the distance a cell has to go through the trajectory space to get from one position to another.
+The way this distance is calculated depends on how two cells are positioned, showcased by an example in Figure \ref{fig:snote1fig_4}:
+
+\begin{itemize}
+	\item \textbf{Both cells are on the same edge in the milestone network.} In this case, the geodesic distance is defined as the product of the difference in milestone percentages and the length of their shared edge.
+For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
+	\item \textbf{Cells reside on different edges in the milestone network.} First, the distance of the cell to all its nearby milestones is calculated, based on its percentage within the edge and the length of the edge.
+These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
+For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
+\end{itemize}
+
+The geodesic distance can be easily extended towards cells within regions of delayed commitment.
+When both cells are part of the same region of delayed commitment, the geodesic distance was defined as the manhattan distances between the milestone percentages weighted by the lengths from the milestone network.
+For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1) = 1.9$.
+The distance between two cells where only one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
+
+Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets.
+For this reason, a set of waypoint cells are defined \textit{a priori}, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories (Figure \ref{fig:snote1fig_5}a).
+These cell waypoints are determined by viewing each milestone, edge and region of delayed commitment as a collection of cells.
+We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection.
+For calculating the $\textit{cor}_{\textit{dist}}$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
+
+To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate $\textit{cor}_{\textit{dist}}$.
+To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the $\textit{cor}_{\textit{dist}}$ at different levels of both cell shuffling and number of cell waypoints (Figure \ref{fig:snote1fig_5}a).
+We found that using cell waypoints does not induce a systematic bias in the $\textit{cor}_{\textit{dist}}$, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_5.pdf}
+	\caption{
+		\textbf{Determination of cell waypoints} \textbf{a)} Illustration of the stratified cell sampling using an example dataset (top).
+Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom).
+\textbf{b)} Accuracy versus time to calculate $\textit{cor}_{\textit{dist}}$.
+Shown are distributions over 100 random waypoint samples.
+The upper whisker of the boxplot extends from the hinge (75\% percentile) to the largest value, no further than 1.5$\times$ the IQR of the hinge.
+The lower whisker extends from the hinge (25\% percentile) to the smallest value, at most 1.5$\times$ the IQR of the hinge.
+	}
+	\label{fig:snote1fig_5}
+\end{figure}
+
+Although the $\textit{cor}_{\textit{dist}}$'s main characteristic is that it looks at the positions of the cells, other features of the trajectory are also (partly) captured.
+To illustrate this, we used the geodesic distances themselves as input for dimensionality reduction (Figure \ref{fig:snote1fig_6}) with varying topologies.
+This reduced space captures the original trajectory structure quite well, including the overall topology and branch lengths.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_6.pdf}
+	\caption{
+		\textbf{Determination of cell waypoints.} 
+		We generated different toy trajectory datasets with varying topologies and calculated the geodesic distances between all cells within the trajectory.
+We then used these distances as input for classical multidimensional scaling.
+This shows that the geodesic distances do not only contain information regarding the cell's positions, but also information on the lengths and wiring of the topology.
+	}
+	\label{fig:snote1fig_6}
+\end{figure}
+
+\subsubsection{$\textit{NMSE}_{\textit{rf}}$ and $\textit{NMSE}_{\textit{lm}}$: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory}
+
+An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory.
+If the cells are at similar positions in the trajectory (relative to its nearby cells), the prediction error should be low.
+
+Specifically, we implemented two metrics which predict the milestone percentages from the reference by using the predicted milestone percentages as features (Figure \ref{fig:snote1fig_7}).
+We did this with two regression methods, linear regression ($\textit{lm}$, using the R lm function) and Random Forest ($\textit{rf}$, implemented in the \textit{ranger} package  \cite{wright_rangerfastimplementation_2017}).
+In both cases, the accuracy of the prediction was measured using the Mean Squared error ($\mathit{MSE}$), in the case of Random forest we used the out-of-bag mean-squared error.
+Next, we calculated $\mathit{MSE}_{worst}$ equal to the $\mathit{MSE}$ when predicting all milestone percentages as the average.
+We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$.
+We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the $\textit{NMSE}_{\textit{rf}}$ and $\textit{NMSE}_{\textit{lm}}$ scores.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_7.pdf}
+	\caption{
+		\textbf{The calculation of $\textit{NMSE}_{\textit{lm}}$ distances on a small example trajectory.} 
+		The milestone percentages of the reference are predicted based on the milestone percentages of the prediction, using regression models such as linear regression or random forests.
+The predicted trajectory is then scored by comparing the mean-squared error (MSE) of this regression model with the baseline MSE where the prediction is the average milestone percentage.
+	}
+	\label{fig:snote1fig_7}
+\end{figure}
+
+\subsubsection{$\textit{cor}_{\textit{features}}$ and $\textit{wcor}_{\textit{features}}$: The accuracy of dynamical differentially expressed features/genes.}
+
+Although most metrics described above already assess some aspects directly relevant to the user, such as whether the method is good at finding the right topology, these metrics do not assess the quality of downstream analyses and hypotheses which can be generated from these models.
+
+Perhaps the main advantage of studying cellular dynamic processes using single-cell -omics data is that the dynamics of gene expression can be studied for the whole transcriptome.
+This can be used to construct other models such as dynamic regulatory networks and gene expression modules.
+Such analyses rely on a "good-enough" cellular ordering, so that it can be used to identify dynamical differentially expressed genes.
+
+To calculate the $\textit{cor}_{\textit{features}}$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory.
+More specifically, we first calculated the geodesic distances for each cell to all milestones in the trajectory.
+Next, we trained a Random Forest regression model (implemented in the R \textit{ranger} package \cite{wright_rangerfastimplementation_2017}, \href{https://github.com/imbs-hl/ranger}{https://github.com/imbs-hl/ranger}) to predict these distances for each milestone, based on the expression of genes within each cell.
+We then extracted feature importances using the Mean Decrease in Impurity (importance = 'impurity' parameter of the ranger function), as illustrated in Figure \ref{fig:snote1fig_8}.
+The overall importance of a feature (gene) was then equal to the mean importance over all milestones.
+Finally, we compared the two rankings by calculating the Pearson correlation, with values between -1 and 0 clipped to 0.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_8.pdf}
+	\caption{
+		\textbf{An illustration of ranking features based on their importance in a trajectory.} 
+		(a) A MDS dimensionality reudction of a real dataset in which mouse embryonic fibroblasts (MEF) differentiate into Neurons and Myocytes.
+(b) The ranking of feature importances from high to low.
+The majority of features have a very low importance.
+(c) Some examples, which were also highlighted in b.
+Higher features in the ranking are clearly specific to certain parts of the trajectory, while features lower on the ranking have a more dispersed expression pattern.
+	}
+	\label{fig:snote1fig_8}
+\end{figure}
+
+Random forest regression has two main hyperparameters.
+The number of trees to be fitted (num\_tree parameter) was fixed to 10000 to provide accurate and stable estimates of the feature importance (Figure \ref{fig:snote1fig_9}.
+The number of features on which can be split (mtry parameter) was set to 1\% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated features, omnipresent in transcriptomics data, are not suppressed in the ranking.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_9.pdf}
+	\caption{
+		\textbf{Effect of the number of trees parameter on the accuracy and variability of the $\textit{cor}_{\textit{features}}$.} 
+		We used the dataset from Figure \ref{fig:snote1fig_8} and calculated the $\textit{cor}_{\textit{features}}$ after shuffling a percentage of cells.
+	}
+	\label{fig:snote1fig_9}
+\end{figure}
+
+For most datasets, only a limited number of features will be differentially expressed in the trajectory.
+For example, in the dataset used in Figure \ref{fig:snote1fig_9} only the top 10\%-20\% show a clear pattern of differential expression.
+The correlation will weight each of these features equally, and will therefore give more weight to the bottom, irrelevant features.
+To prioritise the top differentially expressed features, we also implemented the $\textit{wcor}_{\textit{features}}$, which will weight the correlation using the feature importance scores in the reference so that the top features have relatively more impact on the score (Figure \ref{fig:snote1fig_10}).
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_10.pdf}
+	\caption{
+		\textbf{Effect of weighting the features based on their feature importance in the reference.} 
+		We used the same dataset as in Figure \ref{fig:snote1fig_8}, and calculated the $\textit{cor}_{\textit{features}}$ after shuffling a percentage of cells.
+	}
+	\label{fig:snote1fig_10}
+\end{figure}
+
+% TODO: check floatbarrier?
+%\FloatBarrier
+
+\subsection{Metric conformity}
+
+Although most metrics described in the previous section make sense intuitively, this does not necessarily mean that these metrics are robust and will generate reasonable results when used for benchmarking.
+This is because different methods and datasets will all lead to a varied set of trajectory models:
+
+\begin{itemize}
+	\item Real datasets have all cells grouped onto milestones
+	\item Some methods place all cells in a region of delayed commitment, others never generate a region of delayed commitment
+	\item Some methods always return a linear trajectory, even if a bifurcation is present in the data
+	\item Some methods filter cells
+\end{itemize}
+
+A good metric, especially a good overall metric, should work in all these circumstances.
+To test this, we designed a set of rules to which a good metric should conform, and assessed empirically whether a metric conforms to these rules.
+
+We generated a panel of toy datasets (using our \href{https://github.com/dynverse/dyntoy}{\textit{dyntoy}} package, \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}) with all possible combinations of:
+
+\begin{itemize}
+	\item \# cells: 10, 20, 50, 100, 200, 500
+	\item \# features: 200
+	\item topologies: linear, bifurcation, multifurcating, tree, cycle, connected graph and disconnected graph
+	\item Whether cells are placed on the milestones (as in real data) or on the edges/regions of delayed commitment between the milestones (as in synthetic data)
+\end{itemize}
+
+We then perturbed the trajectories in these datasets in certain ways, and tested whether the scores follow an expected pattern.
+An overview of the conformity of every metric is first given in Table \ref{tab:conformity_overview}.
+The individual rules and metric behaviour are discussed in the Supplementary Material that can be found at \href{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}.
+
+\begin{table}
+	\caption{Overview of whether a particular metric conforms to a particular rule} \label{tab:conformity_overview}
+
+	\centering\begingroup\fontsize{7}{9}\selectfont
+
+	\begin{tabular}{>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}}
+		\toprule
+		\rotatebox{0}{name} & \rotatebox{90}{$\textit{cor}_{\textit{dist}}$} & \rotatebox{90}{$\textit{NMSE}_{\textit{rf}}$} & \rotatebox{90}{$\textit{NMSE}_{\textit{lm}}$} & \rotatebox{90}{$\textit{edgeflip}$} & \rotatebox{90}{$\textit{HIM}$} & \rotatebox{90}{isomorphic} & \rotatebox{90}{$\textit{cor}_{\textit{features}}$} & \rotatebox{90}{$\textit{wcor}_{\textit{features}}$} & \rotatebox{90}{$\textit{F1}_{\textit{branches}}$} & \rotatebox{90}{$\textit{F1}_{\textit{milestones}}$} & \rotatebox{90}{$\textit{mean}_{\textit{geometric}}$}\\
+		\midrule
+		Same score on identity & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Local cell shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Edge shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Local and global cell shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Changing positions locally and/or globally & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+
+		Cell filtering & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Removing divergence regions & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Move cells to start milestone & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Move cells to closest milestone & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Length shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+
+		Cells into small subedges & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		New leaf edges & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		New connecting edges & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Changing topology and cell position & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Bifurcation merging & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+
+		Bifurcation merging and changing cell positions & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Bifurcation concatentation & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Cycle breaking & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Linear joining & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Linear splitting & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+
+		Change of topology & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		Cells on milestones vs edges & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
+		\bottomrule
+	\end{tabular}
+	\endgroup{}
+\end{table}
+
+% \FloatBarrier
+
+\subsection{Score aggregation}
+
+To rank the methods, we need to aggregate on two levels: across \textbf{datasets} and across specific/application metrics to calculate an \textbf{overall metric}.
+
+\subsubsection{Aggregating over datasets}
+
+When combining different datasets, it is important that the biases in the datasets does not influence the overall score.
+In our study, we define three such biases, although there are potentially many more:
+
+\begin{itemize}
+	\item \textbf{Difficulty of the datasets} Some datasets are more difficult than others.
+This can have various reasons, such as the complexity of the topology, the amount of biological and technical noise, or the dimensions of the data.
+It is important that a small increase in performance on a more difficult dataset has an equal impact on the final score as a large increase in performance on easier datasets.
+	\item \textbf{Dataset sources} It is much easier to generate synthetic datasets than real datasets, and this bias is reflected in our set of datasets.
+However, given their higher biological relevance, real datasets should be given at least equal importance than synthetic datasets.
+	\item \textbf{Trajectory types} There are many more linear and disconnected real datasets, and only a limited number of tree or graph datasets.
+This imbalance is there because historically most datasets have been linear datasets, and because it is easy to create disconnected datasets by combining different datasets.
+However, this imbalance in trajectory types does not necessarily reflect the general importance of that trajectory type.
+\end{itemize}
+
+We designed an aggregation scheme which tries to prevent these biases from influencing the ranking of the methods.
+
+The difficulty of a dataset can easily have an impact on how much weight the dataset gets in an overall ranking.
+We illustrate this with a simple example in Figure \ref{fig:snote1fig_11}.
+One method consistently performs well on both the easy and the difficult datasets.
+But because the differences are small in the difficult datasets, the mean would not give this method a high score.
+Meanwhile, a variable method which does not perform well on the difficult dataset gets the highest score, because it scored so high on the easier dataset.
+
+To avoid this bias, we normalise the scores of each dataset by first scaling and centering to $\mu = 0$ and $\sigma = 1$, and then moving the score values back to $[0, 1]$ by applying the unit normal density distribution function.
+This results in scores which are comparable across different datasets (Figure \ref{fig:snote1fig_11}).
+In contrast to other possible normalisation techniques, this will still retain some information on the relative difference between the scores, which would have been lost when using the ranks for normalisation.
+An example of this normalisation, which will also be used in the subsequent aggregation steps, can be seen in Figure \ref{fig:snote1fig_12}.
+
+\begin{figure}
+	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_11.pdf}
+	\caption{
+		\textbf{An illustration of how the difficulty of a dataset can influence the overall ranking.} 
+		A decent method, which consistently ranks high on an easy and difficult dataset, does not get a high score when averaging.
+On the other hand, a method which ranks high on the easy dataset, but very low on the difficult dataset does get a high score on average.
+After normalising the scores (right), this problem disappears.
+	}
+	\label{fig:snote1fig_11}
+\end{figure}
+
+\begin{figure}
+	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_12.pdf}
+	\caption{
+		\textbf{An example of the normalisation procedure.} 
+		Shown are some results of a benchmarking procedure, where every row contains the scores of a particular method (red shading) on a particular dataset (blue shading), with a trajectory type (green shading) and dataset source (orange shading).
+	}
+	\label{fig:snote1fig_12}
+\end{figure}
+
+After normalisation, we aggregate step by step the scores from different datasets.
+We first aggregate the datasets with the same dataset source and trajectory type using an arithmetic mean of their scores (Figure \ref{fig:snote1fig_13}a).
+Next, the scores are averaged over different dataset sources, using a arithmetic mean which was weighted based on how much the synthetic and silver scores correlated with the real gold scores (Figure \ref{fig:snote1fig_13}b).
+Finally, the scores are aggregated over the different trajectory types again using a arithmetic mean (Figure \ref{fig:snote1fig_13}c).
+
+\begin{figure}
+	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_13.pdf}
+	\caption{
+		\textbf{An example of the aggregation procedure.} 
+		In consecutive steps we aggregated across \textbf{(a)} different datasets with the same source and trajectory type, \textbf{(b)} different dataset sources with the same trajectory type (weighted for the correlation of the dataset source with the real gold dataset source) and \textbf{(c)} all trajectory types.
+	}
+	\label{fig:snote1fig_13}
+\end{figure}
+
+\subsubsection{Overall metrics}
+
+Undoubtedly, a single optimal overall metric does not exist for trajectories, as different users may have different priorities:
+
+\begin{itemize}
+	\item A user may be primarily interested in defining the correct topology, and only use the cellular ordering when the topology is correct
+	\item A user may be less interested in how the cells are ordered within a branch, but primarily in which cells are in which branches
+	\item A user may already know the topology, and may be primarily interested in finding good features related to a particular branching point
+	\item ...
+\end{itemize}
+
+Each of these scenarios would require a combinations of \textit{specific} and \textit{application} metrics with different weights.
+To provide an "overall" ranking of the metrics, which is impartial for the scenarios described above, we therefore chose a metric which weighs every aspect of the trajectory equally:
+
+\begin{itemize}
+	\item Its \textbf{ordering}, using the $\textit{cor}_{\textit{dist}}$
+	\item Its \textbf{branch assignment}, using the $\textit{F1}_{\textit{branches}}$
+	\item Its \textbf{topology}, using the $\textit{HIM}$
+	\item The accuracy of \textbf{differentially expressed features}, using the $\textit{wcor}_{\textit{features}}$
+\end{itemize}
+
+Next, we considered three different ways of averaging different scores: the arithmetic mean, geometric mean and harmonic mean.
+Each of these types of mean have different use cases.
+The harmonic mean is most appropriate when the scores would all have a common denominator (as is the case for the $\textit{Recovery}$ and $\textit{Relevance}$ described earlier).
+The arithmetic mean would be most appropriate when all the metrics have the same range.
+For our use case, the geometric mean is the most appropriate, because it is low if one of the values is low.
+For example, this means that if a method is not good at inferring the correct topology, it will get a low overall score, even if it performs better at all other scores.
+This ensures that a high score will only be reached if a prediction has a good ordering, branch assignment, topology, and set of differentially expressed features.
+
+The final overall score (Figure \ref{fig:snote1fig_14}) for a method was thus defined as:
+
+$$\textit{Overall} = \textit{mean}_{\textit{geometric}} = \sqrt[4]{\textit{cor}_{\textit{dist}} \times \textit{F1}_{\textit{branches}} \times \textit{HIM} \times \textit{wcor}_{\textit{features}}}$$
+
+\begin{figure}
+	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_14.pdf}
+	\caption{
+		\textbf{An example of the averaging procedure.} 
+		For each method, we calculated the geometric mean between its normalised and aggregated scores
+	}
+	\label{fig:snote1fig_14}
+\end{figure}
+
+We do however want to stress that different use cases will require a different overall score to order the methods.
+Such a context-dependent ranking of all methods is provided through the dynguidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}).
+
+\clearpage
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-08-09 ch2_dynbenchmark.tex
fix citations and remove bibliography at end of chapter

@@
 	\item \textbf{Component 1}: This method returns the first component of a principal component analysis (PCA) dimensionality reduction as a linear trajectory.
-This method is especially relevant as it has been used in a few studies already \cite{kouno_temporaldynamicstranscriptional_2013,zeng_pseudotemporalorderingsingle_2017}.
+This method is especially relevant as it has been used in a few studies already \cite{Kouno2013-km,zeng_pseudotemporalorderingsingle_2017}.
 
@@
 
-The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdom_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter2011,marbach_wisdom_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
 At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
-One driver of bifurcation seems to be mutual antagonism, where genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
+One driver of bifurcation seems to be mutual antagonism, where genes \cite{Xu2015-kn} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
 Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
-Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
+Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{Yosef2013-mj}.
 
@@
 
-The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{norel_selfassessmenttrap_2011}.
+The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{Norel2011}.
 Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer.
@@
 Such a context-dependent ranking of all methods is provided through the dynguidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}).
-
-\clearpage
-
-\clearpage
-\section{References}
-\printbibliography[heading=none]

### 2019-08-09 ch2_dynbenchmark.tex
small fixes

@@
 	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_2.pdf}
-	Figure caption on next page $\rightarrow$
-\end{figure}
-
-\clearpage % insert a page break
-\captionof{figure}{
+	\caption{
 		\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
@@
 \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation.
-Methods that errored on more than 50\% of the datasets are not included in this figure and are shown instead in Supplementary Fig.
+Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig.
 2.
@@
 	\label{fig:figure_2}
+\end{figure}
 
@@
 	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_3.pdf}
-	Figure caption on next page $\rightarrow$
-\end{figure}
-\clearpage % insert a page break
-\captionof{figure}{	
+	\caption{	
 		\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
@@
 	\label{fig:figure_3}
+\end{figure}
 
@@
 	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_3.pdf}
-	Figure caption on next page $\rightarrow$
-\end{figure}
-
-\clearpage % insert a page break
-\captionof{figure}{
+	\caption{
 		\textbf{Accuracy of trajectory inference methods.}
@@
 	\label{fig:supfigure_3}
+\end{figure}
 
@@
 To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods.
-A top model in this case was defined as a model with an overall score of at least 95\% as the best model.
-On all datasets, using one method resulted in getting a top model about 27\% of the time.
-This increased up to 74\% with the addition of six other methods (Figure \ref{fig:figure_4}a).
+A top model in this case was defined as a model with an overall score of at least 95$\%$ as the best model.
+On all datasets, using one method resulted in getting a top model about 27$\%$ of the time.
+This increased up to 74$\%$ with the addition of six other methods (Figure \ref{fig:figure_4}a).
 The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA.
@@
 		\textbf{Complementarity between different trajectory inference methods.}
-		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95\% of the best model) when applied to all datasets.
+		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95$\%$ of the best model) when applied to all datasets.
 \textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types.
@@
 It is not only important that a method is able to infer an accurate model in a reasonable time frame, but also that it produces a similar model when given very similar input data.
-To test the stability of each method, we executed each method on ten different subsamples of the datasets (95\% of the cells, 95\% of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (Figure \ref{fig:figure_3}d).
+To test the stability of each method, we executed each method on ten different subsamples of the datasets (95$\%$ of the cells, 95$\%$ of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (Figure \ref{fig:figure_3}d).
 
@@
 
-	\begin{tabular}{p{2.5cm}p{7.5cm}p{3cm}}
+	\begin{tabular}{p{3.5cm}p{9cm}p{2.5cm}}
 		\toprule
@@
 We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
-However, ‘good-yet-not-the-best’ methods \cite{norel_selfassessmenttrap_2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases.
+However, ‘good-yet-not-the-best’ methods \cite{Norel2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases.
 This is also supported by our analysis of method complementarity.
@@
 This was then processed as a type 1, direct, method.
-	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25\%.
+	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25$\%$.
 This was then processed as a type 7, cell graph, method.
@@
 	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
-
 \end{itemize}
@@
 We calculated the number of differentially expressed features within a trajectory using a two-way Mann–Whitney U test between every pair of cell groups.
-These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5\% of cells, and had at least a fold-change of 2.
+These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5$\%$ of cells, and had at least a fold-change of 2.
 We also calculated several other parameters, such as drop-out rates and library sizes using the Splatter package \cite{zappia_splattersimulationsinglecell_2017}.
@@
 For each dataset, we used the same number of genes as were differentially expressed in the real datasets.
-5\% of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module).
+5$\%$ of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module).
 We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module.
@@
 To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdom_2012}.
-In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):
+In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter2011} (ODEs):
 
@@
 
-Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows.
+Similar to GeneNetWeaver \cite{schaffter2011}, we sample the different parameters from random distributions, defined as follows.
 $e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.
@@
 The advantages of this pipeline are that it works both with and without spike-ins, and it includes a harsh cell filtering that looks at abnormalities in library sizes, mitochondrial gene expression and the number of genes expressed using median absolute deviations (which we set to 3).
-We required that a gene was expressed in at least 5\% of the cells and that it should have an average expression higher than 0.02.
-Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5\% and a biological component higher than 0.5.
+We required that a gene was expressed in at least 5$\%$ of the cells and that it should have an average expression higher than 0.02.
+Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5$\%$ and a biological component higher than 0.5.
 As a final filter, we removed both all-zero genes and cells until convergence.
@@
 We then extracted feature importance values for each feature and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset to give more weight to large differences.
-As hyperparameters we set the number of trees to 10,000 and the number of features on which to split to 1\% of all available features.
+As hyperparameters we set the number of trees to 10,000 and the number of features on which to split to 1$\%$ of all available features.
 We illustrate this metric and assess the effect of its hyperparameters in Supplementary Note 1.
@@
 
-To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score was equal to or higher than 95\% of the best overall score for that particular dataset.
+To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score was equal to or higher than 95$\%$ of the best overall score for that particular dataset.
 We then calculated for every method the weighted percentage of datasets that fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method.
@@
 However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R or the `random.seed` function in numpy).
-To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25\% of the datasets accounting for 15\% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
-We subsampled each dataset 10 times with 95\% of the original cells and 95\% of the original features.
+To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25$\%$ of the datasets accounting for 15$\%$ of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
+We subsampled each dataset 10 times with 95$\%$ of the original cells and 95$\%$ of the original features.
 We ran every method on each of the bootstraps, and assessed the stability by calculating the benchmarking scores between each pair of subsequent models (run $i$ is compared to run $i+1$).
@@
 Shown are distributions over 100 random waypoint samples.
-The upper whisker of the boxplot extends from the hinge (75\% percentile) to the largest value, no further than 1.5$\times$ the IQR of the hinge.
-The lower whisker extends from the hinge (25\% percentile) to the smallest value, at most 1.5$\times$ the IQR of the hinge.
+The upper whisker of the boxplot extends from the hinge (75$\%$ percentile) to the largest value, no further than 1.5$\times$ the IQR of the hinge.
+The lower whisker extends from the hinge (25$\%$ percentile) to the smallest value, at most 1.5$\times$ the IQR of the hinge.
 	}
@@
 The number of trees to be fitted (num\_tree parameter) was fixed to 10000 to provide accurate and stable estimates of the feature importance (Figure \ref{fig:snote1fig_9}.
-The number of features on which can be split (mtry parameter) was set to 1\% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated features, omnipresent in transcriptomics data, are not suppressed in the ranking.
+The number of features on which can be split (mtry parameter) was set to 1$\%$ of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated features, omnipresent in transcriptomics data, are not suppressed in the ranking.
 
@@
 For most datasets, only a limited number of features will be differentially expressed in the trajectory.
-For example, in the dataset used in Figure \ref{fig:snote1fig_9} only the top 10\%-20\% show a clear pattern of differential expression.
+For example, in the dataset used in Figure \ref{fig:snote1fig_9} only the top 10$\%$-20$\%$ show a clear pattern of differential expression.
 The correlation will weight each of these features equally, and will therefore give more weight to the bottom, irrelevant features.

### 2019-08-11 ch1_introduction.tex
update texs and files
use zotero instead

@@
 \begin{quote}
-	\textbf{Abstract:} Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner.
+	\textbf{Abstract:}
+
+	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner.
 Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following.
@@
 
-\section{Single cell biology}
+\section{The cell}
+The cell is the smallest unit of life, of which all known living organisms are composed.
+Every cell houses a plethora of biomolecular processes that allow it to adapt to changes in their environment continuously.
+It can be very challenging to comprehend the cellular response to a signal due to the dynamic nature of these processes.
+A reductionist approach to understanding a complex biological system is to study the biochemical components which it is comprised of\cite{brigandt_reductionismbiology_2017}.
+
+Reductionist biologists are delighted by recent advances in experimental technologies that permit measuring the abundance of thousands of different biochemical molecules in tens of thousands of individual cells.
+Observing the biomolecular insides of cells in this manner will ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for diagnosing and treating disease.
+Every coin has its flip side, however, and in this case, it is that the amount of data generated from these experiments is not analysable by hand.
+
+For example, the Human Cell Atlas (HCA) consortium\cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
+Experts in the field often metaphorically describe the HCA initiative as aiming to develop a 'Google Maps' of the human body.
+Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs\cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.
+
+The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret.
+Namely, the data inherently also suffers from batch effects arising from differences between donors and labs, and also contains high levels of noise arising from the experimental profiling techniques used\cite{hon_humancellatlas_2018}.
+Biologists thus turn to computer scientists\footnote{or computational biologists turn to themselves} to develop new tools to tackle these problems and help biologists extract meaningful biological insights from the data.
+
+This work makes incremental contributions to the field in order to be able to address the aforementioned problems in a more comprehensive context.
+This chapter first introduces several fundamental concepts in both cell biology and computer science.
+Afterwards, the research objectives and main contributions of this work are outlined.
 
@@
 
-\subsection{Supervised learning}
+\subsection{Dimensionality reduction}
 
-\subsection{Unsupervised learning}
+\subsection{Trajectory inference}
 
 \subsection{Feature selection}
-
-\subsection{Trajectory inference}
 
@@
 
-\section{Outline}
+\section{Outline} % include contributions?

### 2019-08-11 ch2_dynbenchmark.tex
use zotero instead

@@
 
-Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scaling_2017,Etzrodt2014-tr}.
-Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{Trapnell2015,cannoodt_computational_2016,moon_manifoldlearningbasedmethods_2018}.
+Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
+Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}.
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.
-TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process1, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{Aibar2017a}.
-Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_scienceforumhuman_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{Aibar2017a,angerer_singlecellsmake_2017} and user-friendly TI methods.
+TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}.
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
 
 A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1).
-Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecelllistsoftware_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
+Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
 While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect.
-Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{Haghverdi2016,Setty2016-og}) or through parameters provided by the user \cite{Trapnell2014,matsumoto_scoupprobabilisticmodel_2016}.
+Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}.
 These methods therefore mainly focus on correctly ordering the cells along the fixed topology.
-More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2017,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.
+More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.
 
 Given the diversity in TI methods, it is important to quantitatively assess their performance, scalability, robustness and usability.
-Many attempts at tackling this issue have already been made \cite{Haghverdi2016,Ji2016-il,Welch2016-ue,matsumoto_scoupprobabilisticmodel_2016,duverle_celltreebioconductorpackage_2016,Cannoodt2016a,lonnberg_singlecellrnaseqcomputational_2017,campbell_probabilisticmodelingbifurcations_2017,wolf_graphabstractionreconciles_2017}, but a comprehensive comparison of TI methods across a large number of different datasets is still lacking.
+Many attempts at tackling this issue have already been made \cite{haghverdi_diffusionpseudotimerobustly_2016,ji_tscanpseudotimereconstruction_2016,welch_slicerinferringbranched_2016,matsumoto_scoupprobabilisticmodel_2016,duverle_celltreebioconductorpackage_2016,cannoodt_computationalmethodstrajectory_2016,lonnberg_singlecellrnaseqcomputational_2017,campbell_probabilisticmodelingbifurcations_2017,wolf_graphabstractionreconciles_2017}, but a comprehensive comparison of TI methods across a large number of different datasets is still lacking.
 This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea of which would optimally solve their problem.
@@
 All other real datasets were classified as ‘silver standard’.
-For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter2011}.
+For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}.
 For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.
@@
 
-While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_ten_2017}.
+While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_tensimplerules_2017}.
 We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}).
@@
 6 right) and we believe that these should receive extra attention from developers.
-Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_best_2014,artaza_top10metrics_2016}.
+Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_bestpracticesscientific_2014,artaza_top10metrics_2016}.
 Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods.
@@
 
-		Open source & (1) Method's code is freely available (2) The code can be run on a freely available platform & \cite{l_rpackages_2017,wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
-
-		Version control & The code is available on a public version controlled repository, such as Github & \cite{l_rpackages_2017,wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
-
-		Packaging & (1) The code is provided as a "package", exposing functionality through functions or shell commands (2) The code can be easily installed through a repository such as CRAN, Bioconductor, PyPI, CPAN, debian packages, … & \cite{l_rpackages_2017,wickham_packagesorganizetest_2015,jimenez_foursimplerecommendations_2017,silva_generalguidelinesbiomedical_2017}\\*
-
-		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*
-
-		License & (1) The code is licensed (2) License allows academic use & \cite{l_rpackages_2017,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
+		Open source & (1) Method's code is freely available (2) The code can be run on a freely available platform & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
+
+		Version control & The code is available on a public version controlled repository, such as Github & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
+
+		Packaging & (1) The code is provided as a "package", exposing functionality through functions or shell commands (2) The code can be easily installed through a repository such as CRAN, Bioconductor, PyPI, CPAN, debian packages, … & \cite{lee_rpackagespackagedevelopment_2017,wickham_packagesorganizetest_2015,jimenez_foursimplerecommendations_2017,silva_generalguidelinesbiomedical_2017}\\*
+
+		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*
+
+		License & (1) The code is licensed (2) License allows academic use & \cite{lee_rpackagespackagedevelopment_2017,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
 
@@
 
-		Function and object naming & (1) Functions/commands have well chosen names (2) Arguments/parameters have well chosen names & \cite{wilson_best_2014,wickham_packagesorganizetest_2015}\\*
-
-		Code style & (1) Code has a consistent style (2) Code follows (basic) good practices in the programming language of choice, for example PEP8 or the tidyverse style guide & \cite{wilson_best_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016}\\*
-
-		Code duplication & Duplicated code is minimal & \cite{wilson_best_2014,wickham_packagesorganizetest_2015}\\
-
-		Self-contained functions & The method is exposed to the user as self-contained functions or commands & \cite{anderson_writing_2016,taschuk_ten_2017,silva_generalguidelinesbiomedical_2017}\\*
+		Function and object naming & (1) Functions/commands have well chosen names (2) Arguments/parameters have well chosen names & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\*
+
+		Code style & (1) Code has a consistent style (2) Code follows (basic) good practices in the programming language of choice, for example PEP8 or the tidyverse style guide & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016}\\*
+
+		Code duplication & Duplicated code is minimal & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\
+
+		Self-contained functions & The method is exposed to the user as self-contained functions or commands & \cite{anderson_writinggreatscientific_2016,taschuk_tensimplerules_2017,silva_generalguidelinesbiomedical_2017}\\*
 
@@
 		Dummy proofing & Package contains dummy proofing, i.e.
-testing whether the parameters and data supplied by the user make sense and are useful & \cite{l_rpackages_2017,karimzadeh_topconsiderationscreating_}\\
+testing whether the parameters and data supplied by the user make sense and are useful & \cite{lee_rpackagespackagedevelopment_2017,karimzadeh_topconsiderationscreating_}\\
 
@@
 
-		Unit testing & Method is tested using unit tests & \cite{l_rpackages_2017,wilson_best_2014,anderson_writing_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
-
-		Unit testing & Tests are run automatically using functionality from the programming language & \cite{l_rpackages_2017,wilson_best_2014,anderson_writing_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
-
-		Continuous integration & The method uses continuous integration, for example on Travis CI & \cite{beaulieu-jones_reproducibility_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
+		Unit testing & Method is tested using unit tests & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
+
+		Unit testing & Tests are run automatically using functionality from the programming language & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
+
+		Continuous integration & The method uses continuous integration, for example on Travis CI & \cite{beaulieu-jones_reproducibilitycomputationalworkflows_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
 
@@
 
-		Support & (1) There is a support ticket system, for example on Github (2) The authors respond to tickets and issues are resolved within a reasonable time frame & \cite{wilson_best_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
+		Support & (1) There is a support ticket system, for example on Github (2) The authors respond to tickets and issues are resolved within a reasonable time frame & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
 
@@
 
-		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
-
-		Inline documentation & Inline documentation is present in the code & \cite{wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
-
-		Parameter transparency & All important parameters are exposed to the user & \cite{taschuk_ten_2017}\\
+		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
+
+		Inline documentation & Inline documentation is present in the code & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
+
+		Parameter transparency & All important parameters are exposed to the user & \cite{taschuk_tensimplerules_2017}\\
 
@@
 
-		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{jfp_greendiceare_2016}\\*
+		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{puget_greendiceare_2016}\\*
 
@@
 
-Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{Aibar2017a} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}.
+Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}.
 Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}).
@@
 We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
-However, ‘good-yet-not-the-best’ methods \cite{Norel2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases.
+However, ‘good-yet-not-the-best’ methods \cite{norel_selfassessmenttrapcan_2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases.
 This is also supported by our analysis of method complementarity.
@@
 
-We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecelllistsoftware_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellpseudotimeoverviewalgorithms_2018}.
+We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
 We welcome any contributions by creating an issue at https://methods.dynverse.org.
@@
 	\item \textbf{Component 1}: This method returns the first component of a principal component analysis (PCA) dimensionality reduction as a linear trajectory.
-This method is especially relevant as it has been used in a few studies already \cite{Kouno2013-km,zeng_pseudotemporalorderingsingle_2017}.
+This method is especially relevant as it has been used in a few studies already \cite{kouno_temporaldynamicstranscriptional_2013,zeng_pseudotemporalorderingsingle_2017}.
 
@@
 	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
-	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulosPROSSTTProbabilisticSimulation2018}
+	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulos_prossttprobabilisticsimulation_2018}
 	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
@@
 
-The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter2011,marbach_wisdom_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
 At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
-One driver of bifurcation seems to be mutual antagonism, where genes \cite{Xu2015-kn} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
+One driver of bifurcation seems to be mutual antagonism, where genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
 Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
-Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{Yosef2013-mj}.
+Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
 
@@
 
-To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdom_2012}.
-In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter2011} (ODEs):
+To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
+In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):
 
@@
 
-Similar to GeneNetWeaver \cite{schaffter2011}, we sample the different parameters from random distributions, defined as follows.
+Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows.
 $e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.
@@
 
-PROSSTT is a recent data simulator \cite{papadopoulosPROSSTTProbabilisticSimulation2018}, which simulates expression using linear mixtures of expression programs and random walks through the trajectory.
+PROSSTT is a recent data simulator \cite{papadopoulos_prossttprobabilisticsimulation_2018}, which simulates expression using linear mixtures of expression programs and random walks through the trajectory.
 We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets.
@@
 
-The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{Norel2011}.
+The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{norel_selfassessmenttrapcan_2011}.
 Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer.
@@
 
-The HIM metric \cite{jurman_HIM_Glocal_Metric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length).
+The HIM metric \cite{jurman_himglocalmetric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length).
 It is a linear combination of the normalized Hamming distance, which gives an indication of the differences in edge lengths, and the normalized Ipsen–Mikhailov distance, which assesses the similarity in degree distributions.
@@
 The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
-The code assurance category is frequently overlooked, and checks for code testing, continuous integration \cite{beaulieu-jones_reproducibility_2017} and an active support system.
+The code assurance category is frequently overlooked, and checks for code testing, continuous integration \cite{beaulieu-jones_reproducibilitycomputationalworkflows_2017} and an active support system.
 The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
@@
 
-The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_HIM_Glocal_Metric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}).
+The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_himglocalmetric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}).
 It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network.
@@
 \begin{itemize}
-	\item The normalised Hamming distance \cite{doughertyValidationGeneRegulatory2011}, which calculates the distance between two graphs by matching individual edges in the adjacency matrix, but disregards overall structural similarity.
-	\item The normalised Ipsen-Mikhailov distance \cite{ipsenEvolutionaryReconstructionNetworks2002}, which calculates the overall distance of two graphs based on matches between its degree and adjacency matrix, while disregarding local structural similarities.
+	\item The normalised Hamming distance \cite{dougherty_validationgeneregulatory_2011}, which calculates the distance between two graphs by matching individual edges in the adjacency matrix, but disregards overall structural similarity.
+	\item The normalised Ipsen-Mikhailov distance \cite{ipsen_evolutionaryreconstructionnetworks_2002}, which calculates the overall distance of two graphs based on matches between its degree and adjacency matrix, while disregarding local structural similarities.
 It requires a $\gamma$ parameter, which is usually estimated based on the number of nodes in the graph, but which we fixed at $0.1$ so as to make the score comparable across different graph sizes.

### 2019-08-11 ch0_frontmatter.tex
fix logos

@@
 	\centering
-	\includegraphics[width=\textwidth]{fig/DI-eps-converted-to.pdf}
-
-	\vspace{1cm}
-
-	\includegraphics[width=2.25cm]{fig/ITGlogo_new} 
-	\vspace{2.25cm}
-
+	\ 
+	\vfill
 	{\Huge \textsf{Modelling single cell dynamics\\with trajectories and gene regulatory networks}}
@@
 
+	\includegraphics[height=.7cm]{fig/logos/icon_UGent_WE_EN_RGB_color}\hfill
+	\includegraphics[height=1cm]{fig/logos/vib_rf_inflammation_research_rgb_pos} \hfill
+	\includegraphics[height=.8cm]{fig/logos/dambi_logo} \hfill
+	\includegraphics[height=.7cm]{fig/logos/cmgg_myshort} \hfill
+	\includegraphics[height=.7cm]{fig/logos/FWO_Logo_KleurCMYK} 
+	\vspace{1cm}
+
 	{\textsf{Vakgroep Toegepaste Wiskunde, Informatica, en Statistiek\\
@@
 			Krijgslaan 281 - S2, 9000 Gent}}
+
 \end{titlepage}

### 2019-08-11 ch1_introduction.tex
fix logos

@@
 It can be very challenging to comprehend the cellular response to a signal due to the dynamic nature of these processes.
-A reductionist approach to understanding a complex biological system is to study the biochemical components which it is comprised of\cite{brigandt_reductionismbiology_2017}.
+A reductionist approach to understanding a complex biological system is to study the biochemical components which it is comprised of \cite{brigandt_reductionismbiology_2017}.
 
@@
 
-For example, the Human Cell Atlas (HCA) consortium\cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
+For example, the Human Cell Atlas (HCA) consortium \cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
 Experts in the field often metaphorically describe the HCA initiative as aiming to develop a 'Google Maps' of the human body.
-Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs\cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.
+Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs \cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.
 
 The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret.
-Namely, the data inherently also suffers from batch effects arising from differences between donors and labs, and also contains high levels of noise arising from the experimental profiling techniques used\cite{hon_humancellatlas_2018}.
+Namely, the data inherently also suffers from batch effects arising from differences between donors and labs, and also contains high levels of noise arising from the experimental profiling techniques used \cite{hon_humancellatlas_2018}.
 Biologists thus turn to computer scientists\footnote{or computational biologists turn to themselves} to develop new tools to tackle these problems and help biologists extract meaningful biological insights from the data.
@@
 Afterwards, the research objectives and main contributions of this work are outlined.
+
+\section{Single cell biology}
 

### 2019-08-15 ch1_introduction.tex
update

--- 2019-08-11 ch1_introduction.tex
@@
 This work makes incremental contributions to the field in order to be able to address the aforementioned problems in a more comprehensive context.
-This chapter first introduces several fundamental concepts in both cell biology and computer science.
+This chapter first introduces several key concepts in both cell biology and computer science, upon which the remainder of this work relies.
 Afterwards, the research objectives and main contributions of this work are outlined.
+% TODO: explain that only the main concepts are introduced?
+\subsection{The origins of life and the RNA world}
+The discovery of the double helix shape of Deoxyribonucleic Acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution.
+A modern high-school student can plausibly be expected to know that DNA serves as a medium for storing the genetic information required to reproduce a whole organism.
+With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.
+The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).
 
-\subsection{Dynamic processes}
+\begin{figure}
+	\centering\includegraphics[width=.5\linewidth]{fig/introduction/fsvm3} % TODO: change image
+	\caption{
+		\textbf{A prominent display of the double helix shape at the VIB FSVM building.} % TODO: expand description
+	}
+	\label{fig:fsvm}
+\end{figure}
 
-\subsection{Gene regulation}
+Even so, a widely-accepted hypothesis states that life (or cells) did not originate from DNA, but instead was kicked off from its lesser-known cousin, Ribonucleic Acid (RNA).
+According to the RNA world hypothesis \cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}).
+Only later did cells develop the ability to use DNA and proteins to self-sustain in a process commonly referred to as the Central Dogma.
+% TODO: use this reference? \cite{cornell_prebioticaminoacids_2019}
 
-\subsection{Single cell transcriptomics}
+\begin{figure}
+	\centering\includegraphics[width=.5\linewidth]{fig/introduction/rnaworld} % TODO: change image
+	\caption{
+		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life.
+Image from Horning (2011) \cite{horning_rnaworld_2011}.
+% TODO: expand the description
+	}
+	\label{fig:rnaworld}
+\end{figure}
+\subsection{The Central Dogma}
+The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existant living cells.
+In short, it states that DNA codes for RNA, which in turn codes for proteins.
+In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure 3).
 
+\textbf{Replication} is the process of duplicating DNA, which allows a cell to divide such that the resulting daughter cells retain complete copies of its genetic information.
+DNA consists of four different so-called nucleobases named adenine (A), cytosine (C), guanine (G), and thymine (T).
+Each strand of the double helix structure of DNA is a linear chain of nucleobases.
+The two strands are held together by hydrogen bonds since adenine can form three hydrogen bonds with thymine and cytosine can form two hydrogen bonds with guanine.
+Since each nucleobase can only bond with one other nucleobase, one strand is complementary to the other.
+Thus a strand of DNA can be represented by a sequence of letters, for example, "ACTCGGTTTAGCA".
+% This could be part of the DNA explanation.
+
+A stretch of DNA that contains a genetic blueprint for a particular molecule is called a gene, and the collection of an organism's genes is called its genome.
+\textbf{Transcription} is the process of synthesising an RNA molecule from a gene, and the resulting molecule is called a transcript.
+RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
+Due to its single-strandedness, RNA is less stable and will break down faster.
+Single-strandedness also allows some types of RNA (e.g.
+transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand.
+This work only considers messenger RNA (mRNA).
+mRNAs are transcribed from protein-coding genes, meaning that the mRNAs will lead to the production of protein molecules through the process of translation.
+% TODO: mention that there are about 20'000 protein-coding genes in the human body.
+
+\textbf{RNA splicing} is a process that occurs in almost all organisms (but not all) and results in the removal of specific regions in the RNA molecule.
+Introns are the regions removed by this process, and exons the regions that remain.
+The main functionality of splicing is to be able to create multiple variants of the same product, which can affect the enzymatic properties or localisation of the resulting product \cite{kelemen_functionalternativesplicing_2013}.
+
+During \textbf{translation}, a chain of amino acid is synthesised from an mRNA transcript.
+Every three nucleobases are translated into one of 21 different amino acids.
+The resulting chain of amino acids is folded up into a protein, the structure of which is determined by the sequence of different amino acids in the chain.
+In turn, its structure determines the functionality of the protein, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
+\subsection{Cell types and differentiation}
+The functionality provided by a cell is defined (mostly) by the proteins of which it consists.
+One common approach to trying to understand the functionality of a cell is to observe which molecules are present in the cell and to associate those molecules with functionality.
+
+\textit{Homo sapiens} like to categorise everything they encounter and so too have they conceptualised groups of cells called "cell types" according to their functionality.
+The concept of cell types eases reasoning about all aspects of biology, for instance, which cell types turn into (differentiate) or communicate with which other cell types, or how a cell type responds to a specific stimulation.
+Cells can be highly specialised toward performing a particular function (e.g.
+memory B cells accelerate immune response by remembering previously encountered pathogens), while other cells maintain a strong ability to differentiate into other cell types.
+
+% TODO: do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.
+Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task.
+In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.
+% TODO: are dynamic processes defined? a dynamic process is just a biochemical reaction
+\subsection{Cell dynamics and gene regulation}
+If cells are dynamic entities and can gradually produce the molecules needed to acquire new functionality, what is the process by which this happens?
+The mechanism by which this happens is called gene regulation.
+Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate).
+Such proteins are called transcription factors (TFs), and the genes it regulates are called its targets.
+Typically, one TF will regulate the transcription rate of many targets.
+% TODO: download database and calculate some statistics.
+
+Production of a specific molecule might require multiple cascades of gene regulation.
+The collection of all gene regulatory interactions between transcription factors and targets is called a gene regulatory network (GRN).
+Studying the active parts of a cell's gene regulatory network can thus reveal which dynamic processes are taking place.
+
+% TODO: explain regulation mechanisms such as transcription factor binding sites?
+\subsection{Profiling single cells}
+Understanding a biological process and proving hypotheses surrounding its workings require being able to profile (i.e.
+observe) the biomolecular components involved therein.
+The abundance of a particular molecule in a closed environment is called its expression levels.
+Technologies which permit profiling expression levels will typically be limited to only one facet of biomolecular machinery (e.g.
+DNA, RNA, or proteins), and can also be limited in the number of different molecules it can profile.
+
+Until recently, there were two main approaches to profiling biological systems: the "omics" approach, or the single-cell approach.
+
+"omics"\footnote{The etymology of "omics" is actually quite interesting \cite{yadav_wholenesssuffixomics_2007}.
+} is a collective term for profiling in a high-throughput manner.
+Genomics, transcriptomics, and proteomics respectively pertain to studying an organism's DNA (including all of its genes), RNA transcripts, and proteins.
+Traditional omics technologies have a strong trade-off, which is that they only profile an ensemble of cells and only return average expression levels.
+Notable examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts.
+By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
+
+Single-cell technologies, on the other hand, measure the expression levels of a set of molecules on a per-cell basis.
+The downside to traditional single-cell technologies (e.g.
+flow cytometry, mass cytometry, FISH, and PCR) is that they are limited in the number of different molecules which can be profiled simultaneously.
+
+Transformative technological advances have enabled the profiling of single cells in a high-throughput approach, thus creating room for single-cell omics (Figure \ref{fig:profiling_convergence}).
+The first such technologies allowed to profile the whole transcriptome of thousands of single cells.
+Current developments in the field allow for much greater cell counts, to also measure proteomic expression levels, or even employ imaging techniques to also capture the spatial information of each molecule within a cell.
+
+\begin{figure}
+	\centering\includegraphics[width=.5\linewidth]{fig/introduction/profiling_convergence} % TODO: change image
+	\caption{
+		\textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is ex-tending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
+Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}.
+% TODO: rewrite the description
+	}
+	\label{fig:profiling_convergence}
+\end{figure}
+
+The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
+Some of these opportunities lie in efforts like the Human Cell Atlas.
+The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories connecting the different cell types.
+As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells.
 \section{Machine learning}
+The sheer size of the datasets generated by single-cell omics technologies disallows for manual analysis of the individual profiles, which is something that biologists have relied upon for decades.
+Instead, biologists have to rely on computer scientists to develop new software to be able to extract meaningful information from that data.
+Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
 

### 2019-08-16 boek.tex
update document
update manuscript
update texs and files
use zotero instead
fix logos
update
update figs, texs, and other files

@@
-\documentclass[11pt,a4paper]{book}
+\documentclass[11pt,hidelinks]{book}
 
@@
 
+% Use Unicode characters when possible
+%\usepackage{xltxtra}
+
+%\usepackage[default,scale=0.95]{opensans}
+%\usepackage[utf8]{inputenc}
+%\usepackage[T1]{fontenc}
+
 % adjust margins (should be done before importing fancyhdr!)
-\usepackage[margin=2.5cm,a4paper]{geometry} %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm
+\usepackage[paperheight=24cm,paperwidth=16cm,margin=2.5cm,a4paper]{geometry} %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm
 
 % text packages
-\usepackage[british]{babel}
+\usepackage{csquotes}
+\usepackage[english]{babel}
 \usepackage{setspace}
@@
 % style packages
-\usepackage[T1]{fontenc}
 \usepackage{lmodern}
 \usepackage{fancyhdr}
-\usepackage[]{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
+%\usepackage[]{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
 
@@
 % figure packages
-\usepackage{tikz,pgfplots}
+\usepackage{tikz}
 \usetikzlibrary{calc,shapes,arrows,intersections,shadows}
@@
 \usepackage{import}
-\graphicspath{{fig/}}
 \usepackage{xcolor}
+\usepackage{colortbl}
 
@@
 \usepackage{tabulary}
+\usepackage{booktabs}
+
+% sections in allcaps
+\usepackage{titlesec}
+%\titleformat{\section} {\Large\bfseries\center}{}{0em}{\hrule height 0.4pt\vspace{2pt}}[\hrule height 0.4pt] \titlespacing*{\section} {0pt}{0pt}{10pt}
+%\titleformat{\subsection} {\large\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsection} {0pt}{0pt}{0pt}
+%\titleformat{\subsubsection} {\normalsize\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsubsection}  {0pt}{0pt}{0pt}
+%\titleformat{\section} [block]{\Large\bfseries\filcenter}{}{0em}{\MakeUppercase}
+%\titleformat{\subsection} [block]{\large\bfseries\filcenter}{}{0em}{\MakeUppercase}
+%\titleformat{\subsubsection} [block]{\normalsize\bfseries\filcenter}{}{0em}{\MakeUppercase}
 
@@
 \bibliography{library}
+%
+%% substitute in references
+%\DeclareDatafieldSet{setall}{
+%	\member[datatype=literal]
+%	\member[datatype=name]
+%	\member[field=journal]% journal is special since it is
+%	% actually journaltitle
+%}
+%
+%\DeclareSourcemap{
+%	\maps[datatype=bibtex]{
+%		\map[overwrite, foreach={setall}]{
+%			% \`{\i}
+%			\step[fieldsource=\regexp{$MAPLOOP},
+%			match=\regexp{\x{0131}\x{0300}},
+%			replace=\regexp{\x{00EC}}]
+%			% \'{\i}
+%			\step[fieldsource=\regexp{$MAPLOOP},
+%			match=\regexp{\x{0131}\x{0301}},
+%			replace=\regexp{\x{00ED}}]
+%			% \^{\i}
+%			\step[fieldsource=\regexp{$MAPLOOP},
+%			match=\regexp{\x{0131}\x{0302}},
+%			replace=\regexp{\x{00EE}}]
+%			% \"{\i}
+%			\step[fieldsource=\regexp{$MAPLOOP},
+%			match=\regexp{\x{0131}\x{0308}},
+%			replace=\regexp{\x{00EF}}]
+%		}
+%	}
+%}
 
@@
 \usepackage{doi}
+
+% add checkmark commands
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+
+% fix named refs for starred sections
+\makeatletter
+\def\ttl@useclass#1#2{%
+	\@ifstar
+	{\ttl@labeltrue\@dblarg{#1{#2}}}% {\ttl@labelfalse#1{#2}[]}%
+	{\ttl@labeltrue\@dblarg{#1{#2}}}}
+\makeatother
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% DEFINE FIGURE HELPERS %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\hugefigure}{\textwidth}
+\newcommand{\LARGEfigure}{0.8\textwidth} % 1 per col
+\newcommand{\LARgefigure}{0.6\textwidth} % 1 per col
+\newcommand{\Largefigure}{0.5\textwidth} % 1 per col
+\newcommand{\largefigure}{0.37\textwidth} % 2 per col
+\newcommand{\normalfigure}{0.3\textwidth} % 3 per col
+\newcommand{\smallfigure}{0.24\textwidth} % 3 per col
+\newcommand{\Smallfigure}{0.19\textwidth} % 4 per col
+\newcommand{\SMALLfigure}{0.15\textwidth} % 5 per col
+
+\makeatletter
+\newcommand*{\centerfloat}{%
+	\parindent \z@
+	\leftskip \z@ \@plus 1fil \@minus \textwidth
+	\rightskip\leftskip
+	\parfillskip \z@skip}
+\makeatother
+
+\newcommand{\picfig}[4]{
+\begin{figure}[ht!]
+	\centering \centerfloat
+	\includegraphics[width=#2]{#1}
+	\caption{#4}
+	\label{#3}
+\end{figure}}
+
+%\newcommand{\picsubf}[2]{
+%	\begin{subfigure}[b]{#2}
+%		\fbox{\includegraphics[width=\linewidth]{#1}}
+%\end{subfigure}}
+%
+%\newcommand{\picsubfc}[4]{
+%	\begin{subfigure}[b]{#4}
+%		\fbox{\includegraphics[width=\linewidth]{#1}}
+%		\caption{#2}
+%		\label{#3}
+%\end{subfigure}}
+%
+%\newcommand{\picsub}[2]{
+%	\begin{subfigure}[b]{#2}
+%		\includegraphics[width=\linewidth]{#1}
+%\end{subfigure}}
+%
+%\newcommand{\picsubc}[4]{
+%	\begin{subfigure}[b]{#4}
+%		\includegraphics[width=\linewidth]{#1}
+%		\caption{#2}
+%		\label{#3}
+%\end{subfigure}}
+
 %%%%%%%%%%%%%%%%%%%%%
@@
 
-% hyperlinks
+%% hyperlinks
 % load hyperref after fncychap
 \usepackage{hyperref}
-\hypersetup{%
-	pdftitle = {A title},
-	pdfsubject = {A subject},
-	pdfkeywords = {import keywords},
-	pdfauthor = {Me of course},
-	pdfcreator = {\LaTeX},
-}
 
 % Set style variables
-\setlength{\parindent}{0cm} 
+%\setlength{\parindent}{0cm} 
 \renewcommand{\contentsname}{Table of Contents}
@@
 \renewcommand{\appendixname}{}
+\renewcommand{\baselinestretch}{1.25}
+
+\usepackage{fontspec}
+
+\setmainfont [Path = fonts/,
+UprightFont = *-300,
+ItalicFont = *-300-Italic,
+BoldFont = *-700,
+BoldItalicFont = *-700-Italic
+]{MuseoSans}
 
 \begin{document}
-
-%\onehalfspace
 
@@
 
-% geen paginanummering tot we aan de inhoudsopgave komen
 \pagenumbering{alph}
@@
 
-% tabel van afkortingen en symbolen
-
-% hoofdstukken
+% chapters
 \mainmatter
@@
 
-\include{ch1_singlecellbiology}
+\include{ch1_introduction}
+
+% landmark mds
+% princurve
+% GillespieSSA
+% dyngen
 
@@
 
+% include gng ?
+
 \include{ch4_dyno}
 
-\include{ch5_incgraph}
+\include{ch5_brood}
 
-\include{ch6_brood}
+\include{ch6_incgraph}
 
@@
 
-% appendices
-
-\appendix
-
-\printbibliography
-
-\backmatter
-
-% eventueel: lijst van figuren en tabellen
-%\listoffigures
-%\listoftables
-
-% lege pagina (!!)
+\include{chX_backmatter}
 

### 2019-08-16 ch1_introduction.tex
update figs, texs, and other files

@@
 
-Adapted from:\\
+Partially adapted from:\\
 \textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y.
@@
 \doi{10.1002/eji.201646347}.\\
+Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
+A comparison of single-cell trajectory inference methods.
+\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
+\doi{10.1038/s41587-019-0071-9}.\\
 {\footnotesize $^*$ Equal contribution}
@@
 The first such technologies allowed to profile the whole transcriptome of thousands of single cells.
-Current developments in the field allow for much greater cell counts, to also measure proteomic expression levels, or even employ imaging techniques to also capture the spatial information of each molecule within a cell.
+Current developments in the field allow for much greater cell counts, to also measure other levels of information such as proteomic expression and spatial information.
+% maybe use this image? https://github.com/arnavm/multimodal-scRNA-seq
 
@@
 	\caption{
-		\textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is ex-tending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
+		\textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
 Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}.
@@
 As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells.
-\section{Machine learning}
-The sheer size of the datasets generated by single-cell omics technologies disallows for manual analysis of the individual profiles, which is something that biologists have relied upon for decades.
-Instead, biologists have to rely on computer scientists to develop new software to be able to extract meaningful information from that data.
-Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
 
+\section{Computational tools}
+The sheer size of the datasets generated by single-cell omics technologies disallows for manual analysis of the individual profiles.
+However, it opens up novel opportunities for collaborations between biologists and computer scientists to develop new computational tools to infer models of dynamic processes in an automated and unbiased fashion.
+This section introduces various types of computational tools which are of particular interest when analysing single-cell omics data \footnote{Yet their applications and implications are much more comprehensive than only single-cell omics.}.
 \subsection{Dimensionality reduction}
+The dimensionality of single-cell omics datasets is typically much too high for most modelling algorithms to tackle directly.
+Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful representation with fewer dimensions.
+DR methods construct this representation by reducing the gene space, the cell space, or both while attempting to maintain the main characteristics of the original data.
+% Could use a figure here.
+Ideally, the reduced representation should have a dimensionality that corresponds to the intrinsic dimensionality of the biological system.
 
+There are many ways of classifying DR methods \cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
+Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible.
+Examples of commonly used projection-based DR methods in single-cell omics are PCA, MDS, LMDS.
+Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g.
+a graph or a grid), visualising the structure in a lower-dimensional space, and mapping the original samples to the lower-dimensional space.
+Manifold learning can be an iterative optimisation process using a predefined criterion.
+Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.
+
+% For more information, we refer the reader to other reviews for a thorough discussion of the advantages and disadvantages of different dimensionality reduction methods [37, 38].
 \subsection{Trajectory inference}
+Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
+Such dynamic processes can be modelled computationally using trajectory inference (TI) methods which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}.
+The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.
+
+The initial dataset can be either a single snapshot of a mixture of cells in different stages or a set of samples collected at different time points (Figure \ref{fig:trajectory_inference}A). 
+TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017} (Figure \ref{fig:trajectory_inference}B).
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
+
+\begin{figure}
+	\centering\includegraphics[width=\Largefigure]{fig/introduction/trajectory_inference_slim} % TODO: Update figure to match text!!!!
+	\caption{
+		Applications of single-cell trajectory inference methods.
+(A) Single-cell omics data appropriate for TI can be both obtained from an unsynchronised population of single cells (snapshot data) but also from synchronised cell populations.
+(B) UPDATE!
+	}
+	\label{fig:trajectory_inference}
+\end{figure}
+
+A plethora of TI methods have been developed over the past few years, and still, more are created every month (Supplementary Table 1).
+Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the "awesome-single-cell" list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
+Each method has its own unique set of characteristics in terms of the underlying algorithm, required prior information and produced outputs.
+Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect.
+Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}.
+Therefore, these methods mainly focus on correctly ordering the cells along transitions in a fixed topology.
+More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.
 
@@
 \section{Research objectives}
+Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
 

### 2019-08-16 ch2_dynbenchmark.tex
update figs, texs, and other files

@@
 \section{Introduction}
-
-Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
-Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}.
-The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.
-TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}.
-Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
-
-A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1).
-Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
-While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect.
-Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}.
-These methods therefore mainly focus on correctly ordering the cells along the fixed topology.
-More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.
 

### 2019-08-19 ch1_introduction.tex
update

@@
 \doi{10.1038/s41587-019-0071-9}.\\
+%Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y.
+Network Inference from Single-Cell Transcriptomic Data.
+\textit{Gene Regulatory Networks} (2019), 235--249.
+\doi{10.1007/978-1-4939-8882-2_10}.
+\\
 {\footnotesize $^*$ Equal contribution}
+
 \newpage
@@
 % TODO: explain that only the main concepts are introduced?
+
 \subsection{The origins of life and the RNA world}
@@
 In turn, its structure determines the functionality of the protein, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
-\subsection{Cell types and differentiation}
+\subsection{Cell types}
 The functionality provided by a cell is defined (mostly) by the proteins of which it consists.
@@
 % TODO: are dynamic processes defined? a dynamic process is just a biochemical reaction
+
 \subsection{Cell dynamics and gene regulation}
@@
 \subsection{Profiling single cells}
-Understanding a biological process and proving hypotheses surrounding its workings require being able to profile (i.e.
+In order to understand a biological process, it is often quite helpful to be able to profile (i.e.
 observe) the biomolecular components involved therein.
-The abundance of a particular molecule in a closed environment is called its expression levels.
-Technologies which permit profiling expression levels will typically be limited to only one facet of biomolecular machinery (e.g.
-DNA, RNA, or proteins), and can also be limited in the number of different molecules it can profile.
+The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.
 
-Until recently, there were two main approaches to profiling biological systems: the "omics" approach, or the single-cell approach.
+The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
+Since it was developed by Coons et al.
+in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins \cite{coons_immunologicalpropertiesantibody_1941}.
+In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system.
+A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.
 
-"omics"\footnote{The etymology of "omics" is actually quite interesting \cite{yadav_wholenesssuffixomics_2007}.
-} is a collective term for profiling in a high-throughput manner.
-Genomics, transcriptomics, and proteomics respectively pertain to studying an organism's DNA (including all of its genes), RNA transcripts, and proteins.
-Traditional omics technologies have a strong trade-off, which is that they only profile an ensemble of cells and only return average expression levels.
-Notable examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts.
+IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation.
+Using different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously.
+Characterising cells in a quantifiable way is labour intensive; however, since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation).
+While modern implementations of IHC improve the throughput drastically by using robots to automate the image acquisition and computer software to automate cell segmentation, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.
+
+Flow cytometry \cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
+Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies.
+
+Since IHC and flow cytometry, many new technologies have been developed which allow quantifying expression levels of molecules in single cells (e.g.
+mass cytometry, single-cell qPCR, FISH).
+All of these single-cell (non-omics) technologies are limited by the number of different molecules they could measure, however; and thus required handpicking the molecules of interest before performing an experiment, making the experiment biased towards the preconceptions of the experimenter.
+
+On the other side of the spectrum are the so-called "omics" technologies.
+"Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
+There are many types of "omics", but the most commonly used are the following.
+In genomics, all of an organism's genes are studied -- its whole genome.
+Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively.
+A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics.
+If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable.
+Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts.
 By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
+% Demonstrate the masking effect of bulk analyses.
 
-Single-cell technologies, on the other hand, measure the expression levels of a set of molecules on a per-cell basis.
-The downside to traditional single-cell technologies (e.g.
-flow cytometry, mass cytometry, FISH, and PCR) is that they are limited in the number of different molecules which can be profiled simultaneously.
-
-Transformative technological advances have enabled the profiling of single cells in a high-throughput approach, thus creating room for single-cell omics (Figure \ref{fig:profiling_convergence}).
-The first such technologies allowed to profile the whole transcriptome of thousands of single cells.
-Current developments in the field allow for much greater cell counts, to also measure other levels of information such as proteomic expression and spatial information.
-% maybe use this image? https://github.com/arnavm/multimodal-scRNA-seq
+Transformative technological advances in microvolume sequencing allowed Tang et al.
+to analyse the transcriptome at single-cell resolution \cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A).
+During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells (Figure \ref{fig:profiling_convergence}B) and measuring other levels of information such as proteomic expression levels (Figure \ref{fig:profiling_convergence}C).
 
 \begin{figure}
-	\centering\includegraphics[width=.5\linewidth]{fig/introduction/profiling_convergence} % TODO: change image
+	\centering
+	A. \includegraphics[width=.45\linewidth]{fig/introduction/profiling_convergence} 
+	% also figure with cell numbers
+	C. \includegraphics[width=.45\linewidth]{fig/introduction/multimodal} 
+	% TODO: change image
 	\caption{
+		A.
 \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
 Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}.
+		B. 
+		C. scmultiomics \cite{moudgil_multimodalscrnaseq_2019}. 
 		% TODO: rewrite the description
@@
 \section{Computational tools}
-The sheer size of the datasets generated by single-cell omics technologies disallows for manual analysis of the individual profiles.
-However, it opens up novel opportunities for collaborations between biologists and computer scientists to develop new computational tools to infer models of dynamic processes in an automated and unbiased fashion.
-This section introduces various types of computational tools which are of particular interest when analysing single-cell omics data \footnote{Yet their applications and implications are much more comprehensive than only single-cell omics.}.
+The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
+
+Some of these opportunities lie in efforts such as the Human Cell Atlas.
+The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories which connect the different cell types.
+As part of this endeavour, the consortium will perform single-cell omics on tens or even hundreds of millions of cells.
+
+Single-cell omics permits new types of analyse but also come with hitherto unseen data characteristics, the combination of which poses exciting new challenges for the computational community to tackle  (Figure \ref{fig:comp_tools}A)\cite{stegle_computationalanalyticalchallenges_2015,yuan_challengesemergingdirections_2017,chen_singlecellrnaseqtechnologies_2019}.
+These challenges include:
+\begin{itemize}
+	\item dimensionality reduction: providing a visual and informative overview of a given dataset, 
+	\item trajectory inference: identifying and characterising transitions between different cellular states, and
+	\item gene regulatory network inference: inferring regulatory interactions between transcription factors across individual cells,
+	\item normalisation: separating biological noise from technical noise.
+\end{itemize}
+
+\begin{figure}
+	\centering\includegraphics[width=.5\linewidth]{fig/introduction/singlecell_tools}
+	\centering\includegraphics[width=.3\linewidth]{fig/introduction/scrnatools}
+	% TODO: change image
+	\caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016) \cite{wagner_revealingvectorscellular_2016}.
+		B. Zappia et al. (2018) \cite{zappia_exploringsinglecellrnaseq_2018}
+		% TODO: rewrite the description
+	}
+	\label{fig:comp_tools}
+\end{figure}
+
+The computational community has been very excited about trying to tackle these problems, having released more than 450 software tools for processing single-cell omics data in one form of another (Figure \ref{fig:comp_tools}B) \cite{zappia_exploringsinglecellrnaseq_2018}.
+Many of these methods attempt to tackle the abovementioned challenges (Figure \ref{fig:comp_tools}C), which will be discussed in the following sections.
+% TODO: provide better connection to the subsections below
+
 \subsection{Dimensionality reduction}
@@
 Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible.
-Examples of commonly used projection-based DR methods in single-cell omics are PCA, MDS, LMDS.
+Examples of commonly used projection-based DR methods in single-cell omics are PCA and MDS.
 Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g.
@@
 
-\subsection{Feature selection}
+% TI methods often use a DR method
 
-\subsection{Network inference}
+\subsection{Gene regulatory network inference}
+
+Traditionally, regulatory interactions are inferred from bulk transcriptional profiles, generated by pooling together the RNA transcripts of a supposedly homogeneous population of several thousands of cells, and quantifying the transcript abundance through a microarray or RNA sequencing.
+Incorrect assumptions on the homogeneity of the pooled cells may lead to the masking of relevant expression patterns in rare cell populations, as expression values are averaged over the whole population (Figure \ref{fig:bulk_vs_singlecell}).
+In addition, NI methods rely on a diverse set of time-series and perturbation experiments in order to reliably identify causal regulatory interactions.
+However, such experiments are expensive and time-consuming, and an inaccurate choice of time points might result in crucial intermediate stages being missed.
+
+One of the main advantages of single-cell transcriptomics is the ability to quantify the exact cellular state of thousands of cells per experiment.
+The intercellular heterogeneity caused by naturally occurring biological stochasticity \cite{Padovan-Merhar2013} can be exploited to infer regulatory interactions between transcription factors (TFs) and their target genes (see Figure \ref{fig:bulk_vs_singlecell}).
+In this sense, heterogeneity in the cell population will ease the inference of networks, rather than mask condition-specific expression patterns and regulatory interactions.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\Largefigure]{fig/introduction/ni_bulk_vs_sc}
+	\caption{Bulk expression data return the average expressions of genes among large numbers of cells.
+In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are typically used.
+On the other hand, sequencing the transcriptome at the single cell level uncovers the high variability among cells, providing the necessary information to directly infer gene regulatory networks.}
+	\label{fig:bulk_vs_singlecell}
+\end{figure}
+
+While single-cell transcriptomics offers many advantages over traditional bulk profiling methods,
+several computational challenges pertaining to the preprocessing of the data have a big impact on 
+single-cell NI \cite{Stegle2015ComputationalChallenges}.  In this chapter, we will therefore firstly focus on zero inflation, confounding factors and scalability problems (Section \ref{challenges}).
+We will then discuss several recent developments in single-cell transcriptomics analysis that present a high interest to further improve NI methods.
+In the second part, we will focus on novel unsupervised learning methods that have been proposed for inferring the different cellular states within a heterogeneous cell population.
+These methods can help to increase the accuracy of NI by deriving differential, dynamic or profile-specific regulatory networks (Section \ref{structure-learning}).
+Lastly, single-cell transcriptomics has opened up a gateway to performing high-throughput multi-omics and/or perturbation experiments, which could again revolutionise how gene regulatory networks are being inferred at a high-throughput scale (Section \ref{perturbed}).
 
@@
 
+Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.
+
 \section{Outline} % include contributions?

### 2019-08-21 ch1_introduction.tex
update

@@
 \doi{10.1038/s41587-019-0071-9}.\\
-%Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y.
+Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y.
 Network Inference from Single-Cell Transcriptomic Data.
@@
 \begin{itemize}
+	\item normalisation: separating biological noise from technical noise,
 	\item dimensionality reduction: providing a visual and informative overview of a given dataset, 
 	\item trajectory inference: identifying and characterising transitions between different cellular states, and
-	\item gene regulatory network inference: inferring regulatory interactions between transcription factors across individual cells,
-	\item normalisation: separating biological noise from technical noise.
+	\item gene regulatory network inference: inferring regulatory interactions between transcription factors across individual cells.
 \end{itemize}
@@
 
-The computational community has been very excited about trying to tackle these problems, having released more than 450 software tools for processing single-cell omics data in one form of another (Figure \ref{fig:comp_tools}B) \cite{zappia_exploringsinglecellrnaseq_2018}.
-Many of these methods attempt to tackle the abovementioned challenges (Figure \ref{fig:comp_tools}C), which will be discussed in the following sections.
 % TODO: provide better connection to the subsections below
 
+\subsection{Normalisation}
+% TODO: need to write still
+
 \subsection{Dimensionality reduction}
-The dimensionality of single-cell omics datasets is typically much too high for most modelling algorithms to tackle directly.
+Single-cell omics datasets typically have too many dimensions (features) in order to be easily interpretable by humans and even by most computational tools. 
 Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful representation with fewer dimensions.
-DR methods construct this representation by reducing the gene space, the cell space, or both while attempting to maintain the main characteristics of the original data.
-% Could use a figure here.
-Ideally, the reduced representation should have a dimensionality that corresponds to the intrinsic dimensionality of the biological system.
+It is important to note that its usage depends on the target audience: for humans -- to visualise data in a 2-D plane to aid with interpretation by humans, or for computers -- to construct a denser representation of the data such that it mostly contains the same information but with fewer dimensions.
 
@@
 Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.
+% section is not very interestingly written
 
-% For more information, we refer the reader to other reviews for a thorough discussion of the advantages and disadvantages of different dimensionality reduction methods [37, 38].
 \subsection{Trajectory inference}
 Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
-Such dynamic processes can be modelled computationally using trajectory inference (TI) methods which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}.
-The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.
+Trajectory inference (TI) is a new category of computational tools used to offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017,cannoodt_computationalmethodstrajectory_2016}.
 
-The initial dataset can be either a single snapshot of a mixture of cells in different stages or a set of samples collected at different time points (Figure \ref{fig:trajectory_inference}A). 
-TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017} (Figure \ref{fig:trajectory_inference}B).
+The dataset can be a single snapshot of a mixture of cells in different stages, or a set of samples collected at different time points (Figure \ref{fig:trajectory_inference}A).
+Typically, TI methods first analyse similarities between cells, optionally infer the topology of the underlying process, and finally order cells along that trajectory (Figure \ref{fig:trajectory_inference}B).
+The second step can be optional, as some methods assume a specific topology beforehand.
+TI methods allow the identification of new subsets of cells, delineation of a differentiation tree, and characterisation of the main driver genes along a state transition (Figure \ref{fig:trajectory_inference}C).
 Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
@@
 (B) UPDATE!
+(C) UPDATE!
 	}
@@
 
-A plethora of TI methods have been developed over the past few years, and still, more are created every month (Supplementary Table 1).
-Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the "awesome-single-cell" list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
-Each method has its own unique set of characteristics in terms of the underlying algorithm, required prior information and produced outputs.
-Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect.
-Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}.
-Therefore, these methods mainly focus on correctly ordering the cells along transitions in a fixed topology.
-More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.
+\subsection{Gene regulatory network inference}
+Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes.
+This topic already received much attention with the advent of bulk omics (before single-cell omics).
+These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods \cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.
 
-% TI methods often use a DR method
-
-\subsection{Gene regulatory network inference}
-
-Traditionally, regulatory interactions are inferred from bulk transcriptional profiles, generated by pooling together the RNA transcripts of a supposedly homogeneous population of several thousands of cells, and quantifying the transcript abundance through a microarray or RNA sequencing.
-Incorrect assumptions on the homogeneity of the pooled cells may lead to the masking of relevant expression patterns in rare cell populations, as expression values are averaged over the whole population (Figure \ref{fig:bulk_vs_singlecell}).
-In addition, NI methods rely on a diverse set of time-series and perturbation experiments in order to reliably identify causal regulatory interactions.
-However, such experiments are expensive and time-consuming, and an inaccurate choice of time points might result in crucial intermediate stages being missed.
-
-One of the main advantages of single-cell transcriptomics is the ability to quantify the exact cellular state of thousands of cells per experiment.
-The intercellular heterogeneity caused by naturally occurring biological stochasticity \cite{Padovan-Merhar2013} can be exploited to infer regulatory interactions between transcription factors (TFs) and their target genes (see Figure \ref{fig:bulk_vs_singlecell}).
-In this sense, heterogeneity in the cell population will ease the inference of networks, rather than mask condition-specific expression patterns and regulatory interactions.
+After the last DREAM competition, it seemed that interest in NI methodology had declined.
+After all, NI on bulk omics profiles suffered from several crucial issues.
+As mentioned previously, bulk profiles are generated by pooling together the RNA transcripts of a supposedly homogeneous population of thousands of cells.
+Since the expression values are averaged over the whole population, incorrect assumptions on the homogeneity of the pooled cells may lead to the masking of relevant expression patterns in rare cell populations (Figure \ref{fig:bulk_vs_singlecell}).
+Besides, NI methods rely on a diverse set of time-series and perturbation experiments in order to reliably identify causal regulatory interactions.
+Such experiments are expensive and time-consuming, and an inaccurate selection of time points might result in crucial intermediate stages being missed.
 
@@
 	\caption{Bulk expression data return the average expressions of genes among large numbers of cells.
-In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are typically used.
-On the other hand, sequencing the transcriptome at the single cell level uncovers the high variability among cells, providing the necessary information to directly infer gene regulatory networks.}
+In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required.
+On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: update label
 	\label{fig:bulk_vs_singlecell}
@@
 
-While single-cell transcriptomics offers many advantages over traditional bulk profiling methods,
-several computational challenges pertaining to the preprocessing of the data have a big impact on 
-single-cell NI \cite{Stegle2015ComputationalChallenges}.  In this chapter, we will therefore firstly focus on zero inflation, confounding factors and scalability problems (Section \ref{challenges}).
-We will then discuss several recent developments in single-cell transcriptomics analysis that present a high interest to further improve NI methods.
-In the second part, we will focus on novel unsupervised learning methods that have been proposed for inferring the different cellular states within a heterogeneous cell population.
-These methods can help to increase the accuracy of NI by deriving differential, dynamic or profile-specific regulatory networks (Section \ref{structure-learning}).
-Lastly, single-cell transcriptomics has opened up a gateway to performing high-throughput multi-omics and/or perturbation experiments, which could again revolutionise how gene regulatory networks are being inferred at a high-throughput scale (Section \ref{perturbed}).
+The advent of single-cell omics has made scientists wonder whether now is the time to revisit network inference \cite{stegle_computationalanalyticalchallenges_2015}. 
+One of the main advantages of single-cell omics is the ability to quantify the exact cellular state of thousands of cells per experiment.
+The heterogeneity between cells caused by naturally occurring biological randomness \cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs (see Figure \ref{fig:bulk_vs_singlecell}).
+In this setting, heterogeneity in the cell population eases network inference, rather than mask condition-specific expression patterns in regulatory interactions.
 
 \section{Research objectives}
+WIP
+
+The computational community has been very excited about trying to tackle these problems.
+By now, more than 450 software tools for processing single-cell omics data in one form of another have been released (Figure \ref{fig:comp_tools}B) \cite{zappia_exploringsinglecellrnaseq_2018}.
+Many of these methods attempt to tackle the abovementioned challenges (Figure \ref{fig:comp_tools}C).
+
 Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.

### 2019-08-21 ch4_dynbenchmark.tex
update

--- 2019-08-11 ch2_dynbenchmark.tex
@@
 \section{Introduction}
-
 Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.

### 2019-08-21 ch8_incgraph.tex
update

--- 2019-05-22 ch6_incgraph.tex
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Optimising regulatory networks} 
-\chaptermark{Optimising regulatory networks.} % short title
+\chapter{incgraph: Optimising regulatory networks} 
+\chaptermark{incgraph: Optimising regulatory networks.} % short title
 \label{chap:incgraph}
@@
 The advent of high-throughput profiling methods (e.g.
-microarrays and RNA sequencing) have allowed to observe the molecular contents of a cell, which in turn have enabled computational network inference methods to reverse engineer the biochemical interaction networks~\cite{Albert2007}.
-Improving the accuracy of inferred networks has been a long-standing challenge, but the development of ever more sophisticated algorithms and community-wide benchmarking studies have resulted in significant process~\cite{Marbach2010, Narendra2011, Marbach2012, Aijo2017}.
+microarrays and RNA sequencing) have allowed to observe the molecular contents of a cell, which in turn have enabled computational network inference methods to reverse engineer the biochemical interaction networks~\cite{albert_networkinferenceanalysis_2007}.
+Improving the accuracy of inferred networks has been a long-standing challenge, but the development of ever more sophisticated algorithms and community-wide benchmarking studies have resulted in significant process~\cite{marbach_revealingstrengthsweaknesses_2010, narendra_comprehensiveassessmentmethods_2011, marbach_wisdomcrowdsrobust_2012, aijo_biophysicallymotivatedregulatory_2017}.
 
-Several recent developments involve incorporating topological priors, either to guide the inference process~\cite{Lopes2014} or post-process the network~\cite{Ruyssinck2016}.
-A common prior is that biological networks are highly modular~\cite{Rives2003}, as they consist of multiple collections of functionally or physically linked molecules~\cite{Hartwell1999,Barabasi2004}.
-At the lowest level, each module is made up out of biochemical interactions arranged in small topological patterns, which act as fundamental building blocks~\cite{Milo2002}.
+Several recent developments involve incorporating topological priors, either to guide the inference process~\cite{lopes_featureselectiontechnique_2014} or post-process the network~\cite{ruyssinck_netterrerankinggene_2016}.
+A common prior is that biological networks are highly modular~\cite{rives_modularorganizationcellular_2003}, as they consist of multiple collections of functionally or physically linked molecules~\cite{hartwell_molecularmodularcell_1999,barabasi_networkbiologyunderstanding_2004}.
+At the lowest level, each module is made up out of biochemical interactions arranged in small topological patterns, which act as fundamental building blocks~\cite{milo_networkmotifssimple_2002}.
 
-Graphlets~\cite{Przulj2004} are one of the tools which could be used to add a topological prior to a biological network,
+Graphlets~\cite{przulj_modelinginteractomescalefree_2004} are one of the tools which could be used to add a topological prior to a biological network,
 Graphlets are small connected subnetworks which can be counted to identify which low-level topological patterns are present in a network.
@@
 An orbit represents a class of isomorphic (i.e. resulting in the same structure) positions of nodes within a graphlet (Fig \ref{fig_graphlets}A, coloured in red).
-Both graphlets and orbits have been used extensively for predicting the properties of nodes such as protein functionality~\cite{Milenkovic2008,Guerrero2008,Singh2014} and gene oncogenicity~\cite{Milenkovic2010a}, by performing network alignment~\cite{Kuchaiev2010,Milenkovic2013} or using them as a similarity measure in machine learning tasks~\cite{Shervashidze2009,Vacic2010}.
+Both graphlets and orbits have been used extensively for predicting the properties of nodes such as protein functionality~\cite{milenkovic_uncoveringbiologicalnetwork_2008,guerrero_characterizationproteasomeinteraction_2008,singh_graphletsignaturebasedscoring_2014} and gene oncogenicity~\cite{milenkovic_optimalnetworkalignment_2010}, by performing network alignment~\cite{kuchaiev_topologicalnetworkalignment_2010,milenkovic_globalnetworkalignment_2013} or using them as a similarity measure in machine learning tasks~\cite{shervashidze_efficientgraphletkernels_2009,vacic_graphletkernelsprediction_2010}.
 
@@
 Eppstein et al.
-introduced data structures and algorithms for updating the counts of size-three\cite{Eppstein2009} and size-four\cite{Eppstein2012} subgraphs in a dynamic setting.
-The data structures were determined such that the amortised time is $O(h)$ and $O(h^2)$, respectively, where $h$ is the h-index of the network\cite{Hirsch2005}.
+introduced data structures and algorithms for updating the counts of size-three\cite{eppstein_hindexgraphits_2009} and size-four\cite{eppstein_extendeddynamicsubgraph_2012} subgraphs in a dynamic setting.
+The data structures were determined such that the amortised time is $O(h)$ and $O(h^2)$, respectively, where $h$ is the h-index of the network\cite{hirsch_indexquantifyindividual_2005}.
 
@@
 \subsection*{Timing experiments}
-We compared the execution time needed to calculate the graphlet counts in iteratively modified networks between our method and a state-of-the-art non-incremental algorithm, Orca~\cite{Hocevar2014}.
-Orca is a heavily optimised algorithm for counting 5-node graphlets in static networks, and outperforms all other static graphlet counting algorithms by an order of magnitude~\cite{Hocevar2014}.
+We compared the execution time needed to calculate the graphlet counts in iteratively modified networks between our method and a state-of-the-art non-incremental algorithm, Orca~\cite{hocevar_combinatorialapproachgraphlet_2014}.
+Orca is a heavily optimised algorithm for counting 5-node graphlets in static networks, and outperforms all other static graphlet counting algorithms by an order of magnitude~\cite{hocevar_combinatorialapproachgraphlet_2014}.
 
@@
 
-The network models were based on three static network models: Barab\'asi-Albert~\cite{Albert2002}, Erd\H{o}s--R\'enyi~\cite{Erdos1959}, and Geometric~\cite{Appel1997}.
+The network models were based on three static network models: Barab\'asi-Albert~\cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~\cite{erdos_randomgraphs_1959}, and Geometric~\cite{appel_minimumvertexdegree_1997}.
 Pseudo code for these random evolving network models can be found in \nameref{code_ba}, \nameref{code_er}, and \nameref{code_geo} respectively.
@@
 cerevisiae}.
-The predicted networks were generated using the network inference method GENIE3~\cite{Vananh2010} with default parameters.
-Gene expression data was obtained from COLOMBOS~\cite{Moretto2016} and GEO~\cite{Edgar2002}, respectively.
-The predicted networks and the optimised versions thereof were compared against respective lists of known gene regulatory interactions~\cite{Gama-Castro2016,Ma2014}.
+The predicted networks were generated using the network inference method GENIE3~\cite{huynh-thu_inferringregulatorynetworks_2010} with default parameters.
+Gene expression data was obtained from COLOMBOS~\cite{moretto_colombosv3leveraging_2016} and GEO~\cite{edgar_geneexpressionomnibus_2002}, respectively.
+The predicted networks and the optimised versions thereof were compared against respective lists of known gene regulatory interactions~\cite{gama-castro_regulondbversionhighlevel_2016,ma_denovolearninggenomescale_2014}.
 
@@
 Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (\nameref{sfig_hindex}).
-This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{Eppstein2012} for counting three-size and four-size subgraphs respectively.
+This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.
 

### 2019-08-22 ch1_introduction.tex
update

@@
 \section{Research objectives}
-WIP
 
-The computational community has been very excited about trying to tackle these problems.
-By now, more than 450 software tools for processing single-cell omics data in one form of another have been released (Figure \ref{fig:comp_tools}B) \cite{zappia_exploringsinglecellrnaseq_2018}.
-Many of these methods attempt to tackle the abovementioned challenges (Figure \ref{fig:comp_tools}C).
+Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology.
+The new types of analyses made possible warrant the development of computational tools which can solve the problem at hand and deal with the new data characteristics, ultimately in order to create useful and accurate hypotheses for biologists.
 
-Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
+In response, scientists all over the globe have been working hard to tackle these computational challenges, resulting in more than 470 software tools for processing single-cell omics data in one form or another \cite{zappia_exploringsinglecellrnaseq_2018}.
+More than half of these tools were published in peer-reviewed journals.
 
-Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.
+The high number of tools available presents the user choice, but also uncertainty and indecision about which tool to use for the task at hand.
+Choosing the method with most citations might result in a weaker result than picking a method at random, and making an informed decision requires spending an enormous amount of time in comparing the results of numerous methods, on various platforms, on multiple different datasets.
+
+The field of single-cell omics needs quantitative ways to measure the performance of different categories of tools.
+Clear quantitative metrics allow performing comprehensive studies to assess the advantages and disadvantages of existing tools.
+In addition, they could also serve as a minimum performance criterion before the publication of a new tool.
+
+In the computational jungle of single-cell omics, experts act as beacons of hope by sharing guidelines based on comprehensive benchmarking studies.
+Their disseminations (in the form of manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}) are crucial in leading new users, and ultimately the whole field, to better practices for performing single-cell omics analyses.
+
+% todo: provide better segway to the next section
 
 \section{Outline} % include contributions?
+This work aims to tackle multiple computational problems in extracting biological hypotheses from single-cell omics data.
+We introduce a framework for evaluating multiple types of single-cell omics tools, more specifically dimensionality reduction, trajectory inference, and gene regulatory network inference (Chapter \ref{chap:dyngen}).
+Next, we perform a comprehensive evaluation of 45 TI methods and construct guidelines based on our observations (Chapter \ref{chap:dynbenchmark}).
+As part of the evaluation, we developed a framework to infer, visualise, and interpret trajectories, which we made available as a separate tool called dyno (Chapter \ref{chap:dyno}).
+We also introduce two new computational tools of our own.
+SCORPIUS is a TI method specialised in inferring linear trajectories (Chapter \ref{chap:scorpius}).
+bred is a true single-cell NI method which can infer gene regulatory networks for individual cells (Chapter \ref{chap:bred}).
+
+\textbf{Chapter \ref{chap:dyngen}} -- why dyngen, what is it, what can it be used for.
+%framework for evaluating methods with synthetic data generator. if available, real data with gold standards would be most desirable.
+
+\textbf{Chapter \ref{chap:dynbenchmark}} -- why dynbenchmark, what is it, what can it be used for.
+
+\textbf{Chapter \ref{chap:dyno}} -- why dyno, what is it, what can it be used for.
+
+\textbf{Chapter \ref{chap:scorpius}} -- why scorpius, what is it, what can it be used for.
+
+\textbf{Chapter \ref{chap:bred}} -- why bred, what is it, what can it be used for.
+
+%Before embarking on this perilous adventure, I attempt to write a "Hippocratic oath for bioinformaticians", to guide me through this dissertation and not stray from the metaphorical path of righteousness.
+%
+%\subsection{Hippocratic oath for bioinformaticians}
+%We proceed with caution in developing new tools.
+Its results should not only be convincing and easy to interpret, but also accurate, robust, and reproducible.
+We value creating open-source software that works reliably but fails gracefully when it does not.
+We acknowledge that writing automated tests is dull but necessary for maintaining long-term software projects.
+% TODO: improve
+
+\section{List of contributions}
+
+\subsection{First-author publications}
+
+\begin{itemize}
+	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
+	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
+SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
+bioRxiv.
+2016 Oct:079509.
+	\item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
+  \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
+  \item dyno, submitted?
+  \item dyngen, submitted?
+  \item scorpius, submitted?
+  \item bred, submitted?
+ \end{itemize}
+
+*: Equal contribution.
+
+\subsection{Co-author publications}
+\begin{itemize}
+	\item Decock A, Ongenaert M, \textbf{Cannoodt R}, Verniers K, De Wilde B, Laureys G, Van Roy N, Berbegall AP, Bienertova-Vasku J, Bown N, Clément N.
+Methyl-CpG-binding domain sequencing reveals a prognostic methylation signature in neuroblastoma.
+Oncotarget.
+2016 Jan 12;7(2):1960.
+	\item Van Cauwenbergh C, Van Schil K, \textbf{Cannoodt R}, Bauwens M, Van Laethem T, De Jaegere S, Steyaert W, Sante T, Menten B, Leroy BP, Coppieters F.
+arrEYE: a customized platform for high-resolution copy number analysis of coding and noncoding regions of known and candidate retinal dystrophy genes and retinal noncoding RNAs.
+Genetics in Medicine.
+2017 Apr;19(4):457.
+	\item Claeys S, Denecker G, \textbf{Cannoodt R}, Kumps C, Durinck K, Speleman F, De Preter K.
+Early and late effects of pharmacological ALK inhibition on the neuroblastoma transcriptome.
+Oncotarget.
+2017 Dec 5;8(63):106820.
+	\item Depuydt P, Boeva V, Hocking TD, \textbf{Cannoodt R}, Ambros IM, Ambros PF, Asgharzadeh S, Attiyeh EF, Combaret V, Defferrari R, Fischer M.
+Genomic amplifications and distal 6q loss: novel markers for poor survival in high-risk neuroblastoma patients.
+JNCI: Journal of the National Cancer Institute.
+2018 Mar 5;110(10):1084-93.
+	\item Scott CL, T'Jonck W, \ldots, \textbf{Cannoodt R}, Saelens W \ldots, Guilliams M.
+The transcription factor ZEB2 is required to maintain the tissue-specific identities of macrophages.
+Immunity.
+2018 Aug 21;49(2):312-25.
+	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
+	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. InGene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
+	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L.
+Trajectory-based differential expression analysis for single-cell sequencing data.
+BioRxiv.
+2019 Jan 1:623397.
+	\item Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, Robinson MD.
+Essential guidelines for computational method benchmarking.
+Genome biology.
+2019 Dec;20(1):125.
+\end{itemize}
+
+\subsection{Open-source software}
+At the moment, only a subset of the dynverse packages are present on CRAN: dynutils, babelwhale, dynparam and dyndimred.
+We are currently in the process of submitting the next packages in our dependency graph on CRAN as well: dynwrap and dynplot.
+In this work package, we will submit all other dynverse packages as well to CRAN, so that the dynverse toolkit becomes broadly distributed.
+This includes making sure the package code complies with CRAN policy, adding extensive documentation and examples, and updating the codebase based on the comments from the CRAN maintainers.
+
+Furthermore, the work packages also includes maintaining and extending existing software packages on CRAN or Bioconductor on which our packages depend.
+We already have extensive experience with contributing to our dependencies.
+This includes speeding up parts of the dependency (princurve, proxyC), fixing bugs (monocle, splatter, slingshot, rlang, devtools), becoming a maintainer of orphaned packages (princurve, GillespieSSA), extending the documentation (devtools) and adding new functionality (ParamHelpers).
+We will keep doing this within this work package, with a focus on dependencies that help with plotting (ggplot2, ggraph, patchwork) and trajectory inference (monocle, slingshot, PAGA).
+
+\paragraph{Author of the following packages:}
+\ \\
+{
+	\small
+	\begin{tabular}{llp{10cm}}
+		babelwhale    & CRAN   & Interacting with Docker and Singularity containers                        \\
+		dyndimred     & CRAN   & Applying dimensionality reduction methods                                 \\
+		dyneval       & Github & Evaluating trajectory inference methods                                   \\
+		dynfeature    & Github & Calculating feature importance scores from trajectories                   \\
+		dyngen        & Github & Simulating single-cell data using gene regulatory networks                \\
+		dynguidelines & Github & User guidelines for trajectory inference                                  \\
+		dynmethods    & Github & A collection of wrappers for trajectory inference methods                 \\
+		dyno          & Github & A pipeline for inferring, visualising and interpreting trajectories       \\
+		dynparam      & Github & Creating meta-information for parameters                                  \\
+		dynplot       & Github & A simple visualisation library for trajectories                           \\
+		dynplot2      & Github & A fully customisable visualisation library for trajectories               \\
+		dyntoy        & Github & Generating simple toy data of cellular differentiation                    \\
+		dynutils      & CRAN   & Common functionality for the dynverse packages                            \\
+		dynwrap       & Github & A common format for trajectories                                          \\
+		GillespieSSA  & CRAN   & Gillespie's Stochastic Simulation Algorithm (SSA)                         \\
+		GillespieSSA2 & CRAN   & Gillespie's Stochastic Simulation Algorithm for Impatient People          \\
+		gng           & Github & An Rcpp implementation of the Growing Neural Gas algorithm                \\
+		incgraph      & CRAN   & Incremental graphlet counting for network optimisation                    \\
+		princurve     & CRAN   & Fits a principal curve in arbitrary dimension                             \\
+		proxyC        & CRAN   & Computes proximity in large sparse matrices                               \\
+		qsub          & CRAN   & Running commands remotely on gridengine clusters                          \\
+		SCORPIUS      & CRAN   & Inferring developmental chronologies from single-cell RNA sequencing data
+	\end{tabular}
+}
+
+\paragraph{Contributor:}
+\ \\
+{
+	\small
+	\begin{tabular}{llp{10cm}}
+		ClusterSignificance & Bioconductor & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data                                           \\
+		devtools            & CRAN         & Tools to make developing R packages easier                                                                                                                       \\
+		merlot              & Github       & A method for reconstructing lineage-tree topologies from scRNA-seq data                                                                                          \\
+		mlr                 & CRAN         & Machine Learning in R                                                                                                                                            \\
+		monocle             & Bioconductor & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq                                                                             \\
+		ParamHelpers        & CRAN         & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning                                                                                    \\
+		pseudogp            & Github       & Probabilistic pseudotime for single-cell RNA-seq                                                                                                                 \\
+		remotes             & CRAN         & R package installation from remote repositories, including GitHub                                                                                                \\
+		rlang               & CRAN         & Functions for base types and core R and tidyverse features                                                                                                       \\
+		SCope               & Github       & Visualization of large-scale and high dimensional single cell data                                                                                               \\
+		slingshot           & Bioconductor & Tools for ordering single-cell sequencing                                                                                                                        \\
+		splatter            & Bioconductor & Simple simulation of single-cell RNA sequencing data                                                                                                             \\
+		URD                 & Github       & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
+		wishbone            & Github       & identify bifurcating developmental trajectories from single-cell data
+	\end{tabular}
+%\endgroup{}
+}

### 2019-08-23 ch1_introduction.tex
update

@@
   \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
+  \item dyngen, submitted?
   \item dyno, submitted?
-  \item dyngen, submitted?
   \item scorpius, submitted?
@@
 \subsection{Open-source software}
-At the moment, only a subset of the dynverse packages are present on CRAN: dynutils, babelwhale, dynparam and dyndimred.
-We are currently in the process of submitting the next packages in our dependency graph on CRAN as well: dynwrap and dynplot.
-In this work package, we will submit all other dynverse packages as well to CRAN, so that the dynverse toolkit becomes broadly distributed.
-This includes making sure the package code complies with CRAN policy, adding extensive documentation and examples, and updating the codebase based on the comments from the CRAN maintainers.
+As part of this work, many open-source software packages were created and many others were contributed to (Table \ref{tab:packages}).
 
-Furthermore, the work packages also includes maintaining and extending existing software packages on CRAN or Bioconductor on which our packages depend.
-We already have extensive experience with contributing to our dependencies.
-This includes speeding up parts of the dependency (princurve, proxyC), fixing bugs (monocle, splatter, slingshot, rlang, devtools), becoming a maintainer of orphaned packages (princurve, GillespieSSA), extending the documentation (devtools) and adding new functionality (ParamHelpers).
-We will keep doing this within this work package, with a focus on dependencies that help with plotting (ggplot2, ggraph, patchwork) and trajectory inference (monocle, slingshot, PAGA).
+Packages that were created as part of this work are hosted on Github under the username rcannood\footnote{\url{https://github.com/rcannood?tab=repositories}} or the dynverse organisation\footnote{\url{https://github.com/dynverse?tab=repositories}}.
+As part of our standard development practices, we automate execution of unit tests and writing extensive documentation to ensure the code complies with CRAN policy before submission.
+We aim to submit all other packages to CRAN as well.
 
-\paragraph{Author of the following packages:}
-\ \\
-{
-	\small
-	\begin{tabular}{llp{10cm}}
-		babelwhale    & CRAN   & Interacting with Docker and Singularity containers                        \\
-		dyndimred     & CRAN   & Applying dimensionality reduction methods                                 \\
-		dyneval       & Github & Evaluating trajectory inference methods                                   \\
-		dynfeature    & Github & Calculating feature importance scores from trajectories                   \\
-		dyngen        & Github & Simulating single-cell data using gene regulatory networks                \\
-		dynguidelines & Github & User guidelines for trajectory inference                                  \\
-		dynmethods    & Github & A collection of wrappers for trajectory inference methods                 \\
-		dyno          & Github & A pipeline for inferring, visualising and interpreting trajectories       \\
-		dynparam      & Github & Creating meta-information for parameters                                  \\
-		dynplot       & Github & A simple visualisation library for trajectories                           \\
-		dynplot2      & Github & A fully customisable visualisation library for trajectories               \\
-		dyntoy        & Github & Generating simple toy data of cellular differentiation                    \\
-		dynutils      & CRAN   & Common functionality for the dynverse packages                            \\
-		dynwrap       & Github & A common format for trajectories                                          \\
-		GillespieSSA  & CRAN   & Gillespie's Stochastic Simulation Algorithm (SSA)                         \\
-		GillespieSSA2 & CRAN   & Gillespie's Stochastic Simulation Algorithm for Impatient People          \\
-		gng           & Github & An Rcpp implementation of the Growing Neural Gas algorithm                \\
-		incgraph      & CRAN   & Incremental graphlet counting for network optimisation                    \\
-		princurve     & CRAN   & Fits a principal curve in arbitrary dimension                             \\
-		proxyC        & CRAN   & Computes proximity in large sparse matrices                               \\
-		qsub          & CRAN   & Running commands remotely on gridengine clusters                          \\
-		SCORPIUS      & CRAN   & Inferring developmental chronologies from single-cell RNA sequencing data
-	\end{tabular}
-}
+We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
+This includes help speed up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
+Several of these package receive millions of downloads per year (devtools, remotes, rlang).
 
-\paragraph{Contributor:}
-\ \\
-{
-	\small
-	\begin{tabular}{llp{10cm}}
-		ClusterSignificance & Bioconductor & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data                                           \\
-		devtools            & CRAN         & Tools to make developing R packages easier                                                                                                                       \\
-		merlot              & Github       & A method for reconstructing lineage-tree topologies from scRNA-seq data                                                                                          \\
-		mlr                 & CRAN         & Machine Learning in R                                                                                                                                            \\
-		monocle             & Bioconductor & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq                                                                             \\
-		ParamHelpers        & CRAN         & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning                                                                                    \\
-		pseudogp            & Github       & Probabilistic pseudotime for single-cell RNA-seq                                                                                                                 \\
-		remotes             & CRAN         & R package installation from remote repositories, including GitHub                                                                                                \\
-		rlang               & CRAN         & Functions for base types and core R and tidyverse features                                                                                                       \\
-		SCope               & Github       & Visualization of large-scale and high dimensional single cell data                                                                                               \\
-		slingshot           & Bioconductor & Tools for ordering single-cell sequencing                                                                                                                        \\
-		splatter            & Bioconductor & Simple simulation of single-cell RNA sequencing data                                                                                                             \\
-		URD                 & Github       & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
-		wishbone            & Github       & identify bifurcating developmental trajectories from single-cell data
-	\end{tabular}
-%\endgroup{}
-}
+\begin{table}[ht!]
+	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the role within the package: aut Author, ctb Contributor.
+Yearly download statistics are based on the number of downloads between 2018-07-01 and 2019-06-30.
+CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
+For Github repositories, no download statistics could be retrieved (N/A).
+} \label{tab:packages}
+
+	\centering\fontsize{9}{11}\selectfont
+	\begin{tabularx}{\linewidth}{|p{2cm}lllX|}
+		\hline
+		Name & Role & Repo & Downloads p.a. & Description \\ \hline\hline
+		babelwhale & aut & CRAN & N/A & Interacting with Docker and Singularity containers \\
+		dynbenchmark & aut & Github & N/A & Pipeline for benchmarking trajectory inference methods \\
+		dyndimred & aut & CRAN & 560 & Applying dimensionality reduction methods \\
+		dyneval & aut & Github & N/A & Evaluating trajectory inference methods \\
+		dynfeature & aut & Github & N/A & Calculating feature importance scores from trajectories \\
+		dyngen & aut & Github & N/A & Simulating single-cell data using gene regulatory networks \\
+		dynguidelines & aut & Github & N/A & User guidelines for trajectory inference \\
+		dynmethods & aut & Github & N/A & A collection of wrappers for trajectory inference methods \\
+		dyno & aut & Github & N/A & A pipeline for inferring, visualising and interpreting trajectories \\
+		dynparam & aut & CRAN & 790 & Creating meta-information for parameters \\
+		dynplot & aut & Github & N/A & A simple visualisation library for trajectories \\
+		dynplot2 & aut & Github & N/A & A fully customisable visualisation library for trajectories \\
+		dyntoy & aut & Github & N/A & Generating simple toy data of cellular differentiation \\
+		dynutils & aut & CRAN & 7143 & Common functionality for the dynverse packages \\
+		dynwrap & aut & Github & N/A & A common format for trajectories \\
+		GillespieSSA & aut & CRAN & 5260 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		GillespieSSA2 & aut & CRAN & N/A & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
+		gng & aut & Github & N/A & An Rcpp implementation of the Growing Neural Gas algorithm \\
+		incgraph & aut & CRAN & 2863 & Incremental graphlet counting for network optimisation \\
+		princurve & aut & CRAN & 27'589 & Fits a principal curve in arbitrary dimension \\
+		proxyC & aut & CRAN & 3861 & Computes proximity in large sparse matrices \\
+		qsub & aut & CRAN & 3273 & Running commands remotely on gridengine clusters \\
+		SCORPIUS & aut & CRAN & 3448 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
+		Clus\-ter\-Sig\-nif\-i\-cance & ctb & Bioc & 1553 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
+		devtools & ctb & CRAN & 4'711'985 & Tools to make developing R packages easier \\
+		merlot & ctb & Github & N/A & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
+		mlr & ctb & CRAN & 145'913 & Machine Learning in R \\
+		monocle & ctb & Bioc & 38'555 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
+		ParamHelpers & ctb & CRAN & 125'471 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
+		pseudogp & ctb & Github & N/A & Probabilistic pseudotime for single-cell RNA-seq \\
+		remotes & ctb & CRAN & 2'019'623 & R package installation from remote repositories, including GitHub \\
+		rlang & ctb & CRAN & 9'838'123 & Functions for base types and core R and tidyverse features \\
+		SCope & ctb & Github & N/A & Visualization of large-scale and high dimensional single cell data \\
+		slingshot & ctb & Bioc & 3772 & Tools for ordering single-cell sequencing \\
+		splatter & ctb & Bioc & 5360 & Simple simulation of single-cell RNA sequencing data \\
+		URD & ctb & Github & N/A & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
+		wishbone & ctb & Github & N/A & Identify bifurcating developmental trajectories from single-cell data \\\hline
+	\end{tabularx}
+\end{table}

### 2019-08-23 ch1_introduction.tex
update

@@
 
+\newcommand{\cranpkg}[1]{\href{https://cran.r-project.org/package=#1}{#1}}
+\newcommand{\biocpkg}[1]{\href{https://bioconductor.org/packages/#1}{#1}}
+\newcommand{\biocpkgl}[2]{\href{https://bioconductor.org/packages/#1}{#2}}
+\newcommand{\githubpkg}[2]{\href{https://github.com/#1/#2}{#2}}
+
 \begin{table}[ht!]
@@
 		\hline
-		Name & Role & Repo & Downloads p.a. & Description \\ \hline\hline
-		babelwhale & aut & CRAN & N/A & Interacting with Docker and Singularity containers \\
-		dynbenchmark & aut & Github & N/A & Pipeline for benchmarking trajectory inference methods \\
-		dyndimred & aut & CRAN & 560 & Applying dimensionality reduction methods \\
-		dyneval & aut & Github & N/A & Evaluating trajectory inference methods \\
-		dynfeature & aut & Github & N/A & Calculating feature importance scores from trajectories \\
-		dyngen & aut & Github & N/A & Simulating single-cell data using gene regulatory networks \\
-		dynguidelines & aut & Github & N/A & User guidelines for trajectory inference \\
-		dynmethods & aut & Github & N/A & A collection of wrappers for trajectory inference methods \\
-		dyno & aut & Github & N/A & A pipeline for inferring, visualising and interpreting trajectories \\
-		dynparam & aut & CRAN & 790 & Creating meta-information for parameters \\
-		dynplot & aut & Github & N/A & A simple visualisation library for trajectories \\
-		dynplot2 & aut & Github & N/A & A fully customisable visualisation library for trajectories \\
-		dyntoy & aut & Github & N/A & Generating simple toy data of cellular differentiation \\
-		dynutils & aut & CRAN & 7143 & Common functionality for the dynverse packages \\
-		dynwrap & aut & Github & N/A & A common format for trajectories \\
-		GillespieSSA & aut & CRAN & 5260 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
-		GillespieSSA2 & aut & CRAN & N/A & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
-		gng & aut & Github & N/A & An Rcpp implementation of the Growing Neural Gas algorithm \\
-		incgraph & aut & CRAN & 2863 & Incremental graphlet counting for network optimisation \\
-		princurve & aut & CRAN & 27'589 & Fits a principal curve in arbitrary dimension \\
-		proxyC & aut & CRAN & 3861 & Computes proximity in large sparse matrices \\
-		qsub & aut & CRAN & 3273 & Running commands remotely on gridengine clusters \\
-		SCORPIUS & aut & CRAN & 3448 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
-		Clus\-ter\-Sig\-nif\-i\-cance & ctb & Bioc & 1553 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
-		devtools & ctb & CRAN & 4'711'985 & Tools to make developing R packages easier \\
-		merlot & ctb & Github & N/A & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
-		mlr & ctb & CRAN & 145'913 & Machine Learning in R \\
-		monocle & ctb & Bioc & 38'555 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
-		ParamHelpers & ctb & CRAN & 125'471 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
-		pseudogp & ctb & Github & N/A & Probabilistic pseudotime for single-cell RNA-seq \\
-		remotes & ctb & CRAN & 2'019'623 & R package installation from remote repositories, including GitHub \\
-		rlang & ctb & CRAN & 9'838'123 & Functions for base types and core R and tidyverse features \\
-		SCope & ctb & Github & N/A & Visualization of large-scale and high dimensional single cell data \\
-		slingshot & ctb & Bioc & 3772 & Tools for ordering single-cell sequencing \\
-		splatter & ctb & Bioc & 5360 & Simple simulation of single-cell RNA sequencing data \\
-		URD & ctb & Github & N/A & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
-		wishbone & ctb & Github & N/A & Identify bifurcating developmental trajectories from single-cell data \\\hline
+		Name & Role & Host & Downloads p.a. & Description \\ \hline\hline
+		\cranpkg{babelwhale} & aut & CRAN & N/A & Interacting with Docker and Singularity containers \\
+		\githubpkg{dynverse}{dynbenchmark} & aut & Github & N/A & Pipeline for benchmarking trajectory inference methods \\
+		\cranpkg{dyndimred} & aut & CRAN & 560 & Applying dimensionality reduction methods \\
+		\githubpkg{dynverse}{dyneval} & aut & Github & N/A & Evaluating trajectory inference methods \\
+		\githubpkg{dynverse}{dynfeature} & aut & Github & N/A & Calculating feature importance scores from trajectories \\
+		\githubpkg{dynverse}{dyngen} & aut & Github & N/A & Simulating single-cell data using gene regulatory networks \\
+		\githubpkg{dynverse}{dynguidelines} & aut & Github & N/A & User guidelines for trajectory inference \\
+		\githubpkg{dynverse}{dynmethods} & aut & Github & N/A & A collection of wrappers for trajectory inference methods \\
+		\githubpkg{dynverse}{dyno} & aut & Github & N/A & A pipeline for inferring, visualising and interpreting trajectories \\
+		\cranpkg{dynparam} & aut & CRAN & 790 & Creating meta-information for parameters \\
+		\githubpkg{dynverse}{dynplot} & aut & Github & N/A & A simple visualisation library for trajectories \\
+		\githubpkg{dynverse}{dynplot2} & aut & Github & N/A & A fully customisable visualisation library for trajectories \\
+		\githubpkg{dynverse}{dyntoy} & aut & Github & N/A & Generating simple toy data of cellular differentiation \\
+		\cranpkg{dynutils} & aut & CRAN & 7143 & Common functionality for the dynverse packages \\
+		\githubpkg{dynverse}{dynwrap} & aut & Github & N/A & A common format for trajectories \\
+		\cranpkg{GillespieSSA} & aut & CRAN & 5260 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		\cranpkg{GillespieSSA2} & aut & CRAN & N/A & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
+		\githubpkg{dynverse}{gng} & aut & Github & N/A & An Rcpp implementation of the Growing Neural Gas algorithm \\
+		\cranpkg{incgraph} & aut & CRAN & 2863 & Incremental graphlet counting for network optimisation \\
+		\cranpkg{princurve} & aut & CRAN & 27'589 & Fits a principal curve in arbitrary dimension \\
+		\cranpkg{proxyC} & aut & CRAN & 3861 & Computes proximity in large sparse matrices \\
+		\cranpkg{qsub} & aut & CRAN & 3273 & Running commands remotely on gridengine clusters \\
+		\cranpkg{SCORPIUS} & aut & CRAN & 3448 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
+		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} & ctb & Bioc & 1553 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
+		\cranpkg{devtools} & ctb & CRAN & 4'711'985 & Tools to make developing R packages easier \\
+		\githubpkg{soedinglab}{merlot} & ctb & Github & N/A & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
+		\cranpkg{mlr} & ctb & CRAN & 145'913 & Machine Learning in R \\
+		\biocpkg{monocle} & ctb & Bioc & 38'555 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
+		\cranpkg{ParamHelpers} & ctb & CRAN & 125'471 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
+		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & N/A & Probabilistic pseudotime for single-cell RNA-seq \\
+		\cranpkg{remotes} & ctb & CRAN & 2'019'623 & R package installation from remote repositories, including GitHub \\
+		\cranpkg{rlang} & ctb & CRAN & 9'838'123 & Functions for base types and core R and tidyverse features \\
+		\githubpkg{aertslab}{SCope} & ctb & Github & N/A & Visualization of large-scale and high dimensional single cell data \\
+		\biocpkg{slingshot} & ctb & Bioc & 3772 & Tools for ordering single-cell sequencing \\
+		\biocpkg{splatter} & ctb & Bioc & 5360 & Simple simulation of single-cell RNA sequencing data \\
+		\githubpkg{farrelja}{URD} & ctb & Github & N/A & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
+		\githubpkg{ManuSetty}{wishbone} & ctb & Github & N/A & Identify bifurcating developmental trajectories from single-cell data \\\hline
 	\end{tabularx}

### 2019-08-23 ch1_introduction.tex
update

--- 2019-08-23 ch1_introduction.tex
@@
 
+\newcommand{\cranpkg}[1]{\href{https://cran.r-project.org/package=#1}{#1}}
+\newcommand{\biocpkg}[1]{\href{https://bioconductor.org/packages/#1}{#1}}
+\newcommand{\biocpkgl}[2]{\href{https://bioconductor.org/packages/#1}{#2}}
+\newcommand{\githubpkg}[2]{\href{https://github.com/#1/#2}{#2}}
+\newcommand{\notavailable}{}
+
+%library(dlstats)
+%library(tidyverse)
+%
+%cran <- c("babelwhale", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "princurve", "proxyC", "qsub", "SCORPIUS", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
+%bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
+%
+%dl_cran <- dlstats::cran_stats(cran)
+%dl_bioc <- dlstats::bioc_stats(bioc)
+%
+%dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>% 
+%mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
+%
+%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / 23 * 365)
+%
+%summ <- dls %>% filter("2018-07-01" <= start, end <= "2019-07-30") %>% group_by(package) %>% summarise(downloads = sum(downloads))
+
 \begin{table}[ht!]
 	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the role within the package: aut Author, ctb Contributor.
-Yearly download statistics are based on the number of downloads between 2018-07-01 and 2019-06-30.
+Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-08-23.
 CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
-For Github repositories, no download statistics could be retrieved (N/A).
+For Github repositories, no download statistics could be retrieved.
 } \label{tab:packages}
@@
 	\centering\fontsize{9}{11}\selectfont
-	\begin{tabularx}{\linewidth}{|p{2cm}lllX|}
+	\begin{tabularx}{\linewidth}{|p{2cm}llp{1.5cm}X|}
 		\hline
-		Name & Role & Repo & Downloads p.a. & Description \\ \hline\hline
-		babelwhale & aut & CRAN & N/A & Interacting with Docker and Singularity containers \\
-		dynbenchmark & aut & Github & N/A & Pipeline for benchmarking trajectory inference methods \\
-		dyndimred & aut & CRAN & 560 & Applying dimensionality reduction methods \\
-		dyneval & aut & Github & N/A & Evaluating trajectory inference methods \\
-		dynfeature & aut & Github & N/A & Calculating feature importance scores from trajectories \\
-		dyngen & aut & Github & N/A & Simulating single-cell data using gene regulatory networks \\
-		dynguidelines & aut & Github & N/A & User guidelines for trajectory inference \\
-		dynmethods & aut & Github & N/A & A collection of wrappers for trajectory inference methods \\
-		dyno & aut & Github & N/A & A pipeline for inferring, visualising and interpreting trajectories \\
-		dynparam & aut & CRAN & 790 & Creating meta-information for parameters \\
-		dynplot & aut & Github & N/A & A simple visualisation library for trajectories \\
-		dynplot2 & aut & Github & N/A & A fully customisable visualisation library for trajectories \\
-		dyntoy & aut & Github & N/A & Generating simple toy data of cellular differentiation \\
-		dynutils & aut & CRAN & 7143 & Common functionality for the dynverse packages \\
-		dynwrap & aut & Github & N/A & A common format for trajectories \\
-		GillespieSSA & aut & CRAN & 5260 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
-		GillespieSSA2 & aut & CRAN & N/A & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
-		gng & aut & Github & N/A & An Rcpp implementation of the Growing Neural Gas algorithm \\
-		incgraph & aut & CRAN & 2863 & Incremental graphlet counting for network optimisation \\
-		princurve & aut & CRAN & 27'589 & Fits a principal curve in arbitrary dimension \\
-		proxyC & aut & CRAN & 3861 & Computes proximity in large sparse matrices \\
-		qsub & aut & CRAN & 3273 & Running commands remotely on gridengine clusters \\
-		SCORPIUS & aut & CRAN & 3448 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
-		Clus\-ter\-Sig\-nif\-i\-cance & ctb & Bioc & 1553 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
-		devtools & ctb & CRAN & 4'711'985 & Tools to make developing R packages easier \\
-		merlot & ctb & Github & N/A & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
-		mlr & ctb & CRAN & 145'913 & Machine Learning in R \\
-		monocle & ctb & Bioc & 38'555 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
-		ParamHelpers & ctb & CRAN & 125'471 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
-		pseudogp & ctb & Github & N/A & Probabilistic pseudotime for single-cell RNA-seq \\
-		remotes & ctb & CRAN & 2'019'623 & R package installation from remote repositories, including GitHub \\
-		rlang & ctb & CRAN & 9'838'123 & Functions for base types and core R and tidyverse features \\
-		SCope & ctb & Github & N/A & Visualization of large-scale and high dimensional single cell data \\
-		slingshot & ctb & Bioc & 3772 & Tools for ordering single-cell sequencing \\
-		splatter & ctb & Bioc & 5360 & Simple simulation of single-cell RNA sequencing data \\
-		URD & ctb & Github & N/A & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
-		wishbone & ctb & Github & N/A & Identify bifurcating developmental trajectories from single-cell data \\\hline
+		Name & Role & Host & Downloads per year & Description \\ \hline\hline
+		\cranpkg{babelwhale} & aut & CRAN & 3491 & Interacting with Docker and Singularity containers \\
+		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
+		\cranpkg{dyndimred} & aut & CRAN & 571 & Applying dimensionality reduction methods \\
+		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
+		\githubpkg{dynverse}{dynfeature} & aut & Github & \notavailable & Calculating feature importance scores from trajectories \\
+		\githubpkg{dynverse}{dyngen} & aut & Github & \notavailable & Simulating single-cell data using gene regulatory networks \\
+		\githubpkg{dynverse}{dynguidelines} & aut & Github & \notavailable & User guidelines for trajectory inference \\
+		\githubpkg{dynverse}{dynmethods} & aut & Github & \notavailable & A collection of wrappers for trajectory inference methods \\
+		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
+		\cranpkg{dynparam} & aut & CRAN & 2539 & Creating meta-information for parameters \\
+		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
+		\githubpkg{dynverse}{dynplot2} & aut & Github & \notavailable & A fully customisable visualisation library for trajectories \\
+		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
+		\cranpkg{dynutils} & aut & CRAN & 4570 & Common functionality for the dynverse packages \\
+		\githubpkg{dynverse}{dynwrap} & aut & Github & \notavailable & A common format for trajectories \\
+		\cranpkg{GillespieSSA} & aut & CRAN & 7855 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		\cranpkg{GillespieSSA2} & aut & CRAN & 2253 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
+		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & An Rcpp implementation of the Growing Neural Gas algorithm \\
+		\cranpkg{incgraph} & aut & CRAN & 2809 & Incremental graphlet counting for network optimisation \\
+		\cranpkg{princurve} & aut & CRAN & 24'439 & Fits a principal curve in arbitrary dimension \\
+		\cranpkg{proxyC} & aut & CRAN & 112'753 & Computes proximity in large sparse matrices \\
+		\cranpkg{qsub} & aut & CRAN & 2698 & Running commands remotely on gridengine clusters \\
+		\cranpkg{SCORPIUS} & aut & CRAN & 4824 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
+		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} & ctb & Bioc & 571 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
+		\cranpkg{devtools} & ctb & CRAN & 3'254'816 & Tools to make developing R packages easier \\
+		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
+		\cranpkg{mlr} & ctb & CRAN & 141'192 & Machine Learning in R \\
+		\biocpkg{monocle} & ctb & Bioc & 29'121 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
+		\cranpkg{ParamHelpers} & ctb & CRAN & 101'137 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
+		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
+		\cranpkg{remotes} & ctb & CRAN & 3'476'435 & R package installation from remote repositories, including GitHub \\
+		\cranpkg{rlang} & ctb & CRAN & 10'314'725 & Functions for base types and core R and tidyverse features \\
+		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
+		\biocpkg{slingshot} & ctb & Bioc & 9538 & Tools for ordering single-cell sequencing \\
+		\biocpkg{splatter} & ctb & Bioc & 2999 & Simple simulation of single-cell RNA sequencing data \\
+		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
+		\githubpkg{ManuSetty}{wishbone} & ctb & Github & \notavailable & Identify bifurcating developmental trajectories from single-cell data \\\hline
 	\end{tabularx}

### 2019-08-23 ch2_dyngen.tex
update

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyngen: Simulating single-cell data using gene regulatory networks} 
\chaptermark{dyngen: Simulating single-cell data using gene regulatory networks} % short title
\label{chap:dyngen}

\begin{quote}
	\textbf{Abstract:} 
\end{quote}

\vfill

Adapted from:\\

\newpage

\begin{figure}
	\centering
	\includegraphics[width=\LARGEfigure]{fig/dyngen/showcase_2.png} 
	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.}} % TODO: update label
	\label{fig:dyngen}
\end{figure}

### 2019-08-23 fig/dyngen/script.R

@@ -1,0 +1,124 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(10)
+
+backbone <- bblego(
+  bblego_start("A", type = "simple", num_modules = 1),
+  bblego_linear("A", "B", type = "simple", num_modules = 3),
+  bblego_branching("B", c("C", "D")),
+  bblego_end("C", type = "simple", num_modules = 2),
+  bblego_end("D", type = "simple", num_modules = 2)
+)
+model <-
+  initialise_model(
+    num_tfs = 12,
+    num_targets = 30,
+    num_hks = 15,
+    num_cells = 150,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8
+
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+## grn
+plot_backbone_statenet(model) %>% ggsave(filename = "dyngen/statenet.pdf", width = 6, height = 6)
+plot_backbone_modulenet(model) %>% ggsave(filename = "dyngen/modulenet.pdf", width = 8, height = 6)
+
+plot_feature_network(model, show_targets = FALSE)
+
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE)
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = "dyngen/featnet.pdf", width = 8, height = 6)
+plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = "dyngen/featnet_onlytfs.pdf", width = 8, height = 6)
+
+## simulate
+model2 <- model %>%
+  generate_gold_standard() %>%
+  generate_cells() %>%
+  generate_experiment()
+
+g <- plot_gold_simulations(model2) + scale_colour_brewer(palette = "Dark2")
+ggsave("dyngen/gold_simulations.pdf", g, width = 8, height = 6)
+
+g <- plot_gold_expression(model2, what = "x") # mrna
+ggsave("dyngen/gold_mrna.pdf", g, width = 8, height = 6)
+g <- plot_gold_expression(model2, label_changing = FALSE) # premrna, mrna, and protein
+ggsave("dyngen/gold_expression.pdf", g, width = 8, height = 6)
+
[…]

### 2019-08-24 ch1_introduction.tex
update

@@
 \chapter{Introduction} 
-\chaptermark{Introduction.} % short title
+\chaptermark{Introduction} % short title
 \label{chap:introduction}
@@
 \end{figure}
+
 \subsection{The Central Dogma}
@@
 In turn, its structure determines the functionality of the protein, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
+
 \subsection{Cell types}
@@
 \section{Outline} % include contributions?
+% mention focus on datasets which contain developing cells.
+
 This work aims to tackle multiple computational problems in extracting biological hypotheses from single-cell omics data.
+(TODO: rephrase that we focus on datasets that contain developing cells, and that the tools we develop make this assumption.)
 We introduce a framework for evaluating multiple types of single-cell omics tools, more specifically dimensionality reduction, trajectory inference, and gene regulatory network inference (Chapter \ref{chap:dyngen}).

### 2019-08-24 ch2_dyngen.tex
update

@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dyngen: Simulating single-cell data using gene regulatory networks} 
-\chaptermark{dyngen: Simulating single-cell data using gene regulatory networks} % short title
+\chapter{dyngen: simulating single cells} 
+\chaptermark{dyngen: simulating single cells}
 \label{chap:dyngen}
@@
 
+\section{Introduction}
+Continuous technological advancements to high-throughput profiling of single cells
+are having profound effects on how researchers can validate biological hypotheses. 
+For example, single-cell RNA sequencing (scRNA-seq) directly resulted in the development
+of a new type of computational method called trajectory inference (TI). By profiling
+the transcriptomics profiles of developing cells, TI methods attempt to reconstruct 
+and characterise the underlying dynamic processes \cite{cannoodt_computationalmethodstrajectory_2016}.
+While early experimental technologies allowed to profile one single modality (e.g. DNA sequence, 
+RNA or protein expression), recent developments permit profiling multiple modalities simultaneously.
+
+An ideal experiment would be able to observe all aspects of a cell, including a full history of its 
+molecular states, spatial positions and environmental interactions \cite{stuart_integrativesinglecellanalysis_2019}. 
+While this falls outside the reach of current experimental technologies, \textit{in silico} simulations
+of single cells would allow developing the next wave of computational techniques
+in anticipation of new experimental technologies.
+
+A few generators of scRNA-seq profiles have already been developed (e.g.
+splatter \cite{zappia_splattersimulationsinglecell_2017} and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}).
+These can be
+used to evaluate the performance of computational tools, and to explore their strengths and weaknesses.
+A limitation of directly simulating a scRNA-seq profile (instead of a single cell) is that extending the simulation to other aspects of the cell -- such as tracking the full history of molecular states -- becomes difficult.
+
+\section{Results}
+
+dyngen is a simulator for single cells that develop over time.
+Each cell contains an underlying gene regulatory network (GRN) that specifies which chemical reactions can occur at any given point in time, depending on the availability of required molecules (Figure \ref{fig:dyngen}A).
+It was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
+One main improvement is being able to profile multiple aspects of a single cell at any given time point: its molecule abundance levels, which reactions took place, and which parts of the gene regulatory network were active.
+The underlying GRN allows dyngen to be used in a variety of experimental scenarios in order to evaluate many different types of computational tools (Figure \ref{fig:dyngen}B).
+
 \begin{figure}
 	\centering
-	\includegraphics[width=\LARGEfigure]{fig/dyngen/showcase_2.png} 
+	\includegraphics[width=\LARGEfigure]{fig/dyngen/showcase_3.png} 
 	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.}} % TODO: update label
@@
 \end{figure}
+
+\subsection{Example simulation}
+
+\subsection{Applications}
+
+\subsection{Supported backbones}
+
+\subsection{Backbone lego}
+
+\subsection{Additional experimental effects}
+% time series, batch effects, perturbation experiments
+
+\section{Discussion}
+
+\section{Methods}
+
+%dyntoy and PROSSTT are specifically developed to generate datasets
+% containing single cells which develop along a certain trajectory.
+% These generators start a certain topology in mind, 
+% simulate changing gene expressions along each of the branches of the topology, and sample cells
+% from random points in the trajectory. splatter is focused on well replicating the specific noise characteristics of
+% scRNA-seq data, and simulates differentially expressed genes in order to replicate clusters of cells, batch effects
+% and even trajectories. dyngen 1.0 borrows a page from GeneNetWeaver in that it also converts a GRN into an ODE, but
+% the GRN is constructed in such a way that the repeated simulations resemble single cells following a regulatory program
+% (e.g. differentiation into one of two celltypes). BoolODE uses the dyngen 1.0 GRNs but uses boolean models to convert 
+% the network into an ODE, and is used to evaluate the performance of NI methods, not TI methods.
+
+%\begin{table}
+%	\caption{\textbf{Comparison of existing single-cell omics simulators.} TI: trajectory inference, NI: network inference, Cl: clustering, DR: dimensionality reduction.} \label{tab:simulators}
+%	\centering\fontsize{9}{11}\selectfont
+%	\begin{tabularx}{\linewidth}{p{1.6cm}p{2.5cm}p{3cm}X}
+%		Generator & Used to evaluate & Data type(s) & Approach \\ \hline
+%		dyngen 1.0 & TI & mRNA \& protein & GRN, ODE, Euler--Maramuya \\
+%		dyntoy & TI & mRNA & Visualise tree in plane, generate expression in plane, sample cells from tree \\
+%		PROSSTT & TI & mRNA & Random walks along tree, sample cells from tree \\
+%		splatter & Cl \& TI & mRNA & Simulate DE genes, simulate scRNA-seq noise \\
+%		BoolODE & NI & mRNA \& protein & Use dyngen GRN's, ODE, Euler--Maramuya \\
+%		dyngen 2.0 & TI, NI, Cl, DR & pre-mRNA, mRNA, protein & Backbone, GRN, SSA \\
+%		dyngen 3? & TI, NI, Cl, DR & pre-mRNA, mRNA, miRNA, protein, protein complex & Backbone, GRN, SSA
+%	\end{tabularx}
+%\end{table}

### 2019-08-24 fig/dyngen/script.R

@@ -19,8 +19,9 @@ model <-
     backbone = backbone,
     verbose = TRUE,
     download_cache_dir = "~/.cache/dyngen",
-    num_cores = 8
-
+    num_cores = 8,
+    distance_metric = "euclidean",
+    simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
   ) %>%
   generate_tf_network() %>%
   generate_feature_network() %>%
@@ -44,6 +45,11 @@ model2 <- model %>%
   generate_cells() %>%
   generate_experiment()
 
+# recalculate dimred
+space <- prcomp(model2$simulations$counts %>% as.matrix, rank. = 2)$x
+colnames(space) <- c("comp_1", "comp_2")
+model2$simulations$dimred <- space
+
 g <- plot_gold_simulations(model2) + scale_colour_brewer(palette = "Dark2")
 ggsave("dyngen/gold_simulations.pdf", g, width = 8, height = 6)
 
@@ -52,7 +58,7 @@ ggsave("dyngen/gold_mrna.pdf", g, width = 8, height = 6)
 g <- plot_gold_expression(model2, label_changing = FALSE) # premrna, mrna, and protein
 ggsave("dyngen/gold_expression.pdf", g, width = 8, height = 6)
 
-g <- plot_simulations(model2)
+g <- plot_simulations(model2) + theme_classic() + theme(axis.ticks = element_blank(), axis.line = element_blank(), axis.text = element_blank(), axis.title = element_blank()) + labs(colour = "Simulation\ntime")
 ggsave("dyngen/simulations.pdf", g, width = 8, height = 6)
 
 g <- plot_gold_mappings(model2, do_facet = FALSE) + scale_colour_brewer(palette = "Dark2")
@@ -122,3 +128,4 @@ g <- ggplot(as.data.frame(dimred)) + geom_point(aes(comp_1, comp_2)) + coord_equ
   theme(axis.text = element_blank(), axis.ticks = element_blank()) +
   labs(x = "Comp 1", y = "Comp 2")
 ggsave("dyngen/dimred.pdf", g, width = 8, height = 6)
+

### 2019-08-25 ch2_dyngen.tex
update

@@
 
-\section{Results}
-
-dyngen is a simulator for single cells that develop over time.
-Each cell contains an underlying gene regulatory network (GRN) that specifies which chemical reactions can occur at any given point in time, depending on the availability of required molecules (Figure \ref{fig:dyngen}A).
-It was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
-One main improvement is being able to profile multiple aspects of a single cell at any given time point: its molecule abundance levels, which reactions took place, and which parts of the gene regulatory network were active.
-The underlying GRN allows dyngen to be used in a variety of experimental scenarios in order to evaluate many different types of computational tools (Figure \ref{fig:dyngen}B).
+We introduce dyngen, a multi-modality simulator of single cells (Figure \ref{fig:dyngen}).
+dyngen was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
+We demonstrate its flexibility by simulating numerous different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
 
@@
 	\centering
-	\includegraphics[width=\LARGEfigure]{fig/dyngen/showcase_3.png} 
-	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.}} % TODO: update label
+	\includegraphics[width=\linewidth]{fig/dyngen/showcase_3.png} 
+	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.} Remove B?} % TODO: update label
 	\label{fig:dyngen}
@@
 
-\subsection{Example simulation}
+\section{Results}
+dyngen is a simulator for single cells that develop over time.
+Throughout this section, a simple simulation of a cell undergoing a cyclic process is used to illustrate key strengths of dyngen (Figure \ref{fig:simplecyclic}).
+This example only comprises of a single cell containing 5 genes, but dyngen can easily scale up to thousands of simulations containing thousands of genes.
+
+In dyngen, a cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A).
+These reactions are determined from a predefined set of gene regulatory interactions (Figure \ref{fig:simplecyclic}B), henceforth referred to as a gene regulatory network (GRN).
+The likelihood of a reaction occurring at any given point in time is defined by the GRN and by the abundance of molecules involved each reaction.
+
+One of dyngen's main advantages is that through careful engineering of the GRN, different cellular developmental processes can be obtained.
+Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies.
+Multiple simulations with slightly different GRNs can emulate rewiring events in disease or perturbation experiments.
+% could use figure; one with GRNs of different topologies, another with rewiring events
+Multiple simulations with different initial molecule abundance levels can be used to replicate batch effects.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{fig/dyngen/simplecyclic_edited} 
+	\caption{Showcase of dyngen functionality.}
+	\label{fig:simplecyclic}
+\end{figure}
+
+Another advantage is that dyngen returns many modalities throughout the whole simulation: molecular abundance (Figure \ref{fig:simplecyclic}C), cellular state (\ref{fig:simplecyclic}D), number of reaction firings (\ref{fig:simplecyclic}E), reaction likelihoods (not shown), and regulation activations (\ref{fig:simplecyclic}F).
+These modalities can serve as input data and ground truth for benchmarking many types of computational approaches, including clustering, batch effect correction, dimensionality reduction, trajectory inference, trajectory alignment, and gene regulatory network inference.
+
+The final main advantage is that by making alterations to the generation pipeline, multiple types of experiments can be simulated.
+The default settings assumes an asynchronous dynamic process and thus cells are sampled uniformly.
+A time-series experiment of a synchonised dynamic process can be simulated by sampling cells from various intervals of the simulation.
+% TODO: you can even profile the same cell at regular intervals, which is a technology that does not exist yet at a single cell omics scale.
+This could be used to evaluate, for example, RNA velocity approaches.
+
+%\subsection{Example simulation}
+%
+%
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=\linewidth]{fig/dyngen/simplecyclic_edited} 
+%	\caption{Simple cyclic} % TODO: update label
+%	\label{fig:simplecyclic}
+%\end{figure}
 
@@
 \section{Discussion}
+% cell cycle, mitosis
+% cell-cell interaction
+% spatial + cell movement
+% transcriptional bursting
+% add mechanism for triggering reactions only when they are needed?
 
 \section{Methods}
+
+\subsection{Simulating a snapshot experiment with dyngen}
+\subsubsection{Define the module network and the state network}
+\subsubsection{Generate the transcription factor network}
+\subsubsection{Generate targets and housekeeping genes}
+\subsubsection{Convert gene regulatory network to a set of SSA reactions}
+\subsubsection{Determine the gold standard}
+\subsubsection{Running SSA simulations and mapping it to the gold standard}
+\subsubsection{Simulate snapshot experiment}
+\subsection{Predefined backbones}
+
+\subsection{Other types of experiments}
+\subsubsection{Time series experiment}
+\subsubsection{Perturbation experiment}
+\subsubsection{Batch effects}
+\subsection{Evaluating computational tools}
+\subsubsection{Clustering}
+\subsubsection{Dimensionality reduction}
+\subsubsection{Batch effect correction}
+\subsubsection{Trajectory inference}
+\subsubsection{Trajectory alignment}
+\subsubsection{Network inference}
 

### 2019-08-25 fig/dyngen/script.R

@@ -50,6 +50,8 @@ space <- prcomp(model2$simulations$counts %>% as.matrix, rank. = 2)$x
 colnames(space) <- c("comp_1", "comp_2")
 model2$simulations$dimred <- space
 
+write_rds(model2, "dyngen/model.rds", compress = "gzip")
+
 g <- plot_gold_simulations(model2) + scale_colour_brewer(palette = "Dark2")
 ggsave("dyngen/gold_simulations.pdf", g, width = 8, height = 6)
 
@@ -73,7 +75,7 @@ dataset <- wrap_dataset(model2)
 library(dynplot)
 g <- plot_dimred(dataset)
 ggsave("dyngen/traj_dimred.pdf", g, width = 8, height = 6)
-plot_graph(dataset)
+g <- plot_graph(dataset)
 ggsave("dyngen/traj_graph.pdf", g, width = 8, height = 6)
 g <- plot_heatmap(dataset, features_oi = 40)
 ggsave("dyngen/traj_heatmap.pdf", g, width = 8, height = 6)

### 2019-08-25 fig/dyngen/simple.R

@@ -1,0 +1,200 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(1)
+
+backbone <- backbone(
+  module_info = tribble(
+    ~module_id, ~a0, ~burn,
+    "A", 1, TRUE,
+    "B", 1, TRUE,
+    "C", 1, TRUE,
+    "D", 0, TRUE,
+    "E", 0, TRUE
+  ),
+  module_network = tribble(
+    ~from, ~to, ~effect, ~strength, ~cooperativity,
+    "A", "B", -1, 4, 2,
+    "B", "C", -1, 4, 2,
+    "C", "D", 1, 1, 2,
+    "D", "E", 1, 1, 2,
+    "E", "A", -1, 4, 2
+  ),
+  expression_patterns = tribble(
+    ~from, ~to, ~module_progression, ~start, ~burn, ~time,
+    "Burn0", "S1", "+A,+B,+C", TRUE, TRUE, 3,
+    "S1", "S2", "+D,+E,-C", FALSE, FALSE, 4,
+    "S2", "S3", "+C,-A,-B", FALSE, FALSE, 4,
+    "S3", "S1", "+A,+B,-D,-E", FALSE, FALSE, 4
+  )
+)
+census_interval <- .1
+total_time <- 10
+model <-
+  initialise_model(
+    num_tfs = nrow(backbone$module_info) * 1,
+    num_targets = 0,
+    num_hks = 0,
+    num_cells = 300,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    kinetics_params = kinetics_default(
+      sample_wpr = function(n) 100,
+      sample_spl = function(n) 10,
+      sample_xdr = function(n) 5,
+      sample_ypr = function(n) 5,
+      sample_ydr = function(n) 5
+    ),
+    simulation_params = simulation_default(
+      burn_time = 2,
+      total_time = total_time,
+      ssa_algorithm = GillespieSSA2::ssa_exact(),
+      num_simulations = 1,
+      census_interval = census_interval,
+      store_reactions = TRUE
+    )
+  )
+
+model <- model %>%
+  generate_tf_network() %>%
+  generate_feature_network()
+
+notf <- function(x) gsub("_TF1$", "", x)
+model$feature_info <- model$feature_info %>% mutate_at(vars(feature_id), notf)
[…]

### 2019-08-26 ch2_dyngen.tex
update

@@
 splatter \cite{zappia_splattersimulationsinglecell_2017} and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}).
-These can be
-used to evaluate the performance of computational tools, and to explore their strengths and weaknesses.
+% BoolODE and dyntoy are also examples.
+These can be used to evaluate the performance of computational tools, and to explore their strengths and weaknesses.
 A limitation of directly simulating a scRNA-seq profile (instead of a single cell) is that extending the simulation to other aspects of the cell -- such as tracking the full history of molecular states -- becomes difficult.
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{fig/dyngen/simplecyclic_edited} 
+	\caption{Showcase of dyngen functionality.
+A time resolution of 0.1 was used, but this can be increased or decreased without effect on performance of the execution of the simulation.
+\textbf{TODO: perhaps it's better to replace Figure \ref{fig:simplecyclic} with one subfigure for each of the paragraphs in this text.}}
+	\label{fig:simplecyclic}
+\end{figure}
+
 In dyngen, a cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A).
@@
 
-\begin{figure}
-	\centering
-	\includegraphics[width=\linewidth]{fig/dyngen/simplecyclic_edited} 
-	\caption{Showcase of dyngen functionality.}
-	\label{fig:simplecyclic}
-\end{figure}
+Another advantage is that dyngen returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F).
+These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches.
+For example, a network inference method could use mRNA abundance and cellular states as inputs, and its output could be benchmarked against the gold standard GRN.
 
-Another advantage is that dyngen returns many modalities throughout the whole simulation: molecular abundance (Figure \ref{fig:simplecyclic}C), cellular state (\ref{fig:simplecyclic}D), number of reaction firings (\ref{fig:simplecyclic}E), reaction likelihoods (not shown), and regulation activations (\ref{fig:simplecyclic}F).
-These modalities can serve as input data and ground truth for benchmarking many types of computational approaches, including clustering, batch effect correction, dimensionality reduction, trajectory inference, trajectory alignment, and gene regulatory network inference.
-
-The final main advantage is that by making alterations to the generation pipeline, multiple types of experiments can be simulated.
-The default settings assumes an asynchronous dynamic process and thus cells are sampled uniformly.
-A time-series experiment of a synchonised dynamic process can be simulated by sampling cells from various intervals of the simulation.
-% TODO: you can even profile the same cell at regular intervals, which is a technology that does not exist yet at a single cell omics scale.
-This could be used to evaluate, for example, RNA velocity approaches.
-
-%\subsection{Example simulation}
-%
-%
-%\begin{figure}
-%	\centering
-%	\includegraphics[width=\linewidth]{fig/dyngen/simplecyclic_edited} 
-%	\caption{Simple cyclic} % TODO: update label
-%	\label{fig:simplecyclic}
-%\end{figure}
-
-\subsection{Applications}
-
-\subsection{Supported backbones}
-
-\subsection{Backbone lego}
-
-\subsection{Additional experimental effects}
-% time series, batch effects, perturbation experiments
+The final main advantage is that by making alterations to the simulation pipeline, multiple types of experiments (sampling technique or profiling technique) can be simulated.
+By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
+It is possible to implement other experimental protocols (which perhaps do not exist in real life), such as sampling the same cell at regular intervals.
 
 \section{Discussion}
-% cell cycle, mitosis
-% cell-cell interaction
-% spatial + cell movement
-% transcriptional bursting
-% add mechanism for triggering reactions only when they are needed?
+As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
+However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications; several examples are given.
+
+Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
+Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
+Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+
+dyngen ultimately also allows anticipating technological developments in single-cell multi-omics.
+In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
+Similarly, it could also be used to compare which experimental technique will likely produce the most accurate result.
+For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knock down experiments a necessity in order to infer high quality regulatory networks?
+
+Currently, dyngen focuses on simulating cells as standalone entities.
+Future developments include extending the framework to simulate multiple cells in a virtual environment.
+Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.
 
@@
 \subsubsection{Simulate snapshot experiment}
-\subsection{Predefined backbones}
 
-\subsection{Other types of experiments}
+\subsection{Adaptations}
+\subsubsection{Predefined backbones}
+\subsubsection{Backbone lego}
 \subsubsection{Time series experiment}
@@
 \subsubsection{Batch effects}
-\subsection{Evaluating computational tools}
-\subsubsection{Clustering}
-\subsubsection{Dimensionality reduction}
-\subsubsection{Batch effect correction}
-\subsubsection{Trajectory inference}
+
+\subsection{Example use cases}
 \subsubsection{Trajectory alignment}
-\subsubsection{Network inference}
+From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
+\subsubsection{Differential network inference}
+From discussion: Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
+\subsubsection{RNA velocity}
+From discussion: Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
 

### 2019-08-26 fig/dyngen/simple.R

@@ -53,7 +53,8 @@ model <-
       ssa_algorithm = GillespieSSA2::ssa_exact(),
       num_simulations = 1,
       census_interval = census_interval,
-      store_reactions = TRUE
+      store_reaction_firings = TRUE,
+      store_reaction_propensities = TRUE
     )
   )
 
@@ -89,7 +90,7 @@ expr_df <- data.frame(time = time, model$simulations$counts %>% as.matrix()) %>%
     molecule = factor(mol_map[gsub("_.*", "", var)], levels = mol_map)
   ) %>%
   filter(time > 0)
-firings_df <- data.frame(time = time, model$simulations$reactions %>% as.matrix()) %>%
+firings_df <- data.frame(time = time, model$simulations$reaction_firings %>% as.matrix()) %>%
   gather(var, value, -time) %>%
   mutate(
     gene = paste0("Gene ", gsub(".*_", "", var)),
@@ -108,6 +109,13 @@ reg_df <- data.frame(time = time, model$simulations$regulation %>% as.matrix()) 
 state_df <- model$simulations$meta %>%
   mutate(cell_id = paste0("cell_", row_number())) %>%
   filter(sim_time > 0)
+prop_df <- data.frame(time = time, model$simulations$reaction_propensities %>% as.matrix()) %>%
+  gather(var, value, -time) %>%
+  mutate(
+    gene = paste0("Gene ", gsub(".*_", "", var)),
+    reaction = factor(reac_map[sub("_[^_]*$", "", var)], levels = reac_map)
+  ) %>%
+  filter(time > 0)
 
 dummy <-
   dynwrap::wrap_data(
@@ -122,6 +130,9 @@ state_df <- state_df %>%
   full_join(crossing(sim_time = state_df$sim_time, milestone_id = dummy$milestone_ids), by = c("milestone_id", "sim_time")) %>%
   mutate(percentage = ifelse(is.na(percentage), 0, percentage))
 
+#######
+# ABUNDANCE LEVELS
+#######
 max_expr <- max(expr_df$value)
 max_expr <- ceiling(max_expr / 10) * 10
 g1 <- ggplot(expr_df, aes(time, value)) +
@@ -137,42 +148,13 @@ g1 <- ggplot(expr_df, aes(time, value)) +
     legend.margin = margin()
   ) +
   scale_x_continuous(breaks = c(0, 5, 10)) +
[…]

### 2019-08-26 boek.tex
update
refactorise project

@@
-\documentclass[11pt,hidelinks]{book}
+\documentclass[10pt]{book}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-
-% Use Unicode characters when possible
-%\usepackage{xltxtra}
-
-%\usepackage[default,scale=0.95]{opensans}
-%\usepackage[utf8]{inputenc}
-%\usepackage[T1]{fontenc}
 
@@
 % scripting packages
-\usepackage{ifthen}
 \usepackage{blindtext}
-\usepackage{todonotes}
 
@@
 \usepackage{fancyhdr}
-%\usepackage[]{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
+\usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
 
@@
 \bibliography{library}
-%
-%% substitute in references
-%\DeclareDatafieldSet{setall}{
-%	\member[datatype=literal]
-%	\member[datatype=name]
-%	\member[field=journal]% journal is special since it is
-%	% actually journaltitle
-%}
-%
-%\DeclareSourcemap{
-%	\maps[datatype=bibtex]{
-%		\map[overwrite, foreach={setall}]{
-%			% \`{\i}
-%			\step[fieldsource=\regexp{$MAPLOOP},
-%			match=\regexp{\x{0131}\x{0300}},
-%			replace=\regexp{\x{00EC}}]
-%			% \'{\i}
-%			\step[fieldsource=\regexp{$MAPLOOP},
-%			match=\regexp{\x{0131}\x{0301}},
-%			replace=\regexp{\x{00ED}}]
-%			% \^{\i}
-%			\step[fieldsource=\regexp{$MAPLOOP},
-%			match=\regexp{\x{0131}\x{0302}},
-%			replace=\regexp{\x{00EE}}]
-%			% \"{\i}
-%			\step[fieldsource=\regexp{$MAPLOOP},
-%			match=\regexp{\x{0131}\x{0308}},
-%			replace=\regexp{\x{00EF}}]
-%		}
-%	}
-%}
+\usepackage{bibentry}
+\nobibliography*
 
@@
 
+\definecolor{light-gray}{gray}{0.90}
+\definecolor{med-gray}{gray}{0.60}
 \tikzset{
@@
 
-%\newcommand{\picsubf}[2]{
-%	\begin{subfigure}[b]{#2}
-%		\fbox{\includegraphics[width=\linewidth]{#1}}
-%\end{subfigure}}
-%
-%\newcommand{\picsubfc}[4]{
-%	\begin{subfigure}[b]{#4}
-%		\fbox{\includegraphics[width=\linewidth]{#1}}
-%		\caption{#2}
-%		\label{#3}
-%\end{subfigure}}
-%
-%\newcommand{\picsub}[2]{
-%	\begin{subfigure}[b]{#2}
-%		\includegraphics[width=\linewidth]{#1}
-%\end{subfigure}}
-%
-%\newcommand{\picsubc}[4]{
-%	\begin{subfigure}[b]{#4}
-%		\includegraphics[width=\linewidth]{#1}
-%		\caption{#2}
-%		\label{#3}
-%\end{subfigure}}
-
 %%%%%%%%%%%%%%%%%%%%%
@@
 
-\definecolor{light-gray}{gray}{0.90} %not always visible on a dell screen!!!
+% set font after chapter sonny style
+\usepackage{fontspec}
+\setmainfont [Path = fonts/,
+UprightFont = *-300,
+ItalicFont = *-300-Italic,
+BoldFont = *-700,
+BoldItalicFont = *-700-Italic
+]{MuseoSans}
 
@@
 \usepackage{hyperref}
+
+\hypersetup{
+	colorlinks = true,
+%	linkcolor = {blue},
+  urlcolor = {blue},
+	citecolor = {med-gray}
+}
 
@@
 \renewcommand{\baselinestretch}{1.25}
-
-\usepackage{fontspec}
-
-\setmainfont [Path = fonts/,
-UprightFont = *-300,
-ItalicFont = *-300-Italic,
-BoldFont = *-700,
-BoldItalicFont = *-700-Italic
-]{MuseoSans}
 
@@
 
-\include{ch1_introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%     CH1: INTRODUCTION     %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{Introduction} 
+\chaptermark{Introduction} % short title
+\label{chap:introduction}
+\graphicspath{{./ch1_introduction/}}
+\lettergroup{\thechapter}
+\begin{quote}
+	\textbf{Abstract:} 
+	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner.
+Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following.
+In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods.
+We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation.
+As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
+\end{quote}
+\vfill
+Partially adapted from:\\
+\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y.
+Computational methods for trajectory inference from single-cell transcriptomics.
+\textit{European Journal of Immunology} 46, 11 (2016), 2496--2506.
+\doi{10.1002/eji.201646347}.\\
+Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
+A comparison of single-cell trajectory inference methods.
+\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
+\doi{10.1038/s41587-019-0071-9}.\\
+Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y.
+Network Inference from Single-Cell Transcriptomic Data.
+\textit{Gene Regulatory Networks} (2019), 235--249.
+\doi{10.1007/978-1-4939-8882-2_10}.
+\\
+{\footnotesize $^*$ Equal contribution}
+\newpage
+\include{ch1_introduction/content}
 
-% landmark mds
-% princurve
-% GillespieSSA
-% dyngen
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%        CH2: DYNGEN        %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{dyngen: simulating single cells} 
+\chaptermark{dyngen: simulating single cells}
+\label{chap:dyngen}
+\graphicspath{{./ch2_dyngen/}}
+\include{ch2_dyngen/content}
 
-\include{ch2_dynbenchmark}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%     CH3: DYNBENCHMARK     %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
+\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
+\label{chap:dynbenchmark}
+\graphicspath{{./ch3_dynbenchmark/}}
+\begin{quote}
+	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
+Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
+Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
+Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
+Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
+Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
+\end{quote}
+\vfill
+Adapted from:\\
+Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
+A comparison of single-cell trajectory inference methods.
+\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
+\doi{10.1038/s41587-019-0071-9}.\\
+{\footnotesize $^*$ Equal contribution}
+\newpage
+\include{ch3_dynbenchmark/content}
 
-\include{ch3_scorpius}
-
-% include gng ?
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%         CH4: DYNO         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \include{ch4_dyno}
 
-\include{ch5_brood}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%       CH5: SCORPIUS       %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\include{ch5_scorpius}
 
-\include{ch6_incgraph}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%         CH6: BRED         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\include{ch6_brood}
 
-\include{ch7_discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%       CH7: INCGRAPH       %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{incgraph: Optimising regulatory networks} 
+\chaptermark{incgraph: Optimising regulatory networks.} % short title
+\label{chap:incgraph}
+\graphicspath{{./ch7_incgraph/}}
+\begin{quote}
+	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
+	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
+	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
+\\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
+	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
+	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
+	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
+\\
+	IncGraph is freely available as an open-source R package on CRAN (incgraph). The development version is also available on GitHub (\href{http://www.github.com/rcannood/incgraph}{rcannood/incgraph}).
+\end{quote}
+\vfill
+Adapted from:\\
+\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y.
+IncGraph: Incremental graphlet counting for topology optimisation.
+\textit{PLOS ONE} 13, 4 (2018), e0195997.
+\doi{10.1371/journal.pone.0195997}.
+\newpage
+\include{ch7_incgraph/content}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%      CH8: DISCUSSION      %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\include{ch8_discussion}
 

### 2019-08-26 ch1_introduction/content.tex
refactorise project

--- 2019-08-24 ch1_introduction.tex
@@
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Introduction} 
-\chaptermark{Introduction} % short title
-\label{chap:introduction}
-\lettergroup{\thechapter}
-
-\begin{quote}
-	\textbf{Abstract:}
-
-	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner.
-Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following.
-In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods.
-We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation.
-As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
-\end{quote}
-
-\vfill
-
-Partially adapted from:\\
-\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y.
-Computational methods for trajectory inference from single-cell transcriptomics.
-\textit{European Journal of Immunology} 46, 11 (2016), 2496--2506.
-\doi{10.1002/eji.201646347}.\\
-Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
-A comparison of single-cell trajectory inference methods.
-\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
-\doi{10.1038/s41587-019-0071-9}.\\
-Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y.
-Network Inference from Single-Cell Transcriptomic Data.
-\textit{Gene Regulatory Networks} (2019), 235--249.
-\doi{10.1007/978-1-4939-8882-2_10}.
-\\
-{\footnotesize $^*$ Equal contribution}
-
-\newpage
-
 \section{The cell}
@@
 \begin{figure}
-	\centering\includegraphics[width=.5\linewidth]{fig/introduction/fsvm3} % TODO: change image
+	\centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=.5\linewidth]{fig/introduction/rnaworld} % TODO: change image
+	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
 	\caption{
@@
 	\centering
-	A. \includegraphics[width=.45\linewidth]{fig/introduction/profiling_convergence} 
+	A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
 	% also figure with cell numbers
-	C. \includegraphics[width=.45\linewidth]{fig/introduction/multimodal} 
+	C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
 	% TODO: change image
@@
 \begin{figure}
-	\centering\includegraphics[width=.5\linewidth]{fig/introduction/singlecell_tools}
-	\centering\includegraphics[width=.3\linewidth]{fig/introduction/scrnatools}
+	\centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
+	\centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
 	% TODO: change image
@@
 \begin{figure}
-	\centering\includegraphics[width=\Largefigure]{fig/introduction/trajectory_inference_slim} % TODO: Update figure to match text!!!!
+	\centering\includegraphics[width=\Largefigure]{fig/trajectory_inference_slim} % TODO: Update figure to match text!!!!
 	\caption{
@@
 	\centering
-	\includegraphics[width=\Largefigure]{fig/introduction/ni_bulk_vs_sc}
+	\includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
 	\caption{Bulk expression data return the average expressions of genes among large numbers of cells.

### 2019-08-26 ch3_dynbenchmark/content.tex
refactorise project

--- 2019-08-21 ch4_dynbenchmark.tex
@@
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
-\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
-\label{chap:dynbenchmark}
-
-\begin{quote}
-	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
-Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
-Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
-Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
-Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
-Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
-\end{quote}
-
-\vfill
-
-Adapted from:\\
-Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
-A comparison of single-cell trajectory inference methods.
-\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
-\doi{10.1038/s41587-019-0071-9}.\\
-{\footnotesize $^*$ Equal contribution}
-\newpage
-
 \section{Introduction}
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_1.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/figure_1.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_1.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/supfigure_1.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_2.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/figure_2.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_3.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/figure_3.pdf}
 	\caption{	
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_3.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/supfigure_3.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_4.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/figure_4.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=0.9\linewidth]{fig/dynbenchmark/supfigure_4.pdf}
+	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_4.pdf}
 	\caption{
@@
 Altogether, the scalability analysis indicated that the dimensions of the data are an important factor in the choice of method, and that method development should pay more attention to maintaining reasonable running times and memory usage.
-
+§
 \subsection{Stability}
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_5.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/figure_5.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_6.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/figure_6.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_1.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_1.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_2.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_2.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_3.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_3.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_4.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_4.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_5.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_5.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_6.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_6.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_7.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_7.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_8.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_8.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_9.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_9.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_10.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_10.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_11.pdf}
+	\centering\includegraphics[width=\linewidth]{fig/snote1fig_11.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_12.pdf}
+	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_12.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_13.pdf}
+	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_13.pdf}
 	\caption{
@@
 \begin{figure}
-	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_14.pdf}
+	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_14.pdf}
 	\caption{

### 2019-08-26 ch7_incgraph/fig/pdf2eps.sh

@@ -1,0 +1,9 @@
+#!/bin/sh
+# $Id: pdf2eps,v 0.01 2005/10/28 00:55:46 Herbert Voss Exp $
+# Convert PDF to encapsulated PostScript.
+# usage:
+# pdf2eps <page number> <pdf file without ext>
+
+pdfcrop "$1.pdf" "$1-temp"
+pdftops -eps "$1-temp" "$1.eps"
+rm  "$1-temp"

### 2019-08-26 ch7_incgraph/fig/pdfconvert.sh

@@ -1,0 +1,6 @@
+#!/bin/sh
+
+for file in *.pdf; do
+noext=`echo $file | sed 's#.pdf##'`
+./pdf2eps.sh $noext
+done

### 2019-08-26 fig/dyngen/multiple_backbones.R

@@ -1,92 +1,0 @@
-library(tidyverse)
-library(dyngen)
-
-backs <- list_backbones()
-
-for (nb in names(backs)) {
-  cat("=============== SIMULATING ", nb, " ===============\n", sep = "")
-  set.seed(1)
-  back <- backs[[nb]]()
-  model <-
-    initialise_model(
-      num_tfs = nrow(back$module_info) * 1.5,
-      num_targets = 30,
-      num_hks = 15,
-      num_cells = 300,
-      backbone = back,
-      verbose = TRUE,
-      download_cache_dir = "~/.cache/dyngen",
-      num_cores = 8,
-      simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 16, census_interval = .01)
-    )
-
-  out <- generate_dataset(model, make_plots = FALSE)
-
-  # recalc dimred
-  gs_counts <- out$model$gold_standard$counts %>% as.matrix
-  sim_counts <- out$model$simulations$counts[,colnames(gs_counts)] %>% as.matrix
-  counts <- rbind(gs_counts, sim_counts)
-  space <- dyndimred::dimred_landmark_mds(counts)
-  gs_ix <- seq_len(nrow(gs_counts))
-  out$model$gold_standard$dimred <- space[gs_ix,]
-  out$model$simulations$dimred <- space[-gs_ix,]
-
-  # plot_gold_mappings(out$model, do_facet = FALSE)
-
-  # save output
-  out_dir <- paste0("dyngen/bb_", nb, "/")
-  if (!dir.exists(out_dir)) dir.create(out_dir)
-  write_rds(out, paste0(out_dir, "/out.rds"), compress = "gz")
-
-  # plot model
-  model <- out$model
-
-  g <- plot_backbone_statenet(model)
-  ggsave(paste0(out_dir, "statenet.pdf"), g, width = 6, height = 6)
-  g <- plot_backbone_modulenet(model)
-  ggsave(paste0(out_dir, "/modulenet.pdf"), g, width = 8, height = 6)
-
-  g <- plot_feature_network(model, show_hks = TRUE)
-  ggsave(paste0(out_dir, "/featnet.pdf"), g, width = 8, height = 6)
-  g <- plot_feature_network(model, show_targets = FALSE)
-  ggsave(paste0(out_dir, "/featnet_onlytfs.pdf"), g, width = 8, height = 6)
-
[…]

### 2019-08-26 fig/incgraph/pdf2eps.sh

@@ -1,9 +1,0 @@
-#!/bin/sh
-# $Id: pdf2eps,v 0.01 2005/10/28 00:55:46 Herbert Voss Exp $
-# Convert PDF to encapsulated PostScript.
-# usage:
-# pdf2eps <page number> <pdf file without ext>
-
-pdfcrop "$1.pdf" "$1-temp"
-pdftops -eps "$1-temp" "$1.eps"
-rm  "$1-temp"

### 2019-08-26 boek.tex
refactorise

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\include{ch4_dyno}
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
+\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
+\label{chap:dyno}
+\begin{quote}
+	\textbf{Abstract:} \blindtext
+\end{quote}
+\vfill
+Adapted from:\\
+\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.
+\include{ch4_dyno/content}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\include{ch5_scorpius}
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{SCORPIUS: Fast, accurate, and robust single-cell pseudotime} 
+\chaptermark{SCORPIUS: Fast, accurate, and robust single-cell pseudotime.} % short title
+\label{chap:scorpius}
+\begin{quote}
+	\textbf{Abstract:} \blindtext
+\end{quote}
+\vfill
+Adapted from:\\
+\textbf{Cannoodt, R.}, \ldots, De Preter, K., and Saeys, Y.
+SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
+\textit{Journal} vol, issue (2019), page--page.
+\doi{10.1101/079509v2}.
+\newpage
+\include{ch5_scorpius/content}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\include{ch6_brood}
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{bred: Inferring single cell regulatory networks} 
+\chaptermark{bred: Inferring single cell regulatory networks.} % short title
+\label{chap:bred}
+\begin{quote}
+	\textbf{Abstract:} \blindtext
+\end{quote}
+\vfill
+Adapted from:\\
+\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Inferring Single Cell Regulatory Networks. \textit{Journal} vol, issue (2019), page--page. doi.
+\newpage
+\include{ch6_bred/content}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\include{ch8_discussion}
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{General discussion} 
+\chaptermark{General discussion.} % short title
+\label{chap:discussion}
+\include{ch8_discussion/content}
 

### 2019-08-26 ch8_discussion/content.tex
refactorise

\blindtext

\section{Overview and conclusions of the presented work}

% notice that normalisation was absent from this dissertation.
%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
%
%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.

\section{Future research directions}

### 2019-09-10 ch1_introduction/content.tex
update
updaze

@@
 The cell is the smallest unit of life, of which all known living organisms are composed.
-Every cell houses a plethora of biomolecular processes that allow it to adapt to changes in their environment continuously.
+Every cell houses a plethora of biomolecular processes that allow it to adapt to changes in its environment continuously.
 It can be very challenging to comprehend the cellular response to a signal due to the dynamic nature of these processes.
@@
 
-Reductionist biologists are delighted by recent advances in experimental technologies that permit measuring the abundance of thousands of different biochemical molecules in tens of thousands of individual cells.
-Observing the biomolecular insides of cells in this manner will ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for diagnosing and treating disease.
+Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells.
+Observing the biomolecular insides of cells in this manner can ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for diagnosing and treating disease.
 Every coin has its flip side, however, and in this case, it is that the amount of data generated from these experiments is not analysable by hand.
@@
 
+%%%% HT: I feel like you're only going to focus on preprocessing, when reading this section.
+Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way?
+%%% RC: In general, these two paragraphs are not telling the reader a lot about where this thesis is going
+
 The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. 
-Namely, the data inherently also suffers from batch effects arising from differences between donors and labs, and also contains high levels of noise arising from the experimental profiling techniques used \cite{hon_humancellatlas_2018}.
+Namely, the data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques uses, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}.
 Biologists thus turn to computer scientists\footnote{or computational biologists turn to themselves} to develop new tools to tackle these problems and help biologists extract meaningful biological insights from the data.
@@
 The discovery of the double helix shape of Deoxyribonucleic Acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution.
-A modern high-school student can plausibly be expected to know that DNA serves as a medium for storing the genetic information required to reproduce a whole organism.
+By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism.
 With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.
 The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).
+%% HT: I'm not sure if the magnitude of the DNA discovery is best reflected in sayings and double helix monuments it sounds a bit light to me, compared to the fact that you can identify geneticel targets of a disease and Crispr cas it for instance, but that's a personal opinion, feel free to ignore it :)
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
@@
 \subsection{The Central Dogma}
-The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existant living cells.
+The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existent living cells.
 In short, it states that DNA codes for RNA, which in turn codes for proteins.
 In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure 3).
+
+%% TODO: add figure
 
@@
 The two strands are held together by hydrogen bonds since adenine can form three hydrogen bonds with thymine and cytosine can form two hydrogen bonds with guanine.
-Since each nucleobase can only bond with one other nucleobase, one strand is complementary to the other.
-Thus a strand of DNA can be represented by a sequence of letters, for example, "ACTCGGTTTAGCA".
-% This could be part of the DNA explanation.
+Each nucleobase can only bond with one other nucleobase, making the strands complementary to one another.
+Since one strand can be deduced from its complementary strand, DNA can be represented by a sequence of letters, for example, "ACTCGGTTTAGCA".
 
@@
 RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
+% HT: , which is able to exit the nucleus compartment.
+% RC: I haven't introduced the concept of a nucleus yet
 Due to its single-strandedness, RNA is less stable and will break down faster.
@@
 transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand.
-This work only considers messenger RNA (mRNA).
-mRNAs are transcribed from protein-coding genes, meaning that the mRNAs will lead to the production of protein molecules through the process of translation.
+This work only considers messenger RNA (mRNA), which are transcribed from protein-coding genes, meaning that the mRNAs can result in the production of particular proteins.
+A protein-coding gene consists of alternating sections called introns and exons.
+Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
 % TODO: mention that there are about 20'000 protein-coding genes in the human body.
 
-\textbf{RNA splicing} is a process that occurs in almost all organisms (but not all) and results in the removal of specific regions in the RNA molecule.
-Introns are the regions removed by this process, and exons the regions that remain.
-The main functionality of splicing is to be able to create multiple variants of the same product, which can affect the enzymatic properties or localisation of the resulting product \cite{kelemen_functionalternativesplicing_2013}.
+\textbf{RNA splicing} is a process that occurs in almost all organisms and results in the removal of intronic regions in the mRNA molecules, resulting in a mature mRNA. 
+Splicing allows to create multiple variants of the same product, which can affect the enzymatic properties or localisation of the resulting product \cite{kelemen_functionalternativesplicing_2013}.
 
-During \textbf{translation}, a chain of amino acid is synthesised from an mRNA transcript.
+During \textbf{translation}, a chain of amino acids is synthesised from a mature mRNA transcript. 
+% HT: in ribosomes. These units can be seen as small factories in which amino acids are assembled following the pattern contained in an mRNA molecule.
+% RC: I can avoid talking about ribosomes
 Every three nucleobases are translated into one of 21 different amino acids.
@@
 Cells can be highly specialised toward performing a particular function (e.g.
-memory B cells accelerate immune response by remembering previously encountered pathogens), while other cells maintain a strong ability to differentiate into other cell types.
+memory B cells accelerate immune response by remembering previously encountered pathogens), or they can maintain a strong ability to differentiate into other cell types.
 
@@
 Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate).
-Such proteins are called transcription factors (TFs), and the genes it regulates are called its targets.
+Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets.
 Typically, one TF will regulate the transcription rate of many targets.
@@
 The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
+% HT: I wouldn't write that if it was my thesis, but then again, maybe it's only my personal choice
 Since it was developed by Coons et al.
@@
 
+% HT: Is the section above needed? I felt like this 1.1.5 section was quite hard to read compared to the light previous sections.
+
 On the other side of the spectrum are the so-called "omics" technologies.
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\Largefigure]{fig/trajectory_inference_slim} % TODO: Update figure to match text!!!!
@@
 	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
-	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. InGene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
+	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. In Gene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
 	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L.
@@
 %
-%cran <- c("babelwhale", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "princurve", "proxyC", "qsub", "SCORPIUS", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
+%cran <- c("babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "princurve", "proxyC", "qsub", "SCORPIUS", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
 %bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
@@
 %
-%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / 23 * 365)
-%
-%summ <- dls %>% filter("2018-07-01" <= start, end <= "2019-07-30") %>% group_by(package) %>% summarise(downloads = sum(downloads))
+%min_date <- "2019-08-01"
+%num_days <- as.integer(Sys.Date() - as.Date(min_date))
+%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% as.data.frame
 
@@
 	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the role within the package: aut Author, ctb Contributor.
-Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-08-23.
+Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-09-10.
 CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
@@
 		Name & Role & Host & Downloads per year & Description \\ \hline\hline
-		\cranpkg{babelwhale} & aut & CRAN & 3491 & Interacting with Docker and Singularity containers \\
+		\cranpkg{babelwhale} & aut & CRAN & 3996 & Interacting with Docker and Singularity containers \\
+		\cranpkg{diffusionMap} & aut & CRAN & 21361 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
 		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
-		\cranpkg{dyndimred} & aut & CRAN & 571 & Applying dimensionality reduction methods \\
+		\cranpkg{dyndimred} & aut & CRAN & 5511 & Applying dimensionality reduction methods \\
 		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
@@
 		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
-		\cranpkg{dynparam} & aut & CRAN & 2539 & Creating meta-information for parameters \\
+		\cranpkg{dynparam} & aut & CRAN & 3084 & Creating meta-information for parameters \\
 		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
@@
 		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
-		\cranpkg{dynutils} & aut & CRAN & 4570 & Common functionality for the dynverse packages \\
+		\cranpkg{dynutils} & aut & CRAN & 5657 & Common functionality for the dynverse packages \\
 		\githubpkg{dynverse}{dynwrap} & aut & Github & \notavailable & A common format for trajectories \\
-		\cranpkg{GillespieSSA} & aut & CRAN & 7855 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
-		\cranpkg{GillespieSSA2} & aut & CRAN & 2253 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
+		\cranpkg{GillespieSSA} & aut & CRAN & 7546 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		\cranpkg{GillespieSSA2} & aut & CRAN & 6506 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
 		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & An Rcpp implementation of the Growing Neural Gas algorithm \\
-		\cranpkg{incgraph} & aut & CRAN & 2809 & Incremental graphlet counting for network optimisation \\
-		\cranpkg{princurve} & aut & CRAN & 24'439 & Fits a principal curve in arbitrary dimension \\
-		\cranpkg{proxyC} & aut & CRAN & 112'753 & Computes proximity in large sparse matrices \\
-		\cranpkg{qsub} & aut & CRAN & 2698 & Running commands remotely on gridengine clusters \\
-		\cranpkg{SCORPIUS} & aut & CRAN & 4824 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
-		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} & ctb & Bioc & 571 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
-		\cranpkg{devtools} & ctb & CRAN & 3'254'816 & Tools to make developing R packages easier \\
+		\cranpkg{incgraph} & aut & CRAN & 3175 & Incremental graphlet counting for network optimisation \\
+		\githubpkg{rcannood}{lmds} & aut & Github & \notavailable & Landmark Multi-Dimensional Scaling \\
+		\cranpkg{princurve} & aut & CRAN & 26'991 & Fits a principal curve in arbitrary dimension \\
+		\cranpkg{proxyC} & aut & CRAN & 117'484 & Computes proximity in large sparse matrices \\
+		\cranpkg{qsub} & aut & CRAN & 3193 & Running commands remotely on gridengine clusters \\
+		\cranpkg{SCORPIUS} & aut & CRAN & 4772 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
+		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} & ctb & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
+		\cranpkg{devtools} & ctb & CRAN & 3'775'350 & Tools to make developing R packages easier \\
 		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
-		\cranpkg{mlr} & ctb & CRAN & 141'192 & Machine Learning in R \\
-		\biocpkg{monocle} & ctb & Bioc & 29'121 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
-		\cranpkg{ParamHelpers} & ctb & CRAN & 101'137 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
+		\cranpkg{mlr} & ctb & CRAN & 142'605 & Machine Learning in R \\
+		\biocpkg{monocle} & ctb & Bioc & 35'240 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
+		\cranpkg{ParamHelpers} & ctb & CRAN & 109'408 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
 		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
-		\cranpkg{remotes} & ctb & CRAN & 3'476'435 & R package installation from remote repositories, including GitHub \\
-		\cranpkg{rlang} & ctb & CRAN & 10'314'725 & Functions for base types and core R and tidyverse features \\
+		\cranpkg{remotes} & ctb & CRAN & 3'704'594 & R package installation from remote repositories, including GitHub \\
+		\cranpkg{rlang} & ctb & CRAN & 11'470'763 & Functions for base types and core R and tidyverse features \\
 		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
-		\biocpkg{slingshot} & ctb & Bioc & 9538 & Tools for ordering single-cell sequencing \\
-		\biocpkg{splatter} & ctb & Bioc & 2999 & Simple simulation of single-cell RNA sequencing data \\
+		\biocpkg{slingshot} & ctb & Bioc & 11'643 & Tools for ordering single-cell sequencing \\
+		\biocpkg{splatter} & ctb & Bioc & 3741 & Simple simulation of single-cell RNA sequencing data \\
 		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\

### 2019-09-18 ch1_introduction/content.tex
update

@@
 
-%%%% HT: I feel like you're only going to focus on preprocessing, when reading this section.
+\comment{
+HT: I feel like you're only going to focus on preprocessing, when reading this section.
 Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way?
-%%% RC: In general, these two paragraphs are not telling the reader a lot about where this thesis is going
+\\
+RC: Literally but succinctly state what the exact contributions of the thesis are, this is still very vague.
+}
 
@@
 Afterwards, the research objectives and main contributions of this work are outlined.
-% TODO: explain that only the main concepts are introduced?
 
@@
 The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).
-%% HT: I'm not sure if the magnitude of the DNA discovery is best reflected in sayings and double helix monuments it sounds a bit light to me, compared to the fact that you can identify geneticel targets of a disease and Crispr cas it for instance, but that's a personal opinion, feel free to ignore it :)
+
+\comment{
+RC: Use crispr-cas to assert the impact of the discovery of DNA, not the FSVM spiral 
+}
 
@@
 	\caption{
-		\textbf{A prominent display of the double helix shape at the VIB FSVM building.} % TODO: expand description
+		\textbf{A prominent display of the double helix shape at the VIB FSVM building.}
 	}
@@
 Only later did cells develop the ability to use DNA and proteins to self-sustain in a process commonly referred to as the Central Dogma.
-% TODO: use this reference? \cite{cornell_prebioticaminoacids_2019}
+
+\comment{use this reference? \cite{cornell_prebioticaminoacids_2019}}
 
@@
 \subsection{The Central Dogma}
+\comment{HT: feels like an enumeration, should make it more into a story}
+
 The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existent living cells.
 In short, it states that DNA codes for RNA, which in turn codes for proteins.
-In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure 3).
+In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure \ref{fig:centraldogma}).
 
-%% TODO: add figure
+\begin{figure}[htb!]
+	\centering\includegraphics[width=.5\linewidth]{fig/central_dogma} % TODO: change image
+	\caption{
+		\textbf{Central Dogma.} % TODO: expand the description
+	}
+	\label{fig:centraldogma}
+\end{figure}
 
@@
 RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
-% HT: , which is able to exit the nucleus compartment.
+% HT: Add: which is able to exit the nucleus compartment.
 % RC: I haven't introduced the concept of a nucleus yet
@@
 
-% TODO: do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.
+\comment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
 Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task.
 In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.
-% TODO: are dynamic processes defined? a dynamic process is just a biochemical reaction
+
+\comment{are dynamic processes defined? a dynamic process is just a biochemical reaction}
 
@@
 Typically, one TF will regulate the transcription rate of many targets.
-% TODO: download database and calculate some statistics.
+
+\comment{TODO: download database and calculate some statistics?}
 
@@
 
-% TODO: explain regulation mechanisms such as transcription factor binding sites?
+\comment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
 \subsection{Profiling single cells}
+\comment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
 In order to understand a biological process, it is often quite helpful to be able to profile (i.e.
@@
 The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
-% HT: I wouldn't write that if it was my thesis, but then again, maybe it's only my personal choice
 Since it was developed by Coons et al.
@@
 A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.
+
+\comment{Rephrase the microscope sentence}
 
@@
 
-% HT: Is the section above needed? I felt like this 1.1.5 section was quite hard to read compared to the light previous sections.
-
 On the other side of the spectrum are the so-called "omics" technologies.
@@
 By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
-% Demonstrate the masking effect of bulk analyses.
+\comment{Demonstrate the masking effect of bulk analyses.}
 
@@
 
-% TODO: provide better connection to the subsections below
+\comment{Make a better connection to the subsections below}
 
 \subsection{Normalisation}
-% TODO: need to write still
+\comment{TODO}
 
@@
 Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.
-% section is not very interestingly written
+
+\comment{This section is not very interestingly written}
 
@@
 \end{figure}
+
+\comment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}
 
@@
 \section{Outline} % include contributions?
-% mention focus on datasets which contain developing cells.
+
+\comment{Get a better focus for the overall contributions of the thesis. For example, focus on datasets that contain developing cells, and that the tools we develop make this assumption.}
 
 This work aims to tackle multiple computational problems in extracting biological hypotheses from single-cell omics data. 
-(TODO: rephrase that we focus on datasets that contain developing cells, and that the tools we develop make this assumption.)
 We introduce a framework for evaluating multiple types of single-cell omics tools, more specifically dimensionality reduction, trajectory inference, and gene regulatory network inference (Chapter \ref{chap:dyngen}).
@@
 
+\comment{
+	Submitted:\\
+	RNA atlas,\\
+	TRADE-seq: Trajectory-based differential expression analysis,\\
+	SCENIC protocol
+}
+
 \subsection{Open-source software}
@@
 We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
-This includes help speed up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
+This includes help speed up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
 Several of these package receive millions of downloads per year (devtools, remotes, rlang).
@@
 		\cranpkg{babelwhale} & aut & CRAN & 3996 & Interacting with Docker and Singularity containers \\
-		\cranpkg{diffusionMap} & aut & CRAN & 21361 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
+		\cranpkg{diffusionMap} & aut & CRAN & 21'361 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
 		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\

### 2019-09-19 boek.tex
update
updaze

@@
 \newcommand{\xmark}{\ding{55}}
+
+% boxes around sections
+\usepackage{tcolorbox}
 
@@
 \usepackage{hyperref}
+\usepackage[open,openlevel=1]{bookmark}
 
@@
 \renewcommand{\baselinestretch}{1.25}
+
+% si units and chemical stuff
+\usepackage{siunitx}
+\usepackage[version=3]{mhchem}
+
+% command for adding comments
+\newcommand{\mycomment}[1]{\noindent\fbox{\parbox{\textwidth}{\color{red}#1}}}
 
@@
 \pagestyle{mainmatter}
+
+% TIP FROM WOUTER: Per chapter, write 'update' reflecting on what came out of the paper
 
@@
 Adapted from:\\
-\textbf{Cannoodt, R.}, \ldots, De Preter, K., and Saeys, Y.
+\textbf{Cannoodt, R.}, Saelens, W., Sichien, D., Tavernier, S., Janssens, S.
+Guilliams, M., Lambrecht, B., De Preter, K., and Saeys, Y.
 SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
@@
 \newpage
+\graphicspath{{./ch5_scorpius/}}
 \include{ch5_scorpius/content}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%       CH8: Guidelines       %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{Essential guidelines for computational method benchmarking} 
+\chaptermark{Essential guidelines for computational method benchmarking.} % short title
+\label{chap:guidelines}
+\graphicspath{{./ch8_guidelines/}}
+\begin{quote}
+	\textbf{Abstract:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
+Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
+However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results.
+Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
+\end{quote}
+\vfill
+Adapted from:\\
+Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
+Boulesteix, A., Saeys, Y., and Robinson, M. D.. 
+Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.
+\newpage
+\include{ch8_guidelines/content}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%      CH8: DISCUSSION      %%%
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{General discussion} 
-\chaptermark{General discussion.} % short title
+\chapter{Discussion and Future Perspectives} 
+\chaptermark{Discussion and Future Perspectives.} % short title
 \label{chap:discussion}

### 2019-09-19 ch1_introduction/content.tex
update

@@
+% INTRODUCTION CONTENT
+
 \section{The cell}
@@
 
-\comment{
+\mycomment{
 HT: I feel like you're only going to focus on preprocessing, when reading this section.
@@
 
-\comment{
+\mycomment{
 RC: Use crispr-cas to assert the impact of the discovery of DNA, not the FSVM spiral 
@@
 
-\comment{use this reference? \cite{cornell_prebioticaminoacids_2019}}
+\mycomment{use this reference? \cite{cornell_prebioticaminoacids_2019}}
 
@@
 \subsection{The Central Dogma}
-\comment{HT: feels like an enumeration, should make it more into a story}
+\mycomment{HT: feels like an enumeration, should make it more into a story}
 
@@
 
-\comment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
+\mycomment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
 Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task.
@@
 
-\comment{are dynamic processes defined? a dynamic process is just a biochemical reaction}
+\mycomment{are dynamic processes defined? a dynamic process is just a biochemical reaction}
 
@@
 
-\comment{TODO: download database and calculate some statistics?}
+\mycomment{TODO: download database and calculate some statistics?}
 
@@
 
-\comment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
+\mycomment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
 \subsection{Profiling single cells}
-\comment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
+\mycomment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
 In order to understand a biological process, it is often quite helpful to be able to profile (i.e.
@@
 
-\comment{Rephrase the microscope sentence}
+\mycomment{Rephrase the microscope sentence}
 
@@
 By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
-\comment{Demonstrate the masking effect of bulk analyses.}
+\mycomment{Demonstrate the masking effect of bulk analyses.}
 
@@
 
-\comment{Make a better connection to the subsections below}
+\mycomment{Make a better connection to the subsections below}
 
 \subsection{Normalisation}
-\comment{TODO}
+\mycomment{TODO}
 
@@
 
-\comment{This section is not very interestingly written}
+\mycomment{This section is not very interestingly written}
 
@@
 Trajectory inference (TI) is a new category of computational tools used to offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017,cannoodt_computationalmethodstrajectory_2016}.
+
+%Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
+%This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}.
+%Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}.
+%TI has two objectives: to reconstruct the topology of the dynamic process (e.g.
+is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology.
+Some TI methods assumes that the user knows the topology beforehand and only focuses on ordering the cells along a predefined topology.
 
@@
 
-\comment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}
+\mycomment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}
 
@@
 
-\comment{Get a better focus for the overall contributions of the thesis. For example, focus on datasets that contain developing cells, and that the tools we develop make this assumption.}
+\mycomment{Get a better focus for the overall contributions of the thesis. For example, focus on datasets that contain developing cells, and that the tools we develop make this assumption.}
 
@@
 \textbf{Chapter \ref{chap:bred}} -- why bred, what is it, what can it be used for.
-
-%Before embarking on this perilous adventure, I attempt to write a "Hippocratic oath for bioinformaticians", to guide me through this dissertation and not stray from the metaphorical path of righteousness.
-%
-%\subsection{Hippocratic oath for bioinformaticians}
-%We proceed with caution in developing new tools.
-Its results should not only be convincing and easy to interpret, but also accurate, robust, and reproducible.
-We value creating open-source software that works reliably but fails gracefully when it does not.
-We acknowledge that writing automated tests is dull but necessary for maintaining long-term software projects.
-% TODO: improve
 
@@
 
-\comment{
+\mycomment{
 	Submitted:\\
@@
 		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
+		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
 		\cranpkg{remotes} & ctb & CRAN & 3'704'594 & R package installation from remote repositories, including GitHub \\

### 2019-09-19 ch8_discussion/content.tex
update

@@
-\blindtext
+\section{Essential guidelines for computational method benchmarking}
 
-\section{Overview and conclusions of the presented work}
+%Before embarking on this perilous adventure, I attempt to write a "Hippocratic oath for bioinformaticians", to guide me through this dissertation and not stray from the path of righteousness.
+%
+%\subsection{Hippocratic oath for bioinformaticians}
+%We proceed with caution in developing new tools.
+Its results should not only be convincing and easy to interpret, but also accurate, robust, and reproducible.
+We value creating open-source software that works reliably but fails gracefully when it does not.
+We acknowledge that writing automated tests is dull but necessary for maintaining long-term software projects.
+% TODO: improve
 
-% notice that normalisation was absent from this dissertation.
 %Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
@@
 
-\section{Future research directions}
+\section{dyngen discussion}
+
+\section{dynbenchmark discussion}
+\begin{itemize}
+	\item Already outdated when the manuscript was published online
+	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
+	\item Include RNA velocity as inputs
+\end{itemize}
+
+\section{SCORPIUS discussion}
+\begin{itemize}
+	\item Extension to inferring non-linear trajectories, i.e. with principal graphs
+\end{itemize}
+
+\section{bred discussion}
+
+\section{incgraph discussion}
+
+\section{dyno discussion}
+% notice that normalisation was absent from this dissertation.

### 2019-09-19 ch8_guidelines/content.tex
update

\section{Introduction}

Many fields of computational research are characterized by a growing number of available methods for data analysis. For example, at the time of writing, almost 400 methods are available for analyzing data from single-cell RNA-sequencing experiments [1]. For experimental researchers and method users, this represents both an opportunity and a challenge, since method choice can significantly affect conclusions.

Benchmarking studies are carried out by computational researchers to compare the performance of different methods, using reference datasets and a range of evaluation criteria. Benchmarks may be performed by authors of new methods to demonstrate performance improvements or other advantages; by independent groups interested in systematically comparing existing methods; or organized as community challenges. ‘Neutral’ benchmarking studies, i.e., those performed independently of new method development by authors without any perceived bias, and with a focus on the comparison itself, are especially valuable for the research community [2, 3].

From our experience conducting benchmarking studies in computational biology, we have learned several key lessons that we aim to synthesize in this review. A number of previous reviews have addressed this topic from a range of perspectives, including: overall commentaries and recommendations on benchmarking design [2, 4,5,6,7,8,9]; surveys of design practices followed by existing benchmarks [7]; the importance of neutral benchmarking studies [3]; principles for the design of real-data benchmarking studies [10, 11] and simulation studies [12]; the incorporation of meta-analysis techniques into benchmarking [13,14,15,16]; the organization and role of community challenges [17, 18]; and discussions on benchmarking design for specific types of methods [19, 20]. More generally, benchmarking may be viewed as a form of meta-research [21].

Our aim is to complement previous reviews by providing a summary of essential guidelines for designing, performing, and interpreting benchmarks. While all guidelines are essential for a truly excellent benchmark, some are more fundamental than others. Our target audience consists of computational researchers who are interested in performing a benchmarking study, or who have already begun one. Our review spans the full ‘pipeline’ of benchmarking, from defining the scope to best practices for reproducibility. This includes crucial questions regarding design and evaluation principles: for example, using rankings according to evaluation metrics to identify a set of high-performing methods, and then highlighting different strengths and tradeoffs among these.

The review is structured as a series of guidelines (Figure \ref{fig:summary_guidelines}), each explained in detail in the following sections. We use examples from computational biology; however, we expect that most arguments apply equally to other fields. We hope that these guidelines will continue the discussion on benchmarking design, as well as assisting computational researchers to design and implement rigorous, informative, and unbiased benchmarking analyses.

\begin{figure}[htb!]
	\caption{\textbf{Summary of the guidelines.} The guidelines in this review can be summarized in the following set of recommendations. Each recommendation is discussed in more detail in the corresponding section in the text.}
	\begin{tcolorbox}
		\begin{enumerate}
			\item Define the purpose and scope of the benchmark.
			\item Include all relevant methods.
			\item Select (or design) representative dataset.
			\item Choose appropriate parameter values and software versions.
			\item Evaluate methods according to key quantitative performance metrics.
			\item Evaluate secondary measures including computational requirements, user-friendliness, installation procedures, and documentation quality.
			\item Interpret results and provide recommendations from both user and method developer perspectives.
			\item Publish results in an accessible format
			\item Design the benchmark to enable future extensions.
			\item Follow reproducible research best practices, by making code and data publicly available.
		\end{enumerate}
  \end{tcolorbox}
%	\centering
%	\includegraphics[width=\hugefigure]{fig/summary_guidelines} 
%	\caption{\textbf{The workflow of dyngen is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network. \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the mocules at each time step are extracted. Each time step is mapped to a point in the backbone. \textbf{F:} Multiple cells are sampled from each simulation. Molecules are sampled from each cell.}
	\label{fig:summary_guidelines}
\end{figure}

### 2019-09-20 ch8_guidelines/content.tex
added guidelines for benchmarking

@@
 Many fields of computational research are characterized by a growing number of available methods for data analysis.
-For example, at the time of writing, almost 400 methods are available for analyzing data from single-cell RNA-sequencing experiments [1].
+For example, at the time of writing, almost 400 methods are available for analyzing data from single-cell RNA-sequencing experiments \cite{zappia_exploringsinglecellrnaseq_2018}.
 For experimental researchers and method users, this represents both an opportunity and a challenge, since method choice can significantly affect conclusions.
@@
 Benchmarks may be performed by authors of new methods to demonstrate performance improvements or other advantages; by independent groups interested in systematically comparing existing methods; or organized as community challenges.
-‘Neutral’ benchmarking studies, i.e., those performed independently of new method development by authors without any perceived bias, and with a focus on the comparison itself, are especially valuable for the research community [2, 3].
+'Neutral' benchmarking studies, i.e., those performed independently of new method development by authors without any perceived bias, and with a focus on the comparison itself, are especially valuable for the research community \cite{boulesteix_necessitydesignstudies_2018,boulesteix_pleaneutralcomparison_2013}.
 
 From our experience conducting benchmarking studies in computational biology, we have learned several key lessons that we aim to synthesize in this review.
-A number of previous reviews have addressed this topic from a range of perspectives, including: overall commentaries and recommendations on benchmarking design [2, 4,5,6,7,8,9]; surveys of design practices followed by existing benchmarks [7]; the importance of neutral benchmarking studies [3]; principles for the design of real-data benchmarking studies [10, 11] and simulation studies [12]; the incorporation of meta-analysis techniques into benchmarking [13,14,15,16]; the organization and role of community challenges [17, 18]; and discussions on benchmarking design for specific types of methods [19, 20].
-More generally, benchmarking may be viewed as a form of meta-research [21].
+A number of previous reviews have addressed this topic from a range of perspectives, including: overall commentaries and recommendations on benchmarking design \cite{boulesteix_necessitydesignstudies_2018,peters_puttingbenchmarkstheir_2018,boulesteix_tensimplerules_2015,zheng_benchmarkingcontextsdetails_2017,mangul_systematicbenchmarkingomics_2019,norel_selfassessmenttrapcan_2011,aniba_issuesbioinformaticsbenchmarking_2010}; surveys of design practices followed by existing benchmarks \cite{mangul_systematicbenchmarkingomics_2019}; the importance of neutral benchmarking studies \cite{boulesteix_pleaneutralcomparison_2013}; principles for the design of real-data benchmarking studies \cite{boulesteix_evidencebasedcomputationalstatistics_2017,boulesteix_statisticalframeworkhypothesis_2015} and simulation studies \cite{morris_usingsimulationstudies_2019}; the incorporation of meta-analysis techniques into benchmarking \cite{gardner_identifyingaccuratemetagenome_2019,gardner_metaanalysisbioinformaticssoftware_2017,evangelou_metaanalysismethodsgenomewide_2013,hong_comparisonmetaanalysismethods_2008}; the organization and role of community challenges \cite{boutros_betterbenchmarkingchallengebased_2014,friedberg_tensimplerules_2015}; and discussions on benchmarking design for specific types of methods \cite{vanmechelen_benchmarkingclusteranalysis_2018,angers-loustau_challengesdesigningbenchmark_2018}.
+More generally, benchmarking may be viewed as a form of meta-research \cite{ioannidis_metaresearchwhyresearch_2018}.
 
@@
 Our target audience consists of computational researchers who are interested in performing a benchmarking study, or who have already begun one.
-Our review spans the full ‘pipeline’ of benchmarking, from defining the scope to best practices for reproducibility.
+Our review spans the full 'pipeline' of benchmarking, from defining the scope to best practices for reproducibility.
 This includes crucial questions regarding design and evaluation principles: for example, using rankings according to evaluation metrics to identify a set of high-performing methods, and then highlighting different strengths and tradeoffs among these.
@@
 \begin{figure}[htb!]
-	\caption{\textbf{Summary of the guidelines.} The guidelines in this review can be summarized in the following set of recommendations.
-Each recommendation is discussed in more detail in the corresponding section in the text.}
+	\centering
 	\begin{tcolorbox}
@@
   \end{tcolorbox}
-%	\centering
-%	\includegraphics[width=\hugefigure]{fig/summary_guidelines} 
-%	\caption{\textbf{The workflow of dyngen is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network.
-\textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module.
-Additional target genes are generated.
-\textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions.
-\textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
-The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
-\textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
-The counts of each of the mocules at each time step are extracted.
-Each time step is mapped to a point in the backbone.
-\textbf{F:} Multiple cells are sampled from each simulation.
-Molecules are sampled from each cell.}
+	\caption{\textbf{Summary of the guidelines as a set of recommendations.} Each recommendation is discussed in more detail in the corresponding section in the text.}
 	\label{fig:summary_guidelines}
 \end{figure}
+
+\subsection{Defining the purpose and scope}
+
+The purpose and scope of a benchmark should be clearly defined at the beginning of the study, and will fundamentally guide the design and implementation.
+In general, we can define three broad types of benchmarking studies: (i) those by method developers, to demonstrate the merits of their approach (e.g., \cite{weber_diffcytdifferentialdiscovery_2019,nowicka_drimseqdirichletmultinomialframework_2016,levine_datadrivenphenotypicdissection_2015,zhou_robustlydetectingdifferential_2014,law_voomprecisionweights_2014}); (ii) neutral studies performed to systematically compare methods for a certain analysis, either conducted directly by an independent group (e.g., \cite{saelens_comparisonsinglecelltrajectory_2019,duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018,saelens_comprehensiveevaluationmodule_2018,weber_comparisonclusteringmethods_2016,korthauer_practicalguidemethods_2019,freytag_comparisonclusteringtools_2018,baruzzo_simulationbasedcomprehensivebenchmarking_2017,kanitz_comparativeassessmentmethods_2015,soneson_comparisonmethodsdifferential_2013,rapaport_comprehensiveevaluationdifferential_2013,dillies_comprehensiveevaluationnormalization_2013}) or in collaboration with method authors (e.g., \cite{sage_quantitativeevaluationsoftware_2015}); or (iii) those organized in the form of a community challenge, such as those from the DREAM \cite{weirauch_evaluationmethodsmodeling_2013,costello_communityeffortassess_2014,kuffner_crowdsourcedanalysisclinical_2015,ewing_combiningtumorgenome_2015,hill_inferringcausalmolecular_2016}, FlowCAP \cite{aghaeepour_criticalassessmentautomated_2013,aghaeepour_benchmarkevaluationalgorithms_2016}, CASP \cite{moult_criticalassessmentmethods_2018,moult_criticalassessmentmethods_2016}, CAMI \cite{sczyrba_criticalassessmentmetagenome_2017}, Assemblathon \cite{earl_assemblathoncompetitiveassessment_2011,bradnam_assemblathonevaluatingnovo_2013}, MAQC/SEQC \cite{maqcconsortium_microarrayqualitycontrol_2006,shi_microarrayqualitycontrol_2010,seqc_comprehensiveassessmentrnaseq_2014}, and GA4GH \cite{krusche_bestpracticesbenchmarking_2019} consortia.
+
+A neutral benchmark or community challenge should be as comprehensive as possible, although for any benchmark there will be tradeoffs in terms of available resources.
+To minimize perceived bias, a research group conducting a neutral benchmark should be approximately equally familiar with all included methods, reflecting typical usage of the methods by independent researchers \cite{boulesteix_pleaneutralcomparison_2013}.
+Alternatively, the group could include the original method authors, so that each method is evaluated under optimal conditions; methods whose authors decline to take part should be reported.
+In either case, bias due to focusing attention on particular methods should be avoided -- for example, when tuning parameters or fixing bugs.
+Strategies to avoid these types of biases, such as the use of blinding, have been previously proposed \cite{boulesteix_evidencebasedcomputationalstatistics_2017}.
+
+By contrast, when introducing a new method, the focus of the benchmark will be on evaluating the relative merits of the new method.
+This may be sufficiently achieved with a less extensive benchmark, e.g., by comparing against a smaller set of state-of-the-art and baseline methods.
+However, the benchmark must still be carefully designed to avoid disadvantaging any methods; for example, extensively tuning parameters for the new method while using default parameters for competing methods would result in a biased representation.
+Some advantages of a new method may fall outside the scope of a benchmark; for example, a new method may enable more flexible analyses than previous methods (e.g., beyond two-group comparisons in differential analyses \cite{weber_diffcytdifferentialdiscovery_2019}).
+
+Finally, results should be summarized in the context of the original purpose of the benchmark.
+A neutral benchmark or community challenge should provide clear guidelines for method users, and highlight weaknesses in current methods so that these can be addressed by method developers.
+On the other hand, benchmarks performed to introduce a new method should discuss what the new method offers compared with the current state-of-the-art, such as discoveries that would otherwise not be possible.
+
+\subsection{Selection of methods}
+
+The selection of methods to include in the benchmark will be guided by the purpose and scope of the study.
+A neutral benchmark should include all available methods for a certain type of analysis.
+In this case, the publication describing the benchmark will also function as a review of the literature; a summary table describing the methods is a key output (e.g., Figure 2 in \cite{saelens_comparisonsinglecelltrajectory_2019} or Table 1 in \cite{weber_comparisonclusteringmethods_2016}).
+Alternatively, it may make sense to include only a subset of methods, by defining inclusion criteria: for example, all methods that (i) provide freely available software implementations, (ii) are available for commonly used operating systems, and (iii) can successfully be installed without errors following a reasonable amount of trouble-shooting.
+Such criteria should be chosen without favoring any methods, and exclusion of any widely used methods should be justified.
+A useful strategy can be to involve method authors within the process, since they may provide additional details on optimal usage.
+In addition, community involvement can lead to new collaborations and inspire future method development.
+However, the overall neutrality and balance of the resulting research team should be maintained.
+Finally, if the benchmark is organized as a community challenge, the selection of methods will be determined by the participants.
+In this case, it is important to communicate the initiative widely -- for example, through an established network such as DREAM challenges.
+However, some authors may choose not to participate; a summary table documenting non-included methods should be provided in this case.
+
+When developing a new method, it is generally sufficient to select a representative subset of existing methods to compare against.
+For example, this could consist of the current best-performing methods (if known), a simple 'baseline' method, and any methods that are widely used.
+The selection of competing methods should ensure an accurate and unbiased assessment of the relative merits of the new approach, compared with the current state-of-the-art.
+In fast-moving fields, for a truly excellent benchmark, method developers should be prepared to update their benchmarks or design them to easily allow extensions as new methods emerge.
+
+\subsection{Selection (or design) of datasets}
+
+The selection of reference datasets is a critical design choice.
+If suitable publicly accessible datasets cannot be found, they will need to be generated or constructed, either experimentally or by simulation.
+Including a variety of datasets ensures that methods can be evaluated under a wide range of conditions.
+In general, reference datasets can be grouped into two main categories: simulated (or synthetic) and real (or experimental).
+
+Simulated data have the advantage that a known true signal (or 'ground truth') can easily be introduced; for example, whether a gene is differentially expressed.
+Quantitative performance metrics measuring the ability to recover the known truth can then be calculated.
+However, it is important to demonstrate that simulations accurately reflect relevant properties of real data, by inspecting empirical summaries of both simulated and real datasets (e.g., using automated tools \cite{soneson_unifiedqualityverification_2018}).
+The set of empirical summaries to use is context-specific; for example, for single-cell RNA-sequencing, dropout profiles and dispersion-mean relationships should be compared \cite{soneson_biasrobustnessscalability_2018}; for DNA methylation, correlation patterns among neighboring CpG sites should be investigated \cite{korthauer_detectionaccuratefalse_2019}; for comparing mapping algorithms, error profiles of the sequencing platforms should be considered \cite{caboche_comparisonmappingalgorithms_2014}.
+Simplified simulations can also be useful, to evaluate a new method under a basic scenario, or to systematically test aspects such as scalability and stability.
+However, overly simplistic simulations should be avoided, since these will not provide useful information on performance.
+A further advantage of simulated data is that it is possible to generate as much data as required; for example, to study variability and draw statistically valid conclusions.
+
+Experimental data often do not contain a ground truth, making it difficult to calculate performance metrics.
+Instead, methods may be evaluated by comparing them against each other (e.g., overlap between sets of detected differential features \cite{nowicka_drimseqdirichletmultinomialframework_2016}), or against a current widely accepted method or 'gold standard' (e.g., manual gating to define cell populations in high-dimensional cytometry \cite{weber_comparisonclusteringmethods_2016, aghaeepour_criticalassessmentautomated_2013}, or fluorescence in situ hybridization to validate absolute copy number predictions \cite{zheng_benchmarkingcontextsdetails_2017}).
+In the context of supervised learning, the response variable to be predicted is known in the manually labeled training and test data.
+However, individual datasets should not be overused, and using the same dataset for both method development and evaluation should be avoided, due to the risk of overfitting and overly optimistic results \cite{grimm_evaluationtoolsused_2015,jelizarow_overoptimismbioinformaticsillustration_2010}.
+In some cases, it is also possible to design experimental datasets containing a ground truth.
+Examples include: (i) 'spiking in' synthetic RNA molecules at known relative concentrations \cite{jiang_syntheticspikeinstandards_2011} in RNA-sequencing experiments (e.g., \cite{seqc_comprehensiveassessmentrnaseq_2014,garalde_highlyparalleldirect_2018}), (ii) large-scale validation of gene expression measurements by quantitative polymerase chain reaction (e.g., \cite{seqc_comprehensiveassessmentrnaseq_2014}), (iii) using genes located on sex chromosomes as a proxy for silencing of DNA methylation status (e.g., \cite{law_voomprecisionweights_2014,fang_genomiclandscapehuman_2012}), (iv) using fluorescence-activated cell sorting to sort cells into known subpopulations prior to single-cell RNA-sequencing (e.g., \cite{soneson_biasrobustnessscalability_2018, schaum_singlecelltranscriptomics20_2018, zheng_massivelyparalleldigital_2017}), or (v) mixing different cell lines to create 'pseudo-cells' \cite{tian_benchmarkingsinglecell_2019}.
+However, it may be difficult to ensure that the ground truth represents an appropriate level of variability -- for example, the variability of spiked-in material, or whether method performance on cell line data is relevant to outbred populations.
+Alternatively, experimental datasets may be evaluated qualitatively, for example, by judging whether each method can recover previous discoveries, although this strategy relies on the validity of previous results.
+
+A further technique is to design 'semi-simulated' datasets that combine real experimental data with an 'in silico' (i.e., computational) spike-in signal; for example, by combining cells or genes from 'null' (e.g., healthy) samples with a subset of cells or genes from samples expected to contain a true differential signal (examples include \cite{weber_diffcytdifferentialdiscovery_2019,arvaniti_sensitivedetectionrare_2017,rigaill_syntheticdatasets_2018}).
+This strategy can create datasets with more realistic levels of variability and correlation, together with a ground truth.
+
+Overall, there is no perfect reference dataset, and the selection of appropriate datasets will involve tradeoffs, e.g., regarding the level of complexity.
+Both simulated and experimental data should not be too 'simple' (e.g., two of the datasets in the FlowCAP-II challenge \cite{aghaeepour_criticalassessmentautomated_2013} gave perfect performance for several algorithms) or too 'difficult' (e.g., for the third dataset in FlowCAP-II, no algorithms performed well); in these situations, it can be impossible to distinguish performance.
+In some cases, individual datasets have also been found to be unrepresentative, leading to over-optimistic or otherwise biased assessment of methods (e.g., \cite{lowes_bralibasedentatale_2017}).
+Overall, the key to truly excellent benchmarking is diversity of evaluations, i.e., using a range of metrics and datasets that span the range of those that might be encountered in practice, so that performance estimates can be credibly extrapolated.
+
+\subsection{Parameters and software versions}
+
+Parameter settings can have a crucial impact on performance.
+Some methods have a large number of parameters, and tuning parameters to optimal values can require significant effort and expertise.
+For a neutral benchmark, a range of parameter values should ideally be considered for each method, although tradeoffs need to be considered regarding available time and computational resources.
+Importantly, the selection of parameter values should comply with the neutrality principle, i.e., certain methods should not be favored over others through more extensive parameter tuning.
+
+There are three major strategies for choosing parameters.
+The first (and simplest) is to use default values for all parameters.
+Default parameters may be adequate for many methods, although this is difficult to judge in advance.
+While this strategy may be viewed as too simplistic for some neutral benchmarks, it reflects typical usage.
+We used default parameters in several neutral benchmarks where we were interested in performance for untrained users \cite{saelens_comparisonsinglecelltrajectory_2019,couronne_randomforestlogistic_2018,schneider_mortalityriskacute_2016}.
+In addition, for \cite{saelens_comparisonsinglecelltrajectory_2019}, due to the large number of methods and datasets, total runtime was already around a week using 192 processor cores, necessitating judgment in the scope of parameter tuning.
+The second strategy is to choose parameters based on previous experience or published values.
+This relies on familiarity with the methods and the literature, reflecting usage by expert users.
+The third strategy is to use a systematic or automated parameter tuning procedure -- for example, a 'grid search' across ranges of values for multiple parameters or techniques such as cross-validation (e.g., \cite{saelens_comprehensiveevaluationmodule_2018}).
+The strategies may also be combined, e.g., setting non-critical parameters to default values and performing a grid search for key parameters.
+Regardless, neutrality should be maintained: comparing methods with the same strategy makes sense, while comparing one method with default parameters against another with extensive tuning makes for an unfair comparison.
+
+For benchmarks performed to introduce a new method, comparing against a single set of optimal parameter values for competing methods is often sufficient; these values may be selected during initial exploratory work or by consulting documentation.
+However, as outlined above, bias may be introduced by tuning the parameters of the new method more extensively.
+The parameter selection strategy should be transparently discussed during the interpretation of the results, to avoid the risk of over-optimistic reporting due to expending more 'researcher degrees of freedom' on the new method \cite{boulesteix_tensimplerules_2015,hu_parametertuningkey_2019}.
+
+Software versions can also influence results, especially if updates include major changes to methodology (e.g., \cite{vaquero-garcia_leafcuttervsmajiq_2018}).
+Final results should generally be based on the latest available versions, which may require re-running some methods if updates become available during the course of a benchmark.
+
+\subsection{Evaluation criteria: key quantitative performance metrics}
+
+Evaluation of methods will rely on one or more quantitative performance metrics (Figure  \ref{fig:fig2_metrics}A).
+The choice of metric depends on the type of method and data.
+For example, for classification tasks with a ground truth, metrics include the true positive rate (TPR; sensitivity or recall), false positive rate (FPR; 1 - specificity), and false discovery rate (FDR).
+For clustering tasks, common metrics include the F1 score, adjusted Rand index, normalized mutual information, precision, and recall; some of these can be calculated at the cluster level as well as averaged (and optionally weighted) across clusters (e.g., these metrics were used to evaluate clustering methods in our own work \cite{duo_systematicperformanceevaluation_2018,weber_comparisonclusteringmethods_2016}  and by others \cite{freytag_comparisonclusteringtools_2018,aghaeepour_criticalassessmentautomated_2013,wiwie_comparingperformancebiomedical_2015}).
+Several of these metrics can also be compared visually to capture the tradeoff between sensitivity and specificity, e.g., using receiver operating characteristic (ROC) curves (TPR versus FPR), TPR versus FDR curves, or precision-recall (PR) curves (Figure \ref{fig:fig2_metrics}B).
+For imbalanced datasets, PR curves have been shown to be more informative than ROC curves \cite{saito_precisionrecallplotmore_2015,powers_visualizationtradeoffevaluation_2015}.
+These visual metrics can also be summarized as a single number, such as area under the ROC or PR curve; examples from our work include \cite{weber_diffcytdifferentialdiscovery_2019,soneson_biasrobustnessscalability_2018}.
+In addition to the tradeoff between sensitivity and specificity, a method's 'operating point' is important; in particular, whether the threshold used (e.g., 5\% FDR) is calibrated to achieve the specified error rate.
+We often overlay this onto TPR-FDR curves by filled or open circles (e.g., Figure \ref{fig:fig2_metrics}B, generated using the iCOBRA package \cite{soneson_icobraopenreproducible_2016}); examples from our work include \cite{weber_diffcytdifferentialdiscovery_2019,nowicka_drimseqdirichletmultinomialframework_2016,zhou_robustlydetectingdifferential_2014,soneson_differentialanalysesrnaseq_2016}.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\hugefigure]{fig/essential_guidelines_figure2} 
+	\caption{\textbf{Summary and examples of performance metrics.} \textbf{A:} Schematic overview of classes of frequently used performance metrics, including examples (\textit{boxes outlined in gray}).
+\textbf{B:} Examples of popular visualizations of quantitative performance metrics for classification methods, using reference datasets with a ground truth.
+ROC curves (\textit{left}).
+TPR versus FDR curves (\textit{center}); circles represent observed TPR and FDR at typical FDR thresholds of 1, 5, and 10\%, with filled circles indicating observed FDR lower than or equal to the imposed threshold.
+PR curves (\textit{right}).
+Visualizations were generated using iCOBRA R/Bioconductor package \cite{soneson_icobraopenreproducible_2016}.
+\textit{FDR} false discovery rate, \textit{FPR} false positive rate, \textit{PR} precision-recall, \textit{ROC} receiver operating characteristic, \textit{TPR} true positive rate}
+	\label{fig:fig2_metrics}
+\end{figure}
+
+For methods with continuous-valued output (e.g., effect sizes or abundance estimates), metrics include the root mean square error, distance measures, Pearson correlation, sum of absolute log-ratios, log-modulus, and cross-entropy.
+As above, the choice of metric depends on the type of method and data (e.g., \cite{costello_communityeffortassess_2014,lindgreen_evaluationaccuracyspeed_2016}  used correlation, while \cite{moult_criticalassessmentmethods_2016} used root mean square deviation).
+Further classes of methods include those generating graphs, phylogenetic trees, overlapping clusters, or distributions; these require more complex metrics.
+In some cases, custom metrics may need to be developed (e.g., we defined new metrics for topologies of developmental trajectories in \cite{saelens_comparisonsinglecelltrajectory_2019}).
+When designing custom metrics, it is important to assess their reliability across a range of prediction values (e.g., \cite{gurevich_quastqualityassessment_2013,narzisi_comparingnovogenome_2011}).
+For some metrics, it may also be useful to assess uncertainty, e.g., via confidence intervals.
+In the context of supervised learning, classification or prediction accuracy can be evaluated by cross-validation, bootstrapping, or on a separate test dataset (e.g., \cite{gardner_identifyingaccuratemetagenome_2019,aghaeepour_benchmarkevaluationalgorithms_2016}).
+In this case, procedures to split data into training and test sets should be appropriate for the data structure and the prediction task at hand (e.g., leaving out whole samples or chromosomes \cite{schreiber_pitfallmachinelearning_2019}).
+
+Additional metrics that do not rely on a ground truth include measures of stability, stochasticity, and robustness.
+These measures may be quantified by running methods multiple times using different inputs or subsampled data (e.g., we observed substantial variability in performance for some methods in \cite{soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
+'Missing values' may occur if a method does not return any values for a certain metric, e.g., due to a failure to converge or other computational issues such as excessive runtime or memory requirements (e.g., \cite{saelens_comparisonsinglecelltrajectory_2019, soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
+Fallback solutions such as imputation may be considered in this case \cite{bischl_benchmarkinglocalclassification_2013}, although these should be transparently reported.
+For non-deterministic methods (e.g., with random starts or stochastic optimization), variability in performance when using different random seeds or subsampled data should be characterized.
+Null comparisons can be constructed by randomizing group labels such that datasets do not contain any true signal, which can provide information on error rates (e.g., \cite{weber_diffcytdifferentialdiscovery_2019,zhou_robustlydetectingdifferential_2014,law_voomprecisionweights_2014}).
+However, these must be designed carefully to avoid confounding by batch or population structure, and to avoid strong within-group batch effects that are not accounted for.
+
+For most benchmarks, multiple metrics will be relevant.
+Focusing on a single metric can give an incomplete view: methods may not be directly comparable if they are designed for different tasks, and different users may be interested in different aspects of performance.
+Therefore, a crucial design decision is whether to focus on an overall ranking, e.g., by combining or weighting multiple metrics.
+In general, it is unlikely that a single method will perform best across all metrics, and performance differences between top-ranked methods for individual metrics can be small.
+Therefore, a good strategy is to use rankings from multiple metrics to identify a set of consistently high-performing methods, and then highlight the different strengths of these methods.
+For example, in \cite{weber_comparisonclusteringmethods_2016}, we identified methods that gave good clustering performance, and then highlighted differences in runtimes among these.
+In several studies, we have presented results in the form of a graphical summary of performance according to multiple criteria (examples include Figure 3 in \cite{saelens_comparisonsinglecelltrajectory_2019} and Figure 5 in \cite{soneson_biasrobustnessscalability_2018} from our work; and Figure 2 in \cite{sage_quantitativeevaluationsoftware_2015} and Figure 6 in \cite{korthauer_practicalguidemethods_2019} from other authors).
+Identifying methods that consistently underperform can also be useful, to allow readers to avoid these.
+
+\subsection{Evaluation criteria: secondary measures}
+
+In addition to the key quantitative performance metrics, methods should also be evaluated according to secondary measures, including runtime, scalability, and other computational requirements, as well as qualitative aspects such as user-friendliness, installation procedures, code quality, and documentation quality (Figure \ref{fig:fig2_metrics}A).
+From the user perspective, the final choice of method may involve tradeoffs according to these measures: an adequately performing method may be preferable to a top-performing method that is especially difficult to use.
+
+In our experience, runtimes and scalability can vary enormously between methods (e.g., in our work, runtimes for cytometry clustering algorithms \cite{weber_comparisonclusteringmethods_2016} and metagenome analysis tools \cite{lindgreen_evaluationaccuracyspeed_2016} ranged across multiple orders of magnitude for the same datasets).
+Similarly, memory and other computational requirements can vary widely.
+Runtimes and scalability may be investigated systematically, e.g., by varying the number of cells or genes in a single-cell RNA-sequencing dataset \cite{duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018}.
+In many cases, there is a tradeoff between performance and computational requirements.
+In practice, if computational requirements for a top-performing method are prohibitive, then a different method may be preferred by some users.
+
+User-friendliness, installation procedures, and documentation quality can also be highly variable \cite{mangul_improvingusabilityarchival_2019,mangul_challengesrecommendationsimprove_2019}.
+Streamlined installation procedures can be ensured by distributing the method via standard package repositories, such as CRAN and Bioconductor for R, or PyPI for Python.
+Alternative options include GitHub and other code repositories or institutional websites; however, these options do not provide users with the same guarantees regarding reliability and documentation quality.
+Availability across multiple operating systems and within popular programming languages for data analysis is also important.
+Availability of graphical user interfaces can further extend accessibility, although graphical-only methods hinder reproducibility and are thus difficult to include in a systematic benchmark.
+
+For many users, freely available and open source software will be preferred, since it is more broadly accessible and can be adapted by experienced users.
+From the developer perspective, code quality and use of software development best practices, such as unit testing and continuous integration, are also important.
+Similarly, adherence to commonly used data formats (e.g., GFF/GTF files for genomic features, BAM/SAM files for sequence alignment data, or FCS files for flow or mass cytometry data) greatly improves accessibility and extensibility.
+
+High-quality documentation is critical, including help pages and tutorials.
+Ideally, all code examples in the documentation should be continually tested, e.g., as Bioconductor does, or through continuous integration.
+
+\subsection{Interpretation, guidelines, and recommendations}
+
+For a truly excellent benchmark, results must be clearly interpreted from the perspective of the intended audience.
+For method users, results should be summarized in the form of recommendations.
+An overall ranking of methods (or separate rankings for multiple evaluation criteria) can provide a useful overview.
+However, as mentioned above, some methods may not be directly comparable (e.g.
+since they are designed for different tasks), and different users may be interested in different aspects of performance.
+In addition, it is unlikely that there will be a clear 'winner' across all criteria, and performance differences between top-ranked methods can be small.
+Therefore, an informative strategy is to use the rankings to identify a set of high-performing methods, and to highlight the different strengths and tradeoffs among these methods.
+The interpretation may also involve biological or other domain knowledge to establish the scientific relevance of differences in performance.
+Importantly, neutrality principles should be preserved during the interpretation.
+
+For method developers, the conclusions may include guidelines for possible future development of methods.
+By assisting method developers to focus their research efforts, high-quality benchmarks can have significant impact on the progress of methodological research.
+
+Limitations of the benchmark should be transparently discussed.
+For example, in \cite{saelens_comparisonsinglecelltrajectory_2019} we used default parameters for all methods, while in \cite{weber_comparisonclusteringmethods_2016} our datasets relied on manually gated reference cell populations as the ground truth.
+Without a thorough discussion of limitations, a benchmark runs the risk of misleading readers; in extreme cases, this may even harm the broader research field by guiding research efforts in the wrong directions.
+
+\subsection{Publication and reporting of results}
+
+The publication and reporting strategy should emphasize clarity and accessibility.
+Visualizations summarizing multiple performance metrics can be highly informative for method users (examples include Figure 3 in \cite{saelens_comparisonsinglecelltrajectory_2019} and Figure 5 in \cite{soneson_biasrobustnessscalability_2018} from our own work; as well as Figure 6 in \cite{korthauer_practicalguidemethods_2019}).
+Summary tables are also useful as a reference (e.g., \cite{weber_comparisonclusteringmethods_2016,aghaeepour_criticalassessmentautomated_2013}).
+Additional visualizations, such as flow charts to guide the choice of method for different analyses, are a helpful way to engage the reader (e.g., Figure 5 in \cite{saelens_comparisonsinglecelltrajectory_2019}).
+
+For extensive benchmarks, online resources enable readers to interactively explore the results (examples from our work include \cite{saelens_comparisonsinglecelltrajectory_2019, weber_comparisonclusteringmethods_2016}, which allow users to filter metrics and datasets).
+Figure 3 displays an example of an interactive website from one of our benchmarks \cite{saelens_comparisonsinglecelltrajectory_2019}, which facilitates exploration of results and assists users with choosing a suitable method.
+While tradeoffs should be considered in terms of the amount of work required, these efforts are likely to have significant benefit for the community.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\hugefigure]{fig/essential_guidelines_figure3} 
+	\caption{\textbf{Example of an interactive website allowing users to explore the results of one of our benchmarking studies \cite{saelens_comparisonsinglecelltrajectory_2019}.} This website was created using the Shiny framework in R.}
+	\label{fig:fig3_interactive}
+\end{figure}
+
+In most cases, results will be published in a peer-reviewed article.
+For a neutral benchmark, the benchmark will be the main focus of the paper.
+For a benchmark to introduce a new method, the results will form one part of the exposition.
+We highly recommend publishing a preprint prior to peer review (e.g., on bioRxiv or arXiv) to speed up distribution of results, broaden accessibility, and solicit additional feedback.
+In particular, direct consultation with method authors can generate highly useful feedback (examples from our work are described in the acknowledgments in \cite{lindgreen_evaluationaccuracyspeed_2016,freyhult_exploringgenomicdark_2007}).
+Finally, at publication time, considering open access options will further broaden accessibility.
+
+\subsection{Enabling future extensions}
+
+Since new methods are continually emerging \cite{zappia_exploringsinglecellrnaseq_2018}, benchmarks can quickly become out of date.
+To avoid this, a truly excellent benchmark should be extensible.
+For example, creating public repositories containing code and data allows other researchers to build on the results to include new methods or datasets, or to try different parameter settings or pre-processing procedures (examples from our work include \cite{saelens_comparisonsinglecelltrajectory_2019,duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018,saelens_comprehensiveevaluationmodule_2018,weber_comparisonclusteringmethods_2016}).
+In addition to raw data and code, it is useful to distribute pre-processed and/or results data (examples include \cite{duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018,soneson_icobraopenreproducible_2016} from our work and \cite{wiwie_comparingperformancebiomedical_2015,bokulich_mockrobiotapublicresource_2016,conchuir_webresourcestandardized_2015} from others), especially for computationally intensive benchmarks.
+This may be combined with an interactive website, where users can upload results from a new method, to be included in an updated comparison either automatically or by the original authors (e.g., \cite{kanitz_comparativeassessmentmethods_2015,cope_benchmarkaffymetrixgenechip_2004,irizarry_comparisonaffymetrixgenechip_2006}).
+'Continuous' benchmarks, which are continually updated, are especially convenient (e.g., \cite{barton_nucleotidesgenomeassembler_2014}), but may require significant additional effort.
+
+\subsection{Reproducible research best practices}
+
+Reproducibility of research findings has become an increasing concern in numerous areas of study \cite{ioannidis_whymostpublished_2005}.
+In computational sciences, reproducibility of code and data analyses has been recognized as a useful 'minimum standard' that enables other researchers to verify analyses \cite{peng_reproducibleresearchcomputational_2011}.
+Access to code and data has previously enabled method developers to uncover potential errors in published benchmarks due to suboptimal usage of methods \cite{vaquero-garcia_leafcuttervsmajiq_2018,zhou_countbaseddifferentialexpression_2015,zhou_mirnaseqnormalizationcomparisons_2013}.
+Journal publication policies can play a crucial role in encouraging authors to follow these practices \cite{hofner_reproducibleresearchstatistics_2016}; experience shows that statements that code and data are 'available on request' are often insufficient \cite{boulesteix_makingcomplexprediction_2019}.
+In the context of benchmarking, code and data availability also provides further benefits: for method users, code repositories serve as a source of annotated code to run methods and build analysis pipelines, while for developers, code repositories can act as a prototype for future method development work.
+
+Parameter values (including random seeds) and software versions should be clearly reported to ensure complete reproducibility.
+For methods that are run using scripts, these will be recorded within the scripts.
+In R, the command \texttt{sessionInfo()} gives a complete summary of package versions, the version of R, and the operating system.
+For methods only available via graphical interfaces, parameters and versions must be recorded manually.
+Reproducible workflow frameworks, such as the Galaxy platform \cite{afgan_galaxyplatformaccessible_2018}, can also be helpful.
+A summary table or spreadsheet of parameter values and software versions can be published as supplementary information along with the publication describing the benchmark (e.g., Supporting Information Table S1 in our study \cite{weber_comparisonclusteringmethods_2016}).
+
+Automated workflow management tools and specialized tools for organizing benchmarks provide sophisticated options for setting up benchmarks and creating a reproducible record, including software environments, package versions, and parameter values.
+Examples include SummarizedBenchmark \cite{kimes_reproduciblereplicablecomparisons_2019}, DataPackageR \cite{finak_datapackagerreproducibledata_2018}, workflowr \cite{blischak_workflowrframeworkreproducible_2019}, and Dynamic Statistical Comparisons \cite{wang_dscdynamicstatistical_2016}.
+Some tools (e.g., workflowr) also provide streamlined options for publishing results online.
+In machine learning, OpenML provides a platform to organize and share benchmarks \cite{vanschoren_openmlnetworkedscience_2014}.
+More general tools for managing computational workflows, including Snakemake \cite{koster_snakemakescalablebioinformatics_2012}, Make, Bioconda \cite{gruning_biocondasustainablecomprehensive_2018}, and conda, can be customized to capture setup information.
+Containerization tools such as Docker and Singularity may be used to encapsulate a software environment for each method, preserving the package version as well as dependency packages and the operating system, and facilitating distribution of methods to end users (e.g., in our study \cite{saelens_comparisonsinglecelltrajectory_2019}).
+Best practices from software development are also useful, including unit testing and continuous integration.
+
+Many free online resources are available for sharing code and data, including GitHub and Bitbucket, repositories for specific data types (e.g., ArrayExpress \cite{kolesnikov_arrayexpressupdatesimplifying_2015}, the Gene Expression Omnibus \cite{barrett_ncbigeoarchive_2013}, and FlowRepository \cite{spidlen_flowrepositoryresourceannotated_2012}), and more general data repositories (e.g., figshare, Dryad, Zenodo, Bioconductor ExperimentHub, and Mendeley Data).
+Customized resources (examples from our work include \cite{soneson_biasrobustnessscalability_2018, soneson_icobraopenreproducible_2016}) can be designed when additional flexibility is needed.
+Several repositories allow the creation of 'digital object identifiers' (DOIs) for code or data objects.
+In general, preference should be given to publicly funded repositories, which provide greater guarantees for long-term archival stability \cite{mangul_improvingusabilityarchival_2019,mangul_challengesrecommendationsimprove_2019}.
+
+An extensive literature exists on best practices for reproducible computational research (e.g., \cite{sandve_tensimplerules_2013}).
+Some practices (e.g., containerization) may involve significant additional work; however, in our experience, almost all efforts in this area prove useful, especially by facilitating later extensions by ourselves or other researchers.
+
+\section{Discussion}
+In this review, we have described a set of key principles for designing a high-quality computational benchmark.
+In our view, elements of all of these principles are essential.
+However, we have also emphasized that any benchmark will involve tradeoffs, due to limited expertise and resources, and that some principles are less central to the evaluation.
+Table \ref{tab:essential_principles} provides a summary of examples of key tradeoffs and pitfalls related to benchmarking, along with our judgment of how truly 'essential' each principle is.
+
+\newcommand{\mycola}[2][1]{\multirow{#1}{3.5cm}{#2}}
+\newcommand{\mycolb}[2][1]{\multirow{#1}{3.5cm}{#2}}
+
+\begin{table}[ht!]
+	\caption{\textbf{Summary of our views regarding 'how essential' each principle is for a truly excellent benchmark, along with examples of key tradeoffs and potential pitfalls relating to each principle}.
+The higher the number of plus signs, the more central the principle is to the evaluation.} \label{tab:essential_principles}
+	\centering
+	\footnotesize
+	\begin{tabularx}{\textwidth}{|p{3.5cm}p{1cm}p{3.5cm}X|}
+		\hline
+		\textbf{Principle} & \textbf{How} \mbox{\textbf{essential}}  & \textbf{Tradeoffs} & \textbf{Potential pitfalls} \\ \hline \hline
+		\mycola[2]{1. Defining the purpose and score} & +++ & \mycolb[2]{How comprehensive the benchmark should be} & Scope too broad: too much work given available resources \\ 
+		& & & Scope too narrow: unrepresentative and possibly misleading results \\ \hline
+		{2. Selection of methods} & +++ & {Number of methods to include} & Excluding key methods \\
+		\mycola{3.
+Selection (or design) of datasets} & +++ & \mycolb{Number and types of datasets to include} & Subjectivity in the choice of datasets: e.g., selecting datasets that are unrepresentative of real-world applications \\
+		& & & Too few datasets or simulation scenarios Overly simplistic simulations \\ \hline
+		\mycola{4.
+Parameter and software versions} & ++ & \mycolb{Amount of parameter tuning} & Extensive parameter tuning for some methods while using default parameters for others (e.g., competing methods) \\ \hline
+		\mycola[3]{5.
+Evaluation criteria: key quantitative performance metrics} & +++ & \mycolb[3]{Number and types of performance metrics} & Subjectivity in the choice of metrics: e.g., selecting metrics that do not translate to real-world performance \\
+		& & & Metrics that give over-optimistic estimates of performance \\
+		& & & Methods may not be directly comparable according to individual metrics (e.g., if methods are designed for different tasks) \\ \hline
+		\mycola[3]{6.
+Evaluation criteria: secondary measures} & ++ & \mycolb[3]{Number and types of performance metrics} & Subjectivity of qualitative measures such as user-friendliness, installation procedures, and documentation quality \\
+		& & & Subjectivity in relative weighting between multiple metrics \\
+		& & & Measures such as runtime and scalability depend on processor speed and memory \\ \hline
+		\mycola[2]{7.
+Interpretation, guidelines, and recommendations} & ++ & \mycolb[2]{Generality versus specificity of recommendations} & Performance differences between top-ranked methods may be minor \\
+		& & & Different readers may be interested in different aspects of performance \\ \hline
+		\mycola{8.
+Publication and reporting of results} & + & Amount of resources to dedicate to building online resources & Online resources may not be accessible (or may no longer run) several years later \\ \hline
+		\mycola{9.
+Enabling future extensions} & ++ & \mycolb{Amount of resources to dedicate to ensuring extensibility} & Selection of methods or datasets for future extensions may be unrepresentative (e.g., due to requests from method authors) \\ \hline
+		\mycola{10. Reproducible research best practices} & ++ & Amount of resources to dedicate to reproducibility & Some tools may not be compatible or accessible several years later \\ \hline
+	\end{tabularx}
+\end{table}
+
+A number of potential pitfalls may arise from benchmarking studies (Table \ref{tab:essential_principles}).
+For example, subjectivity in the choice of datasets or evaluation metrics could bias the results.
+In particular, a benchmark that relies on unrepresentative data or metrics that do not translate to real-world scenarios may be misleading by showing poor performance for methods that otherwise perform well.
+This could harm method users, who may select an inappropriate method for their analyses, as well as method developers, who may be discouraged from pursuing promising methodological approaches.
+In extreme cases, this could negatively affect the research field by influencing the direction of research efforts.
+A thorough discussion of the limitations of a benchmark can help avoid these issues.
+Over the longer term, critical evaluations of published benchmarks, so-called meta-benchmarks, will also be informative \cite{boulesteix_evidencebasedcomputationalstatistics_2017,gardner_identifyingaccuratemetagenome_2019,gardner_metaanalysisbioinformaticssoftware_2017}.
+
+Well-designed benchmarking studies provide highly valuable information for users and developers of computational methods, but require careful consideration of a number of important design principles.
+In this review, we have discussed a series of guidelines for rigorous benchmarking design and implementation, based on our experiences in computational biology.
+We hope these guidelines will assist computational researchers to design high-quality, informative benchmarks, which will contribute to scientific advances through informed selection of methods by users and targeting of research efforts by developers.

### 2019-09-20 ch8_guidelines/content.tex
markup changes

@@
 Benchmarks may be performed by authors of new methods to demonstrate performance improvements or other advantages; by independent groups interested in systematically comparing existing methods; or organized as community challenges.
-'Neutral' benchmarking studies, i.e., those performed independently of new method development by authors without any perceived bias, and with a focus on the comparison itself, are especially valuable for the research community \cite{boulesteix_necessitydesignstudies_2018,boulesteix_pleaneutralcomparison_2013}.
+Neutral benchmarking studies, i.e., those performed independently of new method development by authors without any perceived bias, and with a focus on the comparison itself, are especially valuable for the research community \cite{boulesteix_necessitydesignstudies_2018,boulesteix_pleaneutralcomparison_2013}.
 
@@
 Our target audience consists of computational researchers who are interested in performing a benchmarking study, or who have already begun one.
-Our review spans the full 'pipeline' of benchmarking, from defining the scope to best practices for reproducibility.
+Our review spans the full pipeline of benchmarking, from defining the scope to best practices for reproducibility.
 This includes crucial questions regarding design and evaluation principles: for example, using rankings according to evaluation metrics to identify a set of high-performing methods, and then highlighting different strengths and tradeoffs among these.
@@
 The purpose and scope of a benchmark should be clearly defined at the beginning of the study, and will fundamentally guide the design and implementation.
-In general, we can define three broad types of benchmarking studies: (i) those by method developers, to demonstrate the merits of their approach (e.g., \cite{weber_diffcytdifferentialdiscovery_2019,nowicka_drimseqdirichletmultinomialframework_2016,levine_datadrivenphenotypicdissection_2015,zhou_robustlydetectingdifferential_2014,law_voomprecisionweights_2014}); (ii) neutral studies performed to systematically compare methods for a certain analysis, either conducted directly by an independent group (e.g., \cite{saelens_comparisonsinglecelltrajectory_2019,duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018,saelens_comprehensiveevaluationmodule_2018,weber_comparisonclusteringmethods_2016,korthauer_practicalguidemethods_2019,freytag_comparisonclusteringtools_2018,baruzzo_simulationbasedcomprehensivebenchmarking_2017,kanitz_comparativeassessmentmethods_2015,soneson_comparisonmethodsdifferential_2013,rapaport_comprehensiveevaluationdifferential_2013,dillies_comprehensiveevaluationnormalization_2013}) or in collaboration with method authors (e.g., \cite{sage_quantitativeevaluationsoftware_2015}); or (iii) those organized in the form of a community challenge, such as those from the DREAM \cite{weirauch_evaluationmethodsmodeling_2013,costello_communityeffortassess_2014,kuffner_crowdsourcedanalysisclinical_2015,ewing_combiningtumorgenome_2015,hill_inferringcausalmolecular_2016}, FlowCAP \cite{aghaeepour_criticalassessmentautomated_2013,aghaeepour_benchmarkevaluationalgorithms_2016}, CASP \cite{moult_criticalassessmentmethods_2018,moult_criticalassessmentmethods_2016}, CAMI \cite{sczyrba_criticalassessmentmetagenome_2017}, Assemblathon \cite{earl_assemblathoncompetitiveassessment_2011,bradnam_assemblathonevaluatingnovo_2013}, MAQC/SEQC \cite{maqcconsortium_microarrayqualitycontrol_2006,shi_microarrayqualitycontrol_2010,seqc_comprehensiveassessmentrnaseq_2014}, and GA4GH \cite{krusche_bestpracticesbenchmarking_2019} consortia.
+In general, we can define three broad types of benchmarking studies: (i) those by method developers, to demonstrate the merits of their approach (e.g.
+\cite{weber_diffcytdifferentialdiscovery_2019,nowicka_drimseqdirichletmultinomialframework_2016,levine_datadrivenphenotypicdissection_2015,zhou_robustlydetectingdifferential_2014,law_voomprecisionweights_2014}); (ii) neutral studies performed to systematically compare methods for a certain analysis, either conducted directly by an independent group (e.g.
+\cite{saelens_comparisonsinglecelltrajectory_2019,duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018,saelens_comprehensiveevaluationmodule_2018,weber_comparisonclusteringmethods_2016,korthauer_practicalguidemethods_2019,freytag_comparisonclusteringtools_2018,baruzzo_simulationbasedcomprehensivebenchmarking_2017,kanitz_comparativeassessmentmethods_2015,soneson_comparisonmethodsdifferential_2013,rapaport_comprehensiveevaluationdifferential_2013,dillies_comprehensiveevaluationnormalization_2013}) or in collaboration with method authors (e.g.
+\cite{sage_quantitativeevaluationsoftware_2015}); or (iii) those organized in the form of a community challenge, such as those from the DREAM \cite{weirauch_evaluationmethodsmodeling_2013,costello_communityeffortassess_2014,kuffner_crowdsourcedanalysisclinical_2015,ewing_combiningtumorgenome_2015,hill_inferringcausalmolecular_2016}, FlowCAP \cite{aghaeepour_criticalassessmentautomated_2013,aghaeepour_benchmarkevaluationalgorithms_2016}, CASP \cite{moult_criticalassessmentmethods_2018,moult_criticalassessmentmethods_2016}, CAMI \cite{sczyrba_criticalassessmentmetagenome_2017}, Assemblathon \cite{earl_assemblathoncompetitiveassessment_2011,bradnam_assemblathonevaluatingnovo_2013}, MAQC/SEQC \cite{maqcconsortium_microarrayqualitycontrol_2006,shi_microarrayqualitycontrol_2010,seqc_comprehensiveassessmentrnaseq_2014}, and GA4GH \cite{krusche_bestpracticesbenchmarking_2019} consortia.
 
@@
 By contrast, when introducing a new method, the focus of the benchmark will be on evaluating the relative merits of the new method.
-This may be sufficiently achieved with a less extensive benchmark, e.g., by comparing against a smaller set of state-of-the-art and baseline methods.
+This may be sufficiently achieved with a less extensive benchmark, for example, by comparing against a smaller set of state-of-the-art and baseline methods.
 However, the benchmark must still be carefully designed to avoid disadvantaging any methods; for example, extensively tuning parameters for the new method while using default parameters for competing methods would result in a biased representation.
-Some advantages of a new method may fall outside the scope of a benchmark; for example, a new method may enable more flexible analyses than previous methods (e.g., beyond two-group comparisons in differential analyses \cite{weber_diffcytdifferentialdiscovery_2019}).
+Some advantages of a new method may fall outside the scope of a benchmark; for example, a new method may enable more flexible analyses than previous methods (e.g.
+beyond two-group comparisons in differential analyses \cite{weber_diffcytdifferentialdiscovery_2019}).
 
@@
 A neutral benchmark should include all available methods for a certain type of analysis.
-In this case, the publication describing the benchmark will also function as a review of the literature; a summary table describing the methods is a key output (e.g., Figure 2 in \cite{saelens_comparisonsinglecelltrajectory_2019} or Table 1 in \cite{weber_comparisonclusteringmethods_2016}).
+In this case, the publication describing the benchmark will also function as a review of the literature; a summary table describing the methods is a key output (e.g.
+Figure 2 in \cite{saelens_comparisonsinglecelltrajectory_2019} or Table 1 in \cite{weber_comparisonclusteringmethods_2016}).
 Alternatively, it may make sense to include only a subset of methods, by defining inclusion criteria: for example, all methods that (i) provide freely available software implementations, (ii) are available for commonly used operating systems, and (iii) can successfully be installed without errors following a reasonable amount of trouble-shooting.
@@
 When developing a new method, it is generally sufficient to select a representative subset of existing methods to compare against.
-For example, this could consist of the current best-performing methods (if known), a simple 'baseline' method, and any methods that are widely used.
+For example, this could consist of the current best-performing methods (if known), a simple baseline method, and any methods that are widely used.
 The selection of competing methods should ensure an accurate and unbiased assessment of the relative merits of the new approach, compared with the current state-of-the-art.
@@
 
-Simulated data have the advantage that a known true signal (or 'ground truth') can easily be introduced; for example, whether a gene is differentially expressed.
+Simulated data have the advantage that a known true signal (or ground truth) can easily be introduced; for example, whether a gene is differentially expressed.
 Quantitative performance metrics measuring the ability to recover the known truth can then be calculated.
-However, it is important to demonstrate that simulations accurately reflect relevant properties of real data, by inspecting empirical summaries of both simulated and real datasets (e.g., using automated tools \cite{soneson_unifiedqualityverification_2018}).
+However, it is important to demonstrate that simulations accurately reflect relevant properties of real data, by inspecting empirical summaries of both simulated and real datasets (e.g.
+using automated tools \cite{soneson_unifiedqualityverification_2018}).
 The set of empirical summaries to use is context-specific; for example, for single-cell RNA-sequencing, dropout profiles and dispersion-mean relationships should be compared \cite{soneson_biasrobustnessscalability_2018}; for DNA methylation, correlation patterns among neighboring CpG sites should be investigated \cite{korthauer_detectionaccuratefalse_2019}; for comparing mapping algorithms, error profiles of the sequencing platforms should be considered \cite{caboche_comparisonmappingalgorithms_2014}.
@@
 Experimental data often do not contain a ground truth, making it difficult to calculate performance metrics.
-Instead, methods may be evaluated by comparing them against each other (e.g., overlap between sets of detected differential features \cite{nowicka_drimseqdirichletmultinomialframework_2016}), or against a current widely accepted method or 'gold standard' (e.g., manual gating to define cell populations in high-dimensional cytometry \cite{weber_comparisonclusteringmethods_2016, aghaeepour_criticalassessmentautomated_2013}, or fluorescence in situ hybridization to validate absolute copy number predictions \cite{zheng_benchmarkingcontextsdetails_2017}).
+Instead, methods may be evaluated by comparing them against each other (e.g.
+overlap between sets of detected differential features \cite{nowicka_drimseqdirichletmultinomialframework_2016}), or against a current widely accepted method or gold standard (e.g.
+manual gating to define cell populations in high-dimensional cytometry \cite{weber_comparisonclusteringmethods_2016, aghaeepour_criticalassessmentautomated_2013}, or fluorescence in situ hybridization to validate absolute copy number predictions \cite{zheng_benchmarkingcontextsdetails_2017}).
 In the context of supervised learning, the response variable to be predicted is known in the manually labeled training and test data.
@@
 In some cases, it is also possible to design experimental datasets containing a ground truth.
-Examples include: (i) 'spiking in' synthetic RNA molecules at known relative concentrations \cite{jiang_syntheticspikeinstandards_2011} in RNA-sequencing experiments (e.g., \cite{seqc_comprehensiveassessmentrnaseq_2014,garalde_highlyparalleldirect_2018}), (ii) large-scale validation of gene expression measurements by quantitative polymerase chain reaction (e.g., \cite{seqc_comprehensiveassessmentrnaseq_2014}), (iii) using genes located on sex chromosomes as a proxy for silencing of DNA methylation status (e.g., \cite{law_voomprecisionweights_2014,fang_genomiclandscapehuman_2012}), (iv) using fluorescence-activated cell sorting to sort cells into known subpopulations prior to single-cell RNA-sequencing (e.g., \cite{soneson_biasrobustnessscalability_2018, schaum_singlecelltranscriptomics20_2018, zheng_massivelyparalleldigital_2017}), or (v) mixing different cell lines to create 'pseudo-cells' \cite{tian_benchmarkingsinglecell_2019}.
+Examples include: (i) spiking in synthetic RNA molecules at known relative concentrations \cite{jiang_syntheticspikeinstandards_2011} in RNA-sequencing experiments (e.g.
+\cite{seqc_comprehensiveassessmentrnaseq_2014,garalde_highlyparalleldirect_2018}), (ii) large-scale validation of gene expression measurements by quantitative polymerase chain reaction (e.g.
+\cite{seqc_comprehensiveassessmentrnaseq_2014}), (iii) using genes located on sex chromosomes as a proxy for silencing of DNA methylation status (e.g.
+\cite{law_voomprecisionweights_2014,fang_genomiclandscapehuman_2012}), (iv) using fluorescence-activated cell sorting to sort cells into known subpopulations prior to single-cell RNA-sequencing (e.g.
+\cite{soneson_biasrobustnessscalability_2018, schaum_singlecelltranscriptomics20_2018, zheng_massivelyparalleldigital_2017}), or (v) mixing different cell lines to create pseudo-cells \cite{tian_benchmarkingsinglecell_2019}.
 However, it may be difficult to ensure that the ground truth represents an appropriate level of variability -- for example, the variability of spiked-in material, or whether method performance on cell line data is relevant to outbred populations.
@@
 
-A further technique is to design 'semi-simulated' datasets that combine real experimental data with an 'in silico' (i.e., computational) spike-in signal; for example, by combining cells or genes from 'null' (e.g., healthy) samples with a subset of cells or genes from samples expected to contain a true differential signal (examples include \cite{weber_diffcytdifferentialdiscovery_2019,arvaniti_sensitivedetectionrare_2017,rigaill_syntheticdatasets_2018}).
+A further technique is to design semi-simulated datasets that combine real experimental data with an in silico (i.e., computational) spike-in signal; for example, by combining cells or genes from null (e.g.
+healthy) samples with a subset of cells or genes from samples expected to contain a true differential signal (examples include \cite{weber_diffcytdifferentialdiscovery_2019,arvaniti_sensitivedetectionrare_2017,rigaill_syntheticdatasets_2018}).
 This strategy can create datasets with more realistic levels of variability and correlation, together with a ground truth.
 
-Overall, there is no perfect reference dataset, and the selection of appropriate datasets will involve tradeoffs, e.g., regarding the level of complexity.
-Both simulated and experimental data should not be too 'simple' (e.g., two of the datasets in the FlowCAP-II challenge \cite{aghaeepour_criticalassessmentautomated_2013} gave perfect performance for several algorithms) or too 'difficult' (e.g., for the third dataset in FlowCAP-II, no algorithms performed well); in these situations, it can be impossible to distinguish performance.
-In some cases, individual datasets have also been found to be unrepresentative, leading to over-optimistic or otherwise biased assessment of methods (e.g., \cite{lowes_bralibasedentatale_2017}).
+Overall, there is no perfect reference dataset, and the selection of appropriate datasets will involve tradeoffs, for example, regarding the level of complexity.
+Both simulated and experimental data should not be too simple (e.g.
+two of the datasets in the FlowCAP-II challenge \cite{aghaeepour_criticalassessmentautomated_2013} gave perfect performance for several algorithms) or too difficult (e.g.
+for the third dataset in FlowCAP-II, no algorithms performed well); in these situations, it can be impossible to distinguish performance.
+In some cases, individual datasets have also been found to be unrepresentative, leading to over-optimistic or otherwise biased assessment of methods (e.g.
+\cite{lowes_bralibasedentatale_2017}).
 Overall, the key to truly excellent benchmarking is diversity of evaluations, i.e., using a range of metrics and datasets that span the range of those that might be encountered in practice, so that performance estimates can be credibly extrapolated.
@@
 This relies on familiarity with the methods and the literature, reflecting usage by expert users.
-The third strategy is to use a systematic or automated parameter tuning procedure -- for example, a 'grid search' across ranges of values for multiple parameters or techniques such as cross-validation (e.g., \cite{saelens_comprehensiveevaluationmodule_2018}).
-The strategies may also be combined, e.g., setting non-critical parameters to default values and performing a grid search for key parameters.
+The third strategy is to use a systematic or automated parameter tuning procedure -- for example, a grid search across ranges of values for multiple parameters or techniques such as cross-validation (e.g.
+\cite{saelens_comprehensiveevaluationmodule_2018}).
+The strategies may also be combined, for example, by setting non-critical parameters to default values and performing a grid search for key parameters.
 Regardless, neutrality should be maintained: comparing methods with the same strategy makes sense, while comparing one method with default parameters against another with extensive tuning makes for an unfair comparison.
@@
 However, as outlined above, bias may be introduced by tuning the parameters of the new method more extensively.
-The parameter selection strategy should be transparently discussed during the interpretation of the results, to avoid the risk of over-optimistic reporting due to expending more 'researcher degrees of freedom' on the new method \cite{boulesteix_tensimplerules_2015,hu_parametertuningkey_2019}.
+The parameter selection strategy should be transparently discussed during the interpretation of the results, to avoid the risk of over-optimistic reporting due to expending more researcher degrees of freedom on the new method \cite{boulesteix_tensimplerules_2015,hu_parametertuningkey_2019}.
 
-Software versions can also influence results, especially if updates include major changes to methodology (e.g., \cite{vaquero-garcia_leafcuttervsmajiq_2018}).
+Software versions can also influence results, especially if updates include major changes to methodology (e.g.
+\cite{vaquero-garcia_leafcuttervsmajiq_2018}).
 Final results should generally be based on the latest available versions, which may require re-running some methods if updates become available during the course of a benchmark.
@@
 For example, for classification tasks with a ground truth, metrics include the true positive rate (TPR; sensitivity or recall), false positive rate (FPR; 1 - specificity), and false discovery rate (FDR).
-For clustering tasks, common metrics include the F1 score, adjusted Rand index, normalized mutual information, precision, and recall; some of these can be calculated at the cluster level as well as averaged (and optionally weighted) across clusters (e.g., these metrics were used to evaluate clustering methods in our own work \cite{duo_systematicperformanceevaluation_2018,weber_comparisonclusteringmethods_2016}  and by others \cite{freytag_comparisonclusteringtools_2018,aghaeepour_criticalassessmentautomated_2013,wiwie_comparingperformancebiomedical_2015}).
-Several of these metrics can also be compared visually to capture the tradeoff between sensitivity and specificity, e.g., using receiver operating characteristic (ROC) curves (TPR versus FPR), TPR versus FDR curves, or precision-recall (PR) curves (Figure \ref{fig:fig2_metrics}B).
+For clustering tasks, common metrics include the F1 score, adjusted Rand index, normalized mutual information, precision, and recall; some of these can be calculated at the cluster level as well as averaged (and optionally weighted) across clusters (e.g.
+these metrics were used to evaluate clustering methods in our own work \cite{duo_systematicperformanceevaluation_2018,weber_comparisonclusteringmethods_2016}  and by others \cite{freytag_comparisonclusteringtools_2018,aghaeepour_criticalassessmentautomated_2013,wiwie_comparingperformancebiomedical_2015}).
+Several of these metrics can also be compared visually to capture the tradeoff between sensitivity and specificity, for example, using receiver operating characteristic (ROC) curves (TPR versus FPR), TPR versus FDR curves, or precision-recall (PR) curves (Figure \ref{fig:fig2_metrics}B).
 For imbalanced datasets, PR curves have been shown to be more informative than ROC curves \cite{saito_precisionrecallplotmore_2015,powers_visualizationtradeoffevaluation_2015}.
 These visual metrics can also be summarized as a single number, such as area under the ROC or PR curve; examples from our work include \cite{weber_diffcytdifferentialdiscovery_2019,soneson_biasrobustnessscalability_2018}.
-In addition to the tradeoff between sensitivity and specificity, a method's 'operating point' is important; in particular, whether the threshold used (e.g., 5\% FDR) is calibrated to achieve the specified error rate.
-We often overlay this onto TPR-FDR curves by filled or open circles (e.g., Figure \ref{fig:fig2_metrics}B, generated using the iCOBRA package \cite{soneson_icobraopenreproducible_2016}); examples from our work include \cite{weber_diffcytdifferentialdiscovery_2019,nowicka_drimseqdirichletmultinomialframework_2016,zhou_robustlydetectingdifferential_2014,soneson_differentialanalysesrnaseq_2016}.
+In addition to the tradeoff between sensitivity and specificity, a methods operating point' is important; in particular, whether the threshold used (e.g.
+5\% FDR) is calibrated to achieve the specified error rate.
+We often overlay this onto TPR-FDR curves by filled or open circles (e.g.
+Figure \ref{fig:fig2_metrics}B, generated using the iCOBRA package \cite{soneson_icobraopenreproducible_2016}); examples from our work include \cite{weber_diffcytdifferentialdiscovery_2019,nowicka_drimseqdirichletmultinomialframework_2016,zhou_robustlydetectingdifferential_2014,soneson_differentialanalysesrnaseq_2016}.
 
@@
 
-For methods with continuous-valued output (e.g., effect sizes or abundance estimates), metrics include the root mean square error, distance measures, Pearson correlation, sum of absolute log-ratios, log-modulus, and cross-entropy.
-As above, the choice of metric depends on the type of method and data (e.g., \cite{costello_communityeffortassess_2014,lindgreen_evaluationaccuracyspeed_2016}  used correlation, while \cite{moult_criticalassessmentmethods_2016} used root mean square deviation).
+For methods with continuous-valued output (e.g.
+effect sizes or abundance estimates), metrics include the root mean square error, distance measures, Pearson correlation, sum of absolute log-ratios, log-modulus, and cross-entropy.
+As above, the choice of metric depends on the type of method and data (e.g.
+\cite{costello_communityeffortassess_2014,lindgreen_evaluationaccuracyspeed_2016}  used correlation, while \cite{moult_criticalassessmentmethods_2016} used root mean square deviation).
 Further classes of methods include those generating graphs, phylogenetic trees, overlapping clusters, or distributions; these require more complex metrics.
-In some cases, custom metrics may need to be developed (e.g., we defined new metrics for topologies of developmental trajectories in \cite{saelens_comparisonsinglecelltrajectory_2019}).
-When designing custom metrics, it is important to assess their reliability across a range of prediction values (e.g., \cite{gurevich_quastqualityassessment_2013,narzisi_comparingnovogenome_2011}).
-For some metrics, it may also be useful to assess uncertainty, e.g., via confidence intervals.
-In the context of supervised learning, classification or prediction accuracy can be evaluated by cross-validation, bootstrapping, or on a separate test dataset (e.g., \cite{gardner_identifyingaccuratemetagenome_2019,aghaeepour_benchmarkevaluationalgorithms_2016}).
-In this case, procedures to split data into training and test sets should be appropriate for the data structure and the prediction task at hand (e.g., leaving out whole samples or chromosomes \cite{schreiber_pitfallmachinelearning_2019}).
+In some cases, custom metrics may need to be developed (e.g.
+we defined new metrics for topologies of developmental trajectories in \cite{saelens_comparisonsinglecelltrajectory_2019}).
+When designing custom metrics, it is important to assess their reliability across a range of prediction values (e.g.
+\cite{gurevich_quastqualityassessment_2013,narzisi_comparingnovogenome_2011}).
+For some metrics, it may also be useful to assess uncertainty, for example, via confidence intervals.
+In the context of supervised learning, classification or prediction accuracy can be evaluated by cross-validation, bootstrapping, or on a separate test dataset (e.g.
+\cite{gardner_identifyingaccuratemetagenome_2019,aghaeepour_benchmarkevaluationalgorithms_2016}).
+In this case, procedures to split data into training and test sets should be appropriate for the data structure and the prediction task at hand (e.g.
+leaving out whole samples or chromosomes \cite{schreiber_pitfallmachinelearning_2019}).
 
 Additional metrics that do not rely on a ground truth include measures of stability, stochasticity, and robustness.
-These measures may be quantified by running methods multiple times using different inputs or subsampled data (e.g., we observed substantial variability in performance for some methods in \cite{soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
-'Missing values' may occur if a method does not return any values for a certain metric, e.g., due to a failure to converge or other computational issues such as excessive runtime or memory requirements (e.g., \cite{saelens_comparisonsinglecelltrajectory_2019, soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
+These measures may be quantified by running methods multiple times using different inputs or subsampled data (e.g.
+we observed substantial variability in performance for some methods in \cite{soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
+Missing values may occur if a method does not return any values for a certain metric, for example, due to a failure to converge or other computational issues such as excessive runtime or memory requirements (e.g.
+\cite{saelens_comparisonsinglecelltrajectory_2019, soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
 Fallback solutions such as imputation may be considered in this case \cite{bischl_benchmarkinglocalclassification_2013}, although these should be transparently reported.
-For non-deterministic methods (e.g., with random starts or stochastic optimization), variability in performance when using different random seeds or subsampled data should be characterized.
-Null comparisons can be constructed by randomizing group labels such that datasets do not contain any true signal, which can provide information on error rates (e.g., \cite{weber_diffcytdifferentialdiscovery_2019,zhou_robustlydetectingdifferential_2014,law_voomprecisionweights_2014}).
+For non-deterministic methods (e.g.
+with random starts or stochastic optimization), variability in performance when using different random seeds or subsampled data should be characterized.
+Null comparisons can be constructed by randomizing group labels such that datasets do not contain any true signal, which can provide information on error rates (e.g.
+\cite{weber_diffcytdifferentialdiscovery_2019,zhou_robustlydetectingdifferential_2014,law_voomprecisionweights_2014}).
 However, these must be designed carefully to avoid confounding by batch or population structure, and to avoid strong within-group batch effects that are not accounted for.
@@
 Focusing on a single metric can give an incomplete view: methods may not be directly comparable if they are designed for different tasks, and different users may be interested in different aspects of performance.
-Therefore, a crucial design decision is whether to focus on an overall ranking, e.g., by combining or weighting multiple metrics.
+Therefore, a crucial design decision is whether to focus on an overall ranking, for example, by combining or weighting multiple metrics.
 In general, it is unlikely that a single method will perform best across all metrics, and performance differences between top-ranked methods for individual metrics can be small.
@@
 
-In our experience, runtimes and scalability can vary enormously between methods (e.g., in our work, runtimes for cytometry clustering algorithms \cite{weber_comparisonclusteringmethods_2016} and metagenome analysis tools \cite{lindgreen_evaluationaccuracyspeed_2016} ranged across multiple orders of magnitude for the same datasets).
+In our experience, runtimes and scalability can vary enormously between methods (e.g.
+in our work, runtimes for cytometry clustering algorithms \cite{weber_comparisonclusteringmethods_2016} and metagenome analysis tools \cite{lindgreen_evaluationaccuracyspeed_2016} ranged across multiple orders of magnitude for the same datasets).
 Similarly, memory and other computational requirements can vary widely.
-Runtimes and scalability may be investigated systematically, e.g., by varying the number of cells or genes in a single-cell RNA-sequencing dataset \cite{duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018}.
+Runtimes and scalability may be investigated systematically, for example, by varying the number of cells or genes in a single-cell RNA-sequencing dataset \cite{duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018}.
 In many cases, there is a tradeoff between performance and computational requirements.
@@
 From the developer perspective, code quality and use of software development best practices, such as unit testing and continuous integration, are also important.
-Similarly, adherence to commonly used data formats (e.g., GFF/GTF files for genomic features, BAM/SAM files for sequence alignment data, or FCS files for flow or mass cytometry data) greatly improves accessibility and extensibility.
+Similarly, adherence to commonly used data formats (e.g.
+GFF/GTF files for genomic features, BAM/SAM files for sequence alignment data, or FCS files for flow or mass cytometry data) greatly improves accessibility and extensibility.
 
 High-quality documentation is critical, including help pages and tutorials.
-Ideally, all code examples in the documentation should be continually tested, e.g., as Bioconductor does, or through continuous integration.
+Ideally, all code examples in the documentation should be continually tested, for example, as Bioconductor does, or through continuous integration.
 
@@
 since they are designed for different tasks), and different users may be interested in different aspects of performance.
-In addition, it is unlikely that there will be a clear 'winner' across all criteria, and performance differences between top-ranked methods can be small.
+In addition, it is unlikely that there will be a clear winner across all criteria, and performance differences between top-ranked methods can be small.
 Therefore, an informative strategy is to use the rankings to identify a set of high-performing methods, and to highlight the different strengths and tradeoffs among these methods.
@@
 Visualizations summarizing multiple performance metrics can be highly informative for method users (examples include Figure 3 in \cite{saelens_comparisonsinglecelltrajectory_2019} and Figure 5 in \cite{soneson_biasrobustnessscalability_2018} from our own work; as well as Figure 6 in \cite{korthauer_practicalguidemethods_2019}).
-Summary tables are also useful as a reference (e.g., \cite{weber_comparisonclusteringmethods_2016,aghaeepour_criticalassessmentautomated_2013}).
-Additional visualizations, such as flow charts to guide the choice of method for different analyses, are a helpful way to engage the reader (e.g., Figure 5 in \cite{saelens_comparisonsinglecelltrajectory_2019}).
+Summary tables are also useful as a reference (e.g.
+\cite{weber_comparisonclusteringmethods_2016,aghaeepour_criticalassessmentautomated_2013}).
+Additional visualizations, such as flow charts to guide the choice of method for different analyses, are a helpful way to engage the reader (e.g.
+Figure 5 in \cite{saelens_comparisonsinglecelltrajectory_2019}).
 
@@
 For a benchmark to introduce a new method, the results will form one part of the exposition.
-We highly recommend publishing a preprint prior to peer review (e.g., on bioRxiv or arXiv) to speed up distribution of results, broaden accessibility, and solicit additional feedback.
+We highly recommend publishing a preprint prior to peer review (e.g.
+on bioRxiv or arXiv) to speed up distribution of results, broaden accessibility, and solicit additional feedback.
 In particular, direct consultation with method authors can generate highly useful feedback (examples from our work are described in the acknowledgments in \cite{lindgreen_evaluationaccuracyspeed_2016,freyhult_exploringgenomicdark_2007}).
@@
 In addition to raw data and code, it is useful to distribute pre-processed and/or results data (examples include \cite{duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018,soneson_icobraopenreproducible_2016} from our work and \cite{wiwie_comparingperformancebiomedical_2015,bokulich_mockrobiotapublicresource_2016,conchuir_webresourcestandardized_2015} from others), especially for computationally intensive benchmarks.
-This may be combined with an interactive website, where users can upload results from a new method, to be included in an updated comparison either automatically or by the original authors (e.g., \cite{kanitz_comparativeassessmentmethods_2015,cope_benchmarkaffymetrixgenechip_2004,irizarry_comparisonaffymetrixgenechip_2006}).
-'Continuous' benchmarks, which are continually updated, are especially convenient (e.g., \cite{barton_nucleotidesgenomeassembler_2014}), but may require significant additional effort.
+This may be combined with an interactive website, where users can upload results from a new method, to be included in an updated comparison either automatically or by the original authors (e.g.
+\cite{kanitz_comparativeassessmentmethods_2015,cope_benchmarkaffymetrixgenechip_2004,irizarry_comparisonaffymetrixgenechip_2006}).
+Continuous benchmarks, which are continually updated, are especially convenient (e.g.
+\cite{barton_nucleotidesgenomeassembler_2014}), but may require significant additional effort.
 
@@
 Reproducibility of research findings has become an increasing concern in numerous areas of study \cite{ioannidis_whymostpublished_2005}.
-In computational sciences, reproducibility of code and data analyses has been recognized as a useful 'minimum standard' that enables other researchers to verify analyses \cite{peng_reproducibleresearchcomputational_2011}.
+In computational sciences, reproducibility of code and data analyses has been recognized as a useful minimum standard that enables other researchers to verify analyses \cite{peng_reproducibleresearchcomputational_2011}.
 Access to code and data has previously enabled method developers to uncover potential errors in published benchmarks due to suboptimal usage of methods \cite{vaquero-garcia_leafcuttervsmajiq_2018,zhou_countbaseddifferentialexpression_2015,zhou_mirnaseqnormalizationcomparisons_2013}.
-Journal publication policies can play a crucial role in encouraging authors to follow these practices \cite{hofner_reproducibleresearchstatistics_2016}; experience shows that statements that code and data are 'available on request' are often insufficient \cite{boulesteix_makingcomplexprediction_2019}.
+Journal publication policies can play a crucial role in encouraging authors to follow these practices \cite{hofner_reproducibleresearchstatistics_2016}; experience shows that statements that code and data are available on request are often insufficient \cite{boulesteix_makingcomplexprediction_2019}.
 In the context of benchmarking, code and data availability also provides further benefits: for method users, code repositories serve as a source of annotated code to run methods and build analysis pipelines, while for developers, code repositories can act as a prototype for future method development work.
@@
 Reproducible workflow frameworks, such as the Galaxy platform \cite{afgan_galaxyplatformaccessible_2018}, can also be helpful.
-A summary table or spreadsheet of parameter values and software versions can be published as supplementary information along with the publication describing the benchmark (e.g., Supporting Information Table S1 in our study \cite{weber_comparisonclusteringmethods_2016}).
+A summary table or spreadsheet of parameter values and software versions can be published as supplementary information along with the publication describing the benchmark (e.g.
+Supporting Information Table S1 in our study \cite{weber_comparisonclusteringmethods_2016}).
 
@@
 Examples include SummarizedBenchmark \cite{kimes_reproduciblereplicablecomparisons_2019}, DataPackageR \cite{finak_datapackagerreproducibledata_2018}, workflowr \cite{blischak_workflowrframeworkreproducible_2019}, and Dynamic Statistical Comparisons \cite{wang_dscdynamicstatistical_2016}.
-Some tools (e.g., workflowr) also provide streamlined options for publishing results online.
+Some tools (e.g.
+workflowr) also provide streamlined options for publishing results online.
 In machine learning, OpenML provides a platform to organize and share benchmarks \cite{vanschoren_openmlnetworkedscience_2014}.
 More general tools for managing computational workflows, including Snakemake \cite{koster_snakemakescalablebioinformatics_2012}, Make, Bioconda \cite{gruning_biocondasustainablecomprehensive_2018}, and conda, can be customized to capture setup information.
-Containerization tools such as Docker and Singularity may be used to encapsulate a software environment for each method, preserving the package version as well as dependency packages and the operating system, and facilitating distribution of methods to end users (e.g., in our study \cite{saelens_comparisonsinglecelltrajectory_2019}).
+Containerization tools such as Docker and Singularity may be used to encapsulate a software environment for each method, preserving the package version as well as dependency packages and the operating system, and facilitating distribution of methods to end users (e.g.
+in our study \cite{saelens_comparisonsinglecelltrajectory_2019}).
 Best practices from software development are also useful, including unit testing and continuous integration.
 
-Many free online resources are available for sharing code and data, including GitHub and Bitbucket, repositories for specific data types (e.g., ArrayExpress \cite{kolesnikov_arrayexpressupdatesimplifying_2015}, the Gene Expression Omnibus \cite{barrett_ncbigeoarchive_2013}, and FlowRepository \cite{spidlen_flowrepositoryresourceannotated_2012}), and more general data repositories (e.g., figshare, Dryad, Zenodo, Bioconductor ExperimentHub, and Mendeley Data).
+Many free online resources are available for sharing code and data, including GitHub and Bitbucket, repositories for specific data types (e.g.
+ArrayExpress \cite{kolesnikov_arrayexpressupdatesimplifying_2015}, the Gene Expression Omnibus \cite{barrett_ncbigeoarchive_2013}, and FlowRepository \cite{spidlen_flowrepositoryresourceannotated_2012}), and more general data repositories (e.g.
+figshare, Dryad, Zenodo, Bioconductor ExperimentHub, and Mendeley Data).
 Customized resources (examples from our work include \cite{soneson_biasrobustnessscalability_2018, soneson_icobraopenreproducible_2016}) can be designed when additional flexibility is needed.
-Several repositories allow the creation of 'digital object identifiers' (DOIs) for code or data objects.
+Several repositories allow the creation of digital object identifiers (DOIs) for code or data objects.
 In general, preference should be given to publicly funded repositories, which provide greater guarantees for long-term archival stability \cite{mangul_improvingusabilityarchival_2019,mangul_challengesrecommendationsimprove_2019}.
 
-An extensive literature exists on best practices for reproducible computational research (e.g., \cite{sandve_tensimplerules_2013}).
-Some practices (e.g., containerization) may involve significant additional work; however, in our experience, almost all efforts in this area prove useful, especially by facilitating later extensions by ourselves or other researchers.
+An extensive literature exists on best practices for reproducible computational research (e.g.
+\cite{sandve_tensimplerules_2013}).
+Some practices (e.g.
+containerization) may involve significant additional work; however, in our experience, almost all efforts in this area prove useful, especially by facilitating later extensions by ourselves or other researchers.
 
@@
 However, we have also emphasized that any benchmark will involve tradeoffs, due to limited expertise and resources, and that some principles are less central to the evaluation.
-Table \ref{tab:essential_principles} provides a summary of examples of key tradeoffs and pitfalls related to benchmarking, along with our judgment of how truly 'essential' each principle is.
+Table \ref{tab:essential_principles} provides a summary of examples of key tradeoffs and pitfalls related to benchmarking, along with our judgment of how truly essential each principle is.
 
@@
 \begin{table}[ht!]
-	\caption{\textbf{Summary of our views regarding 'how essential' each principle is for a truly excellent benchmark, along with examples of key tradeoffs and potential pitfalls relating to each principle}.
+	\caption{\textbf{Summary of our views regarding how essential each principle is for a truly excellent benchmark, along with examples of key tradeoffs and potential pitfalls relating to each principle}.
 The higher the number of plus signs, the more central the principle is to the evaluation.} \label{tab:essential_principles}
@@
 		\mycola{3.
-Selection (or design) of datasets} & +++ & \mycolb{Number and types of datasets to include} & Subjectivity in the choice of datasets: e.g., selecting datasets that are unrepresentative of real-world applications \\
+Selection (or design) of datasets} & +++ & \mycolb{Number and types of datasets to include} & Subjectivity in the choice of datasets: e.g.
+selecting datasets that are unrepresentative of real-world applications \\
 		& & & Too few datasets or simulation scenarios Overly simplistic simulations \\ \hline
 		\mycola{4.
-Parameter and software versions} & ++ & \mycolb{Amount of parameter tuning} & Extensive parameter tuning for some methods while using default parameters for others (e.g., competing methods) \\ \hline
+Parameter and software versions} & ++ & \mycolb{Amount of parameter tuning} & Extensive parameter tuning for some methods while using default parameters for others (e.g.
+competing methods) \\ \hline
 		\mycola[3]{5.
-Evaluation criteria: key quantitative performance metrics} & +++ & \mycolb[3]{Number and types of performance metrics} & Subjectivity in the choice of metrics: e.g., selecting metrics that do not translate to real-world performance \\
+Evaluation criteria: key quantitative performance metrics} & +++ & \mycolb[3]{Number and types of performance metrics} & Subjectivity in the choice of metrics: e.g.
+selecting metrics that do not translate to real-world performance \\
 		& & & Metrics that give over-optimistic estimates of performance \\
-		& & & Methods may not be directly comparable according to individual metrics (e.g., if methods are designed for different tasks) \\ \hline
+		& & & Methods may not be directly comparable according to individual metrics (e.g. if methods are designed for different tasks) \\ \hline
 		\mycola[3]{6.
@@
 		\mycola{9.
-Enabling future extensions} & ++ & \mycolb{Amount of resources to dedicate to ensuring extensibility} & Selection of methods or datasets for future extensions may be unrepresentative (e.g., due to requests from method authors) \\ \hline
+Enabling future extensions} & ++ & \mycolb{Amount of resources to dedicate to ensuring extensibility} & Selection of methods or datasets for future extensions may be unrepresentative (e.g.
+due to requests from method authors) \\ \hline
 		\mycola{10. Reproducible research best practices} & ++ & Amount of resources to dedicate to reproducibility & Some tools may not be compatible or accessible several years later \\ \hline

### 2019-09-20 ch8_guidelines/content.tex
change to uk spelling

@@
 Many fields of computational research are characterized by a growing number of available methods for data analysis.
-For example, at the time of writing, almost 400 methods are available for analyzing data from single-cell RNA-sequencing experiments \cite{zappia_exploringsinglecellrnaseq_2018}.
+For example, at the time of writing, almost 400 methods are available for analysing data from single-cell RNA-sequencing experiments \cite{zappia_exploringsinglecellrnaseq_2018}.
 For experimental researchers and method users, this represents both an opportunity and a challenge, since method choice can significantly affect conclusions.
@@
 Our review spans the full pipeline of benchmarking, from defining the scope to best practices for reproducibility.
-This includes crucial questions regarding design and evaluation principles: for example, using rankings according to evaluation metrics to identify a set of high-performing methods, and then highlighting different strengths and tradeoffs among these.
+This includes crucial questions regarding design and evaluation principles: for example, using rankings according to evaluation metrics to identify a set of high-performing methods, and then highlighting different strengths and trade-offs among these.
 
@@
 
-A neutral benchmark or community challenge should be as comprehensive as possible, although for any benchmark there will be tradeoffs in terms of available resources.
+A neutral benchmark or community challenge should be as comprehensive as possible, although for any benchmark there will be trade-offs in terms of available resources.
 To minimize perceived bias, a research group conducting a neutral benchmark should be approximately equally familiar with all included methods, reflecting typical usage of the methods by independent researchers \cite{boulesteix_pleaneutralcomparison_2013}.
@@
 Alternatively, it may make sense to include only a subset of methods, by defining inclusion criteria: for example, all methods that (i) provide freely available software implementations, (ii) are available for commonly used operating systems, and (iii) can successfully be installed without errors following a reasonable amount of trouble-shooting.
-Such criteria should be chosen without favoring any methods, and exclusion of any widely used methods should be justified.
+Such criteria should be chosen without favouring any methods, and exclusion of any widely used methods should be justified.
 A useful strategy can be to involve method authors within the process, since they may provide additional details on optimal usage.
@@
 using automated tools \cite{soneson_unifiedqualityverification_2018}).
-The set of empirical summaries to use is context-specific; for example, for single-cell RNA-sequencing, dropout profiles and dispersion-mean relationships should be compared \cite{soneson_biasrobustnessscalability_2018}; for DNA methylation, correlation patterns among neighboring CpG sites should be investigated \cite{korthauer_detectionaccuratefalse_2019}; for comparing mapping algorithms, error profiles of the sequencing platforms should be considered \cite{caboche_comparisonmappingalgorithms_2014}.
+The set of empirical summaries to use is context-specific; for example, for single-cell RNA-sequencing, drop-out profiles and dispersion-mean relationships should be compared \cite{soneson_biasrobustnessscalability_2018}; for DNA methylation, correlation patterns among neighbouring CpG sites should be investigated \cite{korthauer_detectionaccuratefalse_2019}; for comparing mapping algorithms, error profiles of the sequencing platforms should be considered \cite{caboche_comparisonmappingalgorithms_2014}.
 Simplified simulations can also be useful, to evaluate a new method under a basic scenario, or to systematically test aspects such as scalability and stability.
@@
 overlap between sets of detected differential features \cite{nowicka_drimseqdirichletmultinomialframework_2016}), or against a current widely accepted method or gold standard (e.g.
-manual gating to define cell populations in high-dimensional cytometry \cite{weber_comparisonclusteringmethods_2016, aghaeepour_criticalassessmentautomated_2013}, or fluorescence in situ hybridization to validate absolute copy number predictions \cite{zheng_benchmarkingcontextsdetails_2017}).
-In the context of supervised learning, the response variable to be predicted is known in the manually labeled training and test data.
+manual gating to define cell populations in high-dimensional cytometry \cite{weber_comparisonclusteringmethods_2016, aghaeepour_criticalassessmentautomated_2013}, or fluorescence \textit{in situ} hybridization to validate absolute copy number predictions \cite{zheng_benchmarkingcontextsdetails_2017}).
+In the context of supervised learning, the response variable to be predicted is known in the manually labelled training and test data.
 However, individual datasets should not be overused, and using the same dataset for both method development and evaluation should be avoided, due to the risk of overfitting and overly optimistic results \cite{grimm_evaluationtoolsused_2015,jelizarow_overoptimismbioinformaticsillustration_2010}.
@@
 \cite{seqc_comprehensiveassessmentrnaseq_2014}), (iii) using genes located on sex chromosomes as a proxy for silencing of DNA methylation status (e.g.
-\cite{law_voomprecisionweights_2014,fang_genomiclandscapehuman_2012}), (iv) using fluorescence-activated cell sorting to sort cells into known subpopulations prior to single-cell RNA-sequencing (e.g.
+\cite{law_voomprecisionweights_2014,fang_genomiclandscapehuman_2012}), (iv) using fluorescence-activated cell sorting to sort cells into known sub-populations prior to single-cell RNA-sequencing (e.g.
 \cite{soneson_biasrobustnessscalability_2018, schaum_singlecelltranscriptomics20_2018, zheng_massivelyparalleldigital_2017}), or (v) mixing different cell lines to create pseudo-cells \cite{tian_benchmarkingsinglecell_2019}.
-However, it may be difficult to ensure that the ground truth represents an appropriate level of variability -- for example, the variability of spiked-in material, or whether method performance on cell line data is relevant to outbred populations.
+However, it may be difficult to ensure that the ground truth represents an appropriate level of variability -- for example, the variability of spiked-in material, or whether method performance on cell line data is relevant to out-bred populations.
 Alternatively, experimental datasets may be evaluated qualitatively, for example, by judging whether each method can recover previous discoveries, although this strategy relies on the validity of previous results.
 
-A further technique is to design semi-simulated datasets that combine real experimental data with an in silico (i.e., computational) spike-in signal; for example, by combining cells or genes from null (e.g.
+A further technique is to design semi-simulated datasets that combine real experimental data with an \textit{in silico} (i.e., computational) spike-in signal; for example, by combining cells or genes from null (e.g.
 healthy) samples with a subset of cells or genes from samples expected to contain a true differential signal (examples include \cite{weber_diffcytdifferentialdiscovery_2019,arvaniti_sensitivedetectionrare_2017,rigaill_syntheticdatasets_2018}).
@@
 
-Overall, there is no perfect reference dataset, and the selection of appropriate datasets will involve tradeoffs, for example, regarding the level of complexity.
+Overall, there is no perfect reference dataset, and the selection of appropriate datasets will involve trade-offs, for example, regarding the level of complexity.
 Both simulated and experimental data should not be too simple (e.g.
@@
 Some methods have a large number of parameters, and tuning parameters to optimal values can require significant effort and expertise.
-For a neutral benchmark, a range of parameter values should ideally be considered for each method, although tradeoffs need to be considered regarding available time and computational resources.
-Importantly, the selection of parameter values should comply with the neutrality principle, i.e., certain methods should not be favored over others through more extensive parameter tuning.
+For a neutral benchmark, a range of parameter values should ideally be considered for each method, although trade-offs need to be considered regarding available time and computational resources.
+Importantly, the selection of parameter values should comply with the neutrality principle, i.e., certain methods should not be favoured over others through more extensive parameter tuning.
 
@@
 We used default parameters in several neutral benchmarks where we were interested in performance for untrained users \cite{saelens_comparisonsinglecelltrajectory_2019,couronne_randomforestlogistic_2018,schneider_mortalityriskacute_2016}.
-In addition, for \cite{saelens_comparisonsinglecelltrajectory_2019}, due to the large number of methods and datasets, total runtime was already around a week using 192 processor cores, necessitating judgment in the scope of parameter tuning.
+In addition, for \cite{saelens_comparisonsinglecelltrajectory_2019}, due to the large number of methods and datasets, total runtime was already around a week using 192 processor cores, necessitating judgement in the scope of parameter tuning.
 The second strategy is to choose parameters based on previous experience or published values.
@@
 these metrics were used to evaluate clustering methods in our own work \cite{duo_systematicperformanceevaluation_2018,weber_comparisonclusteringmethods_2016}  and by others \cite{freytag_comparisonclusteringtools_2018,aghaeepour_criticalassessmentautomated_2013,wiwie_comparingperformancebiomedical_2015}).
-Several of these metrics can also be compared visually to capture the tradeoff between sensitivity and specificity, for example, using receiver operating characteristic (ROC) curves (TPR versus FPR), TPR versus FDR curves, or precision-recall (PR) curves (Figure \ref{fig:fig2_metrics}B).
+Several of these metrics can also be compared visually to capture the trade-off between sensitivity and specificity, for example, using receiver operating characteristic (ROC) curves (TPR versus FPR), TPR versus FDR curves, or precision-recall (PR) curves (Figure \ref{fig:fig2_metrics}B).
 For imbalanced datasets, PR curves have been shown to be more informative than ROC curves \cite{saito_precisionrecallplotmore_2015,powers_visualizationtradeoffevaluation_2015}.
 These visual metrics can also be summarized as a single number, such as area under the ROC or PR curve; examples from our work include \cite{weber_diffcytdifferentialdiscovery_2019,soneson_biasrobustnessscalability_2018}.
-In addition to the tradeoff between sensitivity and specificity, a methods operating point' is important; in particular, whether the threshold used (e.g.
+In addition to the trade-off between sensitivity and specificity, a methods operating point' is important; in particular, whether the threshold used (e.g.
 5\% FDR) is calibrated to achieve the specified error rate.
@@
 ROC curves (\textit{left}).
-TPR versus FDR curves (\textit{center}); circles represent observed TPR and FDR at typical FDR thresholds of 1, 5, and 10\%, with filled circles indicating observed FDR lower than or equal to the imposed threshold.
+TPR versus FDR curves (\textit{centre}); circles represent observed TPR and FDR at typical FDR thresholds of 1, 5, and 10\%, with filled circles indicating observed FDR lower than or equal to the imposed threshold.
 PR curves (\textit{right}).
@@
 Additional metrics that do not rely on a ground truth include measures of stability, stochasticity, and robustness.
-These measures may be quantified by running methods multiple times using different inputs or subsampled data (e.g.
+These measures may be quantified by running methods multiple times using different inputs or sub-sampled data (e.g.
 we observed substantial variability in performance for some methods in \cite{soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
@@
 \cite{saelens_comparisonsinglecelltrajectory_2019, soneson_biasrobustnessscalability_2018,weber_comparisonclusteringmethods_2016}).
-Fallback solutions such as imputation may be considered in this case \cite{bischl_benchmarkinglocalclassification_2013}, although these should be transparently reported.
+Fall-back solutions such as imputation may be considered in this case \cite{bischl_benchmarkinglocalclassification_2013}, although these should be transparently reported.
 For non-deterministic methods (e.g.
-with random starts or stochastic optimization), variability in performance when using different random seeds or subsampled data should be characterized.
+with random starts or stochastic optimization), variability in performance when using different random seeds or sub-sampled data should be characterized.
 Null comparisons can be constructed by randomizing group labels such that datasets do not contain any true signal, which can provide information on error rates (e.g.
@@
 Therefore, a good strategy is to use rankings from multiple metrics to identify a set of consistently high-performing methods, and then highlight the different strengths of these methods.
-For example, in \cite{weber_comparisonclusteringmethods_2016}, we identified methods that gave good clustering performance, and then highlighted differences in runtimes among these.
+For example, in \cite{weber_comparisonclusteringmethods_2016}, we identified methods that gave good clustering performance, and then highlighted differences in run-times among these.
 In several studies, we have presented results in the form of a graphical summary of performance according to multiple criteria (examples include Figure 3 in \cite{saelens_comparisonsinglecelltrajectory_2019} and Figure 5 in \cite{soneson_biasrobustnessscalability_2018} from our work; and Figure 2 in \cite{sage_quantitativeevaluationsoftware_2015} and Figure 6 in \cite{korthauer_practicalguidemethods_2019} from other authors).
-Identifying methods that consistently underperform can also be useful, to allow readers to avoid these.
+Identifying methods that consistently under-perform can also be useful, to allow readers to avoid these.
 
@@
 In addition to the key quantitative performance metrics, methods should also be evaluated according to secondary measures, including runtime, scalability, and other computational requirements, as well as qualitative aspects such as user-friendliness, installation procedures, code quality, and documentation quality (Figure \ref{fig:fig2_metrics}A).
-From the user perspective, the final choice of method may involve tradeoffs according to these measures: an adequately performing method may be preferable to a top-performing method that is especially difficult to use.
+From the user perspective, the final choice of method may involve trade-offs according to these measures: an adequately performing method may be preferable to a top-performing method that is especially difficult to use.
 
-In our experience, runtimes and scalability can vary enormously between methods (e.g.
-in our work, runtimes for cytometry clustering algorithms \cite{weber_comparisonclusteringmethods_2016} and metagenome analysis tools \cite{lindgreen_evaluationaccuracyspeed_2016} ranged across multiple orders of magnitude for the same datasets).
+In our experience, run-times and scalability can vary enormously between methods (e.g.
+in our work, run-times for cytometry clustering algorithms \cite{weber_comparisonclusteringmethods_2016} and meta-genome analysis tools \cite{lindgreen_evaluationaccuracyspeed_2016} ranged across multiple orders of magnitude for the same datasets).
 Similarly, memory and other computational requirements can vary widely.
-Runtimes and scalability may be investigated systematically, for example, by varying the number of cells or genes in a single-cell RNA-sequencing dataset \cite{duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018}.
-In many cases, there is a tradeoff between performance and computational requirements.
+Run-times and scalability may be investigated systematically, for example, by varying the number of cells or genes in a single-cell RNA-sequencing dataset \cite{duo_systematicperformanceevaluation_2018,soneson_biasrobustnessscalability_2018}.
+In many cases, there is a trade-off between performance and computational requirements.
 In practice, if computational requirements for a top-performing method are prohibitive, then a different method may be preferred by some users.
@@
 In addition, it is unlikely that there will be a clear winner across all criteria, and performance differences between top-ranked methods can be small.
-Therefore, an informative strategy is to use the rankings to identify a set of high-performing methods, and to highlight the different strengths and tradeoffs among these methods.
+Therefore, an informative strategy is to use the rankings to identify a set of high-performing methods, and to highlight the different strengths and trade-offs among these methods.
 The interpretation may also involve biological or other domain knowledge to establish the scientific relevance of differences in performance.
@@
 Figure 3 displays an example of an interactive website from one of our benchmarks \cite{saelens_comparisonsinglecelltrajectory_2019}, which facilitates exploration of results and assists users with choosing a suitable method.
-While tradeoffs should be considered in terms of the amount of work required, these efforts are likely to have significant benefit for the community.
+While trade-offs should be considered in terms of the amount of work required, these efforts are likely to have significant benefit for the community.
 
@@
 on bioRxiv or arXiv) to speed up distribution of results, broaden accessibility, and solicit additional feedback.
-In particular, direct consultation with method authors can generate highly useful feedback (examples from our work are described in the acknowledgments in \cite{lindgreen_evaluationaccuracyspeed_2016,freyhult_exploringgenomicdark_2007}).
+In particular, direct consultation with method authors can generate highly useful feedback (examples from our work are described in the acknowledgements in \cite{lindgreen_evaluationaccuracyspeed_2016,freyhult_exploringgenomicdark_2007}).
 Finally, at publication time, considering open access options will further broaden accessibility.
@@
 In our view, elements of all of these principles are essential.
-However, we have also emphasized that any benchmark will involve tradeoffs, due to limited expertise and resources, and that some principles are less central to the evaluation.
-Table \ref{tab:essential_principles} provides a summary of examples of key tradeoffs and pitfalls related to benchmarking, along with our judgment of how truly essential each principle is.
+However, we have also emphasized that any benchmark will involve trade-offs, due to limited expertise and resources, and that some principles are less central to the evaluation.
+Table \ref{tab:essential_principles} provides a summary of examples of key trade-offs and pitfalls related to benchmarking, along with our judgement of how truly essential each principle is.
 
@@
 \begin{table}[ht!]
-	\caption{\textbf{Summary of our views regarding how essential each principle is for a truly excellent benchmark, along with examples of key tradeoffs and potential pitfalls relating to each principle}.
+	\caption{\textbf{Summary of our views regarding how essential each principle is for a truly excellent benchmark, along with examples of key trade-offs and potential pitfalls relating to each principle}.
 The higher the number of plus signs, the more central the principle is to the evaluation.} \label{tab:essential_principles}
@@
 		\hline
-		\textbf{Principle} & \textbf{How} \mbox{\textbf{essential}}  & \textbf{Tradeoffs} & \textbf{Potential pitfalls} \\ \hline \hline
+		\textbf{Principle} & \textbf{How} \mbox{\textbf{essential}}  & \textbf{Trade-offs} & \textbf{Potential pitfalls} \\ \hline \hline
 		\mycola[2]{1. Defining the purpose and score} & +++ & \mycolb[2]{How comprehensive the benchmark should be} & Scope too broad: too much work given available resources \\ 

### 2019-09-20 boek.tex
added guidelines for benchmarking
update
rewrite research context and objectives

@@
 \usepackage{tcolorbox}
+
+% allow to adjust spacing in enumerates and itemizes
+\usepackage{enumitem}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%         CH4: DYNO         %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
-\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
-\label{chap:dyno}
-\begin{quote}
-	\textbf{Abstract:} \blindtext
-\end{quote}
-\vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.
-\include{ch4_dyno/content}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%         CH4: DYNO         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\newpage{\thispagestyle{empty}\cleardoublepage}
+%\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
+%\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
+%\label{chap:dyno}
+%\begin{quote}
+%	\textbf{Abstract:} \blindtext
+%\end{quote}
+%\vfill
+%Adapted from:\\
+%\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.
+%\include{ch4_dyno/content}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH7: INCGRAPH       %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{incgraph: Optimising regulatory networks} 
-\chaptermark{incgraph: Optimising regulatory networks.} % short title
-\label{chap:incgraph}
-\graphicspath{{./ch7_incgraph/}}
-\begin{quote}
-	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
-	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
-	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%       CH7: INCGRAPH       %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\newpage{\thispagestyle{empty}\cleardoublepage}
+%\chapter{incgraph: Optimising regulatory networks} 
+%\chaptermark{incgraph: Optimising regulatory networks.} % short title
+%\label{chap:incgraph}
+%\graphicspath{{./ch7_incgraph/}}
+%\begin{quote}
+%	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
+%	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
+%	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
 \\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
-	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
-	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
-	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
+%	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
+%	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
+%	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
 \\
-	IncGraph is freely available as an open-source R package on CRAN (incgraph). The development version is also available on GitHub (\href{http://www.github.com/rcannood/incgraph}{rcannood/incgraph}).
-\end{quote}
-\vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y.
+%	IncGraph is freely available as an open-source R package on CRAN (incgraph). The development version is also available on GitHub (\href{http://www.github.com/rcannood/incgraph}{rcannood/incgraph}).
+%\end{quote}
+%\vfill
+%Adapted from:\\
+%\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y.
 IncGraph: Incremental graphlet counting for topology optimisation.
@@
 \doi{10.1371/journal.pone.0195997}.
-\newpage
-\include{ch7_incgraph/content}
+%\newpage
+%\include{ch7_incgraph/content}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%      CH9: DISCUSSION      %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{Discussion} 
+\chaptermark{Discussion.} % short title
+\label{chap:discussion}
+\include{ch9_discussion/content}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%      CH8: DISCUSSION      %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Discussion and Future Perspectives} 
-\chaptermark{Discussion and Future Perspectives.} % short title
-\label{chap:discussion}
-\include{ch8_discussion/content}
-
 \include{chX_backmatter}

### 2019-09-20 ch1_introduction/content.tex
rewrite research context and objectives

@@
 
-\section{Research objectives}
+\section{Research context and objectives}
+Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology.
+Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication.
+Hundreds of new software tools were developed \cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g.
+differential expression, dimensionality reduction, normalisation).
 
-Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology.
-The new types of analyses made possible warrant the development of computational tools which can solve the problem at hand and deal with the new data characteristics, ultimately in order to create useful and accurate hypotheses for biologists.
+One major shortcoming during the advent of single cell omics was that majority of the newly developed computational tools were not quantitatively and comparatively evaluated.
+Rather, they relied on anecdotal evidence to demonstrate its usefulness.
+This issue is not the result of the tool developer's malevolence, but instead of the lack of data required to perform such comprehensive benchmarks.
 
-In response, scientists all over the globe have been working hard to tackle these computational challenges, resulting in more than 470 software tools for processing single-cell omics data in one form or another \cite{zappia_exploringsinglecellrnaseq_2018}.
-More than half of these tools were published in peer-reviewed journals.
+Uncontrolled development of software tools without comprehensive benchmarking poses serious problems. 
+For one, it slows down scientific progress.
+Every end-user needs to make a large commitment researching the domain in order to make an informed decision of which tool to use, or risk a higher incidence of false positive discoveries (either way, valuable resources are being wasted).
+It also negatively impacts the credibility of the field and discourages new users.
 
-The high number of tools available presents the user choice, but also uncertainty and indecision about which tool to use for the task at hand.
-Choosing the method with most citations might result in a weaker result than picking a method at random, and making an informed decision requires spending an enormous amount of time in comparing the results of numerous methods, on various platforms, on multiple different datasets.
+In this work, we aim to speed up scientific progress in single-cell omics by providing end-users with guidelines on how to use which state-of-the-art tools, and by providing developers with the necessary tools to assert a mimimum performance criterion when introducing new methods.
 
-The field of single-cell omics needs quantitative ways to measure the performance of different categories of tools.
-Clear quantitative metrics allow performing comprehensive studies to assess the advantages and disadvantages of existing tools.
-In addition, they could also serve as a minimum performance criterion before the publication of a new tool.
+We develop benchmarking strategies for assessing the performance of novel computational tools constrained by low availability of real data (Chapter~\ref{chap:dyngen}).
+\textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
 
-In the computational jungle of single-cell omics, experts act as beacons of hope by sharing guidelines based on comprehensive benchmarking studies.
-Their disseminations (in the form of manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}) are crucial in leading new users, and ultimately the whole field, to better practices for performing single-cell omics analyses.
+We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}).
+Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking.
 
-% todo: provide better segway to the next section
+We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
+Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
 
-\section{Outline} % include contributions?
-
-\mycomment{Get a better focus for the overall contributions of the thesis. For example, focus on datasets that contain developing cells, and that the tools we develop make this assumption.}
-
-This work aims to tackle multiple computational problems in extracting biological hypotheses from single-cell omics data. 
-We introduce a framework for evaluating multiple types of single-cell omics tools, more specifically dimensionality reduction, trajectory inference, and gene regulatory network inference (Chapter \ref{chap:dyngen}).
-Next, we perform a comprehensive evaluation of 45 TI methods and construct guidelines based on our observations (Chapter \ref{chap:dynbenchmark}).
-As part of the evaluation, we developed a framework to infer, visualise, and interpret trajectories, which we made available as a separate tool called dyno (Chapter \ref{chap:dyno}).
-We also introduce two new computational tools of our own.
-SCORPIUS is a TI method specialised in inferring linear trajectories (Chapter \ref{chap:scorpius}).
-bred is a true single-cell NI method which can infer gene regulatory networks for individual cells (Chapter \ref{chap:bred}).
-
-\textbf{Chapter \ref{chap:dyngen}} -- why dyngen, what is it, what can it be used for.
-%framework for evaluating methods with synthetic data generator. if available, real data with gold standards would be most desirable.
-
-\textbf{Chapter \ref{chap:dynbenchmark}} -- why dynbenchmark, what is it, what can it be used for.
-
-\textbf{Chapter \ref{chap:dyno}} -- why dyno, what is it, what can it be used for.
-
-\textbf{Chapter \ref{chap:scorpius}} -- why scorpius, what is it, what can it be used for.
-
-\textbf{Chapter \ref{chap:bred}} -- why bred, what is it, what can it be used for.
+We conclude by inventing a new type of approach aimed at inferring the GRN of individual cells (Chapter~\ref{chap:bred}).
+Due a lack of real data which can be used as a gold standard dataset, we use \textit{in silico} single-cell data to quantitatively evaluate the performance.
 
@@
 \begin{table}[ht!]
-	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the role within the package: aut Author, ctb Contributor.
+	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor.
 Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-09-10.
@@
 		\cranpkg{SCORPIUS} & aut & CRAN & 4772 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
-		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} & ctb & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
+
+		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
 		\cranpkg{devtools} & ctb & CRAN & 3'775'350 & Tools to make developing R packages easier \\

### 2019-09-21 ch1_introduction/content.tex
update

@@
 Every end-user needs to make a large commitment researching the domain in order to make an informed decision of which tool to use, or risk a higher incidence of false positive discoveries (either way, valuable resources are being wasted).
-It also negatively impacts the credibility of the field and discourages new users.
+In addition, it also negatively impacts the credibility of the field, thus discouraging potential users or researchers from entering.
 
@@
   \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
-	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
-SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
-bioRxiv.
-2016 Oct:079509.
   \item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
   \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
-  \item dyngen, submitted?
-  \item dyno, submitted?
-  \item scorpius, submitted?
-  \item bred, submitted?
+  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
+  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
+  \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
+  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. bred: Inferring single cell regulatory networks. In preparation.
  \end{itemize}
@@
 2019 Dec;20(1):125.
+  \item Lorenzi L, \ldots, \textbf{Cannoodt R}, \ldots, Mestdagh P. The RNA-Atlas, a single nucleotide resolution map of the human transcriptome. In preparation.
+  \item Van den Berge K, Roux de Bézieux H, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S. Trajectory-based differential expression analysis. Submitted to Nature Communications.
+  \item Van de Sande Bram, \ldots, \textbf{Cannoodt R}, \ldots, Saeys Y, Aerts S. A scalable SCENIC workflow for single-cell gene regulatory network analysis. Submitted to Nature Protocols.
 \end{itemize}
-
-\mycomment{
-	Submitted:\\
-	RNA atlas,\\
-	TRADE-seq: Trajectory-based differential expression analysis,\\
-	SCENIC protocol
-}
 

### 2019-09-21 ch1_introduction/scripts/eval_checks.R

@@ -1,0 +1,151 @@
+library(tidyverse)
+
+# qc_checks <- read_rds(dynbenchmark::result_file("qc_checks.rds", "03-methods")) %>%
+#   filter(aspect_id == "evaluation", check_id == 48)
+evals <- read_rds(dynbenchmark::result_file("tool_qc.rds", "03-methods")) %>%
+  filter(check_id == 48) %>%
+  select(tool_id, eval_answer = answer)
+tools %>% arrange(manuscript_date) %>% pull(tool_id)
+extra_evals <-
+  tribble(
+    ~tool_id, ~num_methods, ~real_datasets, ~simulated_datasets,
+    "monocle1", 0, 0, 0,
+    "wanderlust", 0, 0, 0,
+    "scuba", 0, 0, 0,
+    "sincell", 0, 0, 0,
+    "nbor", 0, 0, 0,
+    "cycler", 0, 0, 0,
+    "oscope", 0, 0, 0,
+    "waterfall", 0, 0, 0,
+    "gpseudotime", 0, 0, 0,
+    "embeddr", 0, 0, 0,
+    "eclair", 0, 0, 0,
+    "dpt", 2, 1, 0,
+    "pseudogp", 0, 0, 0,
+    "slicer", 0, 0, 0,
+    "scell", 0, 0, 0,
+    "wishbone", 0, 0, 0,
+    "tscan", 5, 3, 0, #sim?
+    "scoup", 3, 2, 0, #sim?
+    "delorean", 0, 0, 0,
+    "raceid_stemid", 0, 0, 0,
+    "ouija", 0, 0, 0,
+    "mpath", 0, 0, 0,
+    "celltree", 3, 1, 0, #sim?
+    "wavecrest", 0, 0, 0,
+    "stemnet", 0, 0, 0,
+    "scimitar", 3, 2, 0, #sim?
+    "scorpius", 4, 10, 0,
+    "scent", 3, 5, 0,
+    "slice", 0, 0, 0,
+    "topslam", 0, 0, 0,
+    "monocle", 4, 1, 0,
+    "gpfates", 0, 0, 0,
+    "mfa", 0, 0, 0,
+    "tasic", 0, 0, 0,
+    "somsc", 0, 0, 0,
+    "slingshot", 5, 0, 3,
+    "sctda", 4, 0, 2,
+    "uncurl", 0, 0, 0,
+    "recat", 6, 10, 0,
[…]

### 2019-09-24 ch1_introduction/content.tex
update

@@
 
-One major shortcoming during the advent of single cell omics was that majority of the newly developed computational tools were not quantitatively and comparatively evaluated.
+One major shortcoming during the advent of single-cell omics was that majority of the newly developed computational tools were not quantitatively and comparatively evaluated.
 Rather, they relied on anecdotal evidence to demonstrate its usefulness.
 This issue is not the result of the tool developer's malevolence, but instead of the lack of data required to perform such comprehensive benchmarks.
+%TODO: also mention guidelines for users.
 

### 2019-09-24 ch9_discussion/content.tex
rewrite research context and objectives
update

--- 2019-09-19 ch8_discussion/content.tex
@@
-\section{Essential guidelines for computational method benchmarking}
+%dus was ik aan het nadenken over de stand van zaken na dynbenchmark ─ oké, ik heb zin om een dynbenchmark2 te doen moet, alles meer en beter
+%maar hoe hebben we een impact gehad? als je kijkt naar course material over trajectory inference zou ik zeggen, veel ─ we zijn een gemakkelijke bron van guidelines
+%maar als ik kijk naar de manuscripten van nieuwe TI methoden zou ik zeggen dat we nog niet zijn waar we wouden geraken
+
+\section{Self-assessment in trajectory inference}
+Many trajectory inference (TI) tools lack quantitative assessment of the accuracy of the method.
+Instead, they rely on anecdotal evidence to demonstrate their added value.
+A brief review of 76 TI articles reveals that only about 35\% contain a self-assessment (Figure \ref{fig:benchmarks_over_time}).
+Peer-reviewed TI articles fared even worse, self-assessing in only 32\% of cases (n=52), whereas articles first published as a pre-print self-assess in 42\% of cases (n=40).
+Only four TI articles feature a decently-sized comparison of at least 5 methods using 5 datasets or more.
+% Note: i did not count how many metrics each of these manuscripts used.
+
+Self-assessments are universally biased in favour of the authors, intentionally or otherwise, but this does not mean they should be skipped out on.
+Norel et al.
+provide a valuable discussion on overcoming overly optimistic self-assessments \cite{norel_selfassessmenttrapcan_2011}.
+% TODO: improve working "this does not mean they should be skipped out on"
+
+In this perspective, we hypothesise that low self-assessment rates are primarily caused by a lack of a standardised problem definition, benchmarking datasets, and metrics. 
+We elaborate on these reasons and viable solutions to easy benchmarking trajectory inference.
+
+%in deze commentary geven we nog een kort overzicht van:
+%* standaard definitie van een trajectory inference probleem
+%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
+%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
+%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.75\linewidth]{fig/self_assessment.pdf} 
+	\caption{\textbf{The number of manuscripts featuring trajectory inference tools over time.} Areas coloured in darker shades of blue represent manuscripts containing self-assessment of the tool, highlighting \textbf{A}: the number of datasets used and \textbf{B}: the number of methods used (including the method of the authors).}
+	\label{fig:benchmarks_over_time}
+\end{figure}
+
+\subsection{Problem definition}
+One main reason why benchmarking TI methods is difficult is due to there being slight variations 
+of the problem a method is attempting to solve (Figure \ref{fig:method_types}A).
+For example, a method might infer linear or cyclic trajectories, or predict the probability of a cell ending up in one of several end states.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.8\linewidth]{fig/method_types.pdf} 
+	\caption{\textbf{Different forms of trajectory inference cause }.}
+	\label{fig:method_types}
+\end{figure}
+
+As a result, it becomes harder to discover similar methods to compare against, as certain articles might only show up with search terms such as "pseudotemporal ordering", "lineage trees" or "fate bias".
+For the discoverability of a new TI method, it is therefore essential to use the term "trajectory inference", or at least list it as one of the keywords.
+
+A more significant and harder to solve problem is that the data formats produced by different methods varies greatly.
+This makes visualising and comparing multiple trajectories difficult, as different output types cannot be compared directly.
+The most commonly used and general is one where cells are positioned along a set of edges connecting milestones (Figure~\ref{fig:method_types}B).
+In practice, this format consists of two data structures: the milestone network specify transition between cell states, and the cell progressions specify how far along each cell has progressed along a transition (Figure~\ref{fig:method_types}C).
+Optionally, a cell can be in a region of "delayed commitment", that is, it is not known which state will end up in but a probability for ending up in either state can be assigned (Figure~\ref{fig:method_types}D).
+
+Saelens et al.
+\cite{saelens_comparisonsinglecelltrajectory_2019} developed an R package, \texttt{dynwrap}, for converting many of the variations of trajectory inference to the common format.
+Using this standardised format allows developing reusable software for interpreting, comparing, and visualising trajectories from any TI method.
+
+\subsection{Benchmarking datasets}
+
+\subsection{Multiple metrics}
+
+%waarom benchmarks maar een laag aantal datasets hebben: 
+%* echte datasets zijn duur om te genereren
+%* het is niet evident om ze te verzamelen
+%* otoh, we zouden naar de meta-informatie van de dynbenchmark/real datasets kunnen kijken en een cumulatieve plot genereren van hoeveel datasets er beschikbaar zijn over tijd, om aan te tonen dat sinds 2016-2017 er feitelijk wel al genoeg datasets zijn om aan self-benchmarking te doen
+
+%waarom benchmarks een laag aantal methoden hebben:
+%* iedere methode heeft zijn eigen set van functies
+%* methodes hebben vaak andere prior inputs (starting cell, time series, etc)
+%* methodes zijn vaak in verschillende talen geschreven (R, python, matlab)
+%* de output van methodes zijn vaak niet rechtstreeks vergelijkbaar, omdat er geen standaard definitie van wat trajectory inference is (zoals bv.
+wel het geval is bij clustering of differential expression)
+%
+%bovendien is trajectory inference een probleem met complexe output, waardoor geen kant-en-klare metrieken bestonden om de predicties mee te evalueren.
+ik zou nog eens door alle manuscripten die een evaluatie hebben gedaan, moeten gaan, maar ik denk dat alle of toch de meeste evaluaties tussen 2016 en 2018 enkel op basis van pseudotime zijn ─ ook voor bv.
+dpt, en dat er maar zelden evaluaties zijn die echt gaan kijken naar de branching
+%
+%
+%kaart aan dat we in 2018 een benchmark preprint hebben gepost van W methoden op X datasets, en in 2019 een artikel hebben gepubliceerd van Y methoden op Z datasets.
+Ondanks dat we in deze manuscripten de grootste problemen (# datasets, # methodes, # metrieken, leg iets beter uit) aanpakken, zien we geen vermindering in het aantal nieuwe publicaties die geen self-benchmark bevatten.
+%in deze commentary geven we nog een kort overzicht van:
+%* standaard definitie van een trajectory inference probleem
+%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
+%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
+%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
+%in conclusie, met al deze informatie melden we dat in 2019 er dus geen enkele reden is om ons te gedragen als gelijk welk ander computationeel / data mining / machine learning veld waarbij er voor quantitatief bewijs wordt gevraagd dat je methode goed werkt als je een nieuwe methode wilt publiceren.
+we vragen vriendelijk aan peer-reviewers om kritisch te zijn over deze kwestie.
+
+%in deze commentary geven we nog een kort overzicht van:
+%* standaard definitie van een trajectory inference probleem
+%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
+%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
+%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
+
+%in conclusie, met al deze informatie melden we dat in 2019 er dus geen enkele reden is om ons te gedragen als gelijk welk ander computationeel / data mining / machine learning veld waarbij er voor quantitatief bewijs wordt gevraagd dat je methode goed werkt als je een nieuwe methode wilt publiceren.
+we vragen vriendelijk aan peer-reviewers om kritisch te zijn over deze kwestie.
+
+%refer to: \cite{norel_selfassessmenttrapcan_2011} 
+%In order to alleviate the overestimation of accuracy from the many bias sources described above, we proposed a few guidelines:
+%* use third‐party validation to test a model with previously unseen data
+%* use more than one metric to evaluate the methods
+%* report well‐performing methods even if they are not the best performers on a particular data set
+%* increase the awareness of editors and reviewers that superior performance in self‐assessment is a biased demonstration of the method's value; instead, impartial assessment should be the preferred evaluation
+%* Establish a scientific culture that values timely, well‐conducted follow‐up studies that confirm or refute previous results
+
+%Before 2016, only a handful of trajectory inference methods had been released. None of these articles contained a self-benchmark simply because
+%
+%In all fairness towards the early TI methods, in 2016 or even 2017 there was a low abundance of good datasets to perform benchmarks with.
+The amount of data required to perform a comprehensive evaluation of their tool did not yet exist and was too expensive to generate.
+In April 2016, we started developing a single-cell simulator that would allow us to perform a comprehensive benchmark of TI methods.
+%
+%
+%It was also non-trivial to compare against many methods as each method has a unique interface where some prior knowledge might be needed. 
+%
+
+% * Single cell omics is expanding at rates faster than any single mind can try to conceive
+%Pioneers developing revolutionary tools to process new types of data are typically faced with the problem on how to quantitatively evaluate the accuracy of their approach.
+After all, the amount of data required to perform a comprehensive evaluation of their tool does not yet exist and is too expensive to generate at the time.
+
+%In the computational jungle of single-cell omics, experts act as beacons of hope by sharing guidelines based on comprehensive benchmarking studies.
+Their disseminations (in the form of manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}) are crucial in leading new users, and ultimately the whole field, to better practices for performing single-cell omics analyses.
 
@@
 Its results should not only be convincing and easy to interpret, but also accurate, robust, and reproducible.
-We value creating open-source software that works reliably but fails gracefully when it does not.
+We open source.
+Our software should work reliably and fail gracefully when it does not.
 We acknowledge that writing automated tests is dull but necessary for maintaining long-term software projects.
@@
 \section{dyngen discussion}
+% felt that upon the development of a new technologies, good quality control practices from
+% other technologies were not being carried over because the data required to evaluate computational
+% tools does not exist yet and is too expensive to develop.
 
@@
 
-\section{incgraph discussion}
+%\section{incgraph discussion}
 
-\section{dyno discussion}
-% notice that normalisation was absent from this dissertation.
+%\section{dyno discussion}

### 2019-09-24 ch1_introduction/scripts/eval_checks.R

@@ -1,151 +1,0 @@
-library(tidyverse)
-
-# qc_checks <- read_rds(dynbenchmark::result_file("qc_checks.rds", "03-methods")) %>%
-#   filter(aspect_id == "evaluation", check_id == 48)
-evals <- read_rds(dynbenchmark::result_file("tool_qc.rds", "03-methods")) %>%
-  filter(check_id == 48) %>%
-  select(tool_id, eval_answer = answer)
-tools %>% arrange(manuscript_date) %>% pull(tool_id)
-extra_evals <-
-  tribble(
-    ~tool_id, ~num_methods, ~real_datasets, ~simulated_datasets,
-    "monocle1", 0, 0, 0,
-    "wanderlust", 0, 0, 0,
-    "scuba", 0, 0, 0,
-    "sincell", 0, 0, 0,
-    "nbor", 0, 0, 0,
-    "cycler", 0, 0, 0,
-    "oscope", 0, 0, 0,
-    "waterfall", 0, 0, 0,
-    "gpseudotime", 0, 0, 0,
-    "embeddr", 0, 0, 0,
-    "eclair", 0, 0, 0,
-    "dpt", 2, 1, 0,
-    "pseudogp", 0, 0, 0,
-    "slicer", 0, 0, 0,
-    "scell", 0, 0, 0,
-    "wishbone", 0, 0, 0,
-    "tscan", 5, 3, 0, #sim?
-    "scoup", 3, 2, 0, #sim?
-    "delorean", 0, 0, 0,
-    "raceid_stemid", 0, 0, 0,
-    "ouija", 0, 0, 0,
-    "mpath", 0, 0, 0,
-    "celltree", 3, 1, 0, #sim?
-    "wavecrest", 0, 0, 0,
-    "stemnet", 0, 0, 0,
-    "scimitar", 3, 2, 0, #sim?
-    "scorpius", 4, 10, 0,
-    "scent", 3, 5, 0,
-    "slice", 0, 0, 0,
-    "topslam", 0, 0, 0,
-    "monocle", 4, 1, 0,
-    "gpfates", 0, 0, 0,
-    "mfa", 0, 0, 0,
-    "tasic", 0, 0, 0,
-    "somsc", 0, 0, 0,
-    "slingshot", 5, 0, 3,
-    "sctda", 4, 0, 2,
-    "uncurl", 0, 0, 0,
-    "recat", 6, 10, 0,
[…]

### 2019-09-24 ch9_discussion/scripts/eval_checks.R

@@ -1,0 +1,300 @@
+library(tidyverse)
+
+extra_evals <-
+  tribble(
+    ~tool_id, ~num_methods, ~real_datasets, ~simulated_datasets,
+    "monocle1", 0, 0, 0,
+    "wanderlust", 0, 0, 0,
+    "scuba", 0, 0, 0,
+    "sincell", 0, 0, 0,
+    "nbor", 0, 0, 0,
+    "cycler", 0, 0, 0,
+    "oscope", 0, 0, 0,
+    "waterfall", 0, 0, 0,
+    "gpseudotime", 0, 0, 0,
+    "embeddr", 0, 0, 0,
+    "eclair", 0, 0, 0,
+    "dpt", 2, 1, 0,
+    "pseudogp", 0, 0, 0,
+    "slicer", 0, 0, 0,
+    "scell", 0, 0, 0,
+    "wishbone", 0, 0, 0,
+    "tscan", 5, 3, 0, #sim?
+    "scoup", 3, 2, 0, #sim?
+    "delorean", 0, 0, 0,
+    "raceid_stemid", 0, 0, 0,
+    "ouija", 0, 0, 0,
+    "mpath", 0, 0, 0,
+    "celltree", 3, 1, 0, #sim?
+    "wavecrest", 0, 0, 0,
+    "stemnet", 0, 0, 0,
+    "scimitar", 3, 2, 0, #sim?
+    "scorpius", 4, 10, 0,
+    "scent", 3, 5, 0,
+    "slice", 0, 0, 0,
+    "topslam", 0, 0, 0,
+    "monocle", 4, 1, 0,
+    "gpfates", 0, 0, 0,
+    "mfa", 0, 0, 0,
+    "tasic", 0, 0, 0,
+    "somsc", 0, 0, 0,
+    "slingshot", 5, 0, 3,
+    "sctda", 4, 0, 2,
+    "uncurl", 0, 0, 0,
+    "recat", 6, 10, 0,
+    "forks", 8, 6, 0,
+    "matcher", 1, 1, 1,
+    "phenopath", 0, 0, 0,
+    "hopland", 7, 6, 5,
+    "soptsc", 3, 4, 0,
+    "pba", 0, 0, 0,
+    "bgp", 3, 0, 1,
+    "brgps", 7, 2, 0,
+    "wot", 0, 0, 0,
+    "treetop", 0, 0, 0,
+    "paga", 0, 0, 0,
[…]

### 2019-09-25 ch9_discussion/content.tex
update

@@
 	\centering
-	\includegraphics[width=.8\linewidth]{fig/method_types.pdf} 
-	\caption{\textbf{Different forms of trajectory inference cause }.}
+	\includegraphics[width=\linewidth]{fig/method_types.pdf} 
+	\caption{
+		\textbf{Different forms of trajectory inference.}
+		\textbf{A:} All TI methods can be categorised in one of seven subtypes in terms of its produced output \cite{saelens_comparisonsinglecelltrajectory_2019}.
+		\textbf{B:} Each of these can be translated into a common format, allowing easier comparison of multiple trajectories.
+		\textbf{C:} The minimum information required to describe a trajectory in this way is the milestone network -- representing transitions between cellular states -- and the cell progressions -- representing the positions of cells along the transitions.
+		\textbf{D:} Optionally, regions of delayed commitment can be defined.
+A region of delayed commitment contain multiple transitions starting from the same milestone.
+This allows a TI method to assign probabilities on how likely a cell is part of one of these transitions.
+	}
 	\label{fig:method_types}
@@
 
-Saelens et al.
-\cite{saelens_comparisonsinglecelltrajectory_2019} developed an R package, \texttt{dynwrap}, for converting many of the variations of trajectory inference to the common format.
-Using this standardised format allows developing reusable software for interpreting, comparing, and visualising trajectories from any TI method.
+\texttt{dynwrap} provides implementations for converting the output from different TI subtypes into the common format.
+Using this standardised format allows developing reusable software for visualising (\texttt{dynplot}) and comparing (\texttt{dyneval}) trajectories from different TI methods.
+All of these software packages are part of the \texttt{dyno} collection, soon to be available on CRAN \cite{dyno}.
+
+\mycomment{examples? elaborate? citation?}
 
 \subsection{Benchmarking datasets}
+Another hurdle in benchmarking trajectory inference methods is collecting datasets to benchmark against.
+Before 2018, there were only a handful of datasets containing complex trajectories (Figure~\ref{fig:datasets}).
+
+To solve this problem, benchmarking studies typically use synthetic data to complement real data. 
+Several simulators have already been used to generate datasets to evaluate TI datasets, namely dyntoy \cite{saelens_comparisonsinglecelltrajectory_2019}}, dyngen \cite{dyngen}, splatter \cite{zappia_splattersimulationsinglecell_2017}, and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}.
+Benefits of synthetic data are that they offer more control over the data characteristics, and that they can be generated in large quantities.
+This allows to evaluate performance of a method in function of a changing parameter (e.g.
+dataset size or noise levels), which provides information on how well the method will work on real datasets.
+
+A common counterargument of synthetic data is that they generate unrealistic datasets and thus provide no additional value in evaluation a method.
+In contrast, we argue that a good set of synthetic datasets should allow benchmarkers to verify that a method should \textit{at least} work well on the synthetic datasets, but good performance on synthetic datasets does not guarantee good performance on real datasets.
+
+By now, already hundreds of suitable real datasets exist (Figure~\ref{fig:datasets}).
+Downloading and pre-processing them requires a significant time investment.
+However, by processing the datasets once and storing them in a single repository allows them to be reused for multiple purposes.
+Readers are invited to reuse and extend the datasets used from our comparison of TI methods hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.75\linewidth]{fig/datasets.pdf} 
+	\caption{\textbf{An overview collection of TI benchmarking datasets in function of their publication date and the topology of the trajectory.} These datasets are readily available on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.}
+	\label{fig:datasets}
+\end{figure}
 
 \subsection{Multiple metrics}
-
-%waarom benchmarks maar een laag aantal datasets hebben: 
-%* echte datasets zijn duur om te genereren
-%* het is niet evident om ze te verzamelen
-%* otoh, we zouden naar de meta-informatie van de dynbenchmark/real datasets kunnen kijken en een cumulatieve plot genereren van hoeveel datasets er beschikbaar zijn over tijd, om aan te tonen dat sinds 2016-2017 er feitelijk wel al genoeg datasets zijn om aan self-benchmarking te doen
 

### 2019-09-25 ch9_discussion/scripts/eval_checks.R

@@ -93,130 +93,7 @@ tools <-
     manuscript_date2 = ifelse(is.na(manuscript_publication_date), manuscript_preprint_date, manuscript_publication_date) %>% as.Date(origin = "1970-01-01")
   )
 
-
-pubs <-
-  tibble(
-    x = as.Date(c("2018-03-05", "2019-04-01"), origin = "1970-01-01"),
-    y = 10,
-    text = c("bioRxiv", "NBT")
-  )
-g <- ggplot() +
-  geom_point(aes(manuscript_date, num_methods, size = num_datasets, colour = peer_reviewed), tools %>% arrange(desc(num_datasets)), shape = 21) +
-  geom_vline(aes(xintercept = x), pubs, linetype = "dashed", colour = "darkgray") +
-  geom_text(aes(x, 10, label = text), pubs, colour = "darkgray", hjust = 1, nudge_x = -15) +
-  theme_classic() +
-  labs(x = "Date", y = "Number of methods", size = "Number of datasets", colour = "Peer reviewed") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
-  scale_y_continuous(limits = c(0, 12), breaks = c(0, 3, 6, 9, 12)) +
-  scale_colour_brewer(palette = "Set1") +
-  scale_size_continuous(breaks = c(0, 3, 6, 9, 12), limits = c(0, 12))
-
-g <- ggplot() +
-  geom_point(aes(manuscript_date, num_datasets, size = num_methods, colour = peer_reviewed), tools %>% arrange(desc(num_methods)), shape = 21) +
-  geom_vline(aes(xintercept = x), pubs, linetype = "dashed", colour = "darkgray") +
-  geom_text(aes(x, 8, label = text), pubs, colour = "darkgray", hjust = 1, nudge_x = -15) +
-  theme_classic() +
-  labs(x = "Date", y = "Number of datasets", size = "Number of methods", colour = "Peer reviewed") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
-  scale_y_continuous(limits = c(0, 12), breaks = c(0, 3, 6, 9, 12)) +
-  scale_colour_brewer(palette = "Set1") +
-  scale_size_continuous(breaks = c(0, 3, 6, 9, 12), limits = c(0, 12))
-
-ggsave("fig/benchmarks.pdf", g, width = 6, height = 3)
[…]

### 2019-09-25 ch9_discussion/content.tex
update

@@
 We elaborate on these reasons and viable solutions to easy benchmarking trajectory inference.
-
-%in deze commentary geven we nog een kort overzicht van:
-%* standaard definitie van een trajectory inference probleem
-%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
-%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
-%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
 
@@
 This makes visualising and comparing multiple trajectories difficult, as different output types cannot be compared directly.
-The most commonly used and general is one where cells are positioned along a set of edges connecting milestones (Figure~\ref{fig:method_types}B).
+The most commonly used and general is one where cells are positioned along a set of edges connecting milestones ("Regular TI", Figure~\ref{fig:method_types}A).
+By adding an extension to regular TI to allow for cells to be part of three or more cellular states, thereby a cell to delay its commitment toward a particular end state (Figure~\ref{fig:method_types}B).
+
+By adding this extension, all TI subtypes can easily be converted into the common format.
+Implementations of these conversions can be found in \texttt{dynwrap}\cite{dyno}.
+Using this standardised format allows developing reusable software for visualising and comparing trajectories from different TI methods.
+
 In practice, this format consists of two data structures: the milestone network specify transition between cell states, and the cell progressions specify how far along each cell has progressed along a transition (Figure~\ref{fig:method_types}C).
-Optionally, a cell can be in a region of "delayed commitment", that is, it is not known which state will end up in but a probability for ending up in either state can be assigned (Figure~\ref{fig:method_types}D).
-
-\texttt{dynwrap} provides implementations for converting the output from different TI subtypes into the common format.
-Using this standardised format allows developing reusable software for visualising (\texttt{dynplot}) and comparing (\texttt{dyneval}) trajectories from different TI methods.
-All of these software packages are part of the \texttt{dyno} collection, soon to be available on CRAN \cite{dyno}.
+In addition, regions of delayed commitment need to be specified, if any (Figure~\ref{fig:method_types}D).
 
@@
 
-To solve this problem, benchmarking studies typically use synthetic data to complement real data. 
-Several simulators have already been used to generate datasets to evaluate TI datasets, namely dyntoy \cite{saelens_comparisonsinglecelltrajectory_2019}}, dyngen \cite{dyngen}, splatter \cite{zappia_splattersimulationsinglecell_2017}, and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}.
+Several authors benchmark their method against synthetic data in order to evaluate their method \cite{haghverdi_diffusionmapshighdimensional_2015, street_slingshotcelllineage_2018,rizvi_singlecelltopologicalrnaseq_2017,welch_matchermanifoldalignment_2017,guo_hoplandsinglecellpseudotime_2017,boukouvalas_bgpbranchedgaussian_2017,gao_calistaclusteringlineage_2018,parra_reconstructingcomplexlineage_2018,chen_singlecelltrajectoriesreconstruction_2019,strauss_gpseudoclustdeconvolutionshared_2019}.
+These simulators are mostly custom made by the authors themselves, thereby increasing the level of bias in the benchmark.
+Instead, readers are encouraged to use of the readily available simulators, namely dyntoy \cite{saelens_comparisonsinglecelltrajectory_2019}, dyngen \cite{dyngen}, splatter \cite{zappia_splattersimulationsinglecell_2017}, and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}.
+
 Benefits of synthetic data are that they offer more control over the data characteristics, and that they can be generated in large quantities.
@@
 
+Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets\cite{cannoodt_scorpiusimprovestrajectory_2016,liu_reconstructingcellcycle_2017,sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,teschendorff_singlecellentropyaccurate_2017,macnair_psupertimesupervisedpseudotime_2019}.
 By now, already hundreds of suitable real datasets exist (Figure~\ref{fig:datasets}).
-Downloading and pre-processing them requires a significant time investment.
-However, by processing the datasets once and storing them in a single repository allows them to be reused for multiple purposes.
-Readers are invited to reuse and extend the datasets used from our comparison of TI methods hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.
+Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes.
+
+Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods.
+The datasets are hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018} and the scripts to process them on GitHub\footnote{\href{https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}{github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}}.
+Note that the ground truths of the datasets are represented using the common data structures format in the previous section.
 
@@
 
-%waarom benchmarks een laag aantal methoden hebben:
-%* iedere methode heeft zijn eigen set van functies
-%* methodes hebben vaak andere prior inputs (starting cell, time series, etc)
-%* methodes zijn vaak in verschillende talen geschreven (R, python, matlab)
-%* de output van methodes zijn vaak niet rechtstreeks vergelijkbaar, omdat er geen standaard definitie van wat trajectory inference is (zoals bv.
-wel het geval is bij clustering of differential expression)
-%
-%bovendien is trajectory inference een probleem met complexe output, waardoor geen kant-en-klare metrieken bestonden om de predicties mee te evalueren.
-ik zou nog eens door alle manuscripten die een evaluatie hebben gedaan, moeten gaan, maar ik denk dat alle of toch de meeste evaluaties tussen 2016 en 2018 enkel op basis van pseudotime zijn ─ ook voor bv.
-dpt, en dat er maar zelden evaluaties zijn die echt gaan kijken naar de branching
-%
-%
-%kaart aan dat we in 2018 een benchmark preprint hebben gepost van W methoden op X datasets, en in 2019 een artikel hebben gepubliceerd van Y methoden op Z datasets.
-Ondanks dat we in deze manuscripten de grootste problemen (# datasets, # methodes, # metrieken, leg iets beter uit) aanpakken, zien we geen vermindering in het aantal nieuwe publicaties die geen self-benchmark bevatten.
 %in deze commentary geven we nog een kort overzicht van:

### 2019-09-25 ch9_discussion/scripts/eval_checks.R

@@ -1,85 +1,89 @@
 library(tidyverse)
 
+
+none <- character(0)
+
+
 extra_evals <-
   tribble(
-    ~tool_id, ~num_methods, ~real_datasets, ~simulated_datasets,
-    "monocle1", 0, 0, 0,
-    "wanderlust", 0, 0, 0,
-    "scuba", 0, 0, 0,
-    "sincell", 0, 0, 0,
-    "nbor", 0, 0, 0,
-    "cycler", 0, 0, 0,
-    "oscope", 0, 0, 0,
-    "waterfall", 0, 0, 0,
-    "gpseudotime", 0, 0, 0,
-    "embeddr", 0, 0, 0,
-    "eclair", 0, 0, 0,
-    "dpt", 2, 1, 0,
-    "pseudogp", 0, 0, 0,
-    "slicer", 0, 0, 0,
-    "scell", 0, 0, 0,
-    "wishbone", 0, 0, 0,
-    "tscan", 5, 3, 0, #sim?
-    "scoup", 3, 2, 0, #sim?
-    "delorean", 0, 0, 0,
-    "raceid_stemid", 0, 0, 0,
-    "ouija", 0, 0, 0,
-    "mpath", 0, 0, 0,
-    "celltree", 3, 1, 0, #sim?
-    "wavecrest", 0, 0, 0,
-    "stemnet", 0, 0, 0,
-    "scimitar", 3, 2, 0, #sim?
-    "scorpius", 4, 10, 0,
-    "scent", 3, 5, 0,
-    "slice", 0, 0, 0,
-    "topslam", 0, 0, 0,
-    "monocle", 4, 1, 0,
-    "gpfates", 0, 0, 0,
-    "mfa", 0, 0, 0,
-    "tasic", 0, 0, 0,
-    "somsc", 0, 0, 0,
-    "slingshot", 5, 0, 3,
-    "sctda", 4, 0, 2,
-    "uncurl", 0, 0, 0,
-    "recat", 6, 10, 0,
-    "forks", 8, 6, 0,
-    "matcher", 1, 1, 1,
-    "phenopath", 0, 0, 0,
-    "hopland", 7, 6, 5,
-    "soptsc", 3, 4, 0,
-    "pba", 0, 0, 0,
-    "bgp", 3, 0, 1,
-    "brgps", 7, 2, 0,
-    "wot", 0, 0, 0,
-    "treetop", 0, 0, 0,
[…]

### 2019-09-25 ch9_discussion/content.tex
update

@@
 Instead, they rely on anecdotal evidence to demonstrate their added value.
-A brief review of 76 TI articles reveals that only about 35\% contain a self-assessment (Figure \ref{fig:benchmarks_over_time}).
-Peer-reviewed TI articles fared even worse, self-assessing in only 32\% of cases (n=52), whereas articles first published as a pre-print self-assess in 42\% of cases (n=40).
-Only four TI articles feature a decently-sized comparison of at least 5 methods using 5 datasets or more.
-% Note: i did not count how many metrics each of these manuscripts used.
+A brief review of 75 TI articles reveals that only about 37\% contain a self-assessment (Figure \ref{fig:benchmarks_over_time}).
+Peer-reviewed TI articles fared even worse, self-assessing in only 34\% of cases (n=55), whereas articles first published as a pre-print self-assess in 43\% of cases (n=39).
+Only three TI articles feature a comparison of at least 5 methods using 5 datasets or more \cite{sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,parra_reconstructingcomplexlineage_2018}.
 
@@
 
-Several authors benchmark their method against synthetic data in order to evaluate their method \cite{haghverdi_diffusionmapshighdimensional_2015, street_slingshotcelllineage_2018,rizvi_singlecelltopologicalrnaseq_2017,welch_matchermanifoldalignment_2017,guo_hoplandsinglecellpseudotime_2017,boukouvalas_bgpbranchedgaussian_2017,gao_calistaclusteringlineage_2018,parra_reconstructingcomplexlineage_2018,chen_singlecelltrajectoriesreconstruction_2019,strauss_gpseudoclustdeconvolutionshared_2019}.
-These simulators are mostly custom made by the authors themselves, thereby increasing the level of bias in the benchmark.
-Instead, readers are encouraged to use of the readily available simulators, namely dyntoy \cite{saelens_comparisonsinglecelltrajectory_2019}, dyngen \cite{dyngen}, splatter \cite{zappia_splattersimulationsinglecell_2017}, and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}.
+When real data is scarce, synthetic data is often used to evaluate computational methods, either standalone (n=5) or to complement real data (n=7).
+Most synthetic data is generated by the authors themselves (n=8), whereas some reuse datasets generated by others (n=3) or use one of the readily available simulators (n=2).
+To avoid introducing self-assessment bias in a benchmark, it is recommended to use readily available simulators if they fit the requirements.
+Examples are dyntoy \cite{saelens_comparisonsinglecelltrajectory_2019}, dyngen \cite{dyngen}, splatter \cite{zappia_splattersimulationsinglecell_2017}, and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}.
 
@@
 
-Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets\cite{cannoodt_scorpiusimprovestrajectory_2016,liu_reconstructingcellcycle_2017,sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,teschendorff_singlecellentropyaccurate_2017,macnair_psupertimesupervisedpseudotime_2019}.
+Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets (n=7). 
 By now, already hundreds of suitable real datasets exist (Figure~\ref{fig:datasets}).

### 2019-09-25 ch9_discussion/scripts/eval_checks.R

@@ -6,7 +6,7 @@ none <- character(0)
 
 extra_evals <-
   tribble(
-    ~tool_id, ~methods, ~real_datasets, ~simulated_datasets, ~metrics,
+    ~tool_id, ~methods, ~real_datasets, ~synthetic_datasets, ~metrics,
     "monocle1", none, none, none, none,
     "wanderlust", none, none, none, none,
     "scuba", none, none, none, none,
@@ -18,7 +18,7 @@ extra_evals <-
     "gpseudotime", none, none, none, none,
     "embeddr", none, none, none, none,
     "eclair", none, none, none, none,
-    "dpt", c("dpt", "wishbone", "monocle1"), c("moignard", "klein", "paul"), c("synthetic_grn"), c("pseudotime_correlation", "robustness"),
+    "dpt", c("dpt", "wishbone", "monocle1"), c("moignard", "klein", "paul"), c("synthetic_dpt"), c("pseudotime_correlation", "robustness"),
     "pseudogp", none, none, none, none,
     "slicer", none, none, none, none,
     "scell", none, none, none, none,
@@ -32,43 +32,43 @@ extra_evals <-
     "celltree", c("monocle1", "tscan", "celltree"), "trapnell", none, "pseudotime_unknown",
     "wavecrest", none, none, none, none,
     "stemnet", none, none, none, none,
-    "scimitar", c("scimitar", "monocle1", "wanderlust"), none, c("synthetic_custom", "synthetic_custom"), "pseudotime_correlation",
-    "scorpius", c("scorpius", "wanderlust", "monocle1", "waterfall"), c("schlitzer", "buettner", "shalek", "trapnell", "kowalczyk"), none, c("pseudotime_cos", "robustness_cva"),
+    "scimitar", c("scimitar", "monocle1", "wanderlust"), none, c("synthetic_scimitar", "synthetic_scimitar"), "pseudotime_correlation",
+    "scorpius", c("scorpius", "wanderlust", "monocle1", "waterfall"), c("schlitzer", "buettner", rep("shalek", 3), "trapnell", rep("kowalczyk", 4)), none, c("pseudotime_cos", "robustness_cva"),
     "scent", c("scent", "slice", "stemid"), c("chu", "trapnell", "treutlein"), none, c("pseudotime_wilcox", "pseudotime_auc"),
     "slice", none, none, none, none,
[…]

### 2019-09-26 ch9_discussion/content.tex
update

@@
-%dus was ik aan het nadenken over de stand van zaken na dynbenchmark ─ oké, ik heb zin om een dynbenchmark2 te doen moet, alles meer en beter
-%maar hoe hebben we een impact gehad? als je kijkt naar course material over trajectory inference zou ik zeggen, veel ─ we zijn een gemakkelijke bron van guidelines
-%maar als ik kijk naar de manuscripten van nieuwe TI methoden zou ik zeggen dat we nog niet zijn waar we wouden geraken
+\mycomment{Note: This text was written very quickly and is only meant as a temporary coathanger.}
 
-\section{Self-assessment in trajectory inference}
-Many trajectory inference (TI) tools lack quantitative assessment of the accuracy of the method.
+This chapter reflects on the impact of this work on the field, referring back to the research objectives that were defined in Chapter \ref{chap:introduction}.
+
+dyngen is a succes, as we created a simulator of single cell data that is already used to evaluate trajectory\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019} and single cell network inference\cite{pratapa_benchmarkingalgorithmsgene_2019}.
+
+dynbenchmark was meant to guide users to better practices for applying trajectory inference methods, and guide developers to adopt better practices for developing accurate and robust TI methods.
+The first part we succeeded in, as the guidelines are commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.
+However, we do not observe an increase in developers of TI methods performing quantitative benchmarks.
+We will discuss this in more detail in section~\ref{sec:selfassessment}.
+
+dynbenchmark was a benchmark of TI methods developed over the course of 4 years involving writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, developing novel metrics for comparing ground truth and predicted trajectories, writing a simulator for synthetic single cell data.
+Along the way, we have learned a lot about how to benchmark computational methods.
+We share our vision and guidelines on benchmarking computational methods in Chapter~\ref{chap:guidelines}.
+
+Something about SCORPIUS.
+
+Something about bred.
+
+Something about dyno.
+
+Something about gng?
+
+\section{Self-assessment in trajectory inference} \label{sec:selfassessment}
+Many articles introducing novel trajectory inference (TI) tools lack quantitative assessment of the accuracy of the method.
 Instead, they rely on anecdotal evidence to demonstrate their added value.
-A brief review of 75 TI articles reveals that only about 37\% contain a self-assessment (Figure \ref{fig:benchmarks_over_time}).
-Peer-reviewed TI articles fared even worse, self-assessing in only 34\% of cases (n=55), whereas articles first published as a pre-print self-assess in 43\% of cases (n=39).
+A brief review of 75 articles reveals that only about 37\% contain a self-assessment (Figure \ref{fig:benchmarks_over_time}A,B).
+Peer-reviewed articles fared even worse, self-assessing in only 34\% of cases (n=55), whereas articles first published as a pre-print self-assess in 43\% of cases (n=39).
+
+The number of datasets used and methods compared against is also below expectations (Figure \ref{fig:benchmarks_over_time}C,D).
 Only three TI articles feature a comparison of at least 5 methods using 5 datasets or more \cite{sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,parra_reconstructingcomplexlineage_2018}.
 
-Self-assessments are universally biased in favour of the authors, intentionally or otherwise, but this does not mean they should be skipped out on.
-Norel et al.
-provide a valuable discussion on overcoming overly optimistic self-assessments \cite{norel_selfassessmenttrapcan_2011}.
-% TODO: improve working "this does not mean they should be skipped out on"
+While self-assessments are universally biased in favour of the authors\cite{norel_selfassessmenttrapcan_2011} (intentially or not), it is dangerous and unusual to publish a computational tool without quantitatively demonstrating its performance compared to state-of-the-art methods.
+Indeed, a recent comparison of 45 TI methods demonstrated that most methods perform worse than a few baseline methods constructed by combining simple off-the-shelf algorithms such as PCA, $k$-means and MST\cite{saelens_comparisonsinglecelltrajectory_2019}.
 
-In this perspective, we hypothesise that low self-assessment rates are primarily caused by a lack of a standardised problem definition, benchmarking datasets, and metrics. 
-We elaborate on these reasons and viable solutions to easy benchmarking trajectory inference.
+In this perspective, we hypothesise that low self-assessment rates are primarily caused by a lack of a standardised problem definition, readily available benchmarking datasets, and suitable metrics. 
+We elaborate on these causal reasons, and provide viable solutions for performing TI benchmarks more easily.
 
@@
 	\includegraphics[width=.75\linewidth]{fig/self_assessment.pdf} 
-	\caption{\textbf{The number of manuscripts featuring trajectory inference tools over time.} Areas coloured in darker shades of blue represent manuscripts containing self-assessment of the tool, highlighting \textbf{A}: the number of datasets used and \textbf{B}: the number of methods used (including the method of the authors).}
+	\caption{
+		\textbf{Less than half of all TI articles perform quantitative self-assessment.} 
+		\textbf{A:} Since 2016, the number of TI articles has been increasing rapidly. Note that TI methods with both a pre-print and a peer-reviewed article only count once in the overall tally.
+		\textbf{B:} Less than 50\% of articles feature a self-assessment. Peer-reviewed articles self-assess only in 34\% of cases.
+		\textbf{C:} The number of datasets used in each benchmark is low.
+		\textbf{D:} The number of methods (inclusing itself) evaluated is low.
+	}
 	\label{fig:benchmarks_over_time}
@@
 For example, a method might infer linear or cyclic trajectories, or predict the probability of a cell ending up in one of several end states.
+
+As a result, it becomes harder to discover similar methods to compare against, as certain articles might only show up with search terms such as "pseudotemporal ordering", "lineage trees" or "fate bias".
+For the discoverability of a new TI method, it is therefore essential to use the term "trajectory inference", or at least list it as one of the keywords.
 
@@
 \end{figure}
-
-As a result, it becomes harder to discover similar methods to compare against, as certain articles might only show up with search terms such as "pseudotemporal ordering", "lineage trees" or "fate bias".
-For the discoverability of a new TI method, it is therefore essential to use the term "trajectory inference", or at least list it as one of the keywords.
 
@@
 
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.75\linewidth]{fig/datasets.pdf} 
+	\caption{\textbf{An overview collection of real TI benchmarking datasets in function of their publication date and the topology of the trajectory.} These datasets are readily available on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.}
+	\label{fig:datasets}
+\end{figure}
+
 Benefits of synthetic data are that they offer more control over the data characteristics, and that they can be generated in large quantities.
@@
 Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets (n=7). 
-By now, already hundreds of suitable real datasets exist (Figure~\ref{fig:datasets}).
+By now, already hundreds of suitable real datasets are available from GEO and ArrayExpress (Figure~\ref{fig:datasets}).
 Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes.
@@
 
+\subsection{Multiple metrics}
+The largest problem, by far, seems to be a lack of standardised metrics to evaluate TI methods.
+None of the benchmarks use a metric directly aimed at comparing the topology or pairwise orderings of two trajectories.
+
+Instead, most benchmarks (n=26) employ an ordering metric (e.g.
+pearson correlation) the pseudotime\footnote{The pseudotime value of is calculated by computing its distance from a pre-defined or inferred start cell.} of the predicted trajectory to ground truth information such as the cell type or time of sampling.
+Several benchmarks (n=5) use a clustering metric (e.g.
+ARI) to compare a cell's transition assignment or milestone assignment to the ground truth cell type.
+In four cases, multiple executions were compared to evaluate the robustness of the predictions.
+In one particular instance, the pseudotime correlation metric was adapted to be suitable for comparing rooted trees\cite{street_slingshotcelllineage_2018}.
+In another, an internal measure is used to quantify the smoothness of gene expression along the pseudotime \cite{darocha_reconstructioncomplexsinglecell_2018}.
+
+While each of these metrics provide some insight the correctness of cell orderings or cell clusterings, they do not evaluate the correctness of the prediction's topology.
+Using only a pseudotime-based metric to evaluate non-linear trajectories is particularly risky, as it
+
 \begin{figure}[htb!]
 	\centering
-	\includegraphics[width=.75\linewidth]{fig/datasets.pdf} 
-	\caption{\textbf{An overview collection of TI benchmarking datasets in function of their publication date and the topology of the trajectory.} These datasets are readily available on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.}
-	\label{fig:datasets}
+	\includegraphics[width=.5\linewidth]{fig/metrics.pdf} 
+	\caption{}
+	\label{fig:metrics}
 \end{figure}
-
-\subsection{Multiple metrics}
 
@@
 %Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality. 
-
-\section{dyngen discussion}
-% felt that upon the development of a new technologies, good quality control practices from
-% other technologies were not being carried over because the data required to evaluate computational
-% tools does not exist yet and is too expensive to develop.
-
-\section{dynbenchmark discussion}
-\begin{itemize}
-	\item Already outdated when the manuscript was published online
-	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
-	\item Include RNA velocity as inputs
-\end{itemize}
-
-\section{SCORPIUS discussion}
-\begin{itemize}
-	\item Extension to inferring non-linear trajectories, i.e. with principal graphs
-\end{itemize}
-
-\section{bred discussion}
-
-%\section{incgraph discussion}
-
-%\section{dyno discussion}
+%
+%\section{dyngen discussion}
+%% felt that upon the development of a new technologies, good quality control practices from
+%% other technologies were not being carried over because the data required to evaluate computational
+%% tools does not exist yet and is too expensive to develop.
+%
+%\section{dynbenchmark discussion}
+%\begin{itemize}
+%	\item Already outdated when the manuscript was published online
+%	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
+%	\item Include RNA velocity as inputs
+%\end{itemize}
+%
+%\section{SCORPIUS discussion}
+%\begin{itemize}
+%	\item Extension to inferring non-linear trajectories, i.e. with principal graphs
+%\end{itemize}
+%
+%\section{bred discussion}
+%
+%%\section{incgraph discussion}
+%
+%%\section{dyno discussion}

### 2019-09-26 ch9_discussion/scripts/eval_checks.R

@@ -1,202 +1,147 @@
 library(tidyverse)
+library(googlesheets)
+library(rlang)
 
 
-none <- character(0)
+# COLLECT DATA ------------------------------------------------------------
 
+# If it's your first time running this script, run this:
+# gs_auth()
+sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE")
 
-extra_evals <-
-  tribble(
-    ~tool_id, ~methods, ~real_datasets, ~synthetic_datasets, ~metrics,
-    "monocle1", none, none, none, none,
-    "wanderlust", none, none, none, none,
-    "scuba", none, none, none, none,
-    "sincell", none, none, none, none,
-    "nbor", none, none, none, none,
-    "cycler", none, none, none, none,
-    "oscope", none, none, none, none,
-    "waterfall", none, none, none, none,
-    "gpseudotime", none, none, none, none,
-    "embeddr", none, none, none, none,
-    "eclair", none, none, none, none,
-    "dpt", c("dpt", "wishbone", "monocle1"), c("moignard", "klein", "paul"), c("synthetic_dpt"), c("pseudotime_correlation", "robustness"),
-    "pseudogp", none, none, none, none,
-    "slicer", none, none, none, none,
-    "scell", none, none, none, none,
-    "wishbone", none, none, none, none,
-    "tscan", c("monocle1", "tscan", "waterfall", "scuba", "wanderlust"), c("trapnell", "amit", "shin"), none, "pseudotime_pos",
-    "scoup", c("scoup", "monocle1", "tscan"), c("kouno", "moignard", "shalek"), none, "pseudotime_pis",
-    "delorean", none, none, none, none,
-    "raceid_stemid", none, none, none, none,
-    "ouija", none, none, none, none,
-    "mpath", none, none, none, none,
-    "celltree", c("monocle1", "tscan", "celltree"), "trapnell", none, "pseudotime_unknown",
-    "wavecrest", none, none, none, none,
-    "stemnet", none, none, none, none,
-    "scimitar", c("scimitar", "monocle1", "wanderlust"), none, c("synthetic_scimitar", "synthetic_scimitar"), "pseudotime_correlation",
[…]

### 2019-09-26 ch7_selfassessment/content.tex
restructure repository

--- 2019-09-26 ch9_discussion/content.tex
@@
-\mycomment{Note: This text was written very quickly and is only meant as a temporary coathanger.}
-
-This chapter reflects on the impact of this work on the field, referring back to the research objectives that were defined in Chapter \ref{chap:introduction}.
-
-dyngen is a succes, as we created a simulator of single cell data that is already used to evaluate trajectory\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019} and single cell network inference\cite{pratapa_benchmarkingalgorithmsgene_2019}.
-
-dynbenchmark was meant to guide users to better practices for applying trajectory inference methods, and guide developers to adopt better practices for developing accurate and robust TI methods.
-The first part we succeeded in, as the guidelines are commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.
-However, we do not observe an increase in developers of TI methods performing quantitative benchmarks.
-We will discuss this in more detail in section~\ref{sec:selfassessment}.
-
-dynbenchmark was a benchmark of TI methods developed over the course of 4 years involving writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, developing novel metrics for comparing ground truth and predicted trajectories, writing a simulator for synthetic single cell data.
-Along the way, we have learned a lot about how to benchmark computational methods.
-We share our vision and guidelines on benchmarking computational methods in Chapter~\ref{chap:guidelines}.
-
-Something about SCORPIUS.
-
-Something about bred.
-
-Something about dyno.
-
-Something about gng?
-
-\section{Self-assessment in trajectory inference} \label{sec:selfassessment}
 Many articles introducing novel trajectory inference (TI) tools lack quantitative assessment of the accuracy of the method.
@@
 
-\subsection{Problem definition}
+\section{Problem definition}
 One main reason why benchmarking TI methods is difficult is due to there being slight variations 
@@
 
-\subsection{Benchmarking datasets}
+\section{Benchmarking datasets}
 Another hurdle in benchmarking trajectory inference methods is collecting datasets to benchmark against.
@@
 
-\subsection{Multiple metrics}
+\section{Multiple metrics}
 The largest problem, by far, seems to be a lack of standardised metrics to evaluate TI methods.
@@
 
-%in deze commentary geven we nog een kort overzicht van:
-%* standaard definitie van een trajectory inference probleem
-%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
-%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
-%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
-%in conclusie, met al deze informatie melden we dat in 2019 er dus geen enkele reden is om ons te gedragen als gelijk welk ander computationeel / data mining / machine learning veld waarbij er voor quantitatief bewijs wordt gevraagd dat je methode goed werkt als je een nieuwe methode wilt publiceren.
-we vragen vriendelijk aan peer-reviewers om kritisch te zijn over deze kwestie.
-
-%in deze commentary geven we nog een kort overzicht van:
-%* standaard definitie van een trajectory inference probleem
-%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
-%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
-%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
-
-%in conclusie, met al deze informatie melden we dat in 2019 er dus geen enkele reden is om ons te gedragen als gelijk welk ander computationeel / data mining / machine learning veld waarbij er voor quantitatief bewijs wordt gevraagd dat je methode goed werkt als je een nieuwe methode wilt publiceren.
-we vragen vriendelijk aan peer-reviewers om kritisch te zijn over deze kwestie.
-
-%refer to: \cite{norel_selfassessmenttrapcan_2011} 
-%In order to alleviate the overestimation of accuracy from the many bias sources described above, we proposed a few guidelines:
-%* use third‐party validation to test a model with previously unseen data
-%* use more than one metric to evaluate the methods
-%* report well‐performing methods even if they are not the best performers on a particular data set
-%* increase the awareness of editors and reviewers that superior performance in self‐assessment is a biased demonstration of the method's value; instead, impartial assessment should be the preferred evaluation
-%* Establish a scientific culture that values timely, well‐conducted follow‐up studies that confirm or refute previous results
-
-%Before 2016, only a handful of trajectory inference methods had been released. None of these articles contained a self-benchmark simply because
-%
-%In all fairness towards the early TI methods, in 2016 or even 2017 there was a low abundance of good datasets to perform benchmarks with.
-The amount of data required to perform a comprehensive evaluation of their tool did not yet exist and was too expensive to generate.
-In April 2016, we started developing a single-cell simulator that would allow us to perform a comprehensive benchmark of TI methods.
-%
-%
-%It was also non-trivial to compare against many methods as each method has a unique interface where some prior knowledge might be needed. 
-%
-
-% * Single cell omics is expanding at rates faster than any single mind can try to conceive
-%Pioneers developing revolutionary tools to process new types of data are typically faced with the problem on how to quantitatively evaluate the accuracy of their approach.
-After all, the amount of data required to perform a comprehensive evaluation of their tool does not yet exist and is too expensive to generate at the time.
-
-%In the computational jungle of single-cell omics, experts act as beacons of hope by sharing guidelines based on comprehensive benchmarking studies.
-Their disseminations (in the form of manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}) are crucial in leading new users, and ultimately the whole field, to better practices for performing single-cell omics analyses.
-
-%Before embarking on this perilous adventure, I attempt to write a "Hippocratic oath for bioinformaticians", to guide me through this dissertation and not stray from the path of righteousness.
-%
-%\subsection{Hippocratic oath for bioinformaticians}
-%We proceed with caution in developing new tools.
-Its results should not only be convincing and easy to interpret, but also accurate, robust, and reproducible.
-We open source.
-Our software should work reliably and fail gracefully when it does not.
-We acknowledge that writing automated tests is dull but necessary for maintaining long-term software projects.
-% TODO: improve
-
-%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
-%
-%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality. 
-%
-%\section{dyngen discussion}
-%% felt that upon the development of a new technologies, good quality control practices from
-%% other technologies were not being carried over because the data required to evaluate computational
-%% tools does not exist yet and is too expensive to develop.
-%
-%\section{dynbenchmark discussion}
-%\begin{itemize}
-%	\item Already outdated when the manuscript was published online
-%	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
-%	\item Include RNA velocity as inputs
-%\end{itemize}
-%
-%\section{SCORPIUS discussion}
-%\begin{itemize}
-%	\item Extension to inferring non-linear trajectories, i.e. with principal graphs
-%\end{itemize}
-%
-%\section{bred discussion}
-%
-%%\section{incgraph discussion}
-%
-%%\section{dyno discussion}
+\section{Further guidelines}

### 2019-09-26 old/ch7_incgraph/content.tex
restructure repository

--- 2019-08-21 ch8_incgraph.tex
@@
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{incgraph: Optimising regulatory networks} 
-\chaptermark{incgraph: Optimising regulatory networks.} % short title
-\label{chap:incgraph}
-
-\begin{quote}
-	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
-	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
-	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
-\\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
-	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
-	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
-	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
-\\
-	IncGraph is freely available as an open-source R package on CRAN (incgraph). The development version is also available on GitHub (\href{http://www.github.com/rcannood/incgraph}{rcannood/incgraph}).
-\end{quote}
-
-\vfill
-
-Adapted from:\\
-\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y.
-IncGraph: Incremental graphlet counting for topology optimisation.
-\textit{PLOS ONE} 13, 4 (2018), e0195997.
-\doi{10.1371/journal.pone.0195997}.
-\newpage
-
 \section*{Introduction}
@@
 %% FIGURE 1
-\picfig{fig/incgraph/figure1.pdf}{\LARGEfigure}{fig_graphlets}{{\bf Graphlet counting in a network characterises its local topologies.}
+\picfig{fig/figure1.pdf}{\LARGEfigure}{fig_graphlets}{{\bf Graphlet counting in a network characterises its local topologies.}
 	(A) In total, there are 30 different graphlets containing 2 to 5 nodes, ranging from $G_0$ to $G_{29}$.
@@
 %% FIGURE 2
-\picfig{fig/incgraph/figure2.pdf}{\LARGEfigure}{fig_efficient}{{\bf Several strategies are employed in order to reduce time and memory consumption.}
+\picfig{fig/figure2.pdf}{\LARGEfigure}{fig_efficient}{{\bf Several strategies are employed in order to reduce time and memory consumption.}
 (A) Only the depth 3 neighbourhood of each modified edge needs to be explored in order to have visited all modified five-node graphlets.
@@
 %% FIGURE 3
-\picfig{fig/incgraph/figure3.pdf}{\LARGEfigure}{fig_method_speedup}{{\bf Static network model generators were modified to generate dynamic networks.}
+\picfig{fig/figure3.pdf}{\LARGEfigure}{fig_method_speedup}{{\bf Static network model generators were modified to generate dynamic networks.}
 	Three network models were used: Barabási–Albert, Erdős–Rényi, and Geometric.
@@
 %% FIGURE 4
-\picfig{fig/incgraph/figure4.pdf}{\hugefigure}{fig_method_reranking}{{\bf Predicted gene regulatory networks of model organisms are optimised to reduce the false positive rate.}
+\picfig{fig/figure4.pdf}{\hugefigure}{fig_method_reranking}{{\bf Predicted gene regulatory networks of model organisms are optimised to reduce the false positive rate.}
 	A)~The number of redundant edges in each graphlet are counted.
@@
 %% FIGURE 5
-\picfig{fig/incgraph/figure5.pdf}{\LARGEfigure}{fig_speedup}{{\bf IncGraph is significantly faster than non-incremental approaches.}
+\picfig{fig/figure5.pdf}{\LARGEfigure}{fig_speedup}{{\bf IncGraph is significantly faster than non-incremental approaches.}
 	For small networks, the execution time of IncGraph $T_{IG}$ is already 50 times less than that of non-incremental approaches $T_{NI}$.
@@
 %% FIGURE 6
-\picfig{fig/incgraph/figure6.pdf}{\Largefigure}{fig_reranking}{{\bf A simple graphlet-based scoring method improves predicted regulatory networks.}
+\picfig{fig/figure6.pdf}{\Largefigure}{fig_reranking}{{\bf A simple graphlet-based scoring method improves predicted regulatory networks.}
 	(A) The F1 score was calculated by calculating the harmonic mean of the AUROC and AUPR scores of the first 1000 interactions.

### 2019-09-26 ch7_incgraph/fig/pdfconvert.sh

@@ -1,6 +1,0 @@
-#!/bin/sh
-
-for file in *.pdf; do
-noext=`echo $file | sed 's#.pdf##'`
-./pdf2eps.sh $noext
-done

### 2019-09-26 ch7_selfassessment/scripts/eval_checks.R

@@ -1,0 +1,208 @@
+library(tidyverse)
+library(googlesheets)
+library(rlang)
+
+
+# COLLECT DATA ------------------------------------------------------------
+
+# If it's your first time running this script, run this:
+# gs_auth()
+sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE")
+
+case_study_data <- gs_read(sheet, ws = "case_study") %>% select(tool_id = id, methods, datasets, metrics)
+
+tools <-
+  read_rds(dynbenchmark::result_file("tools.rds", "03-methods")) %>%
+  mutate(
+    tool_id = ifelse(tool_id == "monocle", "monocle2", tool_id),
+    tool_name = ifelse(tool_id == "monocle2", "Monocle 2", ifelse(is.na(tool_name), method_name, tool_name)),
+    manuscript_doi = ifelse(tool_id == "elpigraph", "arXiv:1804.0758", manuscript_doi)
+  ) %>%
+  inner_join(case_study_data, by = c("tool_id")) %>%
+  mutate_at(c("methods", "datasets", "metrics"), ~ strsplit(. %|% "", ",")) %>%
+  mutate(
+    synthetic_datasets = map(datasets, keep, ~ grepl("synthetic_", .)),
+    real_datasets = map(datasets, discard, ~ grepl("synthetic_", .)),
+    num_methods = map_int(methods, length),
+    num_datasets = map_int(datasets, length),
+    num_metrics = map_int(metrics, length),
+    self_assessment = num_methods > 0,
+    peer_reviewed = !is.na(manuscript_publication_date),
+    has_preprint = !is.na(manuscript_preprint_date),
+    manuscript_date = ifelse(has_preprint, manuscript_preprint_date, manuscript_publication_date) %>% as.Date(origin = "1970-01-01"),
+    manuscript_date2 = ifelse(peer_reviewed, manuscript_publication_date, manuscript_preprint_date) %>% as.Date(origin = "1970-01-01"),
+    metric_group = map_chr(metrics, function(w) {
+      x <- gsub("_.*", "", w)
+      case_when(
+        length(x) == 0 ~ "None",
+        "branch" %in% x && "pseudotime" %in% x ~ "Cluster & pseudotime",
+        "cluster" %in% x ~ "Cluster",
[…]

### 2019-09-26 ch9_discussion/scripts/eval_checks.R

@@ -1,208 +1,0 @@
-library(tidyverse)
-library(googlesheets)
-library(rlang)
-
-
-# COLLECT DATA ------------------------------------------------------------
-
-# If it's your first time running this script, run this:
-# gs_auth()
-sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE")
-
-case_study_data <- gs_read(sheet, ws = "case_study") %>% select(tool_id = id, methods, datasets, metrics)
-
-tools <-
-  read_rds(dynbenchmark::result_file("tools.rds", "03-methods")) %>%
-  mutate(
-    tool_id = ifelse(tool_id == "monocle", "monocle2", tool_id),
-    tool_name = ifelse(tool_id == "monocle2", "Monocle 2", ifelse(is.na(tool_name), method_name, tool_name)),
-    manuscript_doi = ifelse(tool_id == "elpigraph", "arXiv:1804.0758", manuscript_doi)
-  ) %>%
-  inner_join(case_study_data, by = c("tool_id")) %>%
-  mutate_at(c("methods", "datasets", "metrics"), ~ strsplit(. %|% "", ",")) %>%
-  mutate(
-    synthetic_datasets = map(datasets, keep, ~ grepl("synthetic_", .)),
-    real_datasets = map(datasets, discard, ~ grepl("synthetic_", .)),
-    num_methods = map_int(methods, length),
-    num_datasets = map_int(datasets, length),
-    num_metrics = map_int(metrics, length),
-    self_assessment = num_methods > 0,
-    peer_reviewed = !is.na(manuscript_publication_date),
-    has_preprint = !is.na(manuscript_preprint_date),
-    manuscript_date = ifelse(has_preprint, manuscript_preprint_date, manuscript_publication_date) %>% as.Date(origin = "1970-01-01"),
-    manuscript_date2 = ifelse(peer_reviewed, manuscript_publication_date, manuscript_preprint_date) %>% as.Date(origin = "1970-01-01"),
-    metric_group = map_chr(metrics, function(w) {
-      x <- gsub("_.*", "", w)
-      case_when(
-        length(x) == 0 ~ "None",
-        "branch" %in% x && "pseudotime" %in% x ~ "Cluster & pseudotime",
-        "cluster" %in% x ~ "Cluster",
[…]

### 2019-09-26 ch7_selfassessment/content.tex
update

@@
 
-Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods.
+Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods\cite{saelens_comparisonsinglecelltrajectory_2019}.
 The datasets are hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018} and the scripts to process them on GitHub\footnote{\href{https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}{github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}}.
@@
 
-\section{Multiple metrics}
+\section{Metrics}
 The largest problem, by far, seems to be a lack of standardised metrics to evaluate TI methods.
@@
 While each of these metrics provide some insight the correctness of cell orderings or cell clusterings, they do not evaluate the correctness of the prediction's topology.
-Using only a pseudotime-based metric to evaluate non-linear trajectories is particularly risky, as it
+Using only a pseudotime-based metric to evaluate non-linear trajectories is particularly risky, as a \ldots.
 
@@
 
-\section{Further guidelines}
+\mycomment{explain why.}
+
+Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models using only one metric.
+In Supplementary Note 1 in our comparison of TI methods\cite{saelens_comparisonsinglecelltrajectory_2019}, we describe and illustrate 10 very different metrics for comparing trajectories.
+We perform a comprehensive characterisation and validation of these metrics on several test cases.
+
+\mycomment{Explain geodesic distance, him, f1 branches, wcor features}
+%Its ordering, using the
+%371 • Its branch assignment, using the
+%372 • Its topology, using the HIM
+%373 • The accuracy of differentially expressed features, using the
+%374
+%F1 branches
+%wcor features
+
+\section{Guidelines for benchmarking TI tools}

### 2019-09-26 ch7_selfassessment/content.tex
update

--- 2019-09-26 ch7_selfassessment/content.tex
@@
 Only three TI articles feature a comparison of at least 5 methods using 5 datasets or more \cite{sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,parra_reconstructingcomplexlineage_2018}.
+In comparison, our recent benchmark of TI methods evaluated the performance 45 TI methods on 110 real and 229 synthetic datasets\cite{saelens_comparisonsinglecelltrajectory_2019}.
 
 While self-assessments are universally biased in favour of the authors\cite{norel_selfassessmenttrapcan_2011} (intentially or not), it is dangerous and unusual to publish a computational tool without quantitatively demonstrating its performance compared to state-of-the-art methods.
-Indeed, a recent comparison of 45 TI methods demonstrated that most methods perform worse than a few baseline methods constructed by combining simple off-the-shelf algorithms such as PCA, $k$-means and MST\cite{saelens_comparisonsinglecelltrajectory_2019}.
+Indeed, our comparison demonstrated that most methods perform worse than a few baseline methods constructed by combining simple off-the-shelf algorithms such as PCA, $k$-means and MST.
 
@@
 
+\mycomment{Todo: mention containerised methods in dynmethods or simply using them as standalone containers.}
+
 In practice, this format consists of two data structures: the milestone network specify transition between cell states, and the cell progressions specify how far along each cell has progressed along a transition (Figure~\ref{fig:method_types}C).
@@
 Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes.
+Mixture control experiments\cite{tian_benchmarkingsinglecell_2019} are particularly useful in this context.
 
-Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods.
+Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods\cite{saelens_comparisonsinglecelltrajectory_2019}.
 The datasets are hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018} and the scripts to process them on GitHub\footnote{\href{https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}{github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}}.
@@
 
-\section{Multiple metrics}
-The largest problem, by far, seems to be a lack of standardised metrics to evaluate TI methods.
-None of the benchmarks use a metric directly aimed at comparing the topology or pairwise orderings of two trajectories.
+\section{Metrics}
+To evaluate a TI method, a quantitative metric is needed to compare the predicted trajectory to the ground truth trajectory.
+No off-the-shelf metrics exist for comparing complex multilayered data structures such as trajectories to each other.
+To get around this problem, most benchmarks repurpose metrics from other domains (Figure~\ref{fig:metrics}A).
 
-Instead, most benchmarks (n=26) employ an ordering metric (e.g.
-pearson correlation) the pseudotime\footnote{The pseudotime value of is calculated by computing its distance from a pre-defined or inferred start cell.} of the predicted trajectory to ground truth information such as the cell type or time of sampling.
-Several benchmarks (n=5) use a clustering metric (e.g.
-ARI) to compare a cell's transition assignment or milestone assignment to the ground truth cell type.
-In four cases, multiple executions were compared to evaluate the robustness of the predictions.
-In one particular instance, the pseudotime correlation metric was adapted to be suitable for comparing rooted trees\cite{street_slingshotcelllineage_2018}.
-In another, an internal measure is used to quantify the smoothness of gene expression along the pseudotime \cite{darocha_reconstructioncomplexsinglecell_2018}.
-
-While each of these metrics provide some insight the correctness of cell orderings or cell clusterings, they do not evaluate the correctness of the prediction's topology.
-Using only a pseudotime-based metric to evaluate non-linear trajectories is particularly risky, as it
+For example, most benchmarks (n=26) compare the linear pseudotime ordering of a trajectory with ground-truth information such as time time of sampling, quite often by calculating the pearson correlation.
+This is a good approach for comparing linear trajectories, but is not suitable as a metric for comparing non-linear trajectories (e.g.
+by calculating the distance from one end point of the trajectory), as this metric does not capture any differences in topology between the two trajectories (Figure~\ref{fig:metrics}B).
+Several other benchmarks (n=5) use a metric typically used to compare a clustering method, by comparing a cell's assignment to the transitions of the trajectory to ground-truth information such as the cell's cell type.
+While this will provide some information on whether cells are grouped correctly in comparison to the ground-truth, it also does not capture topological differences between trajectories (Figure~\ref{fig:metrics}C).
 
@@
 	\centering
-	\includegraphics[width=.5\linewidth]{fig/metrics.pdf} 
-	\caption{}
+	\includegraphics[width=.6\linewidth]{fig/perturbations.pdf} 
+	\caption{
+		\textbf{A:} Without any off-the-shelf metrics to use for evaluating TI methods, authors use metrics from other domains.
+		\textbf{B:} Comparing these two trajectories using a pseudotime metric (e.g. pearson correlation) would return a perfect score, even though these two trajectories are clearly different.
+		\textbf{C:} Comparing these two trajectories using a clustering metric (e.g. ARI) would also return a perfect score, even though these two trajectories are clearly different.
+	}
 	\label{fig:metrics}
@@
 
-\section{Further guidelines}
+Extra credit should be given to four cases in which the robustness of methods was evaluated by comparing multiple executions of the same method\cite{haghverdi_diffusionpseudotimerobustly_2016,cannoodt_scorpiusimprovestrajectory_2016,_doesyourcode_2018,sharma_forksfindingorderings_2017,jin_scepathenergylandscapebased_2018}.
+Computing the robustness does not replace the necessity of a relevant metric that captures whether a predicted trajectory resembles the ground truth -- that is, a TI method can robustly make incorrect predictions and obtain high robustness scores.
+In another benchmark, an internal metric is used to quantify the smoothness of gene expression along the pseudotime \cite{darocha_reconstructioncomplexsinglecell_2018} -- the idea being that good TI methods would order cells such that gene expression is smooth along the pseudotime.
+
+In our comparison of TI methods\cite{saelens_comparisonsinglecelltrajectory_2019}, we use a metric called the \emph{geodesic correlation}.
+Here, two trajectories are compared by calculating the geodesic distances between pairs of cells and comparing those distance values using a Pearson correlation.
+Note not all pairwise distances are evaluated in this way, or the metric would not scale well to larger datasets.
+We also use the Hamming-Ipsen-Mikhailov (HIM) distance\cite{jurman_himglocalmetric_2015} to compare the topology of two trajectories.
+
+In Supplementary Note 1 of our comparison of TI methods, we describe and illustrate 10 different metrics, including the geodesic correlation metric, the HIM distance, and several clustering and internal metrics.
+We constructed 26 test cases (similar to Figure~\ref{fig:metrics}B,C) to assess whether a metric is able of capturing the desired information.
+We found that the geodesic distance passes nearly all of the test cases, using the geometric mean of different metrics performs much better.
+
+All of these metrics are described in detail in Supplementary Note 1 and implementations are available in \texttt{dyneval}\cite{dyno}.
+
+\section{Guidelines for performing self-assessments}
+Articles introducing novel trajectory inference methods have unfortunately been plagued by low self-assessment rates.
+Most articles passing peer-review do not provide quantitative evidence that their method performs well.
+In the previous sections, we show that this is most likely caused by differing problem statements making methods hard to compare and a lack of benchmarking datasets and relevant metrics.
+We provide the reader with viable solutions to each of the problems, in order to be able to benchmark their method with ease.
+
+We wish to leave the reader with a set of guidelines on how to perform a self-assessment on the performance of their own computational method.
+%refer to: \cite{norel_selfassessmenttrapcan_2011} 
+%In order to alleviate the overestimation of accuracy from the many bias sources described above, we proposed a few guidelines:
+%* use third‐party validation to test a model with previously unseen data
+%* use more than one metric to evaluate the methods
+%* report well‐performing methods even if they are not the best performers on a particular data set
+%* increase the awareness of editors and reviewers that superior performance in self‐assessment is a biased demonstration of the method's value; instead, impartial assessment should be the preferred evaluation
+%* Establish a scientific culture that values timely, well‐conducted follow‐up studies that confirm or refute previous results
+
+\mycomment{Get a few guidelines from \cite{norel_selfassessmenttrapcan_2011} and \cite{weber_essentialguidelinescomputational_2019}.}
+
+\begin{table}[h]
+	\caption{\textbf{Trajectory inference method metadata.} } \label{tab:methods}
+	\centering
+	\tiny
+	\rowcolors{2}{gray!10}{white}
+\begin{tabularx}{\linewidth}{|lp{1cm}p{1cm}lXXX|}
+	\hline
+	id & preprint date & publication date & doi & methods & metrics & datasets \\ \hline \hline
+monocle1 &  & 2014-03-23 & \doi{10.1038/nbt.2859} &  &  &  \\
+wanderlust &  & 2014-04-24 & \doi{10.1016/j.cell.2014.04.005} &  &  &  \\
+scuba &  & 2014-12-30 & \doi{10.1073/pnas.1408993111} &  &  &  \\
+sincell & 2015-01-27 & 2015-06-22 & \doi{10.1093/bioinformatics/btv368} &  &  &  \\
+nbor &  & 2015-06-08 & \doi{10.1038/ni.3200} &  &  &  \\
+oscope &  & 2015-08-24 & \doi{10.1038/nmeth.3549} &  &  &  \\
+cycler &  & 2015-08-24 & \doi{10.1038/nmeth.3545} &  &  &  \\
+waterfall &  & 2015-09-03 & \doi{10.1016/j.stem.2015.07.013} &  &  &  \\
+gpseudotime & 2015-09-15 &  & \doi{10.1101/026872} &  &  &  \\
+embeddr & 2015-09-18 &  & \doi{10.1101/027219} &  &  &  \\
+eclair & 2016-01-12 & 2016-05-20 & \doi{10.1093/nar/gkw452} &  &  &  \\
+dpt & 2016-02-08 & 2016-08-29 & \doi{10.1038/nmeth.3971} & dpt, wishbone, monocle1 & pseudotime\_correlation, robustness\_unknown & moignard, klein, paul, synthetic\_dpt \\
+pseudogp & 2016-04-05 & 2016-11-21 & \doi{10.1371/journal.pcbi.1005212} &  &  &  \\
+slicer & 2016-04-09 & 2016-05-23 & \doi{10.1186/s13059-016-0975-3} &  &  &  \\
+scell &  & 2016-04-19 & \doi{10.1093/bioinformatics/btw201} &  &  &  \\
+wishbone &  & 2016-05-02 & \doi{10.1016/j.cell.2014.04.005} &  &  &  \\
+tscan &  & 2016-05-13 & \doi{10.1093/nar/gkw430} & monocle1, tscan, waterfall, scuba, wanderlust & pseudotime\_pos & trapnell, amit, shin \\
+scoup &  & 2016-06-08 & \doi{10.1186/s12859-016-1109-3} & scoup, monocle1, tscan & pseudotime\_pis & kouno, moignard, shalek \\
+delorean &  & 2016-06-17 & \doi{10.1093/bioinformatics/btw372} &  &  &  \\
+raceid\_stemid &  & 2016-06-21 & \doi{10.1016/j.stem.2016.05.010} &  &  &  \\
+ouija & 2016-06-23 &  & \doi{10.1101/060442} &  &  &  \\
+mpath &  & 2016-06-30 & \doi{10.1038/ncomms11988} &  &  &  \\
+celltree &  & 2016-08-13 & \doi{10.1186/s12859-016-1175-6} & monocle1, tscan, celltree & pseudotime\_unknown & trapnell \\
+wavecrest &  & 2016-08-17 & \doi{10.1186/s13059-016-1033-x} &  &  &  \\
+stemnet &  & 2016-08-25 & \doi{10.1038/ncb3493} &  &  &  \\
+scimitar & 2016-10-04 & 2017-01-04 & \doi{10.1142/9789813207813\_0053} & scimitar, monocle1, wanderlust & pseudotime\_correlation & synthetic\_scimitar, synthetic\_scimitar \\
+scorpius & 2016-10-07 &  & \doi{10.1101/079509} & scorpius, wanderlust, monocle1, waterfall & pseudotime\_cos, robustness\_cva & schlitzer, buettner, shalek, shalek, shalek, trapnell, kowalczyk, kowalczyk, kowalczyk, kowalczyk \\
+scent &  & 2016-10-30 & \doi{10.1038/ncomms15599} & scent, slice, stemid & pseudotime\_wilcox, pseudotime\_auc & chu, trapnell, treutlein \\
+slice &  & 2016-12-19 & \doi{10.1093/nar/gkw1278} &  &  &  \\
+topslam & 2017-02-13 &  & \doi{10.1101/057778} & monocle1, wishbone, topslam & pseudotime\_correlation & synthetic\_topslam \\
+monocle2 & 2017-02-21 & 2017-07-20 & \doi{10.1038/nmeth.4402} & monocle1, monocle2, dpt, wishbone & pseudotime\_correlation, branch\_ari & paul \\
+gpfates &  & 2017-03-03 & \doi{10.1126/sciimmunol.aal2192} &  &  &  \\
+mfa &  & 2017-03-15 & \doi{10.12688/wellcomeopenres.11087.1} &  &  &  \\
+tasic &  & 2017-04-04 & \doi{10.1093/bioinformatics/btx173} &  &  &  \\
+somsc & 2017-04-05 &  & \doi{10.1101/124693} &  &  &  \\
+slingshot & 2017-04-19 & 2018-06-19 & \doi{10.1186/s12864-018-4772-0} & slingshot, monocle1, monocle2, dpt, tscan & pseudotime\_correlation & synthetic\_splatter, synthetic\_splatter, synthetic\_splatter, synthetic\_splatter, synthetic\_splatter \\
+sctda &  & 2017-05-01 & \doi{10.1038/nbt.3854} & sctda, wishbone, slicer, dpt & pseudotime\_correlation & synthetic\_sctda \\
+uncurl & 2017-05-31 &  & \doi{10.1101/142398} &  &  &  \\
+recat &  & 2017-06-19 & \doi{10.1038/s41467-017-00039-z} & recat, scuba, monocle1, tscan, wishbone, dpt & pseudotime\_correlation, pseudotime\_custom & buettner \\
+forks & 2017-06-20 &  & \doi{10.1101/132811} & forks, monocle2, scuba, tscan, waterfall, dpt, gpfates, slicer & pseudotime\_correlation, robustness\_stdev & windram, deng, guo, klein, amit, petropoulos \\
+matcher &  & 2017-06-24 & \doi{10.1186/s13059-017-1269-0} & matcher & pseudotime\_correlation & angelmueller, synthetic\_matcher \\
+phenopath & 2017-07-06 & 2018-06-23 & \doi{10.1101/159913} &  &  &  \\
+hopland &  & 2017-07-12 & \doi{10.1093/bioinformatics/btx232} & hopland, wanderlust, monocle1, topslam, scuba, wishbone, dpt & pseudotime\_correlation & guo, deng, yan, amit, islam, synthetic\_topslam \\
+soptsc & 2017-07-26 & 2019-06-20 & \doi{10.1093/nar/gkz204} & soptsc, monocle2, dpt & pseudotime\_correlation & guo, klein, shalek \\
+pba &  & 2017-07-30 & \doi{10.1073/pnas.1714723115} &  &  &  \\
+brgps & 2017-08-15 &  & \doi{10.1101/167684} & brgps, grandprix, monocle2, scuba, slicer, tscan, wishbone & pseudotime\_correlation & guo, guo \\
+wot & 2017-09-27 & 2019-02-07 & \doi{10.1016/j.cell.2019.01.006} &  &  &  \\
+treetop & 2017-10-10 &  & \doi{10.1101/200923} &  &  &  \\
+paga & 2017-10-27 & 2019-03-19 & \doi{10.1186/s13059-019-1663-x} &  &  &  \\
+fateid & 2017-11-11 & 2018-04-09 & \doi{10.1038/nmeth.4662} &  &  &  \\
+pseudodynamics & 2017-11-14 &  & \doi{10.1101/219188} &  &  &  \\
+pcreode &  & 2017-11-15 & \doi{10.1016/j.cels.2017.10.012} &  &  &  \\
+icpsc &  & 2017-11-30 & \doi{10.1038/s41467-017-01860-2} & icpsc, wishbone, monocle2, dpt & pseudotime\_correlation & sun, trapnell, yao \\
+grandprix & 2017-12-03 & 2018-07-02 & \doi{10.1101/227843} & delorean, grandprix & pseudotime\_correlation & windram \\
+cshmm & 2017-12-03 & 2019-04-30 & \doi{10.1093/bioinformatics/btz296} &  &  &  \\
+calista & 2018-01-31 &  & \doi{10.1101/257550} & monocle2, calista, dpt & pseudotime\_correlation & moignard, bargaje, treutlein, chu, synthetic\_calista \\
+scepath &  & 2018-02-05 & \doi{10.1093/bioinformatics/bty058} & scepath, monocle1, monocle2, tscan, dpt & pseudotime\_correlation, robustness\_correlation & yan, treutlein, treutlein, trapnell \\
+merlot & 2018-02-08 &  & \doi{10.1101/261768} & merlot, dpt, slicer, monocle2, slingshot, tscan & branch\_mi, pseudotime\_correlation & paul, guo, velten, synthetic\_prosstt, synthetic\_prosstt, synthetic\_splatter \\
+gpseudorank & 2018-02-08 & 2018-07-25 & \doi{10.1093/bioinformatics/bty664} &  &  &  \\
+cellrouter &  & 2018-03-01 & \doi{10.1038/s41467-018-03214-y} & monocle2, dpt, wishbone, waterfall & internal\_autocorrelation & paul, olsson \\
+densitypath & 2018-03-05 & 2018-12-07 & \doi{10.1093/bioinformatics/bty1009} & monocle2, wishbone, dpt, densitypath & branch\_ari, pseudotime\_correlation & petropoulos, synthetic\_phate, synthetic\_topslam \\
+topographer & 2018-03-23 &  & \doi{10.1101/251207} &  &  &  \\
+stream & 2018-04-18 & 2019-04-23 & \doi{10.1038/s41467-019-09670-4} & stream, sctda, wishbone, slicer, monocle2, dpt, tscan, scuba, mpath, gpfates & pseudotime\_correlation & synthetic\_sctda \\
+elpigraph & 2018-04-20 &  & arXiv:\href{https://arxiv.org/abs/1804.0758}{1804.0758} &  &  &  \\
+urd &  & 2018-04-26 & \doi{10.1126/science.aar3131} &  &  &  \\
+celltrails &  & 2018-06-05 & \doi{10.1016/j.celrep.2018.05.002} &  &  &  \\
+ddd & 2018-07-12 &  & \doi{10.1101/367789} &  &  &  \\
+palantir & 2018-08-05 & 2019-03-21 & \doi{10.1038/s41587-019-0068-4} &  &  &  \\
+confess & 2018-09-04 &  & \doi{10.1101/407932} &  &  &  \\
+graphddp &  & 2018-09-11 & \doi{10.1038/s41467-018-05988-7} &  &  &  \\
+monocle3 &  & 2019-02-28 & \doi{10.1038/s41586-019-0969-x} &  &  &  \\
+gpseudoclust & 2019-03-05 &  & \doi{10.1101/567115} & gpseudoclust, monocle2, delorian, slicer & cluster\_ari, cluster\_fmi, cluster\_nmi & sasagawa, shalek, synthetic\_gpseudoclust, synthetic\_gpseudoclust \\
+psupertime & 2019-04-29 &  & \doi{10.1101/622001} & monocle2, slingshot, psupertime & pseudotime\_correlation & enge, qiu, petropoulos, li, treutlein \\
+cyclum & 2019-05-02 &  & \doi{10.1101/625566} & cyclum, recat & cluster\_accuracy & buettner, mcdavid, mcdavid, mcdavid \\
+sinova &  & 2019-05-17 & \doi{10.1016/j.celrep.2016.04.043} &  &  &  \\
+\hline
+\end{tabularx}
+\end{table}

### 2019-09-26 ch7_selfassessment/scripts/eval_checks.R

@@ -37,11 +37,9 @@ tools <-
         length(x) == 0 ~ "None",
         "branch" %in% x && "pseudotime" %in% x ~ "Cluster & pseudotime",
         "cluster" %in% x ~ "Cluster",
-        any(c("pseudotime", "internal", "pseudotime") %in% x) ~ "Pseudotime",
-        "cyclepseudotime" %in% x ~ "Pseudotime for cycles",
-        "treepseudotime" %in% x ~ "Pseudotime for trees"
+        any(c("pseudotime", "internal", "pseudotime") %in% x) ~ "Pseudotime"
       )
-    }) %>% factor(levels = c("None", "Pseudotime", "Cluster", "Cluster & pseudotime", "Pseudotime for cycles", "Pseudotime for trees"))
+    }) %>% factor(levels = c("None", "Pseudotime", "Cluster", "Cluster & pseudotime"))
   )
 
 if (save_tsv) {
@@ -63,12 +61,24 @@ if (save_tsv) {
   #   mutate_at(c("methods", "metrics", "datasets"), ~ map_chr(., paste, collapse = ",")) %>%
   #   mutate_at(c("preprint_date", "pub_date"), function(x) ifelse(is.na(x), "", as.character(x)))
   # gs_edit_cells(sheet, ws = "case_study", input = overwrite, anchor = "A1", col_names = TRUE, trim = TRUE)
+  # overwrite %>%
+  #   transmute(
+  #     id, preprint_date, pub_date,
+  #     doi = ifelse(grepl("arXiv",  doi), gsub("arXiv:(.*)", "arXiv:\\\\href{https://arxiv.org/abs/\\1}{\\1}", doi), paste0("\\doi{", doi, "}")),
+  #     methods = methods %>% gsub(",", ", ", .),
+  #     metrics = metrics %>% gsub(",", ", ", .),
+  #     datasets = datasets %>% gsub(",", ", ", .)
+  #   ) %>%
+  #   apply(1, paste, collapse = " & ") %>%
+  #   gsub("_", "\\\\_", .) %>%
+  #   paste0(" \\\\") %>%
+  #   write_lines("~/test.tex")
 }
 
 
 
 # INTRODUCTION ------------------------------------------------------------
-
+tools$metric_group %>% table
 tools
 tools$self_assessment %>% mean
 tools %>% filter(!is.na(manuscript_publication_date)) %>% pull(self_assessment) %>% mean
[…]

### 2019-09-26 ch7_selfassessment/scripts/eval_checks.R
add submodule?

@@ -1,219 +1,0 @@
-library(tidyverse)
-library(googlesheets)
-library(rlang)
-
-
-# COLLECT DATA ------------------------------------------------------------
-
-# If it's your first time running this script, run this:
-# gs_auth()
-sheet <- gs_key("1Mug0yz8BebzWt8cmEW306ie645SBh_tDHwjVw4OFhlE")
-
-case_study_data <- gs_read(sheet, ws = "case_study") %>% select(tool_id = id, methods, datasets, metrics)
-
-tools <-
-  read_rds(dynbenchmark::result_file("tools.rds", "03-methods")) %>%
-  mutate(
-    tool_id = ifelse(tool_id == "monocle", "monocle2", tool_id),
-    tool_name = ifelse(tool_id == "monocle2", "Monocle 2", ifelse(is.na(tool_name), method_name, tool_name)),
-    manuscript_doi = ifelse(tool_id == "elpigraph", "arXiv:1804.0758", manuscript_doi)
-  ) %>%
-  inner_join(case_study_data, by = c("tool_id")) %>%
-  mutate_at(c("methods", "datasets", "metrics"), ~ strsplit(. %|% "", ",")) %>%
-  mutate(
-    synthetic_datasets = map(datasets, keep, ~ grepl("synthetic_", .)),
-    real_datasets = map(datasets, discard, ~ grepl("synthetic_", .)),
-    num_methods = map_int(methods, length),
-    num_datasets = map_int(datasets, length),
-    num_metrics = map_int(metrics, length),
-    self_assessment = num_methods > 0,
-    peer_reviewed = !is.na(manuscript_publication_date),
-    has_preprint = !is.na(manuscript_preprint_date),
-    manuscript_date = ifelse(has_preprint, manuscript_preprint_date, manuscript_publication_date) %>% as.Date(origin = "1970-01-01"),
-    manuscript_date2 = ifelse(peer_reviewed, manuscript_publication_date, manuscript_preprint_date) %>% as.Date(origin = "1970-01-01"),
-    metric_group = map_chr(metrics, function(w) {
-      x <- gsub("_.*", "", w)
-      case_when(
-        length(x) == 0 ~ "None",
-        "branch" %in% x && "pseudotime" %in% x ~ "Cluster & pseudotime",
-        "cluster" %in% x ~ "Cluster",
[…]

### 2019-09-27 boek.tex
update
restructure repository

--- 2019-09-19 boek.tex
@@
+\PassOptionsToPackage{table}{xcolor}
 \documentclass[10pt]{book}
@@
 \usepackage{setspace}
-
-% scripting packages
-\usepackage{blindtext}
 
@@
 \usepackage{colortbl}
+\usepackage{tcolorbox}
 
@@
 
-% boxes around sections
-\usepackage{tcolorbox}
+% allow to adjust spacing in enumerates and itemizes
+\usepackage{enumitem}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%         CH4: DYNO         %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
-\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
-\label{chap:dyno}
-\begin{quote}
-	\textbf{Abstract:} \blindtext
-\end{quote}
-\vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.
-\include{ch4_dyno/content}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%         CH4: DYNO         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\newpage{\thispagestyle{empty}\cleardoublepage}
+%\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
+%\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
+%\label{chap:dyno}
+%\begin{quote}
+%	\textbf{Abstract:} 
+%\end{quote}
+%\vfill
+%Adapted from:\\
+%\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.
+%\include{ch4_dyno/content}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH5: SCORPIUS       %%%
+%%%       CH4: SCORPIUS       %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{quote}
-	\textbf{Abstract:} \blindtext
+	\textbf{Abstract:}
 \end{quote}
@@
 \newpage
-\graphicspath{{./ch5_scorpius/}}
-\include{ch5_scorpius/content}
+\graphicspath{{./ch4_scorpius/}}
+\include{ch4_scorpius/content}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%         CH6: BRED         %%%
+%%%         CH5: BRED         %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newpage
-\include{ch6_bred/content}
+\include{ch5_bred/content}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH7: INCGRAPH       %%%
+%%%       CH5B: INCGRAPH       %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{chap:incgraph}
-\graphicspath{{./ch7_incgraph/}}
+\graphicspath{{./ch5b_incgraph/}}
 \begin{quote}
@@
 \newpage
-\include{ch7_incgraph/content}
+\include{ch5b_incgraph/content}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%      CH6: DISCUSSION      %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{Discussion} 
+\chaptermark{Discussion.} % short title
+\label{chap:discussion}
+\graphicspath{{./ch6_discussion/}}
+\include{ch6_discussion/content}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%   CH7: Self-assessment    %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{Self-assessment in trajectory inference} 
+\chaptermark{Self-assessment in trajectory inference.} % short title
+\label{chap:selfassessment}
+\graphicspath{{./ch7_selfassessment/}}
+%\begin{quote}
+%	\textbf{Abstract:}
+%\end{quote}
+%\vfill
+%Adapted from:\\
+%Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
+%Boulesteix, A., Saeys, Y., and Robinson, M. D.. 
+%Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.
+\newpage
+\include{ch7_selfassessment/content}
 
@@
 Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
-Boulesteix, A., Saeys, Y., and Robinson, M. D.. 
+Boulesteix, A., Saeys, Y., and Robinson, M. D. 
 Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.

### 2019-09-27 ch5b_incgraph/fig/pseudocodeS1.tex
update

\documentclass[a4paper,11pt]{article}

\usepackage[scaled]{helvet}
\renewcommand\familydefault{\sfdefault} 
\usepackage[T1]{fontenc}

\newcommand{\pluseq}{\mathrel{+}=}
\newcommand{\asteq}{\mathrel{*}=}

% Drawing with TiKz
\usepackage{tikz}
\usetikzlibrary[positioning]
\usetikzlibrary{patterns}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{shapes.arrows}
\usetikzlibrary{arrows}
\usetikzlibrary{calc,positioning,shapes,chains,mindmap,trees,decorations}
\usetikzlibrary{matrix,fit}

\usepackage{algorithm}
\usepackage{fixltx2e}
\usepackage{algpseudocode}
\MakeRobust{\Call}

\usepackage[small, bf]{caption}
\usepackage{subcaption}

\usepackage{float}

\usepackage[cm]{fullpage}

\usepackage{pdflscape}

\newcommand{\subsubsec}[1]{{\setlength{\parindent}{0cm}\textbf{#1} \\}}
\newcommand{\specialcell}[2][c]{\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\newcommand\mycaption[4]{%
  \textbf{#1:\ \ \ #3} #4}

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

\pagestyle{empty}
\begin{document}

\begin{algorithmic}
	\Function{CalculateDelta}{Network $G'$, Node $n_0$, Node $n_1$}
		\State \Comment{Assuming an edge $(n_0, n_1)$ has just been added or removed from $G'$}
		\State $\Delta^- \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the removed graphlets}
		\State $\Delta^+ \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the added graphlets}
		\State $B_0 = \{n_0, n_1\}$ \Comment{A blacklist of nodes not to visit anymore}
		\State $e = (n_0, n_1)$ \Comment{Different name for the edge}
		\If{$e \in N'$}
			\State $(m^-, m^+) = (0, 1)$ \Comment{In other words, $m^- = 1$ iff $e \in N$ and $m^+ = 1$ iff $e \in N'$}
		\Else
			\State $(m^-, m^+) = (1, 0)$
		\EndIf
		\State $x_0 = 0$
		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1), x_0, m^-, m^+}$ \Comment{Update delta matrices for current nodes}

		%\For{$n_2 \in \Call{Neighbours}{S_0}$}
		\For{$n_2 \in \bigcup_{i \in \{0, 1\}}\Call{Neighbours}{n_i}$}
			\If{$n_2 \not\in B_0$}
				\State $B_0 = \{n_2\} \bigcup B_0$ \Comment{Add $n_2$ to blacklist $B_0$}
				\State $x_1 = x_0 + \Call{W}{G', (n_0, n_2), 1} + \Call{W}{G', (n_1, n_2), 2}$ \Comment{Calculate edge weights for current nodes}
				\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2), x_1, m^-, m^+}$
				\State $B_1 = B_0$ \Comment{Make a copy for the next iteration depth}
				\For{$n_3 \in \bigcup_{i \in \{0, 1, 2\}}\Call{Neighbours}{n_i}$}
					\If{$n_3 \not\in B_1$}
						\State $B_1 = \{n_3\} \bigcup B_1$
						\State $x_2 = x_1 + \Call{W}{G', (n_0, n_3), 3} + \Call{W}{G', (n_1, n_3), 4} + \Call{W}{G', (n_2, n_3), 5}$
						\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3), x_2, m^-, m^+}$
						\State $B_2 = B_1$

						\For{$n_4 \in \bigcup_{i \in \{0, 1, 2, 3\}}\Call{Neighbours}{n_i}$}
							\If{$n_4 \not\in B_2$}
								\State $B_2 = \{n_4\} \bigcup B_2$
								\State $x_3 = x_2 + \Sigma_{i \in \{0, 1, 2, 3\}}\Call{W}{G', (n_i, n_4), i+6}$
								\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3, n_4), x_3, m^-, m^+}$
							\EndIf
						\EndFor
					\EndIf
				\EndFor
			\EndIf
		\EndFor
	\EndFunction
	\Function{CountOrbits}{$\Delta^+$, $\Delta^-$, Nodes $S$, Edgeweights $x$, Modifier $m^-$, Modifier $m^+$}
  \State $L^- = L[x + m^-]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N$}
			\State $\Delta^-[S, L^-] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^-$ in $\Delta^-$}
  \State $L^+ = L[x + m^+]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N'$}
			\State $\Delta^+[S, L^+] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^+$ in $\Delta^+$}
			\State \Return $(\Delta^-, \Delta^+)$
	\EndFunction
	\Function{W}{Network $G$, Edge $e$, Exponent $i$}
		\State \Return $e \in \Call{Edges}{G} \ ?\  2^i : 0$ \Comment{Return $2^i$ if $G$ contains edge $e$}
	\EndFunction
\end{algorithmic}

\end{document}

### 2019-09-28 ch1_introduction/content.tex
update

@@
 The cell is the smallest unit of life, of which all known living organisms are composed.
-Every cell houses a plethora of biomolecular processes that allow it to adapt to changes in its environment continuously.
-It can be very challenging to comprehend the cellular response to a signal due to the dynamic nature of these processes.
-A reductionist approach to understanding a complex biological system is to study the biochemical components which it is comprised of \cite{brigandt_reductionismbiology_2017}.
+Every cell houses a plethora of biomolecular processes that allow it to continuously adapt to changes in its environment.
+Due to the dynamic nature of these processes, it can be very challenging to comprehend the cellular response to a signal.
+A reductionist approach to understanding a complex biological system is to study the biochemical components of which it is comprised\cite{brigandt_reductionismbiology_2017}.
 
 Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells.
-Observing the biomolecular insides of cells in this manner can ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for diagnosing and treating disease.
-Every coin has its flip side, however, and in this case, it is that the amount of data generated from these experiments is not analysable by hand.
+With it comes the challenge of analysing what scientists refer to as “big data”, an amount of data that is not easily analysable and interpretable by hand.
+The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret.
+Namely, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}.
+Biologists thus turn to computer scientists
+to develop new tools to tackle these problems and help them to extract meaningful biological insights from the data.
+In this work, incremental contributions were made to the field in order to be able to address the aforementioned problems in a more comprehensive context.
 
+Observing the biomolecular insides of cells can ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for disease diagnosis, prognosis, and treatment.
 For example, the Human Cell Atlas (HCA) consortium \cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
@@
 
-\mycomment{
-HT: I feel like you're only going to focus on preprocessing, when reading this section.
-Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way?
-\\
-RC: Literally but succinctly state what the exact contributions of the thesis are, this is still very vague.
-}
-
-The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret. 
-Namely, the data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques uses, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}.
-Biologists thus turn to computer scientists\footnote{or computational biologists turn to themselves} to develop new tools to tackle these problems and help biologists extract meaningful biological insights from the data.
-
-This work makes incremental contributions to the field in order to be able to address the aforementioned problems in a more comprehensive context.
-This chapter first introduces several key concepts in both cell biology and computer science, upon which the remainder of this work relies.
-Afterwards, the research objectives and main contributions of this work are outlined.
+The next part of the chapter highlights several key concepts in both cell biology and computer science, upon which the remainder of this work relies.
 
 \subsection{The origins of life and the RNA world}
-The discovery of the double helix shape of Deoxyribonucleic Acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution.
+The discovery of the double helix shape of deoxyribonucleic acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution.
 By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism.
 With other words, the DNA of an organism contains the complete set of instructions required to build all of the biomolecular machinery present in its body.
-The magnitude of this discovery is reflected in our language and culture alike; with sayings such as "It's in your DNA.", or usage of its shape in countless illustrations or artworks (Figure \ref{fig:fsvm}).
 
-\mycomment{
-RC: Use crispr-cas to assert the impact of the discovery of DNA, not the FSVM spiral 
-}
-
-\begin{figure}[htb!]
-  \centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
-  \caption{
-    \textbf{A prominent display of the double helix shape at the VIB FSVM building.}
-  }
-  \label{fig:fsvm}
-\end{figure}
-
-Even so, a widely-accepted hypothesis states that life (or cells) did not originate from DNA, but instead was kicked off from its lesser-known cousin, Ribonucleic Acid (RNA).
+Even so, a widely-accepted hypothesis states that life (or cells) did not originate from DNA, but instead was kicked off from its lesser-known cousin, ribonucleic acid (RNA).
 According to the RNA world hypothesis \cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}).
-Only later did cells develop the ability to use DNA and proteins to self-sustain in a process commonly referred to as the Central Dogma.
-
-\mycomment{use this reference? \cite{cornell_prebioticaminoacids_2019}}
+Only later did cells develop the ability to use the more chemically stable DNA molecules to self-sustain in a process commonly referred to as the Central Dogma.
 
@@
 
-\subsection{The Central Dogma}
-\mycomment{HT: feels like an enumeration, should make it more into a story}
-
-The Central Dogma is a set of processes present which govern the general flow of genetic information in almost all existent living cells.
-In short, it states that DNA codes for RNA, which in turn codes for proteins.
-In this work, we assume the main processes involved in the Central Dogma are replication, transcription, splicing and translation (Figure \ref{fig:centraldogma}).
+The central dogma describes the general flow of genetic information in almost all existent living cells: DNA is decoded to RNA, which in turn encodes proteins.
+Main processes involved in the central dogma are transcription, splicing and translation (Figure \ref{fig:centraldogma}).
 
@@
 \end{figure}
-
-\textbf{Replication} is the process of duplicating DNA, which allows a cell to divide such that the resulting daughter cells retain complete copies of its genetic information.
-DNA consists of four different so-called nucleobases named adenine (A), cytosine (C), guanine (G), and thymine (T).
-Each strand of the double helix structure of DNA is a linear chain of nucleobases.
-The two strands are held together by hydrogen bonds since adenine can form three hydrogen bonds with thymine and cytosine can form two hydrogen bonds with guanine.
-Each nucleobase can only bond with one other nucleobase, making the strands complementary to one another.
-Since one strand can be deduced from its complementary strand, DNA can be represented by a sequence of letters, for example, "ACTCGGTTTAGCA".
 
@@
 Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
-% TODO: mention that there are about 20'000 protein-coding genes in the human body.
+
+% CVC: Work in progress
+%A gene is a stretch of DNA that contains a genetic blueprint for a RNA molecule. The gene serves as a template for making an RNA molecule in a process called transcription. 
+%The resulting RNA transcript can be protein-coding or non-coding.
+For this work we only consider the protein-coding genes, which consists of alternating sections called introns and exons.
+Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
+%
+%A protein-coding gene encodes protein-coding RNA molecules, the so-called messenger RNA (mRNA).
+%
+%This work only considers the protein-coding RNA molecules, the so-called messenger RNA (mRNA).
+%% TODO: mention that there are about 20'000 protein-coding genes in the human body.
+%
+%The DNA that serves as a template for making a protein-coding RNA molecule is called a gene, and the complete set of an organism's genes is called its genome.
+%
+%The protein-coding part of the DNA is called a gene.
 
@@
 
+\mycomment{Add summaries for IncGraph, discussion, self-assessment and essential guidelines}
+
 \section{List of contributions}
@@
   \item \textbf{Cannoodt R}, Saelens W, Saeys Y. bred: Inferring single cell regulatory networks. In preparation.
+  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
  \end{itemize}
@@
     \cranpkg{incgraph} & aut & CRAN & 3175 & Incremental graphlet counting for network optimisation \\
-    \githubpkg{rcannood}{lmds} & aut & Github & \notavailable & Landmark Multi-Dimensional Scaling \\
+    \cranpkg{lmds} & aut & CRAN & \notavailable & Landmark Multi-Dimensional Scaling \\
     \cranpkg{princurve} & aut & CRAN & 26'991 & Fits a principal curve in arbitrary dimension \\

### 2019-09-29 boek.tex
update

@@
 \usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+\usepackage{parskip} % disable paragraph indent and set paragraph skipping
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
 
@@
 \makeatother
+
+% si units and chemical stuff
+\usepackage{siunitx}
+\usepackage[version=3]{mhchem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%&%%%
+%%%      RMD packages      %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% use upquote for straight quotes in verbatim environments
+\usepackage{upquote}
+
+% fancy verbatim
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+
+% common commands
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+
+% something with figures
+\usepackage{grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+
+% something with making smaller itemizes
+\providecommand{\tightlist}{%
+	\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% booktab tables
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{wrapfig}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
 
@@
 \usepackage{hyperref}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[open,openlevel=1]{bookmark}
@@
 \renewcommand{\baselinestretch}{1.25}
-
-% si units and chemical stuff
-\usepackage{siunitx}
-\usepackage[version=3]{mhchem}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dyngen: simulating single cells} 
-\chaptermark{dyngen: simulating single cells}
+\chapter{dyngen: benchmarking with \textit{in silico} single cells} 
+\chaptermark{dyngen: benchmarking with \textit{in silico} single cells}
 \label{chap:dyngen}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%         CH4: DYNO         %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\newpage{\thispagestyle{empty}\cleardoublepage}
-%\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
-%\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
-%\label{chap:dyno}
-%\begin{quote}
-%	\textbf{Abstract:} 
-%\end{quote}
-%\vfill
-%Adapted from:\\
-%\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.
-%\include{ch4_dyno/content}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%         CH3B: DYNO         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
+\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
+\label{chap:dyno}
+\begin{quote}
+	\textbf{Abstract:} Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+Since 2014, at least 75 tools for trajectory inference have been developed.
+However, for end-users with a dataset of interest, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.
+
+	We developed dyno, a set of software tools that allow a user to easily infer, visualise and interpret single-cell trajectories.
+First, the user can select the most optimal set of methods based on the size of the dataset, the prior knowledge on the trajectories' topology and other user preferences.
+Each selected method can then be easily run within a common interface.
+Because the output of these methods are all converted into a common consistent format, the trajectories can be easily interpreted.
+For this, we provide a complete toolkit: adding a root and labelling the different cell stages, detecting genes which are differentially expressed at different stages of the trajectory, and plotting the trajectory within any dimensionality reduction or heatmap.
+
+	We are currently extending this pipeline to include additional input data, such as RNA velocity measurements, and additional tools for interpreting the trajectories, such as alignment and differential expression.
+\end{quote}
+\vfill
+Adapted from:\\
+\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.\\
+{\footnotesize $^*$ Equal contribution}
+\newpage
+\graphicspath{{./ch3b_dyno/}}
+\include{ch3b_dyno/content}
 

### 2019-09-29 ch1_introduction/content.tex
update

@@
 Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells.
-With it comes the challenge of analysing what scientists refer to as “big data”, an amount of data that is not easily analysable and interpretable by hand.
+With it comes the challenge of analysing a large amounts of data that are not easily interpretable by hand.
 The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret.
-Namely, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}.
+For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}.
 Biologists thus turn to computer scientists
@@
 
-%Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
-%This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}.
-%Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}.
-%TI has two objectives: to reconstruct the topology of the dynamic process (e.g.
+Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
+This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}.
+Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}.
+TI has two objectives: to reconstruct the topology of the dynamic process (e.g.
 is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology.
@@
 
-In this work, we aim to speed up scientific progress in single-cell omics by providing end-users with guidelines on how to use which state-of-the-art tools, and by providing developers with the necessary tools to assert a mimimum performance criterion when introducing new methods.
+In this work, we aim to speed up scientific progress in single-cell omics by providing tools both for end-users and developers alike.
+For developers of computational approaches, we provide tools and guidelines for benchmarking their method on real and synthetic data.
+For end-users we develop new tools and guidelines for analysing dynamic processes by inferring trajectories and gene regulatory networks.
+These contributions are discussed in the following chapters:
 
-We develop benchmarking strategies for assessing the performance of novel computational tools constrained by low availability of real data (Chapter~\ref{chap:dyngen}).
+\begin{itemize}
+  \item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}).
 \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
-
-We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}).
+  \item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}).
 Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking.
-
-We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
+We provide a set of guidelines for end-users wishing to infer trajectories.
+We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
+  \item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
 Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
-
-We conclude by inventing a new type of approach aimed at inferring the GRN of individual cells (Chapter~\ref{chap:bred}).
-Due a lack of real data which can be used as a gold standard dataset, we use \textit{in silico} single-cell data to quantitatively evaluate the performance.
-
-\mycomment{Add summaries for IncGraph, discussion, self-assessment and essential guidelines}
+  \item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
+  \item Every NI method has certain topological biases.
+We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+  \item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}).
+We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
+  \item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
+\end{itemize}
 
@@
   \item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
+  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
   \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
-  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
   \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.

### 2019-09-29 ch7_incgraph/fig/pseudocodeS4.tex
restructure

--- 2019-09-27 ch5b_incgraph/fig/pseudocodeS1.tex
@@
 	\begin{algorithmic}
-	\Function{CalculateDelta}{Network $G'$, Node $n_0$, Node $n_1$}
-		\State \Comment{Assuming an edge $(n_0, n_1)$ has just been added or removed from $G'$}
-		\State $\Delta^- \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the removed graphlets}
-		\State $\Delta^+ \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the added graphlets}
-		\State $B_0 = \{n_0, n_1\}$ \Comment{A blacklist of nodes not to visit anymore}
-		\State $e = (n_0, n_1)$ \Comment{Different name for the edge}
-		\If{$e \in N'$}
-			\State $(m^-, m^+) = (0, 1)$ \Comment{In other words, $m^- = 1$ iff $e \in N$ and $m^+ = 1$ iff $e \in N'$}
-		\Else
-			\State $(m^-, m^+) = (1, 0)$
-		\EndIf
-		\State $x_0 = 0$
-		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1), x_0, m^-, m^+}$ \Comment{Update delta matrices for current nodes}
-
-		%\For{$n_2 \in \Call{Neighbours}{S_0}$}
-		\For{$n_2 \in \bigcup_{i \in \{0, 1\}}\Call{Neighbours}{n_i}$}
-			\If{$n_2 \not\in B_0$}
-				\State $B_0 = \{n_2\} \bigcup B_0$ \Comment{Add $n_2$ to blacklist $B_0$}
-				\State $x_1 = x_0 + \Call{W}{G', (n_0, n_2), 1} + \Call{W}{G', (n_1, n_2), 2}$ \Comment{Calculate edge weights for current nodes}
-				\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2), x_1, m^-, m^+}$
-				\State $B_1 = B_0$ \Comment{Make a copy for the next iteration depth}
-				\For{$n_3 \in \bigcup_{i \in \{0, 1, 2\}}\Call{Neighbours}{n_i}$}
-					\If{$n_3 \not\in B_1$}
-						\State $B_1 = \{n_3\} \bigcup B_1$
-						\State $x_2 = x_1 + \Call{W}{G', (n_0, n_3), 3} + \Call{W}{G', (n_1, n_3), 4} + \Call{W}{G', (n_2, n_3), 5}$
-						\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3), x_2, m^-, m^+}$
-						\State $B_2 = B_1$
-
-						\For{$n_4 \in \bigcup_{i \in \{0, 1, 2, 3\}}\Call{Neighbours}{n_i}$}
-							\If{$n_4 \not\in B_2$}
-								\State $B_2 = \{n_4\} \bigcup B_2$
-								\State $x_3 = x_2 + \Sigma_{i \in \{0, 1, 2, 3\}}\Call{W}{G', (n_i, n_4), i+6}$
-								\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3, n_4), x_3, m^-, m^+}$
-							\EndIf
-						\EndFor
-					\EndIf
-				\EndFor
-			\EndIf
-		\EndFor
-	\EndFunction
-	\Function{CountOrbits}{$\Delta^+$, $\Delta^-$, Nodes $S$, Edgeweights $x$, Modifier $m^-$, Modifier $m^+$}
-  \State $L^- = L[x + m^-]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N$}
-			\State $\Delta^-[S, L^-] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^-$ in $\Delta^-$}
-  \State $L^+ = L[x + m^+]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N'$}
-			\State $\Delta^+[S, L^+] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^+$ in $\Delta^+$}
-			\State \Return $(\Delta^-, \Delta^+)$
-	\EndFunction
-	\Function{W}{Network $G$, Edge $e$, Exponent $i$}
-		\State \Return $e \in \Call{Edges}{G} \ ?\  2^i : 0$ \Comment{Return $2^i$ if $G$ contains edge $e$}
+		\Function{Geometric}{number of nodes $n$, number of edges $e$, \par\hspace*{2.77cm} number of operations $o$, number of dimensions $d=3$}
+			\State $P \gets$ Sample $n$ points from a multivariate continuous uniform distribution $U_{d}((0,1)^{d})$
+			\State $D \gets$ Calculate distance matrix from $P$
+			\State $N_0 \gets \Call{Head}{\Call{ArgSort}{\Call{LowerTriangle}{D}}, e}$ \Comment{Initial network}
+			\State $N_p \gets N_0$
+			\State $O \gets (\,)$ \Comment{Variable to store the generated operations in}
+			\While{$|O| < o$} \Comment{Modify network until $O$ is sufficiently large}
+				\State $i \gets $ Sample 1 index from $\{0\,..\,n-1\}$
+				\State $P[i] \gets$ Sample 1 point from a $U_{d}((0,1)^{d})$ \Comment{Give node $i$ a new location}
+				\State $D[i,] \gets D[,i] \gets $ Calculate new distances between node $i$ and all other nodes
+				\State $N_c \gets \Call{Head}{\Call{ArgSort}{\Call{LowerTriangle}{D}}, e}$ \Comment{New network}
+				\State $O_a \gets \{(\textrm{ADD}, e) \,|\, e \not\in E(N_p) \wedge e \in E(N_c) \}$ \Comment{Gather added edges}
+				\State $O_r \gets \{(\textrm{REM}, e) \,|\, e \in E(N_p) \wedge e \not\in E(N_c) \}$ \Comment{Gather removed edges}
+				\State $O \gets O + \Call{Shuffle}{O_a + O_r}$ \Comment{Append new operations to $O$}
+				\State $N_p \gets N_c$
+			\EndWhile
+			\State $O \gets \Call{Head}{O, o}$ 
+			\State \Return $(N_0, O)$ \Comment{Return the initial geometric network and $o$ operations}
 		\EndFunction

### 2019-09-29 ch7_incgraph/content.tex
minor changes

--- 2019-09-26 old/ch7_incgraph/content.tex
@@
 
-The differences in graphlet counts $\Delta_{G, G'}$ are calculated by iterating recursively over the neighbours surrounding each of the modified edges (See \nameref{code_incgraph}).
+The differences in graphlet counts $\Delta_{G, G'}$ are calculated by iterating recursively over the neighbours surrounding each of the modified edges (Fig~\ref{code_incgraph}).
 Several strategies are used in order to calculate $\Delta_{G,G'}$ as efficiently as possible (Fig~\ref{fig_efficient}).
@@
 The network models were based on three static network models: Barab\'asi-Albert~\cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~\cite{erdos_randomgraphs_1959}, and Geometric~\cite{appel_minimumvertexdegree_1997}.
-Pseudo code for these random evolving network models can be found in \nameref{code_ba}, \nameref{code_er}, and \nameref{code_geo} respectively.
+These models were adapted to generate evolving networks instead (Figs~\ref{code_ba}, \ref{code_er}, and \ref{code_geo}).
 Each model generates an initial network according to the static network model it is based on, and a list of network modifications (removing an edge from or adding an edge to the network).
@@
 
-Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (\nameref{sfig_hindex}).
+Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (Fig~\ref{sfig_hindex}).
 This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.
@@
 
-\section*{Supporting information}
+\section{Supporting information}
 
-%% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Pseudocode.}
+\newcommand{\pluseq}{\mathrel{+}=}
+\newcommand{\asteq}{\mathrel{*}=}
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.8\linewidth]{fig/figureS1.pdf} 
+	\caption{
+		\textbf{Empirical measurements show a strong relation between the execution time of IncGraph and the h-index cubed of the network it was applied to.} This is in line with the findings by Eppstein et al., where counting 3-size subgraphs has an amortised time of $O(h)$ and counting 4-size subgraphs has an amortised time of $O(h^2)$.
+	}
+	\label{sfig_hindex}
+\end{figure}
+
+\begin{figure}[htb!]
+	\begin{algorithmic}
+		\Function{CalculateDelta}{Network $G'$, Node $n_0$, Node $n_1$}
+		\State \Comment{Assuming an edge $(n_0, n_1)$ has just been added or removed from $G'$}
+		\State $\Delta^- \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the removed graphlets}
+		\State $\Delta^+ \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the added graphlets}
+		\State $B_0 = \{n_0, n_1\}$ \Comment{A blacklist of nodes not to visit anymore}
+		\State $e = (n_0, n_1)$ \Comment{Different name for the edge}
+		\If{$e \in N'$}
+		\State $(m^-, m^+) = (0, 1)$ \Comment{In other words, $m^- = 1$ iff $e \in N$ and $m^+ = 1$ iff $e \in N'$}
+		\Else
+		\State $(m^-, m^+) = (1, 0)$
+		\EndIf
+		\State $x_0 = 0$
+		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1), x_0, m^-, m^+}$ \Comment{Update delta matrices for current nodes}
+
+		%\For{$n_2 \in \Call{Neighbours}{S_0}$}
+		\For{$n_2 \in \bigcup_{i \in \{0, 1\}}\Call{Neighbours}{n_i}$}
+		\If{$n_2 \not\in B_0$}
+		\State $B_0 = \{n_2\} \bigcup B_0$ \Comment{Add $n_2$ to blacklist $B_0$}
+		\State $x_1 = x_0 + \Call{W}{G', (n_0, n_2), 1} + \Call{W}{G', (n_1, n_2), 2}$ \Comment{Calculate edge weights for current nodes}
+		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2), x_1, m^-, m^+}$
+		\State $B_1 = B_0$ \Comment{Make a copy for the next iteration depth}
+		\For{$n_3 \in \bigcup_{i \in \{0, 1, 2\}}\Call{Neighbours}{n_i}$}
+		\If{$n_3 \not\in B_1$}
+		\State $B_1 = \{n_3\} \bigcup B_1$
+		\State $x_2 = x_1 + \Call{W}{G', (n_0, n_3), 3} + \Call{W}{G', (n_1, n_3), 4} + \Call{W}{G', (n_2, n_3), 5}$
+		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3), x_2, m^-, m^+}$
+		\State $B_2 = B_1$
+
+		\For{$n_4 \in \bigcup_{i \in \{0, 1, 2, 3\}}\Call{Neighbours}{n_i}$}
+		\If{$n_4 \not\in B_2$}
+		\State $B_2 = \{n_4\} \bigcup B_2$
+		\State $x_3 = x_2 + \Sigma_{i \in \{0, 1, 2, 3\}}\Call{W}{G', (n_i, n_4), i+6}$
+		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3, n_4), x_3, m^-, m^+}$
+		\EndIf
+		\EndFor
+		\EndIf
+		\EndFor
+		\EndIf
+		\EndFor
+		\EndFunction
+		\Function{CountOrbits}{$\Delta^+$, $\Delta^-$, Nodes $S$, Edgeweights $x$, Modifier $m^-$, Modifier $m^+$}
+		\State $L^- = L[x + m^-]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N$}
+		\State $\Delta^-[S, L^-] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^-$ in $\Delta^-$}
+		\State $L^+ = L[x + m^+]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N'$}
+		\State $\Delta^+[S, L^+] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^+$ in $\Delta^+$}
+		\State \Return $(\Delta^-, \Delta^+)$
+		\EndFunction
+		\Function{W}{Network $G$, Edge $e$, Exponent $i$}
+		\State \Return $e \in \Call{Edges}{G} \ ?\  2^i : 0$ \Comment{Return $2^i$ if $G$ contains edge $e$}
+		\EndFunction
+	\end{algorithmic}
+	\caption{
+		\textbf{Pseudocode of IncGraph.} IncGraph calculates $\Delta_{G,G'}$ using a strict branch-and-bound strategy.
+	}
 	\label{code_incgraph}
-{\bf IncGraph calculates $\Delta_{G,G'}$ using a strict branch-and-bound strategy.}
+\end{figure}
 
-\paragraph*{S2 Pseudocode.}
+\begin{figure}[htb!]	
+	\begin{algorithmic}
+		\Function{BarabasiAlbert}{number of nodes $n$, degree $d$, \par\hspace*{3.72cm} number of operations $o$, offset exponent $x=1$}
+		\State $D \gets \Call{Rep}{0, n}$ \Comment{All nodes start with 0 degree}
+		\State $T \gets \Call{Rep}{\{\}, n}$ \Comment{List of targets of each node}
+		\State $S \gets \Call{Rep}{\{\}, n}$ \Comment{List of sources of each node}
+		\State $O \gets (\,)$ \Comment{Variable to store the generated operations in}
+
+		\Procedure{AddEdge}{node $i$, node $j$, add as operation $b$}
+		\State $D[i] \gets D[i] + 1$;\ $D[j] \gets D[j] + 1$ \Comment{Update degrees}
+		\State $T[i] \gets T[i] \cup \{j\}$;\ $S[j] \gets S[j] \cup \{i\}$ \Comment{Update targets and sources}
+		\If{$b$}
+		\State $O \gets O + (\textrm{ADD}, (i, j))$
+		\EndIf
+		\EndProcedure
+		\Procedure{RemoveEdge}{node $i$, node $j$, add as operation $b$}
+		\State $D[i] \gets D[i] - 1$;\ $D[j] \gets D[j] - 1$ \Comment{Update degrees}
+		\State $T[i] \gets T[i] \setminus \{j\}$;\ $S[j] \gets S[j] \setminus \{i\}$ \Comment{Update targets and sources}
+		\If{$b$}
+		\State $O \gets O + (\textrm{REM}, (i, j))$
+		\EndIf
+		\EndProcedure
+		\Procedure{AddNode}{node $i$, add as operations $b$}
+		\State $C \gets \{j \,|\, 0 \leq j < i \wedge j \not\in S[i]\}$ \Comment{Select candidate neighbours}
+		\State $W \gets (D[C] \,/\, \Sigma D[C])^x$ \Comment{Calculate weights for preferred attachment}
+		\State $X \gets $ sample $d$ neighbours from $C$ with weights $W$
+		\For{$j \in X$}
+		\State \Call{AddEdge}{$i$, $j$, $b$}
+		\EndFor
+		\EndProcedure
+		\Procedure{RemoveNode}{node $i$, add as operations $b$}
+		\While{$|T[i]| > 0$}
+		\State $j \gets \Call{Head}{T[i], 1}$
+		\State \Call{RemoveEdge}{$i$, $j$, $b$}
+		\EndWhile
+		\EndProcedure
+		\For{$i \in \{1 \,..\, m\}$}
+		\For{$j \in \{0 \,..\, i-1\}$}
+		\State \Call{AddEdge}{$i$, $j$, false} \Comment{Start with $m+1$ complete graph}
+		\EndFor
+		\EndFor
+		\For{$i \in \{m+1\,..\,n-1\}$}
+		\State \Call{AddNode}{$i$, false} \Comment{Add the rest of the nodes}
+		\EndFor
+		\State $G_0 \gets \{ (i, j) \,|\, i \in \{0\,..\,n-1\} \wedge j \in N[i]\}$ \Comment{Initial network}
+		\While{$|O| < o$} \Comment{Modify network until $O$ is sufficiently large}
+		\State $i \gets$ Sample 1 index from $\{0 \,..\, n-1\}$
+		\State \Call{RemoveNode}{$i$, true}
+		\State \Call{AddNode}{$i$, true}
+		\EndWhile
+		\State $O \gets \Call{Head}{O, o}$ 
+		\State \Return $(N_0, O)$ \Comment{Return the initial network and $o$ operations}
+		\EndFunction
+	\end{algorithmic}
+	\caption{
+		\textbf{Pseudo code for generating an evolving Barab\'asi-Albert (BA) network.} It first generates a BA network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a BA network.
+	}
 	\label{code_ba}
-{\bf Pseudo code for generating an evolving Barab\'asi-Albert (BA) network.} It first generates a BA network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a BA network.
+\end{figure}
 
-\paragraph*{S3 Pseudocode.}
+\begin{figure}[htb!]
+	\begin{algorithmic}
+		\Function{ErdosRenyi}{number of nodes $n$, number of edges $e$, number of operations $o$}
+		\State $P \gets \{(i, j) \,|\, i \in \{1 \,..\, n-1\} \wedge j \in \{0 \,..\, i-1\}\}$ \Comment{All possible interactions}
+		\State $N_0 \gets$ Sample $e$ edges from $P$
+		\State $N_c \gets N_0$
+		\While{$|O|<o$}
+		\State $e_a \gets $ Sample 1 edge from $P \setminus N_c$
+		\State $e_r \gets $ Sample 1 edge from $N_c$
+		\State $N_c \gets (N_c \setminus e_r) \cup \{e_a\}$
+		\State $O \gets O + ((\textrm{ADD}, e_a), (\textrm{REM}, e_r))$
+		\EndWhile
+		\State $O \gets \Call{Head}{O, o}$ 
+		\State \Return $(N_0, O)$ \Comment{Return the initial geometric network and $o$ operations}
+		\EndFunction
+	\end{algorithmic}
+	\caption{
+		\textbf{Pseudo code for generating an evolving Erd\H{o}s--R\'enyi (ER) network.} It first generates an ER network, and then generates $o$ operations such that at any time point, the network is or very closely resembles an ER network.
+	}
 	\label{code_er}
-{\bf Pseudo code for generating an evolving Erd\H{o}s--R\'enyi (ER) network.} It first generates an ER network, and then generates $o$ operations such that at any time point, the network is or very closely resembles an ER network.
+\end{figure}
 
-\paragraph*{S4 Pseudocode.}
+\begin{figure}[htb!]
+	\begin{algorithmic}
+		\Function{Geometric}{number of nodes $n$, number of edges $e$, \par\hspace*{2.77cm} number of operations $o$, number of dimensions $d=3$}
+		\State $P \gets$ Sample $n$ points from a multivariate continuous uniform distribution $U_{d}((0,1)^{d})$
+		\State $D \gets$ Calculate distance matrix from $P$
+		\State $N_0 \gets \Call{Head}{\Call{ArgSort}{\Call{LowerTriangle}{D}}, e}$ \Comment{Initial network}
+		\State $N_p \gets N_0$
+		\State $O \gets (\,)$ \Comment{Variable to store the generated operations in}
+		\While{$|O| < o$} \Comment{Modify network until $O$ is sufficiently large}
+		\State $i \gets $ Sample 1 index from $\{0\,..\,n-1\}$
+		\State $P[i] \gets$ Sample 1 point from a $U_{d}((0,1)^{d})$ \Comment{Give node $i$ a new location}
+		\State $D[i,] \gets D[,i] \gets $ Calculate new distances between node $i$ and all other nodes
+		\State $N_c \gets \Call{Head}{\Call{ArgSort}{\Call{LowerTriangle}{D}}, e}$ \Comment{New network}
+		\State $O_a \gets \{(\textrm{ADD}, e) \,|\, e \not\in E(N_p) \wedge e \in E(N_c) \}$ \Comment{Gather added edges}
+		\State $O_r \gets \{(\textrm{REM}, e) \,|\, e \in E(N_p) \wedge e \not\in E(N_c) \}$ \Comment{Gather removed edges}
+		\State $O \gets O + \Call{Shuffle}{O_a + O_r}$ \Comment{Append new operations to $O$}
+		\State $N_p \gets N_c$
+		\EndWhile
+		\State $O \gets \Call{Head}{O, o}$ 
+		\State \Return $(N_0, O)$ \Comment{Return the initial geometric network and $o$ operations}
+		\EndFunction
+	\end{algorithmic}
+	\caption{
+		\textbf{Pseudo code for generating an evolving geometric network.} It first generates a geometric network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a geometric network.
+	}
 	\label{code_geo}
-{\bf Pseudo code for generating an evolving geometric network.} It first generates a geometric network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a geometric network.
-
-\paragraph*{S1 Fig.}
-\label{sfig_hindex}
-{\bf Empirical measurements show a strong relation between the execution time of IncGraph and the h-index cubed of the network it was applied to.} This is in line with the findings by Eppstein et al., where counting 3-size subgraphs has an amortised time of $O(h)$ and counting 4-size subgraphs has an amortised time of $O(h^2)$.
+\end{figure}

### 2019-09-29 ch8_discussion/content.tex
added guidelines for benchmarking
minor changes

--- 2019-08-26 ch8_discussion/content.tex
@@
-\blindtext
+% DISCUSSION CONTENT
 
-\section{Overview and conclusions of the presented work}
+This chapter reflects on the impact of this work on the field, referring back to the research objectives that were defined in Chapter \ref{chap:introduction}.
 
-% notice that normalisation was absent from this dissertation.
-%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
-%
-%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.
+\section{dyngen: benchmarking with \textit{in silico} single cells}
+We developed a single-cell simulator that can be used to simulate types of single-cell omics data for which the technology does not yet exist (Chapter~\ref{chap:dyngen}).
+The \textit{in silico} single cell data can help kick-start emerging domains with low data availability more safely by making novel computational tools more reproducible.
+Indeed, the simulator has already been used to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and single cell network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
 
-\section{Future research directions}
+\section{dynbenchmark: A comparison of single-cell trajectory inference methods}
+Using this simulator and a collection of real datasets, we performed a comparison of 45 TI methods (Chapter~\ref{chap:dynbenchmark}). 
+Our contributions include writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, and developing novel metrics for comparing ground truth and predicted trajectories.
+
+We constructed a set of guidelines for end-users to help choose a TI method that is accurate, robust and fit for their application. 
+Since such guidelines were hitherto lacking, they are commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.
+
+We made our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
+Despite our best efforts, we do not observe an increase in developers performing quantitative benchmarks.
+We hypothesise causal reasons for this phenomenon and provide solutions in order to spur TI developers to perform more self-assessments (Chapter~\ref{chap:selfassessment}).
+
+Over the course of four years, we developed a lot of practical experience in benchmarking complex computational tools.
+We summarise this experience in a set of essential guidelines for benchmarking computational tools (Chapter~\ref{chap:guidelines}).
+
+%	\item Already outdated when the manuscript was published online
+%	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
+%	\item Include RNA velocity as inputs
+
+\section{dyno: A toolkit for inferring and interpreting trajectories}
+% spinoff from dynbenchmark
+% not yet published, yet a lot of activity on github
+% invest time in publishing all of the subpackages on github
+
+\section{SCORPIUS: Fast, accurate, and robust single-cell pseudotime}
+
+%\item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
+Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
+
+%\section{SCORPIUS discussion}
+%\begin{itemize}
+%	\item Extension to inferring non-linear trajectories, i.e. with principal graphs or GNG
+%\end{itemize}
+
+\section{bred: Inferring single cell regulatory networks}
+
+%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
+
+\section{incgraph: Optimising regulatory networks}
+
+%\item Every NI method has certain topological biases.
+We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).

### 2019-09-29 ch6_bred/content.tex
clean up thesis

--- 2019-08-16 ch2_dynbenchmark.tex
@@
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
-\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
-\label{chap:dynbenchmark}
+% BRED
 
-\begin{quote}
-	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
-Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
-Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
-Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
-Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
-Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
-\end{quote}
-
-\vfill
-
-Adapted from:\\
-Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
-A comparison of single-cell trajectory inference methods.
-\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
-\doi{10.1038/s41587-019-0071-9}.\\
-{\footnotesize $^*$ Equal contribution}
-\newpage
-
-\section{Introduction}
-
-Given the diversity in TI methods, it is important to quantitatively assess their performance, scalability, robustness and usability.
-Many attempts at tackling this issue have already been made \cite{haghverdi_diffusionpseudotimerobustly_2016,ji_tscanpseudotimereconstruction_2016,welch_slicerinferringbranched_2016,matsumoto_scoupprobabilisticmodel_2016,duverle_celltreebioconductorpackage_2016,cannoodt_computationalmethodstrajectory_2016,lonnberg_singlecellrnaseqcomputational_2017,campbell_probabilisticmodelingbifurcations_2017,wolf_graphabstractionreconciles_2017}, but a comprehensive comparison of TI methods across a large number of different datasets is still lacking.
-This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea of which would optimally solve their problem.
-Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.
-
-In this study, we evaluated the accuracy, scalability, stability and usability of 45 TI methods (Figure \ref{fig:figure_1}a).
-We found substantial complementarity between current methods, with different sets of methods performing most optimally depending on the characteristics of the data.
-For method users, we created an interactive set of guidelines (available at \href{http://guidelines.dynverse.org}{guidelines.dynverse.org}), which gives context-specific recommendations for method usage.
-Our evaluation also highlights some challenges for current methods, and our evaluation strategy can be useful to spearhead the development of new tools that accurately infer trajectories on ever more complex use cases.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_1.pdf}
-	\caption{
-		\textbf{Overview of several key aspects of the evaluation.}
-		\textbf{a}, A schematic overview of our evaluation pipeline.
-\textbf{b}, To make the trajectories comparable to each other, a common trajectory model was used to represent reference trajectories from the real and synthetic datasets, as well as any predictions of TI methods.
-\textbf{c}, Trajectories are automatically classified into one of seven trajectory types, with increasing complexity.
-\textbf{d}, We defined four metrics, each assessing the quality of a different aspect of the trajectory.
-The HIM score assesses the similarity between the two topologies, taking into account differences in edge lengths and degree distributions.
-The F1branches assesses the similarity of the assignment of cells onto branches.
-The cordist quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances.
-Finally, wcorfeatures quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.
-	}
-	\label{fig:figure_1}
-\end{figure}
-
-\section{Results}
-
-\subsection{Trajectory inference methods}
-To make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (Figure \ref{fig:figure_1}b).
-In this model, the overall topology is represented by a network of ‘milestones’, and the cells are placed within the space formed by each set of connected milestones.
-Although almost every method returned a unique set of outputs, we were able to classify these outputs into seven distinct groups (Figure \ref{fig:supfigure_1}) and we wrote a common output converter for each of these groups (Figure \ref{fig:figure_2}a).
-When strictly required, we also provided prior information to the method.
-These different priors can range from weak priors that are relatively easy to acquire, such as a start cell, to strong priors, such as a known grouping of cells, that are much harder to know a priori, and which can potentially introduce a large bias into the analysis (Figure \ref{fig:figure_2}a).
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_1.pdf}
-	\caption{
-		\textbf{A common interface for TI methods.}
-		\textbf{a} The input and output of each TI method is standardized.
-As input, each TI method receives either raw or normalized counts, several parameters, and a selection of prior information.
-After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model.
-This common model then allows to perform common analysis functions on trajectory models produced by any TI method.
-\textbf{b} Illustrations of the specific transformations performed by each of the wrapper functions.
-	}
-	\label{fig:supfigure_1}
-\end{figure}
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_2.pdf}
-	\caption{
-		\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
-		\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies.
-The methods are grouped vertically based on the most complex trajectory type they can infer.
-\textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation.
-Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig.
-2.
-	}
-	\label{fig:figure_2}
-\end{figure}
-
-The largest difference between TI methods is whether a method fixes the topology and, if it does not, what kind of topology it can detect.
-We defined seven possible types of topology, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
-Most methods either focus on inferring linear trajectories or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (Figure \ref{fig:figure_2}a).
-
-We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript.
-Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b).
-We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig.
-2), after which we conclude with guidelines for method users and future perspectives for method developers.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_3.pdf}
-	\caption{	
-		\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
-		\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}.
-\textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types.
-The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type.
-\textbf{c}, Predicted execution times for varying numbers of cells and features (no.
-of cells $\times$ no.
-of features).
-Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features.
-k, thousands; m, millions; cor, correlation.
-\textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method.
-\textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category.
-Off-the-shelf methods were directly implemented in R and thus do not have a usability score.
-	}
-	\label{fig:figure_3}
-\end{figure}
-
-\subsection{Accuracy}
-
-We defined several metrics to compare a prediction to a reference trajectory (Supplementary Note 1).
-Based on an analysis of their robustness and conformity to a set of rules (Supplementary Note 1), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures).
-The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance.
-These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table 2).
-Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}.
-All other real datasets were classified as ‘silver standard’.
-For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}.
-For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.
-
-We found that method performance was very variable across datasets, indicating that there is no ‘one-size-fits-all’ method that works well on every dataset (Figure \ref{fig:supfigure_3}a).
-Even methods that can detect most of the trajectory types, such as PAGA, RaceID/StemID and SLICER were not the best methods across all trajectory types (Figure \ref{fig:figure_3}b).
-The overall score between the different dataset sources was moderately to highly correlated (Spearman rank correlation between 0.5–0.9) with the scores on real datasets containing a gold standard (Figure \ref{fig:supfigure_3}b), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data.
-On the other hand, the different metrics frequently disagreed with each other, with Monocle and PAGA Tree scoring better on the topology scores, whereas other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (Figure \ref{fig:figure_3}b).
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_3.pdf}
-	\caption{
-		\textbf{Accuracy of trajectory inference methods.}
-		\textbf{a} Overall score for all methods across 339 datasets, colored by the source of the datasets.
-Black line indicates the mean.
-\textbf{b} Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard, across all methods (n = 46, after filtering out methods that errored too frequently).
-Shown in the top left is the Pearson correlation.
-\textbf{c} Bias in the overall score towards trajectory types for all methods across 339 datasets.
-Black line indicates the mean.
-\textbf{d} Distributions of the difference in size between predicted and reference topologies.
-A positive difference means that the topology predicted by the method is more complex than the one in the reference.
-	}
-	\label{fig:supfigure_3}
-\end{figure}
-
-The performance of a method was strongly dependent on the type of trajectory present in the data (Figure \ref{fig:figure_3}b).
-Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories (Figure \ref{fig:supfigure_3}c).
-This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, whereas those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (Figure \ref{fig:supfigure_3}d).
-This analysis therefore indicates that detecting the right topology is still a difficult task for most of these methods, because methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.
-
-The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods.
-To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods.
-A top model in this case was defined as a model with an overall score of at least 95$\%$ as the best model.
-On all datasets, using one method resulted in getting a top model about 27$\%$ of the time.
-This increased up to 74$\%$ with the addition of six other methods (Figure \ref{fig:figure_4}a).
-The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA.
-We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (Figure \ref{fig:figure_4}b), although in these cases less methods were necessary to obtain at least one top model for a given dataset.
-Altogether, this shows that there is considerable complementarity between the different methods and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori.
-Moreover, it also opens up the possibilities for new ensemble methods that utilize this complementarity.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_4.pdf}
-	\caption{
-		\textbf{Complementarity between different trajectory inference methods.}
-		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95$\%$ of the best model) when applied to all datasets.
-\textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types.
-For this figure, we did not include any methods requiring a cell grouping or a time course as prior information.
-	}
-	\label{fig:figure_4}
-\end{figure}
-
-\subsection{Scalability}
-
-While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million \cite{svensson_exponentialscalingsinglecell_2018}.
-Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features \cite{cao_jointprofilingchromatin_2018}.
-To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets.
-We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a).
-As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Supplementary Fig.
-5), and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig.
-2a).
-
-\begin{figure}
-	\centering\includegraphics[width=0.9\linewidth]{fig/dynbenchmark/supfigure_4.pdf}
-	\caption{
-		\textbf{Scalability of trajectory inference methods.}
-		\textbf{a} Three examples of average observed running times across five datasets (left) and the predicted running time (right).
-\textbf{b} Overview of the scalability results of all methods, ordered by their average predicted running time from (a).
-We predicted execution times and memory usage for each method with increasing number of features or cells, and used these values to classify each method into sublinear, linear, quadratic and superquadratic based on the shape of the curve.
-	}
-	\label{fig:supfigure_4}
-\end{figure}
-
-We found that the scalability of most methods was overall very poor, with most graph and tree methods not finishing within an hour on a dataset with ten thousand cells and ten thousand features (Figure \ref{fig:figure_3}c), which is around the size of a typical droplet-based single-cell dataset \cite{svensson_exponentialscalingsinglecell_2018}.
-Running times increased further with increasing number of cells, with only a handful of graph/tree methods completing within a day on a million cells (PAGA, PAGA Tree, Monocle DDRTree, Stemnet and GrandPrix).
-Some methods, such as Monocle DDRTree and GrandPrix, also suffered from unsatisfactory running times when given a high number of features.
-
-Methods with a low running time typically had two defining aspects: they had a linear time complexity with respect to the features and/or cells, and adding new cells or features led to a relatively low increase in time (Figure \ref{fig:supfigure_4}b).
-We found that more than half of all methods had a quadratic or superquadratic complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (Figure \ref{fig:supfigure_4}b).
-
-We also assessed the memory requirements of each method (Supplementary Fig.
-2c).
-Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features.
-Notably, the memory requirements were very high for several methods on datasets with high numbers of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).
-
-Altogether, the scalability analysis indicated that the dimensions of the data are an important factor in the choice of method, and that method development should pay more attention to maintaining reasonable running times and memory usage.
-
-\subsection{Stability}
-
-It is not only important that a method is able to infer an accurate model in a reasonable time frame, but also that it produces a similar model when given very similar input data.
-To test the stability of each method, we executed each method on ten different subsamples of the datasets (95$\%$ of the cells, 95$\%$ of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (Figure \ref{fig:figure_3}d).
-
-Given that the trajectories of methods that fix the topology either algorithmically or through a parameter are already very constrained, it is to be expected that such methods tend to generate very stable results.
-Nonetheless, some fixed topology methods still produced slightly more stable results, such as SCORPIUS and MATCHER for linear methods and MFA for multifurcating methods.
-Stability was much more diverse among methods with a free topology.
-Slingshot produced more stable models than PAGA (Tree), which in turn produced more stable results than pCreode and Monocle DDRTree.
-
-\subsection{Usability}
-
-While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_tensimplerules_2017}.
-We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}).
-It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB).
-The reasons for not including certain tools are all discussed on our repository (https://github.com/dynverse/dynmethods/issues?q=label:unwrappable).
-Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.
-
-We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Supplementary Fig.
-6).
-While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Supplementary Fig.
-6 right) and we believe that these should receive extra attention from developers.
-Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_bestpracticesscientific_2014,artaza_top10metrics_2016}.
-Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods.
-We observed no clear relation between usability and method accuracy or usability (Figure \ref{fig:figure_2}b).
-
-\begin{table}
-	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}
-
-	\centering\begingroup\fontsize{7}{9}\selectfont
-
-	\begin{tabular}{p{3.5cm}p{9cm}p{2.5cm}}
-		\toprule
-		Aspect & Items & References \\ \midrule
-
-		\multicolumn{3}{l}{\textbf{Availability}}\\*
-
-		Open source & (1) Method's code is freely available (2) The code can be run on a freely available platform & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
-
-		Version control & The code is available on a public version controlled repository, such as Github & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
-
-		Packaging & (1) The code is provided as a "package", exposing functionality through functions or shell commands (2) The code can be easily installed through a repository such as CRAN, Bioconductor, PyPI, CPAN, debian packages, … & \cite{lee_rpackagespackagedevelopment_2017,wickham_packagesorganizetest_2015,jimenez_foursimplerecommendations_2017,silva_generalguidelinesbiomedical_2017}\\*
-
-		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*
-
-		License & (1) The code is licensed (2) License allows academic use & \cite{lee_rpackagespackagedevelopment_2017,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
-
-		Interface & (1) The tool can be run using a graphical user interface, either locally or on a web server (2) The tool can be run through the command line or through a programming language & \cite{silva_generalguidelinesbiomedical_2017}\\
-
-		\midrule
-		\multicolumn{3}{l}{\textbf{Code quality}} \\*
-
-		Function and object naming & (1) Functions/commands have well chosen names (2) Arguments/parameters have well chosen names & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\*
-
-		Code style & (1) Code has a consistent style (2) Code follows (basic) good practices in the programming language of choice, for example PEP8 or the tidyverse style guide & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016}\\*
-
-		Code duplication & Duplicated code is minimal & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\
-
-		Self-contained functions & The method is exposed to the user as self-contained functions or commands & \cite{anderson_writinggreatscientific_2016,taschuk_tensimplerules_2017,silva_generalguidelinesbiomedical_2017}\\*
-
-		Plotting & Plotting functions are provided for the final and/or intermediate results & \\*
-
-		Dummy proofing & Package contains dummy proofing, i.e.
-testing whether the parameters and data supplied by the user make sense and are useful & \cite{lee_rpackagespackagedevelopment_2017,karimzadeh_topconsiderationscreating_}\\
-
-		\midrule
-		\multicolumn{3}{l}{\textbf{Code assurance}} \\*
-
-		Unit testing & Method is tested using unit tests & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
-
-		Unit testing & Tests are run automatically using functionality from the programming language & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*
-
-		Continuous integration & The method uses continuous integration, for example on Travis CI & \cite{beaulieu-jones_reproducibilitycomputationalworkflows_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*
-
-		Code coverage & (1) The code coverage of the repository is assessed. (2) What is the percentage of code coverage & \\
-
-		\midrule
-		\multicolumn{3}{l}{\textbf{Documentation}} \\*
-
-		Support & (1) There is a support ticket system, for example on Github (2) The authors respond to tickets and issues are resolved within a reasonable time frame & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*
-
-		Development model & (1) The repository separates the development code from master code, for example using git master en developer branches (2) The repository has created releases, or several branches corresponding to major releases.
-(3) The repository has branches for the development of separate features.
-& \cite{driessen_successfulgitbranching_2010}\\*
-
-		Tutorial & (1) A tutorial or vignette is available (2) The tutorial has example results (3) The tutorial has real example data (4) The tutorial showcases the method on several datasets (1=0, 2=0.5, >2=1) & \cite{wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017,karimzadeh_topconsiderationscreating_,boulesteix_tensimplerules_2015}\\*
-
-		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
-
-		Inline documentation & Inline documentation is present in the code & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
-
-		Parameter transparency & All important parameters are exposed to the user & \cite{taschuk_tensimplerules_2017}\\
-
-		\midrule
-		\multicolumn{3}{l}{\textbf{Behaviour}} \\*
-
-		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{puget_greendiceare_2016}\\*
-
-		Unexpected output & (1) No unexpected output messages are generated by the method (2) No unexpected files, folders or plots are generated (3) No unexpected warnings during runtime or compilation are generated & \cite{artaza_top10metrics_2016}\\*
-
-		Trajectory format & The postprocessing necessary to extract the relevant output from the method is minimal (1), moderate (0.5) or extensive (0) & \\*
-
-		Prior information & Prior information is required (0), optional (1) or not required (1) & \\*
-
-		\midrule
-		\multicolumn{3}{l}{\textbf{Paper}} \\*
-
-		Publishing & The method is published & \\*
-
-		Peer review & The paper is published in a peer-reviewed journal & \cite{karimzadeh_topconsiderationscreating_,gannon_essentialrolepeer_2001,baldwin_refereeswetrust_2017}\\*
-
-		Evaluation on real data & (1) The paper shows the method's usefulness on several (1), one (0.25) or no real datasets.
-(2) The paper quantifies the accuracy of the method given a gold or silver standard trajectory & \cite{aniba_issuesbioinformaticsbenchmarking_2010,jelizarow_overoptimismbioinformaticsillustration_2010}\\*
-
-		Evaluation of robustness & The paper assessed method robustness (to eg.
-noise, subsampling, parameter changes, stability) in one (0.5) or several (1) ways & \cite{karimzadeh_topconsiderationscreating_,aniba_issuesbioinformaticsbenchmarking_2010,boulesteix_tensimplerules_2015,jelizarow_overoptimismbioinformaticsillustration_2010}\\
-
-		\bottomrule
-	\end{tabular}
-	\endgroup{}
-\end{table}
-
-\section{Discussion}
-
-In this study, we presented a large-scale evaluation of the performance of 45 TI methods.
-By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the accuracy of the methods on more than 200 datasets.
-We also assessed several other important quality measures, such as the quality of the method’s implementation, the scalability to hundreds of thousands of cells and the stability of the output on small variations of the datasets.
-
-Based on the results of our benchmark, we propose a set of practical guidelines for method users (Figure \ref{fig:figure_5} and guidelines.dynverse.org).
-We postulate that, as a method’s performance is heavily dependent on the trajectory type being studied, the choice of method should currently be primarily driven by the anticipated trajectory topology in the data.
-For most use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure.
-In each of these use cases, our evaluation suggests a different set of optimal methods, as shown in Figure \ref{fig:figure_5}.
-Several other factors will also impact the choice of methods, such as the dimensions of the dataset and the prior information that is available.
-These factors and several others can all be dynamically explored in our interactive app (guidelines.dynverse.org).
-This app can also be used to query the results of this evaluation, such as filtering the datasets or changing the importance of the evaluation metrics for the final ranking.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_5.pdf}
-	\caption{
-		\textbf{Practical guidelines for method users.}
-		As the performance of a method mostly depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user’s existing knowledge about the expected topology in the data.
-We therefore devised a set of practical guidelines, which combines the method’s performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory.
-Methods to the right are ranked according to their performance on a particular (set of) trajectory type.
-Further to the right are shown the accuracy (+: scaled performance $\geq$ 0.9, $\pm$: >0.6), usability scores (+:$\geq$0.9, $\pm$ $\geq$0.6), estimated running times and required prior information.
-k, thousands; m, millions.
-	}
-	\label{fig:figure_5}
-\end{figure}
-
-When inferring a trajectory on a dataset of interest, it is important to take two further points into account.
-First, it is critical that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods.
-This is to make sure that the prediction is not biased due to the given parameter setting or the particular algorithm underlying a TI method.
-The value of using different methods is further supported by our analysis indicating substantial complementarity between the different methods.
-Second, even if the expected topology is known, it can be beneficial to also try out methods that make less assumptions about the trajectory topology.
-When the expected topology is confirmed using such a method, it provides additional evidence to the user.
-When a more complex topology is produced, this could indicate that the underlying biology is much more complex than anticipated by the user.
-
-Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}.
-Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}).
-Using our framework, this figure can be recreated using only a couple of lines of R code (https://methods.dynverse.org).
-In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses.
-In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_6.pdf}
-	\caption{
-		\textbf{Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.}
-		Trajectories inferred by each method were projected to a common dimensionality reduction using multidimensional scaling.
-For each dataset, we also calculated a ‘consensus’ prediction, by calculating the cordist between each pair of models and picking the model with the highest score on average.
-\textbf{a}, The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC.
-\textbf{b}, The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts.
-\textbf{c}, A synthetic dataset generated by dyntoy, containing four disconnected trajectories.
-\textbf{d}, A synthetic dataset generated by dyngen, containing a cyclic trajectory.
-	}
-	\label{fig:figure_6}
-\end{figure}
-
-Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories (Figure \ref{fig:figure_6}a,b).
-However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analyzing single-cell omics datasets with complex trajectories.
-Foremost, new methods should focus on improving the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be identified using a dimensionality reduction method (Figure \ref{fig:figure_6}c,d).
-Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell omics field.
-Finally, new tools should be designed to scale well with the increasing number of cells and features.
-We found that only a handful of current methods can handle datasets with more than 10,000 cells within a reasonable time frame.
-To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method on the different measures proposed in this study at \href{https://benchmark.dynverse.org}{https://benchmark.dynverse.org}.
-
-We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods.
-However, ‘good-yet-not-the-best’ methods \cite{norel_selfassessmenttrapcan_2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases.
-This is also supported by our analysis of method complementarity.
-Some examples for the latter include PhenoPath, which can include additional covariates in its model, ouija, which returns a measure of uncertainty of each cell’s position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.
-
-\section{Methods}
-
-\subsection{Trajectory inference methods}
-
-We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
-We welcome any contributions by creating an issue at https://methods.dynverse.org.
-
-Methods were excluded from the evaluation based on several criteria: (1) not freely available; (2) no code available; (3) superseded by another method; (4) requires data types other than expression; (5) no programming interface; (6) unresolved errors during wrapping; (7) too slow (requires more than 1 h on a 100 $\times$ 100 dataset); (8) does not return an ordering; and (9) requires additional user input during the algorithm (other than prior information).
-The discussions on why these methods were excluded can be found at \href{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}.
-In the end, we included 45 methods in the evaluation.
-
-\subsection{Method wrappers}
-
-To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}).
-These containers are automatically built and tested using Travis continuous integration (\href{https://travis-ci.org/dynverse}{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity.
-For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts).
-This script reads in the input data, runs the method and outputs the files required to construct a trajectory.
-We also created a script to generate an example dataset, which is used for automated testing.
-
-We used the Github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores.
-About one-third of the authors responded and we improved the wrappers based on their feedback.
-These discussions can be viewed on Github: \href{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}{https://github.com/dynverse/dynmethods/issues?q=label:method\_discussion}
-
-\subsubsection{Method input}
-
-As input, we provided each method with either the raw count data (after cell and gene filtering) or normalized expression values, based on the description in the method documentation or from the study describing the method.
-A large portion of the methods requires some form of prior information (for example, a start cell) to be executable.
-Other methods optionally allow the exploitation of certain prior information.
-Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes or even a grouping of cells into cell states.
-Providing prior information to a TI method can be both a blessing and a curse.
-In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives.
-On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge.
-Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well-studied systems.
-
-The prior information was extracted from the reference trajectory as follows:
-
-\begin{itemize}
-	\item \textbf{Start cells}: the identity of one or more start cells.
-For both real and synthetic data, a cell was chosen that was the closest (in geodesic distance) to each milestone with only outgoing edges.
-For ties, one random cell was chosen.
-For cyclic datasets, a random cell was chosen.
-	\item \textbf{End cells}: the identity of one or more end cells. This is similar to the start cells, but now for every state with only incoming edges.
-	\item \textbf{No. of end states}: number of terminal states, i.e., the number of milestones with only incoming edges.
-	\item \textbf{Grouping}: for each cell a label showing which state/cluster/branch it belongs to.
-For real data, the states were from the gold/silver standard.
-For synthetic data, each milestone was seen as one group and cells were assigned to their closest milestone.
-	\item \textbf{No.
-of branches}: number of branches/intermediate states.
-For real data, this was the number of states in the gold/silver standard.
-For synthetic data, this was the number of milestones.
-	\item \textbf{Discrete time course}: for each cell a time point from which it was sampled.
-If available, this was directly extracted from the reference trajectory; otherwise the geodesic distance from the root milestone was used.
-For synthetic data, the simulation time was uniformily discretized into four timepoints.
-	\item \textbf{Continuous time course}: for each cell a time point from which it was sampled.
-For real data, this was equal to the discrete time course.
-For synthetic data, we used the internal simulation time of each simulator.
-\end{itemize}
-
-\subsubsection{Common trajectory model}
-
-Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps.
-We therefore post-processed the output of each method into a common probabilistic trajectory model (Figure \ref{fig:supfigure_1}a).
-This model consisted of three parts.
-(1) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them.
-(2) The milestone percentages contain, for each cell, its position between milestones and sums for each cell to one.
-(3) The regions of delayed commitment define connections between three or more milestones.
-These must be explicitly defined in the trajectory model and per region one milestone must be directly connected to all other milestones of the region.
-
-Depending on the output of a method, we used different strategies to convert the output to our model (Figure \ref{fig:supfigure_1}b).
-Special conversions are denoted by an asterisk and will be explained in more detail in the second list below.
-
-\begin{itemize}
-	\item \textbf{Type 1, direct}: CALISTA*, DPT*, ElPiGraph, ElPiGraph cycle, ElPiGraph linear, MERLoT, PAGA, SLICE*, Slingshot, URD* and Wishbone.
-The wrapped method directly returned a network of milestones, the regions of delayed commitment and for each cell it is given to what extent it belongs to a milestone.
-In some cases, this indicates that additional transformations were required for the method, not covered by any of the following output formats.
-Some methods returned a branch network instead of a milestone network and this network was converted by calculating the line graph of the branch network.
-	\item \textbf{Type 2, linear pseudotime}: Component 1, Embeddr, FORKS, MATCHER, ouija, ouijaflow, PhenoPath, pseudogp, SCIMITAR, SCORPIUS, topslam, TSCAN, Wanderlust and Waterfall.
-The method returned a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
-	\item \textbf{Type 3, cyclical pseudotime}: Angle and reCAT.
-The method returned a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones.
-These milestones were positioned at pseudotime 0, 1/3 and 2/3.
-	\item \textbf{Type 4, end state probability}: FateID, GPfates, GrandPrix, MFA*, SCOUP and STEMNET.
-The method returned a pseudotime and for each cell and end state a probability (Pr) for how likely a cell will end up in a certain end state.
-This was translated into a star-shaped milestone network, with one starting milestone (M0) and several outer milestones (Mi), with regions of delayed commitment between all milestones.
-The milestone percentage of a cell to one of the outer milestones was equal to pseudotime$\times$PrMi.
-The milestone percentage to the starting milestone was equal to 1 $-$ pseudotime.
-	\item \textbf{Type 5, cluster assignment}: Mpath and SCUBA.
-The method returned a milestone network and an assignment of each cell to a specific milestone.
-Cells were positioned onto the milestones they are assigned to, with milestone percentage equal to 1.
-	\item \textbf{Type 6, orthogonal projection}: MST, pCreode and RaceID/StemID.
-The method returned a milestone network, and a dimensionality reduction of the cells and milestones.
-The cells were projected to the closest nearest segment, thus determining the cells’ position along the milestone network.
-If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell.
-For these methods, we usually wrote two wrappers, one which included the projection and one without.
-	\item\textbf{ Type 7, cell graph}: CellRouter, CellTrails, cellTree Gibbs, cellTree maptpx, cellTree VEM, Monocle DDRTree, Monocle ICA, Sincell* and SLICER.
-The method returned a network of cells and which cell–cell transitions were part of the ‘backbone’ structure.
-Backbone cells with degree $\neq$ 2 were regarded as milestones and all other cells were placed on transitions between the milestones.
-If a method did not return a distance between pairs of cells, the cells were uniformly positioned between the two milestones.
-Otherwise, we first calculated the distance between two milestones as the sum of the distances between the cells and then divided the distance of each pair of cells with the total distance to get the milestone percentages.
-\end{itemize}
-
-Special conversions were necessary for certain methods:
-
-\begin{itemize}
-	\item \textbf{CALISTA}: We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest.
-The cluster probabilities of the two selected milestones were then used as milestone percentages.
-This was then processed as a type 1, direct, method.
-	\item \textbf{DPT}: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contained the cells that were assigned to the ‘unknown’ branch) and three terminal milestones, each corresponding to a tip point.
-This was then processed as a type 1, direct, method.
-	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25$\%$.
-This was then processed as a type 7, cell graph, method.
-	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\href{https://research.cchmc.org/pbge/slice.html}{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE.
-This was then processed as a type 1, direct, method.
-	\item \textbf{MFA}: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as a type 4, end state probabilities, method.
-	\item \textbf{URD}: We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.
-
-\end{itemize}
-
-More information on how each method was wrapped can be found within the comments of each wrapper script, listed at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.
-
-\subsubsection{Off-the-shelf methods}
-
-For baseline performance, we added several ‘off-the-shelf’ TI methods that can be run using a few lines of code in R.
-
-\begin{itemize}
-
-	\item \textbf{Component 1}: This method returns the first component of a principal component analysis (PCA) dimensionality reduction as a linear trajectory.
-This method is especially relevant as it has been used in a few studies already \cite{kouno_temporaldynamicstranscriptional_2013,zeng_pseudotemporalorderingsingle_2017}.
-
-	\item \textbf{Angle}: Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA and uses this angle as a pseudotime for generating a cyclical trajectory.
-
-	\item \textbf{MST}: This method performs PCA dimensionality reduction, followed by clustering using the R mclust package, after which the clusters are connected using a minimum spanning tree.
-The trees are orthogonally projected to the nearest segment of the tree.
-This baseline is highly relevant as many methods follow the same methodology: dimensionality reduction, clustering, topology inference and project cells to topology.
-
-\end{itemize}
-
-\subsection{Trajectory types}
-
-We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (Figure \ref{fig:figure_1}c).
-These trajectory types start from the most general trajectory type, a disconnected graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types are reached: linear and cyclical.
-A disconnected graph is a graph in which only one edge can exist between two nodes.
-A (connected) graph is a disconnected graph in which all nodes are connected.
-An acyclic graph is a graph containing no cycles.
-A tree is an acyclic graph containing no convergences (no nodes with in-degree higher than 1).
-A convergence is an acyclic graph in which only one node has a degree larger than 1 and this same node has an in-degree of 1.
-A multifurcation is a tree in which only one node has a degree larger than 1.
-A bifurcation is a multifurcation in which only one node has a degree equal to 3.
-A linear topology is a graph in which no node has a degree larger than 3.
-Finally, a cycle is a connected graph in which every node has a degree equal to 2.
-In most cases, a method that was able to detect a complex trajectory type was also able to detect less complex trajectory types, with some exceptions shown in Figure \ref{fig:figure_2}a.
-
-For simplicity, we merged the bifurcation and convergence trajectory type, and the acyclic graph and connected graph trajectory type in the main figures of the paper.
-
-\subsection{Real datasets}
-
-We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table 2).
-The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}).
-Whenever possible, we preferred to start from the raw counts data.
-These raw counts were all normalized and filtered using a common pipeline, as discussed later.
-Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation.
-In the end, we included 110 datasets for this evaluation.
-
-For each dataset, we extracted a reference trajectory, consisting of two parts: the cellular grouping (milestones) and the connections between these groups (milestone network).
-The cellular grouping was provided by the authors of the original study, and we classified it as a gold standard when it was created independently from the expression matrix (such as from cell sorting, the origin of the sample, the time it was sampled or cellular mixing) or as a silver standard otherwise (usually by clustering the expression values).
-To connect these cell groups, we used the original study to determine the network that the authors validated or otherwise found to be the most likely.
-In the end, each group of cells was placed on a milestone, having a percentage of 1 for that particular milestone.
-The known connections between these groups were used to construct the milestone network.
-If there was biological or experimental time data available, we used this as the length of the edge; otherwise we set all the lengths equal to one.
-
-\subsection{Synthetic datasets}
-
-To generate synthetic datasets, we used four different synthetic data simulators:
-
-\begin{itemize}
-	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
-	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
-	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulos_prossttprobabilisticsimulation_2018}
-	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
-\end{itemize}
-
-For every simulator, we took great care to make the datasets as realistic as possible.
-To do this, we extracted several parameters from all real datasets.
-We calculated the number of differentially expressed features within a trajectory using a two-way Mann–Whitney U test between every pair of cell groups.
-These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5$\%$ of cells, and had at least a fold-change of 2.
-We also calculated several other parameters, such as drop-out rates and library sizes using the Splatter package \cite{zappia_splattersimulationsinglecell_2017}.
-These parameters were then given to the simulators when applicable, as described for each simulator below.
-Not every real dataset was selected to serve as a reference for a synthetic dataset.
-Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (with $k = 10$, implemented in the R cluster package) to generate synthetic data.
-
-\subsubsection{dyngen}
-
-The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
-At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
-We simulated a total of 110 datasets, with 11 different topologies.
-
-\hfill\break
-\textit{Network generation}
-\hfill\break
-
-One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
-The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
-One driver of bifurcation seems to be mutual antagonism, where genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
-Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
-Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
-
-To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
-Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
-The structure of these module networks is available at \href{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}.
-
-From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added.
-For each dataset, we used the same number of genes as were differentially expressed in the real datasets.
-5$\%$ of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module).
-We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module.
-To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits \cite{marbach_tissuespecificregulatorycircuits_2016}), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the page\_rank function of the \textit{igraph} package.
-
-\hfill\break
-\textit{Simulation of gene regulatory systems using thermodynamic models}
-\hfill\break
-
-To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
-In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):
-
-$$
-\begin{aligned}
-\label{eq:mrna_ode}
-\frac{dx_i}{dt} &= \underbrace{m \times f(y_1, y_2, ...)}_\text{production} - \underbrace{\lambda \times x_i}_\text{degradation}
-\end{aligned}
-$$
-$$
-\begin{aligned}
-\label{eq:prot_ode}
-\frac{dy_i}{dt} &= \underbrace{r \times x_i}_\text{production} - \underbrace{\Lambda \times y_i}_\text{degradation}
-\end{aligned}
-$$
-
-where $m$, $\lambda$, $r$ and $\Lambda$ represent production and degradation rates, the ratio of which determines the maximal gene and protein expression.
-The two types of equations are coupled because the production of protein $y_i$ depends on the amount of gene expression $x_i$, which in turn depends on the amount of other proteins through the activation function $f(y_1, y_2, ...)$.
-
-The activation function is inspired by a thermodynamic model of gene regulation, in which the promoter of a gene can be bound or unbound by a set of transcription factors, each representing a certain state of the promoter.
-Each state is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state.
-The production rate of the gene is calculated by combining the probabilities of the promoter being in each state with the relative activation:
-
-$$
-\begin{aligned}
-\label{eq:input_function_probabilities}
-f(y_1, y_2, ..., y_n) = \sum_{j \in \{0, 1, ..., n^2\}} \alpha_j \times P_j
-\end{aligned}
-$$
-
-The probability of being in a state is based on the thermodynamics of transcription factor binding. When only one transcription factor is bound in a state:
-$$
-\begin{aligned}
-P_j \propto \nu = \left(\frac{y}{k}\right)^{n}
-\end{aligned}
-$$
-
-where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
-$$
-\begin{aligned}
-P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
-\end{aligned}
-$$
-
-where $\rho$ represents the cooperativity of binding between the different transcription factors.
-
-$P_i$ is only proportional to $\nu$ because $\nu$ is normalized such that $\sum_{i} P_i = 1$.
-
-To each differential equation, we added an additional stochastic term: 
-$$
-\begin{aligned}
-\frac{dx_i}{dt} & = m \times f(y_1, y_2, ...) - \lambda \times x_i + \eta \times \sqrt{x_i} \times \Delta W_t \\
-\frac{dy_i}{dt} & = r \times x_i - \Lambda \times y_i + \eta \times \sqrt{y_i} \times \Delta W_t
-\end{aligned}
-$$
-
-with $\Delta W_t \sim \mathcal{N}(0, h)$.
-
-Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows.
-$e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.
-
-$$
-\begin{aligned}
-r & = \mathcal{U}(10, 200) \\
-d & = \mathcal{U}(2, 8) \\
-p & = \mathcal{U}(2, 8) \\
-q & = \mathcal{U}(1, 5) \\
-a_0 & = \begin{cases}1 & \text{if } |e| = 0 \\ 1 & \text{if } \forall x \in e \text{, } x = -1 \\ 0 & \text{if } \forall x \in e \text{, } x = 1 \\ 0.5 & \text{otherwise}\end{cases} \\
-a_i & = \begin{cases}0 & \text{if } \exists x \in e_i \text{, } x = -1 \\ 1 & \text{otherwise}\end{cases} \\
-s & = \mathcal{U}(1, 20) \\
-k & = y_{max} / (2 * s) \text{,} \\
-\ & \quad \text{where } y_{max} = r / d \times p / q \\
-c & = \mathcal{U}(1, 4)
-\end{aligned}
-$$
-
-We converted each ODE to an SDE by adding a chemical Langevin equation, as described in \cite{schaffter_genenetweaversilicobenchmark_2011}.
-These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 8$.
-The total simulation time varied between 5 for linear and bifurcating datasets, 10 for consecutive bifurcating, trifurcating and converging datasets, 15 for bifurcating converging datasets and 30 for linear long, cycle and bifurcating loop datasets.
-The burn-in period was for each simulation 2.
-Each network was simulated 32 times.
-
-\hfill\break
-\textit{Simulation of the single-cell RNA-seq experiment}
-\hfill\break
-
-For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in.
-These cells were sampled uniformly across the different steps of the 32 simulations.
-Next, we used the Splatter package \cite{zappia_splattersimulationsinglecell_2017} to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
-We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$.
-These were combined with the expression levels of the genes simulated within a trajectory.
-Next, true counts were simulated using $Y'_{i,j} \sim \text{Poisson}(\lambda_{i,j})$.
-Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
-Both $x_0$ (the midpoint for the dropout logistic function) and $k$ (the shape of the dropout logistic function) were estimated by Splatter.
-
-This count matrix was then filtered and normalised using the pipeline described below.
-
-\hfill\break
-\textit{Gold standard extraction}
-\hfill\break
-
-Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward.
-Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable.
-We therefore first constructed a backbone expression profile for each branch within the trajectory.
-To do this, we first defined in which order the expression of the modules is expected to change, and then generated a backbone expression profile in which the expression of these modules increases and decreases smoothly between 0 and 1.
-We also smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
-We used dynamic time warping, implemented in the dtw R package \cite{giorgino_computingvisualizingdynamic_2009,tormene_matchingincompletetime_2009}, with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
-In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.
-
-\subsubsection{dyntoy}
-
-For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (\url{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}) .
-We created 12 topology generators (described below), and with 10 datasets per generator, this lead to a total of 120 datasets.
-
-We created a set of topology generators, were $B(n, p)$ denotes a binomial distribution, and $U(a, b)$ denotes a uniform distribution:
-
-* Linear and cyclic, with number of milestones $\sim B(10, 0.25)$
-* Bifurcating and converging, with four milestones
-* Binary tree, with number of branching points $\sim U(3, 6)$
-* Tree, with number of branching points $\sim U(3, 6)$ and maximal degree $\sim U(3, 6)$
-
-For more complex topologies we first calculated a random number of "modifications" $\sim U(3, 6)$ and a $\textit{deg}_{\textit{max}} \sim B(10, 0.25) + 1$.
-For each type of topology, we defined what kind of modifications are possible: divergences, loops, convergences and divergence-convergence.
-We then iteratively constructed the topology by uniformly sampling from the set of possible modifications, and adding this modification to the existing topology.
-For a divergence, we connected an existing milestone to a number of a new milestones.
-Conversely, for a convergence we connected a number of new nodes to an existing node.
-For a loop, we connected two existing milestones with a number of milestones in between.
-Finally for a divergence-convergence we connected an existing milestone to several new milestones which again converged on a new milestone.
-The number of nodes was sampled from $\sim B(\textit{deg}_{\textit{max}} - 3, 0.25) + 2$
-
-* Looping, allowed loop modifications
-* Diverging-converging, allowed divergence and converging modifications
-* Diverging with loops, allowed divergence and loop modifications
-* Multiple looping, allowed looping modifications
-* Connected, allowed looping, divergence and convergence modifications
-* Disconnected, number of components sampled from $\sim B(5, 0.25) + 2$, for each component we randomly chose a topology from the ones listed above
-
-After generating the topology, we sampled the length of each edge $\sim U(0.5, 1)$.
-We added regions of delayed commitment to a divergence in a random half of the cases.
-We then placed the number of cells (same number as from the reference real dataset), on this topology uniformly, based on the length of the edges in the milestone network.
-
-For each gene (same number as from the reference real dataset), we calculated the Kamada-Kawai layout in 2 dimensions, with edge weight equal to the length of the edge.
-For this gene, we then extracted for each cell a density value using a bivariate normal distribution with $\mu \sim U(x_{\textit{min}}, x_{\textit{min}})$ and $\sigma \sim U(x_{\textit{min}}/10, x_{\textit{min}}/8)$.
-We used this density as input for a zero-inflated negative binomial distribution with $\mu ~ U(100, 1000) \times \textit{density}$, $k ~ U(\mu / 10, \mu / 4)$ and $pi$ from the parameters of the reference real dataset, to get the final count values.
-
-This count matrix was then filtered and normalised using the pipeline described below.
-
-\subsubsection{PROSSTT}
-
-PROSSTT is a recent data simulator \cite{papadopoulos_prossttprobabilisticsimulation_2018}, which simulates expression using linear mixtures of expression programs and random walks through the trajectory.
-We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets.
-However, due to frequent crashes of the tool, only 19 datasets created output and were thus used in the evaluation.
-
-Using the simulate\_lineage function, we simulated the lineage expression, with parameters $a \sim U(0.01, 0.1)$, $\textit{branch-tol}_{\textit{intra}} \sim U(0, 0.9)$ and $\textit{branch-tol}_{\textit{inter}} \sim U(0, 0.9)$.
-These parameter distributions were chosen very broad so as to make sure both easy and difficult datasets are simulated.
-After simulating base gene expression with simulate\_base\_gene\_exp, we used the sample\_density function to finally simulate expression values of a number of cells (the same as from the reference real dataset), with $\alpha \sim \textit{Lognormal}$ ($\mu = 0.3$ and $\sigma = 1.5$) and $\beta \sim \textit{Lognormal}$ ($\mu = 2$ and $\sigma = 1.5$).
-Each of these parameters were centered around the default values of PROSSTT, but with enough variability to ensure a varied set of datasets.
-
-This count matrix was then filtered and normalised using the pipeline described below.
-
-\subsubsection{Splatter}
-
-Splatter \cite{zappia_splattersimulationsinglecell_2017} simulates expression values by constructing non-linear paths between different states, each having a distinct expression profile.
-We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets, leading to a total of 50 datasets.
-
-We used the splatSimulatePaths function from Splatter to simulate datasets, with number of cells and genes equal to those in the reference real dataset, and with parameters  $\textit{nonlinearProb}$, $\textit{sigmaFac}$ and $\textit{skew}$ all sampled from $U(0, 1)$.
-
-\subsection{Dataset filtering and normalization}
-
-We used a standard single-cell RNA-seq preprocessing pipeline that applies parts of the scran and scater Bioconductor packages \cite{lun_stepbystepworkflowlowlevel_2016}.
-The advantages of this pipeline are that it works both with and without spike-ins, and it includes a harsh cell filtering that looks at abnormalities in library sizes, mitochondrial gene expression and the number of genes expressed using median absolute deviations (which we set to 3).
-We required that a gene was expressed in at least 5$\%$ of the cells and that it should have an average expression higher than 0.02.
-Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5$\%$ and a biological component higher than 0.5.
-As a final filter, we removed both all-zero genes and cells until convergence.
-
-\subsection{Benchmark metrics}
-
-The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{norel_selfassessmenttrapcan_2011}.
-Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer.
-We therefore defined several possible metrics for comparing trajectories, each investigating different layers.
-These are all discussed in Supplementary Note 1 along with examples and robustness analyses when appropriate.
-
-Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (Supplementary Note 1).
-Based on this analysis, we chose four metrics for the evaluation, each assessing a different aspect of the trajectory: (1) the HIM measures the topological similarity; (2) the F1branches compares the branch assignment; (3) the cordist assesses the similarity in pairwise cell–cell distances and thus the cellular positions; and (4) the wcorfeatures looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
-
-\subsubsection{The Hamming–Ipsen–Mikhailov metric}
-
-The HIM metric \cite{jurman_himglocalmetric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length).
-It is a linear combination of the normalized Hamming distance, which gives an indication of the differences in edge lengths, and the normalized Ipsen–Mikhailov distance, which assesses the similarity in degree distributions.
-The latter has a parameter $\gamma$, which was fixed at 0.1 to make the scores comparable between datasets.
-We illustrate the metric and discuss alternatives in Supplementary Note 1.
-
-\subsubsection{The F1 between branch assignments}
-
-To compare branch assignment, we used an F1 score, also used used for comparing biclustering methods \cite{saelens_comprehensiveevaluationmodule_2018}.
-To calculate this metric, we first calculated the similarity of all pairs of branches between the two trajectories using the Jaccard similarity.
-Next, we defined the ‘Recovery' (respectively ‘Relevance') as the average maximal similarity of all branches in the reference dataset (respectively prediction).
-The F1branches was then defined as the harmonic mean between Recovery and Relevance.
-We illustrate this metric further in Supplementary Note 1.
-
-\subsubsection{Correlation between geodesic distances}
-
-When the position of a cell is the same in both the reference and the prediction, its relative distances to all other cells in the trajectory should also be the same.
-This observation is the basis for the cordist metric.
-To calculate the cordist, we first sampled 100 waypoint cells in both the prediction and the reference dataset, using stratified sampling between the different milestones, edges and regions of delayed commitment, weighted by the number of cells in each collection.
-We then calculated the geodesic distances between the union of waypoint cells from both datasets and all other cells.
-The calculation of the geodesic distance depended on the location of the two cells within the trajectory, further discussed in Supplementary Note 1, and was weighted by the length of the edge in the milestone network.
-Finally, the cordist was defined as the Spearman rank correlation between the distances of both datasets.
-We illustrate the metric and assess the effect of the number of waypoint cells in Supplementary Note 1.
-
-\subsubsection{The correlation between important features}
-
-The wcorfeatures assesses whether the same differentially expressed features are found using the predicted trajectory as in the known trajectory.
-To calculate this metric, we used Random Forest regression (implemented in the R ranger package \cite{wright_rangerfastimplementation_2017}), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone.
-We then extracted feature importance values for each feature and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset to give more weight to large differences.
-As hyperparameters we set the number of trees to 10,000 and the number of features on which to split to 1$\%$ of all available features.
-We illustrate this metric and assess the effect of its hyperparameters in Supplementary Note 1.
-
-\subsubsection{Score aggregation}
-
-To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics. This aggregation strategy is explained in more detail in Supplementary Note 1.
-
-To ensure that easy and difficult datasets have equal influence on the final score, we first normalized the scores on each dataset across the different methods.
-We shifted and scaled the scores to $\sigma$ = 1 and $\mu$ = 0, and then applied the unit probability density function of a normal distribution on these values to get the scores back into the [0,1] range.
-
-Since there is a bias in dataset source and trajectory type (for example, there are many more linear datasets), we aggregated the scores per method and dataset in multiple steps.
-We first aggregated the datasets with the same dataset source and trajectory type using an arithmetic mean of their scores.
-Next, the scores were averaged over different dataset sources, using an arithmetic mean that was weighted based on how much the synthetic and silver scores correlated with the real gold scores.
-Finally, the scores were aggregated over the different trajectory types again using an arithmetic mean.
-
-Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.
-
-\subsection{Method execution}
-
-Each execution of a method on a dataset was performed in a separate task as part of a gridengine job.
-Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 at 2.40 GHz, and one R session was started for each task.
-During the execution of a method on a dataset, if the time limit (>1 h) or memory limit (16 GB) was exceeded, or an error was produced, a zero score was returned for that execution.
-
-\subsection{Complementarity}
-
-To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score was equal to or higher than 95$\%$ of the best overall score for that particular dataset.
-We then calculated for every method the weighted percentage of datasets that fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method.
-We iteratively added new methods until all methods were selected.
-For this analysis, we did not include any methods that require any strong prior information and only included methods that could detect the trajectory types present in at least one of the datasets.
-
-\subsection{Scalability}
-
-To assess the scalability of each method, we started from five real datasets, selected using the centers from a k-medoids as discussed before.
-We up- and downscaled these datasets between 10 and 100,000 cells and 10 and 100,000 features, while never going higher than 1,000,000 values in total.
-To generate new cells or features, we first generated a 10-nearest-neighbor graph of both the cells and features from the expression space.
-For every new cell or feature, we used a linear combination of one to three existing cells or features, where each cell or feature was given a weight sampled from a uniform distribution between 0 and 1.
-
-We ran each method on each dataset for maximally 1 h and gave each process 10 GB of memory.
-To determine the running time of each method, we started the timer right after data loading and the loading of any packages, and stopped the clock before postprocessing and saving of the output.
-Pre- and postprocessing steps specific to a method, such as dimensionality reduction and gene filtering, were included in the time.
-To estimate the maximal memory usage, we used the max\_vmem value from the qacct command provided by a gridengine cluster.
-We acknowledge, however, that these memory estimates are very noisy and the averages provided in this study are therefore only rough estimates.
-
-The relationship between the dimensions of a dataset and the running time or maximal memory usage was modeled using shape constrained additive models \cite{pya_shapeconstrainedadditive_2015}, with $log_{10}|\textit{cells}|$ and $log_{10}|\textit{features}|$ as predictor variables, and fitted this model using the scam function as implemented in the R scam package, with $log_{10}\textit{time}$ (or $log_{10}\textit{memory}$) as outcome.
-
-To classify the time complexity of each method with respect to the number of cells, we predicted the running time at 10,000 features with increasing number of cells from 100 to 100,000, with steps of 100.
-We trained a generalized linear model with the following function: $y \approxeq \log{x} + \sqrt{x} + x + x^2 + x^3$ with y as running time and x as the number of cells or features.
-The time complexity of a method was then classified using the weights w from this model:
-
-$$
-\begin{cases}
-\text{superquadratic} & \text{if } w_{x^3} > 0.25,\\
-\text{quadratic} & \text{if } w_{x^2} > 0.25,\\
-\text{linear} & \text{if } w_x > 0.25,\\
-\text{sublinear} & \text{if } w_{\text{log}(x)} > 0.25 \text{ or } w_{\text{sqrt}(x)} > 0.25,\\
-\text{case with highest weight} & \text{else.}
-\end{cases}
-$$
-
-This process was repeated for classifying the time complexity with respect to the number of features, and the memory complexity both with respect to the number of cells and features.
-
-\subsection{Stability}
-
-In the ideal case, a method should produce a similar trajectory, even when the input data is slightly different.
-However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R or the `random.seed` function in numpy).
-To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25$\%$ of the datasets accounting for 15$\%$ of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking.
-We subsampled each dataset 10 times with 95$\%$ of the original cells and 95$\%$ of the original features.
-We ran every method on each of the bootstraps, and assessed the stability by calculating the benchmarking scores between each pair of subsequent models (run $i$ is compared to run $i+1$).
-For the cordist and F1branches, we only used the intersection between the cells of two datasets, while the intersection of the features was used for the wcorfeatures.
-
-\subsection{Usability}
-
-We created a transparent scoring scheme to quantify the usability of each method based on several existing tool quality and programming guidelines in the literature and online (Table \ref{tab:scoresheet}).
-The main goal of this quality control is to stimulate the improvement of current methods, and the development of user- and developer-friendly new methods.
-The quality control assessed six categories, each looking at several aspects, which are further divided into individual items.
-The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used.
-The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication).
-The code assurance category is frequently overlooked, and checks for code testing, continuous integration \cite{beaulieu-jones_reproducibilitycomputationalworkflows_2017} and an active support system.
-The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation).
-The behavior category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output.
-Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of datasets in which the usefulness of the method was shown and the scope of method evaluation in the paper.
-
-Each quality aspect received a weight depending on how frequently it was found in several papers and online sources that discuss tool quality (Table \ref{tab:scoresheet}).
-This was to make sure that more important aspects, such as the open source availability of the method, outweighed other less important aspects, such as the availability of a graphical user interface.
-For each aspect, we also assigned a weight to the individual questions being investigated (Table \ref{tab:scoresheet}).
-For calculating the final score, we weighed each of the six categories equally.
-
-\subsection{Guidelines}
-
-For each set of outcomes in the guidelines figure, we selected one to four methods, by first filtering the methods on those that can detect all required trajectory types, and ordering the methods according to their average accuracy score on datasets containing these trajectory types (aggregated according to the scheme presented in the section Accuracy).
-
-We used the same approach for selecting the best set of methods in the guidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}), developed using the R shiny package.
-This app will also filter the methods, among other things, depending on the predicted running time and memory requirements, the prior information available and the preferred execution environment (using the dynmethods package or standalone).
-
-\subsection{Reporting Summary}
-
-Further information on research design is available in the Nature Research Reporting Summary, available at \href{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}
-
-\section{Supplementary Note 1: Metrics to compare two trajectories}
-
-A trajectory, as defined in our evaluation, is a model with multiple abstractions.
-The top abstraction is the topology which contains information about the paths each cell can take from their starting point.
-Deeper abstractions involve the mapping of each cell to a particular branch within this network, and the position (or ordering) of each cells within these branches.
-Internally, the topology is represented by the milestone network and regions of delayed commitment, the branch assignment and cellular positions are represented by the milestone percentages (Figure \ref{fig:snote1fig_1}).
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_1.pdf}
-	\caption{
-		\textbf{An example trajectory that will be used throughout this section.}
-		It contains contains four milestones (W to Z) and five cells (a to e).
-	}
-	\label{fig:snote1fig_1}
-\end{figure}
-
-Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models using only one metric.
-We therefore sought to use different comparison metrics, each serving a different purpose:
-
-\begin{itemize}
-	\item \textbf{Specific metrics} investigate one particular aspect of the trajectory.
-Such metrics make it possible to find particular weak points for methods, e.g.
-that a method is very good at ordering but does not frequently find the correct topology.
-Moreover, having multiple individual metrics allow personalised rankings of methods, for example for users which are primarily interested in using the method correct topology.
-	\item \textbf{Application metrics} focus on the quality of a downstream analysis using the trajectory.
-For example, it measures whether the trajectory can be used to find accurate differentially expressed genes.
-	\item \textbf{Overall metrics} should capture all the different abstractions, in other words such metrics measure whether the resulting trajectory has a good topology, that the cells belong to similar branches \textit{and} that they are ordered correctly.
-\end{itemize}
-
-Here, we first describe and illustrate several possible specific, application and overall metrics.
-Next, we test these metrics on several test cases, to make sure they robustly identify "wrong" trajectory predictions.
-
-All metrics described here were implemented within the \href{https://github.com/dynverse/dyneval}{dyneval R package (https://github.com/dynverse/dyneval)}.
-
-\subsection{Metric characterisation and testing}
-
-\subsubsection{\textit{isomorphic}, \textit{edgeflip} and \textit{HIM}: Edit distance between two trajectory topologies}
-
-We used three different scores to assess the similarity in the topology between two trajectories, irregardless of where the cells were positioned.
-
-For all three scores, we first simplified the topology of the trajectory to make both graph structures comparable:
-
-\begin{itemize}
-	\item As we are only interested in the main structure of the topology without start or end, the graph was made undirected.
-	\item All milestones with degree 2 were removed. For example in the topology A $\Rightarrow$ B $\Rightarrow$ C $\Rightarrow$ D, C $\Rightarrow$ D, the B milestone was removed
-	\item A linear topology was converted to A $\Rightarrow$ B $\Rightarrow$ C
-	\item A cyclical topology such as A $\Rightarrow$ B $\Rightarrow$ C $\Rightarrow$ D or A $\Rightarrow$ B $\Rightarrow$ A were all simplified to A $\Rightarrow$ B $\Rightarrow$ C $\Rightarrow$ A
-	\item Duplicated edges such as A $\Rightarrow$ B, A $\Rightarrow$ B were decoupled to A $\Rightarrow$ B, A $\Rightarrow$ C $\Rightarrow$ B
-\end{itemize}
-
-The \textit{isomorphic} score returns 1 if two graphs are isomorphic, and 0 if they were not.
-For this, we used the used the BLISS algorithm \cite{junttila_engineeringefficientcanonical_2007}, as implemented in the R *igraph* package.
-
-The \textit{edgeflip} score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks.
-This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution \cite{bahiense_maximumcommonedge_2012}.
-We implemented a branch and bound approach for this problem, using several heuristics to speed up the search:
-
-\begin{itemize}
-	\item First check all possible edge additions and removals corresponding to the number of different edges between the two graphs.
-	\item For each possible solution, first check whether: \begin{enumerate}
-		\item The maximal degree is the same
-		\item The minimal degree is the same
-		\item All degrees are the same after sorting
-	\end{enumerate}
-	\item Only then check if the two graphs are isomorphic as described earlier.
-	\item If no solution is found, check all possible solutions with two extra edge additions/removals.
-\end{itemize}
-
-The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_himglocalmetric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}).
-It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network.
-Conceptually, \textit{HIM} is a linear combination of:
-
-\begin{itemize}
-	\item The normalised Hamming distance \cite{dougherty_validationgeneregulatory_2011}, which calculates the distance between two graphs by matching individual edges in the adjacency matrix, but disregards overall structural similarity.
-	\item The normalised Ipsen-Mikhailov distance \cite{ipsen_evolutionaryreconstructionnetworks_2002}, which calculates the overall distance of two graphs based on matches between its degree and adjacency matrix, while disregarding local structural similarities.
-It requires a $\gamma$ parameter, which is usually estimated based on the number of nodes in the graph, but which we fixed at $0.1$ so as to make the score comparable across different graph sizes.
-\end{itemize}
-
-We compared the three scores on several common topologies (Figure \ref{fig:snote1fig_2}a).
-While conceptually very different, the \textit{edgeflip} and \textit{HIM} still produce similar scores (Figure \ref{fig:snote1fig_2}b).
-The \textit{HIM} tends to punish the detection of cycles, while the \textit{edgeflip} is more harsh for differences in the number of bifurcations (Figure \ref{fig:snote1fig_2}b).
-The main difference however is that the \textit{HIM} takes into account edge lengths when comparing two trajectories, as illustrated in (Figure \ref{fig:snote1fig_2}c).
-Short "extra" edges in the topology are less punished by the \textit{HIM} than by the \textit{edgeflip}.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_2.pdf}
-	\caption{
-		\textbf{Showcase of three metrics to evaluate topologies: \textit{isomorphic}, \textit{edgeflip} and \textit{HIM}}
-		\textbf{(a)} The used topologies.
-\textbf{(b)} The scores when comparing each pair of trajectory types.
-(c) Four datasets in which aan extra edge is added and made progressively longer.
-This shows how the HIM can take into account edge lengths.
-	}
-	\label{fig:snote1fig_2}
-\end{figure}
-
-To summarise, the different topology based scores are useful for different scenarios:
-
-\begin{itemize}
-	\item If the two trajectories should only be compared when the topology is exactly the same, the \textit{isomorphic} should be used.
-	\item If it is important that the topologies are similar, but not necessarily isomorphic, the \textit{edgeflip} is most appropriate.
-	\item If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the \textit{HIM} is most appropriate.
-\end{itemize}
-
-\subsubsection{$\textit{F1}_{\textit{branches}}$ and $\textit{F1}_{\textit{milestones}}$: Comparing how well the cells are clustered in the trajectory}
-
-Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone \textit{or} branch \ref{fig:snote1fig_3}.
-These clusters of cells can then be compared using one of the many external cluster evaluation measures \cite{saelens_comprehensiveevaluationmodule_2018}.
-When selecting a cluster evaluation metric, we had two main conditions:
-
-- Because we allow methods to filter cells in the trajectory, the metric should be able to handle "non-exhaustive assignment", where some cells are not assigned to any cluster.
-- The metric should give each cluster equal weight, so that rare cell stages are equally important as large stages.
-
-The $\textit{F1}$ score between the $\textit{Recovery}$ and $\textit{Relevance}$ is a metric which conforms to both these conditions.
-This metric will map two clustersets by using their shared members based on the $\textit{Jaccard}$ similarity.
-It then calculates the $\textit{Recovery}$ as the average maximal $\textit{Jaccard}$ for every cluster in the first set of clusters (in our case the reference trajectory).
-Conversely, the $\textit{Relevance}$ is calculated based on the average maximal similarity in the second set of clusters (in our case the prediction).
-Both the $\textit{Recovery}$ and $\textit{Relevance}$ are then given equal weight in a harmonic mean ($\textit{F1}$).
-Formally, if $C$ and $C'$ are two cell clusters:
-
-\begin{align*}
-\textit{Jaccard}(c, c') &= \frac{|c \cap c'|}{|c \cup c'|} \\
-\textit{Recovery} &= \frac{1}{|C|} \sum_{c \in C}{\max_{c' \in C'}{\textit{Jaccard(c, c')}}} \\
-\textit{Relevance} &= \frac{1}{|C'|} \sum_{c' \in C'}{\max_{c \in C}{\textit{Jaccard(c, c')}}} \\
-\textit{F1} &= \frac{2}{\frac{1}{\textit{Recovery}} + \frac{1}{\textit{Relevance}}}
-\end{align*}
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_3.pdf}
-	\caption{
-		\textbf{Mapping cells to their closest milestone or branch for the calculation of the {$\textit{F1}_{\textit{milestones}}$} and {$\textit{F1}_{\textit{branches}}$} .}
-		To calculate the {$\textit{F1}_{\textit{milestones}}$}, cells are mapped towards the nearest milestone, i.e.
-the milestone with the highest milestone percentage.
-For the {$\textit{F1}_{\textit{branches}}$}, the cells are mapped to the closest edge.
-	}
-	\label{fig:snote1fig_3}
-\end{figure}
-
-\subsubsection{$\textit{cor}_{\textit{dist}}$: Correlation between geodesic distances}
-
-When the position of a cell is the same in both the reference and the prediction, its \textit{relative} distances to all other cells in the trajectory should also be the same.
-This observation is the basis for the $\textit{cor}_{\textit{dist}}$ metric.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_4.pdf}
-	\caption{
-		\textbf{The calculation of geodesic distances on a small example trajectory.}
-		\textbf{a)} A toy example containing four milestones (W to Z) and five cells (a to e).
-\textbf{b)} The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model.
-\textbf{c)} The calculations made for calculating the pairwise geodesic distances.
-\textbf{d)} A heatmap representation of the pairwise geodesic distances.
-	}
-	\label{fig:snote1fig_4}
-\end{figure}
-
-The geodesic distance is the distance a cell has to go through the trajectory space to get from one position to another.
-The way this distance is calculated depends on how two cells are positioned, showcased by an example in Figure \ref{fig:snote1fig_4}:
-
-\begin{itemize}
-	\item \textbf{Both cells are on the same edge in the milestone network.} In this case, the geodesic distance is defined as the product of the difference in milestone percentages and the length of their shared edge.
-For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$.
-	\item \textbf{Cells reside on different edges in the milestone network.} First, the distance of the cell to all its nearby milestones is calculated, based on its percentage within the edge and the length of the edge.
-These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells.
-For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.
-\end{itemize}
-
-The geodesic distance can be easily extended towards cells within regions of delayed commitment.
-When both cells are part of the same region of delayed commitment, the geodesic distance was defined as the manhattan distances between the milestone percentages weighted by the lengths from the milestone network.
-For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1) = 1.9$.
-The distance between two cells where only one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the distance between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.
-
-Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets.
-For this reason, a set of waypoint cells are defined \textit{a priori}, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories (Figure \ref{fig:snote1fig_5}a).
-These cell waypoints are determined by viewing each milestone, edge and region of delayed commitment as a collection of cells.
-We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection.
-For calculating the $\textit{cor}_{\textit{dist}}$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.
-
-To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate $\textit{cor}_{\textit{dist}}$.
-To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the $\textit{cor}_{\textit{dist}}$ at different levels of both cell shuffling and number of cell waypoints (Figure \ref{fig:snote1fig_5}a).
-We found that using cell waypoints does not induce a systematic bias in the $\textit{cor}_{\textit{dist}}$, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_5.pdf}
-	\caption{
-		\textbf{Determination of cell waypoints} \textbf{a)} Illustration of the stratified cell sampling using an example dataset (top).
-Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom).
-\textbf{b)} Accuracy versus time to calculate $\textit{cor}_{\textit{dist}}$.
-Shown are distributions over 100 random waypoint samples.
-The upper whisker of the boxplot extends from the hinge (75$\%$ percentile) to the largest value, no further than 1.5$\times$ the IQR of the hinge.
-The lower whisker extends from the hinge (25$\%$ percentile) to the smallest value, at most 1.5$\times$ the IQR of the hinge.
-	}
-	\label{fig:snote1fig_5}
-\end{figure}
-
-Although the $\textit{cor}_{\textit{dist}}$'s main characteristic is that it looks at the positions of the cells, other features of the trajectory are also (partly) captured.
-To illustrate this, we used the geodesic distances themselves as input for dimensionality reduction (Figure \ref{fig:snote1fig_6}) with varying topologies.
-This reduced space captures the original trajectory structure quite well, including the overall topology and branch lengths.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_6.pdf}
-	\caption{
-		\textbf{Determination of cell waypoints.} 
-		We generated different toy trajectory datasets with varying topologies and calculated the geodesic distances between all cells within the trajectory.
-We then used these distances as input for classical multidimensional scaling.
-This shows that the geodesic distances do not only contain information regarding the cell's positions, but also information on the lengths and wiring of the topology.
-	}
-	\label{fig:snote1fig_6}
-\end{figure}
-
-\subsubsection{$\textit{NMSE}_{\textit{rf}}$ and $\textit{NMSE}_{\textit{lm}}$: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory}
-
-An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory.
-If the cells are at similar positions in the trajectory (relative to its nearby cells), the prediction error should be low.
-
-Specifically, we implemented two metrics which predict the milestone percentages from the reference by using the predicted milestone percentages as features (Figure \ref{fig:snote1fig_7}).
-We did this with two regression methods, linear regression ($\textit{lm}$, using the R lm function) and Random Forest ($\textit{rf}$, implemented in the \textit{ranger} package  \cite{wright_rangerfastimplementation_2017}).
-In both cases, the accuracy of the prediction was measured using the Mean Squared error ($\mathit{MSE}$), in the case of Random forest we used the out-of-bag mean-squared error.
-Next, we calculated $\mathit{MSE}_{worst}$ equal to the $\mathit{MSE}$ when predicting all milestone percentages as the average.
-We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$.
-We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the $\textit{NMSE}_{\textit{rf}}$ and $\textit{NMSE}_{\textit{lm}}$ scores.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_7.pdf}
-	\caption{
-		\textbf{The calculation of $\textit{NMSE}_{\textit{lm}}$ distances on a small example trajectory.} 
-		The milestone percentages of the reference are predicted based on the milestone percentages of the prediction, using regression models such as linear regression or random forests.
-The predicted trajectory is then scored by comparing the mean-squared error (MSE) of this regression model with the baseline MSE where the prediction is the average milestone percentage.
-	}
-	\label{fig:snote1fig_7}
-\end{figure}
-
-\subsubsection{$\textit{cor}_{\textit{features}}$ and $\textit{wcor}_{\textit{features}}$: The accuracy of dynamical differentially expressed features/genes.}
-
-Although most metrics described above already assess some aspects directly relevant to the user, such as whether the method is good at finding the right topology, these metrics do not assess the quality of downstream analyses and hypotheses which can be generated from these models.
-
-Perhaps the main advantage of studying cellular dynamic processes using single-cell -omics data is that the dynamics of gene expression can be studied for the whole transcriptome.
-This can be used to construct other models such as dynamic regulatory networks and gene expression modules.
-Such analyses rely on a "good-enough" cellular ordering, so that it can be used to identify dynamical differentially expressed genes.
-
-To calculate the $\textit{cor}_{\textit{features}}$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory.
-More specifically, we first calculated the geodesic distances for each cell to all milestones in the trajectory.
-Next, we trained a Random Forest regression model (implemented in the R \textit{ranger} package \cite{wright_rangerfastimplementation_2017}, \href{https://github.com/imbs-hl/ranger}{https://github.com/imbs-hl/ranger}) to predict these distances for each milestone, based on the expression of genes within each cell.
-We then extracted feature importances using the Mean Decrease in Impurity (importance = 'impurity' parameter of the ranger function), as illustrated in Figure \ref{fig:snote1fig_8}.
-The overall importance of a feature (gene) was then equal to the mean importance over all milestones.
-Finally, we compared the two rankings by calculating the Pearson correlation, with values between -1 and 0 clipped to 0.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_8.pdf}
-	\caption{
-		\textbf{An illustration of ranking features based on their importance in a trajectory.} 
-		(a) A MDS dimensionality reudction of a real dataset in which mouse embryonic fibroblasts (MEF) differentiate into Neurons and Myocytes.
-(b) The ranking of feature importances from high to low.
-The majority of features have a very low importance.
-(c) Some examples, which were also highlighted in b.
-Higher features in the ranking are clearly specific to certain parts of the trajectory, while features lower on the ranking have a more dispersed expression pattern.
-	}
-	\label{fig:snote1fig_8}
-\end{figure}
-
-Random forest regression has two main hyperparameters.
-The number of trees to be fitted (num\_tree parameter) was fixed to 10000 to provide accurate and stable estimates of the feature importance (Figure \ref{fig:snote1fig_9}.
-The number of features on which can be split (mtry parameter) was set to 1$\%$ of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated features, omnipresent in transcriptomics data, are not suppressed in the ranking.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_9.pdf}
-	\caption{
-		\textbf{Effect of the number of trees parameter on the accuracy and variability of the $\textit{cor}_{\textit{features}}$.} 
-		We used the dataset from Figure \ref{fig:snote1fig_8} and calculated the $\textit{cor}_{\textit{features}}$ after shuffling a percentage of cells.
-	}
-	\label{fig:snote1fig_9}
-\end{figure}
-
-For most datasets, only a limited number of features will be differentially expressed in the trajectory.
-For example, in the dataset used in Figure \ref{fig:snote1fig_9} only the top 10$\%$-20$\%$ show a clear pattern of differential expression.
-The correlation will weight each of these features equally, and will therefore give more weight to the bottom, irrelevant features.
-To prioritise the top differentially expressed features, we also implemented the $\textit{wcor}_{\textit{features}}$, which will weight the correlation using the feature importance scores in the reference so that the top features have relatively more impact on the score (Figure \ref{fig:snote1fig_10}).
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_10.pdf}
-	\caption{
-		\textbf{Effect of weighting the features based on their feature importance in the reference.} 
-		We used the same dataset as in Figure \ref{fig:snote1fig_8}, and calculated the $\textit{cor}_{\textit{features}}$ after shuffling a percentage of cells.
-	}
-	\label{fig:snote1fig_10}
-\end{figure}
-
-% TODO: check floatbarrier?
-%\FloatBarrier
-
-\subsection{Metric conformity}
-
-Although most metrics described in the previous section make sense intuitively, this does not necessarily mean that these metrics are robust and will generate reasonable results when used for benchmarking.
-This is because different methods and datasets will all lead to a varied set of trajectory models:
-
-\begin{itemize}
-	\item Real datasets have all cells grouped onto milestones
-	\item Some methods place all cells in a region of delayed commitment, others never generate a region of delayed commitment
-	\item Some methods always return a linear trajectory, even if a bifurcation is present in the data
-	\item Some methods filter cells
-\end{itemize}
-
-A good metric, especially a good overall metric, should work in all these circumstances.
-To test this, we designed a set of rules to which a good metric should conform, and assessed empirically whether a metric conforms to these rules.
-
-We generated a panel of toy datasets (using our \href{https://github.com/dynverse/dyntoy}{\textit{dyntoy}} package, \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}) with all possible combinations of:
-
-\begin{itemize}
-	\item \# cells: 10, 20, 50, 100, 200, 500
-	\item \# features: 200
-	\item topologies: linear, bifurcation, multifurcating, tree, cycle, connected graph and disconnected graph
-	\item Whether cells are placed on the milestones (as in real data) or on the edges/regions of delayed commitment between the milestones (as in synthetic data)
-\end{itemize}
-
-We then perturbed the trajectories in these datasets in certain ways, and tested whether the scores follow an expected pattern.
-An overview of the conformity of every metric is first given in Table \ref{tab:conformity_overview}.
-The individual rules and metric behaviour are discussed in the Supplementary Material that can be found at \href{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}.
-
-\begin{table}
-	\caption{Overview of whether a particular metric conforms to a particular rule} \label{tab:conformity_overview}
-
-	\centering\begingroup\fontsize{7}{9}\selectfont
-
-	\begin{tabular}{>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}>{\raggedright\arraybackslash}p{1em}}
-		\toprule
-		\rotatebox{0}{name} & \rotatebox{90}{$\textit{cor}_{\textit{dist}}$} & \rotatebox{90}{$\textit{NMSE}_{\textit{rf}}$} & \rotatebox{90}{$\textit{NMSE}_{\textit{lm}}$} & \rotatebox{90}{$\textit{edgeflip}$} & \rotatebox{90}{$\textit{HIM}$} & \rotatebox{90}{isomorphic} & \rotatebox{90}{$\textit{cor}_{\textit{features}}$} & \rotatebox{90}{$\textit{wcor}_{\textit{features}}$} & \rotatebox{90}{$\textit{F1}_{\textit{branches}}$} & \rotatebox{90}{$\textit{F1}_{\textit{milestones}}$} & \rotatebox{90}{$\textit{mean}_{\textit{geometric}}$}\\
-		\midrule
-		Same score on identity & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Local cell shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Edge shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Local and global cell shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Changing positions locally and/or globally & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-
-		Cell filtering & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Removing divergence regions & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Move cells to start milestone & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Move cells to closest milestone & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Length shuffling & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-
-		Cells into small subedges & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		New leaf edges & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		New connecting edges & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Changing topology and cell position & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Bifurcation merging & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-
-		Bifurcation merging and changing cell positions & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Bifurcation concatentation & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Cycle breaking & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Linear joining & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Linear splitting & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-
-		Change of topology & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{FB9A99}{\xmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		Cells on milestones vs edges & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}} & \multicolumn{1}{c}{\cellcolor[HTML]{B2DF8A}{\cmark}}\\
-		\bottomrule
-	\end{tabular}
-	\endgroup{}
-\end{table}
-
-% \FloatBarrier
-
-\subsection{Score aggregation}
-
-To rank the methods, we need to aggregate on two levels: across \textbf{datasets} and across specific/application metrics to calculate an \textbf{overall metric}.
-
-\subsubsection{Aggregating over datasets}
-
-When combining different datasets, it is important that the biases in the datasets does not influence the overall score.
-In our study, we define three such biases, although there are potentially many more:
-
-\begin{itemize}
-	\item \textbf{Difficulty of the datasets} Some datasets are more difficult than others.
-This can have various reasons, such as the complexity of the topology, the amount of biological and technical noise, or the dimensions of the data.
-It is important that a small increase in performance on a more difficult dataset has an equal impact on the final score as a large increase in performance on easier datasets.
-	\item \textbf{Dataset sources} It is much easier to generate synthetic datasets than real datasets, and this bias is reflected in our set of datasets.
-However, given their higher biological relevance, real datasets should be given at least equal importance than synthetic datasets.
-	\item \textbf{Trajectory types} There are many more linear and disconnected real datasets, and only a limited number of tree or graph datasets.
-This imbalance is there because historically most datasets have been linear datasets, and because it is easy to create disconnected datasets by combining different datasets.
-However, this imbalance in trajectory types does not necessarily reflect the general importance of that trajectory type.
-\end{itemize}
-
-We designed an aggregation scheme which tries to prevent these biases from influencing the ranking of the methods.
-
-The difficulty of a dataset can easily have an impact on how much weight the dataset gets in an overall ranking.
-We illustrate this with a simple example in Figure \ref{fig:snote1fig_11}.
-One method consistently performs well on both the easy and the difficult datasets.
-But because the differences are small in the difficult datasets, the mean would not give this method a high score.
-Meanwhile, a variable method which does not perform well on the difficult dataset gets the highest score, because it scored so high on the easier dataset.
-
-To avoid this bias, we normalise the scores of each dataset by first scaling and centering to $\mu = 0$ and $\sigma = 1$, and then moving the score values back to $[0, 1]$ by applying the unit normal density distribution function.
-This results in scores which are comparable across different datasets (Figure \ref{fig:snote1fig_11}).
-In contrast to other possible normalisation techniques, this will still retain some information on the relative difference between the scores, which would have been lost when using the ranks for normalisation.
-An example of this normalisation, which will also be used in the subsequent aggregation steps, can be seen in Figure \ref{fig:snote1fig_12}.
-
-\begin{figure}
-	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_11.pdf}
-	\caption{
-		\textbf{An illustration of how the difficulty of a dataset can influence the overall ranking.} 
-		A decent method, which consistently ranks high on an easy and difficult dataset, does not get a high score when averaging.
-On the other hand, a method which ranks high on the easy dataset, but very low on the difficult dataset does get a high score on average.
-After normalising the scores (right), this problem disappears.
-	}
-	\label{fig:snote1fig_11}
-\end{figure}
-
-\begin{figure}
-	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_12.pdf}
-	\caption{
-		\textbf{An example of the normalisation procedure.} 
-		Shown are some results of a benchmarking procedure, where every row contains the scores of a particular method (red shading) on a particular dataset (blue shading), with a trajectory type (green shading) and dataset source (orange shading).
-	}
-	\label{fig:snote1fig_12}
-\end{figure}
-
-After normalisation, we aggregate step by step the scores from different datasets.
-We first aggregate the datasets with the same dataset source and trajectory type using an arithmetic mean of their scores (Figure \ref{fig:snote1fig_13}a).
-Next, the scores are averaged over different dataset sources, using a arithmetic mean which was weighted based on how much the synthetic and silver scores correlated with the real gold scores (Figure \ref{fig:snote1fig_13}b).
-Finally, the scores are aggregated over the different trajectory types again using a arithmetic mean (Figure \ref{fig:snote1fig_13}c).
-
-\begin{figure}
-	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_13.pdf}
-	\caption{
-		\textbf{An example of the aggregation procedure.} 
-		In consecutive steps we aggregated across \textbf{(a)} different datasets with the same source and trajectory type, \textbf{(b)} different dataset sources with the same trajectory type (weighted for the correlation of the dataset source with the real gold dataset source) and \textbf{(c)} all trajectory types.
-	}
-	\label{fig:snote1fig_13}
-\end{figure}
-
-\subsubsection{Overall metrics}
-
-Undoubtedly, a single optimal overall metric does not exist for trajectories, as different users may have different priorities:
-
-\begin{itemize}
-	\item A user may be primarily interested in defining the correct topology, and only use the cellular ordering when the topology is correct
-	\item A user may be less interested in how the cells are ordered within a branch, but primarily in which cells are in which branches
-	\item A user may already know the topology, and may be primarily interested in finding good features related to a particular branching point
-	\item ...
-\end{itemize}
-
-Each of these scenarios would require a combinations of \textit{specific} and \textit{application} metrics with different weights.
-To provide an "overall" ranking of the metrics, which is impartial for the scenarios described above, we therefore chose a metric which weighs every aspect of the trajectory equally:
-
-\begin{itemize}
-	\item Its \textbf{ordering}, using the $\textit{cor}_{\textit{dist}}$
-	\item Its \textbf{branch assignment}, using the $\textit{F1}_{\textit{branches}}$
-	\item Its \textbf{topology}, using the $\textit{HIM}$
-	\item The accuracy of \textbf{differentially expressed features}, using the $\textit{wcor}_{\textit{features}}$
-\end{itemize}
-
-Next, we considered three different ways of averaging different scores: the arithmetic mean, geometric mean and harmonic mean.
-Each of these types of mean have different use cases.
-The harmonic mean is most appropriate when the scores would all have a common denominator (as is the case for the $\textit{Recovery}$ and $\textit{Relevance}$ described earlier).
-The arithmetic mean would be most appropriate when all the metrics have the same range.
-For our use case, the geometric mean is the most appropriate, because it is low if one of the values is low.
-For example, this means that if a method is not good at inferring the correct topology, it will get a low overall score, even if it performs better at all other scores.
-This ensures that a high score will only be reached if a prediction has a good ordering, branch assignment, topology, and set of differentially expressed features.
-
-The final overall score (Figure \ref{fig:snote1fig_14}) for a method was thus defined as:
-
-$$\textit{Overall} = \textit{mean}_{\textit{geometric}} = \sqrt[4]{\textit{cor}_{\textit{dist}} \times \textit{F1}_{\textit{branches}} \times \textit{HIM} \times \textit{wcor}_{\textit{features}}}$$
-
-\begin{figure}
-	\centering\includegraphics[width=0.8\linewidth]{fig/dynbenchmark/snote1fig_14.pdf}
-	\caption{
-		\textbf{An example of the averaging procedure.} 
-		For each method, we calculated the geometric mean between its normalised and aggregated scores
-	}
-	\label{fig:snote1fig_14}
-\end{figure}
-
-We do however want to stress that different use cases will require a different overall score to order the methods.
-Such a context-dependent ranking of all methods is provided through the dynguidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}).
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-10-01 ch1_introduction/content.tex
clean up thesis
update text

@@
 The cell is the smallest unit of life, of which all known living organisms are composed.
-Every cell houses a plethora of biomolecular processes that allow it to continuously adapt to changes in its environment.
+Every cell houses a plethora of biomolecular processes that allows it to continuously adapt to changes in its environment.
 Due to the dynamic nature of these processes, it can be very challenging to comprehend the cellular response to a signal.
@@
 Recent advances in experimental technologies are playing a crucial role in reductionist biology, allowing to measure the abundance of thousands of different biochemical molecules in tens of thousands of individual cells.
-With it comes the challenge of analysing a large amounts of data that are not easily interpretable by hand.
+With it comes the challenge of analysing large amounts of data that are not easily interpretable by hand.
 The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret.
-For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}.
+For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs\cite{hon_humancellatlas_2018}.
 Biologists thus turn to computer scientists
@@
 
-Observing the biomolecular insides of cells can ultimately provide fundamental insights into the processes that govern these cells and help uncover novel approaches for disease diagnosis, prognosis, and treatment.
-For example, the Human Cell Atlas (HCA) consortium \cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
+Observing the biomolecular insides of cells can ultimately provide fundamental understanding into the processes that govern these cells and help uncover novel approaches for disease diagnosis, prognosis, and treatment.
+For example, the Human Cell Atlas (HCA) consortium\cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
 Experts in the field often metaphorically describe the HCA initiative as aiming to develop a 'Google Maps' of the human body.
-Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs \cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.
+Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs\cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.
 
@@
 
-\subsection{The origins of life and the RNA world}
-The discovery of the double helix shape of deoxyribonucleic acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origins of life and evolution.
+\subsection{The origin of life and the RNA world}
+The discovery of the double helix shape of deoxyribonucleic acid (DNA)\cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origin of life and evolution.
 By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism.
@@
 
-Even so, a widely-accepted hypothesis states that life (or cells) did not originate from DNA, but instead was kicked off from its lesser-known cousin, ribonucleic acid (RNA).
-According to the RNA world hypothesis \cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}).
-Only later did cells develop the ability to use the more chemically stable DNA molecules to self-sustain in a process commonly referred to as the Central Dogma.
+Life (or cells) did not originate from DNA, however.
+A widely-accepted hypothesis states that life originates from its lesser-known cousin, ribonucleic acid (RNA).
+According to the RNA world hypothesis\cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and to perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}).
+Only later did cells develop the ability to use the more chemically stable DNA molecules to self-sustain in a process commonly referred to as the central dogma.
 
@@
 		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life.
-Image from Horning (2011) \cite{horning_rnaworld_2011}.
+Image from Horning (2011)\cite{horning_rnaworld_2011}.
 % TODO: expand the description
@@
 
-The central dogma describes the general flow of genetic information in almost all existent living cells: DNA is decoded to RNA, which in turn encodes proteins.
-Main processes involved in the central dogma are transcription, splicing and translation (Figure \ref{fig:centraldogma}).
+The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins.
+Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).
 
@@
 
-A stretch of DNA that contains a genetic blueprint for a particular molecule is called a gene, and the collection of an organism's genes is called its genome.
-\textbf{Transcription} is the process of synthesising an RNA molecule from a gene, and the resulting molecule is called a transcript.
-RNA is similar to DNA but differs in several ways; most notably all thymine nucleobases are replaced with uracil (U), and RNA molecules consist of only one strand.
-% HT: Add: which is able to exit the nucleus compartment.
-% RC: I haven't introduced the concept of a nucleus yet
-Due to its single-strandedness, RNA is less stable and will break down faster.
-Single-strandedness also allows some types of RNA (e.g.
-transfer RNA, ribosomal RNA) to form more complex three-dimensional structures by having certain regions bind to other regions of the strand.
-This work only considers messenger RNA (mRNA), which are transcribed from protein-coding genes, meaning that the mRNAs can result in the production of particular proteins.
-A protein-coding gene consists of alternating sections called introns and exons.
-Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
+During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA.
+The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA).
+This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons).
+A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms.
+This process of alternative splicing is essential to generate over a 100.000 different proteins starting from a rather small number of genes (20.000-25.000).
 
-% CVC: Work in progress
-%A gene is a stretch of DNA that contains a genetic blueprint for a RNA molecule. The gene serves as a template for making an RNA molecule in a process called transcription. 
-%The resulting RNA transcript can be protein-coding or non-coding.
-For this work we only consider the protein-coding genes, which consists of alternating sections called introns and exons.
-Exonic regions contain code for what a protein should be made up of, whereas the intronic regions can contain information on how to assemble the pieces.
-%
-%A protein-coding gene encodes protein-coding RNA molecules, the so-called messenger RNA (mRNA).
-%
-%This work only considers the protein-coding RNA molecules, the so-called messenger RNA (mRNA).
-%% TODO: mention that there are about 20'000 protein-coding genes in the human body.
-%
-%The DNA that serves as a template for making a protein-coding RNA molecule is called a gene, and the complete set of an organism's genes is called its genome.
-%
-%The protein-coding part of the DNA is called a gene.
-
-\textbf{RNA splicing} is a process that occurs in almost all organisms and results in the removal of intronic regions in the mRNA molecules, resulting in a mature mRNA. 
-Splicing allows to create multiple variants of the same product, which can affect the enzymatic properties or localisation of the resulting product \cite{kelemen_functionalternativesplicing_2013}.
-
-During \textbf{translation}, a chain of amino acids is synthesised from a mature mRNA transcript. 
-% HT: in ribosomes. These units can be seen as small factories in which amino acids are assembled following the pattern contained in an mRNA molecule.
-% RC: I can avoid talking about ribosomes
-Every three nucleobases are translated into one of 21 different amino acids.
-The resulting chain of amino acids is folded up into a protein, the structure of which is determined by the sequence of different amino acids in the chain.
-In turn, its structure determines the functionality of the protein, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
+ The mature mRNA is then transported to the cytoplasm, where it engages with ribosomes to initiate \textbf{translation}.
+During this highly evolutionary conserved process, a chain of amino acids, known as the protein building blocks, is being synthesised.
+Each amino acid is specified by three nucleotides (a codon) in the mRNA, according to a nearly universal genetic code.
+After being released by the ribosomes, the translation product undergoes a variety of chemical modifications to form the final folded protein, the structure of which is determined by the sequence of different amino acids in the chain.
+In addition, polypeptides may be cleaved to yield more than one active polypeptide product.
+The structure of a protein determines its functionality, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
+ % Ref: Matera AG, Wang Z. A day in the life of the spliceosome. Nature reviews Molecular cell biology. 2014 Feb;15(2):108. 
+ % Ref: Book: Human Molecular Genetics, Strachan and Read
 
 \subsection{Cell types}
-The functionality provided by a cell is defined (mostly) by the proteins of which it consists.
-One common approach to trying to understand the functionality of a cell is to observe which molecules are present in the cell and to associate those molecules with functionality.
+\textit{Homo sapiens} like to categorise everything they encounter, and so too have they conceptualised groups of cells called "cell types".
+The human body contains more than 200 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous.
+This however, is a major underestimation of the real number of cell types.
+Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types.
+% Ref: Strachan and Read
 
-\textit{Homo sapiens} like to categorise everything they encounter and so too have they conceptualised groups of cells called "cell types" according to their functionality.
-The concept of cell types eases reasoning about all aspects of biology, for instance, which cell types turn into (differentiate) or communicate with which other cell types, or how a cell type responds to a specific stimulation.
-Cells can be highly specialised toward performing a particular function (e.g.
+The concept of cell types eases reasoning and our understanding about many aspects of biology (e.g.
+the process of cell differentiation, cell-cell communication, cellular response to certain stimuli).
+Some cells are known to be highly specialised toward performing a particular function (e.g.
 memory B cells accelerate immune response by remembering previously encountered pathogens), or they can maintain a strong ability to differentiate into other cell types.
 
-\mycomment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
-Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task.
-In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.
-
-\mycomment{are dynamic processes defined? a dynamic process is just a biochemical reaction}
+One common approach for understanding the functionality of a particular cell is to observe which molecules are present in the cell and to associate those set of molecules with functionality.
+Taking a snap shot of the protein or RNA transcript content in a particular cell, might already provide us with major insights into its functionality.
+However, in order to fulfil a particular task, the biochemical machinery of the cell gradually changes over time.
+Therefore it is highly informative to also consider the transition states between cell types and the dynamic processes involved therein.
 
 \subsection{Cell dynamics and gene regulation}
-If cells are dynamic entities and can gradually produce the molecules needed to acquire new functionality, what is the process by which this happens?
-The mechanism by which this happens is called gene regulation.
+Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality.
+The variation within and between cells happens because of differences in gene expression.
+The mechanism by which this happens is called gene regulation, and takes place primarily at the level of transcription.
+According to the needs of a cell, different genes are being transcribed.
+Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages.
+%ref: Strachan and Read
+% CVC
+
 Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate).
@@
 Typically, one TF will regulate the transcription rate of many targets.
-
-\mycomment{TODO: download database and calculate some statistics?}
 
@@
 
-\mycomment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
 \subsection{Profiling single cells}
-\mycomment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
 In order to understand a biological process, it is often quite helpful to be able to profile (i.e.
@@
 
+\subsubsection{Single-cell}
 The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
 Since it was developed by Coons et al.
-in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins \cite{coons_immunologicalpropertiesantibody_1941}.
+in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins\cite{coons_immunologicalpropertiesantibody_1941}.
 In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system.
@@
 
-\mycomment{Rephrase the microscope sentence}
+\mycomment{Rephrase the microscope sentence (dino joke?)}
 
@@
 
-Flow cytometry \cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
+Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
 Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies.
@@
 
+\subsubsection{Omics}
 On the other side of the spectrum are the so-called "omics" technologies.
-"Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
+"Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
 There are many types of "omics", but the most commonly used are the following.
@@
 
+\subsubsection{Single-cell omics}
 Transformative technological advances in microvolume sequencing allowed Tang et al.
-to analyse the transcriptome at single-cell resolution \cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A).
+to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A).
 During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells (Figure \ref{fig:profiling_convergence}B) and measuring other levels of information such as proteomic expression levels (Figure \ref{fig:profiling_convergence}C).
@@
 \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
-Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}.
+Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}.
     B. 
-    C. scmultiomics \cite{moudgil_multimodalscrnaseq_2019}. 
+    C. scmultiomics\cite{moudgil_multimodalscrnaseq_2019}. 
     % TODO: rewrite the description
@@
   % TODO: change image
-  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016) \cite{wagner_revealingvectorscellular_2016}.
-    B. Zappia et al. (2018) \cite{zappia_exploringsinglecellrnaseq_2018}
+  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016)\cite{wagner_revealingvectorscellular_2016}.
+    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}
     % TODO: rewrite the description
@@
 
-There are many ways of classifying DR methods \cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
+There are many ways of classifying DR methods\cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
 Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible.
@@
 \subsection{Trajectory inference}
-Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
-Trajectory inference (TI) is a new category of computational tools used to offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017,cannoodt_computationalmethodstrajectory_2016}.
+Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation\cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
+Trajectory inference (TI) is a new category of computational tools used to offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017,cannoodt_computationalmethodstrajectory_2016}.
 
 Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
-This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}.
-Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}.
+This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them\cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}.
+Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place\cite{cannoodt_computationalmethodstrajectory_2016}.
 TI has two objectives: to reconstruct the topology of the dynamic process (e.g.
@@
 TI methods allow the identification of new subsets of cells, delineation of a differentiation tree, and characterisation of the main driver genes along a state transition (Figure \ref{fig:trajectory_inference}C).
-Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms\cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable\cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
 
@@
 This topic already received much attention with the advent of bulk omics (before single-cell omics).
-These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods \cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.
+These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods\cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.
 
@@
 
-The advent of single-cell omics has made scientists wonder whether now is the time to revisit network inference \cite{stegle_computationalanalyticalchallenges_2015}. 
+The advent of single-cell omics has made scientists wonder whether now is the time to revisit network inference\cite{stegle_computationalanalyticalchallenges_2015}. 
 One of the main advantages of single-cell omics is the ability to quantify the exact cellular state of thousands of cells per experiment.
-The heterogeneity between cells caused by naturally occurring biological randomness \cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs (see Figure \ref{fig:bulk_vs_singlecell}).
+The heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs (see Figure \ref{fig:bulk_vs_singlecell}).
 In this setting, heterogeneity in the cell population eases network inference, rather than mask condition-specific expression patterns in regulatory interactions.
@@
 Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication.
-Hundreds of new software tools were developed \cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g.
+Hundreds of new software tools were developed\cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g.
 differential expression, dimensionality reduction, normalisation).

### 2019-10-01 ch1_introduction/content.tex
update introduction

@@
 Image from Horning (2011)\cite{horning_rnaworld_2011}.
-% TODO: expand the description
+	  \fbox{\color{red} TODO: combine images, rewrite the description.}
 	}
@@
 
-The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins.
+The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins\cite{strachan_humanmoleculargenetics_2011}.
 Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).
@@
   \caption{
-    \textbf{Central Dogma.} % TODO: expand the description
+    \textbf{Central Dogma.}
+    \fbox{\color{red} TODO: combine images, rewrite the description.}
   }
@@
 A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms.
-This process of alternative splicing is essential to generate over a 100.000 different proteins starting from a rather small number of genes (20.000-25.000).
+This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes\cite{nilsen_expansioneukaryoticproteome_2010}.
 
@@
 The structure of a protein determines its functionality, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
- % Ref: Matera AG, Wang Z. A day in the life of the spliceosome. Nature reviews Molecular cell biology. 2014 Feb;15(2):108. 
- % Ref: Book: Human Molecular Genetics, Strachan and Read
 
@@
 This however, is a major underestimation of the real number of cell types.
-Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types.
-% Ref: Strachan and Read
+Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types\cite{strachan_humanmoleculargenetics_2011}.
 
@@
 According to the needs of a cell, different genes are being transcribed.
-Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages.
-%ref: Strachan and Read
+Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages\cite{strachan_humanmoleculargenetics_2011}.
+
 % CVC
@@
   C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
-  % TODO: change image
   \caption{
@@
 Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}.
-    B. 
+    B. Cell numbers vs. gene count over time.
     C. scmultiomics\cite{moudgil_multimodalscrnaseq_2019}. 
-    % TODO: rewrite the description
+    \fbox{\color{red} TODO: combine images, rewrite the description.}
   }
@@
 \section{Computational tools}
-The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
+The new types of analyses permitted by single-cell omics harbour exceptional opportunities to discover new aspects of biology and redefine existing knowledge (Figure \ref{fig:comp_tools}A).
+These include the following categories\cite{zappia_exploringsinglecellrnaseq_2018}.
+\begin{itemize}
+	\item \textbf{Data imputation}: \ldots.
+	\item \textbf{Data integration}: \ldots.
+	\item \textbf{Clustering}: \ldots.
+	\item \textbf{Dimensionality reduction}: providing a visual and informative overview of a given dataset.
+	\item \textbf{Trajectory inference}: identifying and characterising transitions between different cellular states.
+	\item \textbf{Trajectory alignment}: \ldots.
+	\item \textbf{Network inference}: inferring gene regulatory interactions between transcription factors across individual cells.
+\end{itemize}
+\mycomment{To do: write a section for each of these categories.}
 
-Some of these opportunities lie in efforts such as the Human Cell Atlas.
-The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories which connect the different cell types.
-As part of this endeavour, the consortium will perform single-cell omics on tens or even hundreds of millions of cells.
-
-Single-cell omics permits new types of analyse but also come with hitherto unseen data characteristics, the combination of which poses exciting new challenges for the computational community to tackle (Figure \ref{fig:comp_tools}A)\cite{stegle_computationalanalyticalchallenges_2015,yuan_challengesemergingdirections_2017,chen_singlecellrnaseqtechnologies_2019}.
-These challenges include:
-\begin{itemize}
-  \item normalisation: separating biological noise from technical noise,
-  \item dimensionality reduction: providing a visual and informative overview of a given dataset, 
-  \item trajectory inference: identifying and characterising transitions between different cellular states, and
-  \item gene regulatory network inference: inferring regulatory interactions between transcription factors across individual cells.
-\end{itemize}
+Developing new computational tools to perform these analyses has proven challenging for two main reasons \cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
+Firstly, single-cell omics data suffers from hitherto unseen noise characteristics, including high dropout rates and high levels of transcriptional stochasticity.
+In addition, a low abundance of single-cell omics data that can be used as a ground-truth obstruct quantitative evaluation of such methods.
 
@@
   \centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
-  % TODO: change image
   \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016)\cite{wagner_revealingvectorscellular_2016}.
-    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}
-    % TODO: rewrite the description
+    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}.
+    \fbox{\color{red} TODO: combine images, rewrite the description.}
   }
@@
 \end{figure}
-
-\mycomment{Make a better connection to the subsections below}
-
-\subsection{Normalisation}
-\mycomment{TODO}
 

### 2019-10-01 ch3_dynbenchmark/content.tex
update
clean up thesis
fix table

@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_1.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/supfigure_1.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_2.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_3.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/supfigure_3.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_4.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_4.pdf}
@@
 
-		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{puget_greendiceare_2016}\\*
+		Seed setting & The method does not artificially become deterministic (1), for example by setting some (0.5) or a lot (0) of seeds & \cite{puget_greendiceare_2016}\\*
 
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_5.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_6.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_1.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_2.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_3.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_4.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_5.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_6.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_7.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_8.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_9.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_10.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_11.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_12.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_13.pdf}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
 	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_14.pdf}
@@
 Such a context-dependent ranking of all methods is provided through the dynguidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}).
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-10-01 ch6_bred/scripts/1_genesets.R
working on ch6

@@ -1,0 +1,47 @@
+library(tidyverse)
+
+## CREATE GO tibble
+go.entrezs <- as.list(org.Mm.eg.db::org.Mm.egGO2ALLEGS)
+ontologies <- annotate::Ontology(names(go.entrezs))
+go.entrezs <- go.entrezs[ontologies %in% c("BP", "MF")]
+go.entrezs <- lapply(go.entrezs, function(go) na.omit(unique(go)))
+go.ids <- names(go.entrezs)
+go.terms <- annotate::Term(GO.db::GOTERM[go.ids])
+go <- tibble(database = "GO", id = go.ids, description = go.terms, entrezs = go.entrezs)
+
+## CREATE KEGG OBJECTS
+pws <- pbapply::pblapply(names(KEGGREST::keggList("pathway", "mmu")), function(n) KEGGREST::keggGet(n)[[1]])
+kegg.entrezs <- lapply(pws, function(pw) {
+  genes <- pw$GENE
+  if (length(genes) > 0) unique(genes[seq(1, length(genes)-1, by=2)])
+  else genes
+})
+kegg.ids <- map_chr(pws, ~ .$ENTRY)
+kegg.names <- map_chr(pws, ~ .$NAME)
+kegg <- tibble(
+  database = "KEGG",
+  id = kegg.ids,
+  description = kegg.names %>% gsub(" - [^-]*", "", .),
+  entrezs = kegg.entrezs
+)
+
+## CREATE REACTOME OBJECTS
+PATHID2NAME <- as.list(reactome.db::reactomePATHID2NAME)
+PATHID2NAME <- PATHID2NAME[grepl("Mus musculus", PATHID2NAME)]
+EXTID2PATHID <- as.list(reactome.db::reactomeEXTID2PATHID)
+reactome.entrezs <- pbapply::pblapply(names(PATHID2NAME), cl = 8, function(id) names(EXTID2PATHID)[sapply(EXTID2PATHID, function(x) id %in% x)])
+reactome <- tibble(
+  database = "reactome",
+  id = names(PATHID2NAME),
+  description = PATHID2NAME %>% gsub("^Mus musculus: ", "", .),
+  entrezs = reactome.entrezs
+)
+
+# genesets db
+genesets <- bind_rows(
+  go,
+  kegg,
+  reactome
+)
+
+write_rds(genesets, "derived_files/data_genesets.rds")

### 2019-10-01 ch6_bred/scripts/2_treutlein.R

@@ -1,0 +1,80 @@
+library(tidyverse)
+library(dyno)
+options(dynwrap_backend = 'r_wrapper')
+
+# download and preprocess dataset
+file <- "derived_files/GSE67310.txt.gz"
+url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE67310&format=file&file=GSE67310%5FiN%5Fdata%5Flog2FPKM%5Fannotated%2Etxt%2Egz"
+if (!file.exists(file)) {
+  download.file(url, file)
+}
+data <- read_tsv(file, col_types = cols(
+  cell_name = "c", assignment = "c", experiment = "c", time_point = "c", .default = "d"
+))
+
+cell_info <-
+  data %>%
+  select(cell_name:time_point) %>%
+  rename(cell_id = cell_name)
+grouping <-
+  cell_info %>%
+  select(cell_id, assignment) %>%
+  deframe()
+expression <-
+  data %>%
+  select(-cell_name:-time_point) %>%
+  as.matrix() %>%
+  magrittr::set_rownames(data$cell_name)
+
+alias2eg <- as.list(org.Mm.eg.db::org.Mm.egALIAS2EG) %>% unlist()
+feature_info <-
+  tibble(
+    feature_id = colnames(expression),
+    entrez = ifelse(feature_id %in% names(alias2eg), alias2eg[feature_id], NA_character_)
+  )
+
+counts <- floor(2^expression-1)
+dataset <-
+  wrap_expression(
+    id = "treutlein_2016",
+    counts = counts,
+    expression = expression,
+    cell_info = cell_info,
+    feature_info = feature_info
+  ) %>%
+  add_grouping(
+    grouping = grouping
+  )
+
+# run slingshot and postprocess trajectory
+model <- infer_trajectory(dataset, ti_slingshot())
+plot_dimred(model, label_milestones = TRUE, grouping = dataset$grouping)
+model2 <- model %>%
+  simplify_trajectory() %>%
+  label_milestones(c("3" = "MEF", "2" = "Induced", "5" = "Myocyte", "4" = "Neuron")) %>%
+  add_root(root_milestone_id = "3")
+model2$dimred_segment_points <-
+  model2$dimred_segment_progressions <-
+  NULL
+
+# calculate feature importance scores
+fimp <- dynfeature::calculate_overall_feature_importance(model2, dataset)
+
+# create visualisations
+g <- plot_dimred(model2, label_milestones = TRUE, grouping = dataset$grouping)
[…]

### 2019-10-01 ch6_bred/scripts/3_submit.R

@@ -1,0 +1,294 @@
+library(tidyverse)
+
+list2env(read_rds("derived_files/data.rds"), .GlobalEnv)
+
+expr <- as.matrix(dataset$expression)
+
+# calculate_target_importance <- function(
+#   target_ix,
+#   expr,
+#   samples,
+#   regulators,
+#   targets,
+#   num_trees = 10000,
+#   num_variables_per_split = 100,
+#   num_samples_per_tree = 250,
+#   min_node_size = 10,
+#   interaction_importance_filter = .01,
+#   sigmoid_mean = mean(expr[expr != 0]),
+#   sigmoid_sd = sd(expr[expr != 0]),
+#   num_perms = 10
+# ) {
+#
+#   target <- targets[[target_ix]]
+#
+#   regs <- setdiff(regulators, target)
+#
+#   target_expr <- scale(expr[,target])[,1]
+#
+#   # data <- Matrix::Matrix(expr[,union(target, regs)], sparse = TRUE)
+#   # data[,1] <- target_expr
+#   # colnames(data)[[1]] <- "PREDICT"
+#
+#   data <- data.frame(
+#     PREDICT = target_expr,
+#     expr[,regs],
+#     check.names = FALSE,
+#     stringsAsFactors = FALSE
+#   )
+#   # st1 <- system.time({
+#   rf2 <- ranger::ranger(
+#     data = data,
+#     dependent.variable.name = "PREDICT",
+#     verbose = TRUE,
+#     num.threads = 1,
+#     importance = "impurity",
+#     mtry = num_variables_per_split,
+#     num.trees = num_trees,
+#     min.node.size = min_node_size,
+#     sample.fraction = num_samples_per_tree / nrow(expr),
+#     keep.inbag = TRUE
+#   )
+#   # })
+#
+#   # st2 <- system.time({
+#   # rf <- randomForest::randomForest(
+#   #   expr[,regs],
+#   #   target_expr,
+#   #   mtry = num_variables_per_split,
+#   #   ntree = num_trees,
+#   #   sampsize = num_samples_per_tree,
+#   #   nodesize = min_node_size,
+#   #   importance = FALSE,
+#   #   localImp = FALSE,
+#   #   keep.inbag = TRUE
+#   # )
+#   # })
+#
+#   imp <-
+#     rf2$variable.importance %>%
+#     enframe("feature_id", "importance") %>%
+#     mutate(ix = row_number()) %>%
+#     # rf$importance %>%
[…]

### 2019-10-01 ch1_introduction/content.tex
update

@@
 
-Developing new computational tools to perform these analyses has proven challenging for two main reasons \cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
+Developing new computational tools to perform these analyses has proven challenging for two main reasons\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
 Firstly, single-cell omics data suffers from hitherto unseen noise characteristics, including high dropout rates and high levels of transcriptional stochasticity.
@@
 
-\subsection{Gene regulatory network inference}
+\subsection{Network inference}
 Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes.
@@
 We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
+  \item We developed dyno, a toolkit to easily infer, visualise and interpret single-cell trajectories using more than 50 different TI methods (Chapter~\ref{chap:dyno}).
+dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g.
+adding directionality or adding annotation).
   \item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
@@
   \item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
-  \item \textbf{Cannoodt R}, Saelens W, Saeys Y. bred: Inferring single cell regulatory networks. In preparation.
+  \item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. \fbox{\color{red}order authors?}
   \item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.

### 2019-10-05 ch6_bred/scripts/2_tabula-muris.R
update

@@ -1,0 +1,323 @@
+library(tidyverse)
+
+dest_dir <- "derived_files/tabula_muris/"
+dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
+
+data_rnaseq <- paste0(dest_dir, "data_rnaseq/")
+data_rds <- paste0(dest_dir, "data_rds/")
+
+# zip_file <- paste0(dest_dir, "data_rnaseq.zip")
+# download.file("https://ndownloader.figshare.com/articles/5968960/versions/3", zip_file)
+# unzip(zip_file, exdir = data_rnaseq)
+# unzip(paste0(data_rnaseq, "droplet.zip"), exdir = data_rnaseq)
+
+# cell_info <- read_csv(
+#   paste0(data_rnaseq, "annotations_droplet.csv"),
+#   col_types = cols(
+#     .default = "c",
+#     tissue_tSNE_1 = "d",
+#     tissue_tSNE_2 = "d",
+#     subsetA = "l",
+#     subsetB = "l",
+#     subsetC = "l",
+#     subsetD = "l"
+#   )
+# )
+#
+# write_rds(cell_info, paste0(data_rds, "cell_info.rds"), compress = "gz")
+#
+# droplet <- paste0(data_rnaseq, "/droplet")
+# samples <- list.files(droplet, full.names = TRUE)
+# counts <- pbapply::pblapply(samples, function(folder_10x) {
+#   channel <- basename(folder_10x) %>% gsub(".*-", "", .)
+#
+#   counts <- Seurat::Read10X(folder_10x) %>% Matrix::t()
+#   rownames(counts) <- paste0(channel, "_", rownames(counts))
+#
+#   counts <- counts[rownames(counts) %in% cell_info$cell, ]
+#
+#   counts
+# }) %>% do.call(rbind, .)
+# counts <- counts[cell_info$cell, ] # reorder cells
+#
+# write_rds(counts, paste0(data_rds, "counts.rds"), compress = "gz")
+#
+# tpm <- counts / Matrix::rowSums(counts) * 1000000
+# # write_rds(tpm, paste0(data_rds, "tpm.rds"), compress = "gz")
+# write_rds(tpm, paste0(data_rds, "tpm.rds"))
+#
+# expression <- tpm
+# expression@x <- log2(expression@x + 1)
+# # write_rds(expression, paste0(data_rds, "expression.rds"), compress = "gz")
+# write_rds(expression, paste0(data_rds, "expression.rds"))
+
+
[…]

### 2019-10-05 ch6_bred/scripts/2_treutlein.R

@@ -3,7 +3,10 @@ library(dyno)
 options(dynwrap_backend = 'r_wrapper')
 
 # download and preprocess dataset
-file <- "derived_files/GSE67310.txt.gz"
+dest_dir <- "derived_files/GSE107727/"
+dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
+
+file <- paste0(dest_dir, "data.tar")
 url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE67310&format=file&file=GSE67310%5FiN%5Fdata%5Flog2FPKM%5Fannotated%2Etxt%2Egz"
 if (!file.exists(file)) {
   download.file(url, file)

### 2019-10-05 ch6_bred/scripts/3_submit.R

@@ -247,9 +247,9 @@ pheatmap::pheatmap(
 grouping <- model$progressions %>%
   mutate(group = case_when(
     from == "3" & percentage < .33 ~ "MEF",
-    from == "3" & percentage < .66 ~ "early-intermediate",
-    from == "3" ~ "intermediate",
-    percentage < .33 ~ "intermediate",
+    from == "3" & percentage < .66 ~ "early-induced",
+    from == "3" ~ "induced",
+    percentage < .33 ~ "induced",
     to == "4" & percentage < .66 ~ "early-neuron",
     to == "5" & percentage < .66 ~ "early-myocyte",
     to == "4" ~ "neuron",

### 2019-10-05 ch6_bred/scripts/find_mart.R

@@ -1,0 +1,28 @@
+
+archs <- listEnsemblArchives()
+gene_symbols <- pbapply::pblapply(seq_len(nrow(archs)), function(i) {
+  tryCatch({
+    ensembl <- useMart("ENSEMBL_MART_ENSEMBL", dataset="mmusculus_gene_ensembl", host=archs$url[[i]])
+    attrs <- listAttributes(ensembl)
+    entrez_attr_name <- ifelse("entrezgene" %in% attrs$name, "entrezgene", "entrezgene_id")
+    gene_symbols <- getBM(
+      attributes = c("ensembl_gene_id", "mgi_symbol", entrez_attr_name),
+      filters = "ensembl_gene_id",
+      values = colnames(counts),
+      mart = ensembl
+    )
+    gene_symbols %>%
+      rename(entrez_gene_id = !!entrez_attr_name) %>%
+      as_tibble()
+  }, error = function(e) {
+    NULL
+  })
+})
+
+
+archs %>%
+  mutate(
+    matching_genes = map_int(gene_symbols, function(x) if(is.data.frame(x)) length(unique(x$ensembl_gene_id)) else 0L),
+    num_genes = ncol(counts)
+  ) %>%
+  filter(num_genes == matching_genes)

### 2019-10-05 ch6_bred/scripts/2_dahlin.R
commit

@@ -1,0 +1,211 @@
+library(tidyverse)
+library(dyno)
+
+if (!file.exists("derived_files/dahlin.rds")) {
+  # download and preprocess dataset
+  dest_dir <- "derived_files/GSE107727/"
+  dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
+
+  # COUNTS ------------------------------------------------------------------
+
+  # file <- paste0(dest_dir, "data.tar")
+  # url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE107727&format=file"
+  # if (!file.exists(file)) {
+  #   download.file(url, file)
+  # }
+  # if (!file.exists(paste0(dest_dir, "GSM2877127_SIGAB1_counts.txt.gz"))) {
+  #   untar(file, exdir = dest_dir)
+  # }
+  #
+  # count_files <- list.files(dest_dir, pattern = "counts\\.txt\\.gz", full.names = TRUE)
+  # counts <-
+  #   pbapply::pblapply(count_files, function(file) {
+  #     gsm <- gsub(".*(GSM\\d*)_.*", "\\1", file)
+  #     mat <-
+  #       read.csv(file, header = TRUE, sep = "\t") %>%
+  #       as.matrix %>%
+  #       Matrix::Matrix(sparse = TRUE) %>%
+  #       Matrix::t()
+  #     rownames(mat) <- paste0(gsm, "_", rownames(mat))
+  #     mat
+  #   }) %>% do.call(rbind, .)
+  # write_rds(counts, paste0(dest_dir, "counts.rds"), compress = "gz")
+  counts <- read_rds(paste0(dest_dir, "counts.rds"))
+
+  num_reads_per_gene <- Matrix::colSums(counts)
+  qplot(log10(num_reads_per_gene + 1)) + geom_vline(xintercept = log10(10 + 1))
+  num_reads_per_cell <- Matrix::rowSums(counts)
+  qplot(log10(num_reads_per_cell + 1)) + geom_vline(xintercept = log10(2000 + 1))
+  num_genes_expressed <- Matrix::rowSums(counts > 0)
+  qplot(log10(num_genes_expressed + 1)) + geom_vline(xintercept = log10(2000 + 1))
+
+  counts <- counts[
+    num_reads_per_cell > 2000 & num_genes_expressed > 2000,
+    num_reads_per_gene > 10
+  ]
+
+
[…]

### 2019-10-05 ch6_bred/scripts/2_dahlin2.R

@@ -1,0 +1,406 @@
+library(tidyverse)
+library(dyno)
+
+if (!file.exists("derived_files/dahlin2.rds")) {
+  # download and preprocess dataset
+  dest_dir <- "derived_files/GSE107727/"
+  dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
+
+  # COUNTS ------------------------------------------------------------------
+
+  # file <- paste0(dest_dir, "data.tar")
+  # url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE107727&format=file"
+  # if (!file.exists(file)) {
+  #   download.file(url, file)
+  # }
+  # if (!file.exists(paste0(dest_dir, "GSM2877127_SIGAB1_counts.txt.gz"))) {
+  #   untar(file, exdir = dest_dir)
+  # }
+  #
+  # count_files <- list.files(dest_dir, pattern = "counts\\.txt\\.gz", full.names = TRUE)
+  # counts <-
+  #   pbapply::pblapply(count_files, function(file) {
+  #     gsm <- gsub(".*(GSM\\d*)_.*", "\\1", file)
+  #     mat <-
+  #       read.csv(file, header = TRUE, sep = "\t") %>%
+  #       as.matrix %>%
+  #       Matrix::Matrix(sparse = TRUE) %>%
+  #       Matrix::t()
+  #     rownames(mat) <- paste0(gsm, "_", rownames(mat))
+  #     mat
+  #   }) %>% do.call(rbind, .)
+  # write_rds(counts, paste0(dest_dir, "counts.rds"), compress = "gz")
+  counts <- read_rds(paste0(dest_dir, "counts.rds"))
+
+  # GENE INFO ---------------------------------------------------------------
+  ensembl <- biomaRt::useMart(
+    "ENSEMBL_MART_ENSEMBL",
+    dataset = "mmusculus_gene_ensembl",
+    host = "http://mar2016.archive.ensembl.org"
+  )
+  gene_symbols <- biomaRt::getBM(
+    attributes = c("ensembl_gene_id", "mgi_symbol", "entrezgene"),
+    filters = "ensembl_gene_id",
+    values = colnames(counts),
+    mart = ensembl
+  ) %>%
+    as_tibble()
+
+  feature_info <-
+    gene_symbols %>%
+    group_by(ensembl_gene_id) %>%
+    summarise(
+      symbol = mgi_symbol[[1]],
[…]

### 2019-10-05 ch6_bred/scripts/2_dahlin3.R

@@ -1,0 +1,7 @@
+library(tidyverse)
+
+library(hdf5r)
+fil <- H5File$new("derived_files/scanpy_processed_matrix_both_no_cc.h5ad")
+
+hdf5r:::print_listing(fil, 20)
+

### 2019-10-05 ch6_bred/scripts/2_mouse-cell-atlas.R

@@ -1,0 +1,61 @@
+library(tidyverse)
+library(rlang)
+
+# download and preprocess dataset
+dest_dir <- "derived_files/mouse-cell-atlas/"
+dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
+
+# cell info
+cell_assignment_file <- paste0(dest_dir, "cellassignment.csv")
+if (!file.exists(cell_assignment_file)) download.file("https://ndownloader.figshare.com/files/11083451?private_link=865e694ad06d5857db4b", cell_assignment_file)
+all_cell_info <-
+  read_csv(cell_assignment_file) %>%
+  rename(cell_id = Cell.name, cluster_id = ClusterID, tissue = Tissue, batch = Batch, barcode = Cell.Barcode, group_id = Annotation) %>%
+  mutate(
+    cell_type = group_id %>% str_extract("^[^\\(_]*"),
+    subtype = group_id %>% str_extract("_([^\\()]*)\\(") %>% str_replace_all("[_\\(]", "")
+  )
+all_cell_info %>% select(cell_type) %>% unique %>% arrange(cell_type) %>% write_tsv(paste0(dest_dir, "/clean_cell_types.tsv"))
+all_cell_info %>% select(cell_type, tissue) %>% unique %>% arrange(cell_type) %>% write_tsv(paste0(dest_dir, "/clean_cell_types2.tsv"))
+# counts
+counts_dir <- paste0(dest_dir, "counts/")
+counts_files <- list.files(counts_dir, pattern = ".txt.gz", full.names = TRUE, recursive = TRUE)
+
+if (length(counts_files) == 0) {
+  counts_zip_file <- paste0(dest_dir, "count_batchremove.zip")
+  download.file("https://ndownloader.figshare.com/files/10756795?private_link=865e694ad06d5857db4b", counts_zip_file)
+
+  dir.create(counts_dir, showWarnings = FALSE, recursive = FALSE)
+  unzip(counts_zip_file, exdir = counts_dir)
+  counts_files <- list.files(counts_dir, pattern = ".txt.gz", full.names = TRUE, recursive = TRUE)
+}
+
+counts <- pbapply::pblapply(counts_files, function(file) {
+  rds <- gsub("\\.txt\\.gz$", ".rds", file)
+  if (!file.exists(rds)) {
+    mat <- read.csv(file, sep = " ") %>%
[…]

### 2019-10-05 ch6_bred/scripts/2_dahlin.R
remove

@@ -1,211 +1,0 @@
-library(tidyverse)
-library(dyno)
-
-if (!file.exists("derived_files/dahlin.rds")) {
-  # download and preprocess dataset
-  dest_dir <- "derived_files/GSE107727/"
-  dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
-
-  # COUNTS ------------------------------------------------------------------
-
-  # file <- paste0(dest_dir, "data.tar")
-  # url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE107727&format=file"
-  # if (!file.exists(file)) {
-  #   download.file(url, file)
-  # }
-  # if (!file.exists(paste0(dest_dir, "GSM2877127_SIGAB1_counts.txt.gz"))) {
-  #   untar(file, exdir = dest_dir)
-  # }
-  #
-  # count_files <- list.files(dest_dir, pattern = "counts\\.txt\\.gz", full.names = TRUE)
-  # counts <-
-  #   pbapply::pblapply(count_files, function(file) {
-  #     gsm <- gsub(".*(GSM\\d*)_.*", "\\1", file)
-  #     mat <-
-  #       read.csv(file, header = TRUE, sep = "\t") %>%
-  #       as.matrix %>%
-  #       Matrix::Matrix(sparse = TRUE) %>%
-  #       Matrix::t()
-  #     rownames(mat) <- paste0(gsm, "_", rownames(mat))
-  #     mat
-  #   }) %>% do.call(rbind, .)
-  # write_rds(counts, paste0(dest_dir, "counts.rds"), compress = "gz")
-  counts <- read_rds(paste0(dest_dir, "counts.rds"))
-
-  num_reads_per_gene <- Matrix::colSums(counts)
-  qplot(log10(num_reads_per_gene + 1)) + geom_vline(xintercept = log10(10 + 1))
-  num_reads_per_cell <- Matrix::rowSums(counts)
-  qplot(log10(num_reads_per_cell + 1)) + geom_vline(xintercept = log10(2000 + 1))
-  num_genes_expressed <- Matrix::rowSums(counts > 0)
-  qplot(log10(num_genes_expressed + 1)) + geom_vline(xintercept = log10(2000 + 1))
-
-  counts <- counts[
-    num_reads_per_cell > 2000 & num_genes_expressed > 2000,
-    num_reads_per_gene > 10
-  ]
-
-
[…]

### 2019-10-05 ch6_bred/scripts/2_dahlin2.R

@@ -1,406 +1,0 @@
-library(tidyverse)
-library(dyno)
-
-if (!file.exists("derived_files/dahlin2.rds")) {
-  # download and preprocess dataset
-  dest_dir <- "derived_files/GSE107727/"
-  dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
-
-  # COUNTS ------------------------------------------------------------------
-
-  # file <- paste0(dest_dir, "data.tar")
-  # url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE107727&format=file"
-  # if (!file.exists(file)) {
-  #   download.file(url, file)
-  # }
-  # if (!file.exists(paste0(dest_dir, "GSM2877127_SIGAB1_counts.txt.gz"))) {
-  #   untar(file, exdir = dest_dir)
-  # }
-  #
-  # count_files <- list.files(dest_dir, pattern = "counts\\.txt\\.gz", full.names = TRUE)
-  # counts <-
-  #   pbapply::pblapply(count_files, function(file) {
-  #     gsm <- gsub(".*(GSM\\d*)_.*", "\\1", file)
-  #     mat <-
-  #       read.csv(file, header = TRUE, sep = "\t") %>%
-  #       as.matrix %>%
-  #       Matrix::Matrix(sparse = TRUE) %>%
-  #       Matrix::t()
-  #     rownames(mat) <- paste0(gsm, "_", rownames(mat))
-  #     mat
-  #   }) %>% do.call(rbind, .)
-  # write_rds(counts, paste0(dest_dir, "counts.rds"), compress = "gz")
-  counts <- read_rds(paste0(dest_dir, "counts.rds"))
-
-  # GENE INFO ---------------------------------------------------------------
-  ensembl <- biomaRt::useMart(
-    "ENSEMBL_MART_ENSEMBL",
-    dataset = "mmusculus_gene_ensembl",
-    host = "http://mar2016.archive.ensembl.org"
-  )
-  gene_symbols <- biomaRt::getBM(
-    attributes = c("ensembl_gene_id", "mgi_symbol", "entrezgene"),
-    filters = "ensembl_gene_id",
-    values = colnames(counts),
-    mart = ensembl
-  ) %>%
-    as_tibble()
-
-  feature_info <-
-    gene_symbols %>%
-    group_by(ensembl_gene_id) %>%
-    summarise(
-      symbol = mgi_symbol[[1]],
[…]

### 2019-10-05 ch6_bred/scripts/2_dahlin3.R

@@ -1,7 +1,0 @@
-library(tidyverse)
-
-library(hdf5r)
-fil <- H5File$new("derived_files/scanpy_processed_matrix_both_no_cc.h5ad")
-
-hdf5r:::print_listing(fil, 20)
-

### 2019-10-05 ch6_bred/scripts/2_mouse-cell-atlas.R

@@ -1,61 +1,0 @@
-library(tidyverse)
-library(rlang)
-
-# download and preprocess dataset
-dest_dir <- "derived_files/mouse-cell-atlas/"
-dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
-
-# cell info
-cell_assignment_file <- paste0(dest_dir, "cellassignment.csv")
-if (!file.exists(cell_assignment_file)) download.file("https://ndownloader.figshare.com/files/11083451?private_link=865e694ad06d5857db4b", cell_assignment_file)
-all_cell_info <-
-  read_csv(cell_assignment_file) %>%
-  rename(cell_id = Cell.name, cluster_id = ClusterID, tissue = Tissue, batch = Batch, barcode = Cell.Barcode, group_id = Annotation) %>%
-  mutate(
-    cell_type = group_id %>% str_extract("^[^\\(_]*"),
-    subtype = group_id %>% str_extract("_([^\\()]*)\\(") %>% str_replace_all("[_\\(]", "")
-  )
-all_cell_info %>% select(cell_type) %>% unique %>% arrange(cell_type) %>% write_tsv(paste0(dest_dir, "/clean_cell_types.tsv"))
-all_cell_info %>% select(cell_type, tissue) %>% unique %>% arrange(cell_type) %>% write_tsv(paste0(dest_dir, "/clean_cell_types2.tsv"))
-# counts
-counts_dir <- paste0(dest_dir, "counts/")
-counts_files <- list.files(counts_dir, pattern = ".txt.gz", full.names = TRUE, recursive = TRUE)
-
-if (length(counts_files) == 0) {
-  counts_zip_file <- paste0(dest_dir, "count_batchremove.zip")
-  download.file("https://ndownloader.figshare.com/files/10756795?private_link=865e694ad06d5857db4b", counts_zip_file)
-
-  dir.create(counts_dir, showWarnings = FALSE, recursive = FALSE)
-  unzip(counts_zip_file, exdir = counts_dir)
-  counts_files <- list.files(counts_dir, pattern = ".txt.gz", full.names = TRUE, recursive = TRUE)
-}
-
-counts <- pbapply::pblapply(counts_files, function(file) {
-  rds <- gsub("\\.txt\\.gz$", ".rds", file)
-  if (!file.exists(rds)) {
-    mat <- read.csv(file, sep = " ") %>%
[…]

### 2019-10-06 ch6_bred/scripts/2_tabula-muris.R
wip tabula muris

@@ -3,6 +3,9 @@ library(tidyverse)
 dest_dir <- "derived_files/tabula_muris/"
 dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
 
+cellontf <- paste0(dest_dir, "cl.obo")
+if (!file.exists(cellontf)) download.file("https://raw.githubusercontent.com/obophenotype/cell-ontology/master/cl.obo", cellontf)
+
 data_rnaseq <- paste0(dest_dir, "data_rnaseq/")
 data_rds <- paste0(dest_dir, "data_rds/")
 
@@ -97,227 +100,320 @@ if (!file.exists(paste0(data_rds, "data_filt.rds"))) {
 
 
 # SUBMIT BRED -------------------------------------------------------------
-data_rds <- "derived_files/tabula_muris/data_rds/"
+dest_dir <- "derived_files/tabula_muris/"
+data_rds <- paste0(dest_dir, "data_rds/")
 # list2env(read_rds(paste0(data_rds, "data.rds")), .GlobalEnv)
 list2env(read_rds(paste0(data_rds, "data_filt.rds")), .GlobalEnv)
 
+if (!file.exists(paste0(dest_dir, "qsub_handle.rds"))) {
+  calculate_target_importance <- function(
+    target_ix,
+    expr,
+    samples,
+    regulators,
+    targets,
+    num_trees = 10000,
+    num_variables_per_split = 100,
+    num_samples_per_tree = 250,
+    min_node_size = 10,
+    max_depth = NULL,
+    interaction_importance_filter = .01,
+    sigmoid_mean = mean(expr[expr != 0]),
+    sigmoid_sd = sd(expr[expr != 0])
+  ) {
 
-calculate_target_importance <- function(
-  target_ix,
-  expr,
-  samples,
-  regulators,
-  targets,
-  num_trees = 10000,
-  num_variables_per_split = 100,
-  num_samples_per_tree = 250,
-  min_node_size = 10,
-  max_depth = NULL,
-  interaction_importance_filter = .01,
-  sigmoid_mean = mean(expr[expr != 0]),
-  sigmoid_sd = sd(expr[expr != 0])
-) {
+    target <- targets[[target_ix]]
 
-  target <- targets[[target_ix]]
+    regs <- setdiff(regulators, target)
 
-  regs <- setdiff(regulators, target)
+    target_expr <- scale(expr[,target])[,1]
 
[…]

### 2019-10-07 ch6_bred/scripts/1_cellontology.R
update scripts

@@ -1,0 +1,37 @@
+library(tidyverse)
+library(ontologyIndex)
+
+
+cellontf <- "derived_files/cl.obo"
+
+if (!file.exists(cellontf)) download.file("https://raw.githubusercontent.com/obophenotype/cell-ontology/master/cl.obo", cellontf)
+
+cell_ont <-
+  get_ontology(cellontf) %>%
+  with(tibble(id, name, parents, children, ancestors, obsolete)) %>%
+  filter(!obsolete, grepl("CL:", id))
+
+gr <- cell_ont %>% select(from = id, to = children) %>% unnest(to) %>%
+  igraph::graph_from_data_frame()
+
+dis <- igraph::distances(gr, v = "CL:0000000")
+depth <- dis[1,] %>% enframe("id", "depth")
+
+cell_ont <- cell_ont %>% left_join(depth, by = "id")
+
+cell_ont_up <-
+  cell_ont %>%
+  select(cell_ontology_id = id, upstream = ancestors) %>%
+  unnest(upstream) %>%
+  filter(cell_ontology_id != upstream)
+
+# cids <- unique(sample_info$cell_ontology_id)
+cids <- c(
+  "CL:0000763", "CL:0000583", "CL:0000236", "CL:0000623", "CL:0000084", "CL:0000875", "CL:0000738", "CL:0000860", "CL:0000097", "CL:0000576", "CL:0002191", "CL:0000559", "CL:0000094", "CL:0000765",
+  "CL:0008001", "CL:0000547", "CL:0002048", "CL:0000767", "CL:0000235", "CL:0002046", "CL:0000816", "CL:0002045", "CL:0000453", "CL:0000451", "CL:0000081", "CL:0002420", "CL:0000894"
+)
+ciddis <- igraph::distances(gr, v = cids, to = cids)
+ciddr <- stats::cmdscale(as.dist(ciddis), k = 10)
+
+out <- "derived_files/cell_ontology.rds"
+write_rds(lst(cell_ont, cell_ont_up, ciddr), out, compress = "gz")

### 2019-10-07 ch6_bred/scripts/2_tabula-muris_hematopoietic.R

@@ -1,8 +1,8 @@
 library(tidyverse)
 
 dest_dir <- "derived_files/tabula_muris/"
-dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
-
+hema_dir <- "derived_files/tabula_muris_hema/"
+dir.create(hema_dir, recursive = TRUE, showWarnings = FALSE)
 
 data_rnaseq <- paste0(dest_dir, "data_rnaseq/")
 data_rds <- paste0(dest_dir, "data_rds/")
@@ -55,11 +55,19 @@ data_rds <- paste0(dest_dir, "data_rds/")
 
 # FILTER DATA -------------------------------------------------------------
 data_rds <- "derived_files/tabula_muris/data_rds/"
+hema_rds <- "derived_files/tabula_muris_hema/data_rds/"
+dir.create(hema_rds, recursive = TRUE, showWarnings = FALSE)
 
-if (!file.exists(paste0(data_rds, "data.rds"))) {
+if (!file.exists(paste0(hema_rds, "data.rds"))) {
   counts <- read_rds(paste0(data_rds, "counts.rds"))
-  # expression <- read_rds(paste0(data_rds, "expression.rds"))
   cell_info <- read_rds(paste0(data_rds, "cell_info.rds"))
+
+  list2env(read_rds("derived_files/cell_ontology.rds"), .GlobalEnv)
+  choose <- cell_ont %>% filter(name == "hematopoietic cell") %>% pull(id)
+  choose_down <- cell_ont_up %>% filter(upstream %in% choose) %>% pull(cell_ontology_id) %>% unique()
+
+  cell_info <- cell_info %>% filter(cell_ontology_id %in% c(choose_down, choose))
+  counts <- counts[cell_info$cell, , drop = FALSE]
 
   seu <-
     Seurat::CreateSeuratObject(
@@ -74,10 +82,10 @@ if (!file.exists(paste0(data_rds, "data.rds"))) {
   feature_info <- seu@assays$RNA@meta.features %>% rownames_to_column("feature_id") %>% as_tibble()
   sample_info <- seu@meta.data %>% rownames_to_column("cell_id") %>% as_tibble()
 
-  write_rds(lst(expression, feature_info, sample_info), paste0(data_rds, "data.rds"))
[…]

### 2019-10-10 ch1_introduction/content.tex
update

@@
 Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality.
-The variation within and between cells happens because of differences in gene expression.
-The mechanism by which this happens is called gene regulation, and takes place primarily at the level of transcription.
+The naturally occurring cell-to-cell variability happens at the level of gene expression.
+Gene expression itself can be controlled at different levels (Figx), one of which is gene regulation by transcription.
+%Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P. An Overview of Gene Control. InMolecular Biology of the Cell. 4th edition 2002. Garland Science.
+
+Fig.
+x: \textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e.
+splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity.
+%From: (https://www.ncbi.nlm.nih.gov/books/NBK26885/figure/A1220/?report=objectonly) Meer inhoudelijke info: https://www.ncbi.nlm.nih.gov/books/NBK26885/ Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P.
+An Overview of Gene Control.
+InMolecular Biology of the Cell.
+4th edition 2002.
+Garland Science.
+
 According to the needs of a cell, different genes are being transcribed.
-Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages\cite{strachan_humanmoleculargenetics_2011}.
+Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific or may be expressed in response to developmental and environmental signals\cite{strachan_humanmoleculargenetics_2011}.
 
-% CVC
+Transcription factors (TFs) modulate the rate of gene transcription by binding and recruiting the transcriptional machinery to \emph{cis}-regulatory regions (enhancers, and silencers) that are typically located in the promotor region of target genes.
+These bindings may result in increased or decreased gene expression.
+There are several TF families of which members share structural characteristics (e.g. zink finger, helix-loop-helix).
 
-Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate).
-Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets.
-Typically, one TF will regulate the transcription rate of many targets.
-
-Production of a specific molecule might require multiple cascades of gene regulation.
-The collection of all gene regulatory interactions between transcription factors and targets is called a gene regulatory network (GRN).
-Studying the active parts of a cell's gene regulatory network can thus reveal which dynamic processes are taking place.
+Many TFs are commonly present in virtually all cell types (e.g.
+NF-κB), while others are specific for cells and developmental stages.
+Typically, the same TF can regulate the rate of transcription of many target genes in different cell types, indicating that these gene regulatory networks (GRNs) are dynamic.
+Moreover, the production of a specific molecule might require several gene regulatory cascades.
+Studying the active parts of a cell's GRN can thus reveal which dynamic processes are taking place within a cell.
+% Ref: Lambert SA, Jolma A, Campitelli LF, Das PK, Yin Y, Albu M, Chen X, Taipale J, Hughes TR, Weirauch MT. The human transcription factors. Cell. 2018 Feb 8;172(4):650-65.
+% The kappa in NF-κB wordt niet goed omgezet in pdf doc
 
 \subsection{Profiling single cells}
-In order to understand a biological process, it is often quite helpful to be able to profile (i.e.
-observe) the biomolecular components involved therein.
-The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.
+Several technologies are now available to profile (i.e.
+observe) biomolecular components, allowing us to gain better understanding in the biological processes that take place within a cell.
+The single-cell "omics" technologies originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.
 
 \subsubsection{Single-cell}
-The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
-Since it was developed by Coons et al.
-in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins\cite{coons_immunologicalpropertiesantibody_1941}.
+The earliest approach for measuring the abundance of a particular molecule in \emph{single cells} is the microscope.
+Since its development by Coons et al.
+(1941), immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins\cite{coons_immunologicalpropertiesantibody_1941}.
 In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system.
@@
 
-\mycomment{Rephrase the microscope sentence (dino joke?)}
-
 IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation.
-Using different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously.
-Characterising cells in a quantifiable way is labour intensive; however, since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation).
+The use of different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously.
+Characterising cells in a semi-quantifiable way is labour intensive, however; since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation).
 While modern implementations of IHC improve the throughput drastically by using robots to automate the image acquisition and computer software to automate cell segmentation, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.
@@
 
-Since IHC and flow cytometry, many new technologies have been developed which allow quantifying expression levels of molecules in single cells (e.g.
+Besides IHC and flow cytometry, many new technologies have been developed which allow quantifying expression levels of molecules in single cells (e.g.
 mass cytometry, single-cell qPCR, FISH).
-All of these single-cell (non-omics) technologies are limited by the number of different molecules they could measure, however; and thus required handpicking the molecules of interest before performing an experiment, making the experiment biased towards the preconceptions of the experimenter.
+All of these single-cell (non-omics) technologies are limited by the number of different molecules they could measure, however.
+Selecting molecules of interest prior to analysis, makes the experiment biased towards the preconceptions of the experimenter.
 
@@
 %
-%cran <- c("babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "princurve", "proxyC", "qsub", "SCORPIUS", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
+%cran <- c("babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "lmds", "princurve", "proxyC", "qsub", "SCORPIUS", "badger", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
 %bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
@@
 
+    \cranpkg{badger} & ctb & CRAN & ? & Query information and generate badge for using in README and GitHub Pages \\
     \biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\

### 2019-10-10 ch6_bred/scripts/1_genesets.R

@@ -44,4 +44,58 @@ genesets <- bind_rows(
   reactome
 )
 
-write_rds(genesets, "derived_files/data_genesets.rds")
+write_rds(genesets, "derived_files/data_genesets_mouse.rds")
+
+
+
+
+
+
+
+
+
+## CREATE GO tibble
+go.entrezs <- as.list(org.Hs.eg.db::org.Hs.egGO2ALLEGS)
+ontologies <- annotate::Ontology(names(go.entrezs))
+go.entrezs <- go.entrezs[ontologies %in% c("BP", "MF")]
+go.entrezs <- lapply(go.entrezs, function(go) na.omit(unique(go)))
+go.ids <- names(go.entrezs)
+go.terms <- annotate::Term(GO.db::GOTERM[go.ids])
+go <- tibble(database = "GO", id = go.ids, description = go.terms, entrezs = go.entrezs)
+
+## CREATE KEGG OBJECTS
+pws <- pbapply::pblapply(names(KEGGREST::keggList("pathway", "hsa")), function(n) KEGGREST::keggGet(n)[[1]])
+kegg.entrezs <- lapply(pws, function(pw) {
+  genes <- pw$GENE
+  if (length(genes) > 0) unique(genes[seq(1, length(genes)-1, by=2)])
+  else genes
+})
+kegg.ids <- map_chr(pws, ~ .$ENTRY)
+kegg.names <- map_chr(pws, ~ .$NAME)
+kegg <- tibble(
+  database = "KEGG",
+  id = kegg.ids,
+  description = kegg.names %>% gsub(" - [^-]*", "", .),
+  entrezs = kegg.entrezs
+)
+
+## CREATE REACTOME OBJECTS
+PATHID2NAME <- as.list(reactome.db::reactomePATHID2NAME)
+PATHID2NAME <- PATHID2NAME[grepl("Homo sapiens", PATHID2NAME)]
+EXTID2PATHID <- as.list(reactome.db::reactomeEXTID2PATHID)
+reactome.entrezs <- pbapply::pblapply(names(PATHID2NAME), cl = 8, function(id) names(EXTID2PATHID)[sapply(EXTID2PATHID, function(x) id %in% x)])
+reactome <- tibble(
+  database = "reactome",
+  id = names(PATHID2NAME),
+  description = PATHID2NAME %>% gsub("^Homo sapiens: ", "", .),
+  entrezs = reactome.entrezs
+)
+
+# genesets db
[…]

### 2019-10-10 ch6_bred/scripts/2_tabula-muris_hematopoietic.R

@@ -240,19 +240,19 @@ if (!file.exists(paste0(dest_dir, "qsub_handle.rds"))) {
 }
 
 dest_dir <- "derived_files/tabula_muris_hema/"
-qsub_handle <- read_rds(paste0(dest_dir, "qsub_handle.rds"))
-
-grn <- qsub::qsub_retrieve(
-  qsub_handle,
-  wait = "just_do_it",
-  post_fun = function(i, li) {
-    li$importance <- li$importance %>% filter(importance > .01)
-    li$importance_sc <- li$importance_sc %>% filter(importance_sc > .01) %>%
-      inner_join(li$importance %>% select(regulator, target), by = c("regulator", "target"))
-    li
-  }
-)
-write_rds(grn, paste0(dest_dir, "grn.rds"))
+# qsub_handle <- read_rds(paste0(dest_dir, "qsub_handle.rds"))
+#
+# grn <- qsub::qsub_retrieve(
+#   qsub_handle,
+#   wait = "just_do_it",
+#   post_fun = function(i, li) {
+#     li$importance <- li$importance %>% filter(importance > .01)
+#     li$importance_sc <- li$importance_sc %>% filter(importance_sc > .01) %>%
+#       inner_join(li$importance %>% select(regulator, target), by = c("regulator", "target"))
+#     li
+#   }
+# )
+# write_rds(grn, paste0(dest_dir, "grn.rds"))
 
 grn <- read_rds(paste0(dest_dir, "grn.rds"))
 
@@ -263,13 +263,14 @@ grn[map_lgl(grn, ~ length(.) == 1 && is.na(.))] <- NULL
 importance <-
   grn %>%
   map_df("importance") %>%
+  arrange(desc(importance)) %>%
   mutate(
     i = row_number(),
     name = paste0(regulator, "->", target)
   )
 ggplot(importance) +
   geom_point(aes(effect, importance))
-write_tsv(importance, "~/importance.tsv")
+write_tsv(importance %>% filter(importance > .1), "~/importance.tsv")
 importance_sc <- grn %>%
   map_df("importance_sc") %>%
   inner_join(importance %>% select(-importance, -name), by = c("regulator", "target"))
[…]

### 2019-10-10 ch6_bred/scripts/2_tcga.R

@@ -1,109 +1,196 @@
 library(tidyverse)
 
-dest_dir <- "derived_files/tabula_muris/"
-dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
+tcga_folder <- "~/Workspace/tcga/"
+data_dir <- "derived_files/tcga/"
+dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
 
+tcga_data_file <- paste0(tcga_folder, "data.rds")
+data_file <- paste0(data_dir, "data.rds")
 
-data_rnaseq <- paste0(dest_dir, "data_rnaseq/")
-data_rds <- paste0(dest_dir, "data_rds/")
+if (!file.exists(tcga_data_file)) {
+  sample_info <- read_tsv(paste0(tcga_folder, "sample_sheet.tsv"))
+  colnames(sample_info) <- colnames(sample_info) %>% tolower() %>% gsub(" ", "_", .)
 
-# zip_file <- paste0(dest_dir, "data_rnaseq.zip")
-# download.file("https://ndownloader.figshare.com/articles/5968960/versions/3", zip_file)
-# unzip(zip_file, exdir = data_rnaseq)
-# unzip(paste0(data_rnaseq, "droplet.zip"), exdir = data_rnaseq)
-
-# cell_info <- read_csv(
-#   paste0(data_rnaseq, "annotations_droplet.csv"),
-#   col_types = cols(
-#     .default = "c",
-#     tissue_tSNE_1 = "d",
-#     tissue_tSNE_2 = "d",
-#     subsetA = "l",
-#     subsetB = "l",
-#     subsetC = "l",
-#     subsetD = "l"
-#   )
-# )
-#
-# write_rds(cell_info, paste0(data_rds, "cell_info.rds"), compress = "gz")
-#
-# droplet <- paste0(data_rnaseq, "/droplet")
-# samples <- list.files(droplet, full.names = TRUE)
-# counts <- pbapply::pblapply(samples, function(folder_10x) {
-#   channel <- basename(folder_10x) %>% gsub(".*-", "", .)
-#
-#   counts <- Seurat::Read10X(folder_10x) %>% Matrix::t()
-#   rownames(counts) <- paste0(channel, "_", rownames(counts))
-#
-#   counts <- counts[rownames(counts) %in% cell_info$cell, ]
-#
-#   counts
-# }) %>% do.call(rbind, .)
[…]

### 2019-10-10 ch6_bred/scripts/find_mart.R

@@ -1,11 +1,13 @@
 
-archs <- listEnsemblArchives()
+
+# MOUSE -------------------------------------------------------------------
+archs <- biomaRt::listEnsemblArchives()
 gene_symbols <- pbapply::pblapply(seq_len(nrow(archs)), function(i) {
   tryCatch({
-    ensembl <- useMart("ENSEMBL_MART_ENSEMBL", dataset="mmusculus_gene_ensembl", host=archs$url[[i]])
-    attrs <- listAttributes(ensembl)
+    ensembl <- biomaRt::useMart("ENSEMBL_MART_ENSEMBL", dataset="mmusculus_gene_ensembl", host=archs$url[[i]])
+    attrs <- biomaRt::listAttributes(ensembl)
     entrez_attr_name <- ifelse("entrezgene" %in% attrs$name, "entrezgene", "entrezgene_id")
-    gene_symbols <- getBM(
+    gene_symbols <- biomaRt::getBM(
       attributes = c("ensembl_gene_id", "mgi_symbol", entrez_attr_name),
       filters = "ensembl_gene_id",
       values = colnames(counts),
@@ -24,5 +26,38 @@ archs %>%
   mutate(
     matching_genes = map_int(gene_symbols, function(x) if(is.data.frame(x)) length(unique(x$ensembl_gene_id)) else 0L),
     num_genes = ncol(counts)
+  ) %>%
+  filter(num_genes == matching_genes)
+
+
+
+
+
+# HUMAN -------------------------------------------------------------------
+archs <- biomaRt::listEnsemblArchives()
+gene_symbols <- pbapply::pblapply(seq_len(nrow(archs)), function(i) {
+  tryCatch({
+    ensembl <- biomaRt::useMart("ENSEMBL_MART_ENSEMBL", dataset="hsapiens_gene_ensembl", host=archs$url[[i]])
+    attrs <- biomaRt::listAttributes(ensembl)
+    entrez_attr_name <- ifelse("entrezgene" %in% attrs$name, "entrezgene", "entrezgene_id")
+    gene_symbols <- biomaRt::getBM(
+      attributes = c("ensembl_gene_id", "hgnc_symbol", entrez_attr_name),
+      filters = "ensembl_gene_id",
+      values = feature_info$feature_id %>% gsub("\\..*$", "", .),
+      mart = ensembl
+    )
+    gene_symbols %>%
+      rename(entrez_gene_id = !!entrez_attr_name) %>%
[…]

### 2019-10-11 ch1_introduction/analysis/computational_tools.R
making small comptool figures

@@ -1,0 +1,153 @@
+library(tidyverse)
+
+fig_size <- 3
+
+melt <- function(mat) {
+  mat %>%
+    magrittr::set_colnames(NULL) %>%
+    magrittr::set_rownames(NULL) %>%
+    reshape2::melt(varnames = c("y", "x"), value.name = "value") %>%
+    as_tibble()
+}
+raster <- function(meltdf, legend = FALSE, palette = "RdYlBu") {
+  g <-
+    ggplot(meltdf) +
+    geom_raster(aes(x, y, fill = value)) +
+    scale_fill_distiller(palette = palette, breaks = c(0, 1), labels = c("Min", "Max")) +
+    dynplot::theme_graph() +
+    labs(fill = "Expression") +
+    coord_cartesian(expand = FALSE)
+  if (!legend) {
+    g <- g + theme(legend.position = "none")
+  }
+  g
+}
+
+
+# Generate toy datasets ---------------------------------------------------
+set.seed(1)
+
+# linear
+linear <- dyntoy::generate_dataset(
+  model = "linear",
+  num_cells = 60,
+  num_features = 40,
+  dropout_probability_factor = 0
+)
+
+expr_li <- as.matrix(linear$expression) %>% dynutils::scale_quantile() %>% t() %>% dynutils::scale_quantile()
+
+dropout_weight <- 1 / (as.vector(expr_li)+1)
+rm_ix <- sample(seq_along(dropout_weight), length(dropout_weight) * .2, prob = dropout_weight, replace = FALSE)
+
+expr_li_nodrop <- expr_li
+expr_li[rm_ix] <- 0
+
+ordr_li <- hclust(dist(expr_li))$order
+ordc_li <- hclust(dist(t(expr_li)))$order
+
+# branching
+set.seed(1)
+branching <- dyntoy::generate_dataset(
+  model = dyntoy::model_binary_tree(num_branchpoints = 2L),
+  num_cells = 100,
+  num_features = 100,
+  allow_tented_progressions = FALSE,
+  sample_mean_count = function() rnorm(1, 1000, 100),
+  sample_dispersion_count = function(mean) map_dbl(mean, ~runif(1, ./10, ./4)),
+  dropout_probability_factor = 10
+)
+expr_br <- as.matrix(branching$expression)
+expr_br_sc <- expr_br %>% dynutils::scale_quantile() %>% t() %>% dynutils::scale_quantile()
+
[…]

### 2019-10-11 ch1_introduction/analysis/computational_tools.R
create more figures

@@ -47,9 +47,9 @@ ordr_li <- hclust(dist(expr_li))$order
 ordc_li <- hclust(dist(t(expr_li)))$order
 
 # branching
-set.seed(1)
+set.seed(2)
 branching <- dyntoy::generate_dataset(
-  model = dyntoy::model_binary_tree(num_branchpoints = 2L),
+  model = dyntoy::model_binary_tree(num_branchpoints = 1L),
   num_cells = 100,
   num_features = 100,
   allow_tented_progressions = FALSE,
@@ -60,10 +60,13 @@ branching <- dyntoy::generate_dataset(
 expr_br <- as.matrix(branching$expression)
 expr_br_sc <- expr_br %>% dynutils::scale_quantile() %>% t() %>% dynutils::scale_quantile()
 
-ordr_br <- hclust(dist(expr_br_sc))$order
-ordc_br <- hclust(dist(t(expr_br_sc)))$order
+hclr_br <- hclust(dist(expr_br_sc))
+ordr_br <- hclr_br$order
+hclc_br <- hclust(dist(t(expr_br_sc)))
+ordc_br <- hclc_br$order
 
 # Experiment --------------------------------------------------------------
+set.seed(1)
 ggsave(
   "fig/comptools/1_experiment_expression.pdf",
   expr_li %>% melt() %>% raster(),
@@ -134,7 +137,7 @@ g <-
     dimred = dimred_br,
     color_cells = "none",
     plot_trajectory = FALSE,
-    size_cells = 1
+    size_cells = 2
   ) +
   theme_classic() +
   theme(axis.text = element_blank(), axis.ticks = element_blank()) +
@@ -149,5 +152,493 @@ ggsave(
 )
 
 # Clustering --------------------------------------------------------------
+ggsave(
+  "fig/comptools/5_cluster_input.pdf",
+  expr_br_sc[ordr_br, ordc_br] %>% melt() %>% raster(),
+  width = fig_size,
+  height = fig_size
+)
 
+cluscut_br <- cutree(hclc_br, k = 3)
+
+clus_ord <- unique(cluscut_br[ordc_br])
+rasters <- map(clus_ord, function(cli) {
+  ix <- cluscut_br[ordc_br] == cli
+  expr_br_sc[ordr_br, ordc_br[ix]] %>% melt() %>% raster()
+})
+widths <- map_int(clus_ord, ~ sum(. == cluscut_br))
+g <- patchwork::wrap_plots(rasters, nrow = 1, widths = widths)
+
+ggsave(
[…]

### 2019-10-11 ch1_introduction/content.tex
update

@@
 
+\subsection{Central dogma}
 The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins\cite{strachan_humanmoleculargenetics_2011}.
@@
 
+% KDP: je zegt niets over epigenetica, die min of meer bepaald welke gebieden in het DNA worden vertaald (dit leidt je misschien te ver...)
 During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA.
@@
 The naturally occurring cell-to-cell variability happens at the level of gene expression.
-Gene expression itself can be controlled at different levels (Figx), one of which is gene regulation by transcription.
-%Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P. An Overview of Gene Control. InMolecular Biology of the Cell. 4th edition 2002. Garland Science.
+Gene expression itself can be controlled at different levels (Figure~\ref{fig:controlexpression}), one of which is gene regulation by transcription.
 
-Fig.
-x: \textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e.
-splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity.
-%From: (https://www.ncbi.nlm.nih.gov/books/NBK26885/figure/A1220/?report=objectonly) Meer inhoudelijke info: https://www.ncbi.nlm.nih.gov/books/NBK26885/ Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P.
-An Overview of Gene Control.
-InMolecular Biology of the Cell.
-4th edition 2002.
-Garland Science.
+\begin{figure}[htb!]
+	\centering\includegraphics[width=.8\linewidth]{fig/control_gene_expression} % TODO: change image
+	\caption{
+		\textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e.
+splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity \cite{alberts_overviewgenecontrol_2002}.
+	}
+	\label{fig:controlexpression}
+\end{figure}
+
+% KDP: niet alleen TFs maar volledige epigenetica regelt of een gen al dan niet tot expressie wordt gebracht, dus misschien kort vermelden dat het toch iets complexer is
 
@@
 These bindings may result in increased or decreased gene expression.
-There are several TF families of which members share structural characteristics (e.g. zink finger, helix-loop-helix).
+There are several TF families of which members share structural characteristics (e.g. zinc finger, helix-loop-helix).
 
 Many TFs are commonly present in virtually all cell types (e.g.
-NF-κB), while others are specific for cells and developmental stages.
+NF-$\kappa$B), while others are specific for cells and developmental stages\cite{lambert_humantranscriptionfactors_2018}.
 Typically, the same TF can regulate the rate of transcription of many target genes in different cell types, indicating that these gene regulatory networks (GRNs) are dynamic.
@@
 Studying the active parts of a cell's GRN can thus reveal which dynamic processes are taking place within a cell.
-% Ref: Lambert SA, Jolma A, Campitelli LF, Das PK, Yin Y, Albu M, Chen X, Taipale J, Hughes TR, Weirauch MT. The human transcription factors. Cell. 2018 Feb 8;172(4):650-65.
-% The kappa in NF-κB wordt niet goed omgezet in pdf doc
+
+% KDP: misschien interessant om over tumor heterogeniteit te spreken: waarvan we vroeger dachten dat homogene massa van cellen was, blijkt toch veel complexer te zijn --> nu mogelijk om met sc technieken te bestuderen?
 
@@
 
+% KDP: antigen-antibody: eerder over protein spreken ipv over antigen dan wordt het toch iets duidelijker, denk ik: een antibody herkent en bindt een bepaald proteine/eiwit,...
 IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation.
@@
 Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
+% KDP: circumvents segmentation: if you want to investigate tissues you have to dissociate the tissue 
 Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies. 
+% KDP: vermeld hier hoeveel merkers flow cytometrie tegenwoordig aankan, want je komt er later op terug
 
@@
 "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
+
 There are many types of "omics", but the most commonly used are the following.
@@
 Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively.
-A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics.
-If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable.
+
 Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts.
 By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
-\mycomment{Demonstrate the masking effect of bulk analyses.}
+
+Traditionally, profiling nano-litre volumes was infeasible.
+To capture enough material to generate a profile, an ensemble of cells needs to be pooled and lysed together, thereby granting the technology the name "bulk" omics.
+Bulk omics is a major workhorse in <...> and has applications in <...>. To capture enough material, an ensemble of cells is pooled and profiled all together.
+
+\mycomment{TODO: fill in}
+
+A notable downside of bulk omics arises when the cells are biomolecularly heterogeneous (Figure~\ref{fig:bulk_vs_sc}A).
+Since the profile is a population average (or rather, a summation), information pertaining the individual profiles of cell subpopulations cannot be discerned (Figure~\ref{fig:bulk_vs_sc}B).
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\LARgefigure]{fig/bulk_vs_sc} 
+	\caption{
+		\textbf{The 'masking' effect of bulk omics.} 
+		\textbf{A:} Cells from several subpopulations are incorrectly assumed to be homogeneous and are profiled with a single bulk omics experiment.
+		\textbf{B:} The signals from the different subpopulations are masked. The resulting profile is dissimilar from the majority of cells it is supposed to represent.
+	}
+	\label{fig:bulk_vs_sc}
+\end{figure}
 
@@
 to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A).
-During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells (Figure \ref{fig:profiling_convergence}B) and measuring other levels of information such as proteomic expression levels (Figure \ref{fig:profiling_convergence}C).
+During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells and measuring other levels of information (e.g.
+proteomic expression levels and spatial location).
+% KDP: wat bedoel je met: such as proteomic expr levels
 
@@
   \centering
-  A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
-  % also figure with cell numbers
-  C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
+  \includegraphics[width=\linewidth]{fig/singlecell_omics_evolution} 
   \caption{
-    A.
-\textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
+    \textbf{} %Convergence of "Omics" Biology and Single-Cell Biology.
+Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
 Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}.
-    B. Cell numbers vs. gene count over time.
-    C. scmultiomics\cite{moudgil_multimodalscrnaseq_2019}. 
-    \fbox{\color{red} TODO: combine images, rewrite the description.}
   }
@@
 As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells.
+% KDP: hier heb je het al eens over gehad: zou je dit niet samenbrengen?
 
 \section{Computational tools}
-The new types of analyses permitted by single-cell omics harbour exceptional opportunities to discover new aspects of biology and redefine existing knowledge (Figure \ref{fig:comp_tools}A).
-These include the following categories\cite{zappia_exploringsinglecellrnaseq_2018}.
-\begin{itemize}
-	\item \textbf{Data imputation}: \ldots.
-	\item \textbf{Data integration}: \ldots.
-	\item \textbf{Clustering}: \ldots.
-	\item \textbf{Dimensionality reduction}: providing a visual and informative overview of a given dataset.
-	\item \textbf{Trajectory inference}: identifying and characterising transitions between different cellular states.
-	\item \textbf{Trajectory alignment}: \ldots.
-	\item \textbf{Network inference}: inferring gene regulatory interactions between transcription factors across individual cells.
-\end{itemize}
-\mycomment{To do: write a section for each of these categories.}
+Whole-genome profiling at single-cell level allows new types of analyses with which to study cellular heterogeneity at a hitherto unseen throughput. 
+The new types of analyses permitted by single-cell omics present several computational challenges\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
+This necessitates the development of novel computational tools, either because the problem statement of the performed analysis is completely novel, or to adapt existing methodology to new data characteristics (e.g.
+dimensionality and noise levels).
+Over the past five years, already 450 new tools for analysing single-cell omics data have been developed\cite{zappia_exploringsinglecellrnaseq_2018}.
+These tools typically fall within one of nine categories (Figure \ref{fig:comp_tools}).
+We discuss each category in the following subsections.
 
-Developing new computational tools to perform these analyses has proven challenging for two main reasons\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
-Firstly, single-cell omics data suffers from hitherto unseen noise characteristics, including high dropout rates and high levels of transcriptional stochasticity.
-In addition, a low abundance of single-cell omics data that can be used as a ground-truth obstruct quantitative evaluation of such methods.
+\mycomment{Todo: describe the characteristics of single cell omics data in the subsection on 'single cell omics'.
+-> counts data.
+additional information from the cell can be present, such as the time of sampling in a time-series experiment or a cell type if the cells have been sorted using a FACS experiment.
+information on the features are typically restricted to the gene symbol, or an identifier of a xxx, such as ensembl or entrez.
+in comparison to bulk data, single-cell omics data is typically very sparse, due to technical and biological stochasticity.
+if several transcripts of a low abundance gene are not captured, this will be seen as a dropout.
+dropouts can also be biological, due to transcriptional bursting.
+}
 
 \begin{figure}[htb!]
-  \centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
-  \centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
-  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016)\cite{wagner_revealingvectorscellular_2016}.
-    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}.
-    \fbox{\color{red} TODO: combine images, rewrite the description.}
+	\centering
+	\includegraphics[width=.75\linewidth]{fig/comptools/comptools_3.pdf} 
+  \caption{
+  	\textbf{Computational tools developed for processing single-cell omics data usually lie within one of nine categories.}
+  	\textbf{A:} Normalisation, separating meaningful biological signal from technical noise or irrelevant biological information.
+  	\textbf{B:} Data imputation, masking dropouts.
+  	\textbf{C:} Data integration, combining multiple single-cell modalities.
+  	\textbf{D:} Dimensionality reduction, providing a visual overview of the main cellular heterogeneity in the dataset.
+  	\textbf{E:} Clustering, grouping together similar cells.
+  	\textbf{F:} Trajectory inference, identifying and characterising transitions between different cellular states.
+  	\textbf{G:} Network inference, predicting gene regulatory interactions from cell-to-cell heterogeneity.
+  	\textbf{H:} RNA velocity, inferring the directionality of trajectories by predicting the future state of each cell.
+  	\textbf{I:} Trajectory alignment, aligning multiple trajectories.
   }
@@
 \end{figure}
+
+\subsection{Normalisation}
+
+\subsection{Data imputation}
+
+\subsection{Data integration}
 
@@
 \mycomment{This section is not very interestingly written}
+
+\subsection{Clustering}
 
@@
 In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required.
-On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: update label
+On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: Already used this figure in the previous section
   \label{fig:bulk_vs_singlecell}
@@
 In this setting, heterogeneity in the cell population eases network inference, rather than mask condition-specific expression patterns in regulatory interactions.
+
+% mention differential networks
+
+\subsection{RNA velocity}
+
+\subsection{Trajectory alignment}
 
@@
 We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
-This includes help speed up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
+This includes speeding up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
 Several of these package receive millions of downloads per year (devtools, remotes, rlang).

### 2019-10-11 ch6_bred/scripts/2_tabula-muris.R
update

@@ -387,7 +387,7 @@ backgr <- pbapply::pblapply(ks, cl = 1, function(k) {
         FN = ONTPOS - TP,
         FP = CLUSPOS - TP,
         TN = TOT - TP - FN - FP,
-        TPR = TP %s/% ONTPOS,
+        TPR = TP %s/% (TP + FN),
         TNR = TN %s/% (TN + FP),
         PPV = TP %s/% (TP + FP),
         NPV = TN %s/% (TN + FN),

### 2019-10-11 ch6_bred/scripts/2_tcga.R

@@ -324,7 +324,7 @@ if (!file.exists(paste0(data_dir, "qsub_handle.rds"))) {
 }
 
 
-if (!file.exists(paste0(data_dir, "grn.rds"))) {
+# if (!file.exists(paste0(data_dir, "grn.rds"))) {
   qsub_handle <- read_rds(paste0(data_dir, "qsub_handle.rds"))
 
   grn <- qsub::qsub_retrieve(
@@ -339,7 +339,7 @@ if (!file.exists(paste0(data_dir, "grn.rds"))) {
   )
 
   write_rds(grn, paste0(data_dir, "grn.rds"))
-}
+# }
 
 grn <- read_rds(paste0(data_dir, "grn.rds"))
 
@@ -370,7 +370,8 @@ samples <- levels(importance_sc$cell_id)
 imp_sc_mat <- Matrix::sparseMatrix(
   i = importance_sc$cell_id %>% as.integer,
   j = importance_sc$i,
-  x = importance_sc$importance_sc,
+  # x = importance_sc$importance_sc,
+  x = importance_sc$importance,
   dims = c(length(samples), nrow(importance)),
   dimnames = list(samples, importance$name)
 )
@@ -380,7 +381,9 @@ rm(imp_sc_mat)
 list2env(read_rds(data_file), .GlobalEnv)
 # df <- data.frame(
 #   sample_info,
+#   # dimred
 #   dimred2
+#   # dimred_fr2
 # )
 # ggplot(df) + geom_point(aes(comp_1, comp_2, colour = project_id))
 
@@ -441,145 +444,84 @@ dimred_fr <- map(knns, "dimred2") %>% do.call(cbind, .) %>% dynutils::scale_unif
 dimred_fr2 <- dyndimred::dimred_pca(dimred_fr)
 plot(dimred_fr2, col = clus)
 
-dimred_umap <- dyndimred::dimred_umap(dimred_fr, pca_components = NULL, n_neighbors = 50)
-plot(dimred_umap, col = clus)
-
-df <- data.frame(
-  dimred_umap,
-  # dimred2,
-  # dimred_fr2,
-  sample_info,
-  clus
-)
+# dimred_umap <- dyndimred::dimred_umap(dimred_fr, pca_components = NULL, n_neighbors = 50)
+# plot(dimred_umap, col = clus)
 
 
+
+proj_annot <-
+  read_tsv(paste0(data_dir, "projects.tsv")) %>%
+  mutate(
[…]

### 2019-10-12 ch1_introduction/content.tex
reindent
update

@@
 
-% KDP: niet alleen TFs maar volledige epigenetica regelt of een gen al dan niet tot expressie wordt gebracht, dus misschien kort vermelden dat het toch iets complexer is
+% CVC: weggewerkt --> KDP: niet alleen TFs maar volledige epigenetica regelt of een gen al dan niet tot expressie wordt gebracht, dus misschien kort vermelden dat het toch iets complexer is
 
@@
 The earliest approach for measuring the abundance of a particular molecule in \emph{single cells} is the microscope.
-Since its development by Coons et al.
-(1941), immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins\cite{coons_immunologicalpropertiesantibody_1941}.
-In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system.
-A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.
+Since its development by Coons et al. (1941), immunohistochemistry (IHC) has been instrumental in visualising proteins.\cite{coons_immunologicalpropertiesantibody_1941}. 
+A cell can present a particular type of protein, also called an antigen, on its cell surface.
+In many multicellular organisms, antigens can stimulate the immune system to produce antibodies.
+IHC realises the visualisation of proteins by exploiting the principle of antibodies binding specific to antigens.
 
-% KDP: antigen-antibody: eerder over protein spreken ipv over antigen dan wordt het toch iets duidelijker, denk ik: een antibody herkent en bindt een bepaald proteine/eiwit,...
-IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation.
-The use of different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously.
+% CVC --> comment verwerkt.
+KDP: antigen-antibody: eerder over protein spreken ipv over antigen dan wordt het toch iets duidelijker, denk ik: een antibody herkent en bindt een bepaald proteine/eiwit,...
+IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon excitation.
+The use of several colours (wavelengths) allows measuring expression levels of different antibodies simultaneously.
 Characterising cells in a semi-quantifiable way is labour intensive, however; since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation).
-While modern implementations of IHC improve the throughput drastically by using robots to automate the image acquisition and computer software to automate cell segmentation, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.
+Modern implementations of IHC improve the throughput drastically by using robots and computer software to automate the image acquisition and cell segmentation.
+Nonetheless, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.
 
 Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
-% KDP: circumvents segmentation: if you want to investigate tissues you have to dissociate the tissue 
+A major limitation of cytometry is that it only enables to study a suspension of viable single cells.
+Today, there are several protocols available to gain viable single cells from human tissues and tumours.
+
+% Eventueel dit als ref: Leelatian N, Doxie DB, Greenplate AR, Sinnaeve J, Ihrie RA, Irish JM.
+Preparing viable single cells from human tissue and tumors for cytomic analysis.
+Current protocols in molecular biology.
+2017 Apr;118(1):25C-1.
+% --> CVC: Comment verwerkt KDP: circumvents segmentation: if you want to investigate tissues you have to dissociate the tissue . 
 Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies. 
-% KDP: vermeld hier hoeveel merkers flow cytometrie tegenwoordig aankan, want je komt er later op terug
+% KDP: vermeld hier hoeveel merkers flow cytometrie tegenwoordig aankan, want je komt er later op terug --> CVC zie mail
 
 Besides IHC and flow cytometry, many new technologies have been developed which allow quantifying expression levels of molecules in single cells (e.g.
-mass cytometry, single-cell qPCR, FISH).
+mass cytometry, single-cell quantitative polymerase chain reaction, fluorescence \emph{in situ} hybridization).
 All of these single-cell (non-omics) technologies are limited by the number of different molecules they could measure, however.

### 2019-10-12 ch1_introduction/content.tex
new modifications

@@
 
-% CVC: weggewerkt --> KDP: niet alleen TFs maar volledige epigenetica regelt of een gen al dan niet tot expressie wordt gebracht, dus misschien kort vermelden dat het toch iets complexer is
-
 According to the needs of a cell, different genes are being transcribed.
@@
 
-% CVC --> comment verwerkt.
-KDP: antigen-antibody: eerder over protein spreken ipv over antigen dan wordt het toch iets duidelijker, denk ik: een antibody herkent en bindt een bepaald proteine/eiwit,...
 IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon excitation.
@@
 Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
-A major limitation of cytometry is that it only enables to study a suspension of viable single cells.
-Today, there are several protocols available to gain viable single cells from human tissues and tumours.
-
-% Eventueel dit als ref: Leelatian N, Doxie DB, Greenplate AR, Sinnaeve J, Ihrie RA, Irish JM.
-Preparing viable single cells from human tissue and tumors for cytomic analysis.
-Current protocols in molecular biology.
-2017 Apr;118(1):25C-1.
-% --> CVC: Comment verwerkt KDP: circumvents segmentation: if you want to investigate tissues you have to dissociate the tissue . 
-Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies. 
-% KDP: vermeld hier hoeveel merkers flow cytometrie tegenwoordig aankan, want je komt er later op terug --> CVC zie mail
+Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in the blood.
+However, many protocols already exist to extract viable single cells from tissue as well\cite{leelatian_preparingviablesingle_2017}.
+Common flow cytometry enables to measure proteins expression levels of millions of cells for 8 different antibodies, though state-of-the-art instrumentation allow detecting up to 27 biomarkers simultaneously\cite{cossarizza_guidelinesuseflow_2017}.
 
@@
 We discuss each category in the following subsections.
+Whole-genome profiling at a single-cell level harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
+% KDP: the new type of computational analysis etc... of eerder zeggen dat met de ontwikkeling van single cell technieken ook nieuwe computationele methoden moe(s)ten ontwikkeld worden
+(Figure \ref{fig:comp_tools})
 
@@
 		\cranpkg{dynutils} & aut & CRAN & 5657 & Common functionality for the dynverse packages \\
-		\githubpkg{dynverse}{dynwrap} & aut & Github & \notavailable & A common format for trajectories \\
+		\cranpkg{dynwrap} & aut & CRAN & ? & A common format for trajectories \\
 		\cranpkg{GillespieSSA} & aut & CRAN & 7546 & Gillespie's Stochastic Simulation Algorithm (SSA) \\

### 2019-10-12 ch1_introduction/content.tex
update

@@
 Characterising cells in a semi-quantifiable way is labour intensive, however; since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation).
-Modern implementations of IHC improve the throughput drastically by using robots and computer software to automate the image acquisition and cell segmentation.
-Nonetheless, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.
+Modern implementations of IHC improve the throughput drastically by using robots and computer software to provide semi-automated image acquisition and cell segmentation\cite{theodosiou_automatedanalysisfish_2007}.
 
-Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
+Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by running a steady stream of cells through a laser and measuring the amount of fluorescent light scattered from those cells.
 Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in the blood.

### 2019-10-12 ch1_introduction/content.tex
update

@@
 "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
-
 There are many types of "omics", but the most commonly used are the following.
@@
 
+% Increasing evidence further suggests that gene expression is heterogeneous, even in similar cell types [https://www.nature.com/articles/s12276-018-0071-8, cites 3 other papers as well]
+% Currently, however, the majority of transcriptome analysis experiments continue to be based on the assumption that cells from a given tissue are homogeneous, and thus, these studies are likely to miss important cell-to-cell variability.
+To better understand stochastic biological processes, a more precise understanding of the transcriptome in individual cells will be essential for elucidating their role in cellular functions and understanding how gene expression can promote beneficial or harmful states.
+
 \begin{figure}[htb!]
@@
 Transformative technological advances in microvolume sequencing allowed Tang et al.
-to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A).
-During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells and measuring other levels of information (e.g.
-proteomic expression levels and spatial location).
-% KDP: wat bedoel je met: such as proteomic expr levels
+to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}).
+During the decade that followed, the number of single-cell omics technologies has sky-rocketed, allowing to profile >100'000 cells\cite{svensson_exponentialscalingsinglecell_2018} and measuring other levels of information (e.g.
+protein abundance and spatial location) \cite{moudgil_multimodalscrnaseq_2019}.
 
@@
 	\centering
-	\includegraphics[width=\linewidth]{fig/singlecell_omics_evolution} 
+	\includegraphics[width=.7\linewidth]{fig/singlecell_omics_methodology} 
 	\caption{
-		\textbf{} %Convergence of "Omics" Biology and Single-Cell Biology.
-Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes.
-Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}.
+		\textbf{A:} Convergence of single-cell and omics biology\cite{junker_everycellspecial_2014}.
+		\textbf{B:} Different approaches for sampling cells with decreasing levels of cellular heterogeneity within the different sub-populations: snapshot, time-series, sorted.
+		\textbf{C:} Two common single-cell RNA sequencing technologies.
+Microfluidics systems let cells travel through nanometer scale tubing, capturing individual cells at intersections.
+Droplet-based systems encapsulate individual cells in droplets.
 	}
@@
 \end{figure}
+
+% could be expanded with information on how to isolate single cells and how to profile them
+% see https://www.nature.com/articles/s12276-018-0071-8
+% seems interesting as well: 10.1146/annurev-biodatasci-080917-013452
+
+In this work, unless noted otherwise, we will be working with transcriptomics data resulting from a single-cell RNA-sequencing experiment (scRNA-seq).
+The workflow of generating scRNA-seq profiles is as follows.
+Same as other single-cell (non-omics) profiling methods, cells first need to be isolated (Figure~\ref{fig:profiling_convergence}B).
+Different sampling techniques yield different levels of information about cellular state.
+By now, many protocols for extracting and tagging RNA from single cells have been developed\cite{svensson_exponentialscalingsinglecell_2018}, the most popular of which are based on microfluidics or droplets (Figure~\ref{fig:profiling_convergence}C).
+By sequencing the trancripts and the attached unique cell identifier tags, each read can be mapped and tallied up.
+scRNA-seq data can thus be summarised in a matrix, where each column represent a single cell, each row a gene, and each value represents the number of transcripts that were sequenced for that gene and cell.
 
@@
 The new types of analyses permitted by single-cell omics present several computational challenges\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
-This necessitates the development of novel computational tools, either because the problem statement of the performed analysis is completely novel, or to adapt existing methodology to new data characteristics (e.g.
-dimensionality and noise levels).
+This necessitates the development of novel computational tools, either because the problem statement of the performed analysis is completely novel, or to adapt existing methodology to new data characteristics -- dimensionality and noise.
+
+scRNA-seq data is typically very sparse -- while the human genome has more than 20'000 genes, they only contain non-zero values of a few thousand genes (typically <4'000).
+This is partially due to cells being specialised in particular functions and thus they do not need proteins of every time, but also due to RNA transcription occurring in bursts rather than continuously\cite{nicolas_whatshapeseukaryotic_2017}.
+
 Over the past five years, already 450 new tools for analysing single-cell omics data have been developed\cite{zappia_exploringsinglecellrnaseq_2018}.
@@
 Whole-genome profiling at a single-cell level harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
-% KDP: the new type of computational analysis etc... of eerder zeggen dat met de ontwikkeling van single cell technieken ook nieuwe computationele methoden moe(s)ten ontwikkeld worden
-(Figure \ref{fig:comp_tools})
-
-\mycomment{Todo: describe the characteristics of single cell omics data in the subsection on 'single cell omics'.
--> counts data.
-additional information from the cell can be present, such as the time of sampling in a time-series experiment or a cell type if the cells have been sorted using a FACS experiment.
-information on the features are typically restricted to the gene symbol, or an identifier of a xxx, such as ensembl or entrez.
-in comparison to bulk data, single-cell omics data is typically very sparse, due to technical and biological stochasticity.
-if several transcripts of a low abundance gene are not captured, this will be seen as a dropout.
-dropouts can also be biological, due to transcriptional bursting.
-}
 

### 2019-10-12 ch1_introduction/content.tex
update

@@
 Over the past five years, already 450 new tools for analysing single-cell omics data have been developed\cite{zappia_exploringsinglecellrnaseq_2018}.
-These tools typically fall within one of nine categories (Figure \ref{fig:comp_tools}).
+These tools typically fall within one of seven categories (Figure \ref{fig:comp_tools}).
 We discuss each category in the following subsections.
@@
 	\centering
-	\includegraphics[width=.75\linewidth]{fig/comptools/comptools_3.pdf} 
+	\includegraphics[width=\linewidth]{fig/comptools/comptools_4.pdf} 
 	\caption{
-		\textbf{Computational tools developed for processing single-cell omics data usually lie within one of nine categories.}
-		\textbf{A:} Normalisation, separating meaningful biological signal from technical noise or irrelevant biological information.
-		\textbf{B:} Data imputation, masking dropouts.
-		\textbf{C:} Data integration, combining multiple single-cell modalities.
-		\textbf{D:} Dimensionality reduction, providing a visual overview of the main cellular heterogeneity in the dataset.
-		\textbf{E:} Clustering, grouping together similar cells.
-		\textbf{F:} Trajectory inference, identifying and characterising transitions between different cellular states.
-		\textbf{G:} Network inference, predicting gene regulatory interactions from cell-to-cell heterogeneity.
-		\textbf{H:} RNA velocity, inferring the directionality of trajectories by predicting the future state of each cell.
-		\textbf{I:} Trajectory alignment, aligning multiple trajectories.
+		\textbf{Computational tools developed for processing single-cell omics data usually lie within one of seven categories.}
+		\textbf{A:} Example of a count matrix from a scRNA-seq dataset, showing only 100 genes (rows) and 100 cells (columns).
+		\textbf{B:} Dimensionality reduction, providing a visual overview of the main cellular heterogeneity in the dataset.
+		\textbf{C:} Clustering, grouping together cells with similar omics profiles.
+		\textbf{D:} Trajectory inference, identifying and characterising transitions between different cellular states.
+		\textbf{E:} Differential expression, \ldots.
+		\textbf{F:} Network inference, predicting gene regulatory interactions from cell-to-cell heterogeneity.
+		\textbf{G:} RNA velocity, inferring the directionality of trajectories by predicting the future state of each cell.
+		\textbf{H:} Trajectory alignment, aligning multiple trajectories.
 	}
@@
 \end{figure}
-
-\subsection{Normalisation}
-
-\subsection{Data imputation}
-
-\subsection{Data integration}
 
@@
 Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.
-
-\mycomment{This section is not very interestingly written}
 
@@
 \mycomment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}
+\subsection{Differential expression}
 

### 2019-10-12 ch6_bred/scripts/2_tcga.R
~

@@ -580,21 +580,45 @@ impsc2 <-
   ) %>%
   ungroup() %>%
   mutate(col = col_names[name]) %>%
-  select(regulator, name, target, everything())
+  select(regulator, name, target, everything()) %>%
+  arrange(desc(importance))
 plot(density(impsc2$importance))
 # impsc2f <- impsc2 %>% filter(importance_sc > 5)
-impsc2f <- impsc2 %>% group_by(name) %>% arrange(desc(importance)) %>% slice(1:50) %>% ungroup() %>% filter(importance > .2)
+impsc2f <- impsc2 %>% group_by(name) %>% slice(1:50) %>% ungroup() %>% filter(importance > .15)
 # impsc2f <- impsc2 %>% group_by(name) %>% arrange(desc(importance_sc)) %>% slice(1:50) %>% ungroup() %>% filter(importance > 1)
 impsc2f %>% group_by(name) %>% summarise(n = n()) %>% arrange(desc(n)) %>% print(n = Inf)
 write_tsv(impsc2f, paste0(data_dir, "grouped_interactions.tsv"))
 
 
 
+ggplot(impsc2 %>% filter(importance > .1)) +
+  geom_histogram(aes(importance, fill = name)) +
+  scale_fill_manual(values = set_names(col_names, cell_names))
 
 
+ggplot(impsc2 %>% filter(importance > .1)) +
+  geom_histogram(aes(log10(importance), fill = name)) +
+  scale_fill_manual(values = set_names(col_names, cell_names)) + facet_wrap(~name, scale = "free_y")
+
+ggplot(impsc2) +
+  geom_histogram(aes(log10(importance), fill = name)) +
+  scale_fill_manual(values = set_names(col_names, cell_names)) + facet_wrap(~name, scale = "free_y")
 
 
+ggplot(impsc2 %>% filter(importance > .1)) +
+  geom_point(aes(effect, importance, colour = name)) +
+  scale_colour_manual(values = set_names(col_names, cell_names)) + facet_wrap(~name, scale = "free_y")
 
 
+# individual interactions
+clus_ids <- labels %>% filter(name %in% c("AML", "ALL")) %>% pull(cluster)
+sel_ids <-
+  samples[clus %in% clus_ids] %>% sample(100)
 
-
+sample_network <-
+  importance_sc %>%
+  filter(cell_id %in% sel_ids) %>%
+  mutate(cluster = as.vector(clus[cell_id])) %>%
[…]

### 2019-10-13 ch1_introduction/content.tex
update changes

@@
 In many multicellular organisms, antigens can stimulate the immune system to produce antibodies.
-IHC realises the visualisation of proteins by exploiting the principle of antibodies binding specific to antigens.
+IHC realises the visualisation of proteins by exploiting the principle of antibodies binding to specific antigens.
 
@@
 Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by running a steady stream of cells through a laser and measuring the amount of fluorescent light scattered from those cells.
-Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in the blood.
-However, many protocols already exist to extract viable single cells from tissue as well\cite{leelatian_preparingviablesingle_2017}.
-Common flow cytometry enables to measure proteins expression levels of millions of cells for 8 different antibodies, though state-of-the-art instrumentation allow detecting up to 27 biomarkers simultaneously\cite{cossarizza_guidelinesuseflow_2017}.
+Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in blood.
+However, many protocols already exist to extract viable single cells from tissues and tumours\cite{leelatian_preparingviablesingle_2017}.
+Conventional flow cytometry devices enable to measure protein expression levels of millions of cells using up to eight different antibody fluorochromes simultaneously, while state-of-the-art instrumentation allows detection of up to 27 biomarkers simultaneously\cite{cossarizza_guidelinesuseflow_2017}.
 
@@
 mass cytometry, single-cell quantitative polymerase chain reaction, fluorescence \emph{in situ} hybridization).
-All of these single-cell (non-omics) technologies are limited by the number of different molecules they could measure, however.
+All of these single-cell (non-omics) technologies are limited by the number of different molecules they measure, however.
 Selecting molecules of interest prior to analysis, makes the experiment biased towards the preconceptions of the experimenter.
@@
 "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
-There are many types of "omics", but the most commonly used are the following.
-In genomics, all of an organism's genes are studied -- its whole genome.
-Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively.
+There are many types of "omics", such as genomics, transcriptomics and proteomics.
+Genomics studies the complete DNA sequence of an organism's genome, while transcriptomics and proteomics study the RNA transcripts and proteins, respectively.
 
-Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts.
-By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
+Specific examples of omics technologies are whole genome sequencing to determine the DNA sequence of an organism, and RNA sequencing to profiles the sequence of RNA transcripts, both using next-generation sequencing technologies.
+A gene expression profile can be obtained by mapping the sequences of RNA transcripts to the genome.
 
-Traditionally, profiling nano-litre volumes was infeasible.
-To capture enough material to generate a profile, an ensemble of cells needs to be pooled and lysed together, thereby granting the technology the name "bulk" omics.
-Bulk omics is a major workhorse in <...> and has applications in <...>. To capture enough material, an ensemble of cells is pooled and profiled all together.
+Several high-throughput technologies have been developed to investigate proteomes in depth.
+The most commonly applied are mass spectrometry-based and gel-based techniques (e.g.
+differential in-gel electrophoresis).
 
-\mycomment{TODO: fill in}
+% Je had proteomics technieken niet meer aangehaald.
+-> zie deze paperlink voor eventueel fig 1 of 2.
+Hoewel dit misschien niet je hoofdtopic is.
+https://academic.oup.com/chromsci/article/55/2/182/2333796
 
-A notable downside of bulk omics arises when the cells are biomolecularly heterogeneous (Figure~\ref{fig:bulk_vs_sc}A).
-Since the profile is a population average (or rather, a summation), information pertaining the individual profiles of cell subpopulations cannot be discerned (Figure~\ref{fig:bulk_vs_sc}B).
+Typically for these methods, to capture enough material to generate a profile, numerous cells need to be pooled and lysed together, thereby granting the technology's name "bulk" omics.
+Bulk omics is a major workhorse in molecular genetics and has applications in cancer research and in diagnostic screening of inheritable disorders.
 
-% Increasing evidence further suggests that gene expression is heterogeneous, even in similar cell types [https://www.nature.com/articles/s12276-018-0071-8, cites 3 other papers as well]
-% Currently, however, the majority of transcriptome analysis experiments continue to be based on the assumption that cells from a given tissue are homogeneous, and thus, these studies are likely to miss important cell-to-cell variability.
-To better understand stochastic biological processes, a more precise understanding of the transcriptome in individual cells will be essential for elucidating their role in cellular functions and understanding how gene expression can promote beneficial or harmful states.
+Increasing evidence shows that cells are biomoleculary heterogeneous, even in very similar cell types\cite{hwang_singlecellrnasequencing_2018} (Figure~\ref{fig:bulk_vs_sc}A).
+Since a bulk profile is a population average (or rather, a summation), important cell-to-cell variability is not discernible (Figure~\ref{fig:bulk_vs_sc}B).
 
@@
 \subsubsection{Single-cell omics}
-Transformative technological advances in microvolume sequencing allowed Tang et al.
-to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}).
+Comparing single-cell technologies with omics technologies shows that they have both clear advantages but also significant drawbacks (Figure \ref{fig:profiling_convergence}A).
+Single-cell biology allows profiling thousands or even millions of cells, but only for a select number of genes.
+On the other hand, omics biology provides a broader view -- since genes do not need to be selected beforehand -- but is a profile of ensemble of cells and thus masks important cellular heterogeneity.
+
+Advances in microvolume sequencing allowed profiling the transcriptome at single-cell resolution, thereby bringing single-cell biology and omics together to create single-cell omics.
 During the decade that followed, the number of single-cell omics technologies has sky-rocketed, allowing to profile >100'000 cells\cite{svensson_exponentialscalingsinglecell_2018} and measuring other levels of information (e.g.
@@
 	\centering
-	\includegraphics[width=.7\linewidth]{fig/singlecell_omics_methodology} 
+	\includegraphics[width=.7\linewidth]{fig/scomics_2.pdf} 
 	\caption{
@@
 The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
-Some of these opportunities lie in efforts like the Human Cell Atlas.
-The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories connecting the different cell types.
-As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells.
-% KDP: hier heb je het al eens over gehad: zou je dit niet samenbrengen?
+Some of these opportunities lie in efforts like the HCA consortium.
+They have set out to redefine all human cell types in both terms of their gene expression and location, and as well as the developmental trajectories connecting the different cell types.
+As part of this endeavour, the consortium will likely profile the whole transcriptomes of tens or even hundreds of millions of cells.
 

### 2019-10-13 ch1_introduction/content.tex
update

@@
 	\caption{
-		\textbf{A:} Convergence of single-cell and omics biology\cite{junker_everycellspecial_2014}.
+		\textbf{A:} Convergence of single-cell and omics biology.
 		\textbf{B:} Different approaches for sampling cells with decreasing levels of cellular heterogeneity within the different sub-populations: snapshot, time-series, sorted.
@@
 This is partially due to cells being specialised in particular functions and thus they do not need proteins of every time, but also due to RNA transcription occurring in bursts rather than continuously\cite{nicolas_whatshapeseukaryotic_2017}.
+This contributes to the high levels of noise seen in scRNA-seq data: no two cells have the same set of non-zero genes.
 
-Over the past five years, already 450 new tools for analysing single-cell omics data have been developed\cite{zappia_exploringsinglecellrnaseq_2018}.
-These tools typically fall within one of seven categories (Figure \ref{fig:comp_tools}).
-We discuss each category in the following subsections.
-Whole-genome profiling at a single-cell level harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
+Over the past five years, already 450 new tools have been developed to perform various analyses of single-cell omics data\cite{zappia_exploringsinglecellrnaseq_2018}, taking into account the specific noise characteristics.
+The most frequent types of analyses are detailed in the following subsections.
+
+\subsection{Dimensionality reduction}
+Single-cell omics datasets are usually one or more high-dimensional matrices, containing between M = $10^3$ to $10^5$ cells and typically about N = $10^3$ to $10^4$ genes (Figure~\ref{fig:comp_tools_dimred}A).
+The dimensionality of such datasets is typically too high for humans to interpret manually and for most modelling algorithms to tackle directly.
+Moreover, in reality, the intrinsic dimensionality of biological systems is probably much lower.
+For example, a differentiating hematopoietic cell could be described by just three dimensions: the first two dimensions lays out the hematopoietic lineage tree, and a third dimension allows for reprogramming between branches to occur.
+
+Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful low-dimensional representation.
+DR methods have two main target audiences; computers -- to construct a K-dimensional (with K $\ll$ M) representation of the data such that pairwise distances between different samples are retained as well as possible (Figure~\ref{fig:comp_tools_dimred}B); and humans -- to obtain a visual overview of the data (Figure~\ref{fig:comp_tools_dimred}C).
+
+Each dimension frequently corresponds to one or more modules of co-expressed genes.
+The reduced space can be interpreted in way analogous to Waddington’s epigenetic landscapes\cite{waddington_strategygenesdiscussion_1957,ferrell_bistabilitybifurcationswaddington_2012,rebhahn_animatedlandscaperepresentation_2014}, where the landscape dictates the possible states a cell can reside in, and transitions between states correspond to dynamic cellular processes, such as cell differentiation.
 
@@
 	\centering
-	\includegraphics[width=\linewidth]{fig/comptools/comptools_4.pdf} 
+	\includegraphics[width=\linewidth]{fig/comptools2/comptool_1_dimred.pdf} 
 	\caption{
-		\textbf{Computational tools developed for processing single-cell omics data usually lie within one of seven categories.}
-		\textbf{A:} Example of a count matrix from a scRNA-seq dataset, showing only 100 genes (rows) and 100 cells (columns).
-		\textbf{B:} Dimensionality reduction, providing a visual overview of the main cellular heterogeneity in the dataset.
-		\textbf{C:} Clustering, grouping together cells with similar omics profiles.
-		\textbf{D:} Trajectory inference, identifying and characterising transitions between different cellular states.
-		\textbf{E:} Differential expression, \ldots.
-		\textbf{F:} Network inference, predicting gene regulatory interactions from cell-to-cell heterogeneity.
-		\textbf{G:} RNA velocity, inferring the directionality of trajectories by predicting the future state of each cell.
-		\textbf{H:} Trajectory alignment, aligning multiple trajectories.
+		\textbf{Dimensionality reduction for single-cell omics data.}
+		\textbf{A:} A heatmap visualisation of an scRNA-seq expression dataset of fibroblasts being reprogrammed to neuron cells\cite{treutlein_dissectingdirectreprogramming_2016}. Only the most variable 
+		\textbf{B:} The reduced space is a M$\times$K-dimensional matrix which attempts to conserve the cellular heterogeneity of the original space as well as possible.
+		\textbf{C:} A dot plot of the first two components of the reduced space provides a global overview of the cells in the dataset.
+Colouring the dots according to prior information (e.g.
+cell type) or gene expression provides insight into the cellular heterogeneity within the dataset.
 	}
-	\label{fig:comp_tools}
+	\label{fig:comp_tools_dimred}
 \end{figure}
 
-\subsection{Dimensionality reduction}
-Single-cell omics datasets typically have too many dimensions (features) in order to be easily interpretable by humans and even by most computational tools. 
-Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful representation with fewer dimensions.
-It is important to note that its usage depends on the target audience: for humans -- to visualise data in a 2-D plane to aid with interpretation by humans, or for computers -- to construct a denser representation of the data such that it mostly contains the same information but with fewer dimensions.
+DR methods can be classified into two main categories, feature projection-based and manifold learning\cite{engel_surveydimensionreduction_2012}.
 
-There are many ways of classifying DR methods\cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
-Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible.
-Examples of commonly used projection-based DR methods in single-cell omics are PCA and MDS.
-Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g.
+Projection-based DR methods aim to perform a transformation of the data while preserving the pairwise distances between samples as much as possible.
+Examples of commonly used projection-based DR methods in single-cell omics are Principal Component Analysis\cite{pearson_liiilinesplanes_1901} (PCA) and Multi-Dimensional Scaling\cite{kruskal_multidimensionalscalingoptimizing_1964} (MDS).
+
+Manifold learning DR methods reconstruct a higher-order structure in the original space (e.g.
 a graph or a grid), visualising the structure in a lower-dimensional space, and mapping the original samples to the lower-dimensional space.
 Manifold learning can be an iterative optimisation process using a predefined criterion.
-Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.
+Examples of manifold learning techniques are t-distributed Stochastic Neighbor Embedding\cite{vandermaaten_visualizingdatausing_2008} (t-SNE), Diffusion Maps\cite{nadler_diffusion_2005,coifman_diffusionmaps_2006} and Uniform Manifold Approximation and Projection\cite{mcinnes_umapuniformmanifold_2018} (UMAP).
+
+For scalability reasons, this work mostly makes use of Landmark MDS\cite{silva_globallocalmethods_2002,lee_landmarkmdsensemble_2009} (LMDS) with a Spearman rank correlation distance metric.
+LMDS is an extension of classical MDS, but rather than calculating a complete distance matrix between all pairs of cells, a set of landmark cells is sampled, only the distances between a set of landmarks and the samples are calculated.
 
 \subsection{Clustering}
+To learn about the different cellular states within a population of cells, clustering methods divide up the cells into separate groups of highly similar cells (Figure~\ref{fig:comp_tools_cluster}A). 
+By visualising gene expression of known genes involved in the cell types of interest, the clusters can be annotated (Figure~\ref{fig:comp_tools_cluster}B).
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.6\linewidth]{fig/comptools2/comptool_2_clustering.pdf} 
+	\caption{
+		\textbf{Clustering for single-cell omics data.}
+		\textbf{A:} Clustering methods group cells with similar omics profiles together. 
+		\textbf{B:} By overlaying gene expression levels on a dimensionality reduction, the clusters can be annotated to allow better interpretation of the cellular heterogeneity.
+	}
+	\label{fig:comp_tools_cluster}
+\end{figure}
+
+Usually, the number of clusters is determined by the user, either as a direct parameter (e.g.
+$k$-means\cite{lloyd_leastsquaresquantization_1982}) or an indirect parameter (e.g.
+a height cutoff in hierarchical clustering).
+In some exceptional cases, the number of clusters is strictly determined by the data itself and cannot be altered with a parameter (e.g.
+Louvain clustering\cite{blondel_fastunfoldingcommunities_2008}).
+
+Clustering methods used in this work are mostly restricted to $k$-Means for clustering low-dimensional spaces and Louvain for clustering networks, since both are highly scalable with respect to the number of cells.
 
 \subsection{Trajectory inference}
-Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation\cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
-Trajectory inference (TI) is a new category of computational tools used to offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017,cannoodt_computationalmethodstrajectory_2016}.
-
-Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
-This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them\cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}.
-Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place\cite{cannoodt_computationalmethodstrajectory_2016}.
-TI has two objectives: to reconstruct the topology of the dynamic process (e.g.
-is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology.
-Some TI methods assumes that the user knows the topology beforehand and only focuses on ordering the cells along a predefined topology.
-
-The dataset can be a single snapshot of a mixture of cells in different stages, or a set of samples collected at different time points (Figure \ref{fig:trajectory_inference}A).
-Typically, TI methods first analyse similarities between cells, optionally infer the topology of the underlying process, and finally order cells along that trajectory (Figure \ref{fig:trajectory_inference}B).
-The second step can be optional, as some methods assume a specific topology beforehand.
-TI methods allow the identification of new subsets of cells, delineation of a differentiation tree, and characterisation of the main driver genes along a state transition (Figure \ref{fig:trajectory_inference}C).
-Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms\cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable\cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
+While clustering methods divide cells into distinct groups, trajectory inference (TI) methods acknowledge that cells are dynamic entities which transition from once cellular state to another via various dynamic processes.
+Rather than making distinct groups, TI methods allow studying dynamic processes by reconstructing the topology of a dynamic process as a trajectory, and map the cells onto that trajectory (Figure~\ref{fig:comp_tools_ti}A).
+Similar to clustering, by colouring the cells according to the expression of genes known to be involved in the dynamic process of interest, the milestones in the trajectory can be annotated (Figure~\ref{fig:comp_tools_cluster}B).
 
 \begin{figure}[htb!]
-	\centering\includegraphics[width=\Largefigure]{fig/trajectory_inference_slim} % TODO: Update figure to match text!!!!
+	\centering
+	\includegraphics[width=.6\linewidth]{fig/comptools2/comptool_3_trajectoryinference.pdf} 
 	\caption{
-		Applications of single-cell trajectory inference methods.
-(A) Single-cell omics data appropriate for TI can be both obtained from an unsynchronised population of single cells (snapshot data) but also from synchronised cell populations.
-(B) UPDATE!
-(C) UPDATE!
+		\textbf{Trajectory inference for single-cell omics data.}
+		\textbf{A:} During a dynamic process cells pass through several transitional states, characterized by different waves of transcriptional, morphological, epigenomic and/or surface marker changes\cite{enver_stemcellstates_2009}.
+TI methods provide a straightforward and unbiased approach to identifying and correctly ordering different transitional stages.
+		\textbf{B:} By overlaying gene expression levels on a dimensionality reduction, the milestones can be annotated to allow better interpretation of the cellular heterogeneity.
 	}
-	\label{fig:trajectory_inference}
+	\label{fig:comp_tools_ti}
 \end{figure}
 
-\mycomment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}
+TI methods typically first require some form of dimensionality reduction prior to inferring a trajectory from the dataset (Figure~\ref{fig:ti_metro}).
+Subsequently, a trajectory modelling step then operates in this reduced space, aiming to identify cell states, constructing a trajectory through the different states, and projecting the cells back on to the trajectory.
+
+A common way to classify TI methods is by the types of trajectories they can infer\cite{saelens_comparisonsinglecelltrajectory_2019}.
+About half of TI methods specialise in inferring linear or cyclic trajectories (i.e.
+they order the cells).
+Others model the trajectory as a rooted tree, allowing for one or more bifurcations to occur.
+Only a few methods are able to infer more generalised trajectories containing disconnected subgraphs or cycles.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.9\linewidth]{fig/ti_metro} 
+	\caption{
+		\textbf{TI methods use several common building blocks and can be organized in a unifying modular framework.} Overall, TI consists of two steps.
+In the first step dimensionality reduction techniques such as manifold learning, clustering, or graph-based methods are used to convert the dataset to a more simplified representation.
+This representation of the data then allows the trajectory itself to be more easily modelled in a second step.
+In this step, the trajectory itself is found within the data using both graph-based and curve-based approaches, after which the cells themselves can be ordered using a variety of methods.
+	}
+	\label{fig:ti_metro}
+\end{figure}
+
 \subsection{Differential expression}
+Given that cells are split up into groups differential expression (DE) methods ranks genes based on whether their expression is significantly higher or lower in one group in comparison to the others.
+This grouping can be based on prior information or an upstream clustering method.
+DE methods are useful for summarising the main differences between different groups of cells more compactly (Figure~\ref{fig:comp_tools_de}A) in comparison to when groups are compared without gene prioritisation (Figure~\ref{fig:comp_tools_cluster}A).
+
+Trajectory differential expression (TDE) is an extension of DE where instead genes are prioritised according to whether their gene expression changes smoothly but significantly along a parts of a trajectory (Figure~\ref{fig:comp_tools_de}).
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.6\linewidth]{fig/comptools2/comptool_4_differentialexpression.pdf} 
+	\caption{
+		\textbf{Differential expression for single-cell omics data.}
+		\textbf{A:} Differential expression methods prioritises genes that are expressed significantly higher or lower in particular given groups.
+		\textbf{B:} Trajectory differential expression prioritises genes that change smoothly but significantly along particular transitions in a trajectory.
+	}
+	\label{fig:comp_tools_de}
+\end{figure}
 
 \subsection{Network inference}
+%One of the central cellular processes underlying development is transcriptional regulation.
+During development, changes in transcription factor activity induce chromatin modifications, chromatin remodeling and ultimately a differential recruitment of the basal transcriptional machinery [66].
+Modeling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
+%
+%The dynamics of gene regulation has classically been studied using time series data [23].
+When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population [67–70].
+Alternatively, time series data can also be generated by synchronizing the dynamic process between cells.
+However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed [23].
+Combining single-cell snapshot data with trajectory inference can now in principle construct an accurate time-series dataset for every cell population both in vivo and in vitro, while avoiding issues with heterogeneity.
+Initially, several pioneering studies demonstrated the potential of this approach in single cells using single-cell qPCR data for selected sets of 10 to 50 transcription factors [71–75].
+With the advent of genome wide single-cell technologies, the models of dynamic gene regulation can now be made even more complete while avoiding a prior bias to well studied transcription factors [17,30,76,77] and several methods have already been developed specifically tuned towards single-cell data [78–80].
+%
+%The power of a dynamic network approach in single cells is not only that it allows the identification of transcription factors specifically regulating a specific state, but also how these factors then induce sets of genes necessary for regulating the next wave of transcription changes (Figure 1B).
+Such a gene regulatory cascade was described during cellular reprogramming from mouse embryonic fibroblasts to induced pluripotent stem cells, where SOX2 was shown to activate a series of other transcription factors which, together with SOX2, induces pluripotency [72].
+Transition states are frequently associated with particular subnetworks of transcriptional regulators, as was shown in a direct reprogramming setting from fibroblasts to neurons [77].
+Several studies have highlighted how some regulatory links can be very dynamic while others show evidence of being static during consecutive developmental stages [74,75].
+Regulatory network motifs can also be extracted from the regulatory network [74], some of which can have interesting properties in dynamic systems [81,82].
+In branching trajectories dynamic regulatory networks can also be used to study the regulators important for a particular branch [17,75,79] (Figure 1C).
+%
+%Another advantage of using ordered datasets, either based on bulk time-series or single-cell trajectories, is the improved accuracy of inferring true causative regulatory interactions.
+Assume that during development the expression of regulator A increases and in turn this regulator activates the expression of regulator B and gene C.
+Between the upregulation of regulator A and the upregulation of B and C we expect a time delay, equal to the time it takes for regulator A to be translated, translocated to the nucleus and induce a detectable increase in the expression of B and C.
+On the other hand the expression of regulator B and gene C will be very strongly correlated, because they are both activated at the same time.
+Incorporating this time-delay into the gene regulatory network inference could therefore be used to better discern the true causative regulatory edge between A and B/C with a spurious correlative edge between B and C [83].
+
 Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes.

### 2019-10-13 ch1_introduction/analysis/computational_tools_2.R

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(dyno)
 
 fig_size <- 3
 
@@ -24,115 +25,101 @@ raster <- function(meltdf, legend = FALSE, palette = "RdYlBu") {
 }
 
 
-# Generate toy datasets ---------------------------------------------------
+# Preprocess dataset ------------------------------------------------------
 set.seed(1)
+dat <- dynbenchmark::load_dataset("real/silver/fibroblast-reprogramming_treutlein")
+dat$expression <- dat$expression()
+dat$counts <- dat$counts()
+expr <- dat$expression
 
-# linear
-linear <- dyntoy::generate_dataset(
-  model = "linear",
-  num_cells = 60,
-  num_features = 40,
-  dropout_probability_factor = 0
-)
+gix <- apply(expr, 2, var) %>% order(decreasing = TRUE) %>% head(100)
+exprf <- expr[, gix]
 
-expr_li <- as.matrix(linear$expression) %>% dynutils::scale_quantile() %>% t() %>% dynutils::scale_quantile()
+exprsc <- exprf %>% as.matrix() %>% dynutils::scale_quantile() %>% t() %>% dynutils::scale_quantile() %>% t()
 
-dropout_weight <- 1 / (as.vector(expr_li)+1)
-rm_ix <- sample(seq_along(dropout_weight), length(dropout_weight) * .2, prob = dropout_weight, replace = FALSE)
+hclr <- hclust(dynutils::calculate_distance(exprf, method = "spearman") %>% as.dist())
+ordr <- hclr$order
+ordc <- hclust(dynutils::calculate_distance(Matrix::t(exprf), method = "spearman") %>% as.dist())$order
 
-expr_li_nodrop <- expr_li
-expr_li[rm_ix] <- 0
-
-ordr_li <- hclust(dist(expr_li))$order
-ordc_li <- hclust(dist(t(expr_li)))$order
-
-# branching
-set.seed(2)
-branching <- dyntoy::generate_dataset(
-  model = dyntoy::model_binary_tree(num_branchpoints = 1L),
-  num_cells = 100,
-  num_features = 100,
-  allow_tented_progressions = FALSE,
-  sample_mean_count = function() rnorm(1, 1000, 100),
[…]

### 2019-10-13 ch1_introduction/content.tex
update intro

@@
 \subsection{Network inference}
-%One of the central cellular processes underlying development is transcriptional regulation.
-During development, changes in transcription factor activity induce chromatin modifications, chromatin remodeling and ultimately a differential recruitment of the basal transcriptional machinery [66].
+One of the central cellular processes underlying development is transcriptional regulation. 	
 Modeling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
-%
-%The dynamics of gene regulation has classically been studied using time series data [23].
-When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population [67–70].
-Alternatively, time series data can also be generated by synchronizing the dynamic process between cells.
-However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed [23].
-Combining single-cell snapshot data with trajectory inference can now in principle construct an accurate time-series dataset for every cell population both in vivo and in vitro, while avoiding issues with heterogeneity.
-Initially, several pioneering studies demonstrated the potential of this approach in single cells using single-cell qPCR data for selected sets of 10 to 50 transcription factors [71–75].
-With the advent of genome wide single-cell technologies, the models of dynamic gene regulation can now be made even more complete while avoiding a prior bias to well studied transcription factors [17,30,76,77] and several methods have already been developed specifically tuned towards single-cell data [78–80].
-%
-%The power of a dynamic network approach in single cells is not only that it allows the identification of transcription factors specifically regulating a specific state, but also how these factors then induce sets of genes necessary for regulating the next wave of transcription changes (Figure 1B).
-Such a gene regulatory cascade was described during cellular reprogramming from mouse embryonic fibroblasts to induced pluripotent stem cells, where SOX2 was shown to activate a series of other transcription factors which, together with SOX2, induces pluripotency [72].
-Transition states are frequently associated with particular subnetworks of transcriptional regulators, as was shown in a direct reprogramming setting from fibroblasts to neurons [77].
-Several studies have highlighted how some regulatory links can be very dynamic while others show evidence of being static during consecutive developmental stages [74,75].
-Regulatory network motifs can also be extracted from the regulatory network [74], some of which can have interesting properties in dynamic systems [81,82].
-In branching trajectories dynamic regulatory networks can also be used to study the regulators important for a particular branch [17,75,79] (Figure 1C).
-%
-%Another advantage of using ordered datasets, either based on bulk time-series or single-cell trajectories, is the improved accuracy of inferring true causative regulatory interactions.
-Assume that during development the expression of regulator A increases and in turn this regulator activates the expression of regulator B and gene C.
-Between the upregulation of regulator A and the upregulation of B and C we expect a time delay, equal to the time it takes for regulator A to be translated, translocated to the nucleus and induce a detectable increase in the expression of B and C.
-On the other hand the expression of regulator B and gene C will be very strongly correlated, because they are both activated at the same time.
-Incorporating this time-delay into the gene regulatory network inference could therefore be used to better discern the true causative regulatory edge between A and B/C with a spurious correlative edge between B and C [83].
 
-Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes.
-This topic already received much attention with the advent of bulk omics (before single-cell omics).
-These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods\cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.
+Network inference (NI) methods predict which genes are regulated by which transcription factors (Figure~\ref{fig:comp_tools_ni}).
+Before single-cell omics, these methods rely on multiple experiments, amongst which perturbation and time-series experiments, to predict the effect each transcription factor has on the up- or downregulation of a gene.
 
-After the last DREAM competition, it seemed that interest in NI methodology had declined.
-After all, NI on bulk omics profiles suffered from several crucial issues.
-As mentioned previously, bulk profiles are generated by pooling together the RNA transcripts of a supposedly homogeneous population of thousands of cells.
-Since the expression values are averaged over the whole population, incorrect assumptions on the homogeneity of the pooled cells may lead to the masking of relevant expression patterns in rare cell populations (Figure \ref{fig:bulk_vs_singlecell}).
-Besides, NI methods rely on a diverse set of time-series and perturbation experiments in order to reliably identify causal regulatory interactions.
-Such experiments are expensive and time-consuming, and an inaccurate selection of time points might result in crucial intermediate stages being missed.
+One of the main advantages of single-cell omics is the heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs.
 
@@
 	\centering
-	\includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
-	\caption{Bulk expression data return the average expressions of genes among large numbers of cells.
-In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required.
-On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: Already used this figure in the previous section
-	\label{fig:bulk_vs_singlecell}
+	\includegraphics[width=.6\linewidth]{fig/comptools2/comptool_5_networkinference.pdf}
+	\caption{
+		\textbf{Network inference for single-cell omics data.}
+	}
+	\label{fig:comp_tools_ni}
 \end{figure}
-
-The advent of single-cell omics has made scientists wonder whether now is the time to revisit network inference\cite{stegle_computationalanalyticalchallenges_2015}. 
-One of the main advantages of single-cell omics is the ability to quantify the exact cellular state of thousands of cells per experiment.
-The heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs (see Figure \ref{fig:bulk_vs_singlecell}).
-In this setting, heterogeneity in the cell population eases network inference, rather than mask condition-specific expression patterns in regulatory interactions.
-
-% mention differential networks
-
-\subsection{RNA velocity}
-
-\subsection{Trajectory alignment}
 
@@
 	\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
+	% TODO: fill this part in!!!
 	\item Every NI method has certain topological biases.

### 2019-10-13 ch1_introduction/content.tex
update

@@
 
-%library(dlstats)
-%library(tidyverse)
-%
-%cran <- c("babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "lmds", "princurve", "proxyC", "qsub", "SCORPIUS", "badger", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
-%bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
-%
-%dl_cran <- dlstats::cran_stats(cran)
-%dl_bioc <- dlstats::bioc_stats(bioc)
-%
-%dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>% 
-%mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
-%
-%min_date <- "2019-08-01"
-%num_days <- as.integer(Sys.Date() - as.Date(min_date))
-%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% as.data.frame
-
 \begin{table}[ht!]
@@
 CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
+In addition, many of the dynverse packages have only recently been published on CRAN.
 For Github repositories, no download statistics could be retrieved.
@@
 		Name & Role & Host & Downloads per year & Description \\ \hline\hline
-		\cranpkg{babelwhale} & aut & CRAN & 3996 & Interacting with Docker and Singularity containers \\
-		\cranpkg{diffusionMap} & aut & CRAN & 21'361 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
+		\cranpkg{babelwhale} & aut & CRAN & 5810 & Interacting with Docker and Singularity containers \\
+		\cranpkg{diffusionMap} & aut & CRAN & 30'450 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
 		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
-		\cranpkg{dyndimred} & aut & CRAN & 5511 & Applying dimensionality reduction methods \\
+		\cranpkg{dyndimred} & aut & CRAN & 5535 & Applying dimensionality reduction methods \\
 		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
@@
 		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
-		\cranpkg{dynparam} & aut & CRAN & 3084 & Creating meta-information for parameters \\
+		\cranpkg{dynparam} & aut & CRAN & 3265 & Creating meta-information for parameters \\
 		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
@@
 		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
-		\cranpkg{dynutils} & aut & CRAN & 5657 & Common functionality for the dynverse packages \\
-		\cranpkg{dynwrap} & aut & CRAN & ? & A common format for trajectories \\
-		\cranpkg{GillespieSSA} & aut & CRAN & 7546 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
-		\cranpkg{GillespieSSA2} & aut & CRAN & 6506 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
+		\cranpkg{dynutils} & aut & CRAN & 13'13' & Common functionality for the dynverse packages \\
+		\cranpkg{dynwrap} & aut & CRAN & 990 & A common format for trajectories \\
+		\cranpkg{GillespieSSA} & aut & CRAN & 7880 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		\cranpkg{GillespieSSA2} & aut & CRAN & 4950 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
 		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & An Rcpp implementation of the Growing Neural Gas algorithm \\
-		\cranpkg{incgraph} & aut & CRAN & 3175 & Incremental graphlet counting for network optimisation \\
-		\cranpkg{lmds} & aut & CRAN & \notavailable & Landmark Multi-Dimensional Scaling \\
-		\cranpkg{princurve} & aut & CRAN & 26'991 & Fits a principal curve in arbitrary dimension \\
-		\cranpkg{proxyC} & aut & CRAN & 117'484 & Computes proximity in large sparse matrices \\
-		\cranpkg{qsub} & aut & CRAN & 3193 & Running commands remotely on gridengine clusters \\
-		\cranpkg{SCORPIUS} & aut & CRAN & 4772 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
+		\cranpkg{incgraph} & aut & CRAN & 3565 & Incremental graphlet counting for network optimisation \\
+		\cranpkg{lmds} & aut & CRAN & 815 & Landmark Multi-Dimensional Scaling \\
+		\cranpkg{princurve} & aut & CRAN & 29'100 & Fits a principal curve in arbitrary dimension \\
+		\cranpkg{proxyC} & aut & CRAN & 117'480 & Computes proximity in large sparse matrices \\
+		\cranpkg{qsub} & aut & CRAN & 3585 & Running commands remotely on gridengine clusters \\
+		\cranpkg{SCORPIUS} & aut & CRAN & 4520 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
 
-		\cranpkg{badger} & ctb & CRAN & ? & Query information and generate badge for using in README and GitHub Pages \\
-		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
-		\cranpkg{devtools} & ctb & CRAN & 3'775'350 & Tools to make developing R packages easier \\
+		\cranpkg{badger} & ctb & CRAN & 6240 & Query information and generate badge for using in README and GitHub Pages \\
+		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 935 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
+		\cranpkg{devtools} & ctb & CRAN & 5'918'700 & Tools to make developing R packages easier \\
 		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
-		\cranpkg{mlr} & ctb & CRAN & 142'605 & Machine Learning in R \\
-		\biocpkg{monocle} & ctb & Bioc & 35'240 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
-		\cranpkg{ParamHelpers} & ctb & CRAN & 109'408 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
+		\cranpkg{mlr} & ctb & CRAN & 176'330 & Machine Learning in R \\
+		\biocpkg{monocle} & ctb & Bioc & 34'360 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
+		\cranpkg{ParamHelpers} & ctb & CRAN & 150'775 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
 		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
 		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
-		\cranpkg{remotes} & ctb & CRAN & 3'704'594 & R package installation from remote repositories, including GitHub \\
-		\cranpkg{rlang} & ctb & CRAN & 11'470'763 & Functions for base types and core R and tidyverse features \\
+		\cranpkg{remotes} & ctb & CRAN & 3'944'090 & R package installation from remote repositories, including GitHub \\
+		\cranpkg{rlang} & ctb & CRAN & 13'269'115 & Functions for base types and core R and tidyverse features \\
 		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
-		\biocpkg{slingshot} & ctb & Bioc & 11'643 & Tools for ordering single-cell sequencing \\
-		\biocpkg{splatter} & ctb & Bioc & 3741 & Simple simulation of single-cell RNA sequencing data \\
+		\biocpkg{slingshot} & ctb & Bioc & 12'085 & Tools for ordering single-cell sequencing \\
+		\biocpkg{splatter} & ctb & Bioc & 5015 & Simple simulation of single-cell RNA sequencing data \\
 		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\

### 2019-10-13 ch1_introduction/analysis/computational_tools_2.R

@@ -377,7 +377,7 @@ fimp <- dynfeature::calculate_branch_feature_importance(
   expression_source = dat$expression
 )
 
-de_traj <- fimp_clus %>% group_by(from, to) %>% slice(1:10) %>% ungroup() %>% pull(feature_id) %>% unique()
+de_traj <- fimp %>% group_by(from, to) %>% slice(1:13) %>% ungroup() %>% pull(feature_id) %>% unique()
 expr_de <-
   expr[, de_traj] %>%
   dynutils::scale_quantile() %>% t() %>% dynutils::scale_quantile() %>% t()

### 2019-10-13 ch1_introduction/analysis/download_stats.R

@@ -1,0 +1,20 @@
+library(dlstats)
+library(tidyverse)
+
+cran <- c(
+  "babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "dynwrap",
+  "GillespieSSA", "GillespieSSA2", "incgraph", "lmds", "princurve", "proxyC",
+  "qsub", "SCORPIUS", "badger", "devtools", "mlr", "ParamHelpers", "remotes", "rlang"
+)
+bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
+
+dl_cran <- dlstats::cran_stats(cran)
+dl_bioc <- dlstats::bioc_stats(bioc)
+
+dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>%
+  mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
+
+min_date <- "2019-08-01"
+num_days <- as.integer(Sys.Date() - as.Date(min_date))
+summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365)
+summ %>% print(n = 50)

### 2019-10-13 ch6_bred/scripts/2_tcga.R

@@ -321,14 +321,47 @@ if (!file.exists(paste0(data_dir, "qsub_handle.rds"))) {
     sigmoid_sd = sigmoid_sd
   )
   write_rds(qsub_handle, paste0(data_dir, "qsub_handle.rds"))
+
+  targets_filt <- targets[2001:4000]
+  expr_filt <- expr[, union(targets_filt, regulators_filt)]
+
+  x <- 1
+  qsub_handle2 <- qsub::qsub_lapply(
+    X = seq_along(targets_filt),
+    qsub_config = qsub::override_qsub_config(
+      max_wall_time = "12:00:00",
+      memory = "10G",
+      name = "bred",
+      wait = FALSE,
+      stop_on_error = FALSE,
+      remove_tmp_folder = FALSE#,
+      # execute_before = "#$ -l h=!prismcls08",
+    ),
+    qsub_packages = c("dynutils", "dplyr", "purrr", "magrittr", "tibble"),
+    qsub_environment = c("x"),
+    FUN = calculate_target_importance,
+    # pass data and other parameters
+    expr = expr_filt,
+    samples = samples_filt,
+    regulators = regulators_filt,
+    targets = targets_filt,
+    num_trees = num_trees,
+    num_variables_per_split = num_variables_per_split,
+    min_node_size = min_node_size,
+    max_depth = max_depth,
+    interaction_importance_filter = interaction_importance_filter,
+    sigmoid_mean = sigmoid_mean,
+    sigmoid_sd = sigmoid_sd
+  )
+  write_rds(qsub_handle2, paste0(data_dir, "qsub_handle2.rds"))
 }
 
 
 # if (!file.exists(paste0(data_dir, "grn.rds"))) {
-  qsub_handle <- read_rds(paste0(data_dir, "qsub_handle.rds"))
+  qsub_handle1 <- read_rds(paste0(data_dir, "qsub_handle.rds"))
 
-  grn <- qsub::qsub_retrieve(
-    qsub_handle,
+  grn1 <- qsub::qsub_retrieve(
+    qsub_handle1,
     wait = "just_do_it",
     post_fun = function(i, li) {
       li$importance <- li$importance %>% filter(importance > .01)
@@ -337,6 +370,21 @@ if (!file.exists(paste0(data_dir, "qsub_handle.rds"))) {
       li
     }
   )
+
[…]

### 2019-10-14 boek.tex
update tex
minor changes
clean up thesis
update

@@
 \usepackage{algpseudocode}
+\usepackage{algpseudocode}
+\MakeRobust{\Call}
 
@@
 % references packages
-\usepackage[sorting=none]{biblatex}
+\usepackage[sorting=none,refsection=chapter]{biblatex}
 \bibliography{library}
@@
 \vfill
-Partially adapted from:\\
-\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y.
-Computational methods for trajectory inference from single-cell transcriptomics.
-\textit{European Journal of Immunology} 46, 11 (2016), 2496--2506.
-\doi{10.1002/eji.201646347}.\\
-Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
-A comparison of single-cell trajectory inference methods.
-\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
-\doi{10.1038/s41587-019-0071-9}.\\
-Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y.
-Network Inference from Single-Cell Transcriptomic Data.
-\textit{Gene Regulatory Networks} (2019), 235--249.
-\doi{10.1007/978-1-4939-8882-2_10}.
-\\
-{\footnotesize $^*$ Equal contribution}
 \newpage
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%         CH3B: DYNO         %%%
+%%%         CH4: DYNO         %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Each selected method can then be easily run within a common interface.
-Because the output of these methods are all converted into a common consistent format, the trajectories can be easily interpreted.
+Because the outputs of these methods are all converted into a common consistent format, the trajectories can be easily interpreted.
 For this, we provide a complete toolkit: adding a root and labelling the different cell stages, detecting genes which are differentially expressed at different stages of the trajectory, and plotting the trajectory within any dimensionality reduction or heatmap.
-
-	We are currently extending this pipeline to include additional input data, such as RNA velocity measurements, and additional tools for interpreting the trajectories, such as alignment and differential expression.
 \end{quote}
@@
 \newpage
-\graphicspath{{./ch3b_dyno/}}
-\include{ch3b_dyno/content}
+\graphicspath{{./ch4_dyno/}}
+\include{ch4_dyno/content}
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH4: SCORPIUS       %%%
+%%%       CH5: SCORPIUS       %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newpage
-\graphicspath{{./ch4_scorpius/}}
-\include{ch4_scorpius/content}
+\graphicspath{{./ch5_scorpius/}}
+\include{ch5_scorpius/content}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%         CH5: BRED         %%%
+%%%         CH6: BRED         %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Inferring Single Cell Regulatory Networks. \textit{Journal} vol, issue (2019), page--page. doi.
+\textbf{Cannoodt, R.}, Saelens, W., Saeys, Y, and De Preter K. Inferring Single Cell Regulatory Networks. \textit{Journal} vol, issue (2019), page--page. doi.
 \newpage
-\include{ch5_bred/content}
+\graphicspath{{./ch6_bred/}}
+\include{ch6_bred/content}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH5B: INCGRAPH       %%%
+%%%       CH7: INCGRAPH       %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{chap:incgraph}
-\graphicspath{{./ch5b_incgraph/}}
+\graphicspath{{./ch7_incgraph/}}
 \begin{quote}
@@
 	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
-\\
-	IncGraph is freely available as an open-source R package on CRAN (incgraph). The development version is also available on GitHub (\href{http://www.github.com/rcannood/incgraph}{rcannood/incgraph}).
 \end{quote}
@@
 \newpage
-\include{ch5b_incgraph/content}
+\include{ch7_incgraph/content}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%      CH6: DISCUSSION      %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Discussion} 
-\chaptermark{Discussion.} % short title
-\label{chap:discussion}
-\graphicspath{{./ch6_discussion/}}
-\include{ch6_discussion/content}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%   CH7: Self-assessment    %%%
+%%%   CH9: Self-assessment    %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{chap:selfassessment}
-\graphicspath{{./ch7_selfassessment/}}
-%\begin{quote}
-%	\textbf{Abstract:}
-%\end{quote}
-%\vfill
-%Adapted from:\\
-%Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
-%Boulesteix, A., Saeys, Y., and Robinson, M. D.. 
-%Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.
+\graphicspath{{./ch9_selfassessment/}}
+\begin{quote}
+	\textbf{Abstract:}
+\end{quote}
+\vfill
+Adapted from:\\
+\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Self-assessment in trajectory inference. \textit{Journal} vol, issue (2019), page--page. doi.
 \newpage
-\include{ch7_selfassessment/content}
+\include{ch9_selfassessment/content}
+\clearpage
+\section{References}
+\printbibliography[heading=none]
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH8: Guidelines       %%%
+%%%       CH10: Guidelines       %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{chap:guidelines}
-\graphicspath{{./ch8_guidelines/}}
+\graphicspath{{./ch10_guidelines/}}
 \begin{quote}
@@
 \newpage
-\include{ch8_guidelines/content}
+\include{ch10_guidelines/content}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%      CH8: DISCUSSION      %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{Discussion} 
+\chaptermark{Discussion.} % short title
+\label{chap:discussion}
+\graphicspath{{./ch8_discussion/}}
+\include{ch8_discussion/content}
 

### 2019-10-14 ch1_introduction/content.tex
update

@@
 
-% KDP: je zegt niets over epigenetica, die min of meer bepaald welke gebieden in het DNA worden vertaald (dit leidt je misschien te ver...)
 During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA.
@@
 \subsection{Cell types}
-\textit{Homo sapiens} like to categorise everything they encounter, and so too have they conceptualised groups of cells called "cell types".
-The human body contains more than 200 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous.
+Ever since Robert Hook first described the different structures of cells in 1665, biologists have been classifying cells by form and function.
+The human body is said to contain more than 210 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous.
 This however, is a major underestimation of the real number of cell types.
@@
 
-Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by running a steady stream of cells through a laser and measuring the amount of fluorescent light scattered from those cells.
+Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by measuring fluorescently labelled proteins as cells pass through a fluidic system.
 Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in blood.
@@
 "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
-There are many types of "omics", such as genomics, transcriptomics and proteomics.
+There are at least ten types of "omics".
+In this work, we mostly consider genomics, transcriptomics, proteomics, and regulomics.
 Genomics studies the complete DNA sequence of an organism's genome, while transcriptomics and proteomics study the RNA transcripts and proteins, respectively.
+Regulomics studies the regulatory molecules (e.g.
+genes, RNAs, proteins) which play a role in determining gene regulation.
 
@@
 The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
-Some of these opportunities lie in efforts like the HCA consortium.
+Some of these opportunities lie in efforts like the HCA consortium\cite{regev_humancellatlas_2018}.
 They have set out to redefine all human cell types in both terms of their gene expression and location, and as well as the developmental trajectories connecting the different cell types.
-As part of this endeavour, the consortium will likely profile the whole transcriptomes of tens or even hundreds of millions of cells.
+As part of this endeavour, the consortium will likely profile the whole transcriptomes of tens or even hundreds of millions of cells\cite{humancellatlasconsortium_humancellatlas_2018}.
 
@@
 While clustering methods divide cells into distinct groups, trajectory inference (TI) methods acknowledge that cells are dynamic entities which transition from once cellular state to another via various dynamic processes.
-Rather than making distinct groups, TI methods allow studying dynamic processes by reconstructing the topology of a dynamic process as a trajectory, and map the cells onto that trajectory (Figure~\ref{fig:comp_tools_ti}A).
+Rather than making distinct groups, TI methods allow studying dynamic processes by reconstructing the topology of a dynamic process as a trajectory, and map the cells onto that trajectory.
+A trajectory is a graph where the nodes represent noteworthy cellular states, and each cell is predicted to be progressing along transitions between the different states (Figure~\ref{fig:comp_tools_ti}A).
+
+A trajectory can be visualised as a graph to better highlight the topology of the trajectory(Figure~\ref{fig:comp_tools_ti}A middle), as a heatmap to better depict the changes in gene expression along the different transitions (Figure~\ref{fig:comp_tools_ti}A right), or may be embedded in a dimensionality reduction of the cells to better demonstrate cellular heterogeneity along the trajectory (Figure~\ref{fig:comp_tools_ti}B right).
 Similar to clustering, by colouring the cells according to the expression of genes known to be involved in the dynamic process of interest, the milestones in the trajectory can be annotated (Figure~\ref{fig:comp_tools_cluster}B).
@@
 	\centering
-	\includegraphics[width=.6\linewidth]{fig/comptools2/comptool_3_trajectoryinference.pdf} 
+	\includegraphics[width=.9\linewidth]{fig/comptools2/comptool_3_trajectoryinference.pdf} 
 	\caption{
@@
 		\textbf{A:} During a dynamic process cells pass through several transitional states, characterized by different waves of transcriptional, morphological, epigenomic and/or surface marker changes\cite{enver_stemcellstates_2009}.
-TI methods provide a straightforward and unbiased approach to identifying and correctly ordering different transitional stages.
+TI methods provide an unbiased approach to identifying and correctly ordering different transitional stages.
 		\textbf{B:} By overlaying gene expression levels on a dimensionality reduction, the milestones can be annotated to allow better interpretation of the cellular heterogeneity.
@@
 
-TI methods typically first require some form of dimensionality reduction prior to inferring a trajectory from the dataset (Figure~\ref{fig:ti_metro}).
-Subsequently, a trajectory modelling step then operates in this reduced space, aiming to identify cell states, constructing a trajectory through the different states, and projecting the cells back on to the trajectory.
+A lot of TI methods use similar algorithms to be able to infer a trajectory.
+By breaking down each method into its set of core algorithms, we can create a map of TI methodology\cite{cannoodt_computationalmethodstrajectory_2016} (Figure~\ref{fig:ti_metro}A), which is divided into two major classes.
+In the first step, dimensionality reduction techniques such as manifold learning, clustering, or graph-based methods are used to convert the dataset to a more simplified representation.
+This representation of the data then allows the trajectory itself to be more easily modelled in a second step.
+In this second step, the trajectory is modelled within the data using graph-based or curve-based approaches, after which the cells themselves can be ordered using a variety of methods.
 
-A common way to classify TI methods is by the types of trajectories they can infer\cite{saelens_comparisonsinglecelltrajectory_2019}.
+A common way to classify TI methods is by the types of trajectories they can infer\cite{saelens_comparisonsinglecelltrajectory_2019} (Figure~\ref{fig:ti_metro}B).
 About half of TI methods specialise in inferring linear or cyclic trajectories (i.e.
@@
 	\caption{
-		\textbf{TI methods use several common building blocks and can be organized in a unifying modular framework.} Overall, TI consists of two steps.
-In the first step dimensionality reduction techniques such as manifold learning, clustering, or graph-based methods are used to convert the dataset to a more simplified representation.
-This representation of the data then allows the trajectory itself to be more easily modelled in a second step.
-In this step, the trajectory itself is found within the data using both graph-based and curve-based approaches, after which the cells themselves can be ordered using a variety of methods.
+		\textbf{TI methods use several common building blocks and can be organized in a unifying modular framework.} 
+		\textbf{A:} Most TI methods consist of two major steps, dimensionality reduction and trajectory modelling.
+TI methods   require some form of dimensionality reduction in order to summarise cell heterogeneity in a lower dimensional space.
+Subsequently, a trajectory modelling step then operates in this reduced space, aiming to identify cell states, constructing a trajectory through the different states, and projecting the cells back on to the trajectory.
+		\textbf{B:} TI methods can be classified according to the trajectory topologies they can infer.
 	}
@@
 Network inference (NI) methods predict which genes are regulated by which transcription factors (Figure~\ref{fig:comp_tools_ni}).
+The output of a network inference is thus a graph, where nodes represent genes and edges denote a regulatory interaction between a regulator and a target gene.
+Interactions typically have two properties: its regulatory strength (a positive real value) and its effect (promoting or repressing).
+
 Before single-cell omics, these methods rely on multiple experiments, amongst which perturbation and time-series experiments, to predict the effect each transcription factor has on the up- or downregulation of a gene.
-
 One of the main advantages of single-cell omics is the heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs.
@@
 
-\section{Research context and objectives}
+\section{Research context and objectives} \label{sec:research_objectives}
 Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology.
@@
 
-One major shortcoming during the advent of single-cell omics was that majority of the newly developed computational tools were not quantitatively and comparatively evaluated.
+One major shortcoming during the advent of single-cell omics was that the majority of newly developed computational tools were not quantitatively and comparatively evaluated.
 Rather, they relied on anecdotal evidence to demonstrate its usefulness.
@@
 For end-users we develop new tools and guidelines for analysing dynamic processes by inferring trajectories and gene regulatory networks.
-These contributions are discussed in the following chapters:
+Our specific scientific objectives are discussed in the following chapters.
 
 \begin{itemize}
-	\item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}).
+	\item We \textbf{develop benchmarking strategies} for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}).
 \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
-	\item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}).
+	\item We apply this strategy to \textbf{perform a comparison of TI methods} (Chapter~\ref{chap:dynbenchmark}).
 Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking.
 We provide a set of guidelines for end-users wishing to infer trajectories.
-We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
-	\item We developed dyno, a toolkit to easily infer, visualise and interpret single-cell trajectories using more than 50 different TI methods (Chapter~\ref{chap:dyno}).
+Our pipeline, datasets, metrics, and containerised wrappers of TI methods are made publicly available for developers to use.
+	\item We develop \textbf{a toolkit to infer, visualise and interpret single-cell trajectories} using more than 50 different TI methods (Chapter~\ref{chap:dyno}).
 dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g.
 adding directionality or adding annotation).
-	\item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
+	\item We introduce a TI method specialised in \textbf{inferring linear trajectories} (Chapter~\ref{chap:scorpius}).
 Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
-	\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
-	% TODO: fill this part in!!!
-	\item Every NI method has certain topological biases.
-We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
-	\item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}).
-We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
-	\item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
+	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
+We demonstrate our NI methodology by inferring regulatory interactions for 25'000 hematopoietic cells, as well as for 15'000 cancer-related bulk profiles.
+% TODO: was this chapter written in the end?
+	\item We provide a tool for  \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
+	Reproducability problems of TI methods are partially due to low rates of quantitative self-assessment.
+We look at 75 manuscripts of TI methods and identify key challenges the authors had in benchmarking their tool.
+For each of these challenges, we provide a summary of how we and others have successfully tackled these challenges, in order to spur developers to perform more self-assessments.
+	\item Finally, we \textbf{summarise our experience in benchmarking computational methods} by providing a list of essential guidelines in benchmarking computational tools (Chapter~\ref{chap:guidelines}).
 \end{itemize}
@@
 \begin{itemize}
+  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
+SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
+bioRxiv 079509.
+2016 Oct.
 	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
@@
 	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. \fbox{\color{red}order authors?}
+	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
 	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
@@
 Packages that were created as part of this work are hosted on Github under the username rcannood\footnote{\url{https://github.com/rcannood?tab=repositories}} or the dynverse organisation\footnote{\url{https://github.com/dynverse?tab=repositories}}.
-As part of our standard development practices, we automate execution of unit tests and writing extensive documentation to ensure the code complies with CRAN policy before submission.
+As part of our standard development practices, we automate execution of unit tests and write extensive documentation to ensure the code complies with CRAN policy before submission.
 We aim to submit all other packages to CRAN as well.

### 2019-10-14 ch6_bred/content.tex
update

@@
 % BRED
+
+One of the central cellular processes underlying development is transcriptional regulation.
+During development, changes in transcription factor activity induce chromatin modifications, chromatin remodeling and ultimately a differential recruitment of the basal transcriptional machinery [66].
+Modeling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
+
+The dynamics of gene regulation has classically been studied using time series data [23].
+When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population [67–70].
+Alternatively, time series data can also be generated by synchronizing the dynamic process between cells.
+However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed [23].
+Combining single-cell snapshot data with trajectory inference can now in principle construct an accurate time-series dataset for every cell population both in vivo and in vitro, while avoiding issues with heterogeneity.
+Initially, several pioneering studies demonstrated the potential of this approach in single cells using single-cell qPCR data for selected sets of 10 to 50 transcription factors [71–75].
+With the advent of genome wide single-cell technologies, the models of dynamic gene regulation can now be made even more complete while avoiding a prior bias to well studied transcription factors [17,30,76,77] and several methods have already been developed specifically tuned towards single-cell data [78–80].
+
+The power of a dynamic network approach in single cells is not only that it allows the identification of transcription factors specifically regulating a specific state, but also how these factors then induce sets of genes necessary for regulating the next wave of transcription changes (Figure 1B).
+Such a gene regulatory cascade was described during cellular reprogramming from mouse embryonic fibroblasts to induced pluripotent stem cells, where SOX2 was shown to activate a series of other transcription factors which, together with SOX2, induces pluripotency [72].
+Transition states are frequently associated with particular subnetworks of transcriptional regulators, as was shown in a direct reprogramming setting from fibroblasts to neurons [77].
+Several studies have highlighted how some regulatory links can be very dynamic while others show evidence of being static during consecutive developmental stages [74,75].
+Regulatory network motifs can also be extracted from the regulatory network [74], some of which can have interesting properties in dynamic systems [81,82].
+In branching trajectories dynamic regulatory networks can also be used to study the regulators important for a particular branch [17,75,79] (Figure 1C).
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\linewidth]{fig/tcga/grouped_interactions.pdf} 
+	\caption{
+		A
+	}
+	\label{fig:tcga}
+\end{figure}
 

### 2019-10-14 ch8_discussion/content.tex
clean up thesis
update

--- 2019-08-26 ch8_discussion/content.tex
@@
-\blindtext
+% DISCUSSION CONTENT
 
-\section{Overview and conclusions of the presented work}
+In this work, we aimed to accelerate scientific progress in the field of single-cell omics by providing and evaluating computational tools.
+In section~\ref{sec:research_objectives} we proposed a concrete set of objectives this work will attempt to solve.
+Every chapter respectively tackles one of these research objectives, and each chapter discusses the implications of that part of our research in detail.
+This chapter reflects on the impact of this work on the field, going over each of the objectives and respective chapters.
 
-% notice that normalisation was absent from this dissertation.
-%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
-%
-%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.
+\section{Benchmarking with \textit{in silico} single cells}
 
-\section{Future research directions}
+We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
+It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
+This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
+
+\texttt{dyngen} has already been used successfully to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
+Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.
+
+\section{Comparing 45 trajectory inference methods}
+Using this simulator and a collection of real datasets, we performed a comparison of 45 TI methods (Chapter~\ref{chap:dynbenchmark}).
+Our contributions include writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, and developing novel metrics for comparing ground truth and predicted trajectories.
+With this benchmarking study, we had two clear goals in mind: to help guide users to TI methods that are suitable for their needs, and to provide developers of TI methods with the necessary tools to benchmark their own tool.
+
+We accomplished the first goal by constructing a set of guidelines for end-users to help choose a TI method that is accurate, robust and fit for their application. 
+Since such guidelines were hitherto lacking, they are now commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.
+
+For our second goal, it seems too early to tell whether our research had any effect on self-assessments of TI method developers.
+We did made our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
+However, so far we do not observe an increase in developers performing quantitative benchmarks.
+We hypothesise causal reasons for this phenomenon and provide solutions in order to spur TI developers to perform more self-assessments (Chapter~\ref{chap:selfassessment}).
+
+\section{A toolkit to infer, visualise and interpret single-cell trajectories}
+
+The previous two works necessitated developing a tool for visualising trajectory data.
+We extended this implementation into a full toolkit, named \texttt{dyno}, for inferring and analysing trajectories (Chapter~\ref{chap:dyno}).
+The toolkit allows performing downstream analyses such detecting differentially expressed genes along transitions in the trajectory, annotating the trajectory, and comparing multiple trajectories in a common dimensionality reduction.
+A major benefit of \texttt{dyno} is that it allows utilising any of the 50 TI methods we wrapped as part of the comparison, without having to install dependencies for every TI method separately.
+
+The software packages consist of several subpackages, each with a separate functionality (e.g.
+\texttt{dynplot} visualises trajectories, \texttt{dynmethods} provides wrappers for the TI methods).
+Since most of the packages are currently only hosted on Github, only users familiar with Github have been able to use it.
+To make \texttt{dyno} available to a larger audience, we are publishing each of these packages and developing command-line interfaces for the most important functionalities.
+
+\section{Fast, accurate, and robust single-cell pseudotime}
+
+We developed a TI method, called \texttt{SCORPIUS}, specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
+Since our comparison of TI methods already lists 30 TI methods that can infer trajectories that are more complex than a linear ordering of cells, why would we need a linear TI method?
+
+Being able to study how cells progress over time is an analysis as fundamental as being able to cluster cells or identify differentially expressed genes between two groups.
+We showed that for analysing datasets containing linear trajectories, \texttt{SCORPIUS} outperforms all other TI methods tested as part of our TI benchmark.
+
+Despite \texttt{SCORPIUS} published on bioRxiv in 2016, discussion and usage of the pre-print has largely been limited to comparing the results of a different (linear) TI methods to those of \texttt{SCORPIUS}.
+Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface.
+Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using SCORPIUS as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019}.
+
+\section{bred: Inferring single cell regulatory networks}
+
+%	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
+We demonstrate our NI methodology by inferring regulatory interactions for 25'000 hematopoietic cells, as well as for 15'000 cancer-related bulk profiles.
+% TODO: was this chapter written in the end?
+
+%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
+
+\section{incgraph: Optimising regulatory networks}
+
+%	\item We provide a tool for  \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+
+%\item Every NI method has certain topological biases.
+We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+
+\section{A life without Git, Travis CI, or tidyverse}
+A significant portion of this work involved developing large software libraries, in a collaborative setting, over a time span of about four years.
+Since the results of our comparison of 45 TI methods required three years to develop, we were happily forced to develop a system where experiments could be rerun and results could be updated on-the-fly.
+
+We summarised our experiences in the form of a set of guidelines for benchmarking computational tools (Chapter~\ref{chap:guidelines}).
+However, these guidelines do not touch upon many aspects of good software development practices that we learned to use in order to bring this thesis to a good end.
+In particular, I would like to highlight several  fundamental (open-source) projects without the likes of which our research would have been simply impossible to perform, namely Git, Travis CI, and the tidyverse.
+
+\textbf{Git}\cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
+Since Wouter and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel.
+Only in few cases did we need to intervene manually to merge our changes.
+Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects.
+
+Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}.
+Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package.
+We wrote many, many unit tests (and still too few), and let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}.
+This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing.
+While sometimes it is a hassle to set up, Travis CI has prevented us countless times from generating faulty results due to a faulty underlying function.
+
+Software bugs can be introduced even by incredibly small, seemingly insignificant changes.
+The R programming language seems particularly susceptible towards getting trapped by its many pitfalls.
+However, there are many benefits of using R in bioinformatics research context, such as its extensive community of statisticians develop packages for the R ecosystem.
+The \textbf{tidyverse} packages\cite{wickham_tidytoolsmanifesto_2017}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.
+
+Honourable mentions go to Linux, Fedora, \LaTeX, TeXstudio, (R)Markdown, Bash, Sed, Regular expressions, Rocket Chat, for making bioinformatics software development even more fun and enjoyable.
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-10-14 ch1_introduction/analysis/computational_tools_2.R

@@ -234,6 +234,16 @@ ggsave(
 
 # Trajectory inference ----------------------------------------------------
 traj <- infer_trajectory(dat, ti_slingshot()) %>% simplify_trajectory()
+
+g <- plot_graph(traj, label_milestones = TRUE)
+
+ggsave(
+  "fig/comptools2/6_ti_graph.pdf",
+  g,
+  width = fig_size,
+  height = fig_size
+)
+
 df <-
   tibble(
     edge = dynwrap::group_onto_trajectory_edges(traj),

### 2019-10-14 ch1_introduction/analysis/download_stats.R

@@ -4,7 +4,7 @@ library(tidyverse)
 cran <- c(
   "babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "dynwrap",
   "GillespieSSA", "GillespieSSA2", "incgraph", "lmds", "princurve", "proxyC",
-  "qsub", "SCORPIUS", "badger", "devtools", "mlr", "ParamHelpers", "remotes", "rlang"
+  "qsub", "SCORPIUS", "badger", "devtools", "mlr", "ParamHelpers", "ranger", "remotes", "rlang"
 )
 bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
 
@@ -16,5 +16,5 @@ dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb
 
 min_date <- "2019-08-01"
 num_days <- as.integer(Sys.Date() - as.Date(min_date))
-summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365)
+summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% slice(match(cran, package))
 summ %>% print(n = 50)

### 2019-10-14 ch6_bred/scripts/2_tcga.R

@@ -357,7 +357,7 @@ if (!file.exists(paste0(data_dir, "qsub_handle.rds"))) {
 }
 
 
-# if (!file.exists(paste0(data_dir, "grn.rds"))) {
+if (!file.exists(paste0(data_dir, "grn.rds"))) {
   qsub_handle1 <- read_rds(paste0(data_dir, "qsub_handle.rds"))
 
   grn1 <- qsub::qsub_retrieve(
@@ -385,222 +385,212 @@ if (!file.exists(paste0(data_dir, "qsub_handle.rds"))) {
   )
 
   grn <- c(grn1, grn2)
+  rm(grn1, grn2, qsub_handle1, qsub_handle2)
 
   write_rds(grn, paste0(data_dir, "grn.rds"))
-# }
-
-grn <- read_rds(paste0(data_dir, "grn.rds"))
-
-# remove unfinished executions
-grn[map_lgl(grn, ~ length(.) == 1 && is.na(.))] <- NULL
-
-# return combined results
-importance <-
-  grn %>%
-  map_df("importance") %>%
-  arrange(desc(importance)) %>%
-  mutate(
-    i = row_number(),
-    name = paste0(regulator, "->", target)
-  )
-ggplot(importance) +
-  geom_point(aes(effect, importance))
-
-# write_tsv(importance %>% filter(importance > .1), "~/importance.tsv")
-importance_sc <- grn %>%
-  map_df("importance_sc") %>%
-  inner_join(importance %>% select(-importance, -name), by = c("regulator", "target"))
-
-rm(grn)
-gc()
-
-samples <- levels(importance_sc$cell_id)
-imp_sc_mat <- Matrix::sparseMatrix(
-  i = importance_sc$cell_id %>% as.integer,
-  j = importance_sc$i,
-  # x = importance_sc$importance_sc,
-  x = importance_sc$importance,
-  dims = c(length(samples), nrow(importance)),
-  dimnames = list(samples, importance$name)
-)
-dimred <- dyndimred::dimred_landmark_mds(imp_sc_mat, ndim = 20, distance_method = "spearman")
-rm(imp_sc_mat)
+}
 
 list2env(read_rds(data_file), .GlobalEnv)
-# df <- data.frame(
-#   sample_info,
-#   # dimred
-#   dimred2
-#   # dimred_fr2
-# )
-# ggplot(df) + geom_point(aes(comp_1, comp_2, colour = project_id))
 
[…]

### 2019-10-14 ch8_discussion/content.tex
update description

@@
 Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface.
-Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using SCORPIUS as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019}.
+Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019}.
 
-\section{bred: Inferring single cell regulatory networks}
-
+\section{Inferring single cell regulatory networks}
+\mycomment{TODO: fill in when chapter has been written.}
 %	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
@@
 
-%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
+%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>^
 
-\section{incgraph: Optimising regulatory networks}
+\section{Optimising regulatory networks}
+This project was a spin-off of Netter\cite{ruyssinck_netterrerankinggene_2016}, a tool for reranking GRN predictions using structural network properties. 
+Predicted GRNs often are topologically dissimilar to real networks, and as such, Netter quantifies the topological properties of a network and makes incremental changes in order to arrive at a network with a 'better' topological profile.
 
-%	\item We provide a tool for  \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+The topological 'profile' of a network was defined as the frequencies of particular topological patterns called 'graphlets'.
+Keeping track of the frequencies of graphlets during an iterative process proved difficult at first.
+\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by making assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals)).
+Under this assumption, it is possible to calculate the differences in topological patterns significantly faster in comparison to recomputing the graphlet frequencies from scratch at every iteration.
 
-%\item Every NI method has certain topological biases.
-We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+While \texttt{incgraph} solved a very specific and interesting problem, and was eventually was published as a peer-reviewed article, it has not succeeded in making a direct impact on scientific research as of yet.
+Perhaps, at some point, a researcher will come across \texttt{incgraph} and find that it exactly solves his or her problem.
 

### 2019-10-14 ch8_discussion/content.tex
update discussion

@@
 We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
-It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
-This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
+The \texttt{dyngen} software supports a selection of existing experimental protocols, and can also simulate experiments that are possible with the current technologies, such as generating omics profiles from the same cell at multiple timepoints.
+The generated data can help kick-start emerging domains with low data availability more safely by allowing software developers to test their method before numerous datasets become publicly available.
 
-\texttt{dyngen} has already been used successfully to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
+\texttt{dyngen} has already been successfully used to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
 Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.
+
+% Eventueel kort ook beschrijven waar je denkt dat het nog van toepassing kan zijn? The use if dyngene is not limited to.... but might also,....
+% Eventueel ook: waar zie je nog verbeter/uitbreidingspunten in je methode. 
+% Wat zijn de huidige limitaties. Bv. Is er een tijdgebonden limitatie. bv. Stel je wilt data van 10.000 cellen genereren maar bent een maand bezig, etc.
 
@@
 We hypothesise causal reasons for this phenomenon and provide solutions in order to spur TI developers to perform more self-assessments (Chapter~\ref{chap:selfassessment}).
-
+% "We hypothesise." Where/what do you hypothise? In your paper? Dan zou ik verwijzen naar deze sectie.
+% Eventueel ook: Is there room for improvement for existing methods (deze zullen vermoedelijk onder een andere naam gepubliceerd dienen te worden)
+% Dienen eerdere publicaties in vraag te worden gesteld (indien minder goede methods gebruikt werden?
+% Kan dit uitgebreid worden naar andere methoden?
+% Is dyngen niet ontstaan omdat je nood had aan betere datasets? Zou je in dit opzicht niet dit deeltje eerst plaatsen en verwijzen naar dyngen?
+% Misschien ook vermelden dat de publicatie een aanzienlijke support verwacht van jullie team na publicatie. Dat de hoge nood gepaard gaat met een grote vraag. 
 \section{A toolkit to infer, visualise and interpret single-cell trajectories}
@@
 We extended this implementation into a full toolkit, named \texttt{dyno}, for inferring and analysing trajectories (Chapter~\ref{chap:dyno}).
-The toolkit allows performing downstream analyses such detecting differentially expressed genes along transitions in the trajectory, annotating the trajectory, and comparing multiple trajectories in a common dimensionality reduction.
+The toolkit allows performing downstream analyses such as detecting differentially expressed genes along transitions in the trajectory, annotating the trajectory, and comparing multiple trajectories in a common dimensionality reduction.
 A major benefit of \texttt{dyno} is that it allows utilising any of the 50 TI methods we wrapped as part of the comparison, without having to install dependencies for every TI method separately.
@@
 
+% Are we talking about R-packages?
+
 \section{Fast, accurate, and robust single-cell pseudotime}
@@
 
-Being able to study how cells progress over time is an analysis as fundamental as being able to cluster cells or identify differentially expressed genes between two groups.
+Being able to study how cells progress over time is in terms of analysis as fundamental as being able to cluster cells or to identify differentially expressed genes between two groups.
 We showed that for analysing datasets containing linear trajectories, \texttt{SCORPIUS} outperforms all other TI methods tested as part of our TI benchmark.
@@
 
-While \texttt{incgraph} solved a very specific and interesting problem, and was eventually was published as a peer-reviewed article, it has not succeeded in making a direct impact on scientific research as of yet.
+While \texttt{incgraph} solved a very specific and interesting problem, and was eventually published as a peer-reviewed article, it has not succeeded in making a direct impact on scientific research as of yet.
 Perhaps, at some point, a researcher will come across \texttt{incgraph} and find that it exactly solves his or her problem.
+
+% Of je dient voor te stellen bij welke dataset het goed van toepassing is en het te promotor onder de Vlaamse research gemeenschap?
 
@@
 Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects.
-
+% Misschien ook Wouter zijn achternaam gebruiken. Klinkt wat professioneler :-p
 Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}.
 Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package.
-We wrote many, many unit tests (and still too few), and let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}.
+We wrote many, many unit tests (and still too few), and we let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}.
 This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing.

### 2019-10-14 ch3_dynbenchmark/content.tex
fix minor issues

@@
 
-We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
-We welcome any contributions by creating an issue at https://methods.dynverse.org.
+We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
+We welcome any contributions by creating an issue at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.
 
@@
 
-* Linear and cyclic, with number of milestones $\sim B(10, 0.25)$
-* Bifurcating and converging, with four milestones
-* Binary tree, with number of branching points $\sim U(3, 6)$
-* Tree, with number of branching points $\sim U(3, 6)$ and maximal degree $\sim U(3, 6)$
+\begin{itemize}
+	\item Linear and cyclic, with number of milestones $\sim B(10, 0.25)$
+	\item Bifurcating and converging, with four milestones
+	\item Binary tree, with number of branching points $\sim U(3, 6)$
+	\item Tree, with number of branching points $\sim U(3, 6)$ and maximal degree $\sim U(3, 6)$
+\end{itemize}
 
@@
 
-* Looping, allowed loop modifications
-* Diverging-converging, allowed divergence and converging modifications
-* Diverging with loops, allowed divergence and loop modifications
-* Multiple looping, allowed looping modifications
-* Connected, allowed looping, divergence and convergence modifications
-* Disconnected, number of components sampled from $\sim B(5, 0.25) + 2$, for each component we randomly chose a topology from the ones listed above
+\begin{itemize}
+	\item Looping, allowed loop modifications
+	\item Diverging-converging, allowed divergence and converging modifications
+	\item Diverging with loops, allowed divergence and loop modifications
+	\item Multiple looping, allowed looping modifications
+	\item Connected, allowed looping, divergence and convergence modifications
+	\item Disconnected, number of components sampled from $\sim B(5, 0.25) + 2$, for each component we randomly chose a topology from the ones listed above
+\end{itemize}
 

### 2019-10-14 ch6_bred/content.tex
wip ch6

--- 2019-09-29 ch6_bred/content.tex
@@
 % BRED
+
+\section{Introduction}
+
+One of the central cellular processes underlying development is transcriptional regulation.
+During development, changes in transcription factor activity induce chromatin modifications, chromatin remodelling and ultimately a differential recruitment of the basal transcriptional machinery\cite{coulon_eukaryotictranscriptionaldynamics_2013}.
+Modelling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
+
+The dynamics of gene regulation has classically been studied using time series data\cite{bar-joseph_studyingmodellingdynamic_2012}.
+When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population\cite{novershtern_denselyinterconnectedtranscriptional_2011, may_dynamicanalysisgene_2013, jojic_identificationtranscriptionalregulators_2013, goode_dynamicgeneregulatory_2016}.
+Alternatively, time series data can also be generated by synchronizing the dynamic process between cells.
+However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed\cite{bar-joseph_studyingmodellingdynamic_2012}.
+
+One of the main advantages of single-cell transcriptomics is the ability to quantify the exact cellular state of thousands of cells per experiment.
+The intercellular heterogeneity caused by naturally occurring biological stochasticity \cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to predict regulatory interactions between transcription factors (TFs) and their target genes.
+The computational tools that infer gene regulatory networks (GRNs) from omics datasets are called network inference (NI) methods.
+% TODO: talk about bulk NI
+
+Several studies have highlighted how some regulatory interactions can be very dynamic while others show evidence of being static during consecutive developmental stages\cite{moignard_characterizationtranscriptionalnetworks_2013, pina_singlecellnetworkanalysis_2015}.
+Since regulatory interactions are context-dependent\cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
+
+Case-specific NI (CSNI) methods avoid predicting a global GRN and instead produce one GRN per case in the dataset.
+The sample-specific GRNs -- or 'sample-specific regulomes' -- can be used in much the same way as single-cell transcriptomi
+
+ Kuijjer et al.\cite{kuijjer_estimatingsamplespecificregulatory_2019} LOO.
+Liu et al.
+\cite{personalizedcharacterizationdiseases}.
+Aibar et al.
+SCENIC \cite{aibar_scenicsinglecellregulatory_2017} use global method and post-process.
+
+NI methods that take into account the dynamic aspect of gene regulation instead produce network models with variable regulatory activity.
+
+ To this end, several approaches have been proposed, which can be broadly classified in three different classes, depending on the output structure they produce: differential NI, dynamic NI, and profile-specific NI (Figure \ref{fig:ni_types}).
+
+With each of these methodologies, it should be noted that while they produce networks specific for only certain subsets of the cell's profiles, they still use the information from all available profiles.
+If a method is to infer a network from cells in only a certain condition, it will infer interactions from noise in the data, rather than the changes that separate that condition from any other.
+As such, a context-dependent network inferred from only a subset of the profiles is likely to be less accurate than a static network trained on all profiles.
+
+\section{Results}
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\linewidth]{fig/tcga/grouped_interactions.pdf} 
+	\caption{
+		A
+	}
+	\label{fig:tcga}
+\end{figure}
+
+\section{Discussion}
+
+\section{Methods}
 

### 2019-10-15 ch6_bred/content.tex
update

@@
 Since regulatory interactions are context-dependent\cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
+Case-wise NI methods\footnote{Case-wise NI is sometimes also called sample-specific NI or case-specific NI.} avoid predicting a static GRN and instead infer one GRN per cell (or per sample, for bulk omics data).
+The case-wise GRNs -- or 'case-wise regulomes' -- are analysed similarly to single-cell transcriptomics data; for example by clustering, inferring trajectories, or finding differentially activated interactions.
 
-Case-specific NI (CSNI) methods avoid predicting a global GRN and instead produce one GRN per case in the dataset.
-The sample-specific GRNs -- or 'sample-specific regulomes' -- can be used in much the same way as single-cell transcriptomi
+In order to compute a case-wise GRN for a single sample, Kuijjer et al.\cite{kuijjer_estimatingsamplespecificregulatory_2019} and Liu et al.\cite{liu_personalizedcharacterizationdiseases_2016} employ similar strategies, namely by computing the difference of computing a static GRN for all the cases, and computing a static GRN for all the cases minus one.
+Since this procedure needs to be repeated for every case in the dataset, and because NI methods are already amongst the most computationally intensive analyses to perform on omics data, this methodology is not applicable for large omics datasets.
+Another case-wise NI method, SCENIC\cite{aibar_scenicsinglecellregulatory_2017} infers case-wise GRNs by first inferring a static GRN using GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010}.
+GENIE3 is a static NI method which uses Random Forests\cite{breiman_randomforests_2001} feature importance scores to prioritise candidate regulators for a particular target gene.
+SCENIC then post-processes the static GRN to determine whether an interaction is enriched for particular cases, resulting in a case-wise GRN.
+In short, while several case-wise NI methods thus already exist, their implementation consisted of post-processing a static GRN to arrive at a case-wise GRN.
 
- Kuijjer et al.\cite{kuijjer_estimatingsamplespecificregulatory_2019} LOO.
-Liu et al.
-\cite{personalizedcharacterizationdiseases}.
-Aibar et al.
-SCENIC \cite{aibar_scenicsinglecellregulatory_2017} use global method and post-process.
+In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method.
 
-NI methods that take into account the dynamic aspect of gene regulation instead produce network models with variable regulatory activity.
-
- To this end, several approaches have been proposed, which can be broadly classified in three different classes, depending on the output structure they produce: differential NI, dynamic NI, and profile-specific NI (Figure \ref{fig:ni_types}).
-
-With each of these methodologies, it should be noted that while they produce networks specific for only certain subsets of the cell's profiles, they still use the information from all available profiles.
-If a method is to infer a network from cells in only a certain condition, it will infer interactions from noise in the data, rather than the changes that separate that condition from any other.
-As such, a context-dependent network inferred from only a subset of the profiles is likely to be less accurate than a static network trained on all profiles.
+\texttt{bred} is an extension of GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} in that it uses the case-wise feature importance\cite{xu_casespecificrandomforests_2016} score rather than the regular Random Forest feature importance scores.
+In addition to predicting regulatory strength of an interaction, \texttt{bred} also predicts the regulatory effect of an interaction -- that is, whether a regulator activates or represses the target.
+We demonstrate \texttt{bred} by applying it on a single-cell dataset of 25'000 hematopoietic cells from the Tabula Muris project\cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 15'000 bulk omics samples from The Cancer Genome Atlas project\cite{weinstein_cancergenomeatlas_2013}.
 
@@
 	\centering
-	\includegraphics[width=\linewidth]{fig/tcga/grouped_interactions.pdf} 
+	\includegraphics[width=\linewidth]{fig/example.pdf} 
 	\caption{
-		A
+		\textbf{A schematic overview of the workings of the \texttt{bred} algorithm.}
+	}
+	\label{fig:example_cni}
+\end{figure}
+
+\begin{figure}[htb!]
+	{\raggedright\textbf{A}}
+	\begin{center}\includegraphics[width=.9\linewidth]{fig/tabula_muris_hema/grouped_interactions.pdf}\end{center}
+	{\raggedright\textbf{B}}
+	\begin{center}\includegraphics[width=\linewidth]{fig/tabula_muris_hema/plot_fr_cropped.png}\end{center}
+	\caption{
+		\textbf{Hematopoietic cells from the Tabula Muris project.}
+	}
+	\label{fig:tabmur}
+\end{figure}
+
+\begin{figure}[htb!]
+	{\raggedright\textbf{A}}
+	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/grouped_interactions.pdf}\end{center}
+	{\raggedright\textbf{B}}
+	\begin{center}\includegraphics[width=.6\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
+	\caption{
+		\textbf{The Cancer Genome Atlas.}
 	}
@@
 
+\subsection{Building the case-wise GRNs}
+Having to infer a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely for every target $T$, predict which of the potential regulators regulate $T$ (Figure~\ref{fig:simplify}B).
+This simplification allowed GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's\cite{breiman_randomforests_2001} feature importance scores for inferring GRNs.
+Namely, a Random Forest is trained to predict the expression of a target gene of interest from the expression of potential regulators.
+The resulting Random Forest inherently allows to extract a feature importance score by observing the effect of each regulator in making a good prediction for the target expression.
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.6\linewidth]{fig/methodology/simplify.pdf} 
+	\caption{
+		\textbf{A:} Inferring a gene regulatory network from an omics dataset can be reduced to a simpler problem. 
+		\textbf{B:} Given the expression of a target of interest and a set of potential regulators, predict which regulators regulate the target gene.
+	}
+	\label{fig:simplify}
+\end{figure}
+
+We make the same simplification in order to build case-wise GRNs, also using Random Forests to compute the feature importance scores.
+A Random Forest consists of $K$ trees, each of which produces feature importance scores, and the feature importance scores of a forest is simply the mean feature importance scores of each of the trees.
+
+Computing the case-wise feature importances of a tree consists of the following 8 steps (Figure~\ref{fig:fimp}).
+The 'randomness' of a Random Forest is due to only using a subset of the samples in the dataset in order to build a single decision tree.
+The samples are split into two groups, the 'in-bag' data and the 'out-of-bag' data (Figure~\ref{fig:fimp}A).
+A decision tree\cite{breiman_classificationregressiontrees_1984} is trained on the in-bag expression of the potential regulators in trying to predict the in-bag target gene expression (Figure~\ref{fig:fimp}B).
+The target expression of the out-of-bag samples is predicted using the decision tree (Figure~\ref{fig:fimp}C), and the squared error between the real and target expression is computed (Figure~\ref{fig:fimp}D).
+For each sample in the out-of-bag set, this vector represents how well the decision tree was able to predict the expression of the target gene.
+
+The next few steps are repeated for every potential regulator $R_i$.
+Within the out-of-bag samples, the expression of $R_i$ is randomly shuffled.
+The target expression of the out-of-bag samples is again calculated (Figure~\ref{fig:fimp}F), as well as the squared error between the real target expression and the predicted expression is calculated (Figure~\ref{fig:fimp}G).
+The importance of regulator $R_i$ for an out-of-bag sample $S_j$ is defined as the increase in squared error between the predicted target expression and the real target expression, after perturbing the expression of $R_i$ (Figure~\ref{fig:fimp}H).
+
+Steps F-G are repeated for every potential regulator $R_i$.
+By aggregating all of the feature importance scores over all the samples, regulators and targets, we obtain an $M$-by-$N$-by-$P$ tensor\footnote{This is the origin of the name of the method, "\texttt{bred}".}.
+
+A moderately-sized dataset could contain $M=10'000$ samples, $N=2'000$ regulators, and $P=10'000$ target genes.
+Due to memory constrains, only interactions with an average importance value (across all samples) higher than a minimum threshold are retained.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.9\linewidth]{fig/methodology/fimp.pdf} 
+	\caption{
+		\textbf{Calculating the feature importance score for one decision tree and one target consists of 8 distinct steps.}
+		\textbf{A:} Randomly split the data into two groups, the in-bag data and the out-of-bag data.
+		\textbf{B:} The in-bag data is used to train a decision tree to try to predict the expression of the target gene from the expression values of the regulators.
+		\textbf{C:} The decision tree is used to predict the gene expression of the target gene of the out-of-bag samples.
+		\textbf{D:} Sample-specific squared error values are computed.
+		\textbf{E:} Repeat steps E-H for every regulator $R_i$. Perturb the expression of regulator $R_i$ in the out-of-bag samples.
+		\textbf{F:} Again predict the gene expression of the target gene with the perturbed expression values.
+		\textbf{G:} Again compute the sample-specific squared error values.
+		\textbf{H:} The difference between the prediction error on the perturbed dataset versus the prediction error on the unperturbed is the importance in $R_i$ in predicting $T$ 
+	}
+	\label{fig:fimp}
+\end{figure}
+
+\subsection{Clustering of case-wise GRNs}
+
+%LMDS -> KNN -> FR \& Louvain clustering
+
 \clearpage

### 2019-10-15 ch6_bred/content.tex
update

@@
 In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. 
-
-\texttt{bred} is an extension of GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} in that it uses the case-wise feature importance\cite{xu_casespecificrandomforests_2016} score rather than the regular Random Forest feature importance scores.
+\texttt{bred} uses case-wise feature importance\cite{xu_casespecificrandomforests_2016} score rather than the regular Random Forest feature importance scores.
 In addition to predicting regulatory strength of an interaction, \texttt{bred} also predicts the regulatory effect of an interaction -- that is, whether a regulator activates or represses the target.
@@
 	{\raggedright\textbf{B}}
-	\begin{center}\includegraphics[width=.6\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
+	\begin{center}\includegraphics[width=.8\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
 	\caption{
@@
 The resulting Random Forest inherently allows to extract a feature importance score by observing the effect of each regulator in making a good prediction for the target expression.
+As in GENIE3, the target expression is first scaled to normalise feature importance scores across different targets.
+
 \begin{figure}[htb!]
@@
 
+To compute the case-wise GRNs, we implemented the abovementioned methodology in C++ in a modified version of the \texttt{ranger} R/C++ package\cite{wright_rangerfastimplementation_2017}.
+
+\subsection{Predicting the effect of an interaction}
+To predict the effect of a potential regulator $R_i$ on a target gene $T$ for a given tree, the Pearson correlation is calculated between the difference in regulator expression (before and after shuffling the values), and the difference in target expression prediction.
+
+\begin{align*}
+  \text{effect}(R_i \rightarrow T) & = \text{cor}(x, y), \\
+  \text{with } x & = \text{expr\_shuffled}[:,R_i] - \text{expr}[:,R_i], \\
+  \text{and } y & = \text{predict}(\text{tree}, \text{expr\_shuffled}) - \text{predict}(\text{tree}, \text{expr}).
+\end{align*}
+
+The Pearson correlation between two variables $x$ and $y$ is usually defined as shown in Equation~\ref{eqn:pearson1}.
+Computing $r_{xy}$ for each (regulator, target) pairs, across all trees, would require storing large amounts of data.
+
+\begin{equation}\label{eqn:pearson1}
+	r_{xy} = \frac{\sum^n_{i=1} (x_i - \bar{x}) \times (y_i - \bar{y})}{\sqrt{\sum^n_{i=1}(x_i - \bar{x})^2}\times\sqrt{\sum^n_{i=1}(y_i - \bar{y})^2}}
+\end{equation}
+
+However, by rearranging the formula, it can be defined as Equation~\ref{eqn:pearson2}.
+
+\begin{equation}\label{eqn:pearson2}
+r_{xy} = \frac{\sum(x_i \times y_i) - \sum x \times \sum y / n}{\sqrt{(\sum x_i^2 - (\sum x)^2 / n)} \times \sqrt{(\sum y_i^2 - (\sum y)^2 / n)}}
+\end{equation}
+
+For every regulator $R_i$ during a perturbation in a given tree, only 6 values need to be stored, namely $A = \sum x_i$, $B = \sum y_i$, $C = n$, $D = \sum{x_i \times y_i}$, $E = \sum{x_i \times x_i}$, and $F = \sum{y_i \times y_i}$.
+
+For every (regulator, target) pair, these values are summed, and the $r_{xy}$ is calculated as shown in Equation~\ref{eqn:pearson3}.
+
+\begin{equation}\label{eqn:pearson3}
+r_{xy} = \frac{D - A \times B / C}{\sqrt{(E - A^2 / C)} \times \sqrt{(F - B^2 / C)}}
+\end{equation}
+
+The following cutoffs were used to determine the effect.
+
+\begin{itemize}
+	\item Strong negative: $r_{xy} < -0.4$
+  \item Weak negative: $-0.4 \leq r_{xy} < -0.2$
+  \item Unclear: $-0.2 \leq r_{xy} \leq 0.2$
+  \item Weak positive: $0.2 < r_{xy} \leq 0.4$
+  \item Strong positive: $0.4 < r_{xy}$
+\end{itemize}
+
 \subsection{Clustering of case-wise GRNs}
+To perform downstream analysis on the cases, first a $k$-nearest neighbour ($k$NN) graph of the cases is computed.
+In order for the $k$NN graph to better emphasise similarities in GRNs rather than absolute euclidean distances, we first reduce the dimensionality of the case-by-interaction matrix to case-by-20 matrix using Landmark Multi-Dimensional Scaling\cite{lee_landmarkmdsensemble_2009} with a Spearman rank distance metric.
 
-%LMDS -> KNN -> FR \& Louvain clustering
+Next, KD-trees are used to calculate the $k$NN graph efficiently.
+The cases in the dataset are visualised and clusted using the Fruchterman-Reingold\cite{fruchterman_graphdrawingforcedirected_1991} and Louvain clustering\cite{blondel_fastunfoldingcommunities_2008}, respectively.
+
+The following R packages provided implementations for each of these algorithms: lmds, RANN, igraph\cite{csardi_igraphsoftwarepackage_2006}.
+
+\subsection{Data from the Tabula Muris project}
+
+\subsection{Data from The Canger Genome Atlas project}
 

### 2019-10-15 ch6_bred/content.tex
~

@@
 Case-wise NI methods\footnote{Case-wise NI is sometimes also called sample-specific NI or case-specific NI.} avoid predicting a static GRN and instead infer one GRN per cell (or per sample, for bulk omics data).
-The case-wise GRNs -- or 'case-wise regulomes' -- are analysed similarly to single-cell transcriptomics data; for example by clustering, inferring trajectories, or finding differentially activated interactions.
 
@@
 
-In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. 
-\texttt{bred} uses case-wise feature importance\cite{xu_casespecificrandomforests_2016} score rather than the regular Random Forest feature importance scores.
-In addition to predicting regulatory strength of an interaction, \texttt{bred} also predicts the regulatory effect of an interaction -- that is, whether a regulator activates or represses the target.
-We demonstrate \texttt{bred} by applying it on a single-cell dataset of 25'000 hematopoietic cells from the Tabula Muris project\cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 15'000 bulk omics samples from The Cancer Genome Atlas project\cite{weinstein_cancergenomeatlas_2013}.
+In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. For each interaction in the inferred case-wise GRN, \texttt{bred} predicts both the regulatory strength and its effect. 
+The case-wise GRNs -- or 'case-wise regulomes' -- can be analysed analogously to transcriptomics data; for example by clustering samples, inferring trajectories, or finding differentially activated regions.
+We demonstrate \texttt{bred} by applying it to a single-cell dataset of 22'122 hematopoietic cells from the Tabula Muris project\cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 14'963 bulk omics samples from The Cancer Genome Atlas project\cite{weinstein_cancergenomeatlas_2013}.
 
@@
 
-\begin{figure}[htb!]
-	\centering
-	\includegraphics[width=\linewidth]{fig/example.pdf} 
-	\caption{
-		\textbf{A schematic overview of the workings of the \texttt{bred} algorithm.}
-	}
-	\label{fig:example_cni}
-\end{figure}
+\subsection{Hematopoietic cells from Tabula Muris}
+From the Tabula Muris\cite{schaum_singlecelltranscriptomics20_2018} project, we selected all cells involved in some part of the hematopoietic lineage tree.
+We computed case-wise GRNs between 22'122 remaining single cells, for the 2000 most variable target genes and a subset of transcription factors as regulators.
+
+To summarise the similarity in GRNs across cases, we first visualise a dimensionality reduction of the case-wise GRNs, where every dot represents a single GRN (Figure~\ref{fig:tabmur}A).
+The dimensionality reduction was computed by applying Fruchterman-Reingold\cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour ($k$NN) graph of the case-wise GRN vectors.
+On the same $k$NN graph, the cells were clustered using the Louvain\cite{blondel_fastunfoldingcommunities_2008} clustering algorithm, and the clusters were labelled using prior information from Tabula Muris.
+For each cluster, we retained the 50 interactions with the highest mean importance scores (Figure~\ref{fig:tabmur}B).
+In the visualisation, each node represents a gene, each edge an interaction, and the colour represents which cluster this interaction belongs to.
+
+For the most part, clusters that are proximate in the dimensionality reduction (Figure~\ref{fig:tabmur}A) are also closely connected in the GRN network view (Figure~\ref{fig:tabmur}B).
+Notably, interactions from the different B cell clusters are almost not connected amongst each-other, while in the dimensionality reduction they re very proximate.
+Retaining more edges could result in the different B cells being connected in the GRN network view.
+
+At the centre of the network lie many strongly connected interactions which are not specific for any particular cluster, but instead are common to almost all of the clusters.
+These are likely housekeeping genes or genes involved in key hematopoietic processes.
 
@@
 	{\raggedright\textbf{A}}
+	\begin{center}\includegraphics[width=\linewidth]{fig/tabula_muris_hema/plot_fr_cropped.png}\end{center}
+	{\raggedright\textbf{B}}
 	\begin{center}\includegraphics[width=.9\linewidth]{fig/tabula_muris_hema/grouped_interactions.pdf}\end{center}
-	{\raggedright\textbf{B}}
-	\begin{center}\includegraphics[width=\linewidth]{fig/tabula_muris_hema/plot_fr_cropped.png}\end{center}
 	\caption{
-		\textbf{Hematopoietic cells from the Tabula Muris project.}
+		\textbf{22'122 case-wise GRNs of hematopoietic cells from the Tabula Muris project.}
+		\textbf{A:} Each dot in this dimensionality reduction corresponds to the GRN of a single cell.
+The dimensionality of the cell-specific regulome matix was reduced using Fruchterman-Reingold and were clustered using Louvain clsutering, after which the clusters were relabelled using the meta information from Tabula Muris.
+		\textbf{B:} Per cluster, the 50 interactions with the highest mean importance score are visualised.
 	}
@@
 
+\subsection{The Cancer Genome Atlas}
+We downloaded all available RNA-seq profiles from The Cancer Genome Atlas. In total, we computed case-wise GRNs for 14'963 tumour samples from 50 sub-projects, including 44 different cancer entities.
+
 \begin{figure}[htb!]
 	{\raggedright\textbf{A}}
-	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/grouped_interactions.pdf}\end{center}
+	\begin{center}\includegraphics[width=.8\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
 	{\raggedright\textbf{B}}
-	\begin{center}\includegraphics[width=.8\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
+	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/grouped_interactions_2.pdf}\end{center}
 	\caption{
-		\textbf{The Cancer Genome Atlas.}
+		\textbf{14'963 case-wise GRNs of cancer bulk profiles from The Cancer Genome Atlas project.}
+		\textbf{A:} The different profiles are visualised and clustered according to their regulomic similarities.
+		\textbf{B:} Visualisation of the strongest interactions per cluster shows both pathways distinct to particular cancer entities as well as pathways common to multiple cancer entities.
 	}
@@
 
+% going forward: functional validation; benchmark against in silico datasets; 
+% can also use TI instead of clustering
+
 \section{Methods}
 
-\subsection{Building the case-wise GRNs}
-Having to infer a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely for every target $T$, predict which of the potential regulators regulate $T$ (Figure~\ref{fig:simplify}B).
+\subsection{Inferring case-wise GRNs}
+The task of inferring a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely: for every target $T$, predict which of the potential regulators $R_i$ regulate $T$ (Figure~\ref{fig:simplify}B).
 This simplification allowed GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's\cite{breiman_randomforests_2001} feature importance scores for inferring GRNs.
@@
 
+\subsection{Visualising clustered GRNs}
+After Louvain clustering, the interactions of the 50 interactions with highest mean importance per cluster are retained.
+These interactions are visualised in Cytoscape\cite{shannon_cytoscapesoftwareenvironment_2003}, in which nodes depict genes, edges depict predicted regulatory interactions, coloured according to which cluster they are predicted for.
+The shape of the arrow denotes the predicted effect of the regulatory interaction.
+
 \subsection{Data from the Tabula Muris project}

### 2019-10-15 ch6_bred/content.tex
update bred

@@
 \subsection{The Cancer Genome Atlas}
-We downloaded all available RNA-seq profiles from The Cancer Genome Atlas. In total, we computed case-wise GRNs for 14'963 tumour samples from 50 sub-projects, including 44 different cancer entities.
+We included all available RNA-seq profiles from The Cancer Genome Atlas.
+In total, we computed case-wise GRNs for 14'963 tumour samples from 50 sub-projects, including 44 different cancer entities.
+Clusters of case-wise GRNs were relabelled by the project they originate from, or by the cell types or organs the samples originate from.
+
+Most groups of interactions are highly specific to just one cancer type.
+Samples from the CPTAC project were split into two groups, clustering together with LUAD (lung adenocarcinoma) and kidney carcinoma samples.
+According to the meta-information of CPTAC profiles, these samples consist of lung, kidney, and uterus adenocarcinomas, explaining the split in CPTAC samples.
+
+%TODO: Extend !!
 
@@
 
-% going forward: functional validation; benchmark against in silico datasets; 
-% can also use TI instead of clustering
+The \texttt{bred} algorithm is a novel approach for directly computing case-wise GRNs for single-cell omics and bulk omics profiles alike.
+We applied the method to 22'122 hematopoietic cells and 14'963 cancer profiles, showing that interactions are often grouped together in modules specific for certain cell types, tissue types, or cancer types.
+
+In this work, we only applied clustering methods to the case-specific 'regulome profiles', but other types of computational methods can be used to annotate and explore case-specific GRNs, such as trajectory inference and differential expression.
+
+Going forward, we will provide further functional validation of the results generated by \texttt{bred}, as well as benchmark the algorithm against various real and \textit{in silico} datasets.
 
@@
 
-\subsection{Data from the Tabula Muris project}
-
-\subsection{Data from The Canger Genome Atlas project}
+%\subsection{Data from the Tabula Muris project}
+%
+%\subsection{Data from The Cancer Genome Atlas project}
 

### 2019-10-15 boek.tex
~

@@
 \chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
-\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
+\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods}
 \label{chap:dynbenchmark}
@@
 \begin{quote}
-	\textbf{Abstract:}
+	\textbf{Abstract:} Recent advances in RNA sequencing enable the generation of genome-wide expression data at the single-cell level, opening up new avenues for transcriptomics and systems biology.
+A new application of single-cell transcriptomics is the unbiased ordering of cells according to their progression along a dynamic process of interest.
+SCORPIUS is a toolkit for pseudotemporally ordering single cells along a linear trajectory.
+Comprehensive benchmarking on linear datasets shows that SCORPIUS consistently obtains higher quality linear trajectories in comparison to state-of-the-art trajectory inference methods.
+We used SCORPIUS to generate novel hypotheses regarding dendritic cell development, which were subsequently validated in vivo.
 \end{quote}
@@
 \vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., Saeys, Y, and De Preter K. Inferring Single Cell Regulatory Networks. \textit{Journal} vol, issue (2019), page--page. doi.
+Adapted from:
+\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y.
+Computational methods for trajectory inference from single-cell transcriptomics.
+\textit{European Journal of Immunology} 46, 11 (2016), 2496--2506.
+\doi{10.1002/eji.201646347}.\\
+%Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
+A comparison of single-cell trajectory inference methods.
+\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
+\doi{10.1038/s41587-019-0071-9}.\\
+Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y.
+Network Inference from Single-Cell Transcriptomic Data.
+\textit{Gene Regulatory Networks} (2019), 235--249.
+\doi{10.1007/978-1-4939-8882-2_10}.
+\\
+{\footnotesize $^*$ Equal contribution}
 \newpage

### 2019-10-15 ch1_introduction/content.tex
~

@@
 \begin{figure}[htb!]
-	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
+	\centering
+	\includegraphics[width=.8\linewidth]{fig/rnaworld} % TODO: change image
 	\caption{
@@
 Image from Horning (2011)\cite{horning_rnaworld_2011}.
-		\fbox{\color{red} TODO: combine images, rewrite the description.}
+%		\fbox{\color{red} TODO: combine images, rewrite the description.}
 	}
@@
 \begin{figure}[htb!]
-	\centering\includegraphics[width=.5\linewidth]{fig/central_dogma} % TODO: change image
+	\centering
+	\includegraphics[width=.7\linewidth]{fig/central_dogma} % TODO: change image
 	\caption{
 		\textbf{Central Dogma.}
-		\fbox{\color{red} TODO: combine images, rewrite the description.}
+%		\fbox{\color{red} TODO: combine images, rewrite the description.}
 	}
@@
 	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
-We demonstrate our NI methodology by inferring regulatory interactions for 25'000 hematopoietic cells, as well as for 15'000 cancer-related bulk profiles.
-% TODO: was this chapter written in the end?
+We demonstrate our NI methodology by inferring regulatory interactions for 22'122 hematopoietic cells, as well as for 14'963 bulk profiles of cancer tumours.
 	\item We provide a tool for  \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
@@
 \end{itemize}
-
-\section{List of contributions}
-
-\subsection{First-author publications}
-
-\begin{itemize}
-  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
-SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
-bioRxiv 079509.
-2016 Oct.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
-	\item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
-	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
-	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
-\end{itemize}
-
-*: Equal contribution.
-
-\subsection{Co-author publications}
-\begin{itemize}
-	\item Decock A, Ongenaert M, \textbf{Cannoodt R}, Verniers K, De Wilde B, Laureys G, Van Roy N, Berbegall AP, Bienertova-Vasku J, Bown N, Clément N.
-Methyl-CpG-binding domain sequencing reveals a prognostic methylation signature in neuroblastoma.
-Oncotarget.
-2016 Jan 12;7(2):1960.
-	\item Van Cauwenbergh C, Van Schil K, \textbf{Cannoodt R}, Bauwens M, Van Laethem T, De Jaegere S, Steyaert W, Sante T, Menten B, Leroy BP, Coppieters F.
-arrEYE: a customized platform for high-resolution copy number analysis of coding and noncoding regions of known and candidate retinal dystrophy genes and retinal noncoding RNAs.
-Genetics in Medicine.
-2017 Apr;19(4):457.
-	\item Claeys S, Denecker G, \textbf{Cannoodt R}, Kumps C, Durinck K, Speleman F, De Preter K.
-Early and late effects of pharmacological ALK inhibition on the neuroblastoma transcriptome.
-Oncotarget.
-2017 Dec 5;8(63):106820.
-	\item Depuydt P, Boeva V, Hocking TD, \textbf{Cannoodt R}, Ambros IM, Ambros PF, Asgharzadeh S, Attiyeh EF, Combaret V, Defferrari R, Fischer M.
-Genomic amplifications and distal 6q loss: novel markers for poor survival in high-risk neuroblastoma patients.
-JNCI: Journal of the National Cancer Institute.
-2018 Mar 5;110(10):1084-93.
-	\item Scott CL, T'Jonck W, \ldots, \textbf{Cannoodt R}, Saelens W \ldots, Guilliams M.
-The transcription factor ZEB2 is required to maintain the tissue-specific identities of macrophages.
-Immunity.
-2018 Aug 21;49(2):312-25.
-	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
-	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. In Gene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
-	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L.
-Trajectory-based differential expression analysis for single-cell sequencing data.
-BioRxiv.
-2019 Jan 1:623397.
-	\item Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, Robinson MD.
-Essential guidelines for computational method benchmarking.
-Genome biology.
-2019 Dec;20(1):125.
-	\item Lorenzi L, \ldots, \textbf{Cannoodt R}, \ldots, Mestdagh P. The RNA-Atlas, a single nucleotide resolution map of the human transcriptome. In preparation.
-	\item Van den Berge K, Roux de Bézieux H, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S. Trajectory-based differential expression analysis. Submitted to Nature Communications.
-	\item Van de Sande Bram, \ldots, \textbf{Cannoodt R}, \ldots, Saeys Y, Aerts S. A scalable SCENIC workflow for single-cell gene regulatory network analysis. Submitted to Nature Protocols.
-\end{itemize}
-
-\subsection{Open-source software}
-As part of this work, many open-source software packages were created and many others were contributed to (Table \ref{tab:packages}).
-
-Packages that were created as part of this work are hosted on Github under the username rcannood\footnote{\url{https://github.com/rcannood?tab=repositories}} or the dynverse organisation\footnote{\url{https://github.com/dynverse?tab=repositories}}.
-As part of our standard development practices, we automate execution of unit tests and write extensive documentation to ensure the code complies with CRAN policy before submission.
-We aim to submit all other packages to CRAN as well.
-
-We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
-This includes speeding up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
-Several of these package receive millions of downloads per year (devtools, remotes, rlang).
-
-\newcommand{\cranpkg}[1]{\href{https://cran.r-project.org/package=#1}{#1}}
-\newcommand{\biocpkg}[1]{\href{https://bioconductor.org/packages/#1}{#1}}
-\newcommand{\biocpkgl}[2]{\href{https://bioconductor.org/packages/#1}{#2}}
-\newcommand{\githubpkg}[2]{\href{https://github.com/#1/#2}{#2}}
-\newcommand{\notavailable}{}
-
-\begin{table}[ht!]
-	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor.
-Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-09-10.
-CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
-In addition, many of the dynverse packages have only recently been published on CRAN.
-For Github repositories, no download statistics could be retrieved.
-} \label{tab:packages}
-
-	\centering\fontsize{9}{11}\selectfont
-	\begin{tabularx}{\linewidth}{|p{2cm}llp{1.5cm}X|}
-		\hline
-		Name & Role & Host & Downloads per year & Description \\ \hline\hline
-		\cranpkg{babelwhale} & aut & CRAN & 5810 & Interacting with Docker and Singularity containers \\
-		\cranpkg{diffusionMap} & aut & CRAN & 30'450 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
-		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
-		\cranpkg{dyndimred} & aut & CRAN & 5535 & Applying dimensionality reduction methods \\
-		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
-		\githubpkg{dynverse}{dynfeature} & aut & Github & \notavailable & Calculating feature importance scores from trajectories \\
-		\githubpkg{dynverse}{dyngen} & aut & Github & \notavailable & Simulating single-cell data using gene regulatory networks \\
-		\githubpkg{dynverse}{dynguidelines} & aut & Github & \notavailable & User guidelines for trajectory inference \\
-		\githubpkg{dynverse}{dynmethods} & aut & Github & \notavailable & A collection of wrappers for trajectory inference methods \\
-		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
-		\cranpkg{dynparam} & aut & CRAN & 3265 & Creating meta-information for parameters \\
-		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
-		\githubpkg{dynverse}{dynplot2} & aut & Github & \notavailable & A fully customisable visualisation library for trajectories \\
-		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
-		\cranpkg{dynutils} & aut & CRAN & 13'13' & Common functionality for the dynverse packages \\
-		\cranpkg{dynwrap} & aut & CRAN & 990 & A common format for trajectories \\
-		\cranpkg{GillespieSSA} & aut & CRAN & 7880 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
-		\cranpkg{GillespieSSA2} & aut & CRAN & 4950 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
-		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & An Rcpp implementation of the Growing Neural Gas algorithm \\
-		\cranpkg{incgraph} & aut & CRAN & 3565 & Incremental graphlet counting for network optimisation \\
-		\cranpkg{lmds} & aut & CRAN & 815 & Landmark Multi-Dimensional Scaling \\
-		\cranpkg{princurve} & aut & CRAN & 29'100 & Fits a principal curve in arbitrary dimension \\
-		\cranpkg{proxyC} & aut & CRAN & 117'480 & Computes proximity in large sparse matrices \\
-		\cranpkg{qsub} & aut & CRAN & 3585 & Running commands remotely on gridengine clusters \\
-		\cranpkg{SCORPIUS} & aut & CRAN & 4520 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
-
-		\cranpkg{badger} & ctb & CRAN & 6240 & Query information and generate badge for using in README and GitHub Pages \\
-		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 935 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
-		\cranpkg{devtools} & ctb & CRAN & 5'918'700 & Tools to make developing R packages easier \\
-		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
-		\cranpkg{mlr} & ctb & CRAN & 176'330 & Machine Learning in R \\
-		\biocpkg{monocle} & ctb & Bioc & 34'360 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
-		\cranpkg{ParamHelpers} & ctb & CRAN & 150'775 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
-		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
-		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
-		\cranpkg{remotes} & ctb & CRAN & 3'944'090 & R package installation from remote repositories, including GitHub \\
-		\cranpkg{rlang} & ctb & CRAN & 13'269'115 & Functions for base types and core R and tidyverse features \\
-		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
-		\biocpkg{slingshot} & ctb & Bioc & 12'085 & Tools for ordering single-cell sequencing \\
-		\biocpkg{splatter} & ctb & Bioc & 5015 & Simple simulation of single-cell RNA sequencing data \\
-		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
-		\githubpkg{ManuSetty}{wishbone} & ctb & Github & \notavailable & Identify bifurcating developmental trajectories from single-cell data \\\hline
-	\end{tabularx}
-\end{table}

### 2019-10-15 chX_backmatter.tex
update
~

@@
 
+\subsection{First-author contributions}
+
+\begin{itemize}
+	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
+SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
+bioRxiv 079509.
+2016 Oct.
+	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
+	\item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
+	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
+	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
+	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
+	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
+	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
+	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
+\end{itemize}
+
+*: Equal contribution.
+
+\subsection{Co-author publications}
+\begin{itemize}
+	\item Decock A, Ongenaert M, \textbf{Cannoodt R}, Verniers K, De Wilde B, Laureys G, Van Roy N, Berbegall AP, Bienertova-Vasku J, Bown N, Clément N.
+Methyl-CpG-binding domain sequencing reveals a prognostic methylation signature in neuroblastoma.
+Oncotarget.
+2016 Jan 12;7(2):1960.
+	\item Van Cauwenbergh C, Van Schil K, \textbf{Cannoodt R}, Bauwens M, Van Laethem T, De Jaegere S, Steyaert W, Sante T, Menten B, Leroy BP, Coppieters F.
+arrEYE: a customized platform for high-resolution copy number analysis of coding and noncoding regions of known and candidate retinal dystrophy genes and retinal noncoding RNAs.
+Genetics in Medicine.
+2017 Apr;19(4):457.
+	\item Claeys S, Denecker G, \textbf{Cannoodt R}, Kumps C, Durinck K, Speleman F, De Preter K.
+Early and late effects of pharmacological ALK inhibition on the neuroblastoma transcriptome.
+Oncotarget.
+2017 Dec 5;8(63):106820.
+	\item Depuydt P, Boeva V, Hocking TD, \textbf{Cannoodt R}, Ambros IM, Ambros PF, Asgharzadeh S, Attiyeh EF, Combaret V, Defferrari R, Fischer M.
+Genomic amplifications and distal 6q loss: novel markers for poor survival in high-risk neuroblastoma patients.
+JNCI: Journal of the National Cancer Institute.
+2018 Mar 5;110(10):1084-93.
+	\item Scott CL, T'Jonck W, \ldots, \textbf{Cannoodt R}, Saelens W \ldots, Guilliams M.
+The transcription factor ZEB2 is required to maintain the tissue-specific identities of macrophages.
+Immunity.
+2018 Aug 21;49(2):312-25.
+	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
+	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. In Gene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
+	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L.
+Trajectory-based differential expression analysis for single-cell sequencing data.
+BioRxiv.
+2019 Jan 1:623397.
+	\item Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, Robinson MD.
+Essential guidelines for computational method benchmarking.
+Genome biology.
+2019 Dec;20(1):125.
+	\item Lorenzi L, \ldots, \textbf{Cannoodt R}, \ldots, Mestdagh P. The RNA-Atlas, a single nucleotide resolution map of the human transcriptome. In preparation.
+	\item Van den Berge K, Roux de Bézieux H, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S. Trajectory-based differential expression analysis. Submitted to Nature Communications.
+	\item Van de Sande Bram, \ldots, \textbf{Cannoodt R}, \ldots, Saeys Y, Aerts S. A scalable SCENIC workflow for single-cell gene regulatory network analysis. Submitted to Nature Protocols.
+\end{itemize}
+
+\subsection{Open-source software}
+As part of this work, many open-source software packages were created and many others were contributed to (Table \ref{tab:packages}).
+
+Packages that were created as part of this work are hosted on Github under the username rcannood\footnote{\url{https://github.com/rcannood?tab=repositories}} or the dynverse organisation\footnote{\url{https://github.com/dynverse?tab=repositories}}.
+As part of our standard development practices, we automate execution of unit tests and write extensive documentation to ensure the code complies with CRAN policy before submission.
+We aim to submit all other packages to CRAN as well.
+
+We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
+This includes speeding up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
+Several of these package receive millions of downloads per year (devtools, remotes, rlang).
+
+\newcommand{\cranpkg}[1]{\href{https://cran.r-project.org/package=#1}{#1}}
+\newcommand{\biocpkg}[1]{\href{https://bioconductor.org/packages/#1}{#1}}
+\newcommand{\biocpkgl}[2]{\href{https://bioconductor.org/packages/#1}{#2}}
+\newcommand{\githubpkg}[2]{\href{https://github.com/#1/#2}{#2}}
+\newcommand{\notavailable}{}
+
+\begin{table}[ht!]
+	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor.
+Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-09-10.
+CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
+In addition, many of the dynverse packages have only recently been published on CRAN.
+For Github repositories, no download statistics could be retrieved.
+} \label{tab:packages}
+
+	\centering\fontsize{9}{11}\selectfont
+	\begin{tabularx}{\linewidth}{|p{2cm}llp{1.5cm}X|}
+		\hline
+		Name & Role & Host & Downloads per year & Description \\ \hline\hline
+		\cranpkg{babelwhale} & aut & CRAN & 5810 & Interacting with Docker and Singularity containers \\
+		\cranpkg{diffusionMap} & aut & CRAN & 30'450 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
+		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
+		\cranpkg{dyndimred} & aut & CRAN & 5535 & Applying dimensionality reduction methods \\
+		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
+		\githubpkg{dynverse}{dynfeature} & aut & Github & \notavailable & Calculating feature importance scores from trajectories \\
+		\githubpkg{dynverse}{dyngen} & aut & Github & \notavailable & Simulating single-cell data using gene regulatory networks \\
+		\githubpkg{dynverse}{dynguidelines} & aut & Github & \notavailable & User guidelines for trajectory inference \\
+		\githubpkg{dynverse}{dynmethods} & aut & Github & \notavailable & A collection of wrappers for trajectory inference methods \\
+		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
+		\cranpkg{dynparam} & aut & CRAN & 3265 & Creating meta-information for parameters \\
+		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
+		\githubpkg{dynverse}{dynplot2} & aut & Github & \notavailable & A fully customisable visualisation library for trajectories \\
+		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
+		\cranpkg{dynutils} & aut & CRAN & 13'13' & Common functionality for the dynverse packages \\
+		\cranpkg{dynwrap} & aut & CRAN & 990 & A common format for trajectories \\
+		\cranpkg{GillespieSSA} & aut & CRAN & 7880 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		\cranpkg{GillespieSSA2} & aut & CRAN & 4950 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
+		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & An Rcpp implementation of the Growing Neural Gas algorithm \\
+		\cranpkg{incgraph} & aut & CRAN & 3565 & Incremental graphlet counting for network optimisation \\
+		\cranpkg{lmds} & aut & CRAN & 815 & Landmark Multi-Dimensional Scaling \\
+		\cranpkg{princurve} & aut & CRAN & 29'100 & Fits a principal curve in arbitrary dimension \\
+		\cranpkg{proxyC} & aut & CRAN & 117'480 & Computes proximity in large sparse matrices \\
+		\cranpkg{qsub} & aut & CRAN & 3585 & Running commands remotely on gridengine clusters \\
+		\cranpkg{SCORPIUS} & aut & CRAN & 4520 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
+
+		\cranpkg{badger} & ctb & CRAN & 6240 & Query information and generate badge for using in README and GitHub Pages \\
+		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 935 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
+		\cranpkg{devtools} & ctb & CRAN & 5'918'700 & Tools to make developing R packages easier \\
+		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
+		\cranpkg{mlr} & ctb & CRAN & 176'330 & Machine Learning in R \\
+		\biocpkg{monocle} & ctb & Bioc & 34'360 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
+		\cranpkg{ParamHelpers} & ctb & CRAN & 150'775 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
+		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
+		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
+		\cranpkg{remotes} & ctb & CRAN & 3'944'090 & R package installation from remote repositories, including GitHub \\
+		\cranpkg{rlang} & ctb & CRAN & 13'269'115 & Functions for base types and core R and tidyverse features \\
+		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
+		\biocpkg{slingshot} & ctb & Bioc & 12'085 & Tools for ordering single-cell sequencing \\
+		\biocpkg{splatter} & ctb & Bioc & 5015 & Simple simulation of single-cell RNA sequencing data \\
+		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
+		\githubpkg{ManuSetty}{wishbone} & ctb & Github & \notavailable & Identify bifurcating developmental trajectories from single-cell data \\\hline
+	\end{tabularx}
+\end{table}
+
 \printbibliography
+
+% thanking:
+% the people who brought me soup and sandwiches (look up names)
+% the people who tried to use the PRISM cluster but couldn't

### 2019-10-15 ch1_introduction/content.tex
reshuffle

--- 2019-10-14 ch1_introduction/content.tex
@@
 \section{List of contributions}
-
 \subsection{First-author publications}

### 2019-10-15 boek.tex
reshuffle
update

--- 2019-10-14 boek.tex
@@
 \usepackage{tcolorbox}
+\usepackage[export]{adjustbox}
 
@@
 \lettergroup{\thechapter}
-\begin{quote}
-	\textbf{Abstract:} 
-	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner.
+%\begin{quote}
+%	\textbf{Abstract:} 
+%	Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in a high-throughput and unbiased manner.
 Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following.
@@
 As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
-\end{quote}
+%\end{quote}
 \vfill
@@
 \graphicspath{{./ch2_dyngen/}}
+%\begin{quote}
+%	\textbf{Abstract:}
+%	A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
+We developed a simulator of \textit{in silico} single cells aimed at being flexible and extendible enough to be used to help kick-start emerging computational sub-domains.
+%	
+%\end{quote}
 \include{ch2_dyngen/content}
@@
 \chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
-\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
+\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods}
 \label{chap:dynbenchmark}
 \graphicspath{{./ch3_dynbenchmark/}}
-\begin{quote}
-	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
+%\begin{quote}
+%	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
 Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
@@
 Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
-\end{quote}
+%\end{quote}
 \vfill
-Adapted from:\\
+%Adapted from:\\
 Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
@@
 \label{chap:dyno}
-\begin{quote}
-	\textbf{Abstract:} Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+%\begin{quote}
+%	\textbf{Abstract:} Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
 Since 2014, at least 75 tools for trajectory inference have been developed.
 However, for end-users with a dataset of interest, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.
-
-	We developed dyno, a set of software tools that allow a user to easily infer, visualise and interpret single-cell trajectories.
+%	
+%	We developed dyno, a set of software tools that allow a user to easily infer, visualise and interpret single-cell trajectories.
 First, the user can select the most optimal set of methods based on the size of the dataset, the prior knowledge on the trajectories' topology and other user preferences.
@@
 For this, we provide a complete toolkit: adding a root and labelling the different cell stages, detecting genes which are differentially expressed at different stages of the trajectory, and plotting the trajectory within any dimensionality reduction or heatmap.
-\end{quote}
-\vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.\\
-{\footnotesize $^*$ Equal contribution}
+%\end{quote}
+%\vfill
+%Adapted from:\\
+%\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.\\
+%{\footnotesize $^*$ Equal contribution}
 \newpage
@@
 \label{chap:scorpius}
-\begin{quote}
-	\textbf{Abstract:}
-\end{quote}
-\vfill
+%\begin{quote}
+%	\textbf{Abstract:} Recent advances in RNA sequencing enable the generation of genome-wide expression data at the single-cell level, opening up new avenues for transcriptomics and systems biology.
+A new application of single-cell transcriptomics is the unbiased ordering of cells according to their progression along a dynamic process of interest.
+SCORPIUS is a toolkit for pseudotemporally ordering single cells along a linear trajectory.
+Comprehensive benchmarking on linear datasets shows that SCORPIUS consistently obtains higher quality linear trajectories in comparison to state-of-the-art trajectory inference methods.
+We used SCORPIUS to generate novel hypotheses regarding dendritic cell development, which were subsequently validated in vivo.
+%\end{quote}
+%\vfill
 Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., Sichien, D., Tavernier, S., Janssens, S.
-Guilliams, M., Lambrecht, B., De Preter, K., and Saeys, Y.
+\textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
 SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
-\textit{Journal} vol, issue (2019), page--page.
-\doi{10.1101/079509v2}.
+bioRxiv 079509.
+2016 Oct.
 \newpage
@@
 \label{chap:bred}
-\begin{quote}
-	\textbf{Abstract:}
-\end{quote}
-\vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., Saeys, Y, and De Preter K. Inferring Single Cell Regulatory Networks. \textit{Journal} vol, issue (2019), page--page. doi.
+%\begin{quote}
+%	\textbf{Abstract:}
+%\end{quote}
 \newpage
@@
 \graphicspath{{./ch7_incgraph/}}
-\begin{quote}
-	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
-	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
-	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
+%\begin{quote}
+%	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
+%	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
+%	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
 \\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
-	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
-	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
-	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
-\end{quote}
+%	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
+%	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
+%	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
+%\end{quote}
 \vfill
-Adapted from:\\
+%Adapted from:\\
 \textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y.
@@
 \graphicspath{{./ch9_selfassessment/}}
-\begin{quote}
-	\textbf{Abstract:}
-\end{quote}
-\vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Self-assessment in trajectory inference. \textit{Journal} vol, issue (2019), page--page. doi.
+%\begin{quote}
+%	\textbf{Abstract:}
+%\end{quote}
+%\vfill
+%Adapted from:\\
+%\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Self-assessment in trajectory inference. \textit{Journal} vol, issue (2019), page--page. doi.
 \newpage
@@
 \graphicspath{{./ch10_guidelines/}}
-\begin{quote}
-	\textbf{Abstract:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
+%\begin{quote}
+%	\textbf{Abstract:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
 Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
@@
 Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
-\end{quote}
+%\end{quote}
 \vfill
-Adapted from:\\
+%Adapted from:\\
 Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,

### 2019-10-15 ch0_frontmatter.tex
update
~
reshuffle

@@
 \begin{titlepage}
+
+	\includegraphics[height=1.5cm]{fig/logos/logo_UGent_EN_RGB_color}
+
 	\centering
-	\ 
+
 	\vfill
@@
 
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Acknowledgements}
-\blindtext
+%\newpage{\thispagestyle{empty}\cleardoublepage}
+%\chapter*{Acknowledgements}
 
@@
 
-\newcommand{\grn}{GRN}
-\newcommand{\grns}{GRNs}
-\nomenclature{GRN}{Gene Regulatory Network}
+\nomenclature{HCA}{Human Cell Atlas}
+\nomenclature{TCGA}{The Cancer Genome Atlas}
 
-\newcommand{\im}{IM}
-\newcommand{\ims}{IMs}
-\nomenclature{IM}{Importance Measure}
-
-\newcommand{\tf}{TF}
-\newcommand{\tfs}{TFs}
+\nomenclature{DNA}{Deoxyribonucleic Acid}
+\nomenclature{RNA}{Ribonucleic Acid}
+\nomenclature{mRNA}{Messenger RNA}
 \nomenclature{TF}{Transcription Factor}
 
-\newcommand{\dna}{DNA}
-\nomenclature{DNA}{Deoxyribonucleic Acid}
+\nomenclature{GRN}{Gene Regulatory Network}
+\nomenclature{NI}{Network Inference}
+\nomenclature{CART}{Classification And Regression Trees}
+\nomenclature{RF}{Random Forests}
+\nomenclature{TI}{Trajectory Inference}
 
-\newcommand{\rna}{RNA}
-\nomenclature{RNA}{Ribonucleic Acid}
+\nomenclature{$k$NN}{$k$-nearest-neighbour}
+\nomenclature{DR}{Dimensionality Reduction}
+\nomenclature{PCA}{Principal Component Analysis}
+\nomenclature{MDS}{Multi-Dimensional Scaling}
+\nomenclature{t-SNE}{t-distributed Stochastic Neighbor Embedding}
+\nomenclature{UMAP}{Uniform Manifold Approximation and Projection}
+\nomenclature{LDMS}{Landmark Multi-Dimensional Scaling}
+\nomenclature{DE}{Differential Expression}
+\nomenclature{TDE}{Trajectory Differential Expression}
 
-\newcommand{\mrna}{mRNA}
-\nomenclature{mRNA}{Messenger RNA}
+\chapter*{Summary}
 
-\newcommand{\nin}{NI}
-\nomenclature{NI}{Network Inference}
+Recent technological developments triggered a major milestone in cellular biology by being able to profile and monitor biomolecular contents of single cells at a genome-wide level.
+The rapid advancements in single-cell omics allow for new types of analyses that fundamentally redefine our understanding of cellular heterogeneity and dynamic processes.
+Even now, there are even more exciting advancements on the horizon, with the Human Cell Atlas aiming to create a biomolecular and spatial map of all the different cell types in the body.
 
-\newcommand{\ml}{ML}
-\nomenclature{ML}{Machine Learning}
+Throughout this doctoral project, I did not only get to follow these inspiring advancements first hand, but was able to make scientific contributions to the field by extensively reviewing novel computational tools, and developing several tools and algorithms of my own.
+More specifically, my efforts have resulted in the following contributions.
 
-\newcommand{\rf}{RF}
-\nomenclature{RF}{Random Forests}
+A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
+We developed a simulator of \textit{in silico} single cells aimed at being flexible and extendible enough to be used to help kick-start emerging computational sub-domains (Chapter~\ref{chap:dyngen}).
 
-\newcommand{\cart}{CART}
-\nomenclature{CART}{Classification And Regression Trees}
+We performed a comprehensive comparison of 45 trajectory inference methods (Chapter~\ref{chap:dynbenchmark}).
+Trajectory inference is a new type of computational analysis whose aim is to identify and study transitions between main cellular states.
+While trajectory inference is one of the most rapidly growing new fields in single-cell omics tools, a quantitative study of the advantages and disadvantages was hitherto lacking.
+
+We developed four new software tools ourselves.
+These methods include an extensive toolkit for inferring, visualising and interpreting trajectories (Chapter~\ref{chap:dyno}), a robust tool for inferring linear trajectories (Chapter~\ref{chap:scorpius}), a novel way for studying gene regulation at a single-cell or single-sample level (Chapter~\ref{chap:bred}), and tool for quantifying the topological changes resulting from minor perturbations in a network (Chapter~\ref{chap:incgraph}).
+
+We reflect on current challenges within the field, by providing guidelines for how to self-assess trajectory inference methods (Chapter~\ref{chap:selfassessment}) and, more generally, how to perform large-scale benchmarking experiments of computational methods (Chapter~\ref{chap:guidelines}).
+
+Throughout this project I particularly enjoyed working in an open-source ecosystem, collaborating with scientists around the world to tackle unsolved problems, whilst learning best practices by inspecting the source code of exemplary software projects.
+
+\chapter*{Samenvatting}
+
+Recente technologische ontwikkelingen hebben geleid tot significante ontwikkelingen in celbiologie die ons toelieten om de biomoleculaire inhoud van individuele cellen op een genoomwijde manier te observeren.
+Dankzij snelle ontwikkelingen in \textit{single-cell omics} technologieën wordt ons verstaan van cellulaire heterogeniteit en dynamische processen voortdurend gewijzigd.
+Inmiddels zijn er nieuwe spannende ontwikkelingen onderweg, zoals het \textit{Human Cell Atlas} project, die mikt om alle celtypes in het menselijke lichaam op biomoleculaire en spatiële wijze in kaart te brengen.
+
+Gedurende dit doctoraatsproject heb ik deze inspirerende ontwikkelingen niet alleen van heel dichtbij kunnen volgen, maar ik heb ook wetenschappelijke bijdragen aan het veld kunnen leveren door nieuwe analytische programma’s extensief te onderzoeken, en door zelf verscheidene algoritmes te ontwikkelen.
+Meer specifiek heeft mijn onderzoek geleid tot de volgende bijdragen.
+
+Een recurrent probleem van baanbrekende computationele algoritmes is dat gedurende hun ontwikkeling er helaas vaak onvoldoende datasets beschikbaar zijn om de performantie op een kwantitatieve wijze te kunnen beoordelen.
+We ontwikkelden een programma om individuele cellen te simuleren, dat flexibel en uitbreidbaar genoeg is om toekomstige computationele onderzoeksvelden te ondersteunen (Hoofdstuk~\ref{chap:dyngen}).
+
+We voerden een comprehensieve studie van 45 \textit{trajectvoorspellende} methoden uit (Hoofdstuk~\ref{chap:dynbenchmark}).
+Trajectvoorspellende methoden is een nieuwe soort van computationele analyse waarbij het doel is om transities tussen verscheidene celtypes te identificeren en bestuderen.
+Terwijl deze klasse van methoden een van de snelst groeiende subdomeinen binnen \textit{single-cell omics} was, was er tot op dat moment nog geen kwantitatieve studie van de voor- en nadelen van dergelijke methoden uitgevoerd.
+
+We ontwikkelde ook zelf vier nieuwe computer programma's, waaronder een uitgebreide toolkit om trajecten te voorspellen, visualiseren en interpreteren (Hoofdstuk~\ref{chap:dyno}), een robuust programma om lineaire trajecten te voorspellen (Hoofdstuk~\ref{chap:scorpius}), een nieuwe manier om gen regulatie op het niveau van een individuele cel of staal te bestuderen (Hoofdstuk~\ref{chap:bred}), en een programma om topologische veranderingen te kwantificeren ten gevolge van kleine veranderingen in een netwerk (Hoofdstuk~\ref{chap:incgraph}).
+
+We reflecteren op huidige uitdagingen binnen het veld, door richtlijnen aan te bieden over hoe je je eigen trajectvoorspellende methode kan evalueren (Hoofdstuk~\ref{chap:selfassessment}) en, in het algemeen, hoe je grootschalige beoordelingen van computationele methoden kan uitvoeren (Hoofdstuk~\ref{chap:guidelines}).
+
+Tijdens het verloop van dit project heb ik bijzonder genoten van te werken in een \textit{open-source} omgeving, waarbij ik wereldwijd collaboraties kon aangaan met onderzoekers.
+Hierbij heb ik bijgeleerd door naar de broncode van voorbeeldige softwareprojecten te kunnen kijken.

### 2019-10-15 ch1_introduction/content.tex
update

@@
 	\centering
-	\includegraphics[width=.8\linewidth]{fig/rnaworld} % TODO: change image
+	\includegraphics[width=.8\linewidth]{fig/rnaworld}
 	\caption{
 		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life.
-Image from Horning (2011)\cite{horning_rnaworld_2011}.
-%		\fbox{\color{red} TODO: combine images, rewrite the description.}
+Adapted from Horning (2011)\cite{horning_rnaworld_2011}.
 	}
@@
 	\centering
-	\includegraphics[width=.7\linewidth]{fig/central_dogma} % TODO: change image
+	\includegraphics[width=.8\linewidth]{fig/central_dogma.pdf}
 	\caption{	
-		\textbf{Central Dogma.}
-%		\fbox{\color{red} TODO: combine images, rewrite the description.}
+		\textbf{The central dogma of molecular biology explains the flow of genetic information, from DNA to RNA, to protein.}
 	}
@@
 
-\begin{figure}[htb!]
-	\centering\includegraphics[width=.8\linewidth]{fig/control_gene_expression} % TODO: change image
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=\linewidth]{fig/control_gene_expression}
 	\caption{
 		\textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e.
-splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity \cite{alberts_overviewgenecontrol_2002}.
+splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity\cite{alberts_overviewgenecontrol_2002}.
 	}
@@
 Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
-	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
+	\item We explore a novel algorithm for \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
 We demonstrate our NI methodology by inferring regulatory interactions for 22'122 hematopoietic cells, as well as for 14'963 bulk profiles of cancer tumours.
@@
 	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor.
-Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-09-10.
+Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-10-14.
 CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.

### 2019-10-15 ch3_dynbenchmark/content.tex
update

@@
 \begin{itemize}
-	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
-	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
+	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \url{https://github.com/dynverse/dyngen}
+	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \url{https://github.com/dynverse/dyntoy}
 	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulos_prossttprobabilisticsimulation_2018}
@@
 
-For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (\url{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}) .
+For more simplistic data generation ("toy" datasets), we created the dyntoy workflow (\url{https://github.com/dynverse/dyntoy}).
 We created 12 topology generators (described below), and with 10 datasets per generator, this lead to a total of 120 datasets.
@@
 
-All metrics described here were implemented within the \href{https://github.com/dynverse/dyneval}{dyneval R package (https://github.com/dynverse/dyneval)}.
+All metrics described here were implemented within the dyneval R package (\url{https://github.com/dynverse/dyneval}).
 

### 2019-10-15 ch8_discussion/content.tex
update

@@
 \section{Inferring single cell regulatory networks}
-\mycomment{TODO: fill in when chapter has been written.}
-%	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
-We demonstrate our NI methodology by inferring regulatory interactions for 25'000 hematopoietic cells, as well as for 15'000 cancer-related bulk profiles.
-% TODO: was this chapter written in the end?
 
-%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>^
+We proposed a novel algorithm for inferring case-wise gene regulatory interactions.
+While we demonstrated the methodology by inferring 22'122 cell-specific GRNs and 14'963 sample-specific GRNs from the Tabula Muris and TCGA projects respectively, further interpretation of these results needs to be perform to further validate the usefulness of the method.
+
+Additional avenues for exploration could involve applying trajectory inference on the case-wise regulome data, and quantitatively evaluating the performance of the algorithm using the \texttt{dyngen} simulations.
 
@@
 \textbf{Git}\cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
-Since Wouter and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel.
+Since Wouter Saelens and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel.
 Only in few cases did we need to intervene manually to merge our changes.
 Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects.
-% Misschien ook Wouter zijn achternaam gebruiken. Klinkt wat professioneler :-p
+
 Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}.

### 2019-10-16 boek.tex
update

--- 2019-10-14 boek.tex
@@
 \label{chap:introduction}
-\graphicspath{{./ch1_introduction/}}
+
 \lettergroup{\thechapter}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:} 
-	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner.
+	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in a high-throughput and unbiased manner.
 Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following.
@@
 \end{quote}
+
 \vfill
-\newpage
-\include{ch1_introduction/content}
+%\newpage
+%\graphicspath{{./ch1_introduction/}}
+%\include{ch1_introduction/content}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dyngen: benchmarking with \textit{in silico} single cells} 
-\chaptermark{dyngen: benchmarking with \textit{in silico} single cells}
+\chapter{Paper 1: Benchmarking with \textit{in silico} single cells} 
+\chaptermark{Paper 1: Benchmarking with \textit{in silico} single cells.}
 \label{chap:dyngen}
-\graphicspath{{./ch2_dyngen/}}
-\include{ch2_dyngen/content}
+
+{\large \textbf{Abstract}}
+\begin{quote}
+	\textbf{Purpose:} A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
+\\
+	\textbf{Results:} We developed \texttt{dyngen}, a multi-modality simulator of single cells. 
+	In dyngen, the biomolecular state of an \textit{in silico} cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
+By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures.
+We demonstrate \texttt{dyngen}'s flexibility by simulating snapshot, time-series and perturbation experiments.\\
+	\textbf{Conclusion:} \texttt{dyngen} lays the foundations for benchmarking a wide variety of computational single-cell tools, and can be used to help kick-start the development of future types of analyses.
+\end{quote}
+
+\vfill
+{\large \textbf{Publication status}}
+\begin{quote}
+	Manuscript in preparation. \\
+	\textbf{Cannoodt R}$^*$, Saelens W$^*$, and Saeys Y.  \\
+	{\footnotesize $^*$ Equal contribution}
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.} and W.S. designed the study.\\
+	$\bullet$ \textbf{R.C.} and W.S. performed the experiments and analysed the data.\\
+	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.} and W.S. reviewed and edited the manuscript. \\
+	$\bullet$ Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+%\newpage
+%\graphicspath{{./ch2_dyngen/}}
+%\include{ch2_dyngen/content}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
-\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
+\chapter{Paper 2: A comparison of single-cell trajectory inference methods} 
+\chaptermark{Paper 2: A comparison of single-cell trajectory inference methods.}
 \label{chap:dynbenchmark}
-\graphicspath{{./ch3_dynbenchmark/}}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
-Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
-Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
+	\textbf{Purpose:} Trajectory inference approaches analyse genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
+Although more than 75 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
+\\
+	\textbf{Results:} Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
 Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
-Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
-Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
+We develop a set of guidelines to help users select the best method for their dataset.
+\\
+	\textbf{Conclusion:}  Our freely available data and evaluation pipeline will aid in the development of improved tools designed to analyse increasingly large and complex single-cell datasets.
 \end{quote}
+
 \vfill
-Adapted from:\\
-Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y.
-A comparison of single-cell trajectory inference methods.
-\textit{Nature Biotechnology} 37, 5 (2019), 547--554.
-\doi{10.1038/s41587-019-0071-9}.\\
+{\large \textbf{Publication status}}
+\begin{quote}
+	Published in Nature Biotechnology 37, 5 (2019). \doi{10.1038/s41587-019-0071-9}. \\
+	Saelens W$^*$, \textbf{Cannoodt R}$^*$, Todorov H, and Saeys Y. \\
 	{\footnotesize $^*$ Equal contribution}
-\newpage
-\include{ch3_dynbenchmark/content}
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.}, W.S., H.T. and Y.S. designed the study.\\
+	$\bullet$ \textbf{R.C.} and W.S. performed the experiments and analysed the data.\\
+	$\bullet$ \textbf{R.C.}, W.S., and H.T. implemented software packages.\\
+	$\bullet$ \textbf{R.C.} and W.S. wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.}, W.S., H.T., and Y.S. reviewed and edited the manuscript.\\
+	$\bullet$ Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+%\newpage
+%\graphicspath{{./ch3_dynbenchmark/}}
+%\include{ch3_dynbenchmark/content}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
-\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
+\chapter{Paper 3: A toolkit for inferring and interpreting trajectories} 
+\chaptermark{Paper 3: A toolkit for inferring and interpreting trajectories.} % short title
 \label{chap:dyno}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:} Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory.
+	\textbf{Purpose:} Recent technological advances allow studying dynamic processes by computationally ordering single-cell omics profiles along a trajectory.
 Since 2014, at least 75 tools for trajectory inference have been developed.
-However, for end-users with a dataset of interest, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.
+For end-users, however, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.
+\\
+	\textbf{Results:} We developed \texttt{dyno}, a toolkit for inferring, visualising and interpreting single-cell trajectories, giving access to more than 50 TI methods.
+\texttt{dyno} provides downstream analyses such as visualising the trajectory with dot-plots or heatmaps, automated labelling using gene sets, automated inference of the directionality, and detecting genes that are differentially expressed along the trajectory.
+\\
+	\textbf{Conclusion:} By providing a uniform interface for executing and visualising trajectories of 50 methods, \texttt{dyno} allows inferring and comparing trajectories for a much broader audience than previously possible.
+\end{quote}
 
-	We developed dyno, a set of software tools that allow a user to easily infer, visualise and interpret single-cell trajectories.
-First, the user can select the most optimal set of methods based on the size of the dataset, the prior knowledge on the trajectories' topology and other user preferences.
-Each selected method can then be easily run within a common interface.
-Because the outputs of these methods are all converted into a common consistent format, the trajectories can be easily interpreted.
-For this, we provide a complete toolkit: adding a root and labelling the different cell stages, detecting genes which are differentially expressed at different stages of the trajectory, and plotting the trajectory within any dimensionality reduction or heatmap.
+\vfill
+{\large \textbf{Publication status}}
+\begin{quote}
+	Manuscript in preparation. \\
+	\textbf{Cannoodt R}$^*$, Saelens W$^*$, and Saeys Y.  \\
+	{\footnotesize $^*$ Equal contribution}
 \end{quote}
+
 \vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. dyno: A toolkit for inferring and interpreting trajectories. \textit{Journal} vol, issue (2019), page--page. doi.\\
-{\footnotesize $^*$ Equal contribution}
-\newpage
-\graphicspath{{./ch4_dyno/}}
-\include{ch4_dyno/content}
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.} and W.S. designed the project.\\
+	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
+	$\bullet$ \textbf{R.C.} and W.S. wrote the original manuscript.\\
+	$\bullet$ Y.S. supervised the project.
+\end{quote}
 
-\clearpage
-\section{References}
-\printbibliography[heading=none]
+\vfill
+
+%\newpage
+%\graphicspath{{./ch4_dyno/}}
+%\include{ch4_dyno/content}
+%
+%\clearpage
+%\section{References}
+%\printbibliography[heading=none]
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{SCORPIUS: Fast, accurate, and robust single-cell pseudotime} 
-\chaptermark{SCORPIUS: Fast, accurate, and robust single-cell pseudotime.} % short title
+\chapter{Paper 4: Fast, accurate, and robust single-cell pseudotime} 
+\chaptermark{Paper 4: Fast, accurate, and robust single-cell pseudotime.} % short title
 \label{chap:scorpius}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:}
+	\textbf{Purpose:} 
+	Recent advances in single-cell RNA-sequencing allow modelling progression of single cells along a dynamic process of interest as a trajectory.
+While already more than 75 trajectory inference methods have been developed, comprehensive benchmarking demonstrated that pseudotemporal ordering cells robustly is still a challenging task.\\
+	\textbf{Results:} \texttt{SCORPIUS} is a trajectory inference method specialised in pseudotemporally ordering single-cell profiles.
+Applying \texttt{SCORPIUS} to a large collection of datasets shows that it consistently produces more accurate and robust trajectories in comparison to state-of-the-art trajectory inference methods.
+We used \texttt{SCORPIUS} to generate novel hypotheses regarding dendritic cell development, which were subsequently validated \textit{in vivo}.\\
+	\textbf{Conclusion:} By providing a robust and accurate toolkit for inferring linear trajectories, we aim to make pseudotime analysis as easy-to-use and prevalent as other types of analysis such as differential expression and gene set enrichment.
 \end{quote}
+
 \vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., Sichien, D., Tavernier, S., Janssens, S.
-Guilliams, M., Lambrecht, B., De Preter, K., and Saeys, Y.
-SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
-\textit{Journal} vol, issue (2019), page--page.
-\doi{10.1101/079509v2}.
-\newpage
-\graphicspath{{./ch5_scorpius/}}
-\include{ch5_scorpius/content}
+{\large \textbf{Publication status}}
+\begin{quote}
+	Published in bioRxiv 079509. \doi{10.1038/s41587-019-0071-9}. \\
+	Revised manuscript in preparation. \\
+	\textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, and Saeys Y. 
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.} and Y.S. designed the study.\\
+	$\bullet$ \textbf{R.C.} and W.S. benchmarks and analysed the data.\\
+	$\bullet$ S.T. and S.D. generated experimental data.\\
+	$\bullet$ \textbf{R.C.} implemented the SCORPIUS software package.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.}, J.S., G.M, L.B., K.D.P., and Y.S. reviewed and edited the manuscript. \\
+	$\bullet$ K.D.P. and Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+%\newpage
+%\graphicspath{{./ch5_scorpius/}}
+%\include{ch5_scorpius/content}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{bred: Inferring single cell regulatory networks} 
-\chaptermark{bred: Inferring single cell regulatory networks.} % short title
+\chapter{Paper 5: Inferring single cell regulatory networks} 
+\chaptermark{Paper 5: Inferring single cell regulatory networks.} % short title
 \label{chap:bred}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:}
+	\textbf{Purpose:} \ldots \\
+	\textbf{Results:} \ldots \\
+	\textbf{Conclusion:} \ldots
 \end{quote}
+
 \vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., Saeys, Y, and De Preter K. Inferring Single Cell Regulatory Networks. \textit{Journal} vol, issue (2019), page--page. doi.
-\newpage
-\graphicspath{{./ch6_bred/}}
-\include{ch6_bred/content}
+{\large \textbf{Publication status}}
+\begin{quote}
+	Manuscript in preparation. \\
+	\textbf{Cannoodt R}, Saeys Y$^*$, and De Preter K$^*$. \\
+	{\footnotesize $^*$ Equal contribution}
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.}, K.D.P., and Y.S. designed the study.\\
+	$\bullet$ \textbf{R.C.} performed the experiments and analysed the data.\\
+	$\bullet$ \textbf{R.C.} implemented the bred software package.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.}, K.D.P., and Y.S. reviewed and edited the manuscript. \\
+	$\bullet$ K.D.P. and Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+%\newpage
+%\graphicspath{{./ch6_bred/}}
+%\include{ch6_bred/content}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{incgraph: Optimising regulatory networks} 
-\chaptermark{incgraph: Optimising regulatory networks.} % short title
+\chapter{Paper 6: Optimising regulatory networks} 
+\chaptermark{Paper 6: Optimising regulatory networks.} % short title
 \label{chap:incgraph}
-\graphicspath{{./ch7_incgraph/}}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
-	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
+	\textbf{Purpose:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
+As part of a topology optimisation process, graphlet counts are used to iteratively modify a network in order to achieve certain topological properties.
 	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
-\\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
-	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
-	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
-	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
+\\
+	\textbf{Results:} IncGraph is calculates the differences in graphlet counts with respect to the network in its previous state.
+	IncGraph reduces the execution time by several orders of magnitude when compared to recounting the graphlet counts from scratch at every iteration.
+	We demonstrate the usefulness of this approach optimising gene regulatory networks using a graphlet-based fitness score. \\
+	\textbf{Conclusion:} IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
 \end{quote}
+
 \vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y.
-IncGraph: Incremental graphlet counting for topology optimisation.
-\textit{PLOS ONE} 13, 4 (2018), e0195997.
-\doi{10.1371/journal.pone.0195997}.
-\newpage
-\include{ch7_incgraph/content}
+{\large \textbf{Publication status}}
+\begin{quote}
+	Published in PLOS ONE 13, 4 (2018). \doi{10.1371/journal.pone.0195997}. \\
+	\textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, and Saeys Y. 
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.}, Jo.R., Ja.R., and Y.S. designed the study.\\
+	$\bullet$ \textbf{R.C.}, Jo.R., and Ja.R. performed the experiments and analysed the data.\\
+	$\bullet$ \textbf{R.C.} implemented the incgraph software package.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.}, Jo.R., Ja.R., K.D.P., and Y.S. reviewed and edited the manuscript. \\
+	$\bullet$ Jo.R., K.D.P. and Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+%\newpage
+%\graphicspath{{./ch7_incgraph/}}
+%\include{ch7_incgraph/content}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Self-assessment in trajectory inference} 
-\chaptermark{Self-assessment in trajectory inference.} % short title
+\chapter{Paper 7: Self-assessment in trajectory inference} 
+\chaptermark{Paper 7: Self-assessment in trajectory inference.} % short title
 \label{chap:selfassessment}
-\graphicspath{{./ch9_selfassessment/}}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:}
+	\textbf{Purpose:} \ldots \\
+	\textbf{Results:} \ldots \\
+	\textbf{Conclusion:} \ldots
 \end{quote}
+
 \vfill
-Adapted from:\\
-\textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Self-assessment in trajectory inference. \textit{Journal} vol, issue (2019), page--page. doi.
-\newpage
-\include{ch9_selfassessment/content}
-\clearpage
-\section{References}
-\printbibliography[heading=none]
+{\large \textbf{Publication status}}
+\begin{quote}
+	Manuscript in preparation. \\
+	\textbf{Cannoodt R.}
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.} designed the study.\\
+	$\bullet$ \textbf{R.C.} surveyed 75 manuscripts from literature and analysed the data.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.
+\end{quote}
+
+\vfill
+
+%\newpage
+%\graphicspath{{./ch9_selfassessment/}}
+%\include{ch9_selfassessment/content}
+%\clearpage
+%\section{References}
+%\printbibliography[heading=none]
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Essential guidelines for computational method benchmarking} 
-\chaptermark{Essential guidelines for computational method benchmarking.} % short title
+\chapter{Paper 8: Essential guidelines for computational method benchmarking} 
+\chaptermark{Paper 8: Essential guidelines for computational method benchmarking.} % short title
 \label{chap:guidelines}
-\graphicspath{{./ch10_guidelines/}}
+
+{\large \textbf{Abstract}}
 \begin{quote}
-	\textbf{Abstract:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
+	\textbf{Purpose:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
 Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
 However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results.
-Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
+\\
+	\textbf{Results:}  Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology. \\
+	\textbf{Conclusion:} \ldots
 \end{quote}
+
 \vfill
-Adapted from:\\
-Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
-Boulesteix, A., Saeys, Y., and Robinson, M. D. 
-Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.
-\newpage
-\include{ch10_guidelines/content}
+{\large \textbf{Publication status}}
+\begin{quote}
+	Manuscript published in Genome Biology 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}. \\
+	Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix A, Saeys Y, and Robinson MD 
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.} designed the study.\\
+	$\bullet$ \textbf{R.C.} surveyed 75 manuscripts from literature and analysed the data.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+%\newpage
+%\graphicspath{{./ch10_guidelines/}}
+%\include{ch10_guidelines/content}
 
@@
 \label{chap:discussion}
-\graphicspath{{./ch8_discussion/}}
-\include{ch8_discussion/content}
 
+%\newpage
+%\graphicspath{{./ch8_discussion/}}
+%\include{ch8_discussion/content}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%        BACKMATTER         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \include{chX_backmatter}

### 2019-10-16 ch1_introduction/content.tex
update

@@
 	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
-	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
+%	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
+%	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
+%	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
+%	\item \textbf{Cannoodt R}, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
+%	\item \textbf{Cannoodt R}. Self-assessment in trajectory inference. In preparation.
 \end{itemize}
@@
 		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
-		\cranpkg{dynutils} & aut & CRAN & 13'13' & Common functionality for the dynverse packages \\
+		\cranpkg{dynutils} & aut & CRAN & 13'130 & Common functionality for the dynverse packages \\
 		\cranpkg{dynwrap} & aut & CRAN & 990 & A common format for trajectories \\

### 2019-10-16 boek.tex
include more abstracts

@@
 
-{\large \textbf{Abstract}}
+%{\large \textbf{Abstract}}
 \begin{quote}
@@
 \vfill
-%\newpage
-%\graphicspath{{./ch1_introduction/}}
-%\include{ch1_introduction/content}
+\newpage
+\graphicspath{{./ch1_introduction/}}
+\include{ch1_introduction/content}
 
@@
 
-%\newpage
-%\graphicspath{{./ch2_dyngen/}}
-%\include{ch2_dyngen/content}
+\newpage
+\graphicspath{{./ch2_dyngen/}}
+\include{ch2_dyngen/content}
 
@@
 
-%\newpage
-%\graphicspath{{./ch3_dynbenchmark/}}
-%\include{ch3_dynbenchmark/content}
+\newpage
+\graphicspath{{./ch3_dynbenchmark/}}
+\include{ch3_dynbenchmark/content}
 
@@
 
-%\newpage
-%\graphicspath{{./ch4_dyno/}}
-%\include{ch4_dyno/content}
-%
-%\clearpage
-%\section{References}
-%\printbibliography[heading=none]
+\newpage
+\graphicspath{{./ch4_dyno/}}
+\include{ch4_dyno/content}
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]
 
@@
 
-%\newpage
-%\graphicspath{{./ch5_scorpius/}}
-%\include{ch5_scorpius/content}
+\newpage
+\graphicspath{{./ch5_scorpius/}}
+\include{ch5_scorpius/content}
 
@@
 
-%\newpage
-%\graphicspath{{./ch6_bred/}}
-%\include{ch6_bred/content}
+\newpage
+\graphicspath{{./ch6_bred/}}
+\include{ch6_bred/content}
 
@@
 
-%\newpage
-%\graphicspath{{./ch7_incgraph/}}
-%\include{ch7_incgraph/content}
+\newpage
+\graphicspath{{./ch7_incgraph/}}
+\include{ch7_incgraph/content}
 
@@
 \begin{quote}
-	\textbf{Purpose:} \ldots \\
-	\textbf{Results:} \ldots \\
-	\textbf{Conclusion:} \ldots
+	\textbf{Purpose:} A brief of 75 articles of trajectory tools reveals 
+	that only 37\% of works contain a quantitative self-assessment. 
+	While self-assessments are universally biased in favour of the
+	authors, it is dangerous to publish a computational tool
+	without quantitatively demonstrating its performance compared to
+	state-of-the-art methods. \\
+	\textbf{Results:} We hypothesise that low self-assessment rates are
+	primarily caused by a lack of a standardised problem definition, readily
+	available benchmarking datasets, and suitable metrics. We elaborate on
+	these causal reasons and provide viable solutions for 
+	performing benchmarks more easily.\\
+	\textbf{Conclusion:} With this perspective, we hope to address the
+	elephant in the room, and provide 
+	developers of trajectory inference methods with the necessary 
+	tools for performing benchmarking software quantitatively. 
 \end{quote}
@@
 
-%\newpage
-%\graphicspath{{./ch9_selfassessment/}}
-%\include{ch9_selfassessment/content}
-%\clearpage
-%\section{References}
-%\printbibliography[heading=none]
+\newpage
+\graphicspath{{./ch9_selfassessment/}}
+\include{ch9_selfassessment/content}
+\clearpage
+\section{References}
+\printbibliography[heading=none]
 
@@
 \\
-	\textbf{Results:}  Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology. \\
-	\textbf{Conclusion:} \ldots
+	\textbf{Results:}  Here, we summarise key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
+Our review spans the full pipeline of benchmarking, from defining the scope to best practices for reproducibility.
+This includes crucial questions regarding design and evaluation principles.
+\\
+	\textbf{Conclusion:} With this review, we aim to guide computational researchers in avoiding common pitfalls in designing, performing, and interpreting benchmarks.
 \end{quote}
@@
 \begin{quote}
-	Manuscript published in Genome Biology 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}. \\
+  Published in Genome Biology 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}. \\
 	Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix A, Saeys Y, and Robinson MD 
@@
 \begin{quote}
-	$\bullet$ \textbf{R.C.} designed the study.\\
-	$\bullet$ \textbf{R.C.} surveyed 75 manuscripts from literature and analysed the data.\\
-	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
-	$\bullet$ Y.S. supervised the project.
+	$\bullet$ L.M.W. proposed the project and drafted the manuscript.\\
+	$\bullet$ W.S., \textbf{R.C.}, C.S., A.H., P.P.G., A.L.B., Y.S., and M.D.R. contributed ideas and references and contributed to drafting of the manuscript.\\
+	$\bullet$ Y.S. and M.D.R. supervised the project.
 \end{quote}
@@
 
-%\newpage
-%\graphicspath{{./ch10_guidelines/}}
-%\include{ch10_guidelines/content}
+\newpage
+\graphicspath{{./ch10_guidelines/}}
+\include{ch10_guidelines/content}
 
@@
 
-%\newpage
-%\graphicspath{{./ch8_discussion/}}
-%\include{ch8_discussion/content}
+\newpage
+\graphicspath{{./ch8_discussion/}}
+\include{ch8_discussion/content}
 

### 2019-10-16 ch8_discussion/content.tex
include more abstracts

@@
 Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface.
-Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019}.
+Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019, reimegard_combinedmrnaprotein_2019}.
 

### 2019-10-16 ch0_frontmatter.tex
update thesis

@@
 
-%\newpage{\thispagestyle{empty}\cleardoublepage}
-%\chapter*{Acknowledgements}
-
 \newpage{\thispagestyle{empty}\cleardoublepage}
@@
 \tableofcontents
-
-%\newpage{\thispagestyle{empty}\cleardoublepage}
-%\listoffigures
-
-%\newpage{\thispagestyle{empty}\cleardoublepage}
-%\listoftables
 
@@
 \chapter*{Summary}
-
+\vspace{-4em}
 Recent technological developments triggered a major milestone in cellular biology by being able to profile and monitor biomolecular contents of single cells at a genome-wide level.
@@
 \chapter*{Samenvatting}
-
+\vspace{-4em}
 Recente technologische ontwikkelingen hebben geleid tot significante ontwikkelingen in celbiologie die ons toelieten om de biomoleculaire inhoud van individuele cellen op een genoomwijde manier te observeren.

### 2019-10-16 ch1_introduction/content.tex
update thesis

--- 2019-10-15 ch1_introduction/content.tex
@@
 \end{itemize}
-
-\section{List of contributions}
-\subsection{First-author publications}
-
-\begin{itemize}
-	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
-SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development.
-bioRxiv 079509.
-2016 Oct.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
-	\item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
-	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
-	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
-\end{itemize}
-
-*: Equal contribution.
-
-\subsection{Co-author publications}
-\begin{itemize}
-	\item Decock A, Ongenaert M, \textbf{Cannoodt R}, Verniers K, De Wilde B, Laureys G, Van Roy N, Berbegall AP, Bienertova-Vasku J, Bown N, Clément N.
-Methyl-CpG-binding domain sequencing reveals a prognostic methylation signature in neuroblastoma.
-Oncotarget.
-2016 Jan 12;7(2):1960.
-	\item Van Cauwenbergh C, Van Schil K, \textbf{Cannoodt R}, Bauwens M, Van Laethem T, De Jaegere S, Steyaert W, Sante T, Menten B, Leroy BP, Coppieters F.
-arrEYE: a customized platform for high-resolution copy number analysis of coding and noncoding regions of known and candidate retinal dystrophy genes and retinal noncoding RNAs.
-Genetics in Medicine.
-2017 Apr;19(4):457.
-	\item Claeys S, Denecker G, \textbf{Cannoodt R}, Kumps C, Durinck K, Speleman F, De Preter K.
-Early and late effects of pharmacological ALK inhibition on the neuroblastoma transcriptome.
-Oncotarget.
-2017 Dec 5;8(63):106820.
-	\item Depuydt P, Boeva V, Hocking TD, \textbf{Cannoodt R}, Ambros IM, Ambros PF, Asgharzadeh S, Attiyeh EF, Combaret V, Defferrari R, Fischer M.
-Genomic amplifications and distal 6q loss: novel markers for poor survival in high-risk neuroblastoma patients.
-JNCI: Journal of the National Cancer Institute.
-2018 Mar 5;110(10):1084-93.
-	\item Scott CL, T'Jonck W, \ldots, \textbf{Cannoodt R}, Saelens W \ldots, Guilliams M.
-The transcription factor ZEB2 is required to maintain the tissue-specific identities of macrophages.
-Immunity.
-2018 Aug 21;49(2):312-25.
-	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
-	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. In Gene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
-	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L.
-Trajectory-based differential expression analysis for single-cell sequencing data.
-BioRxiv.
-2019 Jan 1:623397.
-	\item Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, Robinson MD.
-Essential guidelines for computational method benchmarking.
-Genome biology.
-2019 Dec;20(1):125.
-	\item Lorenzi L, \ldots, \textbf{Cannoodt R}, \ldots, Mestdagh P. The RNA-Atlas, a single nucleotide resolution map of the human transcriptome. In preparation.
-	\item Van den Berge K, Roux de Bézieux H, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S. Trajectory-based differential expression analysis. Submitted to Nature Communications.
-	\item Van de Sande Bram, \ldots, \textbf{Cannoodt R}, \ldots, Saeys Y, Aerts S. A scalable SCENIC workflow for single-cell gene regulatory network analysis. Submitted to Nature Protocols.
-\end{itemize}
-
-\subsection{Open-source software}
-As part of this work, many open-source software packages were created and many others were contributed to (Table \ref{tab:packages}).
-
-Packages that were created as part of this work are hosted on Github under the username rcannood\footnote{\url{https://github.com/rcannood?tab=repositories}} or the dynverse organisation\footnote{\url{https://github.com/dynverse?tab=repositories}}.
-As part of our standard development practices, we automate execution of unit tests and write extensive documentation to ensure the code complies with CRAN policy before submission.
-We aim to submit all other packages to CRAN as well.
-
-We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
-This includes speeding up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
-Several of these package receive millions of downloads per year (devtools, remotes, rlang).
-
-\newcommand{\cranpkg}[1]{\href{https://cran.r-project.org/package=#1}{#1}}
-\newcommand{\biocpkg}[1]{\href{https://bioconductor.org/packages/#1}{#1}}
-\newcommand{\biocpkgl}[2]{\href{https://bioconductor.org/packages/#1}{#2}}
-\newcommand{\githubpkg}[2]{\href{https://github.com/#1/#2}{#2}}
-\newcommand{\notavailable}{}
-
-\begin{table}[ht!]
-	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor.
-Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-10-14.
-CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
-In addition, many of the dynverse packages have only recently been published on CRAN.
-For Github repositories, no download statistics could be retrieved.
-} \label{tab:packages}
-
-	\centering\fontsize{9}{11}\selectfont
-	\begin{tabularx}{\linewidth}{|p{2cm}llp{1.5cm}X|}
-		\hline
-		Name & Role & Host & Downloads per year & Description \\ \hline\hline
-		\cranpkg{babelwhale} & aut & CRAN & 5810 & Interacting with Docker and Singularity containers \\
-		\cranpkg{diffusionMap} & aut & CRAN & 30'450 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
-		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
-		\cranpkg{dyndimred} & aut & CRAN & 5535 & Applying dimensionality reduction methods \\
-		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
-		\githubpkg{dynverse}{dynfeature} & aut & Github & \notavailable & Calculating feature importance scores from trajectories \\
-		\githubpkg{dynverse}{dyngen} & aut & Github & \notavailable & Simulating single-cell data using gene regulatory networks \\
-		\githubpkg{dynverse}{dynguidelines} & aut & Github & \notavailable & User guidelines for trajectory inference \\
-		\githubpkg{dynverse}{dynmethods} & aut & Github & \notavailable & A collection of wrappers for trajectory inference methods \\
-		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
-		\cranpkg{dynparam} & aut & CRAN & 3265 & Creating meta-information for parameters \\
-		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
-		\githubpkg{dynverse}{dynplot2} & aut & Github & \notavailable & A fully customisable visualisation library for trajectories \\
-		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
-		\cranpkg{dynutils} & aut & CRAN & 13'13' & Common functionality for the dynverse packages \\
-		\cranpkg{dynwrap} & aut & CRAN & 990 & A common format for trajectories \\
-		\cranpkg{GillespieSSA} & aut & CRAN & 7880 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
-		\cranpkg{GillespieSSA2} & aut & CRAN & 4950 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
-		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & An Rcpp implementation of the Growing Neural Gas algorithm \\
-		\cranpkg{incgraph} & aut & CRAN & 3565 & Incremental graphlet counting for network optimisation \\
-		\cranpkg{lmds} & aut & CRAN & 815 & Landmark Multi-Dimensional Scaling \\
-		\cranpkg{princurve} & aut & CRAN & 29'100 & Fits a principal curve in arbitrary dimension \\
-		\cranpkg{proxyC} & aut & CRAN & 117'480 & Computes proximity in large sparse matrices \\
-		\cranpkg{qsub} & aut & CRAN & 3585 & Running commands remotely on gridengine clusters \\
-		\cranpkg{SCORPIUS} & aut & CRAN & 4520 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
-
-		\cranpkg{badger} & ctb & CRAN & 6240 & Query information and generate badge for using in README and GitHub Pages \\
-		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 935 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
-		\cranpkg{devtools} & ctb & CRAN & 5'918'700 & Tools to make developing R packages easier \\
-		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
-		\cranpkg{mlr} & ctb & CRAN & 176'330 & Machine Learning in R \\
-		\biocpkg{monocle} & ctb & Bioc & 34'360 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
-		\cranpkg{ParamHelpers} & ctb & CRAN & 150'775 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
-		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
-		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
-		\cranpkg{remotes} & ctb & CRAN & 3'944'090 & R package installation from remote repositories, including GitHub \\
-		\cranpkg{rlang} & ctb & CRAN & 13'269'115 & Functions for base types and core R and tidyverse features \\
-		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
-		\biocpkg{slingshot} & ctb & Bioc & 12'085 & Tools for ordering single-cell sequencing \\
-		\biocpkg{splatter} & ctb & Bioc & 5015 & Simple simulation of single-cell RNA sequencing data \\
-		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
-		\githubpkg{ManuSetty}{wishbone} & ctb & Github & \notavailable & Identify bifurcating developmental trajectories from single-cell data \\\hline
-	\end{tabularx}
-\end{table}

### 2019-10-16 ch3_dynbenchmark/content.tex
update thesis

@@
 
-A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1).
+A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}).
 Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_1.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh]
 	\centering\includegraphics[width=\linewidth]{fig/supfigure_1.pdf}
@@
 
-\begin{figure}[htb!]
-	\centering\includegraphics[width=\linewidth]{fig/figure_2.pdf}
+\afterpage{%
+	\clearpage% flush all other floats
+	\ifodd\value{page}
+%	\else% uncomment this else to get odd/even instead of even/odd
+	\expandafter\afterpage% put it on the next page if this one is odd
+	\fi
+	{%
+		\begin{figure}[tbh!]
+			\centering\includegraphics[height=.75\textheight]{fig/figure_2.pdf}
 			\caption{
@@
 \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation.
-Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig.
-2.
+Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}.
 			}
@@
 		\end{figure}
-
-The largest difference between TI methods is whether a method fixes the topology and, if it does not, what kind of topology it can detect.
-We defined seven possible types of topology, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
-Most methods either focus on inferring linear trajectories or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (Figure \ref{fig:figure_2}a).
-
-We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript.
-Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b).
-We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig.
-2), after which we conclude with guidelines for method users and future perspectives for method developers.
-
-\begin{figure}[htb!]
-	\centering\includegraphics[width=\linewidth]{fig/figure_3.pdf}
+		\clearpage
+		\begin{figure}[tbh!]
+			\centering\includegraphics[height=.75\textheight]{fig/figure_3.pdf}
 			\caption{	
@@
 		\end{figure}
-
+		\clearpage
+	}%
+}
+
+The largest difference between TI methods is whether a method fixes the topology and, if it does not, what kind of topology it can detect.
+We defined seven possible types of topology, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs).
+Most methods either focus on inferring linear trajectories or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (Figure \ref{fig:figure_2}a).
+
+We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript.
+Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b).
+We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}), after which we conclude with guidelines for method users and future perspectives for method developers.
+
+\newpage
 \subsection{Accuracy}
 
-We defined several metrics to compare a prediction to a reference trajectory (Supplementary Note 1).
-Based on an analysis of their robustness and conformity to a set of rules (Supplementary Note 1), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures).
+We defined several metrics to compare a prediction to a reference trajectory (Section~\ref{sec:dynb_supn1}).
+Based on an analysis of their robustness and conformity to a set of rules (Section~\ref{sec:dynb_supn1}), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures).
 The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance.
-These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table 2).
+These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}).
 Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}.
@@
 
-\begin{figure}[htb!]
+The performance of a method was strongly dependent on the type of trajectory present in the data (Figure \ref{fig:figure_3}b).
+Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories (Figure \ref{fig:supfigure_3}c).
+This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, whereas those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (Figure \ref{fig:supfigure_3}d).
+This analysis therefore indicates that detecting the right topology is still a difficult task for most of these methods, because methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.
+
+The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods.
+To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods.
+A top model in this case was defined as a model with an overall score of at least 95$\%$ as the best model.
+On all datasets, using one method resulted in getting a top model about 27$\%$ of the time.
+This increased up to 74$\%$ with the addition of six other methods (Figure \ref{fig:figure_4}a).
+The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA.
+We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (Figure \ref{fig:figure_4}b), although in these cases less methods were necessary to obtain at least one top model for a given dataset.
+Altogether, this shows that there is considerable complementarity between the different methods and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori.
+Moreover, it also opens up the possibilities for new ensemble methods that utilize this complementarity.
+
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/supfigure_3.pdf}
@@
 
-The performance of a method was strongly dependent on the type of trajectory present in the data (Figure \ref{fig:figure_3}b).
-Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories (Figure \ref{fig:supfigure_3}c).
-This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, whereas those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (Figure \ref{fig:supfigure_3}d).
-This analysis therefore indicates that detecting the right topology is still a difficult task for most of these methods, because methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.
-
-The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods.
-To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods.
-A top model in this case was defined as a model with an overall score of at least 95$\%$ as the best model.
-On all datasets, using one method resulted in getting a top model about 27$\%$ of the time.
-This increased up to 74$\%$ with the addition of six other methods (Figure \ref{fig:figure_4}a).
-The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA.
-We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (Figure \ref{fig:figure_4}b), although in these cases less methods were necessary to obtain at least one top model for a given dataset.
-Altogether, this shows that there is considerable complementarity between the different methods and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori.
-Moreover, it also opens up the possibilities for new ensemble methods that utilize this complementarity.
-
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_4.pdf}
@@
 We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a).
-As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Supplementary Fig.
-5), and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig.
-2a).
-
-\begin{figure}[htb!]
+As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Figure \ref{fig:supfigure_5}, and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}a).
+
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_4.pdf}
@@
 
+\begin{figure}[tbh!]
+	\centering\includegraphics[width=0.6\linewidth]{fig/supfigure_5.pdf}
+	\caption{
+		\textbf{Agreement between actual values and predictions for execution times and memory usage.}
+		We created a predictive model of the running time and memory usage based on a set of scaling datasets (left), and validated this model based on the similarity of the predictions and actual values on all benchmark datasets (right).
+Shown are the values for each method and dataset (n = 65618 for training, n = 11939 for test).
+Top left indicates the Pearson correlation coefficient.
+	}
+	\label{fig:supfigure_5}
+\end{figure}
+
 We found that the scalability of most methods was overall very poor, with most graph and tree methods not finishing within an hour on a dataset with ten thousand cells and ten thousand features (Figure \ref{fig:figure_3}c), which is around the size of a typical droplet-based single-cell dataset \cite{svensson_exponentialscalingsinglecell_2018}.
@@
 
-We also assessed the memory requirements of each method (Supplementary Fig.
-2c).
+We also assessed the memory requirements of each method (Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}c).
 Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features.
@@
 Altogether, the scalability analysis indicated that the dimensions of the data are an important factor in the choice of method, and that method development should pay more attention to maintaining reasonable running times and memory usage.
-§
+
 \subsection{Stability}
@@
 
-We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Supplementary Fig.
-6).
-While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Supplementary Fig.
-6 right) and we believe that these should receive extra attention from developers.
+We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Figure~\ref{fig:supfigure_6}).
+While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Figure~\ref{fig:supfigure_6} right) and we believe that these should receive extra attention from developers.
 Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_bestpracticesscientific_2014,artaza_top10metrics_2016}.
@@
 
-\begin{table}
+\begin{figure}[tbh!]
+	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_6.pdf}
+	\caption{
+		\textbf{Usability of trajectory inference methods.}
+		Shown is the score given for each method on every item from the usability score sheet(Table~\ref{tab:scoresheet}).
+Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score.
+Within each  category,  each  aspect  also  received  a  weight  depending  on  how  often  it  was mentioned  in  a set  of  papers  discussing  good practices in tool development and evaluation.
+This is represented in the plot as the height on the y-axis.
+Top: Average usability score for each method.
+Right: The average score of each quality control item.
+Shown into more detail are those items which had an average score lower than 0.5.
+	}
+	\label{fig:supfigure_6}
+\end{figure}
+
+\begin{table}[tbh!]
 	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_5.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/figure_6.pdf}
@@
 
-We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
+We gathered a list of 71 trajectory inference tools (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
 We welcome any contributions by creating an issue at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.
@@
 
-We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table 2).
+We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}).
 The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}).
@@
 We therefore defined several possible metrics for comparing trajectories, each investigating different layers.
-These are all discussed in Supplementary Note 1 along with examples and robustness analyses when appropriate.
-
-Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (Supplementary Note 1).
+These are all discussed in Section~\ref{sec:dynb_supn1} along with examples and robustness analyses when appropriate.
+
+Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (Section~\ref{sec:dynb_supn1}).
 Based on this analysis, we chose four metrics for the evaluation, each assessing a different aspect of the trajectory: (1) the HIM measures the topological similarity; (2) the F1branches compares the branch assignment; (3) the cordist assesses the similarity in pairwise cell–cell distances and thus the cellular positions; and (4) the wcorfeatures looks at whether similar important features (genes) are found in both the reference dataset and the prediction.
@@
 The latter has a parameter $\gamma$, which was fixed at 0.1 to make the scores comparable between datasets.
-We illustrate the metric and discuss alternatives in Supplementary Note 1.
+We illustrate the metric and discuss alternatives in Section~\ref{sec:dynb_supn1}.
 
@@
 The F1branches was then defined as the harmonic mean between Recovery and Relevance.
-We illustrate this metric further in Supplementary Note 1.
+We illustrate this metric further in Section~\ref{sec:dynb_supn1}.
 
@@
 We then calculated the geodesic distances between the union of waypoint cells from both datasets and all other cells.
-The calculation of the geodesic distance depended on the location of the two cells within the trajectory, further discussed in Supplementary Note 1, and was weighted by the length of the edge in the milestone network.
+The calculation of the geodesic distance depended on the location of the two cells within the trajectory, further discussed in Section~\ref{sec:dynb_supn1}, and was weighted by the length of the edge in the milestone network.
 Finally, the cordist was defined as the Spearman rank correlation between the distances of both datasets.
-We illustrate the metric and assess the effect of the number of waypoint cells in Supplementary Note 1.
+We illustrate the metric and assess the effect of the number of waypoint cells in Section~\ref{sec:dynb_supn1}.
 
@@
 As hyperparameters we set the number of trees to 10,000 and the number of features on which to split to 1$\%$ of all available features.
-We illustrate this metric and assess the effect of its hyperparameters in Supplementary Note 1.
+We illustrate this metric and assess the effect of its hyperparameters in Section~\ref{sec:dynb_supn1}.
 
@@
 
-To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics. This aggregation strategy is explained in more detail in Supplementary Note 1.
+To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics.
+This aggregation strategy is explained in more detail in Section~\ref{sec:dynb_supn1}.
 
@@
 
-Further information on research design is available in the Nature Research Reporting Summary, available at \href{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}
-
-\section{Supplementary Note 1: Metrics to compare two trajectories}
+Further information on research design is available in the Nature Research Reporting Summary, available at \href{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}.
+
+\section{Supplementary Note 1: Metrics to compare two trajectories} \label{sec:dynb_supn1}
 
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_1.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_2.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_3.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_4.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_5.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_6.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_7.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_8.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_9.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_10.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=\linewidth]{fig/snote1fig_11.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_12.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_13.pdf}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[tbh!]
 	\centering\includegraphics[width=0.8\linewidth]{fig/snote1fig_14.pdf}

### 2019-10-16 chX_backmatter.tex
reshuffle
update
update thesis

@@
 
-\backmatter
+%\backmatter
 
-\chapter{Samenvatting}
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{Curriculum Vitae}
+\section{Personalia}
+\begin{tabular}{ll}
+	Name & Robrecht Cannoodt \\
+	Date of Birth & January 25th, 1990 \\
+	Place of birth & Ghent \\
+	Nationality & Belgian \\
+	Address & Kerkstraat 75, 9070 Destelbergen \\
+	Email & \href{mailto:robrecht.cannoodt@gmail.com}{robrecht.cannoodt@gmail.com}
+\end{tabular}
 
-\chapter{Summary}
+\section{Professional Experience}
+\textit{Ph.D. student in Bioinformatics} \\
+\hspace*{2 ex} Ghent University, Belgium, September 2013 -- now \\
+\textit{Laboratory assistant (summer job)} \\
+\hspace*{2 ex} Bloedtransfusiecentrum Gent, Belgium, July (2009 -- 2012)
 
-\chapter{List of Publications}
+\section{Education} 
+\textit{Master of Science in Computer Science Engineering: Software Engineering} \\
+\hspace*{2 ex} Ghent University, Belgium, 2011 -- 2013 \vspace*{1mm}\\ 
+\textit{Bachelors degree in Informatics}\\
+\hspace*{2 ex} Ghent University, Belgium, 2008 -- 2011 \vspace*{1mm} \\
+\textit{Bachelors degree in Engineering: Architecture} \\
+\hspace*{2 ex} Ghent University, Belgium, 2007 -- 2008  \vspace*{1mm}\\
+\textit{International Baccalaureate} \\
+\hspace*{2 ex} International School of Berne, Switzerland, 2002 -- 2007
 
-\subsection{First-author contributions}
+\section{First-author publications}
 
@@
 	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
-	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
-	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
-	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
+	%	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
+SCORPIUS: Fast, accurate, and robust single-cell pseudotime.
+In preparation.
+	%	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
+	%	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
+	%	\item \textbf{Cannoodt R}, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
+	%	\item \textbf{Cannoodt R}. Self-assessment in trajectory inference. In preparation.
 \end{itemize}
@@
 
-\subsection{Co-author publications}
+\section{Co-author publications}
 \begin{itemize}
@@
 	\item Van den Berge K, Roux de Bézieux H, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S. Trajectory-based differential expression analysis. Submitted to Nature Communications.
-	\item Van de Sande Bram, \ldots, \textbf{Cannoodt R}, \ldots, Saeys Y, Aerts S. A scalable SCENIC workflow for single-cell gene regulatory network analysis. Submitted to Nature Protocols.
+	\item Van de Sande B, \ldots, \textbf{Cannoodt R}, \ldots, Saeys Y, Aerts S. A scalable SCENIC workflow for single-cell gene regulatory network analysis. Submitted to Nature Protocols.
 \end{itemize}
 
-\subsection{Open-source software}
+\section{Open-source software}
 As part of this work, many open-source software packages were created and many others were contributed to (Table \ref{tab:packages}).
@@
 		\githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
+		\cranpkg{ranger} & ctb & CRAN & 413'641 & A Fast Implementation of Random Forests \\
 		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
@@
 
-\printbibliography
+%\subsection{Oral presentations}
+%
+%\subsection{Poster presentations}
+%
+%\subsection{Awards}
+%
+%\subsection{Courses / Workshops}
+%
+%\subsection{Conferences}
+%
+%\subsection{Student supervision}
+%
+%\subsection{Teaching}
+
+\section{Sources of funding}
+This study was supported by the Fonds Wetenschappelijk Onderzoek (11Y6218N).
+
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter*{Acknowledgements}
 

### 2019-10-16 ch1_introduction/content.tex
change cites

@@
 Due to the dynamic nature of these processes, it can be very challenging to comprehend the cellular response to a signal.
-A reductionist approach to understanding a complex biological system is to study the biochemical components of which it is comprised\cite{brigandt_reductionismbiology_2017}.
+A reductionist approach to understanding a complex biological system is to study the biochemical components of which it is comprised \cite{brigandt_reductionismbiology_2017}.
 
@@
 The sheer volume of the data generated from such highly-integrative and high-throughput experiments are not the only reason why they are so challenging to interpret.
-For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs\cite{hon_humancellatlas_2018}.
+For instance, the generated data contains high levels of noise arising from inherent biomolecular stochasticity in the cells and from the experimental profiling techniques used, as well as batch effects arising from differences between donors and labs \cite{hon_humancellatlas_2018}.
 Biologists thus turn to computer scientists
@@
 Observing the biomolecular insides of cells can ultimately provide fundamental understanding into the processes that govern these cells and help uncover novel approaches for disease diagnosis, prognosis, and treatment.
-For example, the Human Cell Atlas (HCA) consortium\cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
+For example, the Human Cell Atlas (HCA) consortium \cite{regev_humancellatlas_2018} has set out to develop a comprehensive reference map of all the different types of cells in the human body.
 Experts in the field often metaphorically describe the HCA initiative as aiming to develop a 'Google Maps' of the human body.
-Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs\cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.
+Even in its infancy, the HCA has profiled 3.8 million cells from 248 donors across 42 labs \cite{humancellatlasconsortium_humancellatlas_2018}, and this number is likely to increase well above one hundred million.
 
@@
 \subsection{The origin of life and the RNA world}
-The discovery of the double helix shape of deoxyribonucleic acid (DNA)\cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origin of life and evolution.
+The discovery of the double helix shape of deoxyribonucleic acid (DNA) \cite{watson_molecularstructurenucleic_1953} is often considered the pivot point in our understanding of the origin of life and evolution.
 By now, it is well known that DNA serves as a medium for storing the genetic information required to reproduce a whole organism.
@@
 A widely-accepted hypothesis states that life originates from its lesser-known cousin, ribonucleic acid (RNA).
-According to the RNA world hypothesis\cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and to perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}).
+According to the RNA world hypothesis \cite{alberts_rnaworldorigins_2002}, the very first primitive cells used RNA both to store genetic information and to perform the chemical reactions required to sustain themselves (Figure \ref{fig:rnaworld}).
 Only later did cells develop the ability to use the more chemically stable DNA molecules to self-sustain in a process commonly referred to as the central dogma.
@@
 		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life.
-Adapted from Horning (2011)\cite{horning_rnaworld_2011}.
+Adapted from Horning (2011) \cite{horning_rnaworld_2011}.
 	}
@@
 \subsection{Central dogma}
-The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins\cite{strachan_humanmoleculargenetics_2011}.
+The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins \cite{strachan_humanmoleculargenetics_2011}.
 Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).
@@
 A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms.
-This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes\cite{nilsen_expansioneukaryoticproteome_2010}.
+This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes \cite{nilsen_expansioneukaryoticproteome_2010}.
 
@@
 This however, is a major underestimation of the real number of cell types.
-Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types\cite{strachan_humanmoleculargenetics_2011}.
+Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types \cite{strachan_humanmoleculargenetics_2011}.
 
@@
 		\textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e.
-splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity\cite{alberts_overviewgenecontrol_2002}.
+splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity \cite{alberts_overviewgenecontrol_2002}.
 	}
@@
 According to the needs of a cell, different genes are being transcribed.
-Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific or may be expressed in response to developmental and environmental signals\cite{strachan_humanmoleculargenetics_2011}.
+Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific or may be expressed in response to developmental and environmental signals \cite{strachan_humanmoleculargenetics_2011}.
 
@@
 Many TFs are commonly present in virtually all cell types (e.g.
-NF-$\kappa$B), while others are specific for cells and developmental stages\cite{lambert_humantranscriptionfactors_2018}.
+NF-$\kappa$B), while others are specific for cells and developmental stages \cite{lambert_humantranscriptionfactors_2018}.
 Typically, the same TF can regulate the rate of transcription of many target genes in different cell types, indicating that these gene regulatory networks (GRNs) are dynamic.
@@
 The earliest approach for measuring the abundance of a particular molecule in \emph{single cells} is the microscope.
-Since its development by Coons et al. (1941), immunohistochemistry (IHC) has been instrumental in visualising proteins.\cite{coons_immunologicalpropertiesantibody_1941}. 
+Since its development by Coons et al. (1941), immunohistochemistry (IHC) has been instrumental in visualising proteins. \cite{coons_immunologicalpropertiesantibody_1941}. 
 A cell can present a particular type of protein, also called an antigen, on its cell surface.
@@
 Characterising cells in a semi-quantifiable way is labour intensive, however; since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation).
-Modern implementations of IHC improve the throughput drastically by using robots and computer software to provide semi-automated image acquisition and cell segmentation\cite{theodosiou_automatedanalysisfish_2007}.
+Modern implementations of IHC improve the throughput drastically by using robots and computer software to provide semi-automated image acquisition and cell segmentation \cite{theodosiou_automatedanalysisfish_2007}.
 
-Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by measuring fluorescently labelled proteins as cells pass through a fluidic system.
+Flow cytometry \cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by measuring fluorescently labelled proteins as cells pass through a fluidic system.
 Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in blood.
-However, many protocols already exist to extract viable single cells from tissues and tumours\cite{leelatian_preparingviablesingle_2017}.
-Conventional flow cytometry devices enable to measure protein expression levels of millions of cells using up to eight different antibody fluorochromes simultaneously, while state-of-the-art instrumentation allows detection of up to 27 biomarkers simultaneously\cite{cossarizza_guidelinesuseflow_2017}.
+However, many protocols already exist to extract viable single cells from tissues and tumours \cite{leelatian_preparingviablesingle_2017}.
+Conventional flow cytometry devices enable to measure protein expression levels of millions of cells using up to eight different antibody fluorochromes simultaneously, while state-of-the-art instrumentation allows detection of up to 27 biomarkers simultaneously \cite{cossarizza_guidelinesuseflow_2017}.
 
@@
 On the other side of the spectrum are the so-called "omics" technologies.
-"Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
+"Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
 There are at least ten types of "omics".
@@
 
-Increasing evidence shows that cells are biomoleculary heterogeneous, even in very similar cell types\cite{hwang_singlecellrnasequencing_2018} (Figure~\ref{fig:bulk_vs_sc}A).
+Increasing evidence shows that cells are biomoleculary heterogeneous, even in very similar cell types \cite{hwang_singlecellrnasequencing_2018} (Figure~\ref{fig:bulk_vs_sc}A).
 Since a bulk profile is a population average (or rather, a summation), important cell-to-cell variability is not discernible (Figure~\ref{fig:bulk_vs_sc}B).
@@
 Advances in microvolume sequencing allowed profiling the transcriptome at single-cell resolution, thereby bringing single-cell biology and omics together to create single-cell omics.
-During the decade that followed, the number of single-cell omics technologies has sky-rocketed, allowing to profile >100'000 cells\cite{svensson_exponentialscalingsinglecell_2018} and measuring other levels of information (e.g.
+During the decade that followed, the number of single-cell omics technologies has sky-rocketed, allowing to profile >100'000 cells \cite{svensson_exponentialscalingsinglecell_2018} and measuring other levels of information (e.g.
 protein abundance and spatial location) \cite{moudgil_multimodalscrnaseq_2019}.
@@
 Different sampling techniques yield different levels of information about cellular state.
-By now, many protocols for extracting and tagging RNA from single cells have been developed\cite{svensson_exponentialscalingsinglecell_2018}, the most popular of which are based on microfluidics or droplets (Figure~\ref{fig:profiling_convergence}C).
+By now, many protocols for extracting and tagging RNA from single cells have been developed \cite{svensson_exponentialscalingsinglecell_2018}, the most popular of which are based on microfluidics or droplets (Figure~\ref{fig:profiling_convergence}C).
 By sequencing the trancripts and the attached unique cell identifier tags, each read can be mapped and tallied up.
@@
 The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.
-Some of these opportunities lie in efforts like the HCA consortium\cite{regev_humancellatlas_2018}.
+Some of these opportunities lie in efforts like the HCA consortium \cite{regev_humancellatlas_2018}.
 They have set out to redefine all human cell types in both terms of their gene expression and location, and as well as the developmental trajectories connecting the different cell types.
-As part of this endeavour, the consortium will likely profile the whole transcriptomes of tens or even hundreds of millions of cells\cite{humancellatlasconsortium_humancellatlas_2018}.
+As part of this endeavour, the consortium will likely profile the whole transcriptomes of tens or even hundreds of millions of cells \cite{humancellatlasconsortium_humancellatlas_2018}.
 
@@
 Whole-genome profiling at single-cell level allows new types of analyses with which to study cellular heterogeneity at a hitherto unseen throughput. 
-The new types of analyses permitted by single-cell omics present several computational challenges\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
+The new types of analyses permitted by single-cell omics present several computational challenges \cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}.
 This necessitates the development of novel computational tools, either because the problem statement of the performed analysis is completely novel, or to adapt existing methodology to new data characteristics -- dimensionality and noise.
@@
 scRNA-seq data is typically very sparse -- while the human genome has more than 20'000 genes, they only contain non-zero values of a few thousand genes (typically <4'000).
-This is partially due to cells being specialised in particular functions and thus they do not need proteins of every time, but also due to RNA transcription occurring in bursts rather than continuously\cite{nicolas_whatshapeseukaryotic_2017}.
+This is partially due to cells being specialised in particular functions and thus they do not need proteins of every time, but also due to RNA transcription occurring in bursts rather than continuously \cite{nicolas_whatshapeseukaryotic_2017}.
 This contributes to the high levels of noise seen in scRNA-seq data: no two cells have the same set of non-zero genes.
 
-Over the past five years, already 450 new tools have been developed to perform various analyses of single-cell omics data\cite{zappia_exploringsinglecellrnaseq_2018}, taking into account the specific noise characteristics.
+Over the past five years, already 450 new tools have been developed to perform various analyses of single-cell omics data \cite{zappia_exploringsinglecellrnaseq_2018}, taking into account the specific noise characteristics.
 The most frequent types of analyses are detailed in the following subsections.
@@
 Each dimension frequently corresponds to one or more modules of co-expressed genes.
-The reduced space can be interpreted in way analogous to Waddington’s epigenetic landscapes\cite{waddington_strategygenesdiscussion_1957,ferrell_bistabilitybifurcationswaddington_2012,rebhahn_animatedlandscaperepresentation_2014}, where the landscape dictates the possible states a cell can reside in, and transitions between states correspond to dynamic cellular processes, such as cell differentiation.
+The reduced space can be interpreted in way analogous to Waddington’s epigenetic landscapes \cite{waddington_strategygenesdiscussion_1957,ferrell_bistabilitybifurcationswaddington_2012,rebhahn_animatedlandscaperepresentation_2014}, where the landscape dictates the possible states a cell can reside in, and transitions between states correspond to dynamic cellular processes, such as cell differentiation.
 
@@
 		\textbf{Dimensionality reduction for single-cell omics data.}
-		\textbf{A:} A heatmap visualisation of an scRNA-seq expression dataset of fibroblasts being reprogrammed to neuron cells\cite{treutlein_dissectingdirectreprogramming_2016}. Only the most variable 
+		\textbf{A:} A heatmap visualisation of an scRNA-seq expression dataset of fibroblasts being reprogrammed to neuron cells \cite{treutlein_dissectingdirectreprogramming_2016}. Only the most variable 
 		\textbf{B:} The reduced space is a M$\times$K-dimensional matrix which attempts to conserve the cellular heterogeneity of the original space as well as possible.
@@
 
-DR methods can be classified into two main categories, feature projection-based and manifold learning\cite{engel_surveydimensionreduction_2012}.
+DR methods can be classified into two main categories, feature projection-based and manifold learning \cite{engel_surveydimensionreduction_2012}.
 
 Projection-based DR methods aim to perform a transformation of the data while preserving the pairwise distances between samples as much as possible.
-Examples of commonly used projection-based DR methods in single-cell omics are Principal Component Analysis\cite{pearson_liiilinesplanes_1901} (PCA) and Multi-Dimensional Scaling\cite{kruskal_multidimensionalscalingoptimizing_1964} (MDS).
+Examples of commonly used projection-based DR methods in single-cell omics are Principal Component Analysis \cite{pearson_liiilinesplanes_1901} (PCA) and Multi-Dimensional Scaling \cite{kruskal_multidimensionalscalingoptimizing_1964} (MDS).
 
@@
 Manifold learning can be an iterative optimisation process using a predefined criterion.
-Examples of manifold learning techniques are t-distributed Stochastic Neighbor Embedding\cite{vandermaaten_visualizingdatausing_2008} (t-SNE), Diffusion Maps\cite{nadler_diffusion_2005,coifman_diffusionmaps_2006} and Uniform Manifold Approximation and Projection\cite{mcinnes_umapuniformmanifold_2018} (UMAP).
+Examples of manifold learning techniques are t-distributed Stochastic Neighbor Embedding \cite{vandermaaten_visualizingdatausing_2008} (t-SNE), Diffusion Maps \cite{nadler_diffusion_2005,coifman_diffusionmaps_2006} and Uniform Manifold Approximation and Projection \cite{mcinnes_umapuniformmanifold_2018} (UMAP).
 
-For scalability reasons, this work mostly makes use of Landmark MDS\cite{silva_globallocalmethods_2002,lee_landmarkmdsensemble_2009} (LMDS) with a Spearman rank correlation distance metric.
+For scalability reasons, this work mostly makes use of Landmark MDS \cite{silva_globallocalmethods_2002,lee_landmarkmdsensemble_2009} (LMDS) with a Spearman rank correlation distance metric.
 LMDS is an extension of classical MDS, but rather than calculating a complete distance matrix between all pairs of cells, a set of landmark cells is sampled, only the distances between a set of landmarks and the samples are calculated.
@@
 Usually, the number of clusters is determined by the user, either as a direct parameter (e.g.
-$k$-means\cite{lloyd_leastsquaresquantization_1982}) or an indirect parameter (e.g.
+$k$-means \cite{lloyd_leastsquaresquantization_1982}) or an indirect parameter (e.g.
 a height cutoff in hierarchical clustering).
 In some exceptional cases, the number of clusters is strictly determined by the data itself and cannot be altered with a parameter (e.g.
-Louvain clustering\cite{blondel_fastunfoldingcommunities_2008}).
+Louvain clustering \cite{blondel_fastunfoldingcommunities_2008}).
 
@@
 		\textbf{Trajectory inference for single-cell omics data.}
-		\textbf{A:} During a dynamic process cells pass through several transitional states, characterized by different waves of transcriptional, morphological, epigenomic and/or surface marker changes\cite{enver_stemcellstates_2009}.
+		\textbf{A:} During a dynamic process cells pass through several transitional states, characterized by different waves of transcriptional, morphological, epigenomic and/or surface marker changes \cite{enver_stemcellstates_2009}.
 TI methods provide an unbiased approach to identifying and correctly ordering different transitional stages.
@@
 A lot of TI methods use similar algorithms to be able to infer a trajectory.
-By breaking down each method into its set of core algorithms, we can create a map of TI methodology\cite{cannoodt_computationalmethodstrajectory_2016} (Figure~\ref{fig:ti_metro}A), which is divided into two major classes.
+By breaking down each method into its set of core algorithms, we can create a map of TI methodology \cite{cannoodt_computationalmethodstrajectory_2016} (Figure~\ref{fig:ti_metro}A), which is divided into two major classes.
 In the first step, dimensionality reduction techniques such as manifold learning, clustering, or graph-based methods are used to convert the dataset to a more simplified representation.
@@
 
-A common way to classify TI methods is by the types of trajectories they can infer\cite{saelens_comparisonsinglecelltrajectory_2019} (Figure~\ref{fig:ti_metro}B).
+A common way to classify TI methods is by the types of trajectories they can infer \cite{saelens_comparisonsinglecelltrajectory_2019} (Figure~\ref{fig:ti_metro}B).
 About half of TI methods specialise in inferring linear or cyclic trajectories (i.e.
@@
 Before single-cell omics, these methods rely on multiple experiments, amongst which perturbation and time-series experiments, to predict the effect each transcription factor has on the up- or downregulation of a gene.
-One of the main advantages of single-cell omics is the heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs.
+One of the main advantages of single-cell omics is the heterogeneity between cells caused by naturally occurring biological randomness \cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs.
 
@@
 Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication.
-Hundreds of new software tools were developed\cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g.
+Hundreds of new software tools were developed \cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g.
 differential expression, dimensionality reduction, normalisation).

### 2019-10-16 ch3_dynbenchmark/content.tex
change cites

@@
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.
-TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}.
+TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}.
 Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

### 2019-10-16 ch6_bred/content.tex
change cites

@@
 One of the central cellular processes underlying development is transcriptional regulation.
-During development, changes in transcription factor activity induce chromatin modifications, chromatin remodelling and ultimately a differential recruitment of the basal transcriptional machinery\cite{coulon_eukaryotictranscriptionaldynamics_2013}.
+During development, changes in transcription factor activity induce chromatin modifications, chromatin remodelling and ultimately a differential recruitment of the basal transcriptional machinery \cite{coulon_eukaryotictranscriptionaldynamics_2013}.
 Modelling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
 
-The dynamics of gene regulation has classically been studied using time series data\cite{bar-joseph_studyingmodellingdynamic_2012}.
-When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population\cite{novershtern_denselyinterconnectedtranscriptional_2011, may_dynamicanalysisgene_2013, jojic_identificationtranscriptionalregulators_2013, goode_dynamicgeneregulatory_2016}.
+The dynamics of gene regulation has classically been studied using time series data \cite{bar-joseph_studyingmodellingdynamic_2012}.
+When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population \cite{novershtern_denselyinterconnectedtranscriptional_2011, may_dynamicanalysisgene_2013, jojic_identificationtranscriptionalregulators_2013, goode_dynamicgeneregulatory_2016}.
 Alternatively, time series data can also be generated by synchronizing the dynamic process between cells.
-However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed\cite{bar-joseph_studyingmodellingdynamic_2012}.
+However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed \cite{bar-joseph_studyingmodellingdynamic_2012}.
 
@@
 
-Several studies have highlighted how some regulatory interactions can be very dynamic while others show evidence of being static during consecutive developmental stages\cite{moignard_characterizationtranscriptionalnetworks_2013, pina_singlecellnetworkanalysis_2015}.
-Since regulatory interactions are context-dependent\cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
+Several studies have highlighted how some regulatory interactions can be very dynamic while others show evidence of being static during consecutive developmental stages \cite{moignard_characterizationtranscriptionalnetworks_2013, pina_singlecellnetworkanalysis_2015}.
+Since regulatory interactions are context-dependent \cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
 Case-wise NI methods\footnote{Case-wise NI is sometimes also called sample-specific NI or case-specific NI.} avoid predicting a static GRN and instead infer one GRN per cell (or per sample, for bulk omics data).
 
-In order to compute a case-wise GRN for a single sample, Kuijjer et al.\cite{kuijjer_estimatingsamplespecificregulatory_2019} and Liu et al.\cite{liu_personalizedcharacterizationdiseases_2016} employ similar strategies, namely by computing the difference of computing a static GRN for all the cases, and computing a static GRN for all the cases minus one.
+In order to compute a case-wise GRN for a single sample, Kuijjer et al.
+\cite{kuijjer_estimatingsamplespecificregulatory_2019} and Liu et al.
+\cite{liu_personalizedcharacterizationdiseases_2016} employ similar strategies, namely by computing the difference of computing a static GRN for all the cases, and computing a static GRN for all the cases minus one.
 Since this procedure needs to be repeated for every case in the dataset, and because NI methods are already amongst the most computationally intensive analyses to perform on omics data, this methodology is not applicable for large omics datasets.
-Another case-wise NI method, SCENIC\cite{aibar_scenicsinglecellregulatory_2017} infers case-wise GRNs by first inferring a static GRN using GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010}.
-GENIE3 is a static NI method which uses Random Forests\cite{breiman_randomforests_2001} feature importance scores to prioritise candidate regulators for a particular target gene.
+Another case-wise NI method, SCENIC \cite{aibar_scenicsinglecellregulatory_2017} infers case-wise GRNs by first inferring a static GRN using GENIE3 \cite{huynh-thu_inferringregulatorynetworks_2010}.
+GENIE3 is a static NI method which uses Random Forests \cite{breiman_randomforests_2001} feature importance scores to prioritise candidate regulators for a particular target gene.
 SCENIC then post-processes the static GRN to determine whether an interaction is enriched for particular cases, resulting in a case-wise GRN.
@@
 The case-wise GRNs -- or 'case-wise regulomes' -- can be analysed analogously to transcriptomics data; for example by clustering samples, inferring trajectories, or finding differentially activated regions.
-We demonstrate \texttt{bred} by applying it to a single-cell dataset of 22'122 hematopoietic cells from the Tabula Muris project\cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 14'963 bulk omics samples from The Cancer Genome Atlas project\cite{weinstein_cancergenomeatlas_2013}.
+We demonstrate \texttt{bred} by applying it to a single-cell dataset of 22'122 hematopoietic cells from the Tabula Muris project \cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 14'963 bulk omics samples from The Cancer Genome Atlas project \cite{weinstein_cancergenomeatlas_2013}.
 
@@
 \subsection{Hematopoietic cells from Tabula Muris}
-From the Tabula Muris\cite{schaum_singlecelltranscriptomics20_2018} project, we selected all cells involved in some part of the hematopoietic lineage tree.
+From the Tabula Muris \cite{schaum_singlecelltranscriptomics20_2018} project, we selected all cells involved in some part of the hematopoietic lineage tree.
 We computed case-wise GRNs between 22'122 remaining single cells, for the 2000 most variable target genes and a subset of transcription factors as regulators.
@@
 To summarise the similarity in GRNs across cases, we first visualise a dimensionality reduction of the case-wise GRNs, where every dot represents a single GRN (Figure~\ref{fig:tabmur}A).
-The dimensionality reduction was computed by applying Fruchterman-Reingold\cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour ($k$NN) graph of the case-wise GRN vectors.
-On the same $k$NN graph, the cells were clustered using the Louvain\cite{blondel_fastunfoldingcommunities_2008} clustering algorithm, and the clusters were labelled using prior information from Tabula Muris.
+The dimensionality reduction was computed by applying Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour ($k$NN) graph of the case-wise GRN vectors.
+On the same $k$NN graph, the cells were clustered using the Louvain \cite{blondel_fastunfoldingcommunities_2008} clustering algorithm, and the clusters were labelled using prior information from Tabula Muris.
 For each cluster, we retained the 50 interactions with the highest mean importance scores (Figure~\ref{fig:tabmur}B).
@@
 The task of inferring a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely: for every target $T$, predict which of the potential regulators $R_i$ regulate $T$ (Figure~\ref{fig:simplify}B).
-This simplification allowed GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's\cite{breiman_randomforests_2001} feature importance scores for inferring GRNs.
+This simplification allowed GENIE3 \cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's \cite{breiman_randomforests_2001} feature importance scores for inferring GRNs.
 Namely, a Random Forest is trained to predict the expression of a target gene of interest from the expression of potential regulators.
@@
 The samples are split into two groups, the 'in-bag' data and the 'out-of-bag' data (Figure~\ref{fig:fimp}A).
-A decision tree\cite{breiman_classificationregressiontrees_1984} is trained on the in-bag expression of the potential regulators in trying to predict the in-bag target gene expression (Figure~\ref{fig:fimp}B).
+A decision tree \cite{breiman_classificationregressiontrees_1984} is trained on the in-bag expression of the potential regulators in trying to predict the in-bag target gene expression (Figure~\ref{fig:fimp}B).
 The target expression of the out-of-bag samples is predicted using the decision tree (Figure~\ref{fig:fimp}C), and the squared error between the real and target expression is computed (Figure~\ref{fig:fimp}D).
@@
 
-To compute the case-wise GRNs, we implemented the abovementioned methodology in C++ in a modified version of the \texttt{ranger} R/C++ package\cite{wright_rangerfastimplementation_2017}.
+To compute the case-wise GRNs, we implemented the abovementioned methodology in C++ in a modified version of the \texttt{ranger} R/C++ package \cite{wright_rangerfastimplementation_2017}.
 
@@
 To perform downstream analysis on the cases, first a $k$-nearest neighbour ($k$NN) graph of the cases is computed.
-In order for the $k$NN graph to better emphasise similarities in GRNs rather than absolute euclidean distances, we first reduce the dimensionality of the case-by-interaction matrix to case-by-20 matrix using Landmark Multi-Dimensional Scaling\cite{lee_landmarkmdsensemble_2009} with a Spearman rank distance metric.
+In order for the $k$NN graph to better emphasise similarities in GRNs rather than absolute euclidean distances, we first reduce the dimensionality of the case-by-interaction matrix to case-by-20 matrix using Landmark Multi-Dimensional Scaling \cite{lee_landmarkmdsensemble_2009} with a Spearman rank distance metric.
 
 Next, KD-trees are used to calculate the $k$NN graph efficiently.
-The cases in the dataset are visualised and clusted using the Fruchterman-Reingold\cite{fruchterman_graphdrawingforcedirected_1991} and Louvain clustering\cite{blondel_fastunfoldingcommunities_2008}, respectively.
+The cases in the dataset are visualised and clusted using the Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} and Louvain clustering \cite{blondel_fastunfoldingcommunities_2008}, respectively.
 
-The following R packages provided implementations for each of these algorithms: lmds, RANN, igraph\cite{csardi_igraphsoftwarepackage_2006}.
+The following R packages provided implementations for each of these algorithms: lmds, RANN, igraph \cite{csardi_igraphsoftwarepackage_2006}.
 
@@
 After Louvain clustering, the interactions of the 50 interactions with highest mean importance per cluster are retained.
-These interactions are visualised in Cytoscape\cite{shannon_cytoscapesoftwareenvironment_2003}, in which nodes depict genes, edges depict predicted regulatory interactions, coloured according to which cluster they are predicted for.
+These interactions are visualised in Cytoscape \cite{shannon_cytoscapesoftwareenvironment_2003}, in which nodes depict genes, edges depict predicted regulatory interactions, coloured according to which cluster they are predicted for.
 The shape of the arrow denotes the predicted effect of the regulatory interaction.

### 2019-10-16 ch7_incgraph/content.tex
clean up thesis
change cites

@@
 The advent of high-throughput profiling methods (e.g.
-microarrays and RNA sequencing) have allowed to observe the molecular contents of a cell, which in turn have enabled computational network inference methods to reverse engineer the biochemical interaction networks~\cite{albert_networkinferenceanalysis_2007}.
-Improving the accuracy of inferred networks has been a long-standing challenge, but the development of ever more sophisticated algorithms and community-wide benchmarking studies have resulted in significant process~\cite{marbach_revealingstrengthsweaknesses_2010, narendra_comprehensiveassessmentmethods_2011, marbach_wisdomcrowdsrobust_2012, aijo_biophysicallymotivatedregulatory_2017}.
+microarrays and RNA sequencing) have allowed to observe the molecular contents of a cell, which in turn have enabled computational network inference methods to reverse engineer the biochemical interaction networks~ \cite{albert_networkinferenceanalysis_2007}.
+Improving the accuracy of inferred networks has been a long-standing challenge, but the development of ever more sophisticated algorithms and community-wide benchmarking studies have resulted in significant process~ \cite{marbach_revealingstrengthsweaknesses_2010, narendra_comprehensiveassessmentmethods_2011, marbach_wisdomcrowdsrobust_2012, aijo_biophysicallymotivatedregulatory_2017}.
 
-Several recent developments involve incorporating topological priors, either to guide the inference process~\cite{lopes_featureselectiontechnique_2014} or post-process the network~\cite{ruyssinck_netterrerankinggene_2016}.
-A common prior is that biological networks are highly modular~\cite{rives_modularorganizationcellular_2003}, as they consist of multiple collections of functionally or physically linked molecules~\cite{hartwell_molecularmodularcell_1999,barabasi_networkbiologyunderstanding_2004}.
-At the lowest level, each module is made up out of biochemical interactions arranged in small topological patterns, which act as fundamental building blocks~\cite{milo_networkmotifssimple_2002}.
+Several recent developments involve incorporating topological priors, either to guide the inference process~ \cite{lopes_featureselectiontechnique_2014} or post-process the network~ \cite{ruyssinck_netterrerankinggene_2016}.
+A common prior is that biological networks are highly modular~ \cite{rives_modularorganizationcellular_2003}, as they consist of multiple collections of functionally or physically linked molecules~ \cite{hartwell_molecularmodularcell_1999,barabasi_networkbiologyunderstanding_2004}.
+At the lowest level, each module is made up out of biochemical interactions arranged in small topological patterns, which act as fundamental building blocks~ \cite{milo_networkmotifssimple_2002}.
 
-Graphlets~\cite{przulj_modelinginteractomescalefree_2004} are one of the tools which could be used to add a topological prior to a biological network,
+Graphlets~ \cite{przulj_modelinginteractomescalefree_2004} are one of the tools which could be used to add a topological prior to a biological network,
 Graphlets are small connected subnetworks which can be counted to identify which low-level topological patterns are present in a network.
@@
 An orbit represents a class of isomorphic (i.e. resulting in the same structure) positions of nodes within a graphlet (Fig \ref{fig_graphlets}A, coloured in red).
-Both graphlets and orbits have been used extensively for predicting the properties of nodes such as protein functionality~\cite{milenkovic_uncoveringbiologicalnetwork_2008,guerrero_characterizationproteasomeinteraction_2008,singh_graphletsignaturebasedscoring_2014} and gene oncogenicity~\cite{milenkovic_optimalnetworkalignment_2010}, by performing network alignment~\cite{kuchaiev_topologicalnetworkalignment_2010,milenkovic_globalnetworkalignment_2013} or using them as a similarity measure in machine learning tasks~\cite{shervashidze_efficientgraphletkernels_2009,vacic_graphletkernelsprediction_2010}.
+Both graphlets and orbits have been used extensively for predicting the properties of nodes such as protein functionality~ \cite{milenkovic_uncoveringbiologicalnetwork_2008,guerrero_characterizationproteasomeinteraction_2008,singh_graphletsignaturebasedscoring_2014} and gene oncogenicity~ \cite{milenkovic_optimalnetworkalignment_2010}, by performing network alignment~ \cite{kuchaiev_topologicalnetworkalignment_2010,milenkovic_globalnetworkalignment_2013} or using them as a similarity measure in machine learning tasks~ \cite{shervashidze_efficientgraphletkernels_2009,vacic_graphletkernelsprediction_2010}.
 
@@
 Eppstein et al.
-introduced data structures and algorithms for updating the counts of size-three\cite{eppstein_hindexgraphits_2009} and size-four\cite{eppstein_extendeddynamicsubgraph_2012} subgraphs in a dynamic setting.
-The data structures were determined such that the amortised time is $O(h)$ and $O(h^2)$, respectively, where $h$ is the h-index of the network\cite{hirsch_indexquantifyindividual_2005}.
+introduced data structures and algorithms for updating the counts of size-three \cite{eppstein_hindexgraphits_2009} and size-four \cite{eppstein_extendeddynamicsubgraph_2012} subgraphs in a dynamic setting.
+The data structures were determined such that the amortised time is $O(h)$ and $O(h^2)$, respectively, where $h$ is the h-index of the network \cite{hirsch_indexquantifyindividual_2005}.
 
@@
 \subsection*{Timing experiments}
-We compared the execution time needed to calculate the graphlet counts in iteratively modified networks between our method and a state-of-the-art non-incremental algorithm, Orca~\cite{hocevar_combinatorialapproachgraphlet_2014}.
-Orca is a heavily optimised algorithm for counting 5-node graphlets in static networks, and outperforms all other static graphlet counting algorithms by an order of magnitude~\cite{hocevar_combinatorialapproachgraphlet_2014}.
+We compared the execution time needed to calculate the graphlet counts in iteratively modified networks between our method and a state-of-the-art non-incremental algorithm, Orca~ \cite{hocevar_combinatorialapproachgraphlet_2014}.
+Orca is a heavily optimised algorithm for counting 5-node graphlets in static networks, and outperforms all other static graphlet counting algorithms by an order of magnitude~ \cite{hocevar_combinatorialapproachgraphlet_2014}.
 
@@
 
-The network models were based on three static network models: Barab\'asi-Albert~\cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~\cite{erdos_randomgraphs_1959}, and Geometric~\cite{appel_minimumvertexdegree_1997}.
+The network models were based on three static network models: Barab\'asi-Albert~ \cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~ \cite{erdos_randomgraphs_1959}, and Geometric~ \cite{appel_minimumvertexdegree_1997}.
 These models were adapted to generate evolving networks instead (Figs~\ref{code_ba}, \ref{code_er}, and \ref{code_geo}).
@@
 cerevisiae}.
-The predicted networks were generated using the network inference method GENIE3~\cite{huynh-thu_inferringregulatorynetworks_2010} with default parameters.
-Gene expression data was obtained from COLOMBOS~\cite{moretto_colombosv3leveraging_2016} and GEO~\cite{edgar_geneexpressionomnibus_2002}, respectively.
-The predicted networks and the optimised versions thereof were compared against respective lists of known gene regulatory interactions~\cite{gama-castro_regulondbversionhighlevel_2016,ma_denovolearninggenomescale_2014}.
+The predicted networks were generated using the network inference method GENIE3~ \cite{huynh-thu_inferringregulatorynetworks_2010} with default parameters.
+Gene expression data was obtained from COLOMBOS~ \cite{moretto_colombosv3leveraging_2016} and GEO~ \cite{edgar_geneexpressionomnibus_2002}, respectively.
+The predicted networks and the optimised versions thereof were compared against respective lists of known gene regulatory interactions~ \cite{gama-castro_regulondbversionhighlevel_2016,ma_denovolearninggenomescale_2014}.
 
@@
 Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (Fig~\ref{sfig_hindex}).
-This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.
+This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.
+\cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.
 
@@
 \end{figure}
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-10-16 ch8_discussion/content.tex
change cites

--- 2019-10-15 ch8_discussion/content.tex
@@
 
-\texttt{dyngen} has already been successfully used to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
+\texttt{dyngen} has already been successfully used to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
 Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.
@@
 Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface.
-Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019}.
+Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results \cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019, reimegard_combinedmrnaprotein_2019}.
 
@@
 \section{Optimising regulatory networks}
-This project was a spin-off of Netter\cite{ruyssinck_netterrerankinggene_2016}, a tool for reranking GRN predictions using structural network properties. 
+This project was a spin-off of Netter \cite{ruyssinck_netterrerankinggene_2016}, a tool for reranking GRN predictions using structural network properties. 
 Predicted GRNs often are topologically dissimilar to real networks, and as such, Netter quantifies the topological properties of a network and makes incremental changes in order to arrive at a network with a 'better' topological profile.
@@
 
-\textbf{Git}\cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
+\textbf{Git} \cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
 Since Wouter Saelens and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel.
@@
 
-Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}.
+Together, we published over 20'000 code contributions (commits) across 50+ software packages \cite{cannoodt_developmentdynverse_2019}.
 Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package.
-We wrote many, many unit tests (and still too few), and we let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}.
+We wrote many, many unit tests (and still too few), and we let these tests automatically be executed on \textbf{Travis CI} \cite{traviscigmbh_traviscitest_2011}.
 This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing.
@@
 However, there are many benefits of using R in bioinformatics research context, such as its extensive community of statisticians develop packages for the R ecosystem.
-The \textbf{tidyverse} packages\cite{wickham_tidytoolsmanifesto_2017}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.
+The \textbf{tidyverse} packages \cite{wickham_tidytoolsmanifesto_2017}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.
 

### 2019-10-16 chX_backmatter.tex
update

@@
 		\cranpkg{babelwhale} & aut & CRAN & 5810 & Interacting with Docker and Singularity containers \\
-		\cranpkg{diffusionMap} & aut & CRAN & 30'450 & Implements diffusion map method of data parameterization, including creation and visualization of diffusion map \\
+		\cranpkg{diffusionMap} & aut & CRAN & 30'450 & Implements diffusion map method of data parameterization \\
 		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
@@
 		\cranpkg{GillespieSSA2} & aut & CRAN & 4950 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
-		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & An Rcpp implementation of the Growing Neural Gas algorithm \\
+		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & Growing Neural Gas implemented in Rcpp \\
 		\cranpkg{incgraph} & aut & CRAN & 3565 & Incremental graphlet counting for network optimisation \\
@@
 
-		\cranpkg{badger} & ctb & CRAN & 6240 & Query information and generate badge for using in README and GitHub Pages \\
+		\cranpkg{badger} & ctb & CRAN & 6240 & Query information and generate badge for using in README \\
 		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 935 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
 		\cranpkg{devtools} & ctb & CRAN & 5'918'700 & Tools to make developing R packages easier \\
-		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
+		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & Reconstructing lineage-tree topologies from scRNA-seq data \\
 		\cranpkg{mlr} & ctb & CRAN & 176'330 & Machine Learning in R \\
@@
 		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
-		\cranpkg{remotes} & ctb & CRAN & 3'944'090 & R package installation from remote repositories, including GitHub \\
+		\cranpkg{remotes} & ctb & CRAN & 3'944'090 & R package installation from remote repositories \\
 		\cranpkg{rlang} & ctb & CRAN & 13'269'115 & Functions for base types and core R and tidyverse features \\
-		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
+		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of high dimensional single cell data \\
 		\biocpkg{slingshot} & ctb & Bioc & 12'085 & Tools for ordering single-cell sequencing \\
 		\biocpkg{splatter} & ctb & Bioc & 5015 & Simple simulation of single-cell RNA sequencing data \\
-		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
+		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & Reconstructing branching trajectories from single-cell RNAseq data \\
 		\githubpkg{ManuSetty}{wishbone} & ctb & Github & \notavailable & Identify bifurcating developmental trajectories from single-cell data \\\hline

### 2019-10-16 ch3_dynbenchmark/content.tex
update

@@
 
-		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*
+		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_2018}\\*
 
@@
 		Dummy proofing & Package contains dummy proofing, i.e.
-testing whether the parameters and data supplied by the user make sense and are useful & \cite{lee_rpackagespackagedevelopment_2017,karimzadeh_topconsiderationscreating_}\\
+testing whether the parameters and data supplied by the user make sense and are useful & \cite{lee_rpackagespackagedevelopment_2017,karimzadeh_topconsiderationscreating_2018}\\
 
@@
 
-		Tutorial & (1) A tutorial or vignette is available (2) The tutorial has example results (3) The tutorial has real example data (4) The tutorial showcases the method on several datasets (1=0, 2=0.5, >2=1) & \cite{wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017,karimzadeh_topconsiderationscreating_,boulesteix_tensimplerules_2015}\\*
-
-		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
-
-		Inline documentation & Inline documentation is present in the code & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*
+		Tutorial & (1) A tutorial or vignette is available (2) The tutorial has example results (3) The tutorial has real example data (4) The tutorial showcases the method on several datasets (1=0, 2=0.5, >2=1) & \cite{wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017,karimzadeh_topconsiderationscreating_2018,boulesteix_tensimplerules_2015}\\*
+
+		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_2018}\\*
+
+		Inline documentation & Inline documentation is present in the code & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_2018}\\*
 
@@
 
-		Peer review & The paper is published in a peer-reviewed journal & \cite{karimzadeh_topconsiderationscreating_,gannon_essentialrolepeer_2001,baldwin_refereeswetrust_2017}\\*
+		Peer review & The paper is published in a peer-reviewed journal & \cite{karimzadeh_topconsiderationscreating_2018,gannon_essentialrolepeer_2001,baldwin_refereeswetrust_2017}\\*
 
@@
 		Evaluation of robustness & The paper assessed method robustness (to eg.
-noise, subsampling, parameter changes, stability) in one (0.5) or several (1) ways & \cite{karimzadeh_topconsiderationscreating_,aniba_issuesbioinformaticsbenchmarking_2010,boulesteix_tensimplerules_2015,jelizarow_overoptimismbioinformaticsillustration_2010}\\
+noise, subsampling, parameter changes, stability) in one (0.5) or several (1) ways & \cite{karimzadeh_topconsiderationscreating_2018,aniba_issuesbioinformaticsbenchmarking_2010,boulesteix_tensimplerules_2015,jelizarow_overoptimismbioinformaticsillustration_2010}\\
 

### 2019-10-18 ch6_bred/content.tex
~

@@
 One of the central cellular processes underlying development is transcriptional regulation. 
-During development, changes in transcription factor activity induce chromatin modifications, chromatin remodelling and ultimately a differential recruitment of the basal transcriptional machinery \cite{coulon_eukaryotictranscriptionaldynamics_2013}.
-Modelling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
+Changes in transcription factor activity induce chromatin modifications, 
+chromatin remodelling, a differential recruitment of the basal transcriptional machinery\cite{coulon_eukaryotictranscriptionaldynamics_2013}.
+Modelling the dynamics of gene regulation is therefore essential to better understand
+why a cellular dynamic processes progresses through several steps,
+and what goes wrong in the case of disease.
 

### 2019-10-18 chX_backmatter.tex
~

@@
 2018 Mar 5;110(10):1084-93.
-	\item Scott CL, T'Jonck W, \ldots, \textbf{Cannoodt R}, Saelens W \ldots, Guilliams M.
+	\item Scott CL*, T'Jonck W*, Martens L, Todorov H, Sichien D, Soen B, Bonnardel J, De Prijck S, Vandamme N, \textbf{Cannoodt R}, Saelens W, Vanneste B, Toussaint W, De Bleser P, Takahashi N, Vandenabeele P, Henri S, Pridans C, Hume DA, Lambrecht BN, De Baetselier P, Milling SWF, Van Ginderachter JA, Malissen B, Berx G, Beschin A, Saeys Y, Guilliams M.
 The transcription factor ZEB2 is required to maintain the tissue-specific identities of macrophages.
@@
 	\item Saelens W, \textbf{Cannoodt R}, Saeys Y. A comprehensive evaluation of module detection methods for gene expression data. Nature communications. 2018 Mar 15;9(1):1090.
-	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. In Gene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY..
+	\item Todorov H, \textbf{Cannoodt R}, Saelens W, Saeys Y. Network Inference from Single-Cell Transcriptomic Data. In Gene Regulatory Networks 2019 (pp. 235-249). Humana Press, New York, NY.
 	\item Van den Berge K, De Bezieux HR, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S, Clement L.
 Trajectory-based differential expression analysis for single-cell sequencing data.
-BioRxiv.
+bioRxiv.
 2019 Jan 1:623397.
@@
 Genome biology.
-2019 Dec;20(1):125.
-	\item Lorenzi L, \ldots, \textbf{Cannoodt R}, \ldots, Mestdagh P. The RNA-Atlas, a single nucleotide resolution map of the human transcriptome. In preparation.
+2019 Jun;20(1):125.
+	\item Lorenzi L*, Hua-Sheng C*, Avila Cobos F, Gross S, Volders PJ, \textbf{Cannoodt R}, Nuytens J, Vanderheyden K, Anckaert J, Lefever S, Goovaerts T, Hansen TB, Kuersten S, Nijs N, Taghon T, Vermaelen K, Brache KR, Saeys Y, De Meyer T, Deshpande N, Anande G, Chen TW, Wilkins MR, Unnikrishnan A, De Preter K, Kjerns J, Koster J, Schroth GP, Vandesompele J, Surnazin P, Mestdagh P.
+The RNA Atlas, a single nucleotide resolution map of the human transcriptome.
+bioRxiv.
+2019 Oct:807529.
+Submitted to Nature.
 	\item Van den Berge K, Roux de Bézieux H, Street K, Saelens W, \textbf{Cannoodt R}, Saeys Y, Dudoit S. Trajectory-based differential expression analysis. Submitted to Nature Communications.
-	\item Van de Sande B, \ldots, \textbf{Cannoodt R}, \ldots, Saeys Y, Aerts S. A scalable SCENIC workflow for single-cell gene regulatory network analysis. Submitted to Nature Protocols.
+	\item Van de Sande B, Flerin C, Davie K, De Waegeneer M, Hulselmans G, Aibar S, Seurinck R, Saelens W, \textbf{Cannoodt R}, Rouchon Q, Verbeiren T, De Maeyer D, Reumers J, Saeys Y, Aerts S.
+A scalable SCENIC workflow for single-cell gene regulatory network analysis.
+Submitted to Nature Protocols.
 \end{itemize}
+
+*: Equal contribution.
 
@@
 As part of our standard development practices, we automate execution of unit tests and write extensive documentation to ensure the code complies with CRAN policy before submission.
-We aim to submit all other packages to CRAN as well.
+Many of the packages are already hosted on CRAN, or are in the process of being prepared for submission.
 
 We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
-This includes speeding up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
-Several of these package receive millions of downloads per year (devtools, remotes, rlang).
+This includes speeding up parts of the dependency (slingshot), implementing new functionality (devtools, ParamHelpers, ranger, rlang), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
+Several of these packages are mainstream R packages which receive millions of downloads per year (devtools, ranger, remotes, rlang).
 

### 2019-10-18 ch6_bred/scripts/1_cellontology.R

@@ -2,6 +2,9 @@ library(tidyverse)
 library(ontologyIndex)
 
 
+# Cell ontology -----------------------------------------------------------
+
+
 cellontf <- "derived_files/cl.obo"
 
 if (!file.exists(cellontf)) download.file("https://raw.githubusercontent.com/obophenotype/cell-ontology/master/cl.obo", cellontf)
@@ -35,3 +38,25 @@ ciddr <- stats::cmdscale(as.dist(ciddis), k = 10)
 
 out <- "derived_files/cell_ontology.rds"
 write_rds(lst(cell_ont, cell_ont_up, ciddr), out, compress = "gz")
+
+
+
+# NCI ---------------------------------------------------------------------
+
+ncif <- "derived_files/nci.obo"
+
+if (!file.exists(ncif)) download.file("http://purl.obolibrary.org/obo/ncit.obo", ncif)
+
+nci_ont <-
+  get_ontology(ncif) %>%
+  with(tibble(id, name, parents, children, ancestors, obsolete)) %>%
+  filter(!obsolete, grepl("NCIT:", id))
+
+
+out <- "derived_files/nci_ontology.rds"
+write_rds(nci_ont, out, compress = "gz")
+
+
+
+ncio <- read_tsv("/home/rcannood/Workspace/tcga/maps/map_out.csv")$nci_ont %>% na.omit %>% unique()
+ncio[!ncio %in% nci_ont$id]

### 2019-10-18 ch6_bred/scripts/2_tcga.R

@@ -8,14 +8,14 @@ tcga_data_file <- paste0(tcga_folder, "data.rds")
 data_file <- paste0(data_dir, "data.rds")
 
 if (!file.exists(tcga_data_file)) {
-  sample_info <- read_tsv(paste0(tcga_folder, "sample_sheet.tsv"))
-  colnames(sample_info) <- colnames(sample_info) %>% tolower() %>% gsub(" ", "_", .)
+  sample_sheet <- read_tsv(paste0(tcga_folder, "sample_sheet.tsv"))
+  colnames(sample_sheet) <- colnames(sample_sheet) %>% tolower() %>% gsub(" ", "_", .)
 
   # preproc metadata
-  sample_info$project_id <- gsub(",.*", "", sample_info$project_id)
-  sample_info$sample_type <- gsub(",.*", "", sample_info$sample_type)
-  sample_info$sample_id <- gsub(",.*", "", sample_info$sample_id)
-  sample_info <- sample_info %>%
+  sample_sheet$project_id <- gsub(",.*", "", sample_sheet$project_id)
+  sample_sheet$sample_type <- gsub(",.*", "", sample_sheet$sample_type)
+  sample_sheet$sample_id <- gsub(",.*", "", sample_sheet$sample_id)
+  sample_sheet <- sample_sheet %>%
     filter(!sample_type %in% c("Control Analyte", "Additional Metastatic", "FFPE Scrolls", "Slides")) %>%
     mutate(
       sample_group = case_when(
@@ -31,16 +31,102 @@ if (!file.exists(tcga_data_file)) {
     ) %>%
     group_by(project_id, sample_group) %>%
     mutate(
-      id = paste0(project_id, "_", tolower(gsub(" ", "-", sample_group)), "_sample", seq_len(n()))
+      old_id = paste0(project_id, "_", tolower(gsub(" ", "-", sample_group)), "_sample", seq_len(n()))
     ) %>%
     ungroup() %>%
-    select(id, project_id, sample_group, everything())
+    filter(!duplicated(file_id))
 
-  # check classes
-  table(sample_info$project_id) %>% sort
-  table(sample_info$sample_type) %>% sort
-  table(sample_info$sample_group) %>% sort
-  table(sample_info$sample_type, sample_info$sample_group)
+  assertthat::assert_that(!any(duplicated(sample_sheet$file_id)))
+
+  # add clinical case id
+  mjs <- c(
[…]

### 2019-10-20 ch6_bred/scripts/2_tcga.R
update bred before cleanup

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(rlang)
 
 tcga_folder <- "~/Workspace/tcga/"
 data_dir <- "derived_files/tcga/"
@@ -79,29 +80,19 @@ if (!file.exists(tcga_data_file)) {
     li <- list(case_id = jsi$case_id)
     for (n in intersect(c("diagnosis", "demographic", "exposure"), names(jsi))) {
       vals <- jsi[[n]]
-      li[paste0(n, ".", names(vals))] <- map(vals, ~ . %||% NA)
+      ix <- names(vals) %in% c("updated_datetime", "created_datetime", "state", "submitter_id")
+      names(vals)[ix] <- paste0(n, ".", names(vals)[ix])
+      li[names(vals)] <- map(vals, ~ . %||% NA)
     }
     as_tibble(li)
-  }) %>% bind_rows()
-
-  clinical2 <- clinical %>%
-    transmute(
-      clinical.case_id = case_id,
-      gender = demographic.gender,
-      race = demographic.race,
-      ethnicity = demographic.ethnicity,
-      vital_status = demographic.vital_status,
-      last_known_disease_status = diagnosis.last_known_disease_status,
-      days_to_birth = demographic.days_to_birth,
-      days_to_diagnosis = diagnosis.days_to_diagnosis,
-      days_to_last_follow_up = diagnosis.days_to_last_follow_up,
-      primary_diagnosis = tolower(diagnosis.primary_diagnosis),
-      tissue_or_organ = tolower(diagnosis.tissue_or_organ_of_origin)
-    )
+  }) %>%
+    bind_rows() %>%
+    rename(clinical.case_id = case_id)
 
   sample_sheet3 <-
     sample_sheet2 %>%
-    left_join(clinical2, by = "clinical.case_id")
+    left_join(clinical, by = "clinical.case_id") %>%
+    mutate(primary_diagnosis = tolower(primary_diagnosis), tissue_or_organ_of_origin = tolower(tissue_or_organ_of_origin))
 
   assertthat::assert_that(!any(duplicated(sample_sheet3$file_id)))
 
@@ -109,7 +100,7 @@ if (!file.exists(tcga_data_file)) {
     proj_meta <- read_tsv(paste0(tcga_folder, "/raw/project_meta.tsv"))
     tabn <-
       clinical %>%
[…]

### 2019-10-21 ch0_frontmatter.tex
update

--- 2019-08-11 ch0_frontmatter.tex
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Acknowledgements}
-\blindtext
-
-\newpage{\thispagestyle{empty}\cleardoublepage}
 
@@
 
-\newcommand{\grn}{GRN}
-\newcommand{\grns}{GRNs}
-\nomenclature{GRN}{Gene Regulatory Network}
+\nomenclature{HCA}{Human Cell Atlas}
+\nomenclature{TCGA}{The Cancer Genome Atlas}
 
-\newcommand{\im}{IM}
-\newcommand{\ims}{IMs}
-\nomenclature{IM}{Importance Measure}
-
-\newcommand{\tf}{TF}
-\newcommand{\tfs}{TFs}
+\nomenclature{DNA}{Deoxyribonucleic Acid}
+\nomenclature{RNA}{Ribonucleic Acid}
+\nomenclature{mRNA}{Messenger RNA}
 \nomenclature{TF}{Transcription Factor}
 
-\newcommand{\dna}{DNA}
-\nomenclature{DNA}{Deoxyribonucleic Acid}
+\nomenclature{GRN}{Gene Regulatory Network}
+\nomenclature{NI}{Network Inference}
+\nomenclature{CART}{Classification And Regression Trees}
+\nomenclature{RF}{Random Forests}
+\nomenclature{TI}{Trajectory Inference}
 
-\newcommand{\rna}{RNA}
-\nomenclature{RNA}{Ribonucleic Acid}
+\nomenclature{$k$NN}{$k$-nearest-neighbour}
+\nomenclature{DR}{Dimensionality Reduction}
+\nomenclature{PCA}{Principal Component Analysis}
+\nomenclature{MDS}{Multi-Dimensional Scaling}
+\nomenclature{t-SNE}{t-distributed Stochastic Neighbor Embedding}
+\nomenclature{UMAP}{Uniform Manifold Approximation and Projection}
+\nomenclature{LDMS}{Landmark Multi-Dimensional Scaling}
+\nomenclature{DE}{Differential Expression}
+\nomenclature{TDE}{Trajectory Differential Expression}
 
-\newcommand{\mrna}{mRNA}
-\nomenclature{mRNA}{Messenger RNA}
+\chapter*{Summary}
+\vspace{-2em}
 
-\newcommand{\nin}{NI}
-\nomenclature{NI}{Network Inference}
+Recent developments in the \textit{single-cell omics} technologies triggered a major milestone in the field of (cell)biology.
+Current techniques can be used to profile both the genome and the transcriptome of individual cells, leading to important insights into cellular heterogeneity and dynamic cell processes.
+This evolution has led, among other things, to the Human Cell Atlas initiative, an international collaboration project with the aim of mapping all human cell types in terms of its biomolecular contents.
 
-\newcommand{\ml}{ML}
-\nomenclature{ML}{Machine Learning}
+During this doctoral project I have been able to follow these inspiring advancements first hand, but I also made scientific contributions myself, by extensively evaluating new computational tools and by developing various tools and algorithms.
 
-\newcommand{\rf}{RF}
-\nomenclature{RF}{Random Forests}
+A common problem of pioneering computational tools is the lack of sufficient available datasets to enable the assessment of performances in a quantitative manner.
+During this PhD, I developed a flexible and extendible \textit{in silico} simulator for individual cells, with the aim of supporting current and future computational research fields and to help kick-start emerging computational sub-domains (Chapter~\ref{chap:dyngen}).
 
-\newcommand{\cart}{CART}
-\nomenclature{CART}{Classification And Regression Trees}
+I conducted an extensive study of 45 \textit{trajectory inference} methods (Chapter~\ref{chap:dynbenchmark}).
+Trajectory inference is a new type of computational analysis with the aim of identifying and studying transitions between main cellular states.
+While this class of methods is one of the fastest growing subdomains within the \textit{single-cell omics}, a quantitative study of the advantages and disadvantages was hitherto lacking.
+
+In addition, I developed four new software tools.
+These methods include an extensive toolkit for inferring, visualising and interpreting trajectories (Chapter~\ref{chap:dyno}), a robust tool for inferring linear trajectories (Chapter~\ref{chap:scorpius}), a novel algorithm for studying gene regulation at a single-cell or single-sample level (Chapter~\ref{chap:bred}), and a tool for quantifying the topological changes resulting from minor perturbations in a network (Chapter~\ref{chap:incgraph}).
+
+Finally, I reflect on current challenges within the field, by providing guidelines for how to self-assess trajectory inference methods (Chapter~\ref{chap:selfassessment}) and, more generally, how to perform large-scale benchmarking experiments of computational methods (Chapter~\ref{chap:guidelines}).
+
+Throughout this project I particularly enjoyed working in an open-source ecosystem, collaborating with scientists around the world to tackle unsolved problems, whilst learning best practices by inspecting the source code of exemplary software projects.
+
+\chapter*{Samenvatting}
+\vspace{-8em}
+
+Recente ontwikkelingen in de \textit{single-cell omics} technologieën vormen een belangrijke mijlpaal in het domein van de (cel)biologie.
+Met de huidige technieken kunnen zowel het genoom als het transcriptoom van individuele cellen geprofileerd worden wat leidt tot belangrijke inzichten in de cellulaire heterogeniteit en de dynamische celprocessen.
+Deze evolutie heeft onder meer geleid tot het Human Cell Atlas initiatief, een internationaal samenwerkingsproject met als doel het in kaart brengen van alle humane cell types.
+
+Gedurende dit doctoraatsproject heb ik niet alleen deze inspirerende ontwikkelingen van nabij kunnen volgen, maar evenzeer zelf wetenschappelijke bijdragen kunnen aanleveren door het uitgebreid evalueren van nieuwe computationele tools en door het ontwikkelen van verscheiden tools en algoritmen.
+
+Een veelvoorkomend probleem bij de ontwikkeling van innovatieve computationele algoritmen is het gebrek aan voldoende beschikbare datasets om de performantie op een kwantitatieve wijze te kunnen beoordelen.
+Tijdens dit doctoraat ontwikkelde ik een flexibele en uitbreidbare \textit{in silico} simulator voor individuele cellen, met als doel huidige en toekomstige computationele onderzoeksvelden te ondersteunen en kick-starter projecten een duwtje in de rug te geven (Chapter~\ref{chap:dyngen}).
+
+Eveneens voerden we een uitgebreide studie van 45 \textit{trajectvoorspellende} methoden uit (Hoofdstuk~\ref{chap:dynbenchmark}).
+Trajectvoorspellende methoden is een nieuwe soort van computationele analyse met als doel transities tussen verscheidene cellulaire toestanden te identificeren en te bestuderen.
+Terwijl deze klasse van methoden een van de snelst groeiende subdomeinen binnen de \textit{single-cell omics} vormt, was er tot op dat moment nog geen kwantitatieve studie van de voor- en nadelen van dergelijke methoden uitgevoerd.
+
+Ik ontwikkelde ook zelf vier nieuwe softwareprogramma's, waaronder een uitgebreide toolkit om trajecten te voorspellen, te visualiseren en te interpreteren (Hoofdstuk~\ref{chap:dyno}), een robuust programma om lineaire trajecten te voorspellen (Hoofdstuk~\ref{chap:scorpius}), een nieuwe manier om genregulatie op het niveau van individuele cellen of stalen te bestuderen (Hoofdstuk~\ref{chap:bred}), en een programma om topologische veranderingen ten gevolge van kleine veranderingen in een netwerk te kwantificeren (Hoofdstuk~\ref{chap:incgraph}).
+
+Ik reflecteer op huidige uitdagingen binnen het veld, door richtlijnen aan te bieden aan ontwikkelaars voor het evalueren van hun trajectvoorspellende methoden (Hoofdstuk~\ref{chap:selfassessment}), en meer algemeen over hoe grootschalige benchmarkingsexperimenten van computationele methoden kunnen worden uitgevoerd (Hoofdstuk~\ref{chap:guidelines}).
+
+Gedurende dit project heb ik vooral genoten van het werken in een textit{open-source} omgeving, en het samenwerken met nationale en internationale collega’s uit het veld om onopgeloste problemen aan te pakken.
+Door de broncode van excellente softwareprojecten te bestuderen, leerde ik mezelf de aanbevolen werkwijzen aan.

### 2019-10-21 ch6_bred/package/R/dimred_and_cluster.R

@@ -1,0 +1,57 @@
+
+dimred_and_cluster <- function(importance_sc, lmds_ndim = 20, knn = 100, use_scaled_imp = TRUE, knn_dist_quantile = .95) {
+  sample_ids <- levels(importance_sc$sample_id)
+  interaction_ids <- levels(importance_sc$interaction_id)
+
+  imp_sc_mat <- Matrix::sparseMatrix(
+    i = importance_sc$sample_id %>% as.integer,
+    j = importance_sc$interaction_id %>% as.integer,
+    x = importance_sc[[if (use_scaled_imp) "importance_sc" else "importance"]],
+    dims = c(length(sample_ids), length(interaction_ids)),
+    dimnames = list(sample_ids, interaction_ids)
+  )
+
+  dimred_lmds <- lmds::lmds(imp_sc_mat, ndim = lmds_ndim, distance_method = "spearman")
+  rm(imp_sc_mat)
+  gc()
+
+  # compute knn
+  knn <- RANN::nn2(dimred_lmds, k = knn + 1)
+  knn$nn.dists <- knn$nn.dists[,-1]
+  knn$nn.idx <- knn$nn.idx[,-1]
+
+  # perform louvain clustering and FR dimred
+  knndf <-
+    inner_join(
+      reshape2::melt(knn$nn.idx, varnames = c("i", "nn"), value.name = "j"),
+      reshape2::melt(knn$nn.dists, varnames = c("i", "nn"), value.name = "dist"),
+      by = c("i", "nn")
+    ) %>%
+    as_tibble() %>%
+    mutate(i2 = pmin(i, j), j2 = pmax(i, j)) %>%
+    select(i = i2, j = j2, dist) %>%
+    unique() %>%
+    arrange(dist) %>%
+    mutate(weight = 1 / dist) %>%
+    filter(dist < quantile(dist, knn_dist_quantile))
+
+  gr <- igraph::graph_from_data_frame(
+    knndf %>% select(i, j, dist),
+    vertices = seq_len(nrow(dimred_lmds)),
+    directed = FALSE
+  )
+  cl <- igraph::cluster_louvain(gr)
+  cluster <- cl$membership
+  names(cluster) <- rownames(dimred_lmds)
+
+  dimred_fr <- igraph::layout_with_fr(gr)
+  rownames(dimred_fr) <- rownames(dimred_lmds)
[…]

### 2019-10-21 ch6_bred/package/R/label_clusters.R

@@ -1,0 +1,58 @@
+`%s/%` <- function(x, y) ifelse(y == 0, 1, x / y)
+
+label_clusters <- function(sample_groupings, cluster, arrange_fun = function(df) arrange(df, desc(F1))) {
+  if (!"group_type" %in% colnames(sample_groupings)) {
+    sample_groupings$group_type <- factor("group")
+  }
+  TOT <- length(unique(sample_groupings$sample_id))
+
+  test <-
+    enframe(cluster, "sample_id", "cluster") %>%
+    group_by(cluster) %>%
+    mutate(NUM_CLUSTER = n()) %>%
+    ungroup() %>%
+    left_join(sample_groupings, by = "sample_id") %>%
+    group_by(group_type, group_id) %>%
+    mutate(NUM_GROUP = n()) %>%
+    group_by(cluster, group_type, group_id) %>%
+    summarise(
+      NUM_CLUSTER = NUM_CLUSTER[[1]],
+      NUM_GROUP = NUM_GROUP[[1]],
+      TP = n()
+    ) %>%
+    ungroup() %>%
+    mutate(
+      FN = NUM_GROUP - TP,
+      FP = NUM_CLUSTER - TP,
+      TN = TOT - TP - FN - FP,
+      TPR = TP %s/% (TP + FN),
+      TNR = TN %s/% (TN + FP),
+      PPV = TP %s/% (TP + FP),
+      NPV = TN %s/% (TN + FN),
+      FNR = FN %s/% (FN + TP),
+      FPR = FP %s/% (FP + TN),
+      FDR = FP %s/% (FP + TP),
+      FOR = FN %s/% (FN + TN),
+      F1 = (2 * PPV * TPR) %s/% (PPV + TPR)
+    )
+
+  labels <-
+    test %>%
+    arrange_fun() %>%
+    group_by(cluster) %>%
+    slice(1) %>%
+    ungroup() %>%
+    rename(name = group_id) %>%
+    group_by(name) %>%
+    mutate(name2 = if (n() == 1) name else paste0(name, " #", row_number())) %>%
+    ungroup() %>%
+    mutate(name = factor(name2)) %>%
+    select(-name2) %>%
+    arrange(name) %>%
+    mutate(col = grDevices::colorRampPalette(RColorBrewer::brewer.pal(8, "Accent"))(n())) %>%
+    arrange(cluster)
+
+  palette <- labels %>% select(name, col) %>% deframe()
+
+  lst(test, labels, palette)
+}

### 2019-10-21 ch6_bred/scripts/2_tabula-muris.R

@@ -233,21 +233,26 @@ if (!file.exists(paste0(dest_dir, "qsub_handle.rds"))) {
 }
 
 dest_dir <- "derived_files/tabula_muris/"
-qsub_handle <- read_rds(paste0(dest_dir, "qsub_handle.rds"))
 
-grn <- qsub::qsub_retrieve(
-  qsub_handle,
-  wait = "just_do_it",
-  post_fun = function(i, li) {
-    li$importance <- li$importance %>% filter(importance > .01)
-    li$importance_sc <- li$importance_sc %>% filter(importance_sc > .01) %>%
-      inner_join(li$importance %>% select(regulator, target), by = c("regulator", "target"))
-    li
-  }
-)
-write_rds(grn, paste0(dest_dir, "grn.rds"))
+if (!file.exists(paste0(dest_dir, "grn.rds"))) {
+  qsub_handle <- read_rds(paste0(dest_dir, "qsub_handle.rds"))
 
+  grn <- qsub::qsub_retrieve(
+    qsub_handle,
+    wait = "just_do_it",
+    post_fun = function(i, li) {
+      li$importance <- li$importance %>% filter(importance > .01)
+      li$importance_sc <- li$importance_sc %>% filter(importance_sc > .01) %>%
+        inner_join(li$importance %>% select(regulator, target), by = c("regulator", "target"))
+      li
+    }
+  )
+  write_rds(grn, paste0(dest_dir, "grn.rds"))
+}
 grn <- read_rds(paste0(dest_dir, "grn.rds"))
+
+
+
 
 # remove unfinished executions
 grn[map_lgl(grn, ~ length(.) == 1 && is.na(.))] <- NULL
@@ -256,285 +261,170 @@ grn[map_lgl(grn, ~ length(.) == 1 && is.na(.))] <- NULL
 importance <-
   grn %>%
   map_df("importance") %>%
+  arrange(desc(importance)) %>%
   mutate(
-    i = row_number(),
-    name = paste0(regulator, "->", target)
+    interaction_id = forcats::fct_inorder(paste0(regulator, "->", target))
   )
-ggplot(importance) +
-  geom_point(aes(effect, importance))
-write_tsv(importance, "~/importance.tsv")
-importance_sc <- grn %>%
+importance_sc <-
+  grn %>%
   map_df("importance_sc") %>%
[…]

### 2019-10-21 ch6_bred/scripts/2_tabula-muris_hematopoietic.R

@@ -265,414 +265,163 @@ importance <-
   map_df("importance") %>%
   arrange(desc(importance)) %>%
   mutate(
-    i = row_number(),
-    name = paste0(regulator, "->", target)
+    interaction_id = forcats::fct_inorder(paste0(regulator, "->", target))
   )
-ggplot(importance) +
-  geom_point(aes(effect, importance))
-write_tsv(importance %>% filter(importance > .1), "~/importance.tsv")
-importance_sc <- grn %>%
+importance_sc <-
+  grn %>%
   map_df("importance_sc") %>%
-  inner_join(importance %>% select(-importance, -name), by = c("regulator", "target"))
+  rename(sample_id = cell_id) %>%
+  left_join(importance %>% select(regulator, target, effect, interaction_id), by = c("regulator", "target"))
 
 rm(grn)
 gc()
 
-samples <- levels(importance_sc$cell_id)
-imp_sc_mat <- Matrix::sparseMatrix(
-  i = importance_sc$cell_id %>% as.integer,
-  j = importance_sc$i,
-  x = importance_sc$importance_sc,
-  dims = c(length(samples), nrow(importance)),
-  dimnames = list(samples, importance$name)
-)
-
-data_rds <- paste0(dest_dir, "data_rds/")
-list2env(read_rds(paste0(data_rds, "data_filt.rds")), .GlobalEnv)
-list2env(read_rds("derived_files/cell_ontology.rds"), .GlobalEnv)
 
 
-ndims <- c(2:30, seq(34, 100, by = 4))
-dist2lm <- lmds::select_landmarks(imp_sc_mat, distance_method = "spearman")
-ix <- sample.int(nrow(imp_sc_mat), 1000)
-df <- bind_rows(pbapply::pblapply(ndims, cl = 8, function(ndim) {
-  dimred <- lmds::cmdscale_landmarks(dist2lm, ndim = ndim)
-  euc <- dynutils::euclidean_distance(dimred[ix,])
-  cor <- dynutils::correlation_distance(imp_sc_mat[ix,])
-  tibble(
-    ndim,
-    cor = cor(as.vector(euc), as.vector(cor))
+if (!file.exists(paste0(dest_dir, "dimred_and_clustering.rds"))) {
[…]

### 2019-10-21 ch6_bred/scripts/2_tcga.R

@@ -316,58 +316,10 @@ importance_sc <-
   select(-i) %>%
   left_join(importance %>% select(regulator, target, interaction_id), by = c("regulator", "target"))
 
-sample_ids <- levels(importance_sc$sample_id)
-interaction_ids <- levels(importance_sc$interaction_id)
-use_scaled_imp <- FALSE
-imp_sc_mat <- Matrix::sparseMatrix(
-  i = importance_sc$sample_id %>% as.integer,
-  j = importance_sc$interaction_id %>% as.integer,
-  x = importance_sc[[if (use_scaled_imp) "importance_sc" else "importance"]],
-  dims = c(length(sample_ids), length(interaction_ids)),
-  dimnames = list(sample_ids, interaction_ids)
-)
-rm(importance_sc)
-gc()
-
 if (!file.exists(paste0(data_dir, "dimred_and_clustering.rds"))) {
-  # dimred_out <- bred::dimred_and_cluster(importance_sc)
-  # dimred_out <- bred::dimred_and_cluster(importance_sc, knn = 11, use_scaled_imp = FALSE)
-  # dimred_out <- bred::dimred_and_cluster(expr, knn = 11, use_scaled_imp = FALSE)
-  dimred_lmds <- lmds::lmds(imp_sc_mat, distance_method = "spearman")
-  # dimred_lmds <- lmds::lmds(expr, distance_method = "spearman")
-  gng_out <- gng::gng(dimred_lmds, max_iter = 100000, max_nodes = 50)
+  dimred_out <- bred:::dimred_and_cluster(importance_sc, knn = 100, use_scaled_imp = FALSE)
 
-  eucl_dist <- as.matrix(stats::dist(gng_out$node_space))
-  pts <-
-    gng_out$edges %>%
-    rowwise() %>%
-    mutate(pct = list(seq(0, 1, length.out = 21))) %>%
-    ungroup() %>%
-    unnest(pct)
-  pts_space <- (1 - pts$pct) * gng_out$node_space[pts$i, ] + pts$pct * gng_out$node_space[pts$j, ]
-  pts$dist <- rowMeans(RANN::nn2(dimred_lmds, pts_space, k = 10)$nn.dist)
-  dendis <-
-    pts %>%
-    group_by(i, j) %>%
-    summarise(den_dist = mean(dist)) %>%
-    ungroup() %>%
-    mutate(
-      eucl_dist = eucl_dist[cbind(i, j)],
-      weight = den_dist * eucl_dist
-    )
-
[…]

### 2019-10-21 ch6_bred/scripts/2_tcga_gng.R

@@ -1,0 +1,539 @@
+library(tidyverse)
+library(rlang)
+
+tcga_folder <- "~/Workspace/tcga/"
+data_dir <- "derived_files/tcga/"
+dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
+
+tcga_data_file <- paste0(tcga_folder, "data.rds")
+data_file <- paste0(data_dir, "data.rds")
+
+if (!file.exists(tcga_data_file)) {
+  sample_sheet <- read_tsv(paste0(tcga_folder, "sample_sheet.tsv"))
+  colnames(sample_sheet) <- colnames(sample_sheet) %>% tolower() %>% gsub(" ", "_", .)
+
+  # preproc metadata
+  sample_sheet$project_id <- gsub(",.*", "", sample_sheet$project_id)
+  sample_sheet$sample_type <- gsub(",.*", "", sample_sheet$sample_type)
+  sample_sheet$sample_id <- gsub(",.*", "", sample_sheet$sample_id)
+  sample_sheet <- sample_sheet %>%
+    filter(!sample_type %in% c("Control Analyte", "Additional Metastatic", "FFPE Scrolls", "Slides")) %>%
+    mutate(
+      sample_group = case_when(
+        sample_type == "Blood Derived Normal" ~ "Normal Blood",
+        sample_type == "Solid Tissue Normal" ~ "Normal Tissue",
+        grepl("Tumor", sample_type) ~ "Cancer Tumor",
+        sample_type == "Additional - New Primary" ~ "Cancer Tumor",
+        grepl("Metastatic", sample_type) ~ "Cancer Metastatic",
+        grepl("Cancer.*Bone Marrow", sample_type) ~ "Cancer BM",
+        grepl("Cancer.*Blood", sample_type) ~ "Cancer Blood"
+      ),
+      file = paste0(tcga_folder, "data/", file_id, "/", file_name)
+    ) %>%
+    group_by(project_id, sample_group) %>%
+    mutate(
+      old_id = paste0(project_id, "_", tolower(gsub(" ", "-", sample_group)), "_sample", seq_len(n()))
+    ) %>%
+    ungroup() %>%
+    filter(!duplicated(file_id))
+
+  assertthat::assert_that(!any(duplicated(sample_sheet$file_id)))
+
+  # add clinical case id
+  mjs <- c(
+    jsonlite::read_json("/home/rcannood/Workspace/tcga/raw/F_metadata.cart.2019-10-08.json"),
[…]

### 2019-10-21 ch6_bred/content.tex
~

@@
 One of the central cellular processes underlying development is transcriptional regulation. 
-Changes in transcription factor activity induce chromatin modifications, 
-chromatin remodelling, a differential recruitment of the basal transcriptional machinery\cite{coulon_eukaryotictranscriptionaldynamics_2013}.
 Modelling the dynamics of gene regulation is therefore essential to better understand
@@
 and what goes wrong in the case of disease. 
+Regulatory dynamics is classically studied using time series data \cite{bar-joseph_studyingmodellingdynamic_2012}.
+When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population \cite{novershtern_denselyinterconnectedtranscriptional_2011, may_dynamicanalysisgene_2013, jojic_identificationtranscriptionalregulators_2013, goode_dynamicgeneregulatory_2016}.
+Alternatively, time series data can also be generated by synchronising the dynamic process between cells.
+However, issues with time-resolution, heterogeneity and good \textit{in vivo} synchronisation models can often limit the predictive power of the dynamic models of gene regulation which can be constructed \cite{bar-joseph_studyingmodellingdynamic_2012}.
 
-The dynamics of gene regulation has classically been studied using time series data \cite{bar-joseph_studyingmodellingdynamic_2012}.
-When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population \cite{novershtern_denselyinterconnectedtranscriptional_2011, may_dynamicanalysisgene_2013, jojic_identificationtranscriptionalregulators_2013, goode_dynamicgeneregulatory_2016}.
-Alternatively, time series data can also be generated by synchronizing the dynamic process between cells.
-However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed \cite{bar-joseph_studyingmodellingdynamic_2012}.
-
-One of the main advantages of single-cell transcriptomics is the ability to quantify the exact cellular state of thousands of cells per experiment.
-The intercellular heterogeneity caused by naturally occurring biological stochasticity \cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to predict regulatory interactions between transcription factors (TFs) and their target genes.
-The computational tools that infer gene regulatory networks (GRNs) from omics datasets are called network inference (NI) methods.
-% TODO: talk about bulk NI
+Network inference (NI) methods are computational tools which use large omics datasets to predict which genes are regulated by which transcription factors.
+While accuracy of the network of predicted regulatory interactions is lower in comparison to experimental validation techniques, NI methods offer an unbiased and high-throughput insight into the regulatory dynamics of a biological system.
+The output of a network inference is thus a graph, where nodes represent genes and edges denote a regulatory interaction between a regulator and a target gene.
+Interactions have two properties: its regulatory strength (a positive real value) and its effect (promoting or repressing).
 
@@
 Another case-wise NI method, SCENIC \cite{aibar_scenicsinglecellregulatory_2017} infers case-wise GRNs by first inferring a static GRN using GENIE3 \cite{huynh-thu_inferringregulatorynetworks_2010}.
-GENIE3 is a static NI method which uses Random Forests \cite{breiman_randomforests_2001} feature importance scores to prioritise candidate regulators for a particular target gene.
+GENIE3 is a static NI method which uses Random Forests (RFs) \cite{breiman_randomforests_2001} variable importance scores to prioritise candidate regulators for a particular target gene.
 SCENIC then post-processes the static GRN to determine whether an interaction is enriched for particular cases, resulting in a case-wise GRN.
-In short, while several case-wise NI methods thus already exist, their implementation consisted of post-processing a static GRN to arrive at a case-wise GRN.
 
-In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. For each interaction in the inferred case-wise GRN, \texttt{bred} predicts both the regulatory strength and its effect. 
-The case-wise GRNs -- or 'case-wise regulomes' -- can be analysed analogously to transcriptomics data; for example by clustering samples, inferring trajectories, or finding differentially activated regions.
-We demonstrate \texttt{bred} by applying it to a single-cell dataset of 22'122 hematopoietic cells from the Tabula Muris project \cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 14'963 bulk omics samples from The Cancer Genome Atlas project \cite{weinstein_cancergenomeatlas_2013}.
+Thus, while several case-wise NI methods have already been proposed, their implementation consisted of post-processing a static GRN to arrive at a case-wise GRN.
+As such, these methods will most likely recover the interactions that are prevalent in the whole population, and will miss interactions that are specific to only a sub-population.
+
+In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method.
+It uses a modified version of the RF variable importance scores used by GENIE3 and SCENIC to compute importance values for each profile and each interaction separately, as well as predict the effect of each interaction (activating or repressing).
+We generate case-wise GRNs -- or case-wise regulomes -- for 14'963 profiles from The Cancer Genome Atlas (TCGA) \cite{weinstein_cancergenomeatlas_2013}, resulting in 6'464'915 predicted case-wise interactions.
+We analyse these case-wise regulomes by clustering them and detecting highly activated interactions.
 
 \section{Results}
-
-\subsection{Hematopoietic cells from Tabula Muris}
-From the Tabula Muris \cite{schaum_singlecelltranscriptomics20_2018} project, we selected all cells involved in some part of the hematopoietic lineage tree.
-We computed case-wise GRNs between 22'122 remaining single cells, for the 2000 most variable target genes and a subset of transcription factors as regulators.
-
-To summarise the similarity in GRNs across cases, we first visualise a dimensionality reduction of the case-wise GRNs, where every dot represents a single GRN (Figure~\ref{fig:tabmur}A).
-The dimensionality reduction was computed by applying Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour ($k$NN) graph of the case-wise GRN vectors.
-On the same $k$NN graph, the cells were clustered using the Louvain \cite{blondel_fastunfoldingcommunities_2008} clustering algorithm, and the clusters were labelled using prior information from Tabula Muris.
-For each cluster, we retained the 50 interactions with the highest mean importance scores (Figure~\ref{fig:tabmur}B).
-In the visualisation, each node represents a gene, each edge an interaction, and the colour represents which cluster this interaction belongs to.
-
-For the most part, clusters that are proximate in the dimensionality reduction (Figure~\ref{fig:tabmur}A) are also closely connected in the GRN network view (Figure~\ref{fig:tabmur}B).
-Notably, interactions from the different B cell clusters are almost not connected amongst each-other, while in the dimensionality reduction they re very proximate.
-Retaining more edges could result in the different B cells being connected in the GRN network view.
-
-At the centre of the network lie many strongly connected interactions which are not specific for any particular cluster, but instead are common to almost all of the clusters.
-These are likely housekeeping genes or genes involved in key hematopoietic processes.
+At the time of writing, the TCGA database contains 14'963 profiles from 44 different cancer entities.
+We used the \texttt{bred} algorithm to infer case-wise regulomes for each of these profiles.
+In total, we detected 73'140 unique interactions and, on average, 7'231 active interactions per profile.
+Dimensionality reduction and clustering of the case-wise regulome data provides a visual overview of the different subpopulations (Figure~\ref{fig:tcga_dimred}).
 
 \begin{figure}[htb!]
-	{\raggedright\textbf{A}}
-	\begin{center}\includegraphics[width=\linewidth]{fig/tabula_muris_hema/plot_fr_cropped.png}\end{center}
-	{\raggedright\textbf{B}}
-	\begin{center}\includegraphics[width=.9\linewidth]{fig/tabula_muris_hema/grouped_interactions.pdf}\end{center}
+	\centering
+	\includegraphics[width=.6\linewidth]{fig/tcga/plot_fr_cropped.png}
 	\caption{
-		\textbf{22'122 case-wise GRNs of hematopoietic cells from the Tabula Muris project.}
-		\textbf{A:} Each dot in this dimensionality reduction corresponds to the GRN of a single cell.
-The dimensionality of the cell-specific regulome matix was reduced using Fruchterman-Reingold and were clustered using Louvain clsutering, after which the clusters were relabelled using the meta information from Tabula Muris.
-		\textbf{B:} Per cluster, the 50 interactions with the highest mean importance score are visualised.
+		\textbf{Visualisation of 14'963 case-wise regulomes of cancer profiles from The Cancer Genome Atlas project.} 
+		Dimensionality reduction of the case-wise regulomes was performed by applying Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour graph ($k=100$) of highly similar regulomes (Spearman correlation).
+The samples were clustered with Louvain clustering \cite{blondel_fastunfoldingcommunities_2008} and each cluster was assigned an ontology term from the NCI thesaurus ontology \cite{sioutos_ncithesaurussemantic_2007} which best fits the sample in the cluster.
+Clusters for which the term has a positive predictive value (PPV) lower than 0.5 are marked with an asterisk (*).
+More detailed information on the enrichment of particular terms or metadata for each of the clusters can be found in Figures~\ref{fig:cluslab1}--\ref{fig:cluslab7}.
 	}
-	\label{fig:tabmur}
+	\label{fig:tcga_dimred}
 \end{figure}
 
-\subsection{The Cancer Genome Atlas}
-We included all available RNA-seq profiles from The Cancer Genome Atlas.
-In total, we computed case-wise GRNs for 14'963 tumour samples from 50 sub-projects, including 44 different cancer entities.
-Clusters of case-wise GRNs were relabelled by the project they originate from, or by the cell types or organs the samples originate from.
-
-Most groups of interactions are highly specific to just one cancer type.
-Samples from the CPTAC project were split into two groups, clustering together with LUAD (lung adenocarcinoma) and kidney carcinoma samples.
-According to the meta-information of CPTAC profiles, these samples consist of lung, kidney, and uterus adenocarcinomas, explaining the split in CPTAC samples.
-
-%TODO: Extend !!
+For each cluster, we computed the average importance of each interaction.
+Retaining the 100 strongest interactions (Figure~\ref{fig:tcga_network}) shows that different cancer entities have vastly different regulomes, yet often have shared regulators and interactions (Table~\ref{tab:oncogenes}).
 
 \begin{figure}[htb!]
-	{\raggedright\textbf{A}}
-	\begin{center}\includegraphics[width=.8\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
-	{\raggedright\textbf{B}}
-	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/grouped_interactions_2.pdf}\end{center}
+	\centering
+	\includegraphics[width=.85\linewidth]{fig/tcga/grouped_interactions.pdf}
 	\caption{
-		\textbf{14'963 case-wise GRNs of cancer bulk profiles from The Cancer Genome Atlas project.}
-		\textbf{A:} The different profiles are visualised and clustered according to their regulomic similarities.
-		\textbf{B:} Visualisation of the strongest interactions per cluster shows both pathways distinct to particular cancer entities as well as pathways common to multiple cancer entities.
+		\textbf{Visualisation of the strongest interactions per cluster} shows both pathways distinct to particular cancer entities as well as pathways common to multiple cancer entities.
 	}
-	\label{fig:tcga}
+	\label{fig:tcga_network}
+\end{figure}
+
+Performing a more in-depth analysis of particular clusters can reveal distinct sub-populations within.
+For example, when comparing samples of normal and healthy tissue within breast carcinoma (Figure~\ref{fig:zoom17}), a significant number if interactions are predicted to have been turned off (top), while others have been turned on (middle, bottom).
+Similar observations can be made for other clusters, such as for melanocytic neoplasm (Figure~\ref{fig:zoom9}) and kidney carcinoma (Figure~\ref{fig:zoom12}).
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.7\linewidth]{fig/tcga/zoom_cluster17_brca_plot.pdf}
+	\includegraphics[width=.7\linewidth]{fig/tcga/zoom_cluster17_brca_heatmap.pdf}
+	\caption{
+		\textbf{In-depth view of cluster 17, breast carcinoma.} 
+		\textbf{Top:} A dimensionality reduction of the samples, coloured according to multiple sources of meta-data. 
+		\textbf{Bottom:} The samples were ordered linearly with SCORPIUS (see trajectory in \textbf{Top}) in order to visualise regulome activity in the form of a heatmap.
+	}
+	\label{fig:zoom17}
 \end{figure}
@@
 
-The \texttt{bred} algorithm is a novel approach for directly computing case-wise GRNs for single-cell omics and bulk omics profiles alike.
-We applied the method to 22'122 hematopoietic cells and 14'963 cancer profiles, showing that interactions are often grouped together in modules specific for certain cell types, tissue types, or cancer types.
+The \texttt{bred} algorithm is a novel approach for directly computing case-wise regulomes for single-cell omics and bulk omics profiles alike.
+We used \texttt{bred} to infer case-wise regulomes for 14'963 cancer profiles from The Cancer Genome Atlas project.
+Analysing the results using common omics algorithms (dimensionality reduction, clustering, enrichment) has shown that regulatory interactions are often specific to particular cancer types, but can also be shared between several cancer types.
+Ultimately, this analysis has resulted in a list of candidate oncogenic genes and interactions per cancer type, though further analysis and validation is required to confirm their oncogenicity.
 
-In this work, we only applied clustering methods to the case-specific 'regulome profiles', but other types of computational methods can be used to annotate and explore case-specific GRNs, such as trajectory inference and differential expression.
-
-Going forward, we will provide further functional validation of the results generated by \texttt{bred}, as well as benchmark the algorithm against various real and \textit{in silico} datasets.
+Future perspectives include applying it on single-cell omics data (e.g.
+from Tabula Muris \cite{schaum_singlecelltranscriptomics20_2018}) and benchmarking the algorithm against \textit{in silico} datasets (e.g.
+produced by \texttt{dyngen} \cite{cannoodt_dyngenbenchmarkingsilico_2019}).
 
@@
 
-\subsection{Inferring case-wise GRNs}
+\subsection{Inferring case-wise regulomes}
 The task of inferring a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely: for every target $T$, predict which of the potential regulators $R_i$ regulate $T$ (Figure~\ref{fig:simplify}B).
@@
 
+% TODO: describe cluster relabelling, see figures fig:cluslab
+
 \subsection{Visualising clustered GRNs}
@@
 
+\section{Supplementary information}
+
+\begin{table}[H]
+	\scriptsize
+	\caption{\textbf{Prioritisation of genes, ranked according to page-rank value.} The majority of these genes are already known to be involved in oncogenesis (data not shown).} \label{tab:oncogenes}
+	\begin{tabularx}{\linewidth}{XX}
+		\hline
+		Cluster name & Top 4 genes, page rank \\
+		\hline\hline
+		Acute Leukemia & HSPA1A, HSPA1B, HOXA10, HOXA9 \\
+		ALL & DDX3Y, XIST, RN7SL128P, TXLNGY \\
+		BL & KIAA0226L, XIST, TCL1A, NUGGC \\
+		BRCA & LINC00993, IRX5, ATP8A2P1, ENSG00000234918 \\
+		Connective and Soft Tissue Disorder * & HOXA11, HOXC9, HOXA10, HOXC10 \\
+		CPTAC \#1 & SLC22A2, CUBN, XIST, UMOD \\
+		CPTAC \#2 & RN7SL752P, SCARNA13, XIST, RNY1 \\
+		Diffuse Large B-Cell Lymphoma & PLA2G2D, ENSG00000224137, CCL25, MS4A1 \\
+		Digestive System Disorder & MEP1A, LGALS4, MUC13, CDX1 \\
+		Glioma & NCAN, GPM6A, MBP, OLIG2 \\
+		Kidney Carcinoma & ACSM2A, SLC22A2, NAT8, UMOD \\
+		LIHC & AGXT, ITIH3, ITIH1, HRG \\
+		LUAD & SFTPA2, SFTPD, NAPSA, SFTA2 \\
+		Melanocytic Neoplasm & MLANA, RPS4Y1, PMEL, TYR \\
+		MM & CD96, GPRC5D, FGFR3, IGLV1-41 \\
+		Peripheral Nervous System Disorder & TH, DBH, PENK, HAND2 \\
+		PRAD & TRGC1, NKX3-1, ACPP, RPL7P16 \\
+		Squamous Cell Carcinoma & TP63, SPRR2F, RPS4Y1, CDKN2B \\
+		THCA & ENSG00000240237, TG, TPO, RPS4Y1 \\
+		Transitional Cell Carcinoma & DHRS2, UPK2, PADI3, VGLL1 \\
+		UCEC & DLX5, MSX1, HOXB5, HOXB8 \\
+		\hline
+		\	\end{tabularx}
+\end{table}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=.95\linewidth,page=1]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=2]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=3]{fig/tcga/cluster_labelling.pdf}
+	\caption{
+		\textbf{Prioritisation of NCI terms and other forms of meta-data used to annotate clusters 1--3.} Cluster 2 is mainly made up of disorders in female reproductive organs, and should have been labelled as such.
+	}
+	\label{fig:cluslab1}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=.95\linewidth,page=4]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=5]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=6]{fig/tcga/cluster_labelling.pdf}
+	\caption{
+		\textbf{Prioritisation of NCI terms and other forms of meta-data used to annotate clusters 4--6.} Cluster 5 is mainly made up of renal cell carcinoma, and should have been labelled as such.
+Cluster 6 is aptly named, but mainly consists of neuroblastoma (NB), pheochromocytoma (PC), and paraganglioma (PG) disorders.
+	}
+	\label{fig:cluslab2}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=.95\linewidth,page=7]{fig/tcga/cluster_labelling.pdf}
+\includegraphics[width=.95\linewidth,page=8]{fig/tcga/cluster_labelling.pdf}
+\includegraphics[width=.95\linewidth,page=9]{fig/tcga/cluster_labelling.pdf}
+\caption{
+	\textbf{Prioritisation of NCI terms and other forms of meta-data used to annotate clusters 7--9.} 
+}
+\label{fig:cluslab3}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=.95\linewidth,page=10]{fig/tcga/cluster_labelling.pdf}
+\includegraphics[width=.95\linewidth,page=11]{fig/tcga/cluster_labelling.pdf}
+\includegraphics[width=.95\linewidth,page=12]{fig/tcga/cluster_labelling.pdf}
+\caption{
+	\textbf{Prioritisation of NCI terms and other forms of meta-data used to annotate clusters 10--12.} 
+	Cluster 11 contains mostly Lung Adenocarcinoma (LUAD), but also Lung Squamous Cell Carcinoma (LUSC), and thus should have been labelled Lung Carcinoma instead.
+}
+\label{fig:cluslab4}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=.95\linewidth,page=13]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=14]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=15]{fig/tcga/cluster_labelling.pdf}
+	\caption{
+		\textbf{Prioritisation of NCI terms and other forms of meta-data used to annotate clusters 13--15.} 
+		Cluster 13 contains a mixture of cancer types, and should have been labelled 'Mixed'.
+	}
+	\label{fig:cluslab5}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=.95\linewidth,page=16]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=17]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=18]{fig/tcga/cluster_labelling.pdf}
+	\caption{
+		\textbf{Prioritisation of NCI terms and other forms of meta-data used to annotate clusters 16--18.} 
+	}
+	\label{fig:cluslab6}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=.95\linewidth,page=19]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=20]{fig/tcga/cluster_labelling.pdf}
+	\includegraphics[width=.95\linewidth,page=21]{fig/tcga/cluster_labelling.pdf}
+	\caption{
+		\textbf{Prioritisation of NCI terms and other forms of meta-data used to annotate clusters 19--21.} 
+		Cluster 21 contains a mixture of cancer types, mostly lung carcinoma and endometrial carcinoma.
+	}
+	\label{fig:cluslab7}
+\end{figure}
+
+\subsection{Melanocytic neoplasm}
+
+\begin{figure}[H]
+	A
+	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/zoom_cluster9_melanocytic-neoplasm_plot.pdf}\end{center}
+	B
+	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/zoom_cluster9_melanocytic-neoplasm_heatmap.pdf}\end{center}
+	\caption{
+		\textbf{In-depth view of cluster 9, melanocytic neoplasm.} 
+		\textbf{A:} A dimensionality reduction of the samples, coloured according to multiple sources of meta-data. 
+		\textbf{B:} The samples were ordered linearly with SCORPIUS (see trajectory in \textbf{A}) in order to visualise regulome activity in the form of a heatmap.
+	}
+	\label{fig:zoom9}
+\end{figure}
+
+\subsection{Kidney carcinoma}
+
+\begin{figure}[H]
+	A
+	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/zoom_cluster12_kidney-carcinoma_plot.pdf}\end{center}
+	B
+	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/zoom_cluster12_kidney-carcinoma_heatmap.pdf}\end{center}
+	\caption{
+		\textbf{In-depth view of cluster 12, kidney carcinoma.} 
+		\textbf{A:} A dimensionality reduction of the samples, coloured according to multiple sources of meta-data. 
+		\textbf{B:} The samples were ordered linearly with SCORPIUS (see trajectory in \textbf{A}) in order to visualise regulome activity in the form of a heatmap.
+	}
+	\label{fig:zoom12}
+\end{figure}
+
 \clearpage

### 2019-10-21 ch6_bred/scripts/2_tcga.R

@@ -266,6 +266,7 @@ if (!file.exists(data_file)) {
 
 # SUBMIT BRED -------------------------------------------------------------
 list2env(read_rds(data_file), .GlobalEnv)
+write_lines(colnames(expr), paste0(data_dir, "genes.txt"))
 
 if (!file.exists(paste0(data_dir, "qsub_handle.rds"))) {
   regulators_filt <- regulators
@@ -386,227 +387,165 @@ if (!file.exists(paste0(data_dir, "dimred_and_clustering.rds"))) {
 # list2env(c(dimred_out, relabel_out), .GlobalEnv)
 list2env(read_rds(paste0(data_dir, "dimred_and_clustering.rds")), .GlobalEnv)
 
-palette <- labels %>% select(name, col) %>% deframe()
+if (!file.exists(paste0(data_dir, "grouped_interactions.rds"))) {
+  palette <- labels %>% select(name, col) %>% deframe()
 
-sample_info_df <- data.frame(
-  dimred_fr[sample_info$id,],
-  sample_info,
-  cluster = cluster[sample_info$id],
-  check.names = FALSE,
-  stringsAsFactors = FALSE
-) %>%
-  left_join(labels, by = "cluster") %>%
-  as_tibble()
-labeldf <- sample_info_df %>% group_by(name) %>% summarise(comp_1 = mean(comp_1), comp_2 = mean(comp_2))
-
-# make heatmap
-tab <- sample_info_df %>%
-  group_by(project_id, name) %>%
-  summarise(n = n()) %>%
-  mutate(pct = n / sum(n)) %>%
-  ungroup() %>%
-  reshape2::acast(name~project_id, value.var = "pct", fill = 0)
-tab <- tab[, order(apply(tab, 2, which.max))]
-gaps_col <- which(diff(apply(tab, 2, which.max)) != 0)
-pheatmap::pheatmap(tab, gaps_col = gaps_col, angle_col = 315, cluster_rows = FALSE, cluster_cols = FALSE, filename = paste0(data_dir, "cluster_heatmap.pdf"), width = 16, height = 8)
-
-# save visualisation
-g <-
-  ggplot(sample_info_df, aes(comp_1, comp_2, col = name)) +
-  geom_point(size = .5) +
-  shadowtext::geom_shadowtext(aes(label = name), labeldf, bg.colour = "white", size = 5) +
-  dynplot::theme_graph() +
-  coord_equal() +
-  scale_colour_manual(values = palette) +
[…]

### 2019-10-21 boek.tex
update thesis
change cites
update

@@
 \usepackage{float}
-\usepackage{multirow}
 \usepackage{import}
@@
 \usepackage{tcolorbox}
-\usepackage[export]{adjustbox}
+\usepackage[export]{adjustbox} % for top alignment
+\usepackage{afterpage} % for figure placement
 
@@
 % table packages
+\usepackage{multirow}
 \usepackage{tabularx}
@@
 \usepackage{bibentry}
-\nobibliography*
+%\nobibliography*
 
@@
 
+% if switching for quick renders
+\newif\ifrenderall
+
+\renderalltrue
+
 \begin{document}
@@
 \pagestyle{mainmatter}
-
-% TIP FROM WOUTER: Per chapter, write 'update' reflecting on what came out of the paper
 
@@
 
+\ifrenderall
 \newpage
@@
 \include{ch2_dyngen/content}
-
+\fi
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \include{ch3_dynbenchmark/content}
-
+\fi
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \printbibliography[heading=none]
+\fi
 
@@
 \include{ch5_scorpius/content}
+\fi
 
@@
 \begin{quote}
-	\textbf{Purpose:} \ldots \\
-	\textbf{Results:} \ldots \\
-	\textbf{Conclusion:} \ldots
+	\textbf{Purpose:} Network inference methods are computational tools which use large omics datasets to predict which genes are regulated by which transcription factors.
+	Since regulatory interactions are context-dependent, attempting to model regulatory dynamics in the form of a single regulatory network may have little relevance. \\
+	\textbf{Results:} In this work, we introduce \texttt{bred}, the first algorithm for inferring case-wise regulatory networks.
+We generate .
+Analysing case-wise regulatory networks of 14'963 profiles from The Cancer Genome Atlas project demonstrated that regulatory interactions are often specific to particular cancer types, but can also be shared between several cancer types.
+\\
+	\textbf{Conclusion:} Ultimately, this analysis has resulted in a list of candidate oncogenic genes and interactions per cancer type, though further analysis and validation is required to confirm their oncogenicity.
 \end{quote}
@@
 \include{ch6_bred/content}
-
+\fi 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \include{ch7_incgraph/content}
+\fi
 
@@
 \include{ch10_guidelines/content}
+\fi
 
@@
 \include{ch8_discussion/content}
-
+\fi 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-10-21 ch8_discussion/content.tex
~

@@
 
-We proposed a novel algorithm for inferring case-wise gene regulatory interactions.
-While we demonstrated the methodology by inferring 22'122 cell-specific GRNs and 14'963 sample-specific GRNs from the Tabula Muris and TCGA projects respectively, further interpretation of these results needs to be perform to further validate the usefulness of the method.
+We proposed a novel algorithm for inferring case-wise gene regulatory interactions (Chapter~\ref{chap:bred}).
+While we demonstrated the methodology by inferring 14'963 sample-specific GRNs from the TCGA project, further interpretation of these results needs to be perform to validate the usefulness of the method.
 

### 2019-10-21 chX_backmatter.tex
update
~

@@
 \section{Professional Experience}
-\textit{Ph.D. student in Bioinformatics} \\
+\textit{Ph.D. student in Computer Science} \\
 \hspace*{2 ex} Ghent University, Belgium, September 2013 -- now \\
@@
 	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor.
-Yearly download statistics are based on the number of downloads between 2019-08-01 and 2019-10-14.
+Yearly download statistics are based on the number of downloads between 2019-10-01 and 2019-10-21.
 CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
@@
 		Name & Role & Host & Downloads per year & Description \\ \hline\hline
-		\cranpkg{babelwhale} & aut & CRAN & 5810 & Interacting with Docker and Singularity containers \\
-		\cranpkg{diffusionMap} & aut & CRAN & 30'450 & Implements diffusion map method of data parameterization \\
+		\cranpkg{babelwhale} & aut & CRAN & 9800 & Interacting with Docker and Singularity containers \\
+		\cranpkg{diffusionMap} & aut & CRAN & 27'850 & Implements diffusion map method of data parameterization \\
 		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
-		\cranpkg{dyndimred} & aut & CRAN & 5535 & Applying dimensionality reduction methods \\
+		\cranpkg{dyndimred} & aut & CRAN & 4708 & Applying dimensionality reduction methods \\
 		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
@@
 		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
-		\cranpkg{dynparam} & aut & CRAN & 3265 & Creating meta-information for parameters \\
+		\cranpkg{dynparam} & aut & CRAN & 2810 & Creating meta-information for parameters \\
 		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
@@
 		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
-		\cranpkg{dynutils} & aut & CRAN & 13'130 & Common functionality for the dynverse packages \\
-		\cranpkg{dynwrap} & aut & CRAN & 990 & A common format for trajectories \\
-		\cranpkg{GillespieSSA} & aut & CRAN & 7880 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		\cranpkg{dynutils} & aut & CRAN & 32'668 & Common functionality for the dynverse packages \\
+		\cranpkg{dynwrap} & aut & CRAN & 4636 & A common format for trajectories \\
+		\cranpkg{GillespieSSA} & aut & CRAN & 7118 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
 		\cranpkg{GillespieSSA2} & aut & CRAN & 4950 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
@@
 		\cranpkg{incgraph} & aut & CRAN & 3565 & Incremental graphlet counting for network optimisation \\
-		\cranpkg{lmds} & aut & CRAN & 815 & Landmark Multi-Dimensional Scaling \\
+		\cranpkg{lmds} & aut & CRAN & 23'' & Landmark Multi-Dimensional Scaling \\
 		\cranpkg{princurve} & aut & CRAN & 29'100 & Fits a principal curve in arbitrary dimension \\
@@
 \section{Sources of funding}
-This study was supported by the Fonds Wetenschappelijk Onderzoek (11Y6218N).
+Robrecht Cannoodt was supported by the Fonds Wetenschappelijk Onderzoek (11Y6218N).
 

### 2019-10-21 ch1_introduction/analysis/download_stats.R

@@ -16,5 +16,5 @@ dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb
 
 min_date <- "2019-08-01"
 num_days <- as.integer(Sys.Date() - as.Date(min_date))
-summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% slice(match(cran, package))
+summ <- dls %>% filter(min_date <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% slice(match(cran, package))
 summ %>% print(n = 50)

### 2019-10-30 ch2_dyngen/analysis/example_methods_overview.R
removed dyngen submodule

@@ -1,0 +1,377 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(4)
+
+dir.create("fig/example_methods", showWarnings = FALSE, recursive = TRUE)
+
+backbone <- backbone(
+  module_info = tribble(
+    ~module_id, ~ba, ~burn,
+    "A", 1, TRUE,
+    "B", 0, FALSE,
+    "C", 0, FALSE,
+    "D", 0, FALSE,
+  ),
+  module_network = tribble(
+    ~from, ~to, ~effect, ~strength, ~cooperativity,
+    "A", "B", 1, 1, 2,
+    "B", "C", 1, 1, 2,
+    "B", "D", 1, 1, 2,
+    "C", "D", -1, 10, 2,
+    "D", "C", -1, 10, 2
+  ),
+  expression_patterns = tribble(
+    ~from, ~to, ~module_progression, ~start, ~burn, ~time,
+    "Burn0", "S1", "+A", TRUE, TRUE, .5,
+    "S1", "S2", "+B", FALSE, FALSE, .5,
+    "S2", "S3", "+C", FALSE, FALSE, 1,
+    "S2", "S4", "+D", FALSE, FALSE, 1,
+  )
+)
+backbone$module_info$color <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3")
+model <-
+  initialise_model(
+    num_tfs = 8,
+    num_targets = 10,
+    num_hks = 0,
+    num_cells = 100,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    distance_metric = "euclidean",
+    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
+    simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+## grn
+plot_backbone_statenet(model) %>% ggsave(filename = "fig/example_methods/statenet.pdf", width = 6, height = 6)
+plot_backbone_modulenet(model) %>% ggsave(filename = "fig/example_methods/modulenet.pdf", width = 8, height = 6)
+
+plot_feature_network(model, show_targets = FALSE)
+
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE)
+plot_feature_network(model)
[…]

### 2019-10-30 ch2_dyngen/analysis/multiple_backbones.R

@@ -1,0 +1,191 @@
+library(tidyverse)
+library(dyngen)
+library(dynplot)
+
+backs <- list_backbones()
+
+oks <- c("linear", "bifurcating", "converging", "cycle", "bifurcating_loop", "consecutive_bifurcating")
+oks <- c(oks, "bifurcating_cycle", "bifurcating_converging", "binary_tree", "branching", "disconnected")
+oks <- c("binary_tree", "branching", "disconnected")
+
+for (seed in seq_len(10))
+for (nb in oks) {
+  cat("=============== SIMULATING ", nb, " seed ", seed, " ===============\n", sep = "")
+  set.seed(seed)
+
+  out_dir <- paste0("fig/example/bb_", nb, "_", seed, "/")
+  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
+
+  if (!file.exists(paste0(out_dir, "traj_dimred.pdf"))) {
+    back <- backs[[nb]]()
+    model <-
+      initialise_model(
+        num_tfs = nrow(back$module_info) * 2,
+        num_targets = 100,
+        num_hks = 20,
+        num_cells = 1000,
+        backbone = back,
+        verbose = TRUE,
+        download_cache_dir = "~/.cache/dyngen",
+        num_cores = 4,
+        gold_standard_params = gold_standard_default(
+          tau = .001,
+          census_interval = .02
+        ),
+        simulation_params = simulation_default(
+          ssa_algorithm = GillespieSSA2::ssa_etl(tau = .001),
+          num_simulations = ifelse(nb %in% c("binary_tree", "branching", "disconnected"), 80, 16),
+          census_interval = .05,
+          burn_time = simtime_from_backbone(back),
+          total_time = simtime_from_backbone(back)
+        )
+      )
+
+    out <- generate_dataset(model, make_plots = TRUE)
+
+    # write_rds(out, paste0(out_dir, "/out.rds"), compress = "gz")
+    ggsave(paste0(out_dir, "plot.pdf"), out$plot, width = 20, height = 15)
+
+    dataset <- out$dataset
+    dimred <- dyndimred::dimred_landmark_mds(dataset$expression, distance_method = "pearson")
[…]

### 2019-11-09 ch9_guidelines/content.tex
rename chapters

--- 2019-09-20 ch8_guidelines/content.tex
@@
+% GUIDELINES FOR BENCHMARKING
+
 \section{Introduction}
-
 Many fields of computational research are characterized by a growing number of available methods for data analysis.
@@
 
+\section{Ten essential guidelines}
 The review is structured as a series of guidelines (Figure \ref{fig:summary_guidelines}), each explained in detail in the following sections.
@@
 We hope these guidelines will assist computational researchers to design high-quality, informative benchmarks, which will contribute to scientific advances through informed selection of methods by users and targeting of research efforts by developers.
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-11-09 boek.tex
technical fixes

@@
 	\textbf{Results:} We developed \texttt{dyngen}, a multi-modality simulator of single cells. 
-	In dyngen, the biomolecular state of an \textit{in silico} cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
+	In \texttt{dyngen}, the biomolecular state of an \textit{in silico} cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
 By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%   CH9: Self-assessment    %%%
+%%%   CH8: Self-assessment    %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newpage
-\graphicspath{{./ch9_selfassessment/}}
-\include{ch9_selfassessment/content}
+\graphicspath{{./ch8_selfassessment/}}
+\include{ch8_selfassessment/content}
 \clearpage
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH10: Guidelines    %%%
+%%%       CH9: Guidelines    %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newpage
-\graphicspath{{./ch10_guidelines/}}
-\include{ch10_guidelines/content}
+\graphicspath{{./ch9_guidelines/}}
+\include{ch9_guidelines/content}
 \fi
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%      CH8: DISCUSSION      %%%
+%%%      CH10: DISCUSSION      %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newpage
-\graphicspath{{./ch8_discussion/}}
-\include{ch8_discussion/content}
+\graphicspath{{./ch10_discussion/}}
+\include{ch10_discussion/content}
 \fi 

### 2019-11-09 ch2_dyngen/content.tex
technical fixes

% DYNGEN CONTENT

%\begin{quote}
%	\textbf{Abstract:} 
%\end{quote}

\section{Introduction}
Continuous technological advancements to single-cells omics are having profound 
effects on how researchers can validate biological hypotheses. 
Early experimental technologies typically only allowed profiling a single modality (e.g. DNA sequence, 
RNA or protein expression). However, recent developments permit profiling multiple modalities simultaneously,
and every modality added allows for new types of analyses that can be performed.

This presents method developers with a problem. The majority of the 250+ peer-reviewed computational tools for analysing single cell omics data were published without a quantitative assessment of the accuracy of the tool. This is partially due to low availability of suitable benchmarking datasets; even if there are sufficient suitable input datasets available, these are often not accompanied by the necessary metadata to serve as ground-truth for a benchmark.

Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data. Generators of scRNA-seq data (e.g. splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers \cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers \cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.

An ideal experiment would be able to observe all aspects of a cell, including a full history of its 
molecular states, spatial positions and environmental interactions \cite{stuart_integrativesinglecellanalysis_2019}.
While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow already developing the next wave of computational tools.

We introduce \texttt{dyngen}, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:showcase}A). A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} where a cell consists of a set of molecules. Throughout a simulation, reactions (e.g. transcription, splicing) modify the abundance of these molecules, and the likelihood of a reaction occurring is again dependent on molecule abundance.

By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:showcase}B). We demonstrate \texttt{dyngen}'s flexibility by simulating snapshot, time-series and pertubation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/showcase_4} 
	\caption{
		\textbf{Showcase of \texttt{dyngen} functionality.} 
		\textbf{A:} 
		\textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
	}
	\label{fig:showcase}
\end{figure}

\section{Results}

A cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A). 
A gene regulatory network (GRN) define the reactions that are allowed to occur (Figure \ref{fig:simplecyclic}B), and is constructed such that a cell will develop over time (Figure \ref{fig:simplecyclic}C--D).
The likelihood of a reaction occurring is a function of the abundance of key molecules involved in each reaction (Figure \ref{fig:simplecyclic}E).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/simplecyclic_edited} 
	\caption{
		\textbf{Key features of \texttt{dyngen} are illustrated using a cyclic toy example.}
		While this example comprises only of a single cell and 5 genes, \texttt{dyngen} is able to simulate thousands of cells with GRNs containing thousands of genes.
	}
	\label{fig:simplecyclic}
\end{figure}

\texttt{dyngen} returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F). These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches. For example, a network inference method could use mRNA abundance and cellular states as inputs, and its output could be benchmarked against the gold standard GRN.

Depending on how the GRN is designed, different cellular developmental processes can be simulated.
\texttt{dyngen} includes generators of GRNs which result in many different developmental topologies (Figure~\ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.

\begin{figure}[ht]
	\centering
	\includegraphics[width=\LARgefigure]{fig/example_runs_2} 
	\caption{
		\textbf{Multiple executions of \texttt{dyngen} with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled. 
		\textbf{A:}~Linear. \textbf{B:}~Bifurcating. \textbf{C:}~Converging.
		\textbf{D:}~Cyclic. \textbf{E:}~Bifurcating loop. \textbf{F:}~Bifurcating converging.
		\textbf{G:}~Consecutive branching. \textbf{H:}~Binary tree. \textbf{I:}~Disconnected.
	}
	\label{fig:example_runs}
\end{figure}

Custom-defined GRNs offer more fine-grained control over the simulation.
Aside from simulating topologies currently not supported by \texttt{dyngen}, this has several other important use-cases. Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure~\ref{fig:usecases}). 
Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.

\begin{figure}[ht]
	\centering
	\includegraphics[width=\LARgefigure]{fig/usecases} 
	\caption{
		\textbf{Examples of simulations with perturbed GRNs.}  
		\textbf{A:} The cells in the top half were simulated with the same GRN as the cells on the bottom half, except that all parameters (e.g. strength of the interaction) were randomised.
		\textbf{B:} Only 10 interactions in the GRN were randomised. In this example, the effect of the GRN perturbation is more subtle.
	}
	\label{fig:usecases}
\end{figure}

\texttt{dyngen} can be used to simulate different experimental conditions. By default, \texttt{dyngen} supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation). 
However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point. This would allow evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

% YSA: RNA velocity: Onduidelijk hier, mensen kennen dit niet.

%\subsection{Example use cases}
%\subsubsection{Trajectory alignment}
%\subsubsection{Differential network inference}
%\subsubsection{RNA velocity}
%\subsubsection{Perturbation experiment}

\section{Discussion}
As is, \texttt{dyngen}'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications. This may include methods that integrate clustering, network inference and trajectory inference. In this respect, \texttt{dyngen} may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.

\texttt{dyngen} ultimately allows anticipating technological developments in single-cell multi-omics. In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing the qualitative and robust results in downstream analysis.

Currently, \texttt{dyngen} focuses on simulating cells as standalone entities that are well mixed.
Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration \cite{smith_spatialstochasticintracellular_2019}.

% We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics.
%It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
%This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
%\texttt{dyngen} has already been used successfully to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods. Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods. % TODO: And RNA velocity?

\section{Methods}
The workflow to generate \textit{in silico} single cell data consists of six main steps (Figure \ref{fig:explain_methods}).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\hugefigure]{fig/explain_methods} 
	\caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network. \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the molecules at each time step are extracted. Each time step is mapped to a point in the backbone. \textbf{F:} Multiple cells are sampled from each simulation. Molecules are sampled from each cell.}
	\label{fig:explain_methods}
\end{figure}

\subsection{Defining the backbone: modules and states} \label{sec:backbone}

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}. One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other \cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015}, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}. Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}. Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.

In \texttt{dyngen}, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules. A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen_ssa}). At the end of the \texttt{dyngen} process, cells will not be sampled from the burn phase simulations. Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).

Several examples of module networks are given (Figure \ref{fig:example_backbones}). 
A simple chain of modules (where one module upregulates the next) results in a \emph{linear} process. By having the last module repress the first module, the process becomes \emph{cyclic}. Two modules repressing eachother is the basis of a \emph{bifurcating} process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process. Finally, a \emph{converging} process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.

Note that these examples represent the bare minimum in terms of number of modules used. Using longer chains of modules is typically desired. In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path. If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen_backbone_lego}).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\LARGEfigure]{fig/example_backbones} 
	\caption{Example module networks}
	\label{fig:example_backbones}
\end{figure}

In addition to the module network, the user also needs to define a network of cellular states called the "backbone". Before simulating any cells, each transition in the backbone is simulated separately to obtain the average changes in expression along that transition (Figure \ref{fig:explain_methods}D). As part of the backbone, the user needs to specify which modules are allowed to alter its expression from one state to another. For example, in order to transition from state S0 to S1 in the cyclic example, gene modules A, B and C are turned on and a simulation is allowed to run. To transition from S1 to S2, gene modules D and E are turned on, and expression of gene module C is kept constant. To transition from S2 to S3, C is turned on again and now A and B are fixed. Finally, to transition from S3 to S1 again, A and B are turned on again and D and E are fixed again. Demonstrations of the backbone will be explained in more detail in section \ref{sec:dyngen_sim_backbone}.

\subsubsection{Backbone lego} \label{sec:dyngen_backbone_lego}
The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure \ref{fig:backbone_lego}). A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure \ref{fig:backbone_lego}A). Parameters allow determining the number of modules involved in the process and the number of outputs. Multiple BBLs can be chained together in order to intuitively create milestone networks and corresponding state networks (Figure \ref{fig:backbone_lego}B). Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\LARGEfigure]{fig/backbone_lego} 
	\caption{Backbone lego}
	\label{fig:backbone_lego}
\end{figure}

When the input node of a \textbf{linear BBL} (Figure \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated. A \emph{simple chain} is a set of modules where a module upregulates the next. A \emph{chain with double repression} has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression. A \emph{grid with double repression} is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain. Finally, a \emph{flip flop} consists of a simple chain where first the modules (except the last) are upregulated. Once the second to last module is upregulated, that module upregulates itself and the first module is strongly repressed, causing all other modules to lose expression and finally the last module to be upregulated. The \emph{flip flop} retains this output state, even when the input changes.

When the input node of a \textbf{branching BBL} (Figure \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated. A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules. Due to its simplicity, however, multiple output modules might be upregulated simultaneously, and over long periods of simulation time it might be possible that the choice of upregulated module changes. A \emph{robust branching} improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.

A \textbf{leaf BBL} (Figure \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs. A \emph{start} BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation (Section \ref{sec:dyngen_sim_backbone}). An \emph{end} BBL is also a linear BBL with its output regulating one final module.

\subsection{Generating the gene regulatory network}
The GRN is generated based on the given backbome in four main steps (Figure~\ref{fig:gen_feature_network}).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/gen_feature_network} 
	\caption{
		\textbf{Generating the feature network from a backbone consists of four main steps.}
	}
	\label{fig:gen_feature_network}
\end{figure}

\paragraph{Step 1, sampling the transcription factors (TF).} The TFs are the main drivers of the molecular changes in the simulation. The user provides a backbone and the number of TFs to generate. Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$). A TF inherits the 'burn' and 'basal expression' from the module in belongs to.

\paragraph{Step 2, generating the TF interactions.} Let each TF be regulated according to the interactions in the backbone. These interactions inherit the effect, strength, and cooperativity parameters from the interactions in the backbone. A TF can only be regulated by other TFs or itself.

\paragraph{Step 3, sampling the target subnetwork.} 
A user-defined number of target genes are added to the GRN. Target genes regulated by a TF or another target gene, but is always downstream of at least one TF. To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently existing TFs are mapped to regulators in the FANTOM5 GRN. The targets are drawn from the FANTOM5 GRN, weighted by their page rank value. For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$). The interactions connecting a target gene and its regulators are added the GRN.

\paragraph{Step 4, sampling the housekeeping subnetwork.}
Housekeeping genes are completely separate from any TFs or target genes. A user-defined set of housekeeping genes are also sampled from the FANTOM5 GRN. The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$). A random gene is sampled, and a breadth-first-search is performed to sample the desired number of housekeeping genes.

\subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
% paper on transcriptional bursting
% doi: 10.1039/C7MB00154A
% interesting to try to replicate in dyngen! :)
\newcommand{\w}[1]{\text{w}_{#1}}
\newcommand{\x}[1]{\text{x}_{#1}}
\newcommand{\y}[1]{\text{y}_{#1}}

\newcommand{\rs}[1]{\text{R}_{#1}}
\newcommand{\rp}[1]{\text{R}^+_{#1}}
\newcommand{\rn}[1]{\text{R}^-_{#1}}

\newcommand{\wpr}[1]{\text{wpr}_{#1}}
\newcommand{\wsr}[1]{\text{wsr}_{#1}}
\newcommand{\xdr}[1]{\text{xdr}_{#1}}
\newcommand{\ypr}[1]{\text{ypr}_{#1}}
\newcommand{\ydr}[1]{\text{ydr}_{#1}}

\newcommand{\str}[1]{\text{str}_{#1}}
\newcommand{\co}[1]{\text{co}_{#1}}
\newcommand{\ind}[1]{\text{ind}_{#1}}
\newcommand{\hmy}[1]{\text{hmy}_{#1}}
\newcommand{\reg}[1]{\text{reg}_{#1}}
\newcommand{\ba}[1]{\text{ba}_{#1}}

For every gene $G$ in the GRN, the simulation will keep track of the abundance levels of three molecules, a pre-mRNA, a mature mRNA and a protein. These abundance levels are represented as $\w G$, $\x G$ and $\y G$ respectively.

Throughout a simulation, the abundance levels of molecules are affected by five reactions, namely transcription, splicing, mRNA degradation, translation, and protein degradation. Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. The effects and propensity functions of these reactions are defined in Table \ref{tab:reaction_def}.

%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
% TODO: add step-by-step  derivation of transcription formula
%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

% YSA: Graag wat meer uitleg in de tekst over al deze formules, definitie van notaties en zo.

\begin{table}[ht]
	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
	\centering
	\begin{tabular}{|lcc|}
		\hline
		Reaction & Effect & Propensity \\ \hline \hline
		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \reg{G,H})}$ \\
		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
		mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
		Translation & $\x G \rightarrow \w G + \y G$ & $\ypr G \times \x G$ \\
		Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
	\end{tabular}
\end{table}

% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]). y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well. 
% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
% are independent and result in multiplicative effect.

% according to https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Proportion_of_ligand-bound_receptors:
% P_H = (y_H ^ N) / (degradation_H / production_H + y_H ^ N)
% ... 
% ... 
% might need to go back to things like:
% https://kar.kent.ac.uk/24077/1/myzabetpaper.pdf

\begin{table}[ht]
	\caption{\textbf{Default parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
	\centering
	\begin{tabular}{|lrl|}
		\hline
		Parameter & Symbol & Definition \\ \hline \hline
		Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
		Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
		mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
		Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
		Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
		Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\ 
		Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
    Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
		TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\ 
		Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
		Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
		\multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
	\end{tabular}
\end{table}

\subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state. It is assumed the root state has no modules active and has no expression of any molecules. To get to the next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition. To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.

% mention that this is basically an ode?

\subsection{Simulate single cells} \label{sec:dyngen_ssa}
%TODO: STOCKS \cite{kierzek_stocksstochastickinetic_2002} uses GSSA to simulate something.
% maybe we could have a look at that! 
% or https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x

% or even https://academic.oup.com/bioinformatics/article/24/10/1318/178687 ('AND' and 'OR')
% and https://academic.oup.com/bioinformatics/article/25/9/1205/203870 (regulation motifs, kinetics)
% https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x (modules of genes)

%TODO: improve this section a lot. Basically, explain SSA and GillespieSSA2.

\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.

\texttt{GillespieSSA2} is an optimised library for performing SSA simulations. The propensity functions are compiled to C++, and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration. The framework also allows to store the abundance levels of molecules only after a specific interval has passed since the previous census. By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information. In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.

\subsubsection{Map SSA simulations to backbone}
We compute the Pearson correlation between the state vectors in the simulation and the average expression levels along a transition in the backbone. Each timepoint in the SSA simulation is mapped to the point in the backbone that has the highest correlation value.

\subsection{Simulate experiment}
From the SSA simulation we obtain the abundance levels of all the molecules at the different time points. We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data. For this, the cells are sampled from the simulations, and molecules are sampled for each of the cells. Real datasets are used in order to achieve similar data characteristics.

\subsubsection{Sample cells}
Cells can be sampled from an unsynchronised population of single cells (snapshot) or at multiple time points in a synchronised population (time series).

\paragraph{Snapshot} Cells are sampled uniformly from the different time points in the simulation.

\paragraph{Time series} The timeline of the simulations is divided into multiple chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.

\subsubsection{Sample molecules} 
Molecules are sampled from the simulation to replicate how molecules are experimentally sampled. A real dataset is downloaded from a repository of single-cell RNA-seq datasets \cite{cannoodt_singlecellomicsdatasets_2018}. For each \textit{in silico} cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset. The capture rate $cr_j$ of each \textit{in silico} molecule type $j$ is drawn from $N(1, 0.05)$. Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.

\clearpage
\section{References}
\printbibliography[heading=none]

### 2019-11-09 ch7_incgraph/content.tex
technical fixes

@@
 %% FIGURE 6
-\picfig{fig/figure6.pdf}{\Largefigure}{fig_reranking}{{\bf A simple graphlet-based scoring method improves predicted regulatory networks.}
+\picfig{fig/figure6.png}{\Largefigure}{fig_reranking}{{\bf A simple graphlet-based scoring method improves predicted regulatory networks.}
 	(A) The F1 score was calculated by calculating the harmonic mean of the AUROC and AUPR scores of the first 1000 interactions.

### 2019-11-20 rebuttal/rebuttal.tex
work on rebuttal

\PassOptionsToPackage{table}{xcolor}
\documentclass[10pt]{article}

%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORT PACKAGES %%%
%%%%%%%%%%%%%%%%%%%%%%%

% adjust margins (should be done before importing fancyhdr!)
\usepackage[margin=2.5cm,a4paper]{geometry}

% text packages
\usepackage{csquotes}
\usepackage[english]{babel}
\usepackage{setspace}

% style packages
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
\usepackage{parskip} % disable paragraph indent and set paragraph skipping
\setlength{\emergencystretch}{3em}  % prevent overfull lines

% used for defining abbreviations, see nomenclature.tex
\usepackage{nomencl}

% figure packages
\usepackage{graphicx}
\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
\usepackage{float}
\usepackage{import}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{tcolorbox}
\usepackage[export]{adjustbox} % for top alignment
\usepackage{afterpage} % for figure placement

% symbols packages
\usepackage{amssymb}
\usepackage{amsmath}

% references packages
\usepackage[sorting=none]{biblatex}
\bibliography{library}
\usepackage{bibentry}
%\nobibliography*

% link packages
\usepackage{url}
\usepackage{doi}

%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE STYLES %%%
%%%%%%%%%%%%%%%%%%%%%

% set font after chapter sonny style
\usepackage{fontspec}
\setmainfont [Path = ../fonts/,
UprightFont = *-300,
ItalicFont = *-300-Italic,
BoldFont = *-700,
BoldItalicFont = *-700-Italic
]{MuseoSans}

\hypersetup{
	colorlinks = true,
%	linkcolor = {blue},
  urlcolor = {blue},
	citecolor = {med-gray}
}
\newcommand{\exam}[2][\  ]{\hspace{0pt}\marginpar{\color{black}#1}$\triangleright$ \textit{\color{gray}#2}}
%\newcommand{\exam}[2][\  ]{\begin{displayquote}\hspace{0pt}\marginpar{\color{black}#1}\textit{\color{gray}#2}\end{displayquote}}
\newcommand{\imp}[1]{{\color{blue} #1}}
\newcommand{\answ}[1]{{\color{red} Answer: #1}}

\usepackage[geometry]{ifsym}

\newcommand{\bigexclaim}{\raisebox{-0.1em}{\BigTriangleUp}\hspace{-0.32em}\llap{\small\textbf{!}}\hspace{0.32em}}

\usepackage{fdsymbol}
\newcommand{\tagimp}{\bigexclaim}
\newcommand{\tagtime}{{\Large $\hourglass$}}

\begin{document}

\section{Joeri Ruyssinck}

\subsection{Strengths of the thesis}

\exam{The research domain of this thesis is young and technical evolutions in the
	encompassing field move fast. This makes it especially challenging to do
	research in the field, as often the original PhD thesis subject is obsolete at the
	end. The candidate has done exceptionally well to not only suggest and create
	novel algorithms which are at the state-of-the-art but also to grasp and tackle
	the greater high level challenges in the field, resulting in a potential seminal
	benchmarking paper.}

\exam{The thesis as a whole contains many relevant contributions, of which several
	have been published in high quality journals.}

\exam{The figures in the thesis are of exceptional quality. They are creative and able
	to convey large volumes of information to the reader in a visually pleasing
	format.}

\exam{The introduction chapter of the thesis demonstrates that the candidate
	understands and is able to convey the background of both (computational)
	biology and computer science.}

\exam{The passion of the candidate to bring modern good software development
	practices to the field is present throughout the work and might contribute to a
	more mature algorithm development in computational biology.}

\subsection{Weaknesses of the thesis}
\exam[\tagimp \tagtime]{\imp{The quality of writing, layout and graphical presentation is of a lower quality in the non-published parts of the thesis. Certain unpublished chapters feel rushed and unfinished. Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.}}

\exam[\tagimp \tagtime]{\imp{The Introduction and Discussion chapter is missing a more broad context
	description. After reading the work, I gained very little insight how other
	researchers are using these tools to advance science.}}

\exam{The main contribution of this work is the excellent benchmarking study
	published in Nature Biotechnology. \imp{However, this also effects the overall
	cohesiveness of the work, as information is repeated throughout the book at
	several levels of detail.} This however a common side-effect of grouping
	published manuscripts into a thesis and I’m aware there is no (quick) solution
	to this.}

\exam[\tagimp]{It is very rare to see that research groups (or individual researchers) invest this
	much effort and time in creating qualitative software. I agree with the ideas of
	the candidates that these practices to adopt software to accompany research
	ideas should be encouraged. \imp{Yet I wonder how the current academic system
	can support such concepts. Academics often hold different positions in
	different research groups around the world, who should be the maintainer in
	these cases? The original PhD students that developed the work? The original
	research group? Will such higher standards not lead to barriers which might
	be harder to overcome for PhD students which might be constrained by limited
	support or funding?}}

\subsection{Summary and chapter 1}

\exam[\tagimp \tagtime]{As an interdisciplinary thesis, it is very much enjoyable that the introduction
	chapter contains a concise, yet precise overview of the concepts which are
	needed to understand the work of the candidate. I believe the chapter is nicely
	written and especially the flow of information is well chosen. However, the
	candidate only briefly touches upon the societal value of the technological
	innovations he describes. For a PhD thesis, \imp{I think it would be an added value
	if the candidate could briefly describe the practical implications of the
	technological advances (e.g. precision medicine, new treatments) and
	motivate the research from a non-technical view point (How is/can this PhD
	change the world?)}}

\exam{\imp{Small details / typo's:} [\ldots]} \\
\answ{These issues have been fixed, unless noted below.}

\exam{\imp{Figure 1.9 does not contain SCORPIUS (probably published before,
	but still a bit weird for a PhD containing the manuscript)}} \\
\answ{Indeed, this figure was adapted from Cannoodt et al. 2016 and serves as an illustration that
TI methods often have common building blocks. Extending this figure to >75 TI methods would decrease the interpretability of the figure.}

\exam{\imp{Bold is only used in 1.1.2 and research challenges}} \\
\answ{Usage of the bold font style in 1.1.2 and 1.3 serve clear, but different, purposes. In 1.1.2, three words are highlighted to establish a link between each term's introduction in the first paragraph, and their explanations in the second paragraph. In 1.3, important sections of each research objective are highlighted in order to highlight the structure of this section.}

\exam{\imp{Use of we/I in summary}} \\
\answ{Usage of first person pronouns in this dissertation is very limited. In the summary of my contributions to the field, I believe usage of the first person improves legibility in comparison to when the phrases in question are converted to a passive form.}

\subsection{Chapter 2}
\exam{\imp{It is not clear to me what is meant with ‘such as sampling a cell at a certain
	time point and once more at a later point’ and why this is currently not
	possible?}}

\exam{\imp{Figure 2.1.A. What does the figure in ‘Combine simulations’ represent? Are
		these single simulations which are mapped continuously to the backbone?}}

\exam{\imp{Figure 2.5. It is not explained what Subfigure F represents.} Only after the
	reader has completed the chapter, it becomes clear that the red lines probably
	represent snapshots and the lines, curves of gene expression.}

\exam{\imp{Figure 2.6: Consider changing the dashed lines}, as they are not (clearly)
		visible printed}

\exam{\imp{Figure 2.6 and the example text for the cyclical example is unclear.}}

\exam{\imp{While I
		understand that the instructions result in the desired state change, I also see
		many other configurations which would result in a cyclical example.}}

\exam{\imp{Why is it currently needed to explicitly keep certain module expressions constant to
		determine the backbone? Would the same result not be achieved by simply
		simulating X times and averaging expression levels?}}

\exam{\imp{Figure 2.7. The figure legend could be extended}}

\exam{\imp{The introduction of FANTOM5 is abrupt and it would be better if the context
		would be briefly explained to the reader.}}

\exam{\imp{The paragraph describing Step 3 is unclear: -> Target genes ARE? Regulated
		... but is-> are ?}}

\exam{\imp{Figure 2.8 can incorrectly suggest that target genes can only have a single
		regulator assigned}}

\exam{\imp{Table 2.2. contains many default ‘magic’ numbers?} It would be nice to know
		how in practice you ended up selecting them and which problems you
		encountered during design.}

\exam{\imp{Since these are default values, it would suggest
		that the user can modify them? Why would a user do that and what would be
		the impact?}}

\exam{\imp{Am I correct that currently splicing offers little extra value as the formula’s
		would imply that alternative splicing is not supported?}}

\exam[\tagimp \tagtime]{\imp{During a first read, it was confusing at times to read specifically about the
	strategy to determine the backbone while it had not been described how the
	simulation was performed.} At a certain time I was under the impression that
	dyngen did not simulate cell evolution end-to-end but only between states.
	Would the readability be improved if the order was changed in which the
	concepts are explained? The backbone is only needed for the visualisation.}

\exam{\imp{2.4.6. is a bit worrying. Could I reproduce dyngen without knowing these
		details? Does the sampling process not deserve more explanation?}}

\exam{\imp{Small details / typo's:} [\ldots]} \\
\answ{These issues have been fixed.}

\subsection{Chapter 3}

\exam{There are some artefacts of converting the paper into a PhD chapter: e.g.
	references to Supplementary tables/figures should be changed to the actual
	table number.} \\
\answ{Certain supplementary figures and files had not been included in the dissertation, but instead the reference contained a clickable URL which directs the reader to a downloadable file. Since this information is lost to readers of the printed dissertation, explicit captions have been added in a section for supplementary figures and tables, and contain a printable URL which direct the reader towards the file.}

\exam{Some words are marked with an asterisk but the footnote? is
	missing.} \\
\answ{Section 3.4.2 uses an asterisk to denote special cases and does not refer to a missing footnote. This notation is introduced in the paragraph leading up to the usage thereof. An actual asterisk has been added to the introduction of the notation to improve readability of this section.}

\exam{\imp{You mention consensus predictions, is this something you investigated more
		in depth?} Would ‘ensemble’ predictions result in better performance? Is there
		a good way to aggregate the predictions?}

\exam{Monocle DDRTree seems to overestimate the topology based on Figure 3.5, \imp{is
	this somehow related to the way you have to force the algorithm into the
	common format?}}

\exam{\imp{The dyngen described here seems to be less advanced than the one from
		Chapter 2?}}

\exam{\imp{Small details / typo's:} [\ldots]} \\
\answ{These issues have been fixed, unless noted below.}

\exam{\imp{Consistency: this chapter is US English}}

\subsection{Chapter 4}

\exam{Figure 4.7: \imp{The figure is lacking what the colors represent.}}

\exam{Figure 4.7: I see that even if you use the same dimensionality
	reduction; there is still a lot of variation between the methods on a very simple
	example. \imp{Could you clarify
	how often these methods disagree on trivial examples?}}

\exam{Dyno offers a lot of
	guidance with respect to how to choose your methods, run them, etc. \imp{Did you
	consider support for features such as interpretation or automated error
	detection after prediction?}}

\exam[\tagimp \tagtime]{\imp{This chapter does not really read as a scientific paper (yet).} There are few
	comments I have since this is mainly an instruction manual + feature overview.}

\exam{\imp{Small details / typo's:} [\ldots]} \\
\answ{These issues have been fixed, unless noted below.}

\exam{\imp{Figure 4.7 is missing slingshot and paga label in part B}} \\
\answ{Figure 4.7 does contain labels for Slingshot and PAGA. A small gap between part A and part B has been added, to improve interpretability.}

\subsection{Chapter 5}

\exam[\tagimp]{\imp{Comparing the preprint with the chapter in the thesis. Some minor changes
	seem to have been made to SCORPIUS: e.g. I believe in the pre-print
	classical Torgerson MDS is mentioned, while the more optimized Landmark
	MDS is used in the chapter and final version. There is also no mention of
	outlier removal. The main result Figure 5.4 is however identical. Did this
	change or other potential changes have no (measurable) effect on the initial
	analysis? I also noticed that the biological conclusions or hypothesis have
	been reduced in claim strength in the paper or have been removed. It would
	be interesting to know if this is due to new insights that have been found in the
	period 2016-> 2019 or mainly because the paper’s target audience has
	shifted.}} \\
\answ{As mentioned, the preprint was originally published in 2016. In order to submit this work to a peer-reviewed journal in 2019, the manuscript and the software was rewritten to reflect developments within the field.

Regarding the methodology. Classical MDS has indeed been replaced with Landmark MDS, since this allows scaling beyond 10'000 cells. Principal curves has been modified to also allow scaling beyond 10'000 cells. These changes have little to no effect on any results. Outlier removal was removed because this step is now part of mainstream preprocessing pipelines such as Seurat. Initialisation of the curve (k-means clustering + shortest path) has been removed -- while this part of the algorithm showed improved performance when applied to the datasets included in the bioRxiv benchmark, this was not the case when applied to a larger set of datasets.

Regarding reducing the biological conclusions. Section 5.2.2 regarding the functional modules found by SCORPIUS was reduced in character count, yet it makes the same claims. The main differences are that gene names and explanations of the biological functions are not discussed in detail, but are instead summarised in figure 5.4. The reasoning for these changes are indeed to adapt the text to the target audience.
For example, detailed explanation of what actin polymerisation is, was removed; because immunologist readers already know this information and data scientist readers are presented with sufficient information when told that particular genes recovered by SCORPIUS are related to a biological process that is highly relevant to dendritic cell development, that is, actin polymerisation.}

\exam{This chapter would offer information on how the developed algorithms
	are being used to discover new insights, so I’m unsure if it is beneficial if this
	chapter evolves more into a technical description of Scorpius as the former is
	missing from the thesis.}

\exam{\imp{Small details / typo's:} [\ldots]} \\
\answ{These issues have been fixed, unless noted below.}

\exam{\imp{It would help to list the title of the manuscript on the pre-print server}} \\
\answ{I fixed the DOI. This should be sufficient to find back the manuscript on the pre-print server.}

\subsection{Chapter 6}

\exam{\imp{The abstract seems to be missing a sentence.}}

\exam{\imp{A main challenge in network inference has always been choosing a suitable
		cut-off to determine if something is interacting or not.} There are examples of
		regions in the network which are accurate but come at a much later stage in
		the ranking, at which time many spurious edges have been added to other
		parts of the network. I can only assume that this problem becomes even more
		challenging if one wishes to create case-wise GRN.}

\exam{\imp{The added value of the last paragraph in 6.3. seems small}}

\exam{\imp{Why is Scorpius used to generate Figure 6.3? Is this not confusing, since I
		assume we are not working with single cell data?}}

\exam{\imp{I’m afraid to ask, but I don’t get the reference to the name bred.}}

\exam{\imp{I don’t understand how the approach in figure 6.5 is different from GENIE3 and
		how it results in a tensor instead of matrix/ranking of ‘shuffle’ variable
		importance scores.} I believe I’m missing important information on how
		SCENIC works to be able to understand the modifications. I really struggle
		with this chapter: is the method executed for each profile or for all profiles
		together? Do you not end up with a single GRN for each of the profiles and as
		such why would bred be a true case-wise NI method and the Scenic approach
		would not? It would be nice to discuss this more during the defense.}

\exam{\imp{The readability of Figure 6.1 could be improved by using a black font}}

\subsection{Chapter 7}
\exam{\imp{Small details / typo's:} [\ldots]} \\
\answ{These issues have been fixed.}

\subsection{Chapter 8}
\exam{\imp{Does this contribution not better fit in an overall discussion of the thesis?}}

\exam{\imp{Small details / typo's:} [\ldots]} \\
\answ{These issues have been fixed.}

\subsection{Chapter 9}
\exam[\tagimp]{\imp{As the candidate is not first author on this paper, I’m not sure if this paper
		should be included in the PhD book verbatim. Perhaps it should not be
		included or it could be merged with chapters 8-9-10 into a single discussion
		chapter. Maybe at the very least, the candidate should state a bit more
		explicitly why the content was included in the PhD book and what the personal
		contribution was.}}

\exam{\imp{Author contributions is not correct ( CS)}} \\
\answ{???}

\subsection{Chapter 10}
\exam{\imp{The chapter contains a concise recap and critical view on the contributions in
		literature. It however does not discuss Chapter 9 (see comment above).}}

\exam{\imp{The discussion could be improved by stating how the tools are allowing new
		biological insights. Only a small hint is provided in the last sentence of 10.4.}}

\end{document}

### 2019-11-20 ch0_frontmatter.tex
fix textit

@@
 
-Gedurende dit project heb ik vooral genoten van het werken in een textit{open-source} omgeving, en het samenwerken met nationale en internationale collega’s uit het veld om onopgeloste problemen aan te pakken.
+Gedurende dit project heb ik vooral genoten van het werken in een \textit{open-source} omgeving, en het samenwerken met nationale en internationale collega’s uit het veld om onopgeloste problemen aan te pakken.
 Door de broncode van excellente softwareprojecten te bestuderen, leerde ik mezelf de aanbevolen werkwijzen aan.

### 2019-11-20 ch1_introduction/content.tex
update
technical fixes
various typo fixes

@@
 The earliest approach for measuring the abundance of a particular molecule in \emph{single cells} is the microscope.
-Since its development by Coons et al. (1941), immunohistochemistry (IHC) has been instrumental in visualising proteins. \cite{coons_immunologicalpropertiesantibody_1941}. 
+Since its development by Coons et al. (1941), immunohistochemistry (IHC) has been instrumental in visualising proteins \cite{coons_immunologicalpropertiesantibody_1941}. 
 A cell can present a particular type of protein, also called an antigen, on its cell surface.
@@
 By now, many protocols for extracting and tagging RNA from single cells have been developed \cite{svensson_exponentialscalingsinglecell_2018}, the most popular of which are based on microfluidics or droplets (Figure~\ref{fig:profiling_convergence}C).
-By sequencing the trancripts and the attached unique cell identifier tags, each read can be mapped and tallied up.
+By sequencing the transcripts and the attached unique cell identifier tags, each read can be mapped and tallied up.
 scRNA-seq data can thus be summarised in a matrix, where each column represent a single cell, each row a gene, and each value represents the number of transcripts that were sequenced for that gene and cell.
@@
 \subsection{Differential expression}
-Given that cells are split up into groups differential expression (DE) methods ranks genes based on whether their expression is significantly higher or lower in one group in comparison to the others.
+Given that cells are split up two or more groups, differential expression (DE) methods ranks genes based on whether their expression is significantly higher or lower in one group in comparison to the others.
 This grouping can be based on prior information or an upstream clustering method.
@@
 One of the central cellular processes underlying development is transcriptional regulation. 	
-Modeling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
+Modelling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
 
@@
 	\item We develop \textbf{a toolkit to infer, visualise and interpret single-cell trajectories} using more than 50 different TI methods (Chapter~\ref{chap:dyno}).
-dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g.
+This toolkit provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the trajectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g.
 adding directionality or adding annotation).
@@
 	\item We explore a novel algorithm for \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
-We demonstrate our NI methodology by inferring regulatory interactions for 22'122 hematopoietic cells, as well as for 14'963 bulk profiles of cancer tumours.
+We demonstrate our NI methodology by inferring regulatory interactions for 14'963 profiles of cancer patients.
 	\item We provide a tool for \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
 	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
-	Reproducability problems of TI methods are partially due to low rates of quantitative self-assessment.
+	Reproducibility problems of TI methods are partially due to low rates of quantitative self-assessment.
 We look at 75 manuscripts of TI methods and identify key challenges the authors had in benchmarking their tool.
@@
 \end{itemize}
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-11-20 ch3_dynbenchmark/content.tex
technical fixes
various fixes

@@
 
-A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}).
+A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table~\hyperref[tab:supptable_1]{3.1}).
 Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories.
@@
 \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation.
-Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}.
+Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}.
 			}
@@
 Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b).
-We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}), after which we conclude with guidelines for method users and future perspectives for method developers.
+We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}), after which we conclude with guidelines for method users and future perspectives for method developers.
 
@@
 The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance.
-These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}).
+These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table~\hyperref[tab:supptable_2]{3.2}).
 Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}.
@@
 We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a).
-As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Figure \ref{fig:supfigure_5}, and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}a).
+As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Figure \ref{fig:supfigure_5}, and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}a).
 
@@
 \begin{figure}[tbh!]
-	\centering\includegraphics[width=0.6\linewidth]{fig/supfigure_5.pdf}
+	\centering\includegraphics[width=0.6\linewidth]{fig/supfigure_5.png}
 	\caption{
@@
 
-We also assessed the memory requirements of each method (Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}c).
+We also assessed the memory requirements of each method (Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}c).
 Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features.
@@
 It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB).
-The reasons for not including certain tools are all discussed on our repository (https://github.com/dynverse/dynmethods/issues?q=label:unwrappable).
+The reasons for not including certain tools are all discussed on our repository (\url{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}).
 Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.
@@
 
-Based on the results of our benchmark, we propose a set of practical guidelines for method users (Figure \ref{fig:figure_5} and guidelines.dynverse.org).
+Based on the results of our benchmark, we propose a set of practical guidelines for method users (Figure \ref{fig:figure_5} and \url{https://guidelines.dynverse.org}).
 We postulate that, as a method’s performance is heavily dependent on the trajectory type being studied, the choice of method should currently be primarily driven by the anticipated trajectory topology in the data.
@@
 Several other factors will also impact the choice of methods, such as the dimensions of the dataset and the prior information that is available.
-These factors and several others can all be dynamically explored in our interactive app (guidelines.dynverse.org).
-This app can also be used to query the results of this evaluation, such as filtering the datasets or changing the importance of the evaluation metrics for the final ranking.
+These factors and several others can all be dynamically explored in our interactive application (\url{https://guidelines.dynverse.org}).
+This application can also be used to query the results of this evaluation, such as filtering the datasets or changing the importance of the evaluation metrics for the final ranking.
 
@@
 Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}).
-Using our framework, this figure can be recreated using only a couple of lines of R code (https://methods.dynverse.org).
+Using our framework, this figure can be recreated using only a couple of lines of R code (\url{https://methods.dynverse.org}).
 In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses.
@@
 
-We gathered a list of 71 trajectory inference tools (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
+We gathered a list of 71 trajectory inference tools (Supplementary Table~\hyperref[tab:supptable_1]{3.1}) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}.
 We welcome any contributions by creating an issue at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.
@@
 
-To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}).
+To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and Singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}).
 These containers are automatically built and tested using Travis continuous integration (\href{https://travis-ci.org/dynverse}{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity.
@@
 Depending on the output of a method, we used different strategies to convert the output to our model (Figure \ref{fig:supfigure_1}b).
-Special conversions are denoted by an asterisk and will be explained in more detail in the second list below.
+Special conversions are denoted by an asterisk (*) and will be explained in more detail in the second list below.
 
@@
 
-We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}).
+We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\hyperref[tab:supptable_2]{3.2}).
 The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}).
@@
 
+\section{Supplementary Figures and Tables}
+
+\paragraph{Supplementary Table 3.1: Overview of available trajectory inference tools, and whether they were included in	this study.} \label{tab:supptable_1}
+Download link: \url{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}.
+
+\paragraph{Supplementary Table 3.2: Overview of the real datasets used in this study.} \label{tab:supptable_2}
+Download link:
+\url{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}.
+
+\paragraph{Supplementary Figure 3.1: Results from the evaluation, for all methods and across all evaluation criteria.} \label{fig:suppfig_2}
+(a) We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies.
+The methods are grouped vertically based on the most complex trajectory type the y can infer.
+(b) The overall results of the evaluation on four criteria: benchmarking using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples, and quality  of  the  implementation.
+(c)  Accuracy  of  trajectory  inference  methods  across metrics,  dataset sources  and  dataset  trajectory types.
+The performance of a method is generally more stable across dataset sources, but very  variable depending on the metric and trajectory  type.
+(d)  Predicted  execution  times  and memory  usage  for  varying  numbers  of  cells  and  features  (\# cells $\times$ \# features).
+Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features.
+(e) Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method.
+(f) Usability scores of the tool and corresponding manuscript, grouped per category.
+Download link: \url{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}.
+
 \section{Supplementary Note 1: Metrics to compare two trajectories} \label{sec:dynb_supn1}
@@
 The \textit{isomorphic} score returns 1 if two graphs are isomorphic, and 0 if they were not.
-For this, we used the used the BLISS algorithm \cite{junttila_engineeringefficientcanonical_2007}, as implemented in the R *igraph* package.
+For this, we used the used the BLISS algorithm \cite{junttila_engineeringefficientcanonical_2007}, as implemented in the R igraph package.
 
@@
 		\textbf{An illustration of ranking features based on their importance in a trajectory.} 
-		(a) A MDS dimensionality reudction of a real dataset in which mouse embryonic fibroblasts (MEF) differentiate into Neurons and Myocytes.
+		(a) A MDS dimensionality reduction of a real dataset in which mouse embryonic fibroblasts (MEF) differentiate into Neurons and Myocytes.
 (b) The ranking of feature importances from high to low.

### 2019-11-20 ch7_incgraph/content.tex
fix typo

@@
 
-Graphlets~ \cite{przulj_modelinginteractomescalefree_2004} are one of the tools which could be used to add a topological prior to a biological network,
+Graphlets~ \cite{przulj_modelinginteractomescalefree_2004} are one of the tools which could be used to add a topological prior to a biological network.
 Graphlets are small connected subnetworks which can be counted to identify which low-level topological patterns are present in a network.

### 2019-11-20 boek.tex
update

@@
 \begin{quote}
-	Published in bioRxiv 079509. \doi{10.1038/s41587-019-0071-9}. \\
+	Published in bioRxiv 079509. \doi{10.1101/079509}. \\
 	Revised manuscript in preparation. \\
@@
 	$\bullet$ \textbf{R.C.} and W.S. benchmarks and analysed the data.\\
-	$\bullet$ S.T. and S.D. generated experimental data.\\
+	$\bullet$ S.T. and D.S. generated experimental data.\\
 	$\bullet$ \textbf{R.C.} implemented the SCORPIUS software package.\\
 	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
-	$\bullet$ \textbf{R.C.}, J.S., G.M, L.B., K.D.P., and Y.S. reviewed and edited the manuscript. \\
+	$\bullet$ \textbf{R.C.}, S.J., M.G., B.L., K.D.P., and Y.S. reviewed and edited the manuscript. \\
 	$\bullet$ K.D.P. and Y.S. supervised the project.
@@
 \end{quote}
-
+% TODO: fix abstract
 \vfill
@@
 \\
-	\textbf{Results:} IncGraph is calculates the differences in graphlet counts with respect to the network in its previous state.
+	\textbf{Results:} IncGraph calculates the differences in graphlet counts with respect to the network in its previous state.
 	IncGraph reduces the execution time by several orders of magnitude when compared to recounting the graphlet counts from scratch at every iteration.
@@
   Published in Genome Biology 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}. \\
-	Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix A, Saeys Y, and Robinson MD 
+	Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, and Robinson MD 
 \end{quote}

### 2019-11-20 rebuttal/rebuttal.tex
update

@@
 
-% used for defining abbreviations, see nomenclature.tex
-\usepackage{nomencl}
-
 % figure packages
-\usepackage{graphicx}
-\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
-\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
-\usepackage{float}
-\usepackage{import}
 \usepackage{xcolor}
-\usepackage{colortbl}
-\usepackage{tcolorbox}
-\usepackage[export]{adjustbox} % for top alignment
-\usepackage{afterpage} % for figure placement
-
-% symbols packages
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-% references packages
-\usepackage[sorting=none]{biblatex}
-\bibliography{library}
-\usepackage{bibentry}
-%\nobibliography*
 
@@
 }
-\newcommand{\exam}[2][\  ]{\hspace{0pt}\marginpar{\color{black}#1}$\triangleright$ \textit{\color{gray}#2}}
-%\newcommand{\exam}[2][\  ]{\begin{displayquote}\hspace{0pt}\marginpar{\color{black}#1}\textit{\color{gray}#2}\end{displayquote}}
-\newcommand{\imp}[1]{{\color{blue} #1}}
-\newcommand{\answ}[1]{{\color{red} Answer: #1}}
+
+\newcommand{\exam}[2][\  ]{\hspace{0pt}\marginpar{\color{red}#1}$\bullet$ \textit{#2}}
+\newcommand{\imp}[1]{{\color{red} #1}}
+\newcommand{\nimp}[1]{{\color{gray} #1}}
+\newcommand{\answ}[1]{{\color{blue} $\triangleright$ #1}}
+
+\usepackage{tikz}
+\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
+		\node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
 
@@
 \newcommand{\bigexclaim}{\raisebox{-0.1em}{\BigTriangleUp}\hspace{-0.32em}\llap{\small\textbf{!}}\hspace{0.32em}}
+\newcommand{\bigquest}{\circled{?}}
 
@@
 
+\section{Summary of most important comments, per chapter}
+
+\textbf{Overall:} The quality of writing, layout and graphical presentation is of a lower quality in the non-published parts of the thesis.
+Certain unpublished chapters feel rushed and unfinished.
+Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.
+
+\textbf{Overall:} Merge chapters 8-9-10 into one discussion.
+
+\textbf{Overall:} Merge chapter 8 into chapter 2.
+
+\textbf{Overall:} Random Forest variable importances are often used to detect differentially expressed genes. Why RF and not more standard statistical tests?
+
+\textbf{Introduction:} {Describe the practical implications of the
+		technological advances (e.g. precision medicine, new treatments) and
+		motivate the research from a non-technical view point (How is/can this PhD
+		change the world?)
+
+\textbf{dyngen:} {Table 2.2: explain default values and formulae}
+
+\textbf{dyngen:} {How crucial are certain parts of the GRN model, e.g. mRNA production and degradation?}
+
+\textbf{dyngen:} {Compare against other scRNA-seq generators}
+
+\textbf{dyngen:} {It is claimed dyngen is a multi-modality simulator, yet it produces only transcriptomics data}
+
+\textbf{dynbenchmark:} {Comparisons focus mainly on relative differences between methods. It would have been interesting to also assess the absolute performance of each method.}
+
+\textbf{dynbenchmark:} {No statistical tests were performed to assess whether methods are significantly different from one another}
+
+\textbf{dynbenchmark:} Turn into UK English to be consistent with the rest of the dissertation.
+
+\textbf{dyno:} {Does not really read as a scientific paper (yet).}
+
+\textbf{SCORPIUS:} {The methodology has changed between the bioRxiv preprint and this chapter, yet Figure 5.4 is exactly the same. Did the changes have no (measurable) effect on this analysis?}
+
+\textbf{SCORPIUS:} {Discuss the computational complexity of SCORPIUS.}
+
+\textbf{bred:} {Better explain how bred is different from GENIE3.}
+
+\textbf{bred:} {What makes bred a true case-wise NI method? Provide a more formal definition of what a case-wise regulatory network is.}
+
+\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles}
+
+\textbf{bred:} {Compare against other methods, e.g. GENIE3 + regulon specificity score}
+
+\textbf{bred:} {Discussion is very short, please elaborate}
+
+\textbf{IncGraph:} {On optimising GRNs, compare a simple MST with IncGraph}
+
+\textbf{Benchmarking guidelines:} {As the candidate is not first author on this paper, I’m not sure if this paper
+	should be included in the PhD book verbatim.}
+
+\textbf{Discussion:} {The Discussion chapter is missing a more broad context
+			description. How are other researchers using these tools to gain new biological insights? Where is the field of modelling single-cell dynamics going?}
+
+\textbf{Discussion:} {Discuss the links between different chapters}
+
+\newpage
+\section{Legend}
+\exam{Question or comment by jury member. \nimp{Description of dissertation.}}
+
+\exam[\tagimp]{\imp{Important matter} $\rightarrow$}
+
+\exam[\tagtime]{Matter which will take some time to reply to / fix $\rightarrow$}
+
+\answ{Response or statement by candidate.}
+
 \section{Joeri Ruyssinck}
+
+\answ{Small changes and typos have been fixed, unless noted otherwise.}
 
@@
 
-\exam{The research domain of this thesis is young and technical evolutions in the
+\exam{\nimp{The research domain of this thesis is young and technical evolutions in the
 	encompassing field move fast. This makes it especially challenging to do
@@
 	the greater high level challenges in the field, resulting in a potential seminal
-	benchmarking paper.}
+	benchmarking paper.}}
 
-\exam{The thesis as a whole contains many relevant contributions, of which several
-	have been published in high quality journals.}
+\exam{\nimp{The thesis as a whole contains many relevant contributions, of which several
+	have been published in high quality journals.}}
 
-\exam{The figures in the thesis are of exceptional quality. They are creative and able
+\exam{\nimp{The figures in the thesis are of exceptional quality. They are creative and able
 	to convey large volumes of information to the reader in a visually pleasing
-	format.}
+	format.}}
 
-\exam{The introduction chapter of the thesis demonstrates that the candidate
+\exam{\nimp{The introduction chapter of the thesis demonstrates that the candidate
 	understands and is able to convey the background of both (computational)
-	biology and computer science.}
+	biology and computer science.}}
 
-\exam{The passion of the candidate to bring modern good software development
+\exam{\nimp{The passion of the candidate to bring modern good software development
 	practices to the field is present throughout the work and might contribute to a
-	more mature algorithm development in computational biology.}
+	more mature algorithm development in computational biology.}}
 
@@
 
-\exam{The main contribution of this work is the excellent benchmarking study
-	published in Nature Biotechnology. \imp{However, this also effects the overall
+\exam{\nimp{The main contribution of this work is the excellent benchmarking study
+	published in Nature Biotechnology.} However, this also effects the overall
 	cohesiveness of the work, as information is repeated throughout the book at
-	several levels of detail.} This however a common side-effect of grouping
+	several levels of detail. \nimp{This however a common side-effect of grouping
 	published manuscripts into a thesis and I’m aware there is no (quick) solution
-	to this.}
+	to this.}}
 
-\exam[\tagimp]{It is very rare to see that research groups (or individual researchers) invest this
+\exam[\tagimp]{\nimp{It is very rare to see that research groups (or individual researchers) invest this
 	much effort and time in creating qualitative software. I agree with the ideas of
 	the candidates that these practices to adopt software to accompany research
-	ideas should be encouraged. \imp{Yet I wonder how the current academic system
+	ideas should be encouraged.} Yet I wonder how the current academic system
 	can support such concepts. Academics often hold different positions in
@@
 	be harder to overcome for PhD students which might be constrained by limited
-	support or funding?}}
+	support or funding?}
 
@@
 
-\exam[\tagimp \tagtime]{As an interdisciplinary thesis, it is very much enjoyable that the introduction
+\exam[\tagimp \tagtime]{\nimp{As an interdisciplinary thesis, it is very much enjoyable that the introduction
 	chapter contains a concise, yet precise overview of the concepts which are
@@
 	candidate only briefly touches upon the societal value of the technological
-	innovations he describes. For a PhD thesis, \imp{I think it would be an added value
+	innovations he describes. For a PhD thesis, } \imp{I think it would be an added value
 	if the candidate could briefly describe the practical implications of the
@@
 
-\exam{\imp{Small details / typo's:} [\ldots]} \\
-\answ{These issues have been fixed, unless noted below.}
-
-\exam{\imp{Figure 1.9 does not contain SCORPIUS (probably published before,
-	but still a bit weird for a PhD containing the manuscript)}} \\
+\exam{Figure 1.9 does not contain SCORPIUS (probably published before,
+	but still a bit weird for a PhD containing the manuscript)} \\
 \answ{Indeed, this figure was adapted from Cannoodt et al. 2016 and serves as an illustration that
@@
 
-\exam{\imp{Bold is only used in 1.1.2 and research challenges}} \\
+\exam{Bold is only used in 1.1.2 and research challenges} \\
 \answ{Usage of the bold font style in 1.1.2 and 1.3 serve clear, but different, purposes.
@@
 
-\exam{\imp{Use of we/I in summary}} \\
+\exam{Use of we/I in summary} \\
 \answ{Usage of first person pronouns in this dissertation is very limited.
@@
 \subsection{Chapter 2}
-\exam{\imp{It is not clear to me what is meant with ‘such as sampling a cell at a certain
+
+\exam{It is not clear to me what is meant with ‘such as sampling a cell at a certain
 	time point and once more at a later point’ and why this is currently not
-	possible?}}
+	possible?}
 
-\exam{\imp{Figure 2.1.A. What does the figure in ‘Combine simulations’ represent? Are
-		these single simulations which are mapped continuously to the backbone?}}
+\exam{Figure 2.1.A. What does the figure in ‘Combine simulations’ represent? Are
+		these single simulations which are mapped continuously to the backbone?}
 
-\exam{\imp{Figure 2.5. It is not explained what Subfigure F represents.} Only after the
+\exam{Figure 2.5. It is not explained what Subfigure F represents. Only after the
 	reader has completed the chapter, it becomes clear that the red lines probably
@@
 
-\exam{\imp{Figure 2.6: Consider changing the dashed lines}, as they are not (clearly)
+\exam{Figure 2.6: Consider changing the dashed lines, as they are not (clearly)
 		visible printed}
 
-\exam{\imp{Figure 2.6 and the example text for the cyclical example is unclear.}}
+\exam{Figure 2.6 and the example text for the cyclical example is unclear.}
 
-\exam{\imp{While I
+\exam{While I
 		understand that the instructions result in the desired state change, I also see
-		many other configurations which would result in a cyclical example.}}
+		many other configurations which would result in a cyclical example.}
 
-\exam{\imp{Why is it currently needed to explicitly keep certain module expressions constant to
+\exam{Why is it currently needed to explicitly keep certain module expressions constant to
 		determine the backbone? Would the same result not be achieved by simply
-		simulating X times and averaging expression levels?}}
+		simulating X times and averaging expression levels?}
 
-\exam{\imp{Figure 2.7. The figure legend could be extended}}
+\exam{Figure 2.7. The figure legend could be extended}
 
-\exam{\imp{The introduction of FANTOM5 is abrupt and it would be better if the context
-		would be briefly explained to the reader.}}
+\exam{The introduction of FANTOM5 is abrupt and it would be better if the context
+		would be briefly explained to the reader.}
 
-\exam{\imp{The paragraph describing Step 3 is unclear: -> Target genes ARE? Regulated
-		... but is-> are ?}}
+\exam{The paragraph describing Step 3 is unclear: -> Target genes ARE? Regulated
+		... but is-> are ?}
 
-\exam{\imp{Figure 2.8 can incorrectly suggest that target genes can only have a single
-		regulator assigned}}
+\exam{Figure 2.8 can incorrectly suggest that target genes can only have a single
+		regulator assigned}
 
-\exam{\imp{Table 2.2. contains many default ‘magic’ numbers?} It would be nice to know
+\exam[\tagimp \tagtime]{\imp{Table 2.2. contains many default ‘magic’ numbers?} It would be nice to know
 		how in practice you ended up selecting them and which problems you
@@
 
-\exam{\imp{Since these are default values, it would suggest
+\exam{Since these are default values, it would suggest
 		that the user can modify them? Why would a user do that and what would be
-		the impact?}}
+		the impact?}
 
-\exam{\imp{Am I correct that currently splicing offers little extra value as the formula’s
-		would imply that alternative splicing is not supported?}}
+\exam{Am I correct that currently splicing offers little extra value as the formula’s
+		would imply that alternative splicing is not supported?}
 
@@
 
-\exam{\imp{2.4.6. is a bit worrying. Could I reproduce dyngen without knowing these
-		details? Does the sampling process not deserve more explanation?}}
-
-\exam{\imp{Small details / typo's:} [\ldots]} \\
-\answ{These issues have been fixed.}
+\exam{2.4.6. is a bit worrying. Could I reproduce dyngen without knowing these
+		details? Does the sampling process not deserve more explanation?}
 
@@
 
-\exam{\imp{You mention consensus predictions, is this something you investigated more
-		in depth?} Would ‘ensemble’ predictions result in better performance? Is there
+\exam{You mention consensus predictions, is this something you investigated more
+		in depth? Would ‘ensemble’ predictions result in better performance? Is there
 		a good way to aggregate the predictions?}
 
-\exam{Monocle DDRTree seems to overestimate the topology based on Figure 3.5, \imp{is
+\exam{Monocle DDRTree seems to overestimate the topology based on Figure 3.5, is
 	this somehow related to the way you have to force the algorithm into the
-	common format?}}
+	common format?}
 
-\exam{\imp{The dyngen described here seems to be less advanced than the one from
-		Chapter 2?}}
+\exam{The dyngen described here seems to be less advanced than the one from
+		Chapter 2?}
 
-\exam{\imp{Small details / typo's:} [\ldots]} \\
-\answ{These issues have been fixed, unless noted below.}
-
-\exam{\imp{Consistency: this chapter is US English}}
+\exam[\tagtime]{Consistency: this chapter is US English}
 
@@
 
-\exam{Figure 4.7: \imp{The figure is lacking what the colors represent.}}
+\exam{Figure 4.7: The figure is lacking what the colors represent.}
 
@@
 	reduction; there is still a lot of variation between the methods on a very simple
-	example. \imp{Could you clarify
-	how often these methods disagree on trivial examples?}}
+	example. Could you clarify
+	how often these methods disagree on trivial examples?}
 
 \exam{Dyno offers a lot of
-	guidance with respect to how to choose your methods, run them, etc. \imp{Did you
+	guidance with respect to how to choose your methods, run them, etc. Did you
 	consider support for features such as interpretation or automated error
-	detection after prediction?}}
+	detection after prediction?}
 
@@
 
-\exam{\imp{Small details / typo's:} [\ldots]} \\
-\answ{These issues have been fixed, unless noted below.}
-
-\exam{\imp{Figure 4.7 is missing slingshot and paga label in part B}} \\
+\exam{Figure 4.7 is missing slingshot and paga label in part B} \\
 \answ{Figure 4.7 does contain labels for Slingshot and PAGA. A small gap between part A and part B has been added, to improve interpretability.}
@@
 
-\exam[\tagimp]{\imp{Comparing the preprint with the chapter in the thesis. Some minor changes
+\exam[\tagimp]{Comparing the preprint with the chapter in the thesis. Some minor changes
 	seem to have been made to SCORPIUS: e.g. I believe in the pre-print
@@
 	MDS is used in the chapter and final version. There is also no mention of
-	outlier removal. The main result Figure 5.4 is however identical. Did this
+	outlier removal. The main result Figure 5.4 is however identical. \imp{Did this
 	change or other potential changes have no (measurable) effect on the initial
 	analysis? I also noticed that the biological conclusions or hypothesis have
-	been reduced in claim strength in the paper or have been removed. It would
+	been reduced in claim strength in the paper or have been removed.} It would
 	be interesting to know if this is due to new insights that have been found in the
 	period 2016-> 2019 or mainly because the paper’s target audience has
-	shifted.}} \\
+	shifted.} \\
+
 \answ{As mentioned, the preprint was originally published in 2016.
@@
 	missing from the thesis.}
+\answ{???}
 
-\exam{\imp{Small details / typo's:} [\ldots]} \\
-\answ{These issues have been fixed, unless noted below.}
-
-\exam{\imp{It would help to list the title of the manuscript on the pre-print server}} \\
+\exam{It would help to list the title of the manuscript on the pre-print server} \\
 \answ{I fixed the DOI. This should be sufficient to find back the manuscript on the pre-print server.}
@@
 
-\exam{\imp{The abstract seems to be missing a sentence.}}
+\exam{The abstract seems to be missing a sentence.}
 
@@
 
-\exam{\imp{The added value of the last paragraph in 6.3. seems small}}
+\exam{The added value of the last paragraph in 6.3. seems small}
 
-\exam{\imp{Why is Scorpius used to generate Figure 6.3? Is this not confusing, since I
-		assume we are not working with single cell data?}}
+\exam{Why is Scorpius used to generate Figure 6.3? Is this not confusing, since I
+		assume we are not working with single cell data?}
 
-\exam{\imp{I’m afraid to ask, but I don’t get the reference to the name bred.}}
+\exam{I’m afraid to ask, but I don’t get the reference to the name bred.}
 
@@
 
-\exam{\imp{The readability of Figure 6.1 could be improved by using a black font}}
-
-\subsection{Chapter 7}
-\exam{\imp{Small details / typo's:} [\ldots]} \\
-\answ{These issues have been fixed.}
+\exam{The readability of Figure 6.1 could be improved by using a black font}
 
 \subsection{Chapter 8}
-\exam{\imp{Does this contribution not better fit in an overall discussion of the thesis?}}
-
-\exam{\imp{Small details / typo's:} [\ldots]} \\
-\answ{These issues have been fixed.}
+\exam{Does this contribution not better fit in an overall discussion of the thesis?}
 
@@
 \exam[\tagimp]{\imp{As the candidate is not first author on this paper, I’m not sure if this paper
-		should be included in the PhD book verbatim. Perhaps it should not be
+		should be included in the PhD book verbatim.} Perhaps it should not be
 		included or it could be merged with chapters 8-9-10 into a single discussion
@@
 		explicitly why the content was included in the PhD book and what the personal
-		contribution was.}}
+		contribution was.}
 
-\exam{\imp{Author contributions is not correct ( CS)}} \\
+\exam{Author contributions is not correct (CS)} \\
 \answ{???}
@@
 \subsection{Chapter 10}
-\exam{\imp{The chapter contains a concise recap and critical view on the contributions in
-		literature. It however does not discuss Chapter 9 (see comment above).}}
+
+\exam{\nimp{The chapter contains a concise recap and critical view on the contributions in
+		literature.} It however does not discuss Chapter 9 (see comment above).}
 
 \exam{\imp{The discussion could be improved by stating how the tools are allowing new
-		biological insights. Only a small hint is provided in the last sentence of 10.4.}}
+		biological insights.} Only a small hint is provided in the last sentence of 10.4.}
+
+\section{Pierre Geurts}
+
+\subsection{Remarks alongside summaries}
+\exam{Chapter 1 does not go into the technical details of the computational tools, but it gives a gentle and nicely written introduction to the field.}
+
+\exam{The results
+	section of Chapter 2 is however limited. I would have appreciated to see more experiments to illustrate the
+	quality of the generated data and also to have a comparison with other simulators from the
+	literature. But to be fair, it is not clear how data simulators could be compared to each other or
+	even how the quality of a given simulator could be quantitatively evaluated.}
+
+\exam{Chapter 4 is a description of the features offered by the toolkit, more than a technical presentation
+	or an evaluation of the tools, but more details about every step can be found in the other papers.}
+
+\exam{Regarding Chapter 6: Further experiments
+	would be needed to really evaluate the performance of the method, notably on artificial datasets
+	where the true case-wise networks would be known, but also on real single-cell omics dataset,
+	which seemed to be the original objective of the authors.}
+
+\subsection{Questions / remarks}
+
+\exam[\tagimp]{The focus of the comparison in Paper 2 is on ranking the different methods. This is extremely
+	useful obviously. \imp{I think it would have been interesting to take this opportunity to also assess
+	the methods on a more absolute scale.} I don’t think the paper provides answer to such
+	questions for example: Is a given method good enough, in general or for a given type of
+	trajectory? Is the problem of inferring a given type of trajectory solved in a satisfactory way?
+	On which kind of trajectories should the research community focus? What are the cases that
+	no method can tackle well?}
+
+\exam{Similarly, in Paper 4, the SCORPIUS method is only assessed relatively to other methods.
+	There is no doubt that SCORPIUS is a very good method but I’m missing an absolute
+	assessment on how well it performs. It would have been interesting to report the correlation
+	between the predicted and true pseudo-time values, in addition to the more general score of
+	Paper 2 and to show datasets where it performs poorly, if any.}
+
+\exam[\tagimp]{\imp{In Paper 2, I was surprised that no statistical test was performed to assess whether methods
+	are statistically different from one another} (or better than random), as this is commonly done
+	in machine learning for example. In addition to method ranking, it would have been
+	interesting to be able to highlight groups of methods that are not statistically different from
+	each other in terms of accuracy.}
+
+\exam[\tagimp]{In several papers, the authors use Random Forests variable importances to find differentially
+	expressed genes (along a trajectory, between branches, at some branching point, etc). \imp{I’m
+	wondering why they use RF instead of more standard statistical tests for finding these genes.}
+	RF are useful when some multivariate effects need to be found but I’m not sure it is the case
+	in all situations where RF is used in the thesis. In Paper 2 for example, it is said that mtry is
+	set to 1\% to make sure that highly correlated features are not suppressed in the ranking. In
+	Paper 4, Figure 5.5, there is a mention that “shallow” random forests are used, which suggest
+	that they don’t want to find complex interaction effects. What is the expected benefit of
+	using RF instead of statistical tests?}
+
+\exam[\tagimp]{In Paper 4, \imp{I’m missing a discussion of the computational complexity of the SCORPIUS
+	method.} In Section 5.2, the authors say that MDS and principal curves have been adapted so
+	that they scale linearly but given that the approximated principal curves algorithm is
+	initialized by k-means, isn’t the algorithm quadratic? This is confirmed by Figure 3.7 in
+	Paper 2 that shows that SCORPIUS scales quadratically with the number of cells.}
+
+\exam[\tagimp]{In Paper 5, it is argued that most case-wise network inference methods consist in post-
+	processing a static GRN, unlike the bred method which is a true case-wise NI method. My
+	feeling is that bred is not that different in spirit from existing post-processing methods, as it
+	starts from a model trained on the full dataset of samples, which is trying to find regulators
+	that are relevant for most of the samples. \imp{What makes bred a true case-wise NI method?}
+	More fundamentally, \imp{I’m missing a more formal definition of what a case-wise regulatory
+	network is.} A network can obviously not be derived from a single sample (as interactions
+	between genes cannot be retrieved from a single static expression vector).}
+
+\exam[\tagimp \tagtime]{As mentioned in the conclusion of Paper 5, the bred method needs to be assessed on more
+	problems and against ground truth networks. One baseline experiment that I would have
+	been interest in at this stage is \imp{a comparison of the clustering in Figure 6.1 obtained from
+	the regulomes with a clustering obtained from the original expression profiles.} Does using
+	the regulomes really make a difference in terms of the final clustering?}
+
+\exam[\tagtime]{In Paper 7, the improvement obtained with IncGraph on the two gene networks is interesting.
+	However, \imp{I would have liked to see a comparison with a very simple heuristic that goes
+	through the ranked list of edges and only adds a new edge if it is not redundant with a
+	previously added edge} (as in minimum spanning tree construction algorithms).}
+
+\subsection{Suggested modifications}
+\answ{Small suggestions have been accepted, unless noted otherwise.}
+
+\exam{During my first read of the thesis, I had difficulty understanding what the author meant by
+	“milestone” and by “region of delayed commitment”. Maybe introduce a definition of these
+	notions somewhere in Chapter 1, if it fits.}
+
+\exam{I don’t understand footnote 2 on page 108 explaining why the method in Chapter 4 is called
+	‘bred’. Maybe clarify this (sorry if I miss something obvious).}
+
+\exam{The title of Chapter 6 is misleading as the method is actually not applied to single-cell data.
+	I would suggest to maybe change it into “inferring case-wise regulatory networks”.}
+
+\exam{In Paper 5, Section 6.4.1: how is the minimum threshold on average importance set? This is
+	important as this threshold conditions the number of interactions found.}
+
+\exam{Page 5, the sentence “IHC realizes the visualization...to specific antigens” seems to be
+	redundant with the next one.}
+
+\section{Stein Aerts}
+
+\exam{The introduction is relatively brief, but well
+written.}
+
+\exam{The methods of Chapter 2 are extensive, but the results are very brief. I’m wondering whether
+some parts that are now in the Methods section would not fit better in the Results section. Testing how
+representative the generated data are is of course a huge challenge, this could have been addressed
+perhaps in the Discussion.}
+
+\exam[\tagimp \tagtime]{\imp{It would also be interesting te evaluate the impact of the gene regulatory
+models in the simulation. How crucial is it to use such an extensive model of mRNA production and
+degradation, and how do these parameters affect the simulated single-cell data?}}
+
+\exam{Chapter 4 is a bit opposite
+of chapter 2 - here the results are described, but without methods. For example, how are features
+(genes) selected on the branch points, in a branch, etc (which methods are used for this).}
+
+\exam{The marker gene idenfitication is very interesting, but is not described in detail in the Methods, how
+is this RF built, and trained? Is it RF regression or classification? What is the topology, why use a RF,
+has it been compared to other techniques?}
+
+\exam{In the benchmark of SCORPIUS compared to other
+methods, it is mentioned that many data sets are used, but there are no details on which data sets.}
+
+\exam[\tagimp \tagtime]{\imp{In Chapter 6, it would have been interesting
+to compare the results of bred with such methods (for example GENIE3 + regulon specificity score).}}
+
+\exam[\tagimp \tagtime]{\imp{The methods could be a little bit more clear on the specific differences between bred and GENIE3.}}
+
+\exam{In Chapter 7, it may have
+been interesting to discuss the relationship with the classical network motifs from Uri Alon.}
+
+\exam{Chapter 9 could have
+contained a bit more in-depth discussion about trajectory inference in general, biology, and discussion
+of the results in the context of other work.}
+
+\section{Vanessa Vermeirssen}
+
+\subsection{Chapter 1}
+
+\exam{\nimp{The introduction is concise and to the point. This is also the case for the introduction sections of the
+		different paper chapters. Therefore,} some sections could have been more elaborated on e.g. briefly
+		describe and include the differences between the different dimensionality reduction methods and
+		define different network inference approaches. \nimp{At the end of the introduction, the objectives are
+		clearly put into the research context.}}
+
+\subsection{Chapter 2}
+
+\exam[\tagimp \tagtime]{In the introduction the PhD candidate mentions \imp{other scRNA-seq profile generators i.e. the
+state of the art, but in this chapter no comparison is made with those in order to analyze how
+well dyngen performs on the generation of scRNA-seq profiles.} In the introduction I also lack
+a short description of the workflows/underlying algorithms of those scRNA-seq profile
+generators as opposed to that of dyngen. Can you also comment on how dyngen differs from
+simulators of gene regulatory networks such as GeneNetWeaver and Neural Gene Network
+Constructor?}
+
+\exam[\tagimp]{Dyngen is put forward as a multi-modality simulator as opposed to other scRNA-seq profile
+generators. \imp{This multi-modality is mostly reflected in the results as different simulations of
+experiments and extracted data, all based on transcriptomics, whereas in the introduction
+this multi-modality is defined as the profiling of another data type than scRNA-seq.} Please
+clarify this in this chapter.}
+
+\exam{p. 26 The backbone is defined as a network of cellular states. The connection between gene
+modules and those cellular states could be made more explicitly clear. Do I understand
+correctly that in a specific cellular state specific gene modules of the module network
+become/are active?}
+
+\exam{A legend is missing for Fig. 2.7}
+
+\exam{p. 27 Step 1: “from the module IT belongs to”.} \\
+\answ{This has been fixed.}
+
+\exam{Write more clearly in step 2: “These
+interactions...” Specify.}
+
+\exam{And step 3:”Target genes...” Verb missing in sentence.} \\
+\answ{???}
+
+\subsection{Chapter 3}
+
+\exam{p.37 Last part missing in sentence? “The overall score between the different dataset sources
+	was moderately to highly correlated ...”} \\
+\answ{???}
+
+\exam{dyngen is somewhat differently explained here in the Methods section. Please integrate with
+	paper 1 chapter and remove redundancy.}
+
+\subsection{Chapter 4}
+
+\exam{It is recommended to clarify and extend the discussion on future improvements that could be
+	added to the tool, e.g. alignment and differential expression. What is the biological
+	significance? What is meant by feature importance tools/score?}
+
+\subsection{Chapter 6}
+
+\exam{Is this really the first algorithm for inferring case-wise regulatory networks?}
+
+\exam[\tagimp]{p. 107 Delete in Results: We generate.}
+
+\exam{\nimp{p. 108 “While accuracy of the network of predicted regulatory interactions is lower in
+	comparison to experimental validation techniques, network inference methods offer an
+	unbiased and high-throughput insight into the regulatory dynamics of a biological system.”} I
+	do not agree. Network inference methods do have a bias, depending on the underlying
+	algorithm (cfr Marbach et al. 2012, ...), but so do specific experimental techniques that
+	measure only an aspect of gene regulation. It is true that precision-recall for network
+	inference predictions is not high, so they can only be viewed as hypothesis generators, but
+	are nevertheless valid since they offer, relatively easy, a genome-wide view on gene
+	regulation.}
+
+\exam{There are significant differences between the methods of Kuijer and Liu, please describe
+	these methods in more detail in the introduction.}
+
+\exam{What about case-specific random forests like in the R package ranger, csfr() ?}
+
+\exam[\tagimp]{The sentence “As such, these methods will most likely recover the interactions that are
+	prevalent in the whole population, and will miss interactions that are specific to only a sub-
+	population” is incorrect. \imp{I would suggest you read the relevant method papers again, see
+	above.}}
+
+\exam{p. 114, 6.4.3. Can you include a motivation for the specific clustering algorithms used here?}
+
+\exam[\tagimp \tagtime]{\imp{The discussion is very short, please elaborate on}: possible applications in the field of
+	personalized precision medicine and comparison to state of the art.}
+
+\subsection{Chapter 7}
+
+\exam{Please define graphlet and orbit more precisely in the introduction and differentiate between
+	the concepts network motif, subgraph and graphlet.}
+
+\exam{Can you also brieflly introduce the non-incremental approach Orca?}
+
+\subsection{Chapter 8}
+
+\exam{A large part of this chapter is already incorporated in previous chapters, especially in paper2.
+	I would suggest to integrate both and avoid redundancy. A large part of this chapter would fit
+	in the discussion of paper2.}
+
+\subsection{Chapter 10}
+\exam[\tagimp \tagtime]{\imp{The general discussion on the thesis is only 3 pages and consists of small, but critical, discussion
+		paragraphs on each of the different papers. The general discussion is too brief and lacks future perspectives on the field. It provides some links between the different chapters,
+		but this bridging could be increased further.
+		I miss an overarching discussion on where the field of modelling single cell dynamics is going.}}
 

### 2019-11-20 ch10_discussion/content.tex
update

--- 2019-10-21 ch8_discussion/content.tex
@@
 We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
-The \texttt{dyngen} software supports a selection of existing experimental protocols, and can also simulate experiments that are possible with the current technologies, such as generating omics profiles from the same cell at multiple timepoints.
+The \texttt{dyngen} software supports a selection of existing experimental protocols, and can also simulate experiments that are not possible with the current technologies, such as generating omics profiles from the same cell at multiple timepoints.
 The generated data can help kick-start emerging domains with low data availability more safely by allowing software developers to test their method before numerous datasets become publicly available.
@@
 Keeping track of the frequencies of graphlets during an iterative process proved difficult at first.
-\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by making assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals)).
+\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by  assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals)).
 Under this assumption, it is possible to calculate the differences in topological patterns significantly faster in comparison to recomputing the graphlet frequencies from scratch at every iteration.

### 2019-11-20 ch1_introduction/content.tex
update

@@
 
-Specific examples of omics technologies are whole genome sequencing to determine the DNA sequence of an organism, and RNA sequencing to profiles the sequence of RNA transcripts, both using next-generation sequencing technologies.
+Specific examples of omics technologies are whole genome sequencing to determine the DNA sequence of an organism, and RNA sequencing to profile the sequence of RNA transcripts, both using next-generation sequencing technologies.
 A gene expression profile can be obtained by mapping the sequences of RNA transcripts to the genome.
@@
 By sequencing the transcripts and the attached unique cell identifier tags, each read can be mapped and tallied up.
-scRNA-seq data can thus be summarised in a matrix, where each column represent a single cell, each row a gene, and each value represents the number of transcripts that were sequenced for that gene and cell.
+scRNA-seq data can thus be summarised in a matrix, where each column represents a single cell, each row a gene, and each value represents the number of transcripts that were sequenced for that gene and cell.
 
@@
 Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful low-dimensional representation.
-DR methods have two main target audiences; computers -- to construct a K-dimensional (with K $\ll$ M) representation of the data such that pairwise distances between different samples are retained as well as possible (Figure~\ref{fig:comp_tools_dimred}B); and humans -- to obtain a visual overview of the data (Figure~\ref{fig:comp_tools_dimred}C).
+DR methods have two main target audiences; computers -- to construct a K-dimensional (with K $\ll$ N) representation of the data such that pairwise distances between different samples are retained as well as possible (Figure~\ref{fig:comp_tools_dimred}B); and humans -- to obtain a visual overview of the data (Figure~\ref{fig:comp_tools_dimred}C).
 
@@
 \subsection{Differential expression}
-Given that cells are split up two or more groups, differential expression (DE) methods ranks genes based on whether their expression is significantly higher or lower in one group in comparison to the others.
+Given that cells are split up two or more groups, differential expression (DE) methods rank genes based on whether their expression is significantly higher or lower in one group in comparison to the others.
 This grouping can be based on prior information or an upstream clustering method.

### 2019-11-20 ch2_dyngen/content.tex
update

@@
 A cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A). 
-A gene regulatory network (GRN) define the reactions that are allowed to occur (Figure \ref{fig:simplecyclic}B), and is constructed such that a cell will develop over time (Figure \ref{fig:simplecyclic}C--D).
+A gene regulatory network (GRN) defines the reactions that are allowed to occur (Figure \ref{fig:simplecyclic}B), and is constructed such that a cell will develop over time (Figure \ref{fig:simplecyclic}C--D).
 The likelihood of a reaction occurring is a function of the abundance of key molecules involved in each reaction (Figure \ref{fig:simplecyclic}E).
@@
 Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$).
-A TF inherits the 'burn' and 'basal expression' from the module in belongs to.
+A TF inherits the 'burn' and 'basal expression' from the module it belongs to.
 
@@
 A user-defined number of target genes are added to the GRN.
-Target genes regulated by a TF or another target gene, but is always downstream of at least one TF.
+Target genes are regulated by a TF or another target gene, but is always downstream of at least one TF.
 To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.

### 2019-11-20 ch6_bred/content.tex
update

@@
 Computing the case-wise feature importances of a tree consists of the following 8 steps (Figure~\ref{fig:fimp}).
-The 'randomness' of a Random Forest is due to only using a subset of the samples in the dataset in order to build a single decision tree.
+The 'randomness' of a Random Forest is in part due to using a subset of the samples in the dataset in order to build a single decision tree.
 The samples are split into two groups, the 'in-bag' data and the 'out-of-bag' data (Figure~\ref{fig:fimp}A).
@@
 A moderately-sized dataset could contain $M=10'000$ samples, $N=2'000$ regulators, and $P=10'000$ target genes.
-Due to memory constrains, only interactions with an average importance value (across all samples) higher than a minimum threshold are retained.
+Due to memory constraints, only interactions with an average importance value (across all samples) higher than a minimum threshold are retained.
 
@@
 Next, KD-trees are used to calculate the $k$NN graph efficiently.
-The cases in the dataset are visualised and clusted using the Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} and Louvain clustering \cite{blondel_fastunfoldingcommunities_2008}, respectively.
+The cases in the dataset are visualised and clustered using the Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} and Louvain clustering \cite{blondel_fastunfoldingcommunities_2008}, respectively.
 

### 2019-11-20 rebuttal/rebuttal.tex
~

@@
 
-\section{Summary of most important comments, per chapter}
+\section{Summary of most important comments}
 
@@
 
-\textbf{Overall:} Merge chapters 8-9-10 into one discussion.
-
-\textbf{Overall:} Merge chapter 8 into chapter 2.
+\textbf{Overall:} From three jury members: Merge chapters 8-9-10 into one discussion, or merge chapter 8 into chapter 3.
 
@@
 
+\textbf{dyngen:} Chapter does not seem completely finished; may need a working example to demonstrate the steps a user needs to take to use the software, and to illustrate how the results look like.
+No references to user documentation, source code (open source?).
+No discussion about the performance of the framework.
+Since this is a PhD in
+computer science, I might also have expected a class diagram on the software, as a proof that the framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter.
+
 \textbf{dynbenchmark:} {Comparisons focus mainly on relative differences between methods. It would have been interesting to also assess the absolute performance of each method.}
@@
 
+\textbf{dyno:} Chapter does not seem completely finished; especially the results section (4.2) read more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study.
+
 \textbf{SCORPIUS:} {The methodology has changed between the bioRxiv preprint and this chapter, yet Figure 5.4 is exactly the same. Did the changes have no (measurable) effect on this analysis?}
@@
 \textbf{SCORPIUS:} {Discuss the computational complexity of SCORPIUS.}
+
+\textbf{SCORPIUS:} How objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?
 
@@
 
-\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles}
+\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles. Compare against other methods, e.g. GENIE3 + r	egulon specificity score.}
 
-\textbf{bred:} {Compare against other methods, e.g. GENIE3 + regulon specificity score}
-
-\textbf{bred:} {Discussion is very short, please elaborate}
+\textbf{bred:} {Discussion is very short, please elaborate. Paper needs finalisation to make it publication-ready.
 
@@
 
-\textbf{Discussion:} {Discuss the links between different chapters}
+\textbf{Discussion:} {Reads more like a summary. Discuss the links between different chapters.}
 
@@
 
+\section{Peter Dawyndt}
+
+\exam{Not all chapters have been published and some are still in stage where some work needs to be done before they are publication-ready}
+
+\exam{Not always a logical connection between the successive chapters}
+
+% TODO: provide visual schematic of what topic this dissertation tries to tackle
+
+\exam{Drop the paper X prefix from most of the chapters}
+
+\subsection{Chapter 2}
+
+\exam[\tagimp \tagtime]{chapter does not seem completely finished; \imp{may need a working example to demonstrate the steps
+	a users needs to take to use the software, and to illustrate how the results look like}; for now, only
+	the inner workings of the simulator are discussed}
+
+\exam{no information about implementation language
+	(since Gillespie SSA uses a C++ library, this may hint the framework is also implemented in C++)}
+
+\exam{case
+	study may also serve as user documentation}
+
+\exam{no references to user documentation, source code
+	(open source?)}
+
+\exam{no discussion about the performance of the framework}
+
+\exam[\tagtime]{since this is a PhD in
+	computer science, I might also have expected a class diagram on the software, as a proof that the
+	framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter}
+
+\exam{p24-25 how are networks specified by user (ergonomy \& standards); support for SBML?}
+
+\exam{p26: has the BBL-ontology been invented for this application or an existing ontology?}
+
+\exam{maybe this chapter should come after the next chapter}
+
+\subsection{Chapter 3}
+
+\exam{what reactions have the authors received from the developers of the tools that were reviewed in
+	this benchmark, or from users of the TI tools that have used the benchmark results to decide what
+	tools to use}
+
+\exam{could this study also yield some standard ways for data formats that tools could use to input/output
+	data (already hinted in the text); what approach would the authors choose towards a standards
+	proposal?}
+
+\subsection{Chapter 4}
+
+\exam{Some users filed an issue on GitHub reporting problems with running Docker on Windows; a
+	promise was made by the authors that this issue would be looked at, but no solution has been
+	provided so far; any progress in resolving this issue?}
+
+\exam{as with CH2, this chapter does not seem completely finished; especially the results section (4.2) read more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study, e.g.
+how to run the analysis one the command line, might benefit the readership}
+
+\subsection{Chapter 5}
+
+\exam[\tagimp]{fundamental question: \imp{how objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?}}
+
+\subsection{Chapter 6}
+
+\exam{paper needs finalisation to make it publication-ready}
+
+\subsection{Chapter 7}
+
+\exam{performance measurements show that method is faster than state of the art, but no formal time complexity is derived and no formal proof is given that the method still computes the correct number of graphlets; strange behaviour in time measurements is not diagnosed in depth}
+
+\subsection{Chapter 8}
+
+\exam{I would argue against having this as a separate chapter, but would use its content to supplement the introduction and CH3}
+
+\subsection{Discussion}
+
+\exam{points of discussion, open source: upon publication or release-early-release-often?}
+
+\exam[\tagimp]{\imp{reads more like a summary} + listing some use case that made use of contributions made in the manuscript, rather than a real discussion}
+
+\exam{10.7 might better go into a preface of the manuscript}
+
 \section{Pierre Geurts}

### 2019-11-21 rebuttal/rebuttal.tex
add back dyngen submodule, apply small changes

@@
 
-\textbf{Overall:} The quality of writing, layout and graphical presentation is of a lower quality in the non-published parts of the thesis.
-Certain unpublished chapters feel rushed and unfinished.
-Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.
-
 \textbf{Overall:} From three jury members: Merge chapters 8-9-10 into one discussion, or merge chapter 8 into chapter 3.
@@
 
-\textbf{dyngen:} Chapter does not seem completely finished; may need a working example to demonstrate the steps a user needs to take to use the software, and to illustrate how the results look like.
+\textbf{dyngen:} May need a working example to demonstrate the steps a user needs to take to use the software, and to illustrate how the results look like.
 No references to user documentation, source code (open source?).
 No discussion about the performance of the framework.
-Since this is a PhD in
-computer science, I might also have expected a class diagram on the software, as a proof that the framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter.
+I might also have expected a class diagram on the software, as a proof that the framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter.
 
@@
 
-\textbf{dyno:} {Does not really read as a scientific paper (yet).}
-
-\textbf{dyno:} Chapter does not seem completely finished; especially the results section (4.2) read more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study.
+\textbf{dyno:} Does not really read as a scientific paper (yet).
+The results section should be improved, it currently reads more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study.
 
@@
 
-\textbf{bred:} {Discussion is very short, please elaborate. Paper needs finalisation to make it publication-ready.
+\textbf{bred:} {Discussion is very short, please elaborate.}
 

### 2019-11-21 ch2_dyngen/analysis/example_methods_overview.R

@@ -1,377 +1,0 @@
-library(tidyverse)
-library(dyngen)
-
-set.seed(4)
-
-dir.create("fig/example_methods", showWarnings = FALSE, recursive = TRUE)
-
-backbone <- backbone(
-  module_info = tribble(
-    ~module_id, ~ba, ~burn,
-    "A", 1, TRUE,
-    "B", 0, FALSE,
-    "C", 0, FALSE,
-    "D", 0, FALSE,
-  ),
-  module_network = tribble(
-    ~from, ~to, ~effect, ~strength, ~cooperativity,
-    "A", "B", 1, 1, 2,
-    "B", "C", 1, 1, 2,
-    "B", "D", 1, 1, 2,
-    "C", "D", -1, 10, 2,
-    "D", "C", -1, 10, 2
-  ),
-  expression_patterns = tribble(
-    ~from, ~to, ~module_progression, ~start, ~burn, ~time,
-    "Burn0", "S1", "+A", TRUE, TRUE, .5,
-    "S1", "S2", "+B", FALSE, FALSE, .5,
-    "S2", "S3", "+C", FALSE, FALSE, 1,
-    "S2", "S4", "+D", FALSE, FALSE, 1,
-  )
-)
-backbone$module_info$color <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3")
-model <-
-  initialise_model(
-    num_tfs = 8,
-    num_targets = 10,
-    num_hks = 0,
-    num_cells = 100,
-    backbone = backbone,
-    verbose = TRUE,
-    download_cache_dir = "~/.cache/dyngen",
-    num_cores = 8,
-    distance_metric = "euclidean",
-    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
-    simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
-  ) %>%
-  generate_tf_network() %>%
-  generate_feature_network() %>%
-  generate_kinetics()
-
-## grn
-plot_backbone_statenet(model) %>% ggsave(filename = "fig/example_methods/statenet.pdf", width = 6, height = 6)
-plot_backbone_modulenet(model) %>% ggsave(filename = "fig/example_methods/modulenet.pdf", width = 8, height = 6)
-
-plot_feature_network(model, show_targets = FALSE)
-
-plot_feature_network(model)
-plot_feature_network(model, show_hks = TRUE)
-plot_feature_network(model)
[…]

### 2019-11-21 ch2_dyngen/analysis/simplecyclic.R

@@ -1,253 +1,0 @@
-library(tidyverse)
-library(dyngen)
-
-set.seed(1)
-
-backbone <- backbone(
-  module_info = tribble(
-    ~module_id, ~a0, ~burn,
-    "A", 1, TRUE,
-    "B", 1, TRUE,
-    "C", 1, TRUE,
-    "D", 0, TRUE,
-    "E", 0, TRUE
-  ),
-  module_network = tribble(
-    ~from, ~to, ~effect, ~strength, ~cooperativity,
-    "A", "B", -1, 4, 2,
-    "B", "C", -1, 4, 2,
-    "C", "D", 1, 1, 2,
-    "D", "E", 1, 1, 2,
-    "E", "A", -1, 4, 2
-  ),
-  expression_patterns = tribble(
-    ~from, ~to, ~module_progression, ~start, ~burn, ~time,
-    "Burn0", "S1", "+A,+B,+C", TRUE, TRUE, 3,
-    "S1", "S2", "+D,+E,-C", FALSE, FALSE, 4,
-    "S2", "S3", "+C,-A,-B", FALSE, FALSE, 4,
-    "S3", "S1", "+A,+B,-D,-E", FALSE, FALSE, 4
-  )
-)
-census_interval <- .1
-total_time <- 10
-model <-
-  initialise_model(
-    num_tfs = nrow(backbone$module_info) * 1,
-    num_targets = 0,
-    num_hks = 0,
-    num_cells = 300,
-    backbone = backbone,
-    verbose = TRUE,
-    download_cache_dir = "~/.cache/dyngen",
-    num_cores = 8,
-    kinetics_params = kinetics_default(
-      sample_wpr = function(n) 100,
-      sample_spl = function(n) 10,
-      sample_xdr = function(n) 5,
-      sample_ypr = function(n) 5,
-      sample_ydr = function(n) 5
-    ),
-    simulation_params = simulation_default(
-      burn_time = 2,
-      total_time = total_time,
-      ssa_algorithm = GillespieSSA2::ssa_exact(),
-      num_simulations = 1,
-      census_interval = census_interval,
-      store_reaction_firings = TRUE,
-      store_reaction_propensities = TRUE
-    )
-  )
-
-model <- model %>%
-  generate_tf_network() %>%
-  generate_feature_network()
-
-notf <- function(x) gsub("_TF1$", "", x)
-model$feature_info <- model$feature_info %>% mutate_at(vars(feature_id), notf)
[…]

### 2019-11-21 ch2_dyngen/analysis/usecase_trajectory_alignment.R

@@ -1,144 +1,0 @@
-library(tidyverse)
-library(dyngen)
-
-set.seed(1)
-
-out_dir <- "fig/usecase/trajectory_alignment/"
-dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
-
-# Generate base model -----------------------------------------------------
-backbone <- bblego(
-  bblego_start("A", type = "simple", num_modules = 4),
-  bblego_linear("A", "B", type = "doublerep1", num_modules = 6),
-  bblego_linear("B", "C", type = "doublerep2", num_modules = 6),
-  bblego_end("C", type = "simple", num_modules = 4)
-)
-model <-
-  initialise_model(
-    num_tfs = 100,
-    num_targets = 200,
-    num_hks = 200,
-    num_cells = 1000,
-    backbone = backbone,
-    verbose = TRUE,
-    download_cache_dir = "~/.cache/dyngen",
-    num_cores = 8,
-    distance_metric = "pearson",
-    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
-    simulation_params = simulation_default(num_simulations = 10, census_interval = .01)
-  ) %>%
-  generate_tf_network() %>%
-  generate_feature_network()
-
-
-# Show GRN ----------------------------------------------------------------
-plot_backbone_statenet(model)
-plot_backbone_modulenet(model)
-plot_feature_network(model, show_targets = FALSE)
-plot_feature_network(model)
-plot_feature_network(model, show_hks = TRUE)
-
-
-# Generate 2 variations with slightly randomised kinetics -----------------
-model1 <- model
-model1$feature_info <- model1$feature_info %>%
-  mutate_at(c("ba", "ind"), ~ . * rnorm(length(.), mean = 1, sd = .01)) %>%
-  mutate_at(c("ba", "ind"), ~ pmin(., 1))
-model1$feature_network <- model1$feature_network %>%
-  mutate_at(c("strength", "cooperativity"), ~ . * rnorm(length(.), mean = 1, sd = .01))
-model1 <- model1 %>%
-  generate_kinetics() %>%
-  generate_gold_standard() %>%
-  generate_cells()
-
-model2 <- model
-# by_backbone <- model2$feature_network$from %>% grepl("_TF", .)
[…]

### 2019-11-21 ch6_bred/scripts/2_tcga.R

@@ -41,8 +41,8 @@ if (!file.exists(tcga_data_file)) {
 
   # add clinical case id
   mjs <- c(
-    jsonlite::read_json("/home/rcannood/Workspace/tcga/raw/F_metadata.cart.2019-10-08.json"),
-    jsonlite::read_json("/home/rcannood/Workspace/tcga/raw/M_metadata.cart.2019-10-08.json")
+    jsonlite::read_json(paste0(tcga_folder, "raw/F_metadata.cart.2019-10-08.json")),
+    jsonlite::read_json(paste0(tcga_folder, "raw/M_metadata.cart.2019-10-08.json"))
   )
 
   meta <- pbapply::pblapply(mjs, function(mji) {
@@ -62,8 +62,8 @@ if (!file.exists(tcga_data_file)) {
 
   # add clinical data
   cjs <- c(
-    jsonlite::read_json("/home/rcannood/Workspace/tcga/raw/F_clinical/clinical.json"),
-    jsonlite::read_json("/home/rcannood/Workspace/tcga/raw/M_clinical/clinical.json")
+    jsonlite::read_json(paste0(tcga_folder, "raw/F_clinical/clinical.json")),
+    jsonlite::read_json(paste0(tcga_folder, "raw/M_clinical/clinical.json"))
   )
 
   clinical <- pbapply::pblapply(seq_along(cjs), cl = 8, function(i) {

### 2019-11-22 defence_internal/pres_v1.tex
~

\documentclass[usepdftitle=false, compress]{beamer}

\usepackage{calc}
\newlength{\leftstart}
\newlength{\topstart}
\setlength{\leftstart}{\hoffset + \oddsidemargin + 1in}
\setlength{\topstart}{\voffset + \topmargin + 1in}

\newlength{\figfour}

\usepackage{etex}
\usepackage{etoolbox}
\usepackage[absolute,overlay]{textpos}
\usepackage{fancybox}

\usepackage{xcolor}
\definecolor{myblue}{rgb}{0,0.4,0.7}
\definecolor{myred}{rgb}{0.7,.1,.1}
\definecolor{mygray}{rgb}{0.7,0.7,0.7}
\definecolor{mygreen}{rgb}{0,0.8,0.1}

\useinnertheme{circles}

\DeclareMathOperator*{\argmin}{arg\,min}

\mode<presentation>{
  \setbeamertemplate{navigation symbols}{}
	\usepackage{lmodern}
	\setbeamercolor{title}{fg=myblue}
	\setbeamercolor{frametitle}{fg=black}
}

\newcommand{\myemph}[1]{{\color{myblue}#1}}

% Drawing with TiKz
\usepackage{tikz}
\usetikzlibrary[positioning]
\usetikzlibrary{patterns}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{shapes.arrows}
\usetikzlibrary{arrows}
\usetikzlibrary{calc,positioning,shapes,chains,mindmap,trees,decorations}

\usepackage{rotating}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{graphicx,graphics}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{tabularx}
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\newcolumntype{Z}{>{\raggedright\arraybackslash}X}
\newcolumntype{W}{>{\raggedleft\arraybackslash}X}

\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usepackage{pgfgantt}

\usepackage[pdf]{pstricks}

\usepackage[absolute,overlay]{textpos}

\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\color{mygreen}\ding{51}}%
\newcommand{\xmark}{\color{myred}\ding{55}}%
\newcommand{\mehhh}{\color{mygray}$\sim$}%

\newcommand{\prog}{ $\vartriangleright$ }

\usepackage{enumitem}
\setitemize{label=\usebeamerfont*{itemize item}%
  \usebeamercolor[fg]{itemize item}
  \usebeamertemplate{itemize item}}

\newenvironment{compitem}{\begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]}{\end{itemize}}

\newcommand{\wholeslide}[1]{{
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{#1}}
\begin{frame}[plain]
\end{frame}
}}
\usepackage{soul}

\makeatletter
\def\imod#1{\allowbreak\mkern7mu{\operator@font mod}\,\,#1}
\def\m#1{\allowbreak\mkern10mu({\operator@font mod}\,\,#1)}
\makeatother

\newcommand{\summary}[5]{%
	\begin{frame}
		\begin{itemize}
			\item {\color{#1}Single-cell biology}
			\item {\color{#2}Trajectory inference}
			\item {\color{#3}Network inference}
			\item {\color{#4}Benchmarking}
			\item {\color{#5}Practical implications}
		\end{itemize}
	\end{frame}
}

\title{Modelling single-cell dynamics\\with trajectories and gene regulatory networks}

\author{Robrecht Cannoodt}

\date{}

\begin{document}

\begin{frame}[plain]
  \titlepage
  \hfill
  \includegraphics[height=.42cm]{../fig/logos/icon_UGent_WE_EN_RGB_color}\hfill
  \includegraphics[height=.60cm]{../fig/logos/vib_rf_inflammation_research_rgb_pos} \hfill
  \includegraphics[height=.48cm]{../fig/logos/dambi_logo} \hfill
  \includegraphics[height=.42cm]{../fig/logos/cmgg_myshort} \hfill
  \includegraphics[height=.42cm]{../fig/logos/FWO_Logo_KleurCMYK} 
  \hfill
\end{frame}

\begin{frame}
	\frametitle{Single-cell biology \uncover<2->{is to study an organism}\\\uncover<3->{by studying the behaviour of its cells}}
	\begin{center}
		\only<2>{\includegraphics[width=\linewidth]{figures/1a_singlecellbiology}}%
		\only<3>{\includegraphics[width=\linewidth]{figures/1b_singlecellbiology}}%
		\only<4>{\includegraphics[width=\linewidth]{figures/1c_singlecellbiology}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Technological advances allow high-throughput\\single-cell biology}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/2_part_scaling.pdf}\\
		\flushright 		{\vspace*{-1em}\footnotesize Regev et al. 2018}
	\end{center}
  \vspace*{-2.2em}
  \uncover<2->{
	  \rotatebox[origin=c]{180}{$\Lsh$} Trajectory inference \\
	  \rotatebox[origin=c]{180}{$\Lsh$} Network inference \\
	}
\end{frame}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{black}{lightgray}{lightgray}{lightgray}

\begin{frame}
	\frametitle{Cells are highly dynamic entities}
	\begin{center}
		\includegraphics[width=.6\linewidth]{figures/3_dynamiccells.pdf}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{A typical trajectory inference analysis}
	\begin{center}
		\only<1>{\includegraphics[width=\linewidth]{figures/4a_trajectoryinference.pdf}}%
		\only<2>{\includegraphics[width=\linewidth]{figures/4b_trajectoryinference.pdf}}%
		\only<3>{\includegraphics[width=\linewidth]{figures/4c_trajectoryinference.pdf}}%
		\only<4>{\includegraphics[width=\linewidth]{figures/4d_trajectoryinference.pdf}}%
		\only<5>{\includegraphics[width=\linewidth]{figures/4e_trajectoryinference.pdf}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Benchmark of 45 trajectory inference methods}
	\begin{center}
		\only<1>{\includegraphics[width=.9\linewidth]{figures/5a_benchmark.pdf}}%
		\only<2>{\includegraphics[width=.9\linewidth]{figures/5b_benchmark.pdf}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Interactive guidelines}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/6_guidelines.pdf}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{SCORPIUS: Linear trajectory inference}
	\begin{center}
		\includegraphics[width=.8\linewidth]{figures/end_part_scorpius.pdf}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{dyno: A toolkit for inferring and interpreting trajectories}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/end_part_dyno.pdf}
	\end{center}
\end{frame}
\begin{frame}
	\frametitle{Perspective: self-assessment in trajectory inference}
	\begin{center}
		\includegraphics[width=.8\linewidth]{figures/end_part_selfassessment.pdf}
	\end{center}
\end{frame}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{lightgray}{black}{lightgray}{lightgray}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{lightgray}{lightgray}{black}{lightgray}

\begin{frame}
	\frametitle{dyngen}
	\begin{center}
		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
	\end{center}
\end{frame}

\begin{frame}
\frametitle{Guidelines for benchmarking computational tools}
\begin{center}
	\includegraphics[width=\linewidth]{figures/end_part_guidelines.pdf}
\end{center}
\end{frame}

%%%%%%%%%%%%%% SUMMARY
\summary{lightgray}{lightgray}{lightgray}{lightgray}{black}

\end{document}

### 2019-11-22 ch6_bred/scripts/2_tcga.R

@@ -270,7 +270,7 @@ write_lines(colnames(expr), paste0(data_dir, "genes.txt"))
 
 if (!file.exists(paste0(data_dir, "qsub_handle.rds"))) {
   regulators_filt <- regulators
-  targets_filt <- targets[1:5000]
+  targets_filt <- targets[1:10000]
   expr_filt <- expr[, union(targets_filt, regulators_filt)]
   samples_filt <- rownames(expr_filt)
 

### 2019-11-22 defence_internal/figures/script.R

@@ -1,0 +1,18 @@
+library(tidyverse)
+library(scales)
+
+set.seed(1)
+df <- tibble(
+  x = round(10^runif(6, .5, )),
+  n = paste0("Molecule ", seq_along(x)) %>% forcats::fct_rev()
+)
+
+ggplot(df) +
+  geom_bar(aes(n, x), stat = "identity") +
+  theme_classic() +
+  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                labels = trans_format("log10", math_format(10^.x)),
+                limits = 10^c(0, 5)) +
+  labs(x = NULL, y = "Abundance") +
+  coord_flip()
+

### 2019-11-23 defence_internal/pres_v1.tex
wip persentation

@@
 }
+\newcommand{\topright}[1]{%
+\begin{textblock*}{.15\linewidth}(.98\linewidth,.05\linewidth)
+	\includegraphics[width=\linewidth]{figures/overview_#1.pdf}
+\end{textblock*}}
 
@@
 
+%%%%%%%%%%%%%% SINGLE CELL BIOLOGY %%%%%%%%%%%%%%
 \begin{frame}
@@
   \vspace*{-2.2em}
-  \uncover<2->{
-	  \rotatebox[origin=c]{180}{$\Lsh$} Trajectory inference \\
-	  \rotatebox[origin=c]{180}{$\Lsh$} Network inference \\
-	}
 \end{frame}
 
-%%%%%%%%%%%%%% SUMMARY
-\summary{lightgray}{black}{lightgray}{lightgray}{lightgray}
+\begin{frame}
+	\frametitle{Research questions}
+	\begin{center}
+		\only<2>{\includegraphics[width=\linewidth]{figures/11a_fundamental.pdf}}%
+		\only<3>{\includegraphics[width=\linewidth]{figures/11b_fundamental.pdf}}%
+		\only<4>{\includegraphics[width=\linewidth]{figures/11c_fundamental.pdf}}%
+		\only<5>{\includegraphics[width=\linewidth]{figures/11d_fundamental.pdf}}%
+	\end{center}
+\end{frame}
+
+\begin{frame}
+\includegraphics[width=\linewidth]{figures/overview.pdf}
+\end{frame}
+
+%%%%%%%%%%%%%% TRAJECTORY INFERENCE %%%%%%%%%%%%%%
+
+\begin{frame}
+	\vfill
+	\begin{center}
+		\LARGE Trajectory inference
+	\end{center}
+	\vfill
+\end{frame}
 
@@
 \begin{frame}
+	\frametitle{dyngen: Benchmarking with \textit{in silico} cells}
+	\begin{center}
+		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
+	\end{center}
+	\topright{ch2}
+\end{frame}
+
+\begin{frame}
 	\frametitle{Benchmark of 45 trajectory inference methods}
-	\begin{center}
+%	\begin{center}
 		\only<1>{\includegraphics[width=.9\linewidth]{figures/5a_benchmark.pdf}}%
 		\only<2>{\includegraphics[width=.9\linewidth]{figures/5b_benchmark.pdf}}%
-	\end{center}
+%	\end{center}
+	\topright{ch3}
 \end{frame}
@@
 	\end{center}
+	\topright{ch3}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Perspective: self-assessment in trajectory inference}
+	\begin{center}
+		\includegraphics[width=.8\linewidth]{figures/end_part_selfassessment.pdf}
+	\end{center}
+  \topright{ch8}
 \end{frame}
@@
 	\end{center}
+	\topright{ch5}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{dyno: A toolkit for inferring and interpreting trajectories}
+	\frametitle{dyno: A toolkit for inferring and\\interpreting trajectories}
 	\begin{center}
-		\includegraphics[width=\linewidth]{figures/end_part_dyno.pdf}
+		\includegraphics[width=.9\linewidth]{figures/end_part_dyno.pdf}
 	\end{center}
-\end{frame}
-\begin{frame}
-	\frametitle{Perspective: self-assessment in trajectory inference}
-	\begin{center}
-		\includegraphics[width=.8\linewidth]{figures/end_part_selfassessment.pdf}
-	\end{center}
+	\topright{ch4}
 \end{frame}
 
-%%%%%%%%%%%%%% SUMMARY
-\summary{lightgray}{lightgray}{black}{lightgray}{lightgray}
-
-%%%%%%%%%%%%%% SUMMARY
-\summary{lightgray}{lightgray}{lightgray}{black}{lightgray}
+%%%%%%%%%%%%%% NETWORK INFERENCE %%%%%%%%%%%%%%
 
 \begin{frame}
-	\frametitle{dyngen}
+	\vfill
 	\begin{center}
-		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
+		\LARGE Network inference
+	\end{center}
+	\vfill
+\end{frame}
+
+\begin{frame}
+	\frametitle{Cellular dynamics is (i.a.) driven by gene regulation}
+	\begin{center}
+		\only<1>{\includegraphics[width=.8\linewidth]{figures/7a_generegulation.pdf}}%
+		\only<2>{\includegraphics[width=.8\linewidth]{figures/7b_generegulation.pdf}}%
+		\only<3>{\includegraphics[width=.8\linewidth]{figures/7c_generegulation.pdf}}%
+		\only<4>{\includegraphics[width=.8\linewidth]{figures/7d_generegulation.pdf}}%
+		\only<5>{\includegraphics[width=.8\linewidth]{figures/7e_generegulation.pdf}}%
+		\only<6>{\includegraphics[width=.8\linewidth]{figures/7f_generegulation.pdf}}%
 	\end{center}
@@
 \begin{frame}
-\frametitle{Guidelines for benchmarking computational tools}
+	\frametitle{Classical network inference}
+	\begin{center}
+		\includegraphics[width=.9\linewidth]{figures/8_networkinference.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Case-specific network inference}
+	\begin{center}
+		\only<1>{\includegraphics[width=\linewidth]{figures/9a_csni_example.pdf}}%
+		\only<2>{\includegraphics[width=\linewidth]{figures/9b_csni_example.pdf}}%
+		\only<3>{\includegraphics[width=\linewidth]{figures/9c_csni_example.pdf}}%
+		\only<4>{\includegraphics[width=\linewidth]{figures/9d_csni_example.pdf}}%
+	\end{center}
+	\topright{ch6}
+\end{frame}
+
+\begin{frame}
+	\frametitle{bred applied to TCGA}
+	\begin{center}
+		\includegraphics[width=.85\linewidth]{figures/10b_tcga.pdf}
+%		\only<1>{\includegraphics[width=.6\linewidth]{figures/10a_tcga.pdf}}%
+%		\only<2>{\includegraphics[width=.85\linewidth]{figures/10b_tcga.pdf}}%
+	\end{center}
+	\topright{ch6}
+\end{frame}
+
+\begin{frame}
+	\frametitle{incgraph: Optimising regulatory networks}
+	\begin{center}
+		\includegraphics[width=\linewidth]{figures/end_part_incgraph.pdf}
+	\end{center}
+	\topright{ch7}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%% BENCHMARKING
+\begin{frame}
+	\vfill
+	\begin{center}
+		\LARGE Benchmarking
+	\end{center}
+	\vfill
+\end{frame}
+
+\begin{frame}
+	\frametitle{Essential guidelines for benchmarking\\computational tools}
 	\begin{center}
@@
 	\end{center}
+	\topright{ch9}
 \end{frame}
 
-%%%%%%%%%%%%%% SUMMARY
-\summary{lightgray}{lightgray}{lightgray}{lightgray}{black}
+%%%%%%%%%%%%%% IMPLICATIONS %%%%%%%%%%%%%%
+
+\begin{frame}
+	\frametitle{Practical implications}
+	\begin{itemize}
+		\item<2-> Fundamental research in single-cell analyses \\
+		  \uncover<3->{\begin{center}\includegraphics[width=.7\linewidth]{figures/11d_fundamental.pdf}\end{center}}
+		\item<4-> Benchmarking
+		\item<5-> R \& tidyverse
+	\end{itemize}
+  \topright{all}
+\end{frame}
 

### 2019-11-23 defence_internal/figures/script.R

@@ -15,4 +15,94 @@ ggplot(df) +
                 limits = 10^c(0, 5)) +
   labs(x = NULL, y = "Abundance") +
   coord_flip()
+ggsave("defence_internal/figures/1_part_abundance.pdf", width = 4, height = 3)
 
+
+
+df <- tribble(
+  ~x, ~y, ~name, ~hjust, ~vjust,
+  0, 0, "§2: dyngen", .5, 1,
+  .4, 1, "§3: dynbenchmark", 0, .5,
+  .6, 1, "§8: self-assessment\nin trajectory inference", 0, .5,
+  .8, 1, "§5: SCORPIUS", 0, .5,
+  1, 1, "§4: dyno", 0, .5,
+  1.8, 1, "§6: bred", 1, .5,
+  2, 1, "§7: incgraph", 1, .5,
+  3, 1, "§9: guidelines\nfor benchmarking", .5, 0
+)
+titdf <- tibble(
+  x = 0:2,
+  y = c(1.6, 1.8, 1.8),
+  # y = 1.6,
+  # y = c(1.6, 1.5, 1.5),
+  name = c("Benchmarking", "Trajectory\ninference", "Network\ninference"),
+  short = c("Benchmarking", "TI", "NI"),
+  col = RColorBrewer::brewer.pal(3, "Set1"),
+  hjust = c(.5, 0, 1)
+)
+
+rgbs <- grDevices::col2rgb(titdf$col)
+mix_cols <- function(df) {
+  rgb <-
+    map2(df$x, df$y, function(x, y) {
+      case_when(
+        y == 0 ~ c(1, 1, 1) / 3,
+        x <= 1 ~ c(1 - x, x, 0),
+        x <= 2 ~ c(0, 2 - x, x - 1),
+        TRUE ~ c(x - 2, 0, 3 - x)
+      )
+    }) %>% do.call(rbind, .)
+  rgb2 <- round(rgbs %*% t(rgb))
+  df$col <- grDevices::rgb(rgb2[1,], rgb2[2,], rgb2[3,], maxColorValue = 255)
+  df
+}
+
+df <- df %>% mix_cols()
+
+arrdf <- tribble(
+  ~gr, ~x, ~y,
+  2, .4, 1,
+  2, .76, 1,
+  3, 1, 1,
+  3, 1.575, 1,
+  4, 2, 1,
+  4, 2.5, 1
+)
+
+g <- ggplot(df) +
[…]

### 2019-11-23 rebuttal/rebuttal.tex
~
update

@@
 
-\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles. Compare against other methods, e.g. GENIE3 + r	egulon specificity score.}
+\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles. Compare against other methods, e.g. GENIE3 + regulon specificity score.}
 
@@
 \answ{Response or statement by candidate.}
+
+\section{Dries De Maeyer}
+
+\subsection{Chapter 1}
+
+\exam{The scope of the complete thesis is to define state transitions in cell types and the
+	definition of cell type/state transition is not clear cut. I would frame the definition of a cell
+	type more based on our current understanding and technology of biology. I.e. where do you
+	draw the line of a new cell type?}
+
+\exam{Could you make the main outline of the thesis could even be more clear by
+	providing a graphical overview of the different topics addressed in this thesis?}
+
+\subsection{Chapter 2}
+\exam{Would you be able to extend this method to include Citeseq data? How would
+	you see this?}
+
+\exam{Can you use the generated models for validation with existing cellular
+	heterogeneity from scRNA-seq data?}
+
+\subsection{Chapter 3}
+
+\exam{Since an ensemble of methods is proposed as the best approach to perform TI,
+	how would this be addressed? Is there an objective means of selecting good performing
+	methods?}
+
+\exam{Can TI methods be used for detecting and quantifying technical variation in
+	between different samples? Be used as a metric for assessing batch integration?}
+
+\subsection{Chapter 7}
+
+\exam{Based on the method presented here, we could “learn” the topology of biological
+	networks. What is to your insight the transposability of these topologies?}
+
+\subsection{Chapter 8}
+\exam{\nimp{The eight chapter is a manuscript in preparation about the assessment of TI methods. The
+		main discussion topic is the lack of a quantitative measure for the scoring of new TI
+		methods. This leads to the publication and development of new methods that do not show
+		a clear performance improvement compared to other methods. The main topic touches
+		upon some sore points that are inherent in bioinformatics. The main thought I have with
+		this chapter is that defining clear goals in an emerging technology is difficult and therefore is
+		sometimes ignored by fellow bioinformaticians. This because the lack of standardized
+		performance metrics, missing public datasets that are well described and an ad hoc problem
+		definition. I follow the idea of defining a clear cut approach to benchmarking (as described
+		in chapter nine) and I like the proposition that is started in this manuscript for a discussion
+		on good TI metrics.}}
+
+\exam{Would approaches like e.g. DREAM challenges be a means of tackling these issues
+	to your opinion?}
+
+\exam{Do you expect that all methods which are mainly developed to answer specific
+	biological questions would adhere to the same biological problem definition? Is the fact that
+	people are trying to answer different question not a prior to define more generalized
+	questions?}
 

### 2019-11-23 defence_internal/figures/script.R

@@ -83,6 +83,29 @@ g <- ggplot(df) +
 ggsave("defence_internal/figures/overview.pdf", g, width = 6, height = 6)
 
 
+
+g <- ggplot(df) +
+  geom_segment(aes(x = x, xend = x, y = 0, yend = 1.05), titdf, colour = "lightgray", linetype = "dashed") +
+  geom_segment(aes(x = 0, xend = 3, y = 1, yend = 1), tibble(1), colour = "lightgray", linetype = "dashed") +
+  geom_text(aes(x, y, label = name, colour = col), titdf, size = 6, fontface = "bold") +
+  coord_polar() +
+  dynplot::theme_graph() +
+  theme(plot.margin=grid::unit(c(0,0,0,0), "mm")) +
+  scale_colour_identity()
+ggsave("defence_internal/figures/overview_categories.pdf", g, width = 6, height = 6)
+
+
+g <- ggplot(df) +
+  geom_segment(aes(x = x, xend = x, y = 0, yend = 1.05), titdf, colour = "lightgray", linetype = "dashed") +
+  geom_segment(aes(x = 0, xend = 3, y = 1, yend = 1), tibble(1), colour = "lightgray", linetype = "dashed") +
+  geom_text(aes(x, y, label = name, colour = col), titdf, size = 6, fontface = "bold") +
+  coord_polar() +
+  dynplot::theme_graph() +
+  theme(plot.margin=grid::unit(c(0,0,0,0), "mm")) +
+  scale_colour_identity()
+ggsave("defence_internal/figures/overview_categories_dots.pdf", g, width = 6, height = 6)
+
+
 for (i in 2:9) {
   g <- ggplot(df) +
     geom_segment(aes(x = x, xend = x, y = 0, yend = 1.05), titdf, colour = "lightgray", linetype = "dashed") +

### 2019-11-23 defence_internal/pres_v1.tex
update
~

@@
 \definecolor{mygreen}{rgb}{0,0.8,0.1}
+
+\definecolor{set1red}{HTML}{E41A1C}
+\definecolor{set1blue}{HTML}{377EB8}
+\definecolor{set1green}{HTML}{4DAF4A}
 
@@
 \begin{frame}
-	\frametitle{Research questions}
+	\frametitle{Research questions regarding single-cell biology}
 	\begin{center}
-		\only<2>{\includegraphics[width=\linewidth]{figures/11a_fundamental.pdf}}%
-		\only<3>{\includegraphics[width=\linewidth]{figures/11b_fundamental.pdf}}%
-		\only<4>{\includegraphics[width=\linewidth]{figures/11c_fundamental.pdf}}%
-		\only<5>{\includegraphics[width=\linewidth]{figures/11d_fundamental.pdf}}%
+		\only<1>{\includegraphics[width=\linewidth]{figures/11a_fundamental.pdf}}%
+		\only<2>{\includegraphics[width=\linewidth]{figures/11b_fundamental.pdf}}%
+		\only<3>{\includegraphics[width=\linewidth]{figures/11c_fundamental.pdf}}%
+		\only<4>{\includegraphics[width=\linewidth]{figures/11d_fundamental.pdf}}%
+		\only<5>{\includegraphics[width=\linewidth]{figures/11e_fundamental.pdf}}%
+		\only<6>{\includegraphics[width=\linewidth]{figures/11f_fundamental.pdf}}%
 	\end{center}
@@
 \begin{frame}
-\includegraphics[width=\linewidth]{figures/overview.pdf}
+  \only<1>{\includegraphics[width=\linewidth]{figures/overview_categories.pdf}}
+  \only<2>{\includegraphics[width=\linewidth]{figures/overview_categories_dots.pdf}}
+  \only<3>{\includegraphics[width=\linewidth]{figures/overview.pdf}}
 \end{frame}
@@
 	\begin{center}
-		\LARGE Trajectory inference
+		\color{set1blue}
+		{\LARGE Trajectory inference} \\
+		How do cells change over time?
 	\end{center}
@@
 		\only<5>{\includegraphics[width=\linewidth]{figures/4e_trajectoryinference.pdf}}%
+		\only<6>{\includegraphics[width=\linewidth]{figures/4f_trajectoryinference.pdf}}%
 	\end{center}
@@
 	\begin{center}
-		\LARGE Network inference
+		\color{set1green}
+		{\LARGE Network inference}\\
+		What drives cells to change?
 	\end{center}
@@
 
-\begin{frame}
-	\frametitle{Cellular dynamics is (i.a.) driven by gene regulation}
-	\begin{center}
-		\only<1>{\includegraphics[width=.8\linewidth]{figures/7a_generegulation.pdf}}%
-		\only<2>{\includegraphics[width=.8\linewidth]{figures/7b_generegulation.pdf}}%
-		\only<3>{\includegraphics[width=.8\linewidth]{figures/7c_generegulation.pdf}}%
-		\only<4>{\includegraphics[width=.8\linewidth]{figures/7d_generegulation.pdf}}%
-		\only<5>{\includegraphics[width=.8\linewidth]{figures/7e_generegulation.pdf}}%
-		\only<6>{\includegraphics[width=.8\linewidth]{figures/7f_generegulation.pdf}}%
-	\end{center}
-\end{frame}
+%\begin{frame}
+%	\frametitle{Cellular dynamics is (i.a.) driven by gene regulation}
+%	\begin{center}
+%		\only<1>{\includegraphics[width=.8\linewidth]{figures/7a_generegulation.pdf}}%
+%		\only<2>{\includegraphics[width=.8\linewidth]{figures/7b_generegulation.pdf}}%
+%		\only<3>{\includegraphics[width=.8\linewidth]{figures/7c_generegulation.pdf}}%
+%		\only<4>{\includegraphics[width=.8\linewidth]{figures/7d_generegulation.pdf}}%
+%		\only<5>{\includegraphics[width=.8\linewidth]{figures/7e_generegulation.pdf}}%
+%		\only<6>{\includegraphics[width=.8\linewidth]{figures/7f_generegulation.pdf}}%
+%	\end{center}
+%\end{frame}
 
@@
 \begin{frame}
-	\frametitle{Case-specific network inference}
+	\frametitle{Classical vs. sample-specific network inference}
+	\begin{center}
+		\includegraphics[width=.9\linewidth]{figures/12_nitypes.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Sample-specific network inference}
 	\begin{center}
@@
 	\begin{center}
-		\LARGE Benchmarking
+		\color{set1blue}
+		{\LARGE Benchmarking} \\
+		How to quantify the accuracy of computational tools?
 	\end{center}
 	\vfill
+\end{frame}
+
+\begin{frame}
+	\frametitle{dyngen: Benchmarking with \textit{in silico} cells}
+	\begin{center}
+		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
+	\end{center}
+	\topright{ch2}
 \end{frame}
@@
 	\end{center}
+	{\scriptsize 
+	\begin{flushright}
+		In collaboration with the labs of Mark Robinson,\\Anne-Laure Boulesteix, and Paul P. Gardner.
+	\end{flushright}}
 	\topright{ch9}
@@
 	\frametitle{Practical implications}
-	\begin{itemize}
-		\item<2-> Fundamental research in single-cell analyses \\
-		  \uncover<3->{\begin{center}\includegraphics[width=.7\linewidth]{figures/11d_fundamental.pdf}\end{center}}
-		\item<4-> Benchmarking
-		\item<5-> R \& tidyverse
-	\end{itemize}
+  \uncover<2->{\begin{center}\includegraphics[width=.7\linewidth]{figures/11f_fundamental.pdf}\end{center}}
+	\uncover<3->{Contributions to open-source software in the R ecosystem}
   \topright{all}

### 2019-11-24 defence_internal/pres_v1.tex
~

@@
 		\only<5>{\includegraphics[width=\linewidth]{figures/11e_fundamental.pdf}}%
-		\only<6>{\includegraphics[width=\linewidth]{figures/11f_fundamental.pdf}}%
 	\end{center}
@@
 
-\begin{frame}
-	\frametitle{dyngen: Benchmarking with \textit{in silico} cells}
-	\begin{center}
-		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
-	\end{center}
-	\topright{ch2}
-\end{frame}
+%\begin{frame}
+%	\frametitle{dyngen: Benchmarking with \textit{in silico} cells}
+%	\begin{center}
+%		\includegraphics[width=.6\linewidth]{figures/end_part_dyngen.pdf}
+%	\end{center}
+%	\topright{ch2}
+%\end{frame}
 
@@
 
+\begin{frame}
+	\includegraphics[width=\linewidth]{figures/overview.pdf}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%% EXTRA
+\begin{frame}
+	\vfill
+	\begin{center}
+		\color{mygray}
+		{\LARGE Extra slides} 
+	\end{center}
+	\vfill
+\end{frame}
+
+\begin{frame}
+	\frametitle{Sample-specific network inference}
+	\begin{center}
+		\only<1>{\includegraphics[width=.7\linewidth]{figures/extra_bred1_simplify.pdf}}%
+		\only<2>{\includegraphics[width=\linewidth]{figures/extra_bred2a_fimp.pdf}}%
+		\only<3>{\includegraphics[width=\linewidth]{figures/extra_bred2b_fimp.pdf}}%
+		\only<4>{\includegraphics[width=\linewidth]{figures/extra_bred2c_fimp.pdf}}%
+		\only<5>{\includegraphics[width=\linewidth]{figures/extra_bred2d_fimp.pdf}}%
+		\only<6>{\includegraphics[width=\linewidth]{figures/extra_bred2e_fimp.pdf}}%
+		\only<7>{\includegraphics[width=\linewidth]{figures/extra_bred2f_fimp.pdf}}%
+		\only<8>{\includegraphics[width=\linewidth]{figures/extra_bred2g_fimp.pdf}}%
+		\only<9>{\includegraphics[width=\linewidth]{figures/extra_bred2h_fimp.pdf}}%
+		\only<10>{\includegraphics[width=\linewidth]{figures/extra_bred2i_fimp.pdf}}%\\
+		\only<11>{\includegraphics[width=\linewidth]{figures/extra_bred2j_fimp.pdf}}%
+		\only<12>{\includegraphics[width=\linewidth]{figures/extra_bred2k_fimp.pdf}}%
+		\only<13>{\includegraphics[width=\linewidth]{figures/extra_bred2l_fimp.pdf}}%
+	\end{center}
+	\topright{ch6}
+\end{frame}
+
 \end{document}

### 2019-11-25 rebuttal/rebuttal.tex
update

@@
 
+% checks and boxes
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}%
+\newcommand{\xmark}{\ding{55}}%
+\newcommand{\todo}{$\square$}
+\newcommand{\done}{\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\checkmark$}}%
+
 %%%%%%%%%%%%%%%%%%%%%
@@
 }
+\definecolor{myblue}{rgb}{0,0.4,0.7}
+\definecolor{myred}{rgb}{0.7,.1,.1}
+\definecolor{mygray}{rgb}{0.7,0.7,0.7}
+\definecolor{mygreen}{rgb}{0,0.8,0.1}
 
-\newcommand{\exam}[2][\  ]{\hspace{0pt}\marginpar{\color{red}#1}$\bullet$ \textit{#2}}
-\newcommand{\imp}[1]{{\color{red} #1}}
-\newcommand{\nimp}[1]{{\color{gray} #1}}
-\newcommand{\answ}[1]{{\color{blue} $\triangleright$ #1}}
+\newcommand{\exam}[2][\  ]{\hspace{0pt}\marginpar{\color{myred}#1}$\bullet$ \textit{#2}}
+%\newcommand{\imp}[1]{{\color{myred} #1}}
+\newcommand{\imp}[1]{\textbf{#1}}
+\newcommand{\nimp}[1]{{\color{mygray} #1}}
+\newcommand{\answ}[1]{{\color{myblue} $\triangleright$ #1}}
+\newcommand{\task}[2][\todo]{{\color{myblue} #1 #2}}
 
@@
 
-\section{Summary of most important comments}
-
-\textbf{Overall:} From three jury members: Merge chapters 8-9-10 into one discussion, or merge chapter 8 into chapter 3.
-
-\textbf{Overall:} Random Forest variable importances are often used to detect differentially expressed genes. Why RF and not more standard statistical tests?
-
-\textbf{Introduction:} {Describe the practical implications of the
-		technological advances (e.g. precision medicine, new treatments) and
-		motivate the research from a non-technical view point (How is/can this PhD
-		change the world?)
-
-\textbf{dyngen:} {Table 2.2: explain default values and formulae}
-
-\textbf{dyngen:} {How crucial are certain parts of the GRN model, e.g. mRNA production and degradation?}
-
-\textbf{dyngen:} {Compare against other scRNA-seq generators}
-
-\textbf{dyngen:} {It is claimed dyngen is a multi-modality simulator, yet it produces only transcriptomics data}
-
-\textbf{dyngen:} May need a working example to demonstrate the steps a user needs to take to use the software, and to illustrate how the results look like.
+%\section{Summary of most important comments}
+%
+%\textbf{Overall:} From three jury members: Merge chapters 8-9-10 into one discussion, or merge chapter 8 into chapter 3.
+%
+%\textbf{Overall:} Random Forest variable importances are often used to detect differentially expressed genes. Why RF and not more standard statistical tests?
+%
+%\textbf{Introduction:} {Describe the practical implications of the
+%		technological advances (e.g. precision medicine, new treatments) and
+%		motivate the research from a non-technical view point (How is/can this PhD
+%		change the world?)
+%	
+%\textbf{dyngen:} {Table 2.2: explain default values and formulae}
+%
+%\textbf{dyngen:} {How crucial are certain parts of the GRN model, e.g. mRNA production and degradation?}
+%
+%\textbf{dyngen:} {Compare against other scRNA-seq generators}
+%
+%\textbf{dyngen:} {It is claimed dyngen is a multi-modality simulator, yet it produces only transcriptomics data}
+%
+%\textbf{dyngen:} May need a working example to demonstrate the steps a user needs to take to use the software, and to illustrate how the results look like.
 No references to user documentation, source code (open source?).
 No discussion about the performance of the framework.
-I might also have expected a class diagram on the software, as a proof that the framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter.
+%I might also have expected a class diagram on the software, as a proof that the framework has been designed with extensibility in mind, as is claimed in the introduction of the chapter. 
+%
+%\textbf{dynbenchmark:} {Comparisons focus mainly on relative differences between methods. It would have been interesting to also assess the absolute performance of each method.}
+%
+%\textbf{dynbenchmark:} {No statistical tests were performed to assess whether methods are significantly different from one another}
+%
+%\textbf{dynbenchmark:} Turn into UK English to be consistent with the rest of the dissertation.
+%
+%\textbf{dyno:} Does not really read as a scientific paper (yet).
+The results section should be improved, it currently reads more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study.
+%
+%\textbf{SCORPIUS:} {The methodology has changed between the bioRxiv preprint and this chapter, yet Figure 5.4 is exactly the same. Did the changes have no (measurable) effect on this analysis?}
+%
+%\textbf{SCORPIUS:} {Discuss the computational complexity of SCORPIUS.}
+%
+%\textbf{SCORPIUS:} How objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?
+%
+%\textbf{bred:} {Better explain how bred is different from GENIE3.}
+%
+%\textbf{bred:} {What makes bred a true case-wise NI method? Provide a more formal definition of what a case-wise regulatory network is.}
+%
+%\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles. Compare against other methods, e.g. GENIE3 + regulon specificity score.}
+%
+%\textbf{bred:} {Discussion is very short, please elaborate.}
+%
+%\textbf{IncGraph:} {On optimising GRNs, compare a simple MST with IncGraph}
+%
+%\textbf{Benchmarking guidelines:} {As the candidate is not first author on this paper, I’m not sure if this paper
+%	should be included in the PhD book verbatim.}
+%
+%\textbf{Discussion:} {The Discussion chapter is missing a more broad context
+%			description. How are other researchers using these tools to gain new biological insights? Where is the field of modelling single-cell dynamics going?}
+%		
+%\textbf{Discussion:} {Reads more like a summary. Discuss the links between different chapters.}
 
-\textbf{dynbenchmark:} {Comparisons focus mainly on relative differences between methods. It would have been interesting to also assess the absolute performance of each method.}
-
-\textbf{dynbenchmark:} {No statistical tests were performed to assess whether methods are significantly different from one another}
-
-\textbf{dynbenchmark:} Turn into UK English to be consistent with the rest of the dissertation.
-
-\textbf{dyno:} Does not really read as a scientific paper (yet).
-The results section should be improved, it currently reads more as an incoherent listing of features of dyno, where an overall flow of this section is missing for the readership; adding some details about the case study.
-
-\textbf{SCORPIUS:} {The methodology has changed between the bioRxiv preprint and this chapter, yet Figure 5.4 is exactly the same. Did the changes have no (measurable) effect on this analysis?}
-
-\textbf{SCORPIUS:} {Discuss the computational complexity of SCORPIUS.}
-
-\textbf{SCORPIUS:} How objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?
-
-\textbf{bred:} {Better explain how bred is different from GENIE3.}
-
-\textbf{bred:} {What makes bred a true case-wise NI method? Provide a more formal definition of what a case-wise regulatory network is.}
-
-\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles. Compare against other methods, e.g. GENIE3 + regulon specificity score.}
-
-\textbf{bred:} {Discussion is very short, please elaborate.}
-
-\textbf{IncGraph:} {On optimising GRNs, compare a simple MST with IncGraph}
-
-\textbf{Benchmarking guidelines:} {As the candidate is not first author on this paper, I’m not sure if this paper
-	should be included in the PhD book verbatim.}
-
-\textbf{Discussion:} {The Discussion chapter is missing a more broad context
-			description. How are other researchers using these tools to gain new biological insights? Where is the field of modelling single-cell dynamics going?}
-
-\textbf{Discussion:} {Reads more like a summary. Discuss the links between different chapters.}
-
-\newpage
+%\newpage
 \section{Legend}
@@
 
+\task{Unfinished task.}
+
+\task[\done]{Finished task.}
+
 \section{Dries De Maeyer}
@@
 	type more based on our current understanding and technology of biology. I.e. where do you
-	draw the line of a new cell type?}
+	draw the line of a new cell type?} \\
+\task{Define cell type in introduction.}
 
 \exam{Could you make the main outline of the thesis could even be more clear by
-	providing a graphical overview of the different topics addressed in this thesis?}
+	providing a graphical overview of the different topics addressed in this thesis?} \\
+\task{Add outline to thesis}
 
@@
 \exam{Would you be able to extend this method to include Citeseq data? How would
-	you see this?}
+	you see this?}\\
+\answ{Yes.
+dyngen already generates protein abundance information, but more work is needed to compare characteristics of distributions of genes of citeseq or cytometry data to dyngen data, to ensure that the outputted information is realistic enough.
+As such, this feature is not included in the list of features in the current version of dyngen.} \\
+\task{Add this to discussion in dyngen?}
 
 \exam{Can you use the generated models for validation with existing cellular
-	heterogeneity from scRNA-seq data?}
+	heterogeneity from scRNA-seq data?} \\
+\task{Ask for explanation pertaining this question.}
 
@@
 	how would this be addressed? Is there an objective means of selecting good performing
-	methods?}
+	methods?} \\
+\answ{An internal measure for quantifying how well a given trajectory fits the original data would be extremely useful but also non-trivial.
+One internal measure could be: to what extent is the expression of some of the genes smooth along the trajectory.
+Variations of this approach should be tried and compared to gold standard metrics.}
 
@@
 	between different samples? Be used as a metric for assessing batch integration?}
+\answ{In some cases, yes.
+If two similar but disconnected trajectories are detected, it is possible to integrate those samples by performing trajectory alignment.
+Up until now, I have not encountered datasets where this approach would likely yield acceptable results, and instead would recommend performing batch correction separately from trajectory inference.}
 
@@
 \exam{Based on the method presented here, we could “learn” the topology of biological
-	networks. What is to your insight the transposability of these topologies?}
+	networks. What is to your insight the transposability of these topologies?}\\
+\task{Ask for explanation pertaining this question.}
 
 \subsection{Chapter 8}
-\exam{\nimp{The eight chapter is a manuscript in preparation about the assessment of TI methods. The
-		main discussion topic is the lack of a quantitative measure for the scoring of new TI
-		methods. This leads to the publication and development of new methods that do not show
-		a clear performance improvement compared to other methods. The main topic touches
+\exam{\nimp{The eight chapter is a manuscript in preparation about the assessment of TI methods.
+The main discussion topic is the lack of a quantitative measure for the scoring of new TI methods.
+This leads to the publication and development of new methods that do not show a clear performance improvement compared to other methods.
+The main topic touches
 		upon some sore points that are inherent in bioinformatics. The main thought I have with
-		this chapter is that defining clear goals in an emerging technology is difficult and therefore is
-		sometimes ignored by fellow bioinformaticians. This because the lack of standardized
-		performance metrics, missing public datasets that are well described and an ad hoc problem
-		definition. I follow the idea of defining a clear cut approach to benchmarking (as described
-		in chapter nine) and I like the proposition that is started in this manuscript for a discussion
-		on good TI metrics.}}
-
-\exam{Would approaches like e.g. DREAM challenges be a means of tackling these issues
-	to your opinion?}
+		this chapter is that defining clear goals in an emerging technology is difficult and therefore is sometimes ignored by fellow bioinformaticians.
+This because the lack of standardized performance metrics, missing public datasets that are well described and an \textit{ad hoc} problem definition.
+I follow the idea of defining a clear cut approach to benchmarking (as described in chapter nine) and I like the proposition that is started in this manuscript for a discussion on good TI metrics.} Would approaches like e.g.
+DREAM challenges be a means of tackling these issues to your opinion?} \\
+\answ{Yes and no.
+I believe any discussions into quantifying performance in any subdomain of bioinformatics through efforts such as benchmarking studies or challenges are crucial.
+However, \ldots chicken and egg problem; \textit{in silico} simulators.} \\
+\task{Finish explanation.}
 
 \exam{Do you expect that all methods which are mainly developed to answer specific
-	biological questions would adhere to the same biological problem definition? Is the fact that
-	people are trying to answer different question not a prior to define more generalized
-	questions?}
+	biological questions would adhere to the same biological problem definition? Is the fact that people are trying to answer different question not a prior to define more generalized
+	questions?} \\
+\answ{Indeed, methods designed to answer specific biological questions might have (slightly) differing problem definitions.
+This is one of the central issues we encountered in performing our benchmark, because the differences in problem definitions resulted in very different data structures in the inputs and outputs of these methods.}
 
@@
 
-\subsection{Weaknesses of the thesis}
+\subsection{Weaknesses of the thesis} \label{sec:joeriweakness}
 \exam[\tagimp \tagtime]{\imp{The quality of writing, layout and graphical presentation is of a lower quality in the non-published parts of the thesis.
-Certain unpublished chapters feel rushed and unfinished.
-Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.}}
+Certain unpublished chapters feel rushed and unfinished.}} \\
+\task{Rewrite sections of dyno and of bred.}
+
+\exam{Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.} \\
+\task{Reorder chapters in a more logical order. Rewrite abstracts to better introduce the context of each of the chapters.}
 
@@
 	description. After reading the work, I gained very little insight how other
-	researchers are using these tools to advance science.}}
+	researchers are using these tools to advance science.}} \\
+\task{Rewrite introduction in terms of the three main research questions.
+Discuss how answering these PhD questions can change the world.
+Better motivate the significance of certain technological advancements.} \\
+\task{Rewrite discussion in terms of the three topics: Trajectory inference, network inference and benchmarking.
+Discuss the practical implications of this work, and future perspectives in the field.
+}
 
@@
 	published manuscripts into a thesis and I’m aware there is no (quick) solution
-	to this.}}
+	to this.}} \\
+\answ{Indeed, this is a result of bundling multiple manuscripts into one thesis.
+While there is a redundancy in information presented to the reader, these redundancies are largely confined to the introduction sections of each chapter.}
 
-\exam[\tagimp]{\nimp{It is very rare to see that research groups (or individual researchers) invest this
+\exam{\nimp{It is very rare to see that research groups (or individual researchers) invest this
 	much effort and time in creating qualitative software. I agree with the ideas of
@@
 	these cases? The original PhD students that developed the work? The original
-	research group? Will such higher standards not lead to barriers which might
+	research group?} \\
+\answ{I would argue that in order to develop software projects of this scale, adapting good software development practices (e.g.
+using code revision systems, unit testing, automated integration testing) 'pays for itself'.
+That is, making use of these practices requires some time investment at the start.
+However, developing and maintaining a large code base without these practices would require much greater effort, as it can lead to more frequent unexpected debugging expeditions as a result of long undetected bugs.}
+
+\exam{Will such higher standards not lead to barriers which might
 	be harder to overcome for PhD students which might be constrained by limited
-	support or funding?}
+	support or funding?} \\
+\answ{Most of the good programming practices which are advocated for throughout this dissertation (Table~3.1, Figure~8.5, Figure~9.1) are skills that are not difficult to acquire given that the PhD student in question is able to program.
+A two-day crash course in good software development practices is likely sufficient to obtain the basics of developing qualitative software for a given programming language.
+If a PhD student is aiming to develop software on which other research will depend, they should familiarise themselves with the proper 'quality control' tools of software development.}
 
@@
 	motivate the research from a non-technical view point (How is/can this PhD
-	change the world?)}}
+	change the world?)}} \\
+\answ{See section~\ref{sec:joeriweakness}.}
 
@@
 
-\exam[\tagimp \tagtime]{\imp{Table 2.2. contains many default ‘magic’ numbers?} It would be nice to know
+\exam{Table 2.2. contains many default ‘magic’ numbers? It would be nice to know
 		how in practice you ended up selecting them and which problems you
-		encountered during design.}
+		encountered during design.} \\
+\answ{The default values (or rather, default distributions) are determined based our understanding of biology; for example transcription occurs at a much faster rate than splicing or translation, and this is reflected in the default distributions.}
 
@@
 	Would the readability be improved if the order was changed in which the
-	concepts are explained? The backbone is only needed for the visualisation.}
+	concepts are explained? The backbone is only needed for the visualisation.} \\
+\answ{The backbone is an essential part of defining the gene regulatory network that is used for the simulation, not just for visualisation purposes.} \\
+\task{I will modify the text to clarify the purpose of the backbone better.}
 
 \exam{2.4.6. is a bit worrying. Could I reproduce dyngen without knowing these
-		details? Does the sampling process not deserve more explanation?}
+		details? Does the sampling process not deserve more explanation?} \\
+\task{I will better describe the different sampling procedures, with the help of one or more illustrations.}
 
@@
 \answ{Certain supplementary figures and files had not been included in the dissertation, but instead the reference contained a clickable URL which directs the reader to a downloadable file.
-Since this information is lost to readers of the printed dissertation, explicit captions have been added in a section for supplementary figures and tables, and contain a printable URL which direct the reader towards the file.}
+This information is lost to readers of the printed dissertation.} \\ \task[\done]{Explicit captions have been added in a section for supplementary figures and tables, and contain a printable URL which direct the reader towards the file.}
 
@@
 \exam[\tagimp \tagtime]{\imp{This chapter does not really read as a scientific paper (yet).} There are few
-	comments I have since this is mainly an instruction manual + feature overview.}
+	comments I have since this is mainly an instruction manual + feature overview.} \\
+\task{I will add a methods section to this chapter, as well as rewrite the results section to make it read more like a manuscript.}
 
 \exam{Figure 4.7 is missing slingshot and paga label in part B} \\
-\answ{Figure 4.7 does contain labels for Slingshot and PAGA. A small gap between part A and part B has been added, to improve interpretability.}
+\answ{Figure 4.7 does contain labels for Slingshot and PAGA.}
+\task[\done]{ small gap between part A and part B has been added, to improve interpretability.}
 
@@
 	change or other potential changes have no (measurable) effect on the initial
-	analysis? I also noticed that the biological conclusions or hypothesis have
+	analysis?}} \\
+\answ{The SCORPIUS software was originally developed and the preprint published in 2016.
+Since then, the source code has been modified to be more scalable (LMDS + changes to principal curves), but these changes do not have a noticeable effect on the results of the method.
+The outlier removal was removed from the methodology because this step has since become a standard practice in single-cell omics preprocessing pipelines.} \\
+\task{I will generate a comparison between the old and new version of SCORPIUS on the Schlitzer dataset, and quantify the differences in outcome.}
+
+\exam[\tagimp]{\imp{I also noticed that the biological conclusions or hypothesis have
 	been reduced in claim strength in the paper or have been removed.} It would
@@
 shifted.} \\
-
-\answ{As mentioned, the preprint was originally published in 2016.
-In order to submit this work to a peer-reviewed journal in 2019, the manuscript and the software was rewritten to reflect developments within the field.
-
-Regarding the methodology.
-Classical MDS has indeed been replaced with Landmark MDS, since this allows scaling beyond 10'000 cells.
-Principal curves has been modified to also allow scaling beyond 10'000 cells.
-These changes have little to no effect on any results.
-Outlier removal was removed because this step is now part of mainstream preprocessing pipelines such as Seurat.
-Initialisation of the curve (k-means clustering + shortest path) has been removed -- while this part of the algorithm showed improved performance when applied to the datasets included in the bioRxiv benchmark, this was not the case when applied to a larger set of datasets.
-
-Regarding reducing the biological conclusions.
-Section 5.2.2 regarding the functional modules found by SCORPIUS was reduced in character count, yet it makes the same claims.
+\answ{Section 5.2.2 regarding the functional modules found by SCORPIUS was reduced in character count, yet it makes the same claims.
 The main differences are that gene names and explanations of the biological functions are not discussed in detail, but are instead summarised in figure 5.4.
@@
 	missing from the thesis.}
-\answ{???}
 
 \exam{It would help to list the title of the manuscript on the pre-print server} \\
-\answ{I fixed the DOI. This should be sufficient to find back the manuscript on the pre-print server.}
+\task[\done]{I fixed the DOI. This should be sufficient to find back the manuscript on the pre-print server.}
 
@@
 		such why would bred be a true case-wise NI method and the Scenic approach
-		would not? It would be nice to discuss this more during the defense.}
+		would not? It would be nice to discuss this more during the defense.} \\
+\answ{bred indeed produces a single GRN for every individual sample in the dataset, and this by running more-or-less the same computations as a typical execution of GENIE3 with permutation importances.
+The distinction between bred and SCENIC is that SCENIC infers a single GRN for the whole dataset using GENIE3, and then uses a separate metric to predict to what extent an interaction was active in one of the samples.} \\
+\task{I will expand improve figure 6.5 to better explain the workings of bred and the differences with GENIE3.}
 
@@
 		explicitly why the content was included in the PhD book and what the personal
-		contribution was.}
+		contribution was.} \\
+\answ{This chapter was a collaboration between multiple labs.
+Each lab had independently had the idea of developing a set of guidelines on how to benchmark computational tools.
+Similar to other authors, I had written my own set of guidelines and reasoning behind the guidelines.
+The first author, Lukas M.
+Weber, pooled all of our inputs together and generated a first draft of the manuscript, which was subsequently reviewed and edited by all authors.
+I will not modify the contribution list as modifying it would make it contradict the author contributions list of the publication.} \\
+\task{I will modify the abstract to better discuss and motivate how this chapter fits in the dissertation.}
 
 \exam{Author contributions is not correct (CS)} \\
-\answ{???}
+\answ{Unless I am mistaken, I believe they are correct.}
 
@@
 \exam{\imp{The discussion could be improved by stating how the tools are allowing new
-		biological insights.} Only a small hint is provided in the last sentence of 10.4.}
+		biological insights.} Only a small hint is provided in the last sentence of 10.4.} \\
+\answ{See section~\ref{sec:joeriweakness}.}
 
@@
 
-\exam{Not all chapters have been published and some are still in stage where some work needs to be done before they are publication-ready}
+\exam{Not all chapters have been published and some are still in stage where some work needs to be done before they are publication-ready. }
 
-\exam{Not always a logical connection between the successive chapters}
-
-% TODO: provide visual schematic of what topic this dissertation tries to tackle
+\exam{\imp{Not always a logical connection between the successive chapters}} \\
+\task{I will better discuss the connection between the chapters in the introduction. I will reorder the chapters according to the new overview figure that was featured in the presentation.}
 
@@
 	a users needs to take to use the software, and to illustrate how the results look like}; for now, only
-	the inner workings of the simulator are discussed}
+	the inner workings of the simulator are discussed} \\
+\task{I am actively working on adding more detailed use cases to the results section of this chapter.}
 
@@
 
-\exam[\tagimp]{fundamental question: \imp{how objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?}}
+\exam[\tagimp]{fundamental question: \imp{how objective is a benchmark of a tool, if developers have designed both the tool and the benchmark?}} \\
+\answ{The software and original preprint of tool pre-dates the benchmark.
+Since then, two modifications have been made to improve the scalability without changing the methodology, and one step in the methodology was removed because it was decreasing the stability of the method without improving the accuracy significantly.
+As such, the benchmark did not have a (strong) impact on the methodology of the tool.} \\
+\answ{Throughout the development of the benchmark, we were acutely aware that we could be unintentionally biasing our benchmark to favour SCORPIUS.
+We paid attention to the following matters when designing our benchmark, hoping to have mitigated the largest sources of bias in this way.
+\begin{itemize}
+	\item Our main hypotheses and methodology were determined before implementation of any code or performing any of the experiments.
+	\item When including synthetic datasets we also included synthetic datasets generated by simulators other than dyngen. 
+	\item We performed extensive testing of our metrics to verify that each metric quantifies what we aim for it to quantify.
+	\item When wrapping SCORPIUS vs.
+TI methods from other authors, we could be biased in our interaction with the tool simply because we better understand our own software.
+To mitigate this, we contacted every author of every TI method for comments on our usage of their software and also on the quality control scores that we assigned to their software based on our question checklist.
+Several authors made contributions to the wrappers, and also improved their own software based on certain QC criterion that we proposed.
+\end{itemize}
+}
 
@@
 
-\exam[\tagimp]{\imp{reads more like a summary} + listing some use case that made use of contributions made in the manuscript, rather than a real discussion}
+\exam{\imp{reads more like a summary} + listing some use case that made use of contributions made in the manuscript, rather than a real discussion} \\
+\answ{See section~\ref{sec:joeriweakness}.}
 
@@
 representative the generated data are is of course a huge challenge, this could have been addressed
-perhaps in the Discussion.}
+perhaps in the Discussion.} \\
+\task{I am working on providing use-cases in the results section of this chapter.}
 
@@
 models in the simulation. How crucial is it to use such an extensive model of mRNA production and
-degradation, and how do these parameters affect the simulated single-cell data?}}
+degradation, and how do these parameters affect the simulated single-cell data?}} \\
+%\answ{Includin}
 
@@
 of chapter 2 - here the results are described, but without methods. For example, how are features
-(genes) selected on the branch points, in a branch, etc (which methods are used for this).}
+(genes) selected on the branch points, in a branch, etc (which methods are used for this).} \\
+\task{I will elaborate on the methodology of this chapter.}
 
@@
 \exam[\tagimp \tagtime]{\imp{In Chapter 6, it would have been interesting
-to compare the results of bred with such methods (for example GENIE3 + regulon specificity score).}}
+to compare the results of bred with such methods (for example GENIE3 + regulon specificity score).}} \\
+\task{I will attempt to finish a quantitative comparison between bred and GENIE3 + regulon specificity score.
+If not in the PhD dissertation, this will definitely be included in the publication of this work.}
 
-\exam[\tagimp \tagtime]{\imp{The methods could be a little bit more clear on the specific differences between bred and GENIE3.}}
+\exam[\tagimp \tagtime]{\imp{The methods could be a little bit more clear on the specific differences between bred and GENIE3.}} \\
+\task{I elaborate on the difference between bred and GENIE3 better.}
 
@@
 
-\exam{Is this really the first algorithm for inferring case-wise regulatory networks?}
+\exam{Is this really the first algorithm for inferring case-wise regulatory networks?} \\
+\answ{No, as mentioned in the introduction: "Thus, while several case-wise NI methods have already been proposed, their implementation consisted of post-processing a static GRN to arrive at a case-wise GRN." I thus argue that bred is the first \textit{true} method for inferring case-wise regulatory networks.}
 
@@
 	are nevertheless valid since they offer, relatively easy, a genome-wide view on gene
-	regulation.}
+	regulation.} \\
+\answ{I meant to use the term 'data-driven', rather than 'unbiased'.} \\
+\task{Change 'unbiased' into 'data-driven'.}
 
-\exam{There are significant differences between the methods of Kuijer and Liu, please describe
-	these methods in more detail in the introduction.}
+\exam{There are significant differences between the methods of Kuijer and Liu, please describe these methods in more detail in the introduction.} \\
+\answ{I respectfully disagree.
+In essence, the core concept of how a case-specific GRN is derived is exactly the same (See Kuijjer Fig 1B and Liu Fig 1C), namely by performing classical (static) NI twice, once all of the samples and once more after removing one of the samples.
+The differences between the two NI executions results in the desired case-specific network.} \\
+\task{Regardless, I will revisit the description of these methods and expand the explanation, if necessary.}
 
-\exam{What about case-specific random forests like in the R package ranger, csfr() ?}
+\exam{What about case-specific random forests like in the R package ranger, csfr() ?} \\
+\answ{case-specific random forest is a similar sounding but entirely different concept.
+From the documentation of the \texttt{csrf()} function: In case-specific random forests (CSRF), random forests are built specific to the cases of interest.
+Instead of using equal probabilities, the cases are weighted according to their difference to the case of interest.}
 
 \exam[\tagimp]{The sentence “As such, these methods will most likely recover the interactions that are
-	prevalent in the whole population, and will miss interactions that are specific to only a sub-
-	population” is incorrect. \imp{I would suggest you read the relevant method papers again, see
-	above.}}
+	prevalent in the whole population, and will miss interactions that are specific to only a subpopulation” is incorrect. \imp{I would suggest you read the relevant method papers again, see
+	above.}} \\
+\answ{I respectfully disagree.
+I believe detecting interactions which are only prevalent in a portion of the population will make it harder for classical (static) NI methods to detect them.
+Similar to how it is possible to detect differential expression in a sub-population when using single-cell transcriptomics vs.
+bulk transcriptomics, sample-specific NI will make it easier to detect context-dependent interactions.
+}
+\task{An experiment would need to be performed to provide an answer to this issue. }
 
@@
 
-\exam[\tagimp \tagtime]{\imp{The discussion is very short, please elaborate on}: possible applications in the field of
-	personalized precision medicine and comparison to state of the art.}
+\exam[\tagimp \tagtime]{\imp{The discussion is very short, please elaborate on}: possible applications in the field of personalized precision medicine and comparison to state of the art.} \\
+\task{I elaborate on the implications and application of the method and refer to the current state of the art.}
 
@@
 		but this bridging could be increased further.
-		I miss an overarching discussion on where the field of modelling single cell dynamics is going.}}
+		I miss an overarching discussion on where the field of modelling single cell dynamics is going.}} \\
+\answ{See section~\ref{sec:joeriweakness}.}
 

### 2019-11-25 ch10_discussion/content.tex
update

@@
 Keeping track of the frequencies of graphlets during an iterative process proved difficult at first.
-\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by  assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals)).
+\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by  assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals).
 Under this assumption, it is possible to calculate the differences in topological patterns significantly faster in comparison to recomputing the graphlet frequencies from scratch at every iteration.
@@
 
+% ideas:
+%  Yet I wonder how the current academic system
+%can support such concepts. Academics often hold different positions in
+%different research groups around the world, who should be the maintainer in
+%these cases? The original PhD students that developed the work? The original
+%research group?
+
+%The eight chapter is a manuscript in preparation about the assessment of TI methods.
+The main discussion topic is the lack of a quantitative measure for the scoring of new TI methods.
+This leads to the publication and development of new methods that do not show a clear performance improvement compared to other methods.
+The main topic touches
+%upon some sore points that are inherent in bioinformatics. The main thought I have with
+%this chapter is that defining clear goals in an emerging technology is difficult and therefore is sometimes ignored by fellow bioinformaticians.
+This because the lack of standardized performance metrics, missing public datasets that are well described and an \textit{ad hoc} problem definition.
+I follow the idea of defining a clear cut approach to benchmarking (as described in chapter nine) and I like the proposition that is started in this manuscript for a discussion on good TI metrics.} Would approaches like e.g.
+DREAM challenges be a means of tackling these issues to your opinion?
+
+% zie peter dawyndt, chapter 5
 \clearpage

### 2019-11-28 boek.tex
add back dyngen submodule, apply small changes
Reordered chapters.

@@
 \PassOptionsToPackage{table}{xcolor}
-\documentclass[10pt]{book}
+\documentclass[8pt]{book}
 
@@
 
-% adjust margins (should be done before importing fancyhdr!)
-\usepackage[paperheight=24cm,paperwidth=16cm,margin=2.5cm,a4paper]{geometry} %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm
+\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm]{geometry}
+
+\usepackage[
+	% set width and height to a4 width and height + 6mm
+	%width=16truecm, height=24truecm,
+	a4,
+	% use any combination of these options to add different cut markings
+	cam, axes, frame, cross,
+	% set the type of TeX renderer you use
+	pdftex,
+	% center the contents
+	center
+]{crop}
+
+% adjust margins (should be done before importing fancyhdr!) %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm
 
@@
 \newlength{\thumbwidth}
-\setlength{\thumbwidth}{1.5cm}
+\setlength{\thumbwidth}{1cm}
 
@@
 
-\renderalltrue
+%\renderalltrue
 
 \begin{document}
-
-% thesis obligatory content: 
-% * an abstract in Dutch
-% * a list of abbreviations
-% * one uniform (compiling) reference list
-% * a resume or Curriculum Vitae of the candidate
-% * an introductory chapter and additional linking texts between chapters (papers) if the dissertation consists of a compilation of papers
-% * acknowledgements of financing institutions
-% * apply for an ISBN number for your thesis. This takes 1-2 days.
 
@@
 %{\large \textbf{Abstract}}
-\begin{quote}
+%\begin{quote}
 	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in a high-throughput and unbiased manner.
@@
 As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
-\end{quote}
+%\end{quote}
 
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 1: Benchmarking with \textit{in silico} single cells} 
-\chaptermark{Paper 1: Benchmarking with \textit{in silico} single cells.}
+\chapter{dyngen: Benchmarking with \textit{in silico} single cells} 
+\chaptermark{dyngen: Benchmarking with \textit{in silico} single cells.}
 \label{chap:dyngen}
@@
 	Manuscript in preparation. \\
-	\textbf{Cannoodt R}$^*$, Saelens W$^*$, and Saeys Y.  \\
+	\textbf{Cannoodt R}$^*$, Saelens W$^*$, Deconinck L, and Saeys Y.  \\
 	{\footnotesize $^*$ Equal contribution}
@@
 	$\bullet$ \textbf{R.C.} and W.S. designed the study.\\
-	$\bullet$ \textbf{R.C.} and W.S. performed the experiments and analysed the data.\\
+	$\bullet$ \textbf{R.C.}, W.S., and L.D. performed the experiments and analysed the data.\\
 	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
 	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
-	$\bullet$ \textbf{R.C.} and W.S. reviewed and edited the manuscript. \\
+	$\bullet$ \textbf{R.C.}, W.S., L.D., and Y.S. reviewed and edited the manuscript. \\
 	$\bullet$ Y.S. supervised the project.
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 2: A comparison of single-cell trajectory inference methods} 
-\chaptermark{Paper 2: A comparison of single-cell trajectory inference methods.}
+\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
+\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.}
 \label{chap:dynbenchmark}
@@
 \fi
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%         CH4: DYNO         %%%
+%%%   CH4: Self-assessment    %%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 3: A toolkit for inferring and interpreting trajectories} 
-\chaptermark{Paper 3: A toolkit for inferring and interpreting trajectories.} % short title
-\label{chap:dyno}
-
-{\large \textbf{Abstract}}
-\begin{quote}
-	\textbf{Purpose:} Recent technological advances allow studying dynamic processes by computationally ordering single-cell omics profiles along a trajectory.
-Since 2014, at least 75 tools for trajectory inference have been developed.
-For end-users, however, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.
-\\
-	\textbf{Results:} We developed \texttt{dyno}, a toolkit for inferring, visualising and interpreting single-cell trajectories, giving access to more than 50 TI methods.
-\texttt{dyno} provides downstream analyses such as visualising the trajectory with dot-plots or heatmaps, automated labelling using gene sets, automated inference of the directionality, and detecting genes that are differentially expressed along the trajectory.
-\\
-	\textbf{Conclusion:} By providing a uniform interface for executing and visualising trajectories of 50 methods, \texttt{dyno} allows inferring and comparing trajectories for a much broader audience than previously possible.
-\end{quote}
-
-\vfill
-{\large \textbf{Publication status}}
-\begin{quote}
-	Manuscript in preparation. \\
-	\textbf{Cannoodt R}$^*$, Saelens W$^*$, and Saeys Y.  \\
-	{\footnotesize $^*$ Equal contribution}
-\end{quote}
-
-\vfill
-{\large \textbf{Author contributions}}
-\begin{quote}
-	$\bullet$ \textbf{R.C.} and W.S. designed the project.\\
-	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
-	$\bullet$ \textbf{R.C.} and W.S. wrote the original manuscript.\\
-	$\bullet$ Y.S. supervised the project.
-\end{quote}
-
-\vfill
-
-\ifrenderall
-\newpage
-\graphicspath{{./ch4_dyno/}}
-\include{ch4_dyno/content}
-
-\clearpage
-\section{References}
-\printbibliography[heading=none]
-\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH5: SCORPIUS       %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 4: Fast, accurate, and robust single-cell pseudotime} 
-\chaptermark{Paper 4: Fast, accurate, and robust single-cell pseudotime.} % short title
-\label{chap:scorpius}
-
-{\large \textbf{Abstract}}
-\begin{quote}
-	\textbf{Purpose:} 
-	Recent advances in single-cell RNA-sequencing allow modelling progression of single cells along a dynamic process of interest as a trajectory.
-While already more than 75 trajectory inference methods have been developed, comprehensive benchmarking demonstrated that pseudotemporal ordering cells robustly is still a challenging task.\\
-	\textbf{Results:} \texttt{SCORPIUS} is a trajectory inference method specialised in pseudotemporally ordering single-cell profiles.
-Applying \texttt{SCORPIUS} to a large collection of datasets shows that it consistently produces more accurate and robust trajectories in comparison to state-of-the-art trajectory inference methods.
-We used \texttt{SCORPIUS} to generate novel hypotheses regarding dendritic cell development, which were subsequently validated \textit{in vivo}.\\
-	\textbf{Conclusion:} By providing a robust and accurate toolkit for inferring linear trajectories, we aim to make pseudotime analysis as easy-to-use and prevalent as other types of analysis such as differential expression and gene set enrichment.
-\end{quote}
-
-\vfill
-{\large \textbf{Publication status}}
-\begin{quote}
-	Published in bioRxiv 079509. \doi{10.1101/079509}. \\
-	Revised manuscript in preparation. \\
-	\textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, and Saeys Y. 
-\end{quote}
-
-\vfill
-{\large \textbf{Author contributions}}
-\begin{quote}
-	$\bullet$ \textbf{R.C.} and Y.S. designed the study.\\
-	$\bullet$ \textbf{R.C.} and W.S. benchmarks and analysed the data.\\
-	$\bullet$ S.T. and D.S. generated experimental data.\\
-	$\bullet$ \textbf{R.C.} implemented the SCORPIUS software package.\\
-	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
-	$\bullet$ \textbf{R.C.}, S.J., M.G., B.L., K.D.P., and Y.S. reviewed and edited the manuscript. \\
-	$\bullet$ K.D.P. and Y.S. supervised the project.
-\end{quote}
-
-\vfill
-
-\ifrenderall
-\newpage
-\graphicspath{{./ch5_scorpius/}}
-\include{ch5_scorpius/content}
-\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%         CH6: BRED         %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 5: Inferring single cell regulatory networks} 
-\chaptermark{Paper 5: Inferring single cell regulatory networks.} % short title
-\label{chap:bred}
-
-{\large \textbf{Abstract}}
-\begin{quote}
-	\textbf{Purpose:} Network inference methods are computational tools which use large omics datasets to predict which genes are regulated by which transcription factors.
-	Since regulatory interactions are context-dependent, attempting to model regulatory dynamics in the form of a single regulatory network may have little relevance. \\
-	\textbf{Results:} In this work, we introduce \texttt{bred}, the first algorithm for inferring case-wise regulatory networks.
-We generate .
-Analysing case-wise regulatory networks of 14'963 profiles from The Cancer Genome Atlas project demonstrated that regulatory interactions are often specific to particular cancer types, but can also be shared between several cancer types.
-\\
-	\textbf{Conclusion:} Ultimately, this analysis has resulted in a list of candidate oncogenic genes and interactions per cancer type, though further analysis and validation is required to confirm their oncogenicity.
-\end{quote}
-% TODO: fix abstract
-\vfill
-{\large \textbf{Publication status}}
-\begin{quote}
-	Manuscript in preparation. \\
-	\textbf{Cannoodt R}, Saeys Y$^*$, and De Preter K$^*$. \\
-	{\footnotesize $^*$ Equal contribution}
-\end{quote}
-
-\vfill
-{\large \textbf{Author contributions}}
-\begin{quote}
-	$\bullet$ \textbf{R.C.}, K.D.P., and Y.S. designed the study.\\
-	$\bullet$ \textbf{R.C.} performed the experiments and analysed the data.\\
-	$\bullet$ \textbf{R.C.} implemented the bred software package.\\
-	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
-	$\bullet$ \textbf{R.C.}, K.D.P., and Y.S. reviewed and edited the manuscript. \\
-	$\bullet$ K.D.P. and Y.S. supervised the project.
-\end{quote}
-
-\vfill
-
-\ifrenderall
-\newpage
-\graphicspath{{./ch6_bred/}}
-\include{ch6_bred/content}
-\fi 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%       CH7: INCGRAPH       %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 6: Optimising regulatory networks} 
-\chaptermark{Paper 6: Optimising regulatory networks.} % short title
-\label{chap:incgraph}
-
-{\large \textbf{Abstract}}
-\begin{quote}
-	\textbf{Purpose:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
-As part of a topology optimisation process, graphlet counts are used to iteratively modify a network in order to achieve certain topological properties.
-	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
-\\
-	\textbf{Results:} IncGraph calculates the differences in graphlet counts with respect to the network in its previous state.
-	IncGraph reduces the execution time by several orders of magnitude when compared to recounting the graphlet counts from scratch at every iteration.
-	We demonstrate the usefulness of this approach optimising gene regulatory networks using a graphlet-based fitness score. \\
-	\textbf{Conclusion:} IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
-\end{quote}
-
-\vfill
-{\large \textbf{Publication status}}
-\begin{quote}
-	Published in PLOS ONE 13, 4 (2018). \doi{10.1371/journal.pone.0195997}. \\
-	\textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, and Saeys Y. 
-\end{quote}
-
-\vfill
-{\large \textbf{Author contributions}}
-\begin{quote}
-	$\bullet$ \textbf{R.C.}, Jo.R., Ja.R., and Y.S. designed the study.\\
-	$\bullet$ \textbf{R.C.}, Jo.R., and Ja.R. performed the experiments and analysed the data.\\
-	$\bullet$ \textbf{R.C.} implemented the incgraph software package.\\
-	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
-	$\bullet$ \textbf{R.C.}, Jo.R., Ja.R., K.D.P., and Y.S. reviewed and edited the manuscript. \\
-	$\bullet$ Jo.R., K.D.P. and Y.S. supervised the project.
-\end{quote}
-
-\vfill
-
-\ifrenderall
-\newpage
-\graphicspath{{./ch7_incgraph/}}
-\include{ch7_incgraph/content}
-\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%   CH8: Self-assessment    %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 7: Self-assessment in trajectory inference} 
-\chaptermark{Paper 7: Self-assessment in trajectory inference.} % short title
+\chapter{Self-assessment in trajectory inference} 
+\chaptermark{Self-assessment in trajectory inference.} % short title
 \label{chap:selfassessment}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%       CH5: SCORPIUS       %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{SCORPIUS: Fast, accurate, and robust single-cell pseudotime} 
+\chaptermark{SCORPIUS: Fast, accurate, and robust single-cell pseudotime.} % short title
+\label{chap:scorpius}
+
+{\large \textbf{Abstract}}
+\begin{quote}
+	\textbf{Purpose:} 
+	Recent advances in single-cell RNA-sequencing allow modelling progression of single cells along a dynamic process of interest as a trajectory.
+While already more than 75 trajectory inference methods have been developed, comprehensive benchmarking demonstrated that pseudotemporal ordering cells robustly is still a challenging task.\\
+	\textbf{Results:} \texttt{SCORPIUS} is a trajectory inference method specialised in pseudotemporally ordering single-cell profiles.
+Applying \texttt{SCORPIUS} to a large collection of datasets shows that it consistently produces more accurate and robust trajectories in comparison to state-of-the-art trajectory inference methods.
+We used \texttt{SCORPIUS} to generate novel hypotheses regarding dendritic cell development, which were subsequently validated \textit{in vivo}.\\
+	\textbf{Conclusion:} By providing a robust and accurate toolkit for inferring linear trajectories, we aim to make pseudotime analysis as easy-to-use and prevalent as other types of analysis such as differential expression and gene set enrichment.
+\end{quote}
+
+\vfill
+{\large \textbf{Publication status}}
+\begin{quote}
+	Published in bioRxiv 079509. \doi{10.1101/079509}. \\
+	Revised manuscript in preparation. \\
+	\textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, and Saeys Y. 
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.} and Y.S. designed the study.\\
+	$\bullet$ \textbf{R.C.} and W.S. benchmarks and analysed the data.\\
+	$\bullet$ S.T. and D.S. generated experimental data.\\
+	$\bullet$ \textbf{R.C.} implemented the SCORPIUS software package.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.}, S.J., M.G., B.L., K.D.P., and Y.S. reviewed and edited the manuscript. \\
+	$\bullet$ K.D.P. and Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+\ifrenderall
+\newpage
+\graphicspath{{./ch5_scorpius/}}
+\include{ch5_scorpius/content}
+\fi
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%         CH6: DYNO         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{dyno: A toolkit for inferring and interpreting trajectories} 
+\chaptermark{dyno: A toolkit for inferring and interpreting trajectories.} % short title
+\label{chap:dyno}
+
+{\large \textbf{Abstract}}
+\begin{quote}
+	\textbf{Purpose:} Recent technological advances allow studying dynamic processes by computationally ordering single-cell omics profiles along a trajectory.
+Since 2014, at least 75 tools for trajectory inference have been developed.
+For end-users, however, these methods are difficult to execute and compare, mainly due to high variability in input/output data structures, software requirements, and programming interfaces.
+\\
+	\textbf{Results:} We developed \texttt{dyno}, a toolkit for inferring, visualising and interpreting single-cell trajectories, giving access to more than 50 TI methods.
+\texttt{dyno} provides downstream analyses such as visualising the trajectory with dot-plots or heatmaps, automated labelling using gene sets, automated inference of the directionality, and detecting genes that are differentially expressed along the trajectory.
+\\
+	\textbf{Conclusion:} By providing a uniform interface for executing and visualising trajectories of 50 methods, \texttt{dyno} allows inferring and comparing trajectories for a much broader audience than previously possible.
+\end{quote}
+
+\vfill
+{\large \textbf{Publication status}}
+\begin{quote}
+	Manuscript in preparation. \\
+	\textbf{Cannoodt R}$^*$, Saelens W$^*$, and Saeys Y.  \\
+	{\footnotesize $^*$ Equal contribution}
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.} and W.S. designed the project.\\
+	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
+	$\bullet$ \textbf{R.C.} and W.S. wrote the original manuscript.\\
+	$\bullet$ Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+\ifrenderall
+\newpage
+\graphicspath{{./ch4_dyno/}}
+\include{ch4_dyno/content}
+
+\clearpage
+\section{References}
+\printbibliography[heading=none]
+\fi
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%         CH7: BRED         %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{bred: Inferring single cell regulatory networks} 
+\chaptermark{bred: Inferring single cell regulatory networks.} % short title
+\label{chap:bred}
+
+{\large \textbf{Abstract}}
+\begin{quote}
+	\textbf{Purpose:} Network inference methods are computational tools which use large omics datasets to predict which genes are regulated by which transcription factors.
+	Since regulatory interactions are context-dependent, attempting to model regulatory dynamics in the form of a single regulatory network may have little relevance. \\
+	\textbf{Results:} In this work, we introduce \texttt{bred}, the first algorithm for inferring case-wise regulatory networks.
+We generate .
+Analysing case-wise regulatory networks of 14'963 profiles from The Cancer Genome Atlas project demonstrated that regulatory interactions are often specific to particular cancer types, but can also be shared between several cancer types.
+\\
+	\textbf{Conclusion:} Ultimately, this analysis has resulted in a list of candidate oncogenic genes and interactions per cancer type, though further analysis and validation is required to confirm their oncogenicity.
+\end{quote}
+% TODO: fix abstract
+\vfill
+{\large \textbf{Publication status}}
+\begin{quote}
+	Manuscript in preparation. \\
+	\textbf{Cannoodt R}, Saeys Y$^*$, and De Preter K$^*$. \\
+	{\footnotesize $^*$ Equal contribution}
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.}, K.D.P., and Y.S. designed the study.\\
+	$\bullet$ \textbf{R.C.} performed the experiments and analysed the data.\\
+	$\bullet$ \textbf{R.C.} implemented the bred software package.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.}, K.D.P., and Y.S. reviewed and edited the manuscript. \\
+	$\bullet$ K.D.P. and Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+\ifrenderall
+\newpage
+\graphicspath{{./ch6_bred/}}
+\include{ch6_bred/content}
+\fi 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%       CH8: INCGRAPH       %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage{\thispagestyle{empty}\cleardoublepage}
+\chapter{incgraph: Optimising regulatory networks} 
+\chaptermark{incgraph: Optimising regulatory networks.} % short title
+\label{chap:incgraph}
+
+{\large \textbf{Abstract}}
+\begin{quote}
+	\textbf{Purpose:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
+As part of a topology optimisation process, graphlet counts are used to iteratively modify a network in order to achieve certain topological properties.
+	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
+\\
+	\textbf{Results:} IncGraph calculates the differences in graphlet counts with respect to the network in its previous state.
+	IncGraph reduces the execution time by several orders of magnitude when compared to recounting the graphlet counts from scratch at every iteration.
+	We demonstrate the usefulness of this approach optimising gene regulatory networks using a graphlet-based fitness score. \\
+	\textbf{Conclusion:} IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
+\end{quote}
+
+\vfill
+{\large \textbf{Publication status}}
+\begin{quote}
+	Published in PLOS ONE 13, 4 (2018). \doi{10.1371/journal.pone.0195997}. \\
+	\textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, and Saeys Y. 
+\end{quote}
+
+\vfill
+{\large \textbf{Author contributions}}
+\begin{quote}
+	$\bullet$ \textbf{R.C.}, Jo.R., Ja.R., and Y.S. designed the study.\\
+	$\bullet$ \textbf{R.C.}, Jo.R., and Ja.R. performed the experiments and analysed the data.\\
+	$\bullet$ \textbf{R.C.} implemented the incgraph software package.\\
+	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.}, Jo.R., Ja.R., K.D.P., and Y.S. reviewed and edited the manuscript. \\
+	$\bullet$ Jo.R., K.D.P. and Y.S. supervised the project.
+\end{quote}
+
+\vfill
+
+\ifrenderall
+\newpage
+\graphicspath{{./ch7_incgraph/}}
+\include{ch7_incgraph/content}
+\fi
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%       CH9: Guidelines    %%%
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Paper 8: Essential guidelines for computational method benchmarking} 
-\chaptermark{Paper 8: Essential guidelines for computational method benchmarking.} % short title
+\chapter{Essential guidelines for computational method benchmarking} 
+\chaptermark{Essential guidelines for computational method benchmarking.} % short title
 \label{chap:guidelines}

### 2019-11-28 chX_backmatter.tex
removed dyngen submodule
Added conferences, oral presentations, poster presentations and master student supervisions to backmatter.

@@
 	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
-	%	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y.
-SCORPIUS: Fast, accurate, and robust single-cell pseudotime.
-In preparation.
-	%	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
-	%	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
-	%	\item \textbf{Cannoodt R}, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
-	%	\item \textbf{Cannoodt R}. Self-assessment in trajectory inference. In preparation.
 \end{itemize}
@@
 
+\section{Conferences and meetings}
+
+\begin{itemize}
+  \item \textbf{Differential Network Medicine}, Antwerp, Belgium, 4-6 December 2013.
+  \item \textbf{BeNeLux Bioinformatics Conference}, Brussels, Belgium, 9-10 December 2013.
+  \item \textbf{OncoPoint}, Ghent, Belgium, 6 February 2014.
+  \item \textbf{AISTATS}, Reykjavik, Iceland, 22-25 April 2014.
+  \item \textbf{N2N Annual Symposium}, Ghent, Belgium, 21 May 2014.
+  \item \textbf{Bioinformatics N2N Conference}, Ghent, Belgium, 21 May 2014.
+  \item \textbf{BENELEARN}, Brussels, Belgium, 6 June 2014.
+  \item \textbf{BeNeLux Bioinformatics Conference}, Luxembourg, 8-9 December 2014.
+  \item \textbf{OncoPoint}, Ghent, Belgium, 11 February 2015.
+  \item \textbf{Big Data to Bedside}, Ghent, Belgium, 1-2 April 2015.
+  \item \textbf{BIG N2N symposium}, Ghent, Belgium, 21 May 2015.
+  \item \textbf{BeNeLux Bioinformatics Conference}, Antwerp, Belgium, 7-8 December 2015
+  \item \textbf{Single Cell Biology Workshop}, Ghent, Belgium, 15 January 2016.
+  \item \textbf{Single Cell Biology}, Hinxton, United Kingdom, 8-10 March 2016.
+  \item \textbf{BIG N2N symposium}, Ghent, Belgium, 19 May 2016.
+  \item \textbf{CYTO}, Seattle, United States of America, 11-15 June 2016.
+  \item \textbf{BENELEARN}, Kortrijk, Belgium, 12-13 September 2016.
+  \item \textbf{Single Cell Genomics}, Hinxton, United Kingdom, 14-16 September 2016.
+  \item \textbf{Single Cell Biology Workshop}, Ghent, Belgium, 15 January 2016.
+  \item \textbf{VIB Seminar}, Veldhoven, Netherlands, 27-28 April 2017.
+  \item \textbf{Keystone Symposia: Single Cell Omics}, Stockholm, Zweden, 26-30 May 2017. 
+  \item \textbf{BeNeLux Bioinformatics Conference}, Louvain, België, 13-14 December 2017.
+  \item \textbf{Single Cell Biology}, Hinxton, United Kingdom, 8-10 March 2018.
+  \item \textbf{Keystone Symposia: Single Cell Biology}, Colorado, USA, 13-17 January 2019.
+  \item \textbf{Human Cell Atlas}, Toronto, Canada, 29-31 July 2019.
+\end{itemize}
+
+\section{Courses / workshops}
+
+\begin{itemize}
+  \item \textbf{Differential Network Medicine}, Antwerp, Belgium, 4-6 September 2013.
+  \item \textbf{Basics of Biology for Engineers}, Louvain, Belgium, 18-20 September 2013.
+  \item \textbf{Metric Learning}, Louvain, Belgium, 15 October 2013.
+  \item \textbf{Machine learning Summer School}, Reykjavik, Iceland, 25 April - 4 May 2014.
+  \item \textbf{Effective Oral Presentations by Jean-Luc Doumont}, Ghent, Belgium, 3 February 2015.
+  \item \textbf{RNA-Seq analysis for differential expression}, Ghent, Belgium, 24-27 April 2015.
+  \item \textbf{BigData@UGent in practice}, Ghent, Belgium, 4 May 2015.
+\end{itemize}
+
+\section{Oral presentations}
+
+\begin{itemize}
+  \item Cannoodt R., Ruyssinck J., De Preter K., Dhaene T., Saeys Y.
+: Network inference by integrating biclustering and feature selection.
+\textbf{BeNeLux Bioinformatics Conference}, Brussels, Belgium, 9-10 December 2013.
+  \item Cannoodt R., Ruyssinck J., De Preter K., Dhaene T., Saeys Y.
+: Network inference by integrating biclustering and feature selection.
+\textbf{IRC bioinformatics day}, Ghent, Belgium, 23 November 2013.
+  \item Cannoodt R., De Preter K., Saeys Y. : Differential network inference for pediatric cancers. \textbf{Maestra meeting}, Ghent, Belgium, 14 March 2014.
+  \item Cannoodt R., Beckers A., Van Cauwenbergh C., Speleman F., Saeys Y., De Preter K.
+: Differential module analysis in neuroblastoma regulatory networks.
+\textbf{Oncopoint}, Ghent, Belgium, 11 February 2015.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.: Inferring developmental chronologies from single cell RNA, \textbf{BeNeLux Bioinformatics Conference}, Antwerp, Belgium, 7-8 December 2015.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.: Inferring trajectories along dynamic processes from single-cell RNA-seq data.
+\textbf{Jean-Luc Doumont workshop}, Ghent, Belgium, 17 December 2015.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.
+: SCORPIUS: Inferring trajectories along dynamic processes form single-cell RNA-seq data.
+\textbf{Single Cell Biology Workshop}, Ghent, Belgium, 15 January 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.
+: Improving marker gene discovery from high-dimensional single-cell snapshot data.
+\textbf{CYTO}, Seattle, United States of America, 11-15 June 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. \textbf{BENELEARN}, Kortrijk, Belgium, 12-13 September 2016.
+  \item Cannoodt R., Saelens W., Sichien D., Tavernier S., Janssens S., Guilliams M., Lambrecht B., De Preter K., Saeys Y.
+: Unbiased modelling of dynamic processes with SCORPIUS identifies novel modules in dendritic cell development.
+\textbf{VIB Seminar}, Veldhoven, Nederland, 27-28 april 2017.
+  \item Cannoodt R., Saelens W. : Automated building and unit testing, Docker and Singularity. \textbf{VIB Developers Meeting}, Ghent, Belgium, 25 January 2019.
+  \item Cannoodt R., Saelens W., Todorov H., Saeys Y.
+: dynbenchmark, Assessing Accuracy, Robustness and Usability of Single-Cell Trajectory Inference methods.
+\textbf{Keystone Symposia: Single Cell Biology}, Colorado, USA, 13-17 January 2019.
+\end{itemize}
+
+\section{Poster presentations}
+
+\begin{itemize}
+  \item Cannoodt R., Ruyssinck J., De Preter K., Dhaene T., Saeys Y.: Network Inference by Integrating Biclustering and Feature Selection. \textbf{N2N Annual Symposium}, Ghent, Belgium, 21 May 2014.
+  \item Cannoodt R., Van Cauwenbergh C., Beckers A., Speleman F., De Preter K., Saeys Y.: Differential Module Analysis in Neuroblastoma Regulatory Networks.
+\textbf{BeNeLux Bioinformatics Conference} 2014, Luxembourg, Belgium, 8-9 December 2014.
+  \item Cannoodt R., Beckers A., Van Cauwenbergh C., Speleman F., De Preter K., Saeys Y.: Differential Module Analysis in Neuroblastoma Regulatory Networks, \textbf{BIG N2N symposium}, Ghent, Belgium, 21 May 2015.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. Single Cell Biology, Hinxton, United Kingdom, 8-10 March 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.
+: Unbiased modelling of dynamic processes with single-cell RNA-sequencing.
+\textbf{BIG N2N symposium}, \textbf{BIG N2N symposium}, Ghent, Belgium, 19 May 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Improving marker gene discovery from high-dimensional single-cell snapshot data. CYTO, Seattle, United States of America, 11-15 June 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. BENELEARN 2016, Kortrijk, Belgium, 12-13 September 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. Single Cell Genomics, Hinxton, United Kingdom, 14-16 September 2016.
+  \item Cannoodt R., Saelens W., Sichien D., Tavernier S., Janssens S., Guilliams M., Lambrecht B., De Preter K., Saeys Y.: Unbiased modelling of dynamic processes with SCORPIUS identifies novel modules in dendritic cell development.
+\textbf{VIB Seminar}, Veldhoven, Netherlands, 27-28 April 2017.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.: True single cell network inference: Modelling gene regulation of individual cells.
+\textbf{Keystone Symposia: Single Cell Omics}, Stockholm, Zweden, 26-30 May 2017.
+  \item Cannoodt R., Saelens W., Todorov H., Saeys Y.: Generalised framework for and comparison of 24 trajectory inference methods.
+\textbf{BeNeLux Bioinformatics Conference}, Louvain, België, 13-14 December 2017.
+  \item Cannoodt R., Saelens W., Todorov H., Saeys Y.
+: A comparison of single-cell trajectory inference methods: towards more accurate and robust tools.
+\textbf{Single Cell Biology}, Hinxton, United Kingdom, 8-10 March 2018.
+  \item Cannoodt R., Saelens W.
+: dynbenchmark, Assessing Accuracy, Robustness and Usability of Single-Cell Trajectory Inference methods.
+\textbf{Keystone Symposia: Single Cell Biology}, Colorado, USA, 13-17 January 2019.
+\end{itemize}
+
+\section{Master student supervision}
+
+\begin{itemize}
+  \item \textbf{Leen De Baets}. Identificatie van nieuwe kankergenen voor neuroblastoomonderzoek met machine learning. September 2013 -- June 2014.
+  \item \textbf{Wouter Saelens}. Locale cel-type specifieke genexpressie in het myeloïde transcriptoom. September 2013 -- June 2014.
+  \item \textbf{Charlotte De Vogelaere}. Quantitative evaluation of network inference methods for single-cell cancer regulomes. September 2015 -- June 2016.
+  \item \textbf{Sofie Veys}. Comparative review of dimensionality reduction methods for high-throughput single-cell transcriptomics. September 2016 -- June 2017.
+  \item \textbf{Chloë Guidi}. Het afleiden van dynamische grafen op basis van snapshot data. September 2016 -- June 2017.
+  \item \textbf{Jarre Knockaert}. Inferentie van cellontwikkelingstrajecten met machine learning. September 2018 -- June 2019.
+\end{itemize}
+
 \section{Open-source software}
@@
 We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends.
-This includes speeding up parts of the dependency (slingshot), implementing new functionality (devtools, ParamHelpers, ranger, rlang), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes).
-Several of these packages are mainstream R packages which receive millions of downloads per year (devtools, ranger, remotes, rlang).
+This includes speeding up parts of the dependency (slingshot), implementing new functionality (devtools, ParamHelpers, ranger, rlang), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes, tidyverse).
+Several of these packages are mainstream R packages which receive millions of downloads per year (devtools, ranger, remotes, rlang, tidyverse).
 
@@
 		\biocpkg{splatter} & ctb & Bioc & 5015 & Simple simulation of single-cell RNA sequencing data \\
+		\cranpkg{tidyverse} & ctb & CRAN &  & Easily install and load packages from the tidyverse \\
 		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & Reconstructing branching trajectories from single-cell RNAseq data \\
@@
 
-%\subsection{Oral presentations}
-%
-%\subsection{Poster presentations}
-%
-%\subsection{Awards}
-%
-%\subsection{Courses / Workshops}
-%
-%\subsection{Conferences}
-%
-%\subsection{Student supervision}
-%
-%\subsection{Teaching}
-
 \section{Sources of funding}
 Robrecht Cannoodt was supported by the Fonds Wetenschappelijk Onderzoek (11Y6218N).
+
+% TODO: copyright
 

### 2019-11-28 ch10_discussion/content.tex
small update discussion

@@
 This chapter reflects on the impact of this work on the field, going over each of the objectives and respective chapters.
+
+\section{Trajectory inference}
+
+\section{Network inference}
+
+\section{Benchmarking in bioinformatics}
 
@@
 However, there are many benefits of using R in bioinformatics research context, such as its extensive community of statisticians develop packages for the R ecosystem.
-The \textbf{tidyverse} packages \cite{wickham_tidytoolsmanifesto_2017}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.
+The \textbf{tidyverse} packages \cite{wickham_welcometidyverse_2019}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.
 

### 2019-11-28 ch1_introduction/analysis/overview.R
add overview figure

@@ -1,0 +1,67 @@
+library(tidyverse)
+library(scales)
+
+set.seed(1)
+
+df <- tribble(
+  ~x, ~y, ~name, ~hjust, ~vjust,
+  0, 0, "§2: dyngen", .5, 1,
+  .4, 1, "§3: dynbenchmark", 0, .5,
+  .6, 1, "§4: self-assessment\nin trajectory inference", 0, .5,
+  .8, 1, "§5: SCORPIUS", 0, .5,
+  1, 1, "§6: dyno", 0, .5,
+  1.8, 1, "§7: bred", 1, .5,
+  2, 1, "§8: incgraph", 1, .5,
+  3, 1, "§9: guidelines\nfor benchmarking", .5, 0
+)
+titdf <- tibble(
+  x = 0:2,
+  y = c(1.6, 1.8, 1.8),
+  name = c("Benchmarking", "Trajectory\ninference", "Network\ninference"),
+  short = c("Benchmarking", "TI", "NI"),
+  col = RColorBrewer::brewer.pal(3, "Set1"),
+  hjust = c(.5, 0, 1)
+)
+
+rgbs <- grDevices::col2rgb(titdf$col)
+mix_cols <- function(df) {
+  rgb <-
+    map2(df$x, df$y, function(x, y) {
+      case_when(
+        y == 0 ~ c(1, 1, 1) / 3,
+        x <= 1 ~ c(1 - x, x, 0),
+        x <= 2 ~ c(0, 2 - x, x - 1),
+        TRUE ~ c(x - 2, 0, 3 - x)
+      )
+    }) %>% do.call(rbind, .)
+  rgb2 <- round(rgbs %*% t(rgb))
+  df$col <- grDevices::rgb(rgb2[1,], rgb2[2,], rgb2[3,], maxColorValue = 255)
+  df
+}
+
+df <- df %>% mix_cols()
+
+arrdf <- tribble(
+  ~gr, ~x, ~y,
+  2, .4, 1,
+  2, .76, 1,
+  3, 1, 1,
+  3, 1.575, 1,
+  4, 2, 1,
+  4, 2.5, 1
+)
+
+g <- ggplot(df) +
+  geom_segment(aes(x = x, xend = x, y = 0, yend = 1.05), titdf, colour = "lightgray", linetype = "dashed") +
+  geom_path(aes(x, y, group = gr), arrdf, arrow = grid::arrow(type = "closed", length = grid::unit(6, "mm"), angle = 25), col = "lightgray") +
[…]

### 2019-12-05 boek.tex
~

@@
 
-% use Sonny chapter style
-\usepackage[Sonny]{fncychap} %Sonny
+\usepackage{titlesec}
+\newcommand{\hsp}{\hspace{20pt}}	
+\titleformat{\chapter}[hang]{\LARGE\bfseries}{\thechapter\hsp\textcolor{light-gray}{|}\hsp}{0pt}{\LARGE\bfseries}
 
-% set font after chapter sonny style
+% set fonts
 \usepackage{fontspec}
@@
 	colorlinks = true,
-%	linkcolor = {blue},
-  urlcolor = {blue},
+	linkcolor = {black},
+  urlcolor = {black},
 	citecolor = {med-gray}
@@
 
-%\renderalltrue
+\renderalltrue
 
@@
 \label{chap:dynbenchmark}
-
+\vspace*{-3em}
 {\large \textbf{Abstract}}
@@
 \label{chap:scorpius}
+\vspace*{-4em}
 
@@
 	\textbf{Purpose:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
-As part of a topology optimisation process, graphlet counts are used to iteratively modify a network in order to achieve certain topological properties.
+Counting graphlet can used to iteratively optimise a network in order to achieve certain topological properties.
 	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
@@
 \begin{quote}
-  Published in Genome Biology 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}. \\
+  Published in Genome Biology 20 (2019). \doi{10.1186/s13059-019-1738-8}. \\
 	Weber LM, Saelens W, \textbf{Cannoodt R}, Soneson C, Hapfelmeier A, Gardner PP, Boulesteix AL, Saeys Y, and Robinson MD 

### 2019-12-05 ch0_frontmatter.tex
added github & orcid
~

@@
 }
+
+\newcommand{\orcid}[1]{	\includegraphics[width=3mm]{fig/icons/orcid.png}\;\href{https://orcid.org/#1}{{https://orcid.org/}{\color{black}#1}}}
+\newcommand{\github}[1]{	\includegraphics[width=3mm]{fig/icons/github.png}\;\href{https://github.com/#1}{{https://github.com/}{\color{black}#1}}}
 
@@
 
-	{\LARGE \textsf{Robrecht Cannoodt}}
+	{\LARGE \textsf{Robrecht Cannoodt}} \\
+	\orcid{0000-0003-3641-729X} \\
+	\github{rcannood}
 
@@
 
-\chapter*{Summary}
-\vspace{-2em}
-
+\newpage
+\ 
+\vfill
+{\bfseries\LARGE Summary}
+\vfill
 Recent developments in the \textit{single-cell omics} technologies triggered a major milestone in the field of (cell)biology.
@@
 
-\chapter*{Samenvatting}
-\vspace{-8em}
+\newpage
+%\ 
+%\vfill
+{\bfseries\LARGE Samenvatting}
+%\vfill
 

### 2019-12-05 chX_backmatter.tex
~

@@
 	Address & Kerkstraat 75, 9070 Destelbergen \\
-	Email & \href{mailto:robrecht.cannoodt@gmail.com}{robrecht.cannoodt@gmail.com}
+	Email & \href{mailto:robrecht.cannoodt@gmail.com}{robrecht.cannoodt@gmail.com} \\
+	Webpage & \url{https://www.cannoodt.dev}
 \end{tabular}
@@
 \textit{Laboratory assistant (summer job)} \\
-\hspace*{2 ex} Bloedtransfusiecentrum Gent, Belgium, July (2009 -- 2012)
+\hspace*{2 ex} Bloedtransfusiecentrum Gent, Belgium, July (2009, 2010, 2011, 2012)
 
@@
 	\caption{\textbf{Contibutions to open-source software.} Following abbreviations denote the relation with respect to the package: \textit{aut} Author, \textit{ctb} Contributor.
-Yearly download statistics are based on the number of downloads between 2019-10-01 and 2019-10-21.
+Yearly download statistics are based on the number of downloads between 2019-10-01 and 2019-11-28.
 CRAN download statistics are retrieved from the Rstudio CRAN mirror only; other CRAN mirrors do not track download statistics.
@@
 
-	\centering\fontsize{9}{11}\selectfont
+	\centering\fontsize{7}{9}\selectfont
 	\begin{tabularx}{\linewidth}{|p{2cm}llp{1.5cm}X|}
@@
 		Name & Role & Host & Downloads per year & Description \\ \hline\hline
-		\cranpkg{babelwhale} & aut & CRAN & 9800 & Interacting with Docker and Singularity containers \\
-		\cranpkg{diffusionMap} & aut & CRAN & 27'850 & Implements diffusion map method of data parameterization \\
+		\cranpkg{babelwhale} & aut & CRAN & 6110 & Interacting with Docker and Singularity containers \\
+		\cranpkg{diffusionMap} & aut & CRAN & 30'123 & Implements diffusion map method of data parameterization \\
 		\githubpkg{dynverse}{dynbenchmark} & aut & Github & \notavailable & Pipeline for benchmarking trajectory inference methods \\
-		\cranpkg{dyndimred} & aut & CRAN & 4708 & Applying dimensionality reduction methods \\
+		\cranpkg{dyndimred} & aut & CRAN & 5116 & Applying dimensionality reduction methods \\
 		\githubpkg{dynverse}{dyneval} & aut & Github & \notavailable & Evaluating trajectory inference methods \\
@@
 		\githubpkg{dynverse}{dyno} & aut & Github & \notavailable & A pipeline for inferring, visualising and interpreting trajectories \\
-		\cranpkg{dynparam} & aut & CRAN & 2810 & Creating meta-information for parameters \\
+		\cranpkg{dynparam} & aut & CRAN & 3816 & Creating meta-information for parameters \\
 		\githubpkg{dynverse}{dynplot} & aut & Github & \notavailable & A simple visualisation library for trajectories \\
@@
 		\githubpkg{dynverse}{dyntoy} & aut & Github & \notavailable & Generating simple toy data of cellular differentiation \\
-		\cranpkg{dynutils} & aut & CRAN & 32'668 & Common functionality for the dynverse packages \\
-		\cranpkg{dynwrap} & aut & CRAN & 4636 & A common format for trajectories \\
-		\cranpkg{GillespieSSA} & aut & CRAN & 7118 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
-		\cranpkg{GillespieSSA2} & aut & CRAN & 4950 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
+		\cranpkg{dynutils} & aut & CRAN & 16'999 & Common functionality for the dynverse packages \\
+		\cranpkg{dynwrap} & aut & CRAN & 4009 & A common format for trajectories \\
+		\cranpkg{GillespieSSA} & aut & CRAN & 7763 & Gillespie's Stochastic Simulation Algorithm (SSA) \\
+		\cranpkg{GillespieSSA2} & aut & CRAN & 4181 & Gillespie's Stochastic Simulation Algorithm for Impatient People \\
 		\githubpkg{dynverse}{gng} & aut & Github & \notavailable & Growing Neural Gas implemented in Rcpp \\
-		\cranpkg{incgraph} & aut & CRAN & 3565 & Incremental graphlet counting for network optimisation \\
-		\cranpkg{lmds} & aut & CRAN & 23'' & Landmark Multi-Dimensional Scaling \\
-		\cranpkg{princurve} & aut & CRAN & 29'100 & Fits a principal curve in arbitrary dimension \\
-		\cranpkg{proxyC} & aut & CRAN & 117'480 & Computes proximity in large sparse matrices \\
-		\cranpkg{qsub} & aut & CRAN & 3585 & Running commands remotely on gridengine clusters \\
-		\cranpkg{SCORPIUS} & aut & CRAN & 4520 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
+		\cranpkg{incgraph} & aut & CRAN & 3570 & Incremental graphlet counting for network optimisation \\
+		\cranpkg{lmds} & aut & CRAN & 1742 & Landmark Multi-Dimensional Scaling \\
+		\cranpkg{princurve} & aut & CRAN & 28'869 & Fits a principal curve in arbitrary dimension \\
+		\cranpkg{proxyC} & aut & CRAN & 122'858 & Computes proximity in large sparse matrices \\
+		\cranpkg{qsub} & aut & CRAN & 3622 & Running commands remotely on gridengine clusters \\
+		\cranpkg{SCORPIUS} & aut & CRAN & 4285 & Inferring developmental chronologies from single-cell RNA sequencing data \\ \hline\hline
 
-		\cranpkg{badger} & ctb & CRAN & 6240 & Query information and generate badge for using in README \\
+		\cranpkg{badger} & ctb & CRAN & 6472 & Query information and generate badge for using in README \\
 		\biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 935 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
 		\cranpkg{devtools} & ctb & CRAN & 5'918'700 & Tools to make developing R packages easier \\
+		\cranpkg{ggrepel} & ctb & CRAN & 2'018'030 & Repel overlapping text labels away from each other \\
 		\githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & Reconstructing lineage-tree topologies from scRNA-seq data \\
@@
 		\githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of high dimensional single cell data \\
+		\cranpkg{shadowtext} & ctb & CRAN & 6822 & shadow text for grid and ggplot2 \\
 		\biocpkg{slingshot} & ctb & Bioc & 12'085 & Tools for ordering single-cell sequencing \\
 		\biocpkg{splatter} & ctb & Bioc & 5015 & Simple simulation of single-cell RNA sequencing data \\
-		\cranpkg{tidyverse} & ctb & CRAN &  & Easily install and load packages from the tidyverse \\
+		\cranpkg{tidyverse} & ctb & CRAN & 5'079'398 & Easily install and load packages from the tidyverse \\
 		\githubpkg{farrelja}{URD} & ctb & Github & \notavailable & Reconstructing branching trajectories from single-cell RNAseq data \\

### 2019-12-05 ch1_introduction/analysis/download_stats.R

@@ -4,7 +4,7 @@ library(tidyverse)
 cran <- c(
   "babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "dynwrap",
   "GillespieSSA", "GillespieSSA2", "incgraph", "lmds", "princurve", "proxyC",
-  "qsub", "SCORPIUS", "badger", "devtools", "mlr", "ParamHelpers", "ranger", "remotes", "rlang"
+  "qsub", "SCORPIUS", "badger", "devtools", "ggrepel", "mlr", "ParamHelpers", "ranger", "remotes", "rlang", "shadowtext", "tidyverse"
 )
 bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
 
@@ -12,9 +12,9 @@ dl_cran <- dlstats::cran_stats(cran)
 dl_bioc <- dlstats::bioc_stats(bioc)
 
 dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>%
-  mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
+  mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", "shadowtext", "ggrepel", "tidyverse", bioc), "Contributor", "Author"))
 
-min_date <- "2019-08-01"
+min_date <- "2019-10-01"
 num_days <- as.integer(Sys.Date() - as.Date(min_date))
 summ <- dls %>% filter(min_date <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% slice(match(cran, package))
 summ %>% print(n = 50)

### 2019-12-05 ch6_bred/scripts/2_tcga.R

@@ -294,11 +294,16 @@ if (!file.exists(paste0(data_dir, "importances.rds"))) {
     qsub_handle,
     wait = "just_do_it",
     post_fun = function(i, li) {
-      li$importance <- li$importance %>% filter(importance > .01)
+      li$importance <- li$importance %>% filter(importance > .05)
       li$importance_sc <- li$importance_sc %>% filter(importance_sc > .01) %>%
         inner_join(li$importance %>% select(regulator, target), by = c("regulator", "target"))
       li
     }
+  )
+
+  out <- list(
+    importance = map_df(out, "importance"),
+    importance_sc = map_df(out, "importance_sc")
   )
 
   write_rds(out, paste0(data_dir, "importances.rds"))
@@ -307,14 +312,9 @@ if (!file.exists(paste0(data_dir, "importances.rds"))) {
 list2env(read_rds(paste0(data_dir, "importances.rds")), .GlobalEnv)
 
 # fixes for old results
-assertthat::assert_that(all(sample_info$old_id %in% levels(importance_sc$cell_id)))
-importance <- importance %>% select(-i) %>% rename(interaction_id = name)
+importance <- importance %>% rename(interaction_id = name)
 importance_sc <-
   importance_sc %>%
-  rename(old_id = cell_id) %>%
-  inner_join(sample_info %>% transmute(sample_id = factor(id), old_id = factor(old_id, levels = levels(importance_sc$cell_id))), "old_id") %>%
-  filter(!is.na(sample_id)) %>%
-  select(-i) %>%
   left_join(importance %>% select(regulator, target, interaction_id), by = c("regulator", "target"))
 
 if (!file.exists(paste0(data_dir, "dimred_and_clustering.rds"))) {

### 2019-12-05 ch7_incgraph/content.tex
fix fontsize for page layout
~

@@
 %% FIGURE 2
-\picfig{fig/figure2.pdf}{\LARGEfigure}{fig_efficient}{{\bf Several strategies are employed in order to reduce time and memory consumption.}
+\picfig{fig/figure2.pdf}{.7\linewidth}{fig_efficient}{{\bf Several strategies are employed in order to reduce time and memory consumption.}
 (A) Only the depth 3 neighbourhood of each modified edge needs to be explored in order to have visited all modified five-node graphlets.
@@
 %% FIGURE 4
-\picfig{fig/figure4.pdf}{\hugefigure}{fig_method_reranking}{{\bf Predicted gene regulatory networks of model organisms are optimised to reduce the false positive rate.}
+\picfig{fig/figure4.pdf}{.9\linewidth}{fig_method_reranking}{{\bf Predicted gene regulatory networks of model organisms are optimised to reduce the false positive rate.}
 	A)~The number of redundant edges in each graphlet are counted.
@@
 
-\section{Supporting information}
+\section{Supplemental information}
 
@@
 \begin{figure}[htb!]
+	\small
 	\begin{algorithmic}
@@
 \begin{figure}[htb!]	
+	\small
 	\begin{algorithmic}

### 2019-12-06 boek.tex
updates

@@
 
-\renderalltrue
+%\renderalltrue
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%   CH4: Self-assessment    %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Self-assessment in trajectory inference} 
-\chaptermark{Self-assessment in trajectory inference.} % short title
-\label{chap:selfassessment}
-
-{\large \textbf{Abstract}}
-\begin{quote}
-	\textbf{Purpose:} A brief of 75 articles of trajectory tools reveals 
-	that only 37\% of works contain a quantitative self-assessment. 
-	While self-assessments are universally biased in favour of the
-	authors, it is dangerous to publish a computational tool
-	without quantitatively demonstrating its performance compared to
-	state-of-the-art methods. \\
-	\textbf{Results:} We hypothesise that low self-assessment rates are
-	primarily caused by a lack of a standardised problem definition, readily
-	available benchmarking datasets, and suitable metrics. We elaborate on
-	these causal reasons and provide viable solutions for 
-	performing benchmarks more easily.\\
-	\textbf{Conclusion:} With this perspective, we hope to address the
-	elephant in the room, and provide 
-	developers of trajectory inference methods with the necessary 
-	tools for performing benchmarking software quantitatively. 
-\end{quote}
-
-\vfill
-{\large \textbf{Publication status}}
-\begin{quote}
-	Manuscript in preparation. \\
-	\textbf{Cannoodt R.}
-\end{quote}
-
-\vfill
-{\large \textbf{Author contributions}}
-\begin{quote}
-	$\bullet$ \textbf{R.C.} designed the study.\\
-	$\bullet$ \textbf{R.C.} surveyed 75 manuscripts from literature and analysed the data.\\
-	$\bullet$ \textbf{R.C.} wrote the original manuscript.
-\end{quote}
-
-\vfill
-
-\ifrenderall
-\newpage
-\graphicspath{{./ch8_selfassessment/}}
-\include{ch8_selfassessment/content}
-\clearpage
-\section{References}
-\printbibliography[heading=none]
-\fi
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%   CH4: Self-assessment    %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\newpage{\thispagestyle{empty}\cleardoublepage}
+%\chapter{Self-assessment in trajectory inference} 
+%\chaptermark{Self-assessment in trajectory inference.} % short title
+%\label{chap:selfassessment}
+%
+%
+%{\large \textbf{Abstract}}
+%\begin{quote}
+%	\textbf{Purpose:} A brief of 75 articles of trajectory tools reveals 
+%	that only 37\% of works contain a quantitative self-assessment. 
+%	While self-assessments are universally biased in favour of the
+%	authors, it is dangerous to publish a computational tool
+%	without quantitatively demonstrating its performance compared to
+%	state-of-the-art methods. \\
+%	\textbf{Results:} We hypothesise that low self-assessment rates are
+%	primarily caused by a lack of a standardised problem definition, readily
+%	available benchmarking datasets, and suitable metrics. We elaborate on
+%	these causal reasons and provide viable solutions for 
+%	performing benchmarks more easily.\\
+%	\textbf{Conclusion:} With this perspective, we hope to address the
+%	elephant in the room, and provide 
+%	developers of trajectory inference methods with the necessary 
+%	tools for performing benchmarking software quantitatively. 
+%\end{quote}
+%
+%\vfill
+%{\large \textbf{Publication status}}
+%\begin{quote}
+%	Manuscript in preparation. \\
+%	\textbf{Cannoodt R.}
+%\end{quote}
+%
+%\vfill
+%{\large \textbf{Author contributions}}
+%\begin{quote}
+%	$\bullet$ \textbf{R.C.} designed the study.\\
+%	$\bullet$ \textbf{R.C.} surveyed 75 manuscripts from literature and analysed the data.\\
+%	$\bullet$ \textbf{R.C.} wrote the original manuscript.
+%\end{quote}
+%
+%\vfill
+%
+%\ifrenderall
+%\newpage
+%\graphicspath{{./ch8_selfassessment/}}
+%\include{ch8_selfassessment/content}
+%\clearpage
+%\section{References}
+%\printbibliography[heading=none]
+%\fi
 

### 2019-12-06 ch0_frontmatter.tex
updates

@@
 
-	{\LARGE \textsf{Robrecht Cannoodt}} \\
-	\orcid{0000-0003-3641-729X} \\
-	\github{rcannood}
+	{\LARGE \textsf{Robrecht Cannoodt}} % \\
+%	\orcid{0000-0003-3641-729X} \\
+%	\github{rcannood}
 
@@
 
-	\includegraphics[height=.7cm]{fig/logos/icon_UGent_WE_EN_RGB_color}\hfill
-	\includegraphics[height=1cm]{fig/logos/vib_rf_inflammation_research_rgb_pos} \hfill
-	\includegraphics[height=.8cm]{fig/logos/dambi_logo} \hfill
-	\includegraphics[height=.7cm]{fig/logos/cmgg_myshort} \hfill
-	\includegraphics[height=.7cm]{fig/logos/FWO_Logo_KleurCMYK} 
+	\includegraphics[width=\linewidth]{fig/logos/mashup}
+%	\includegraphics[height=.7cm]{fig/logos/icon_UGent_WE_EN_RGB_color}\hfill
+%	\includegraphics[height=1cm]{fig/logos/vib_rf_inflammation_research_rgb_pos} \hfill
+%	\includegraphics[height=.8cm]{fig/logos/dambi_logo} \hfill
+%	\includegraphics[height=.7cm]{fig/logos/cmgg_myshort} \hfill
+%	\includegraphics[height=.7cm]{fig/logos/FWO_Logo_KleurCMYK} 
 	\vspace{1cm}

### 2019-12-06 ch1_introduction/content.tex
updates

@@
 	\caption{
-		\textbf{A:} Convergence of single-cell and omics biology.
+		\textbf{A:} Technological improvements in microfluidics and low-volume sequencing were instrumental in the convergence of the fields of single-cell biology and omics biology.
 		\textbf{B:} Different approaches for sampling cells with decreasing levels of cellular heterogeneity within the different sub-populations: snapshot, time-series, sorted.
@@
 
-scRNA-seq data is typically very sparse -- while the human genome has more than 20'000 genes, they only contain non-zero values of a few thousand genes (typically <4'000).
+scRNA-seq data is typically very sparse -- while the human genome has more than 20'000 genes, they only contain non-zero values of a few thousand genes (typically less than 4'000).
 This is partially due to cells being specialised in particular functions and thus they do not need proteins of every time, but also due to RNA transcription occurring in bursts rather than continuously \cite{nicolas_whatshapeseukaryotic_2017}.
@@
 
-Over the past five years, already 450 new tools have been developed to perform various analyses of single-cell omics data \cite{zappia_exploringsinglecellrnaseq_2018}, taking into account the specific noise characteristics.
-The most frequent types of analyses are detailed in the following subsections.
+Over the past five years, already 450 new tools have been developed to perform various analyses of single-cell omics data \cite{zappia_exploringsinglecellrnaseq_2018}.
+Most of these tools belong to one (or more) of the five main classes of computational methods used within single-cell omics (Figure~\ref{fig:scapplications}).
+The main concepts and applications of these types of analyses are discussed in the following sections.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\linewidth]{fig/singlecell_technologies_onlycore} 
+	\caption{
+		\textbf{Five main types of analyses typically performed with single-cell omics.}
+	}
+	\label{fig:scapplications}
+\end{figure}
 
@@
 The dimensionality of such datasets is typically too high for humans to interpret manually and for most modelling algorithms to tackle directly.
-Moreover, in reality, the intrinsic dimensionality of biological systems is probably much lower.
+Moreover, in reality, the intrinsic dimensionality of biological systems is much lower.
 For example, a differentiating hematopoietic cell could be described by just three dimensions: the first two dimensions lays out the hematopoietic lineage tree, and a third dimension allows for reprogramming between branches to occur.
@@
 Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful low-dimensional representation.
+DR methods are typically one of the first methods to apply on the data as a visual quality control by asserting that cells group together in ways that are expected.
+
 DR methods have two main target audiences; computers -- to construct a K-dimensional (with K $\ll$ N) representation of the data such that pairwise distances between different samples are retained as well as possible (Figure~\ref{fig:comp_tools_dimred}B); and humans -- to obtain a visual overview of the data (Figure~\ref{fig:comp_tools_dimred}C).
+DR methods are also commonly used to denoise the data, which is particularly useful since single-cell omics data is inherently noisy.
 
@@
 \subsection{Clustering}
-To learn about the different cellular states within a population of cells, clustering methods divide up the cells into separate groups of highly similar cells (Figure~\ref{fig:comp_tools_cluster}A). 
+Clustering methods divide up the cells into separate groups of highly similar cells in order to delineate cell populations (Figure~\ref{fig:comp_tools_cluster}A).
 By visualising gene expression of known genes involved in the cell types of interest, the clusters can be annotated (Figure~\ref{fig:comp_tools_cluster}B).
+Particular clustering methods might be developed to better deal with imbalanced data in order to better identify rare cell type populations.
 
@@
 While clustering methods divide cells into distinct groups, trajectory inference (TI) methods acknowledge that cells are dynamic entities which transition from once cellular state to another via various dynamic processes.
-Rather than making distinct groups, TI methods allow studying dynamic processes by reconstructing the topology of a dynamic process as a trajectory, and map the cells onto that trajectory.
+Rather than making distinct groups, TI methods attempt to predict how cells develop or undergo dynamic processes.
+The main applications of trajectory inference are to identify cellular states and branchpoints, reconstruct the topology of dynamic processes, and identify state-specific markers.
+
+TI methods try to reconstruct the topology of a dynamic process as a trajectory and subsequently map the cells onto that trajectory.
 A trajectory is a graph where the nodes represent noteworthy cellular states, and each cell is predicted to be progressing along transitions between the different states (Figure~\ref{fig:comp_tools_ti}A).
@@
 A trajectory can be visualised as a graph to better highlight the topology of the trajectory(Figure~\ref{fig:comp_tools_ti}A middle), as a heatmap to better depict the changes in gene expression along the different transitions (Figure~\ref{fig:comp_tools_ti}A right), or may be embedded in a dimensionality reduction of the cells to better demonstrate cellular heterogeneity along the trajectory (Figure~\ref{fig:comp_tools_ti}B right).
-Similar to clustering, by colouring the cells according to the expression of genes known to be involved in the dynamic process of interest, the milestones in the trajectory can be annotated (Figure~\ref{fig:comp_tools_cluster}B).
+Similar to clustering, by colouring the cells according to the expression of genes known to be involved in the dynamic process of interest, the milestones in the trajectory can be annotated (Figure~\ref{fig:comp_tools_ti}B).
 
@@
 
-A lot of TI methods use similar algorithms to be able to infer a trajectory.
+Groups of TI methods use similar algorithms to infer a trajectory.
 By breaking down each method into its set of core algorithms, we can create a map of TI methodology \cite{cannoodt_computationalmethodstrajectory_2016} (Figure~\ref{fig:ti_metro}A), which is divided into two major classes.
@@
 \subsection{Differential expression}
-Given that cells are split up two or more groups, differential expression (DE) methods rank genes based on whether their expression is significantly higher or lower in one group in comparison to the others.
-This grouping can be based on prior information or an upstream clustering method.
-DE methods are useful for summarising the main differences between different groups of cells more compactly (Figure~\ref{fig:comp_tools_de}A) in comparison to when groups are compared without gene prioritisation (Figure~\ref{fig:comp_tools_cluster}A).
+Given that cells are split up two or more groups -- either through clustering or prior knowledge -- differential expression (DE) methods attempt to identify a group of genes that are expressed significantly higher or lower in one group in comparison to the others.
+Main applications of DE methods are the development of population signatures or simply the visualisation of the main difference between several groups in a compact manner (Figure~\ref{fig:comp_tools_de}A).
 
@@
 One of the central cellular processes underlying development is transcriptional regulation. 	
-Modelling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
+Modelling gene regulation is essential to better understand what drives cells to develop in the way that they do and to identify what goes wrong in the case of disease.
 
-Network inference (NI) methods predict which genes are regulated by which transcription factors (Figure~\ref{fig:comp_tools_ni}).
-The output of a network inference is thus a graph, where nodes represent genes and edges denote a regulatory interaction between a regulator and a target gene.
-Interactions typically have two properties: its regulatory strength (a positive real value) and its effect (promoting or repressing).
+Network inference (NI) methods predict for every gene (or a subset thereof) by which transcription factors they are regulated (Figure~\ref{fig:comp_tools_ni}).
+The output of a network inference is a graph in which nodes represent genes and edges denote a regulatory interaction between a regulator and a target gene.
+Interactions can have two properties: its regulatory strength (a positive real value) and its effect (promoting or repressing).
 
@@
 \end{figure}
+
+\section{Benchmarking in bioinformatics}
+
+Many articles introducing novel computational tools for analysing single-cell omics data
+lack a quantitative assessment of the accuracy of the method. Instead, they
+rely on anecdotal evidence to demonstrate their added value.
+
+As a demonstration, we perform a brief review of 75 articles 
+introducing new TI methods, but the findings presented are
+generalisable to single-cell omics in general.
+Out of 75 articles, only about 37\% of articles contain a
+self-assessment (Figure \ref{fig:benchmarks_over_time}).
+Peer-reviewed articles fared even worse, self-assessing in only 35\% of
+cases (n=55), whereas articles first published as a pre-print
+self-assess in 44\% of cases (n=39).
+
+While self-assessments are universally biased in favour of the
+authors \cite{norel_selfassessmenttrapcan_2011} (intentionally or
+not), it is dangerous and unusual to publish a computational tool
+without quantitatively demonstrating its performance compared to
+state-of-the-art methods. Indeed, our comparison demonstrated that most
+methods perform worse than a few baseline methods constructed by
+combining simple off-the-shelf algorithms such as PCA, \(k\)-means and
+MST.
+
+We hypothesise that low self-assessment rates are
+primarily caused by a lack of a standardised problem definition, readily
+available benchmarking datasets, and suitable metrics.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.75\linewidth]{fig/selfassessment/self_assessment.pdf}
+	\caption{
+		\textbf{Less than half of all TI articles perform quantitative self-assessment.}
+		\textbf{A:} Since 2016, the number of TI articles has been increasing rapidly. Note that TI methods with both a pre-print and a peer-reviewed article only count once in the overall tally.
+		\textbf{B:} Less than 50\% of articles feature a self-assessment. Peer-reviewed articles self-assess only in 34\% of cases.
+	}
+	\label{fig:benchmarks_over_time}
+\end{figure}
+
+\subsection{Problem definition}
+
+One main reason why benchmarking TI methods is difficult is due to there being slight variations of the problem a method is attempting to solve.
+For example, a TI method might infer linear or cyclic trajectories, or predict the probability of a cell
+ending up in one of several end states. As a result, it becomes harder to discover similar methods to compare against, as certain TI articles might only show up with search terms such
+as ``pseudotemporal ordering'', ``lineage trees'' or ``fate bias''. For
+the discoverability of a new TI method, it is therefore essential to use
+the term ``trajectory inference'', or at least list it as one of the
+keywords.
+
+\subsection{Datasets} 
+During the infancy of the relatively domain 
+of single-cell omics, large well-annotated datasets that can be used to 
+benchmark novel computational tools are scarce and expensive to generate. 
+When adequate benchmarking datasets are lowly abundant,
+ synthetic data is often used to evaluate
+computational methods, either standalone or to complement real
+data.
+
+Benefits of synthetic data are that they offer more control over the
+data characteristics and that they can be generated in large quantities.
+This allows evaluating the performance of a method in function of a
+changing parameter (e.g.~dataset size or noise levels), which provides
+information on how well the method will work on real datasets.
+
+A common counterargument of synthetic data is that they generate
+unrealistic datasets and thus provide no additional value in evaluating
+a method. In contrast, we argue that a good set of synthetic datasets
+should allow benchmarkers to verify that a method should
+\textit{at least} work well on the synthetic datasets, but good
+performance on synthetic datasets does not guarantee good performance on
+real datasets.
+
+\subsection{Metrics}
+
+To evaluate a computational tools, a quantitative metric is needed to compare 
+the often complex multi-layered data structures of the ground-truth dataset
+to predictions thereof, or to compare the data structures of multiple predictions.
+When adequate metrics are lacking, it is advisable to re-purpose metrics from other 
+domains and adapt them, if necessary.
+
+A second important criterion is to evaluate the robustness of predictions by comparing
+multiple executions of the same
+method. Computing the robustness does not replace the necessity of a
+relevant metric that captures whether a predicted trajectory resembles
+the ground truth.
 
@@
 
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=.75\linewidth]{fig/overview.pdf}
+	\caption{Overview}
+	\label{fig:overview_thesis}
+\end{figure}
+
 \begin{itemize}
@@
 	\item We provide a tool for \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
-	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
-	Reproducibility problems of TI methods are partially due to low rates of quantitative self-assessment.
+%	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
+%	Reproducibility problems of TI methods are partially due to low rates of quantitative self-assessment.
 We look at 75 manuscripts of TI methods and identify key challenges the authors had in benchmarking their tool.
@@
 
+\section{Hippocratic oath for method developers}
+Before setting out on this perilous journey, I write a Hippocratic oath 
+which I believe all developers of computational tools should swear by 
+before haphazardly creating new tools.
+
+\begin{figure}[htb!]
+	\centering
+	\begin{tcolorbox}
+		\begin{itemize}
+			\item In developing a new computational tool, I swear to fulfil, to the best of my ability, the following items:
+			\item I will write software that works reliably but fails gracefully when it does not.
+			\item I will assert that my method produces accurate and reproducible results.
+			\item I will validate the performance of my tool with previously unseen data from third-party sources.
+			\item I will use suitable quantitative performance metrics. If possible, I use multiple metrics to avoid overfitting.
+			\item I will compare the performance to relevant methods.
+			\item I will report positive achievements of my method, but also be critical of it, and discuss areas in which it performs suboptimally.
+			\item I will write unit tests to ensure that each component in my tool works as intended.
+If the tool produces unexpected errors when tested on various edge cases, my tool will be perceived as dysfunctional by others.
+			\item I will follow reproducible research best practices, by making code and data publicly available.
+		\end{itemize}
+	\end{tcolorbox}
+	\caption{
+		\textbf{A hippocratic oath for computational method developers.} Inspired by Laplante\cite{laplante_firstnoharm_2004} and lists of guidelines for benchmarking\cite{norel_selfassessmenttrapcan_2011,weber_essentialguidelinescomputational_2019}
+	}
+	\label{fig:hippocratic_oath}
+\end{figure}
+
 \clearpage

### 2019-12-06 ch1_introduction/analysis/overview.R

@@ -7,12 +7,12 @@ df <- tribble(
   ~x, ~y, ~name, ~hjust, ~vjust,
   0, 0, "§2: dyngen", .5, 1,
   .4, 1, "§3: dynbenchmark", 0, .5,
-  .6, 1, "§4: self-assessment\nin trajectory inference", 0, .5,
-  .8, 1, "§5: SCORPIUS", 0, .5,
-  1, 1, "§6: dyno", 0, .5,
-  1.8, 1, "§7: bred", 1, .5,
-  2, 1, "§8: incgraph", 1, .5,
-  3, 1, "§9: guidelines\nfor benchmarking", .5, 0
+  # .6, 1, "§4: self-assessment\nin trajectory inference", 0, .5,
+  .7, 1, "§4: SCORPIUS", 0, .5,
+  1, 1, "§5: dyno", 0, .5,
+  1.8, 1, "§6: bred", 1, .5,
+  2, 1, "§7: incgraph", 1, .5,
+  3, 1, "§8: guidelines\nfor benchmarking", .5, 0
 )
 titdf <- tibble(
   x = 0:2,
@@ -44,7 +44,7 @@ df <- df %>% mix_cols()
 arrdf <- tribble(
   ~gr, ~x, ~y,
   2, .4, 1,
-  2, .76, 1,
+  2, .6, 1,
   3, 1, 1,
   3, 1.575, 1,
   4, 2, 1,
@@ -60,8 +60,8 @@ g <- ggplot(df) +
   geom_text(aes(x, y, label = name, colour = col), titdf, size = 6, fontface = "bold") +
   coord_polar() +
   dynplot::theme_graph() +
-  theme(plot.margin=grid::unit(c(0,0,0,0), "mm")) +
+  theme(plot.margin=margin(0,0,0,0)) +
   scale_colour_identity()
-ggsave("fig/overview.pdf", g, width = 6, height = 6)
+ggsave("fig/overview.pdf", g, width = 6, height = 4)
 
 

### 2019-12-07 ch1_introduction/content.tex
update intro

@@
 	\caption{
-		\textbf{Five main types of analyses typically performed with single-cell omics.}
+		\textbf{Five main types of analyses typically performed with single-cell omics, and the applications thereof.}
 	}
@@
 
-\section{Benchmarking in bioinformatics}
+\section{Benchmarking computational tools}
 
-Many articles introducing novel computational tools for analysing single-cell omics data
-lack a quantitative assessment of the accuracy of the method. Instead, they
-rely on anecdotal evidence to demonstrate their added value.
+Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology.
+Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication.
+Hundreds of new software tools were developed \cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g.
+differential expression, dimensionality reduction, normalisation).
 
-As a demonstration, we perform a brief review of 75 articles 
-introducing new TI methods, but the findings presented are
-generalisable to single-cell omics in general.
-Out of 75 articles, only about 37\% of articles contain a
-self-assessment (Figure \ref{fig:benchmarks_over_time}).
-Peer-reviewed articles fared even worse, self-assessing in only 35\% of
-cases (n=55), whereas articles first published as a pre-print
-self-assess in 44\% of cases (n=39).
+One major shortcoming during the advent of single-cell omics was that the majority of newly developed computational tools were not quantitatively and comparatively evaluated.
+Rather, they relied on anecdotal evidence to demonstrate its usefulness.
 
-While self-assessments are universally biased in favour of the
-authors \cite{norel_selfassessmenttrapcan_2011} (intentionally or
-not), it is dangerous and unusual to publish a computational tool
-without quantitatively demonstrating its performance compared to
-state-of-the-art methods. Indeed, our comparison demonstrated that most
-methods perform worse than a few baseline methods constructed by
-combining simple off-the-shelf algorithms such as PCA, \(k\)-means and
-MST.
+As a demonstration, we perform a brief review of 75 articles introducing new TI methods, but the findings presented are generalisable to single-cell omics in general.
+Out of 75 articles, only about 37\% of articles contain a self-assessment (Figure \ref{fig:benchmarks_over_time}).
+Peer-reviewed articles fared even worse, self-assessing in only 35\% of cases, whereas articles first published as a pre-print self-assess in 44\% of cases.
 
-We hypothesise that low self-assessment rates are
-primarily caused by a lack of a standardised problem definition, readily
-available benchmarking datasets, and suitable metrics.
+While self-assessments are universally biased in favour of the authors \cite{norel_selfassessmenttrapcan_2011} (intentionally or not), it is dangerous and unusual to publish a computational tool without quantitatively demonstrating its performance compared to state-of-the-art methods.
+Uncontrolled development of software tools without comprehensive benchmarking poses serious problems.
+For one, it slows down scientific progress.
+Every end-user needs to make a large commitment researching the domain in order to make an informed decision of which tool to use, or risk a higher incidence of false positive discoveries (either way, valuable resources are being wasted).
+In addition, it also negatively impacts the credibility of the field, thus discouraging potential users or researchers from entering.
+
+This issue is likely not due to the tool developer's malevolence, but instead due to the absence of benchmarking methodology for this application, which can include standardised problem definitions, readily available benchmarking datasets, and suitable metrics.
+Each of these topics is discussed in more detail below.
 
@@
 \subsection{Problem definition} 
+One main reason why benchmarking TI methods is difficult is due to there being slight variations of the problem a method is attempting to solve.
+For example, a TI method might infer linear or cyclic trajectories, or predict the probability of a cell ending up in one of several end states.
 
-One main reason why benchmarking TI methods is difficult is due to there being slight variations of the problem a method is attempting to solve.
-For example, a TI method might infer linear or cyclic trajectories, or predict the probability of a cell
-ending up in one of several end states. As a result, it becomes harder to discover similar methods to compare against, as certain TI articles might only show up with search terms such
-as ``pseudotemporal ordering'', ``lineage trees'' or ``fate bias''. For
-the discoverability of a new TI method, it is therefore essential to use
-the term ``trajectory inference'', or at least list it as one of the
-keywords.
+These differences reflect themselves in the input and output interfaces of each method, making it harder to directly compare methods directly.
+By constructing a generalised version that encompasses the different sub-problems and wrapping each method accordingly allows to fix the interface to make methods directly comparable.
+
+Additionally, differences in problem definitions might complicate knowledge discovery to find similar methods, as certain articles might only show up with specific search terms.
+For the discoverability of a new TI method, it is therefore essential to use include generalised terminology, or at least list it as one of the keywords.
 
 \subsection{Datasets} 
-During the infancy of the relatively domain 
-of single-cell omics, large well-annotated datasets that can be used to 
-benchmark novel computational tools are scarce and expensive to generate. 
-When adequate benchmarking datasets are lowly abundant,
- synthetic data is often used to evaluate
-computational methods, either standalone or to complement real
-data.
+During the infancy of the relatively domain of single-cell omics, large well-annotated datasets that can be used to benchmark novel computational tools are scarce and expensive to generate.
+When adequate benchmarking datasets are lowly abundant, synthetic data is often used to evaluate computational methods, either standalone or to complement real data.
 
-Benefits of synthetic data are that they offer more control over the
-data characteristics and that they can be generated in large quantities.
-This allows evaluating the performance of a method in function of a
-changing parameter (e.g.~dataset size or noise levels), which provides
-information on how well the method will work on real datasets.
+Benefits of synthetic data are that they offer more control over the data characteristics and that they can be generated in large quantities.
+This allows evaluating the performance of a method in function of a changing parameter (e.g.~dataset size or noise levels), which provides information on how well the method will work on real datasets.
 
-A common counterargument of synthetic data is that they generate
-unrealistic datasets and thus provide no additional value in evaluating
-a method. In contrast, we argue that a good set of synthetic datasets
-should allow benchmarkers to verify that a method should
-\textit{at least} work well on the synthetic datasets, but good
-performance on synthetic datasets does not guarantee good performance on
-real datasets.
+A common counterargument of synthetic data is that they generate unrealistic datasets and thus provide no additional value in evaluating a method.
+In contrast, we argue that a good set of synthetic datasets should allow benchmarkers to verify that a method should \textit{at least} work well on the synthetic datasets, but good performance on synthetic datasets does not guarantee good performance on real datasets.
 
 \subsection{Metrics}
+To evaluate a computational tools, a quantitative metric is needed to compare  the often complex multi-layered data structures of the ground-truth dataset to predictions thereof, or to compare the data structures of multiple predictions.
+When adequate metrics are lacking, it is advisable to re-purpose metrics from other domains and adapt them, if necessary.
 
-To evaluate a computational tools, a quantitative metric is needed to compare 
-the often complex multi-layered data structures of the ground-truth dataset
-to predictions thereof, or to compare the data structures of multiple predictions.
-When adequate metrics are lacking, it is advisable to re-purpose metrics from other 
-domains and adapt them, if necessary.
-
-A second important criterion is to evaluate the robustness of predictions by comparing
-multiple executions of the same
-method. Computing the robustness does not replace the necessity of a
-relevant metric that captures whether a predicted trajectory resembles
-the ground truth.
+A second important criterion is to evaluate the robustness of predictions by comparing multiple executions of the same method.
+Computing the robustness does not replace the necessity of a relevant metric that captures whether a predicted trajectory resembles the ground truth.
 
 \section{Research context and objectives} \label{sec:research_objectives}
-Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology.
-Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication.
-Hundreds of new software tools were developed \cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g.
-differential expression, dimensionality reduction, normalisation).
+Technological breakthroughs in single-cell omics profiling have resulted in a wealth of information on the biomolecular contents of individual cells.
+In this work, we aim to speed up scientific discovery by developing and benchmarking computational tools for generating high-quality hypotheses in a high-throughput and data-driven manner.
 
-One major shortcoming during the advent of single-cell omics was that the majority of newly developed computational tools were not quantitatively and comparatively evaluated.
-Rather, they relied on anecdotal evidence to demonstrate its usefulness.
-This issue is not the result of the tool developer's malevolence, but instead of the lack of data required to perform such comprehensive benchmarks.
-%TODO: also mention guidelines for users.
+The objectives of this work can be reduced to three main questions. How do cells change over time? What drives cells to change? How good are these predictions?
 
-Uncontrolled development of software tools without comprehensive benchmarking poses serious problems. 
-For one, it slows down scientific progress.
-Every end-user needs to make a large commitment researching the domain in order to make an informed decision of which tool to use, or risk a higher incidence of false positive discoveries (either way, valuable resources are being wasted).
-In addition, it also negatively impacts the credibility of the field, thus discouraging potential users or researchers from entering.
-
-In this work, we aim to speed up scientific progress in single-cell omics by providing tools both for end-users and developers alike.
-For developers of computational approaches, we provide tools and guidelines for benchmarking their method on real and synthetic data.
-For end-users we develop new tools and guidelines for analysing dynamic processes by inferring trajectories and gene regulatory networks.
-Our specific scientific objectives are discussed in the following chapters.
+We tackle these three questions using three respective approaches, namely trajectory inference, network inference, and benchmarking (Figure~\ref{fig:overview_thesis}).
+Each of the chapters in this work contributes to answering a part of one or more of these questions.
 
@@
 
-\begin{itemize}
-	\item We \textbf{develop benchmarking strategies} for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}).
+\paragraph{Trajectory inference} We develop a simulator of developing single cells (Chapter~\ref{chap:dyngen}).
+We use synthetic data to complement real data in benchmarking TI methods in terms of accuracy, stability, scalability and usability (Chapter~\ref{chap:dynbenchmark}).
+Based on these results, we provide guidelines to end-users on how to perform trajectory inference analysis.
+We introduce a novel TI method specialised in inferring linear trajectories which outperforms state-of-the-art existing methods (Chapter~\ref{chap:scorpius}).
+Finally, we provide a full-stack toolkit for performing inferring, visualising and interpreting trajectories from more than 50 different TI methods (Chapter~\ref{chap:dyno}).
+
+\paragraph{Network inference} We explore a novel algorithm for inferring gene regulatory interactions at a single-cell level (Chapter~\ref{chap:bred}).
+We demonstrate our NI methodology by inferring regulatory interactions for 14'963 profiles of cancer patients.
+We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+
+\paragraph{Benchmarking} Using our synthetic single-cell generator, we develop benchmarking strategies for assessing the performance of novel types of single-cell omics tools which are currently in a conceptual or prototypical state (Chapter~\ref{chap:dyngen}).
+We summarise our experience in benchmarking computational methods by providing a list of essential guidelines in how to benchmark computational tools (Chapter~\ref{chap:guidelines}).
+
+%\begin{itemize}
+%	\item We \textbf{develop benchmarking strategies} for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}).
 \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
-	\item We apply this strategy to \textbf{perform a comparison of TI methods} (Chapter~\ref{chap:dynbenchmark}).
+%	\item We apply this strategy to \textbf{perform a comparison of TI methods} (Chapter~\ref{chap:dynbenchmark}).
 Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking.
@@
 Our pipeline, datasets, metrics, and containerised wrappers of TI methods are made publicly available for developers to use.
-	\item We develop \textbf{a toolkit to infer, visualise and interpret single-cell trajectories} using more than 50 different TI methods (Chapter~\ref{chap:dyno}).
+%	\item We introduce a TI method specialised in \textbf{inferring linear trajectories} (Chapter~\ref{chap:scorpius}).
+Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
+%	\item We develop \textbf{a toolkit to infer, visualise and interpret single-cell trajectories} using more than 50 different TI methods (Chapter~\ref{chap:dyno}).
 This toolkit provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the trajectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g.
 adding directionality or adding annotation).
-	\item We introduce a TI method specialised in \textbf{inferring linear trajectories} (Chapter~\ref{chap:scorpius}).
-Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
-	\item We explore a novel algorithm for \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
+%	\item We explore a novel algorithm for \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}).
 We demonstrate our NI methodology by inferring regulatory interactions for 14'963 profiles of cancer patients.
-	\item We provide a tool for \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
-%	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
-%	Reproducibility problems of TI methods are partially due to low rates of quantitative self-assessment.
-We look at 75 manuscripts of TI methods and identify key challenges the authors had in benchmarking their tool.
-For each of these challenges, we provide a summary of how we and others have successfully tackled these challenges, in order to spur developers to perform more self-assessments.
-	\item Finally, we \textbf{summarise our experience in benchmarking computational methods} by providing a list of essential guidelines in benchmarking computational tools (Chapter~\ref{chap:guidelines}).
-\end{itemize}
+%	\item We provide a tool for \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
+%	\item Finally, we \textbf{summarise our experience in benchmarking computational methods} by providing a list of essential guidelines in benchmarking computational tools (Chapter~\ref{chap:guidelines}).
+%\end{itemize}
 
+\newpage 
 \section{Hippocratic oath for method developers}

### 2019-12-09 ch10_discussion/content.tex
~
updates
work on discussion

@@
 
-In this work, we aimed to accelerate scientific progress in the field of single-cell omics by providing and evaluating computational tools.
-In section~\ref{sec:research_objectives} we proposed a concrete set of objectives this work will attempt to solve.
-Every chapter respectively tackles one of these research objectives, and each chapter discusses the implications of that part of our research in detail.
-This chapter reflects on the impact of this work on the field, going over each of the objectives and respective chapters.
+\section{Impact of this work}
+In this work, we made scientific contributions to the field of computational biology applied to single-cell omics.
+More specifically, we developed approaches for performing trajectory inference and network inference analyses for single-cell omics, as well as approaches for assessing the performance of such methods in a quantitative way.
 
-\section{Trajectory inference}
+The contribution with the largest impact in the field is the large-scale comparison of 45 TI methods.
+Based on our results, we provided guidelines on how to perform a TI analysis and developed a toolkit for performing a inferring and interpreting trajectories using any of these methods.
+Since such guidelines were hitherto lacking, they are now commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018, luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019, martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.
+These disseminations are having a significant impact in how TI analyses are being performed in academia and industry alike.
 
-\section{Network inference}
+This dissertation commences with a rant on low self-assessment rates of method developers.
+As a result, we developed a simulator of \textit{in silico} single cells, that allow quantifying the accuracy the prediction of a tool, even if the real data required to perform this analysis does not exist yet.
+While the manuscript is not published yet, dyngen has already been used to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory-based differential expression \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
 
-\section{Benchmarking in bioinformatics}
+% has used dyno detomaso_functionalinterpretationsingle_2019
 
-\section{Benchmarking with \textit{in silico} single cells}
+\section{Outlook}
+The types of computational analyses presented in Section~\ref{sec:comptools} (Clustering, DE, DR, TI, NI) all become routine methodologies for analysing and interpreting single-cell omics data.
+Here, we discuss several promising categories of computational methods that are currently still in a exploratory stage (Figure~\ref{fig:scapplications_extended}).
 
-We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
-The \texttt{dyngen} software supports a selection of existing experimental protocols, and can also simulate experiments that are not possible with the current technologies, such as generating omics profiles from the same cell at multiple timepoints.
-The generated data can help kick-start emerging domains with low data availability more safely by allowing software developers to test their method before numerous datasets become publicly available.
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\linewidth]{fig/singlecell_technologies_v8.pdf}
+	\caption{Developments in computational methods for single-cell omics. Adapted from presentations by Wouter Saelens and myself.}
+	\label{fig:scapplications_extended}
+\end{figure}
 
-\texttt{dyngen} has already been successfully used to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
-Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.
+\subsection{Trajectory differential expression}
+Trajectory inference methods have made transformative changes in single-cell omics by allowing to study how cells change during a dynamic process of interest in a high-throughput and unsupervised approach.
+A crucial aspect in analysing and interpreting the resulting trajectories is the discovery of genes whose expression significantly shifts in a region of interest within the trajectory, called Trajectory Differential Expression (TDE).
 
-% Eventueel kort ook beschrijven waar je denkt dat het nog van toepassing kan zijn? The use if dyngene is not limited to.... but might also,....
-% Eventueel ook: waar zie je nog verbeter/uitbreidingspunten in je methode. 
-% Wat zijn de huidige limitaties. Bv. Is er een tijdgebonden limitatie. bv. Stel je wilt data van 10.000 cellen genereren maar bent een maand bezig, etc.
+Several trajectory inference methods offer TDE functionality as part of downstream analysis and interpretation of the outputted trajectories \cite{cannoodt_scorpiusimprovestrajectory_2016,qiu_reversedgraphembedding_2017,lonnberg_singlecellrnaseqcomputational_2017,wolf_pagagraphabstraction_2019}.
+However, current approaches typically cluster cells and perform differential expression analyses between the clusters, or are otherwise limited towards detecting differential expression of a gene along linearly ordered cells along one lineage path in the trajectory.
 
-\section{Comparing 45 trajectory inference methods}
-Using this simulator and a collection of real datasets, we performed a comparison of 45 TI methods (Chapter~\ref{chap:dynbenchmark}).
-Our contributions include writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, and developing novel metrics for comparing ground truth and predicted trajectories.
-With this benchmarking study, we had two clear goals in mind: to help guide users to TI methods that are suitable for their needs, and to provide developers of TI methods with the necessary tools to benchmark their own tool.
+tradeSeq \cite{vandenberge_trajectorybaseddifferentialexpression_2019} is a generalised TDE method which can be used to discover different types of differential expression along a trajectory, which can be applied as a downstream analysis to trajectories inferred from any TI method.
+By including a comparative benchmark of TDE methods on real and synthetic data, Van den Berge et al.
+started a rational dialogue on TDE methodology.
 
-We accomplished the first goal by constructing a set of guidelines for end-users to help choose a TI method that is accurate, robust and fit for their application. 
-Since such guidelines were hitherto lacking, they are now commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.
+\subsection{Trajectory alignment}
+Trajectory alignment allows studying the differences between multiple trajectories that are mostly similar.
+For example, the cell developmental process of a patient could be compared to that of a healthy control to detect the transcriptomic differences of a particular lineage branch.
 
-For our second goal, it seems too early to tell whether our research had any effect on self-assessments of TI method developers.
-We did made our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
-However, so far we do not observe an increase in developers performing quantitative benchmarks.
-We hypothesise causal reasons for this phenomenon and provide solutions in order to spur TI developers to perform more self-assessments (Chapter~\ref{chap:selfassessment}).
-% "We hypothesise." Where/what do you hypothise? In your paper? Dan zou ik verwijzen naar deze sectie.
-% Eventueel ook: Is there room for improvement for existing methods (deze zullen vermoedelijk onder een andere naam gepubliceerd dienen te worden)
-% Dienen eerdere publicaties in vraag te worden gesteld (indien minder goede methods gebruikt werden?
-% Kan dit uitgebreid worden naar andere methoden?
-% Is dyngen niet ontstaan omdat je nood had aan betere datasets? Zou je in dit opzicht niet dit deeltje eerst plaatsen en verwijzen naar dyngen?
-% Misschien ook vermelden dat de publicatie een aanzienlijke support verwacht van jullie team na publicatie. Dat de hoge nood gepaard gaat met een grote vraag. 
-\section{A toolkit to infer, visualise and interpret single-cell trajectories}
+Trajectory alignment has been used to compare gene expression kinetics resulting from different biological processes\cite{cacchiarelli_aligningsinglecelldevelopmental_2018}, to compare human, chimpanzee and macaque neuronal development\cite{kanton_organoidsinglecellgenomic_2019}, to find differences in gene regulation in the presence of certain growth factors\cite{mcfaline-figueroa_pooledsinglecellgenetic_2019}, and to compare human and mouse embryogenesis\cite{alpert_alignmentsinglecelltrajectories_2018}.
 
-The previous two works necessitated developing a tool for visualising trajectory data.
-We extended this implementation into a full toolkit, named \texttt{dyno}, for inferring and analysing trajectories (Chapter~\ref{chap:dyno}).
-The toolkit allows performing downstream analyses such as detecting differentially expressed genes along transitions in the trajectory, annotating the trajectory, and comparing multiple trajectories in a common dimensionality reduction.
-A major benefit of \texttt{dyno} is that it allows utilising any of the 50 TI methods we wrapped as part of the comparison, without having to install dependencies for every TI method separately.
+However, aligning trajectories becomes exponentially more difficult as the complexity of compared trajectories increases.
+For this reason, trajectory alignment remains a mostly unexplored territory within single-cell omics.
 
-The software packages consist of several subpackages, each with a separate functionality (e.g.
-\texttt{dynplot} visualises trajectories, \texttt{dynmethods} provides wrappers for the TI methods).
-Since most of the packages are currently only hosted on Github, only users familiar with Github have been able to use it.
-To make \texttt{dyno} available to a larger audience, we are publishing each of these packages and developing command-line interfaces for the most important functionalities.
+Dynamic Time Warping (DTW) \cite{giorgino_computingvisualizingdynamic_2009} is most commonly used to align linear trajectories.
+DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between two sequences that indicates which sequence must be dilated or contracted to best match the other one.
 
-% Are we talking about R-packages?
+Over time, as the affordability of single-cell omics technologies improves and the abundance of available datasets increases, we hypothesize that these methods will become highly relevant in comparing dynamic processes in samples from multiple donors.
 
-\section{Fast, accurate, and robust single-cell pseudotime}
+\subsection{Differential network inference}
+Differential NI methods aim to reconstruct one network for each of the given conditions amongst the transcriptomic profiles.
+The conditions can be different cellular states or changes in environment, and profiles can be grouped according to prior knowledge or derived through unsupervised clustering.
+From the resulting networks, one can then investigate the pathways that are differentially activated between conditions (e.g.
+deregulated pathways between a diseased and healthy condition), or those that are similarly activated between conditions (e.g.
+similarly activated pathways between two different disease conditions).
 
-We developed a TI method, called \texttt{SCORPIUS}, specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}).
-Since our comparison of TI methods already lists 30 TI methods that can infer trajectories that are more complex than a linear ordering of cells, why would we need a linear TI method?
+Differential NI methods have already been described for bulk -omics data \cite{Ideker2012}, where they have been used, for instance, to elucidate deregulated mechanisms in different subtypes of leukaemia \cite{Gill2014}.
 
-Being able to study how cells progress over time is in terms of analysis as fundamental as being able to cluster cells or to identify differentially expressed genes between two groups.
-We showed that for analysing datasets containing linear trajectories, \texttt{SCORPIUS} outperforms all other TI methods tested as part of our TI benchmark.
+\subsection{Dynamic network inference}
+Advances in single-cell transcriptomics have fostered the development of trajectory inference methods \cite{Cannoodt2016}, which are computational tools to reconstruct dynamic processes such as cellular differentiation and cell cycle (Figure \ref{fig:ti_with_ni}).
+These methods usually recover the underlying structure of the data by focusing on the similarities between cells.
+The cells can then be ordered along a trajectory that optimally models the data structure.
+This bears resemblance to the more traditional time-series experiments with bulk transcriptomics data, albeit at a much higher temporal resolution.
+Unlike time-series experiments, trajectories are ideally inferred in an unsupervised manner, as not to introduce human bias.
 
-Despite \texttt{SCORPIUS} published on bioRxiv in 2016, discussion and usage of the pre-print has largely been limited to comparing the results of a different (linear) TI methods to those of \texttt{SCORPIUS}.
-Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface.
-Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results \cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019, reimegard_combinedmrnaprotein_2019}.
+As in time-series experiments, NI methods can exploit the information provided by trajectory inference methods in order to improve the resulting regulatory network.
+Several NI methods use this information in order to determine the directionality and the type of an interaction.
+This has helped to construct more accurate boolean networks (SCNS \cite{Moignard2015}), by selecting the network that most reliably describes the observed cell ordering.
+Several NI methods use this cell-ordering information to construct ordinary differential equations that are again improved to optimally describe the observed data \cite{Ocone2015,Wei2016}.
+While such methods are generally able to improve the accuracy of the produced networks by using information from the trajectory, they suffer from the same limitations as discussed at the start of the section since they produce a single regulatory network that is supposed to be a model for all of the cells in the population.
 
-\section{Inferring single cell regulatory networks}
+Instead, dynamic NI methods produce a network of interactions with variable activation levels across the trajectory.
+They provide useful information on which transcription factors are expressed at the beginning of a developmental process, and which gene interactions occur at later stages for instance.
+Dynamic networks are more complex models in comparison to differential and static networks.
 
-We proposed a novel algorithm for inferring case-wise gene regulatory interactions (Chapter~\ref{chap:bred}).
-While we demonstrated the methodology by inferring 14'963 sample-specific GRNs from the TCGA project, further interpretation of these results needs to be perform to validate the usefulness of the method.
+\subsection{Casewise network inference}
+Casewise network inference (CNI)
+methods\footnote{Other terms are commonly used when dealing with data from a particular source.
+For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.}
+predict not only which transcription factors regulate which target genes
+(Figure~\ref{fig:scgrn}A, top left), but also how active each
+interaction is in every case (Figure~\ref{fig:scgrn}A). In CNI, a `case'
+might be a cell, but it might also refer to a bulk sample.
 
-Additional avenues for exploration could involve applying trajectory inference on the case-wise regulome data, and quantitatively evaluating the performance of the algorithm using the \texttt{dyngen} simulations.
+While a few pioneering CNI approaches have already been
+developed\autocites{aibar_scenicsinglecellregulatory_2017}{kuijjer_estimatingsamplespecificregulatory_2019}{liu_personalizedcharacterizationdiseases_2016},
+a quantitative assessment of the performance is hitherto lacking. This
+is not surprising, as neither real nor in silico datasets of
+cell-specific or even cell-type-specific interactions exists that is
+large enough so that it can be used as a ground-truth for evaluating CNI
+methods.
 
-\section{Optimising regulatory networks}
-This project was a spin-off of Netter \cite{ruyssinck_netterrerankinggene_2016}, a tool for reranking GRN predictions using structural network properties. 
-Predicted GRNs often are topologically dissimilar to real networks, and as such, Netter quantifies the topological properties of a network and makes incremental changes in order to arrive at a network with a 'better' topological profile.
+\subsection{Intercellular network inference}
+Cells are influenced by extracellular signals produced by cells
+in their microenvironment. One way of studying intercellular
+communication is to profile gene expression in interacting cells.
+Especially single-cell-omics technologies are promising because
+of their ability to analyze tissue composition in high resolution \cite{yosef_writlargegenomic_2016,tanay_scalingsinglecellgenomics_2017} .
+Furthermore, spatial and sophisticated single-cell transcriptomics
+technologies have recently been developed to assay the spatial con-
+text of a cell \cite{stahl_visualizationanalysisgene_2016,rodriques_slideseqscalabletechnology_2019,eng_transcriptomescalesuperresolvedimaging_2019,vickovic_highdefinitionspatialtranscriptomics_2019} , characterize cellular niches 7 or construct physical
+cell–cell interaction maps 8 .
+Several studies have recently demonstrated the potential of
+single-cell transcriptomics to investigate intercellular communica-
+tion 9–14 . There, computational methods, such as CellPhoneDB 14 ,
+were applied to infer links between ligands (that is, extracellular
+protein signals) expressed by sender cells and receptors expressed
+by receiver cells. Thus, these computational approaches elucidate
+by which extracellular signals cells can communicate. Nevertheless,
+a functional understanding of cell–cell communication requires
+knowledge about the influence of these ligand–receptor interac-
+tions on target gene expression. Hence, there is a need for compu-
+tational methodologies that use expression data of interacting cells
+to infer the effects of sender-cell ligands on receiver-cell expression.
+Although a few basic methods were developed for bulk data (for
+example, CCCExplorer) 15,16 , comprehensive and predictive methods
+that are generally applicable to both bulk and single-cell expression
+data are lacking.
 
-The topological 'profile' of a network was defined as the frequencies of particular topological patterns called 'graphlets'.
-Keeping track of the frequencies of graphlets during an iterative process proved difficult at first.
-\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by  assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals).
-Under this assumption, it is possible to calculate the differences in topological patterns significantly faster in comparison to recomputing the graphlet frequencies from scratch at every iteration.
-
-While \texttt{incgraph} solved a very specific and interesting problem, and was eventually published as a peer-reviewed article, it has not succeeded in making a direct impact on scientific research as of yet.
-Perhaps, at some point, a researcher will come across \texttt{incgraph} and find that it exactly solves his or her problem.
-
-% Of je dient voor te stellen bij welke dataset het goed van toepassing is en het te promotor onder de Vlaamse research gemeenschap?
+% When does a cluster represent a new cell type? \cite{kiselev_challengesunsupervisedclustering_2019}
 

### 2019-12-09 ch1_introduction/content.tex
remove usage of particular commands

@@
 	\centering
-	\includegraphics[width=\LARgefigure]{fig/bulk_vs_sc} 
+	\includegraphics[width=0.6\textwidth]{fig/bulk_vs_sc} 
 	\caption{
@@
 
-\section{Computational tools}
+\section{Computational tools} \label{sec:comptools}
 Whole-genome profiling at single-cell level allows new types of analyses with which to study cellular heterogeneity at a hitherto unseen throughput. 

### 2019-12-09 ch7_incgraph/content.tex
remove usage of particular commands

@@
 %% FIGURE 1
-\picfig{fig/figure1.pdf}{\LARGEfigure}{fig_graphlets}{{\bf Graphlet counting in a network characterises its local topologies.}
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=0.8\textwidth]{fig/figure1.pdf}
+	\caption{%
+		{\bf Graphlet counting in a network characterises its local topologies.}
 		(A) In total, there are 30 different graphlets containing 2 to 5 nodes, ranging from $G_0$ to $G_{29}$.
@@
 The 73 different orbits are coloured in red.
-(B) By counting the occurrences of these graphlets in the network, the local topology surrounding a node can be quantified.}
+(B) By counting the occurrences of these graphlets in the network, the local topology surrounding a node can be quantified.
+	}
+	\label{fig_graphlets}
+\end{figure}
 
@@
 %% FIGURE 2
-\picfig{fig/figure2.pdf}{.7\linewidth}{fig_efficient}{{\bf Several strategies are employed in order to reduce time and memory consumption.}
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=.7\linewidth]{fig/figure2.pdf}
+	\caption{{\bf Several strategies are employed in order to reduce time and memory consumption.}
 		(A) Only the depth 3 neighbourhood of each modified edge needs to be explored in order to have visited all modified five-node graphlets.
@@
 (D) The adjacency matrix of the network is compressed in order to reduce memory usage.}
+	\label{fig_efficient}
+\end{figure}
 
@@
 %% FIGURE 3
-\picfig{fig/figure3.pdf}{\LARGEfigure}{fig_method_speedup}{{\bf Static network model generators were modified to generate dynamic networks.}
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=0.8\textwidth]{fig/figure3.pdf}
+	\caption{{\bf Static network model generators were modified to generate dynamic networks.}
 		Three network models were used: Barabási–Albert, Erdős–Rényi, and Geometric.
@@
 Steps 1 to 3 are repeated until all modifications generated at step 0 are processed, or until an execution time threshold has been reached.}
+	\label{fig_method_speedup}
+\end{figure}
 
@@
 %% FIGURE 4
-\picfig{fig/figure4.pdf}{.9\linewidth}{fig_method_reranking}{{\bf Predicted gene regulatory networks of model organisms are optimised to reduce the false positive rate.}
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=.9\linewidth]{fig/figure4.pdf}
+	\caption{{\bf Predicted gene regulatory networks of model organisms are optimised to reduce the false positive rate.}
 		A)~The number of redundant edges in each graphlet are counted.
@@
 C) Starting from an empty network, the interactions are modified by iteratively evaluating the increase in redundancy of the next $k$ interactions, and adding the first edge for which its redundancy is less than the $90^\textrm{th}$ percentile redundancy.}
+	\label{fig_method_reranking}
+\end{figure}
 
@@
 %% FIGURE 5
-\picfig{fig/figure5.pdf}{\LARGEfigure}{fig_speedup}{{\bf IncGraph is significantly faster than non-incremental approaches.}
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=.8\textwidth]{fig/figure5.pdf}
+	\caption{{\bf IncGraph is significantly faster than non-incremental approaches.}
 		For small networks, the execution time of IncGraph $T_{IG}$ is already 50 times less than that of non-incremental approaches $T_{NI}$.
 This ratio increases even further for networks with higher numbers of nodes or higher average degrees.}
+	\label{fig_speedup}
+\end{figure}
 
@@
 %% FIGURE 6
-\picfig{fig/figure6.png}{\Largefigure}{fig_reranking}{{\bf A simple graphlet-based scoring method improves predicted regulatory networks.}
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=0.5\textwidth]{fig/figure6.png}
+	\caption{{\bf A simple graphlet-based scoring method improves predicted regulatory networks.}
 		(A) The F1 score was calculated by calculating the harmonic mean of the AUROC and AUPR scores of the first 1000 interactions.
@@
 Note that for each interaction added to the network, the graphlet counts of 100 putative interactions were evaluated.}
+	\label{fig_reranking}
+\end{figure}
 

### 2019-12-09 chX_backmatter.tex
fixes to backmatter

@@
   \item \textbf{BIG N2N symposium}, Ghent, Belgium, 19 May 2016.
-  \item \textbf{CYTO}, Seattle, United States of America, 11-15 June 2016.
+  \item \textbf{CYTO}, Seattle, USA, 11-15 June 2016.
   \item \textbf{BENELEARN}, Kortrijk, Belgium, 12-13 September 2016.
@@
   \item \textbf{VIB Seminar}, Veldhoven, Netherlands, 27-28 April 2017.
-  \item \textbf{Keystone Symposia: Single Cell Omics}, Stockholm, Zweden, 26-30 May 2017. 
-  \item \textbf{BeNeLux Bioinformatics Conference}, Louvain, België, 13-14 December 2017.
+  \item \textbf{Keystone Symposia: Single Cell Omics}, Stockholm, Sweden, 26-30 May 2017. 
+  \item \textbf{BeNeLux Bioinformatics Conference}, Louvain, Belgium, 13-14 December 2017.
   \item \textbf{Single Cell Biology}, Hinxton, United Kingdom, 8-10 March 2018.
@@
   \item Cannoodt R., Saelens W., De Preter K., Saeys Y.
-: SCORPIUS: Inferring trajectories along dynamic processes form single-cell RNA-seq data.
+: SCORPIUS: Inferring trajectories along dynamic processes from single-cell RNA-seq data.
 \textbf{Single Cell Biology Workshop}, Ghent, Belgium, 15 January 2016.
-  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.
-: Improving marker gene discovery from high-dimensional single-cell snapshot data.
-\textbf{CYTO}, Seattle, United States of America, 11-15 June 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Improving marker gene discovery from high-dimensional single-cell snapshot data. \textbf{CYTO}, Seattle, USA, 11-15 June 2016.
   \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. \textbf{BENELEARN}, Kortrijk, Belgium, 12-13 September 2016.
@@
 : Unbiased modelling of dynamic processes with SCORPIUS identifies novel modules in dendritic cell development.
-\textbf{VIB Seminar}, Veldhoven, Nederland, 27-28 april 2017.
+\textbf{VIB Seminar}, Veldhoven, Netherlands, 27-28 april 2017.
   \item Cannoodt R., Saelens W. : Automated building and unit testing, Docker and Singularity. \textbf{VIB Developers Meeting}, Ghent, Belgium, 25 January 2019.
@@
   \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. Single Cell Biology, Hinxton, United Kingdom, 8-10 March 2016.
-  \item Cannoodt R., Saelens W., De Preter K., Saeys Y.
-: Unbiased modelling of dynamic processes with single-cell RNA-sequencing.
-\textbf{BIG N2N symposium}, \textbf{BIG N2N symposium}, Ghent, Belgium, 19 May 2016.
-  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Improving marker gene discovery from high-dimensional single-cell snapshot data. CYTO, Seattle, United States of America, 11-15 June 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. \textbf{BIG N2N symposium}, Ghent, Belgium, 19 May 2016.
+  \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Improving marker gene discovery from high-dimensional single-cell snapshot data. CYTO, Seattle, USA, 11-15 June 2016.
   \item Cannoodt R., Saelens W., De Preter K., Saeys Y. : Unbiased modelling of dynamic processes with single-cell RNA-sequencing. BENELEARN 2016, Kortrijk, Belgium, 12-13 September 2016.
@@
   \item Cannoodt R., Saelens W., De Preter K., Saeys Y.: True single cell network inference: Modelling gene regulation of individual cells.
-\textbf{Keystone Symposia: Single Cell Omics}, Stockholm, Zweden, 26-30 May 2017.
+\textbf{Keystone Symposia: Single Cell Omics}, Stockholm, Sweden, 26-30 May 2017.
   \item Cannoodt R., Saelens W., Todorov H., Saeys Y.: Generalised framework for and comparison of 24 trajectory inference methods.
-\textbf{BeNeLux Bioinformatics Conference}, Louvain, België, 13-14 December 2017.
+\textbf{BeNeLux Bioinformatics Conference}, Louvain, Belgium, 13-14 December 2017.
   \item Cannoodt R., Saelens W., Todorov H., Saeys Y.
@@
 
-% thanking:
-% the people who brought me soup and sandwiches (look up names)
-% the people who tried to use the PRISM cluster but couldn't
+I was told that the end of the dissertation is the right place to thank friends, family, and colleagues for their continued support during my PhD.
+Nobody ever specified \textit{which} end of the book.
+This is not the right end you're looking at.
+
+Hello there! Did you stumble across this phrase by accident, 
+or are you looking for what I think you're looking for?

### 2019-12-09 ch0_frontmatter.tex
fixes to front matter

@@
 \hypersetup{%
-	pdftitle = {Modelling single cell dynamics with trajectories and gene regulatory networks},
+	pdftitle = {Modelling single-cell dynamics with trajectories and gene regulatory networks},
 	pdfsubject = {A subject},
@@
 	\includegraphics[width=\linewidth]{fig/logos/mashup}
-%	\includegraphics[height=.7cm]{fig/logos/icon_UGent_WE_EN_RGB_color}\hfill
-%	\includegraphics[height=1cm]{fig/logos/vib_rf_inflammation_research_rgb_pos} \hfill
-%	\includegraphics[height=.8cm]{fig/logos/dambi_logo} \hfill
-%	\includegraphics[height=.7cm]{fig/logos/cmgg_myshort} \hfill
-%	\includegraphics[height=.7cm]{fig/logos/FWO_Logo_KleurCMYK} 
 	\vspace{1cm}
@@
 
-Finally, I reflect on current challenges within the field, by providing guidelines for how to self-assess trajectory inference methods (Chapter~\ref{chap:selfassessment}) and, more generally, how to perform large-scale benchmarking experiments of computational methods (Chapter~\ref{chap:guidelines}).
+Finally, I reflect on current challenges within the field, by providing guidelines on how to perform large-scale benchmarking experiments of computational methods (Chapter~\ref{chap:guidelines}).
 
@@
 
-Ik reflecteer op huidige uitdagingen binnen het veld, door richtlijnen aan te bieden aan ontwikkelaars voor het evalueren van hun trajectvoorspellende methoden (Hoofdstuk~\ref{chap:selfassessment}), en meer algemeen over hoe grootschalige benchmarkingsexperimenten van computationele methoden kunnen worden uitgevoerd (Hoofdstuk~\ref{chap:guidelines}).
+Ik reflecteer op huidige uitdagingen binnen het veld, door richtlijnen aan te bieden aan ontwikkelaars over hoe grootschalige benchmarkingsexperimenten van computationele methoden kunnen worden uitgevoerd (Hoofdstuk~\ref{chap:guidelines}).
 

### 2019-12-09 ch10_discussion/content.tex
update

@@
 \section{Outlook}
-The types of computational analyses presented in Section~\ref{sec:comptools} (Clustering, DE, DR, TI, NI) all become routine methodologies for analysing and interpreting single-cell omics data.
-Here, we discuss several promising categories of computational methods that are currently still in a exploratory stage (Figure~\ref{fig:scapplications_extended}).
+The types of computational analyses presented in Section~\ref{sec:comptools} (Clustering, DE, DR, TI, NI) have all become routine methodologies for analysing and interpreting single-cell omics data.
+With projects like the Human Cell Atlas\cite{humancellatlasconsortium_humancellatlas_2018} soon to be churning out millions of single-cell profiles on a regular basis, research in high-throughput computational tools to analyse these data in an unsupervised approach becomes increasingly relevant.
+
+Here, we discuss several promising categories of computational methods (Figure~\ref{fig:scapplications_extended}) that are mostly in an exploratory stage but show promising results towards becoming another major computational workhorse in single-cell biology.
 
@@
 
-\subsection{Differential network inference}
-Differential NI methods aim to reconstruct one network for each of the given conditions amongst the transcriptomic profiles.
-The conditions can be different cellular states or changes in environment, and profiles can be grouped according to prior knowledge or derived through unsupervised clustering.
-From the resulting networks, one can then investigate the pathways that are differentially activated between conditions (e.g.
-deregulated pathways between a diseased and healthy condition), or those that are similarly activated between conditions (e.g.
-similarly activated pathways between two different disease conditions).
+\subsection{Variations on network inference}
+Using a collection of transcriptomic profiles, network inference (NI) methods predict which genes are the regulators of a target gene.
+The output of an NI method is called a gene regulatory network in which each predicted interaction can be assigned a predicted strength or effect (activating or inhibitory).
+However, cells are very molecularly heterogeneous, and only a fraction of all possible gene regulatory interactions is active at any point in time.
+Unfortunately, 'bulk' NI does not permit studying the diversity of regulatory activity within and between cell populations, and will thus suffer from similar drawbacks as bulk omics when compared to single-cell omics.
+Several adaptations of NI have been developed to address this problem, namely Differential NI, Dynamic NI, Casewise NI and Intercellular NI.
 
-Differential NI methods have already been described for bulk -omics data \cite{Ideker2012}, where they have been used, for instance, to elucidate deregulated mechanisms in different subtypes of leukaemia \cite{Gill2014}.
+\textbf{Differential NI} methods reconstruct one network per group of cells.
+The grouping can either be defined by prior knowledge, for example from gene expression, clustering, or by sorting the cells \textit{a priori}.
+The resulting networks can be investigate to detect differentially active (groups of) interactions between conditions (e.g.
+deregulated pathways between healthy and diseased), or (groups of) interactions that are highly active in between conditions (e.g.
+for investigating common pathways between different conditions).
+Differential NI methodology had already been developed for bulk omics \cite{ideker_differentialnetworkbiology_2012} to, for example, infer deregulated regulatory mechanisms in different subtypes of leukaemia \cite{gill_differentialnetworkanalysis_2014}.
+Recently, similar methodology has been discussed in the context of single-cell omics as well \cite{chiu_scdnetcomputationaltool_2018}.
 
-\subsection{Dynamic network inference}
-Advances in single-cell transcriptomics have fostered the development of trajectory inference methods \cite{Cannoodt2016}, which are computational tools to reconstruct dynamic processes such as cellular differentiation and cell cycle (Figure \ref{fig:ti_with_ni}).
-These methods usually recover the underlying structure of the data by focusing on the similarities between cells.
-The cells can then be ordered along a trajectory that optimally models the data structure.
-This bears resemblance to the more traditional time-series experiments with bulk transcriptomics data, albeit at a much higher temporal resolution.
-Unlike time-series experiments, trajectories are ideally inferred in an unsupervised manner, as not to introduce human bias.
+\textbf{Dynamic NI} methods exploit cell ordering information to improve the resulting regulatory network.
+Prior ordering information can be obtained experimentally (e.g.
+by performing lineage tracing or time-series experiments) or computationally (e.g.
+by performing trajectory inference or RNA velocity experiments).
+Unfortunately, most methods which are labelled as dynamic NI methods use time-series information to augment the inference of a static network.
+Ideally, dynamic NI methods would predict how the activity of interactions change along (pseudo-)time, as doing so would allow to detect regulatory cascades or interactions which act as the main drivers behind crucial dynamic processes.
 
-As in time-series experiments, NI methods can exploit the information provided by trajectory inference methods in order to improve the resulting regulatory network.
-Several NI methods use this information in order to determine the directionality and the type of an interaction.
-This has helped to construct more accurate boolean networks (SCNS \cite{Moignard2015}), by selecting the network that most reliably describes the observed cell ordering.
-Several NI methods use this cell-ordering information to construct ordinary differential equations that are again improved to optimally describe the observed data \cite{Ocone2015,Wei2016}.
-While such methods are generally able to improve the accuracy of the produced networks by using information from the trajectory, they suffer from the same limitations as discussed at the start of the section since they produce a single regulatory network that is supposed to be a model for all of the cells in the population.
+In \textbf{Casewise NI}, one regulatory network is predicted for every single-cell in the dataset.
+Doing so allows studying the heterogeneity (or homogeneity) between and within different cell populations.
+In essence, most analyses that can be run on single-cell omics (e.g.
+diffential expression) can also be applied on casewise regulatory networks (e.g.
+to predict differentially expressed interactions).
+It is the most generalised form of NI of the variants discussed above, as a casewise regulatory network can be clustered to create a differential network, or trajectory inference can be applied to derive a dynamic regulatory network.
+Examples of casewise NI methods include SCENIC \cite{aibar_scenicsinglecellregulatory_2017}, SSN \cite{liu_personalizedcharacterizationdiseases_2016}, and LIONESS \cite{kuijjer_estimatingsamplespecificregulatory_2019}.
 
-Instead, dynamic NI methods produce a network of interactions with variable activation levels across the trajectory.
-They provide useful information on which transcription factors are expressed at the beginning of a developmental process, and which gene interactions occur at later stages for instance.
-Dynamic networks are more complex models in comparison to differential and static networks.
+Instead of inferring gene regulatory mechanisms occurring within a cell, \textbf{intercellular NI} methods study communication between cells by predicting interactions between the ligands of one type of cells and the receptors of another group of cells.
 
-\subsection{Casewise network inference}
-Casewise network inference (CNI)
-methods\footnote{Other terms are commonly used when dealing with data from a particular source.
-For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.}
-predict not only which transcription factors regulate which target genes
-(Figure~\ref{fig:scgrn}A, top left), but also how active each
-interaction is in every case (Figure~\ref{fig:scgrn}A). In CNI, a `case'
-might be a cell, but it might also refer to a bulk sample.
+% browaeys_nichenetmodelinginterce
 
-While a few pioneering CNI approaches have already been
-developed\autocites{aibar_scenicsinglecellregulatory_2017}{kuijjer_estimatingsamplespecificregulatory_2019}{liu_personalizedcharacterizationdiseases_2016},
-a quantitative assessment of the performance is hitherto lacking. This
-is not surprising, as neither real nor in silico datasets of
-cell-specific or even cell-type-specific interactions exists that is
-large enough so that it can be used as a ground-truth for evaluating CNI
-methods.
-
-\subsection{Intercellular network inference}
-Cells are influenced by extracellular signals produced by cells
-in their microenvironment. One way of studying intercellular
-communication is to profile gene expression in interacting cells.
-Especially single-cell-omics technologies are promising because
-of their ability to analyze tissue composition in high resolution \cite{yosef_writlargegenomic_2016,tanay_scalingsinglecellgenomics_2017} .
-Furthermore, spatial and sophisticated single-cell transcriptomics
-technologies have recently been developed to assay the spatial con-
-text of a cell \cite{stahl_visualizationanalysisgene_2016,rodriques_slideseqscalabletechnology_2019,eng_transcriptomescalesuperresolvedimaging_2019,vickovic_highdefinitionspatialtranscriptomics_2019} , characterize cellular niches 7 or construct physical
-cell–cell interaction maps 8 .
-Several studies have recently demonstrated the potential of
-single-cell transcriptomics to investigate intercellular communica-
-tion 9–14 . There, computational methods, such as CellPhoneDB 14 ,
-were applied to infer links between ligands (that is, extracellular
-protein signals) expressed by sender cells and receptors expressed
-by receiver cells. Thus, these computational approaches elucidate
-by which extracellular signals cells can communicate. Nevertheless,
-a functional understanding of cell–cell communication requires
-knowledge about the influence of these ligand–receptor interac-
-tions on target gene expression. Hence, there is a need for compu-
+Hence, there is a need for compu-
 tational methodologies that use expression data of interacting cells

### 2019-12-09 ch10_discussion/content.tex
update

@@
 
-In \textbf{Casewise NI}, one regulatory network is predicted for every single-cell in the dataset.
+In \textbf{casewise NI}, one regulatory network is predicted for every single-cell in the dataset.
 Doing so allows studying the heterogeneity (or homogeneity) between and within different cell populations.
@@
 
-Instead of inferring gene regulatory mechanisms occurring within a cell, \textbf{intercellular NI} methods study communication between cells by predicting interactions between the ligands of one type of cells and the receptors of another group of cells.
-
-% browaeys_nichenetmodelinginterce
-
-Hence, there is a need for compu-
-tational methodologies that use expression data of interacting cells
-to infer the effects of sender-cell ligands on receiver-cell expression.
-Although a few basic methods were developed for bulk data (for
-example, CCCExplorer) 15,16 , comprehensive and predictive methods
-that are generally applicable to both bulk and single-cell expression
-data are lacking.
-
-% When does a cluster represent a new cell type? \cite{kiselev_challengesunsupervisedclustering_2019}
+Instead of inferring gene regulatory mechanisms occurring within a cell, \textbf{intercellular NI} methods \cite{efremova_cellphonedbv2inferring_2019,browaeys_nichenetmodelingintercellular_2019} study communication between cells by predicting interactions between the ligands of one type of cells and the receptors of another group of cells.
+Such methods typically require much more prior data (e.g.
+which cells are communicating with which cells, what are the ligands and receptors), but allows studying how the cells react to their environment and investigate intercellular feedback mechanisms.
 
@@
 
-Honourable mentions go to Linux, Fedora, \LaTeX, TeXstudio, (R)Markdown, Bash, Sed, Regular expressions, Rocket Chat, for making bioinformatics software development even more fun and enjoyable.
+Honourable mentions are Linux, Fedora, \LaTeX, TeXstudio, (R)Markdown, Bash, Sed, Regular expressions, Rocket Chat, for making bioinformatics software development even more fun and enjoyable.
 
-% ideas:
-%  Yet I wonder how the current academic system
-%can support such concepts. Academics often hold different positions in
-%different research groups around the world, who should be the maintainer in
-%these cases? The original PhD students that developed the work? The original
-%research group?
-
-%The eight chapter is a manuscript in preparation about the assessment of TI methods.
-The main discussion topic is the lack of a quantitative measure for the scoring of new TI methods.
-This leads to the publication and development of new methods that do not show a clear performance improvement compared to other methods.
-The main topic touches
-%upon some sore points that are inherent in bioinformatics. The main thought I have with
-%this chapter is that defining clear goals in an emerging technology is difficult and therefore is sometimes ignored by fellow bioinformaticians.
-This because the lack of standardized performance metrics, missing public datasets that are well described and an \textit{ad hoc} problem definition.
-I follow the idea of defining a clear cut approach to benchmarking (as described in chapter nine) and I like the proposition that is started in this manuscript for a discussion on good TI metrics.} Would approaches like e.g.
-DREAM challenges be a means of tackling these issues to your opinion?
-
-% zie peter dawyndt, chapter 5
 \clearpage

### 2019-12-09 chX_backmatter.tex
update

@@
 
-I was told that the end of the dissertation is the right place to thank friends, family, and colleagues for their continued support during my PhD.
-Nobody ever specified \textit{which} end of the book.
-This is not the right end you're looking at.
-
-Hello there! Did you stumble across this phrase by accident, 
-or are you looking for what I think you're looking for?
+This dissertation would never have seen the light of day were it not for the continued support of many friends, family, and colleagues.
+While writing the dissertation, Caro told me that the right place to thank everybody was at the end of the book.
+It was not specified \textit{which} end of the book I should be using for this.
+This isn't the Acknowledgement section you're looking for.

### 2019-12-10 boek.tex
updates

--- 2019-12-05 boek.tex
@@
 \PassOptionsToPackage{table}{xcolor}
-\documentclass[8pt]{book}
+\documentclass[10pt]{book}
 
@@
 
-\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm]{geometry}
+%\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm,inner=2cm,outer=1.5cm]{geometry}
+%\usepackage[
+%	% set width and height to a4 width and height + 6mm
+%	%width=16truecm, height=24truecm,
+%	a4,
+%	% use any combination of these options to add different cut markings
+%	cam, axes, frame, cross,
+%	% set the type of TeX renderer you use
+%	pdftex,
+%	% center the contents
+%	center
+%]{crop}
 
-\usepackage[
-	% set width and height to a4 width and height + 6mm
-	%width=16truecm, height=24truecm,
-	a4,
-	% use any combination of these options to add different cut markings
-	cam, axes, frame, cross,
-	% set the type of TeX renderer you use
-	pdftex,
-	% center the contents
-	center
-]{crop}
-
+\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm,inner=2cm,outer=1.5cm]{geometry}
 % adjust margins (should be done before importing fancyhdr!) %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm
@@
 \usepackage{titlesec}
-%\titleformat{\section} {\Large\bfseries\center}{}{0em}{\hrule height 0.4pt\vspace{2pt}}[\hrule height 0.4pt] \titlespacing*{\section} {0pt}{0pt}{10pt}
-%\titleformat{\subsection} {\large\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsection} {0pt}{0pt}{0pt}
-%\titleformat{\subsubsection} {\normalsize\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsubsection}  {0pt}{0pt}{0pt}
-%\titleformat{\section} [block]{\Large\bfseries\filcenter}{}{0em}{\MakeUppercase}
-%\titleformat{\subsection} [block]{\large\bfseries\filcenter}{}{0em}{\MakeUppercase}
-%\titleformat{\subsubsection} [block]{\normalsize\bfseries\filcenter}{}{0em}{\MakeUppercase}
 
@@
 
+% magic for reading parts of a file
+\makeatletter
+\newread\pin@file
+\newcounter{pinlineno}
+\newcommand\pin@accu{}
+\newcommand\pin@ext{pintmp}
+% inputs #3, selecting only lines #1 to #2 (inclusive)
+\newcommand*\partialinput [3] {%
+	\IfFileExists{#3}{%
+		\openin\pin@file #3
+		% skip lines 1 to #1 (exclusive)
+		\setcounter{pinlineno}{1}
+		\@whilenum\value{pinlineno}<#1 \do{%
+			\read\pin@file to\pin@line
+			\stepcounter{pinlineno}%
+		}
+		% prepare reading lines #1 to #2 inclusive
+		\addtocounter{pinlineno}{-1}
+		\let\pin@accu\empty
+		\begingroup
+		\endlinechar\newlinechar
+		\@whilenum\value{pinlineno}<#2 \do{%
+			% use safe catcodes provided by e-TeX's \readline
+			\readline\pin@file to\pin@line
+			\edef\pin@accu{\pin@accu\pin@line}%
+			\stepcounter{pinlineno}%
+		}
+		\closein\pin@file
+		\expandafter\endgroup
+		\scantokens\expandafter{\pin@accu}%
+	}{%
+		\errmessage{File `#3' doesn't exist!}%
+	}%
+}
+\makeatother
+
+\newcounter{acki}
+\setcounter{acki}{1}
+\newcommand{\doack}{%
+%	{\footnotesize %
+%    %\#\theacki\ | % LINENUMBER DEBUG
+%		\partialinput{\theacki}{\theacki}{ack}
+%	}%
+	\stepcounter{acki}%
+}
+
 %%%%%%%%%%%%%%%%%%%%%
@@
 }
+
 \fancypagestyle{mainmatter}{%    
@@
 	\fancyhead[RO,LE]{\nouppercase{\small \textsf{\leftmark}}}
-	\fancyfoot[RO,LE]{\textsf{Page \thepage}} 
+	\fancyfoot[RO,LE]{\textsf{{\footnotesize Page \thepage}}}
+	\fancyfoot[LO]{\ifodd\theacki\doack\fi}
+	\fancyfoot[RE]{\ifodd\theacki\else\doack\fi}
+
 }
@@
 		\chead{\oddthumb{#1}}% chapters start only on odd pages
-		\fancyfoot[RO,LE]{\textsf{Page \thepage}} %\fancyfoot[RO,LE]{\textsf{Page \thepage}}
+		\fancyfoot[RO,LE]{\textsf{Page \thepage}}
 	}
@@
 }
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% DEFINE FIGURE HELPERS %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\hugefigure}{\textwidth}
-\newcommand{\LARGEfigure}{0.8\textwidth} % 1 per col
-\newcommand{\LARgefigure}{0.6\textwidth} % 1 per col
-\newcommand{\Largefigure}{0.5\textwidth} % 1 per col
-\newcommand{\largefigure}{0.37\textwidth} % 2 per col
-\newcommand{\normalfigure}{0.3\textwidth} % 3 per col
-\newcommand{\smallfigure}{0.24\textwidth} % 3 per col
-\newcommand{\Smallfigure}{0.19\textwidth} % 4 per col
-\newcommand{\SMALLfigure}{0.15\textwidth} % 5 per col
-
-\makeatletter
-\newcommand*{\centerfloat}{%
-	\parindent \z@
-	\leftskip \z@ \@plus 1fil \@minus \textwidth
-	\rightskip\leftskip
-	\parfillskip \z@skip}
-\makeatother
-
-\newcommand{\picfig}[4]{
-\begin{figure}[ht!]
-	\centering \centerfloat
-	\includegraphics[width=#2]{#1}
-	\caption{#4}
-	\label{#3}
-\end{figure}}
 
@@
 \begin{document}
-
 \pagenumbering{alph}
-
+\pagestyle{frontmatter}
 \frontmatter
-\pagestyle{empty}
 
@@
 
-%{\large \textbf{Abstract}}
-%\begin{quote}
-	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in a high-throughput and unbiased manner.
-Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following.
-In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods.
-We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation.
-As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
-%\end{quote}
+\vfill
 
-\vfill
 \newpage
@@
 	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
-	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
+	$\bullet$ \textbf{R.C.} and W.S. wrote the original manuscript.\\
 	$\bullet$ \textbf{R.C.}, W.S., L.D., and Y.S. reviewed and edited the manuscript. \\
@@
 \include{ch3_dynbenchmark/content}
-\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%   CH4: Self-assessment    %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{Self-assessment in trajectory inference} 
-\chaptermark{Self-assessment in trajectory inference.} % short title
-\label{chap:selfassessment}
-
-{\large \textbf{Abstract}}
-\begin{quote}
-	\textbf{Purpose:} A brief of 75 articles of trajectory tools reveals 
-	that only 37\% of works contain a quantitative self-assessment. 
-	While self-assessments are universally biased in favour of the
-	authors, it is dangerous to publish a computational tool
-	without quantitatively demonstrating its performance compared to
-	state-of-the-art methods. \\
-	\textbf{Results:} We hypothesise that low self-assessment rates are
-	primarily caused by a lack of a standardised problem definition, readily
-	available benchmarking datasets, and suitable metrics. We elaborate on
-	these causal reasons and provide viable solutions for 
-	performing benchmarks more easily.\\
-	\textbf{Conclusion:} With this perspective, we hope to address the
-	elephant in the room, and provide 
-	developers of trajectory inference methods with the necessary 
-	tools for performing benchmarking software quantitatively. 
-\end{quote}
-
-\vfill
-{\large \textbf{Publication status}}
-\begin{quote}
-	Manuscript in preparation. \\
-	\textbf{Cannoodt R.}
-\end{quote}
-
-\vfill
-{\large \textbf{Author contributions}}
-\begin{quote}
-	$\bullet$ \textbf{R.C.} designed the study.\\
-	$\bullet$ \textbf{R.C.} surveyed 75 manuscripts from literature and analysed the data.\\
-	$\bullet$ \textbf{R.C.} wrote the original manuscript.
-\end{quote}
-
-\vfill
-
-\ifrenderall
-\newpage
-\graphicspath{{./ch8_selfassessment/}}
-\include{ch8_selfassessment/content}
-\clearpage
-\section{References}
-\printbibliography[heading=none]
 \fi

### 2019-12-10 ch0_frontmatter.tex
updates

@@
 \vfill
+{
+\fontsize{9}{10}\selectfont
 Recent developments in the \textit{single-cell omics} technologies triggered a major milestone in the field of (cell)biology.
@@
 Throughout this project I particularly enjoyed working in an open-source ecosystem, collaborating with scientists around the world to tackle unsolved problems, whilst learning best practices by inspecting the source code of exemplary software projects.
+\vfill
+}
 
 \newpage
-%\ 
-%\vfill
+\ 
+\vfill
 {\bfseries\LARGE Samenvatting}
-%\vfill
+\vfill
 
+{
+\fontsize{9}{10}\selectfont
 Recente ontwikkelingen in de \textit{single-cell omics} technologieën vormen een belangrijke mijlpaal in het domein van de (cel)biologie.
@@
 Door de broncode van excellente softwareprojecten te bestuderen, leerde ik mezelf de aanbevolen werkwijzen aan.
+\vfill
+}

### 2019-12-10 ch3_dynbenchmark/content.tex
updates

@@
 
-\newpage
 \subsection{Accuracy}
@@
 \begin{table}[tbh!]
-	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}
+	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.
+Part 1.} \label{tab:scoresheet}
 
@@
 
-	\begin{tabular}{p{3.5cm}p{9cm}p{2.5cm}}
+	\begin{tabular}{p{2.5cm}p{8cm}p{1.5cm}}
 		\toprule
@@
 
-		\midrule
+		\bottomrule
+	\end{tabular}
+	\endgroup{}
+\end{table}
+
+\begin{table}[tbh!]
+	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Part 2.} \label{tab:scoresheet2}
+
+	\centering\begingroup\fontsize{7}{9}\selectfont
+
+	\begin{tabular}{p{2.5cm}p{8cm}p{1.5cm}}
+		\toprule
+		Aspect & Items & References \\ \midrule
+
 		\multicolumn{3}{l}{\textbf{Documentation}} \\*
@@
 Using our framework, this figure can be recreated using only a couple of lines of R code (\url{https://methods.dynverse.org}).
-In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses.
+In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{lamanno_rnavelocitysingle_2018}, and easier downstream analyses.
 In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.
@@
 
-The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
+The dyngen (Chapter~\ref{chap:dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
 At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
 We simulated a total of 110 datasets, with 11 different topologies.
-
-\hfill\break
-\textit{Network generation}
-\hfill\break
-
-One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
-The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo.
-One driver of bifurcation seems to be mutual antagonism, where genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
-Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
-Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
-
-To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
-Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle).
-The structure of these module networks is available at \href{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}.
-
-From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added.
-For each dataset, we used the same number of genes as were differentially expressed in the real datasets.
-5$\%$ of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module).
-We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module.
-To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits \cite{marbach_tissuespecificregulatorycircuits_2016}), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the page\_rank function of the \textit{igraph} package.
-
-\hfill\break
-\textit{Simulation of gene regulatory systems using thermodynamic models}
-\hfill\break
-
-To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
-In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):
-
-$$
-\begin{aligned}
-\label{eq:mrna_ode}
-\frac{dx_i}{dt} &= \underbrace{m \times f(y_1, y_2, ...)}_\text{production} - \underbrace{\lambda \times x_i}_\text{degradation}
-\end{aligned}
-$$
-$$
-\begin{aligned}
-\label{eq:prot_ode}
-\frac{dy_i}{dt} &= \underbrace{r \times x_i}_\text{production} - \underbrace{\Lambda \times y_i}_\text{degradation}
-\end{aligned}
-$$
-
-where $m$, $\lambda$, $r$ and $\Lambda$ represent production and degradation rates, the ratio of which determines the maximal gene and protein expression.
-The two types of equations are coupled because the production of protein $y_i$ depends on the amount of gene expression $x_i$, which in turn depends on the amount of other proteins through the activation function $f(y_1, y_2, ...)$.
-
-The activation function is inspired by a thermodynamic model of gene regulation, in which the promoter of a gene can be bound or unbound by a set of transcription factors, each representing a certain state of the promoter.
-Each state is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state.
-The production rate of the gene is calculated by combining the probabilities of the promoter being in each state with the relative activation:
-
-$$
-\begin{aligned}
-\label{eq:input_function_probabilities}
-f(y_1, y_2, ..., y_n) = \sum_{j \in \{0, 1, ..., n^2\}} \alpha_j \times P_j
-\end{aligned}
-$$
-
-The probability of being in a state is based on the thermodynamics of transcription factor binding. When only one transcription factor is bound in a state:
-$$
-\begin{aligned}
-P_j \propto \nu = \left(\frac{y}{k}\right)^{n}
-\end{aligned}
-$$
-
-where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
-$$
-\begin{aligned}
-P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
-\end{aligned}
-$$
-
-where $\rho$ represents the cooperativity of binding between the different transcription factors.
-
-$P_i$ is only proportional to $\nu$ because $\nu$ is normalized such that $\sum_{i} P_i = 1$.
-
-To each differential equation, we added an additional stochastic term: 
-$$
-\begin{aligned}
-\frac{dx_i}{dt} & = m \times f(y_1, y_2, ...) - \lambda \times x_i + \eta \times \sqrt{x_i} \times \Delta W_t \\
-\frac{dy_i}{dt} & = r \times x_i - \Lambda \times y_i + \eta \times \sqrt{y_i} \times \Delta W_t
-\end{aligned}
-$$
-
-with $\Delta W_t \sim \mathcal{N}(0, h)$.
-
-Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows.
-$e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.
-
-$$
-\begin{aligned}
-r & = \mathcal{U}(10, 200) \\
-d & = \mathcal{U}(2, 8) \\
-p & = \mathcal{U}(2, 8) \\
-q & = \mathcal{U}(1, 5) \\
-a_0 & = \begin{cases}1 & \text{if } |e| = 0 \\ 1 & \text{if } \forall x \in e \text{, } x = -1 \\ 0 & \text{if } \forall x \in e \text{, } x = 1 \\ 0.5 & \text{otherwise}\end{cases} \\
-a_i & = \begin{cases}0 & \text{if } \exists x \in e_i \text{, } x = -1 \\ 1 & \text{otherwise}\end{cases} \\
-s & = \mathcal{U}(1, 20) \\
-k & = y_{max} / (2 * s) \text{,} \\
-\ & \quad \text{where } y_{max} = r / d \times p / q \\
-c & = \mathcal{U}(1, 4)
-\end{aligned}
-$$
-
-We converted each ODE to an SDE by adding a chemical Langevin equation, as described in \cite{schaffter_genenetweaversilicobenchmark_2011}.
-These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 8$.
-The total simulation time varied between 5 for linear and bifurcating datasets, 10 for consecutive bifurcating, trifurcating and converging datasets, 15 for bifurcating converging datasets and 30 for linear long, cycle and bifurcating loop datasets.
-The burn-in period was for each simulation 2.
-Each network was simulated 32 times.
-
-\hfill\break
-\textit{Simulation of the single-cell RNA-seq experiment}
-\hfill\break
-
-For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in.
-These cells were sampled uniformly across the different steps of the 32 simulations.
-Next, we used the Splatter package \cite{zappia_splattersimulationsinglecell_2017} to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities.
-We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$.
-These were combined with the expression levels of the genes simulated within a trajectory.
-Next, true counts were simulated using $Y'_{i,j} \sim \text{Poisson}(\lambda_{i,j})$.
-Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.
-Both $x_0$ (the midpoint for the dropout logistic function) and $k$ (the shape of the dropout logistic function) were estimated by Splatter.
-
-This count matrix was then filtered and normalised using the pipeline described below.
-
-\hfill\break
-\textit{Gold standard extraction}
-\hfill\break
-
-Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward.
-Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable.
-We therefore first constructed a backbone expression profile for each branch within the trajectory.
-To do this, we first defined in which order the expression of the modules is expected to change, and then generated a backbone expression profile in which the expression of these modules increases and decreases smoothly between 0 and 1.
-We also smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.
-We used dynamic time warping, implemented in the dtw R package \cite{giorgino_computingvisualizingdynamic_2009,tormene_matchingincompletetime_2009}, with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone.
-In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.
 
@@
 	\item A user may already know the topology, and may be primarily interested in finding good features related to a particular branching point
-	\item ...
 \end{itemize}

### 2019-12-10 ack.tex
update

Hello there! Did you stumble across this phrase by accident, 
or are you looking for what I think you're looking for?
For every page you turn over in this book,
I feel that there is at least one person who deserves to be thanked 
a thousand times over for their continued love, sacrifices and support.
So here we go, until the end of the book: Thank you! Thank you!
Thank you! Thank you! Thank you! Thank you! Thank you! Thank you! 
Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
Thank you! Thank you! Okay, enough with the silly business.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Yvan, you are an incredible supervisor and group leader. 
You always tell the baby PhD students that the first step in a PhD
is not learning to write well, but learning to like to write.
Those words are amazing PhD training wheels.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Katleen, thank you for your dedication and perseverance
in guiding me and making sure I got off to a good start in my PhD.
I see exciting times ahead for the KDP group, which is overflowing
with enthusiasm and inspiration. I wonder what the future will bring!
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Frank, thank you for giving me the opportunity to start on this journey.
You are an inexhaustible source of new ideas
which often explore the very thin line
between ingenuity and completely insane.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Jo, sorry for spilling coffee on your shoes on my first day in Blok B.
My strategy seemed to have worked, for suddenly I had a fourth promotor!
I thoroughly enjoyed my time as an unofficial member of your group.
You are a true people's person; you bring out the best in everybody.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Joeri, we've come full circle. You started me on this adventure
with your incredible friendliness and that darned oversized pumpkin;
and now, with you as my jury member, you helped me reach my goal.
Let's stay in touch :)
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Wouter, I cannot imagine a better partner to do a PhD with.
If we were genes, I'm sure we would reciprocal activators,
constantly stimulating each other to do just a little bit more!
Yet, our knowledge was often so perfectly complementary;
if we were a PCA plot, you'd be PC1 and I'd be PC2.
Or maybe the other way around? Let's take turns being PC1 and PC2.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Sofie, you saved my rear-end from embarrassing moments more frequently than I dare to admit.
The perfect example was when I walked in the MMM on 2014-01-06,
I arrived a little bit too late, and you send me a message:
"You do know that you have to present today, right?"
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Helena, thank you for being such a lovely colleague and an even better friend.
I enjoy our scientific talks and our non-scientific walks. 
Together with you, I played the longest ongoing game of hide-and-seek,
on page 22-9 you probably already found the truc that I am referring to.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Paulina, like a soft summer wind you passed through our lives.
Thanks for putting so many healthy vege-tables and strange Polish delicacies in our fridge,
for cycling together to work, and for making weird noises at Remi.
I'm looking forward to the next time we meet, so you can be amazed at how much Remi has grown.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Soup: Dani, Helena, Caroline, Robin B, Maxim, Annelies, Quentin, Katrien, Arne G, Ruth, Joris
Dambi
Dambi
Dambi
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Miriam
Miriam
Miriam
Miriam
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Laure, Michael, Jasper en Lize (spelling??)
Mara, Seppe \& Alice
Marta, Jan, Tristan, Rafael
Daan, Mieke \& Lou
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Cedric
Cedric
Cedric
Cedric
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Familie
Familie
Familie
Familie
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Victor, humor
Victor
Victor
Victor
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Hendrik, getting into PC
Hendrik 
Hendrik
Hendrik
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Papa Chris, babbel voor het slapen gaan, de vele schaakspelletjes
Chris
Chris
Chris
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Mama en papa Jean-Paul: opvoeding, zorgeloosheid
Mama
Mama
Mama
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Amor, as I write this, you are sitting on my lap and drooling on the table.
I guesstimate about 50\% of this dissertation was written like this.
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Hello Petit Remi! 
Remi
Remi
Remi
\rule{10cm}{0.4pt}
\rule{10cm}{0.4pt}
Dear Caro,
I cannot imagine a better partner and mother than you.
You are a strange combination of being extremely thoughtful and caring,
yet you roll over laughing when I stub my toe against the side of the table.
Let's get back on track collecting doorsteps,
and spending time doing the things we love, just us three.
I look forward to the rest of our lives together,
my strong independent woman.

### 2019-12-10 ack.tex
update acks

@@
 Thank you! Thank you! Okay, enough with the silly business.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+\newperson
+\newperson
 Yvan, you are an incredible supervisor and group leader.
@@
 Those words are amazing PhD training wheels.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+\newperson
+\newperson
 Katleen, thank you for your dedication and perseverance
@@
 with enthusiasm and inspiration. I wonder what the future will bring!
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+\newperson
+\newperson
 Frank, thank you for giving me the opportunity to start on this journey.
@@
 which often explore the very thin line
-between ingenuity and completely insane.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+between ingenuity and complete insanity.
+\newperson
+\newperson
 Jo, sorry for spilling coffee on your shoes on my first day in Blok B.
@@
 You are a true people's person; you bring out the best in everybody.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Joeri, we've come full circle. You started me on this adventure
+\newperson
+\newperson
+Joeri, we've come full circle. You brought me into this whole adventure
 with your incredible friendliness and that darned oversized pumpkin;
@@
 Let's stay in touch :)
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+\newperson
+\newperson
 Wouter, I cannot imagine a better partner to do a PhD with.
-If we were genes, I'm sure we would reciprocal activators,
+If we were genes, I'm sure we would be reciprocal activators,
 constantly stimulating each other to do just a little bit more!
@@
 Or maybe the other way around? Let's take turns being PC1 and PC2.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+\newperson
+\newperson
 Sofie, you saved my rear-end from embarrassing moments more frequently than I dare to admit.
@@
 "You do know that you have to present today, right?"
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+\newperson
+\newperson
 Helena, thank you for being such a lovely colleague and an even better friend.
@@
 on page 22-9 you probably already found the truc that I am referring to.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+\newperson
+\newperson
 Paulina, like a soft summer wind you passed through our lives.
-Thanks for putting so many healthy vege-tables and strange Polish delicacies in our fridge,
+Thanks for putting so many healthy “vege-tables” and strange Polish delicacies in our fridge,
 for cycling together to work, and for making weird noises at Remi.
 I'm looking forward to the next time we meet, so you can be amazed at how much Remi has grown.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Soup: Dani, Helena, Caroline, Robin B, Maxim, Annelies, Quentin, Katrien, Arne G, Ruth, Joris
-Dambi
-Dambi
-Dambi
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Miriam
-Miriam
-Miriam
-Miriam
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Laure, Michael, Jasper en Lize (spelling??)
-Mara, Seppe \& Alice
-Marta, Jan, Tristan, Rafael
+\newperson
+\newperson
+A big thank you to all the Dambini's for taking care of me while I was hard at work:
+An,ArS,ArG,Art,C,D,H,I,Ja,Jon,Jor,Ka,Ke,Li,Lo,M,N,Pa,Pi,Q,RVd,RB,Ru,Sa,Sl,So,St,W,Y.
+I'm grateful to be part of such a dynamic and fun group;
+you all make hard work at the VIB feel like a walk in the park.
+\newperson
+\newperson
+Miriam, thanks for always being there for us, 
+during the good times but also during times of need.
+As part of our family, there will always be a spot for you in our house.
+Let it be a while until one of us has to organise another reception.
+\newperson
+\newperson
+Because relaxing is as important as work,
+a special thanks to a selection of the finest friends.
+Laure, Michael, Jasper \& Lize for haut cuisine fun veggie nights.
+Joke, Niels \& Kai for exploring dungeons with trolls and lizards together.
+Marta, Jan, Tristan \& Rafaël for the many tasty brunches and classy parties.
+Mara Marrainne, Sepperd \& Alice for all the unforgettable family moments with our cute kids.
 Daan, Mieke \& Lou
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Cedric
-Cedric
-Cedric
-Cedric
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Familie
-Familie
-Familie
-Familie
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Victor, humor
-Victor
-Victor
-Victor
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Hendrik, getting into PC
-Hendrik 
-Hendrik
-Hendrik
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Papa Chris, babbel voor het slapen gaan, de vele schaakspelletjes
-Chris
-Chris
-Chris
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Mama en papa Jean-Paul: opvoeding, zorgeloosheid
-Mama
-Mama
-Mama
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Amor, as I write this, you are sitting on my lap and drooling on the table.
-I guesstimate about 50\% of this dissertation was written like this.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+Sophie and Marc for entertaining Caro while Elias, Charlotte and I play Minecraft together.
+\newperson
+\newperson
+Cedric, no time difference or distance can keep us apart,
+or change our habit to forget the world during our gaming and conversations. 
+놀라운 사람과 최고의 친구가되어 주셔서 감사합니다
+\newperson
+\newperson
+My warm thoughts go to:
+Vanessa, Marilou, Franco, Alain, Luc, Malika, Sien, Silke, Liam
+Anne, Patrick, Evelyne, Stefaan, Steven, Joke, Paul, Tania, Arne, Laure, Lukas, Lise, Dorian
+\newperson
+\newperson
+To mama and papa Jean Paul, thank you for all your love.
+You gave me a carefree youth and a good set of morals,
+and allowed me to study for as long as I liked.
+For all these things and so much more, I can never repay you.
+\newperson
+\newperson
+Papa Chris, I always enjoy talking to you, such as the countless
+chess games that we played, and the late night discussions by my bedside.
+You made me think for myself and have an opinion of my own.
+At the right time, I will pass on the torch to Remi.
+\newperson
+\newperson
+Victor, you're the most adventurous of us three brothers.
+Maybe you don't know yet what life will bring, and that's okay. None of us do.
+I enjoy driving other people absolutely nuts with our crazy talk.
+Thanks for all the memes :)
+\newperson
+\newperson
+Hendrik, as the one who I always looked up to, I blame you for getting me interested
+in computers. With you taking computers apart on the table, how could I resist.
+I know you paid dearly for this, though, since you would continuously have to
+repair the \textit{Windows Me} installation that I just destroyed.
+\newperson
+\newperson
+Ingeborg, Elke and Gallina, thank you for your hospitality and tasty food.
+You are all amazing women
+for all I know, dealing with a Cannoodt,
+requires a special kind of patience.
+\newperson
+\newperson
+To Amor, a big thank you for contributing countless cuddles and endless drool.
+I guesstimate at least 50\% of the dissertation was written with you on my lap.
+\newperson
+\newperson
 Hello Petit Remi! 
-Remi
-Remi
-Remi
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
+You're only seven months old, but it feels like I've known you for much longer already.
+Your spontaneous happiness when you see me come downstairs 
+from a hard day of work immediately makes my day.
+\newperson
+\newperson
 Dear Caro,
@@
 my strong independent woman.
+
+1

### 2019-12-10 ack.tex
update

--- 2019-12-10 ack.tex
@@
-Hello there! Did you stumble across this phrase by accident, 
-or are you looking for what I think you're looking for?
-For every page you turn over in this book,
-I feel that there is at least one person who deserves to be thanked 
-a thousand times over for their continued love, sacrifices and support.
-So here we go, until the end of the book: Thank you! Thank you!
-Thank you! Thank you! Thank you! Thank you! Thank you! Thank you! 
-Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
-Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
-Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
-Thank you! Thank you! Okay, enough with the silly business.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Yvan, you are an incredible supervisor and group leader. 
-You always tell the baby PhD students that the first step in a PhD
-is not learning to write well, but learning to like to write.
-Those words are amazing PhD training wheels.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Katleen, thank you for your dedication and perseverance
-in guiding me and making sure I got off to a good start in my PhD.
-I see exciting times ahead for the KDP group, which is overflowing
-with enthusiasm and inspiration. I wonder what the future will bring!
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Frank, thank you for giving me the opportunity to start on this journey.
-You are an inexhaustible source of new ideas
-which often explore the very thin line
-between ingenuity and completely insane.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Jo, sorry for spilling coffee on your shoes on my first day in Blok B.
-My strategy seemed to have worked, for suddenly I had a fourth promotor!
-I thoroughly enjoyed my time as an unofficial member of your group.
-You are a true people's person; you bring out the best in everybody.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Joeri, we've come full circle. You started me on this adventure
-with your incredible friendliness and that darned oversized pumpkin;
-and now, with you as my jury member, you helped me reach my goal.
-Let's stay in touch :)
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Wouter, I cannot imagine a better partner to do a PhD with.
-If we were genes, I'm sure we would reciprocal activators,
-constantly stimulating each other to do just a little bit more!
-Yet, our knowledge was often so perfectly complementary;
-if we were a PCA plot, you'd be PC1 and I'd be PC2.
-Or maybe the other way around? Let's take turns being PC1 and PC2.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Sofie, you saved my rear-end from embarrassing moments more frequently than I dare to admit.
-The perfect example was when I walked in the MMM on 2014-01-06,
-I arrived a little bit too late, and you send me a message:
-"You do know that you have to present today, right?"
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Helena, thank you for being such a lovely colleague and an even better friend.
-I enjoy our scientific talks and our non-scientific walks. 
-Together with you, I played the longest ongoing game of hide-and-seek,
-on page 22-9 you probably already found the truc that I am referring to.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Paulina, like a soft summer wind you passed through our lives.
-Thanks for putting so many healthy vege-tables and strange Polish delicacies in our fridge,
-for cycling together to work, and for making weird noises at Remi.
-I'm looking forward to the next time we meet, so you can be amazed at how much Remi has grown.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Soup: Dani, Helena, Caroline, Robin B, Maxim, Annelies, Quentin, Katrien, Arne G, Ruth, Joris
-Dambi
-Dambi
-Dambi
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Miriam
-Miriam
-Miriam
-Miriam
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Laure, Michael, Jasper en Lize (spelling??)
-Mara, Seppe \& Alice
-Marta, Jan, Tristan, Rafael
-Daan, Mieke \& Lou
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Cedric
-Cedric
-Cedric
-Cedric
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Familie
-Familie
-Familie
-Familie
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Victor, humor
-Victor
-Victor
-Victor
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Hendrik, getting into PC
-Hendrik 
-Hendrik
-Hendrik
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Papa Chris, babbel voor het slapen gaan, de vele schaakspelletjes
-Chris
-Chris
-Chris
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Mama en papa Jean-Paul: opvoeding, zorgeloosheid
-Mama
-Mama
-Mama
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Amor, as I write this, you are sitting on my lap and drooling on the table.
-I guesstimate about 50\% of this dissertation was written like this.
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Hello Petit Remi! 
-Remi
-Remi
-Remi
-\rule{10cm}{0.4pt}
-\rule{10cm}{0.4pt}
-Dear Caro,
-I cannot imagine a better partner and mother than you.
-You are a strange combination of being extremely thoughtful and caring,
-yet you roll over laughing when I stub my toe against the side of the table.
-Let's get back on track collecting doorsteps,
-and spending time doing the things we love, just us three.
-I look forward to the rest of our lives together,
-my strong independent woman.
+Hello there! Did you stumble across this phrase by accident,\\or are you looking for what I think you're looking for?
+For every page you turn over in this book,\\I feel that there is at least one person who deserves to be thanked 
+a thousand times over for their continued love, sacrifices and support.\\So here we go, until the end of the book: Thank you! Thank you!
+Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!\\Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
+Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!\\Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!
+Thank you! Thank you! Thank you! Thank you! Thank you! Thank you!\\Okay, enough with the silly business. 
+Go to the next chapter for more acknowledgements.
+
+Yvan, you are an incredible supervisor and group leader.\\You always tell the baby PhD students that the first step in a PhD
+is not learning to write well, but learning to like to write.\\Those words are amazing PhD training wheels.
+Katleen, thank you for your dedication and perseverance\\in guiding me and making sure I got off to a good start in my PhD.
+I see exciting times ahead for the KDP group, which is overflowing\\with enthusiasm and inspiration. I wonder what the future will bring!
+Frank, thank you for giving me the opportunity to start on this journey.\\You are an inexhaustible source of new ideas
+which often explore the very thin line\\between ingenuity and complete insanity.
+Jo, sorry for spilling coffee on your shoes on my first day in Blok B.\\My strategy seemed to have worked, for suddenly I had a fourth promotor!
+I thoroughly enjoyed my time as an unofficial member of your group.\\You are a true people's person; you bring out the best in everybody.
+Joeri, you brought me into this whole adventure\\with your incredible friendliness and that darned oversized pumpkin;
+and now, with you as my jury member, you helped me finish my PhD adventure.\\We've truly come full circle.
+Go to the next chapter for more acknowledgements.
+
+Wouter, I cannot imagine a better partner to do a PhD with. If we were genes,\\we'd be reciprocal activators, constantly stimulating each other to do just a little bit more!
+Yet, our knowledge was often so perfectly complementary; if we were a PCA plot,\\you'd be PC1 and I'd be PC2. Or the other way around? Let's take turns being PC1 and PC2.
+Sofie, you saved my rear-end from embarrassing moments more frequently than I dare\\ to admit. The perfect example was when I walked in the MMM on 2014-01-06,
+I arrived a little bit too late, and you send me a message:\\"You do know that you have to present today, right?"
+Helena, thank you for being such a lovely colleague and an even better friend.\\I enjoy our scientific talks and our non-scientific walks.
+Together with you, I played the longest ongoing game of hide-and-seek,\\on page 22-9 you probably already found the truc that I am referring to.
+Paulina, like a soft summer wind you passed through our lives.\\Thanks for putting so many healthy “vege-tables” and strange Polish delicacies in our fridge,
+for cycling together to work, and for making weird noises at Remi.\\I look forward to the next time we meet; you will be amazed at how much Remi has grown.
+A big thank you to all the Dambini's for taking care of me while I was hard at work:\\An,ArS,ArG,Art,C,D,H,I,Ja,Jon,Jor,Ka,Ke,Li,Lo,M,N,Pa,Pi,Q,RVd,RB,Ru,Sa,Sl,So,St,W,Y.
+I'm grateful to be part of such a dynamic and fun group;\\you make hard work at the VIB feel like a walk in the park.
+Go to the next chapter for more acknowledgements.
+
+Because relaxing is as important as work,\\a special thanks to a selection of the friends and family.
+Laure, Michael, Jasper \& Lize, thank you for the haut cuisine fun veggie nights.
+Joke, Niels \& Kai, thank you for exploring dungeons with trolls and lizards together.
+Marta, Jan, Tristan \& Rafaël, thank you for the many tasty brunches and classy parties.
+Mara Marrainne, Sepperd \& Alice, thank you for all\\the unforgettable family moments with our cute kids.
+Daan, Mieke \& Lou, thank you for taking matters in your own hands\\and upgrading our friendship status into family.
+Sophie and Marc, thank youfor entertaining Caro\\while Elias, Charlotte and I play Minecraft together. 
+Bart, thank you for being such a good friend\\and listening to me when I needed it the most.
+Miriam, thank you for always being there for us,\\during the good times but also during times of need.
+As part of our family, there will always be a spot for you in our house.\\Let it be a while until one of us has to organise another reception.
+Cedric, no time difference or distance can keep us apart,\\or change our habit to forget the world during our gaming and conversations. 
+놀라운 사람과 최고의 친구가되어 주셔서 감사합니다
+Big hugs for my family who supported me (Vanessa, Marilou, Franco, Alain,\\Luc, Malika, Sien, Silke, Liam, Anne, Patrick, Evelyne, Stefaan, Arne, Laure, 
+Paul, Tania, Dorian, Steven, Joke, Lukas, Lise), and often had to miss me at family parties.
+To mama and papa Jean Paul, thank you for all your love.\\You gave me a carefree youth and a good set of morals,
+and allowed me to study for as long as I liked.\\For all these things and so much more, I can never repay you.
+Papa Chris, I always enjoy talking to you, such as the countless\\chess games that we played, and the late night discussions by my bedside.
+You made me think for myself and have an opinion of my own.\\At the right time, I will pass on the torch to Remi.
+Victor, you're the most adventurous of us three brothers.\\Maybe you don't know yet what life will bring, and that's okay. None of us do.
+I enjoy driving other people absolutely nuts with our crazy talk.\\Thanks for all the memes :)
+Hendrik, as the one who I always looked up to, I blame you for getting me interested\\in computers. With you taking computers apart on the table, how could I resist.
+I know you paid dearly for this, though, since you would continuously have to\\repair the \textit{Windows Me} installation that I just destroyed.
+Ingeborg, Elke and Gallina, thank you for your hospitality and tasty food.\\You are all amazing women
+for all I know, dealing with a Cannoodt,\\requires a special kind of patience.
+Go to the next chapter for more acknowledgements.
+
+To Amor, a big thank you for contributing countless cuddles and endless drool.\\I guesstimate at least 50\% of the dissertation was written with you on my lap.
+
+Hello Petit Remi! You're only seven months old,\\but it feels like I've known you for much longer already.
+Your spontaneous happiness when you see me come downstairs\\from a hard day of work immediately makes my day.
+Dear Caro, I cannot imagine a better partner and mother than you.\\You make me want to be the best version of myself.
+Let's get back on track collecting doorsteps, and spending time doing the things we love,\\just us three. I look forward to the rest of my life together with my strong independent woman.
+
+1

### 2019-12-10 boek.tex
update
update acks

@@
 
+\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm,inner=2cm,outer=1.5cm]{geometry}
+\usepackage[
+	% set width and height to a4 width and height + 6mm
+	%width=16truecm, height=24truecm,
+	a4,
+	% use any combination of these options to add different cut markings
+	cam, axes, frame, cross,
+	% set the type of TeX renderer you use
+	pdftex,
+	% center the contents
+	center
+]{crop}
+
 %\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm,inner=2cm,outer=1.5cm]{geometry}
-%\usepackage[
-%	% set width and height to a4 width and height + 6mm
-%	%width=16truecm, height=24truecm,
-%	a4,
-%	% use any combination of these options to add different cut markings
-%	cam, axes, frame, cross,
-%	% set the type of TeX renderer you use
-%	pdftex,
-%	% center the contents
-%	center
-%]{crop}
-
-\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm,inner=2cm,outer=1.5cm]{geometry}
 % adjust margins (should be done before importing fancyhdr!) %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm
@@
 \usepackage{setspace}
+\usepackage{kotex} % hangul
 
@@
 
+%\#\theacki\ | % LINENUMBER DEBUG
 \newcounter{acki}
 \setcounter{acki}{1}
-\newcommand{\doack}{%
-%	{\footnotesize %
-%    %\#\theacki\ | % LINENUMBER DEBUG
-%		\partialinput{\theacki}{\theacki}{ack}
-%	}%
+
+\usepackage{makecell}
+\newcommand{\doackl}{%
+	\textcolor{med-gray}{\footnotesize\makecell[l]{\hspace*{-1.06em}\partialinput{\theacki}{\theacki}{ack}}}%
 	\stepcounter{acki}%
 }
+\newcommand{\doackr}{%
+\textcolor{med-gray}{\footnotesize\makecell[r]{\partialinput{\theacki}{\theacki}{ack}}}%
+\stepcounter{acki}%
+}
+
+%\newcommand{\doack}[1]{%
+%	\footnotesize\makecell[#1]{\partialinput{\theacki}{\theacki}{ack}}%
+%	\stepcounter{acki}%
+%}
+
+\newcommand{\newperson}{\rule{10cm}{0.4pt}}
 
@@
 	\fancyfoot[RO,LE]{\textsf{{\footnotesize Page \thepage}}}
-	\fancyfoot[LO]{\ifodd\theacki\doack\fi}
-	\fancyfoot[RE]{\ifodd\theacki\else\doack\fi}
-
+	\fancyfoot[LO]{\ifodd\theacki\else\doackl\fi}
+	\fancyfoot[RE]{\ifodd\theacki\doackr\fi}
 }
@@
 \\
-	\textbf{Results:} We developed \texttt{dyngen}, a multi-modality simulator of single cells. 
-	In \texttt{dyngen}, the biomolecular state of an \textit{in silico} cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
-By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures.
-We demonstrate \texttt{dyngen}'s flexibility by simulating snapshot, time-series and perturbation experiments.\\
-	\textbf{Conclusion:} \texttt{dyngen} lays the foundations for benchmarking a wide variety of computational single-cell tools, and can be used to help kick-start the development of future types of analyses.
+	\textbf{Results:} We developed dyngen, a multi-modality simulator of single cells.
+In dyngen, the biomolecular state of an \textit{in silico} cell changes over time according to a gene regulatory network.
+The simulator is extendible for adding new modalities or experimental procedures.
+We demonstrate dyngen's flexibility by applying it to benchmark RNA velocity and casewise network inference methods.
+\\
+	\textbf{Conclusion:} dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.
 \end{quote}
-
-\vfill
+%
+%\vfill
 {\large \textbf{Publication status}}
@@
 \end{quote}
-
-\vfill
+%
+%\vfill
 {\large \textbf{Author contributions}}
@@
 While already more than 75 trajectory inference methods have been developed, comprehensive benchmarking demonstrated that pseudotemporal ordering cells robustly is still a challenging task.\\
-	\textbf{Results:} \texttt{SCORPIUS} is a trajectory inference method specialised in pseudotemporally ordering single-cell profiles.
-Applying \texttt{SCORPIUS} to a large collection of datasets shows that it consistently produces more accurate and robust trajectories in comparison to state-of-the-art trajectory inference methods.
-We used \texttt{SCORPIUS} to generate novel hypotheses regarding dendritic cell development, which were subsequently validated \textit{in vivo}.\\
+	\textbf{Results:} {SCORPIUS} is a trajectory inference method specialised in pseudotemporally ordering single-cell profiles.
+Applying {SCORPIUS} to a large collection of datasets shows that it consistently produces more accurate and robust trajectories in comparison to state-of-the-art trajectory inference methods.
+We used {SCORPIUS} to generate novel hypotheses regarding dendritic cell development, which were subsequently validated \textit{in vivo}.\\
 	\textbf{Conclusion:} By providing a robust and accurate toolkit for inferring linear trajectories, we aim to make pseudotime analysis as easy-to-use and prevalent as other types of analysis such as differential expression and gene set enrichment.
@@
 \\
-	\textbf{Results:} We developed \texttt{dyno}, a toolkit for inferring, visualising and interpreting single-cell trajectories, giving access to more than 50 TI methods.
-\texttt{dyno} provides downstream analyses such as visualising the trajectory with dot-plots or heatmaps, automated labelling using gene sets, automated inference of the directionality, and detecting genes that are differentially expressed along the trajectory.
+	\textbf{Results:} We developed {dyno}, a toolkit for inferring, visualising and interpreting single-cell trajectories, giving access to more than 50 TI methods.
+{dyno} provides downstream analyses such as visualising the trajectory with dot-plots or heatmaps, automated labelling using gene sets, automated inference of the directionality, and detecting genes that are differentially expressed along the trajectory.
 \\
-	\textbf{Conclusion:} By providing a uniform interface for executing and visualising trajectories of 50 methods, \texttt{dyno} allows inferring and comparing trajectories for a much broader audience than previously possible.
+	\textbf{Conclusion:} By providing a uniform interface for executing and visualising trajectories of 50 methods, {dyno} allows inferring and comparing trajectories for a much broader audience than previously possible.
 \end{quote}
@@
 	Since regulatory interactions are context-dependent, attempting to model regulatory dynamics in the form of a single regulatory network may have little relevance. \\
-	\textbf{Results:} In this work, we introduce \texttt{bred}, the first algorithm for inferring case-wise regulatory networks.
+	\textbf{Results:} In this work, we introduce {bred}, the first algorithm for inferring case-wise regulatory networks.
 We generate .
@@
 \newpage{\thispagestyle{empty}\cleardoublepage}
-\chapter{incgraph: Optimising regulatory networks} 
+\chapter{incgraph: Optimising regulatory\\networks} 
 \chaptermark{incgraph: Optimising regulatory networks.} % short title
@@
 \begin{quote}
-	\textbf{Purpose:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
+	\textbf{Purpose:} Graphlets are small network patterns that can be counted in order to characterise the topology of a network.
 Counting graphlet can used to iteratively optimise a network in order to achieve certain topological properties.
-	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications.
-\\
-	\textbf{Results:} IncGraph calculates the differences in graphlet counts with respect to the network in its previous state.
-	IncGraph reduces the execution time by several orders of magnitude when compared to recounting the graphlet counts from scratch at every iteration.
+	Hitherto graphlets were not suited as a metric for performing topology optimisation due to the computational complexity of recalculating all the graphlet counts for each of the edge modifications. \\
+	\textbf{Results:} IncGraph calculates the differences in graphlet counts with respect to the network in its previous state, thereby reducing execution time by several orders of magnitude when compared to classic approaches.
 	We demonstrate the usefulness of this approach optimising gene regulatory networks using a graphlet-based fitness score. \\
-	\textbf{Conclusion:} IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation.
+	\textbf{Conclusion:} IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving networks, or develop graphlet-based criteria for topology optimisation.
 \end{quote}
@@
 	\textbf{Purpose:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
-Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
+Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
 However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results.
 \\
-	\textbf{Results:}  Here, we summarise key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
+	\textbf{Results:}  Here, we summarise key guidelines and recommendations for performing high-quality benchmarking analyses.
 Our review spans the full pipeline of benchmarking, from defining the scope to best practices for reproducibility.

### 2019-12-10 ch3_dynbenchmark/content.tex
update

@@
 
-We generated a panel of toy datasets (using our \href{https://github.com/dynverse/dyntoy}{\textit{dyntoy}} package, \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}) with all possible combinations of:
+We generated a panel of toy datasets (using our \href{https://github.com/dynverse/dyntoy}{\textit{dyntoy}} package, \url{https://github.com/dynverse/dyntoy}) with all possible combinations of:
 

### 2019-12-11 ack.tex
update

@@
 
-To Amor, a big thank you for contributing countless cuddles and endless drool.\\I guesstimate at least 50\% of the dissertation was written with you on my lap.
-
+To Amor, a big thank you for contributing countless cuddles and endless drool.
+I guesstimate at least 50\% of the dissertation was written with you on my lap.
 Hello Petit Remi! You're only seven months old,\\but it feels like I've known you for much longer already.
@@
 Dear Caro, I cannot imagine a better partner and mother than you.\\You make me want to be the best version of myself.
-Let's get back on track collecting doorsteps, and spending time doing the things we love,\\just us three. I look forward to the rest of my life together with my strong independent woman.
+Let's get back on track collecting doorsteps,\\and doing the things we love doing, just us three.
 

### 2019-12-11 ch10_discussion/content.tex
update

@@
 
-In \textbf{casewise NI}, one regulatory network is predicted for every single-cell in the dataset.
-Doing so allows studying the heterogeneity (or homogeneity) between and within different cell populations.
-In essence, most analyses that can be run on single-cell omics (e.g.
+In \textbf{casewise NI}, one regulatory network is predicted for every case (sample, patient, single cell) in the dataset.
+Such methodology allows generating hypotheses of deregulated interactions for individual patients, which could be used for drug re-purposing and personalised medicine.
+Furthermore, most analyses that can be run on single-cell omics (e.g.
 diffential expression) can also be applied on casewise regulatory networks (e.g.

### 2019-12-11 ch1_introduction/content.tex
update

@@
 \subsection{Cell types}
-Ever since Robert Hook first described the different structures of cells in 1665, biologists have been classifying cells by form and function.
+Ever since Robert Hook first described the different structures of cells in 1665, biologists have been classifying cells into different "cell types", by grouping them according to form and function.
 The human body is said to contain more than 210 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous.
@@
 One common approach for understanding the functionality of a particular cell is to observe which molecules are present in the cell and to associate those set of molecules with functionality.
-Taking a snap shot of the protein or RNA transcript content in a particular cell, might already provide us with major insights into its functionality.
+Taking a snapshot of the protein or RNA transcript content in a particular cell, might already provide us with major insights into its functionality.
 However, in order to fulfil a particular task, the biochemical machinery of the cell gradually changes over time.
@@
 	\centering
-	\includegraphics[width=.75\linewidth]{fig/selfassessment/self_assessment.pdf}
+	\includegraphics[width=.75\linewidth]{fig/selfassessment/self_assessment_edited.pdf}
 	\caption{
@@
 	\centering
-	\includegraphics[width=.75\linewidth]{fig/overview.pdf}
+	\includegraphics[width=.75\linewidth]{fig/overview_cropped.pdf}
 	\caption{Overview}

### 2019-12-11 ch3_dynbenchmark/content.tex
update

@@
 We found substantial complementarity between current methods, with different sets of methods performing most optimally depending on the characteristics of the data.
-For method users, we created an interactive set of guidelines (available at \href{http://guidelines.dynverse.org}{guidelines.dynverse.org}), which gives context-specific recommendations for method usage.
+For method users, we created an interactive set of guidelines (available at \url{http://guidelines.dynverse.org}), which gives context-specific recommendations for method usage.
 Our evaluation also highlights some challenges for current methods, and our evaluation strategy can be useful to spearhead the development of new tools that accurately infer trajectories on ever more complex use cases.
@@
 We found that only a handful of current methods can handle datasets with more than 10,000 cells within a reasonable time frame.
-To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method on the different measures proposed in this study at \href{https://benchmark.dynverse.org}{https://benchmark.dynverse.org}.
+To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method on the different measures proposed in this study at \url{https://benchmark.dynverse.org}.
 
@@
 Methods were excluded from the evaluation based on several criteria: (1) not freely available; (2) no code available; (3) superseded by another method; (4) requires data types other than expression; (5) no programming interface; (6) unresolved errors during wrapping; (7) too slow (requires more than 1 h on a 100 $\times$ 100 dataset); (8) does not return an ordering; and (9) requires additional user input during the algorithm (other than prior information).
-The discussions on why these methods were excluded can be found at \href{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}.
+The discussions on why these methods were excluded can be found at \url{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}.
 In the end, we included 45 methods in the evaluation.
@@
 
-To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and Singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}).
-These containers are automatically built and tested using Travis continuous integration (\href{https://travis-ci.org/dynverse}{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity.
+To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and Singularity containers (available at \url{https://methods.dynverse.org}).
+These containers are automatically built and tested using Travis continuous integration (\url{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity.
 For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts).
@@
 About one-third of the authors responded and we improved the wrappers based on their feedback.
-These discussions can be viewed on Github: \href{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}{https://github.com/dynverse/dynmethods/issues?q=label:method\_discussion}
+These discussions can be viewed on Github: \url{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}
 
@@
 This was then processed as a type 7, cell graph, method.
-	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\href{https://research.cchmc.org/pbge/slice.html}{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE.
+	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\url{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE.
 This was then processed as a type 1, direct, method.
@@
 
-More information on how each method was wrapped can be found within the comments of each wrapper script, listed at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.
+More information on how each method was wrapped can be found within the comments of each wrapper script, listed at \url{https://methods.dynverse.org}.
 
@@
 We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\hyperref[tab:supptable_2]{3.2}).
-The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}).
+The scripts to download and process these datasets are available on our repository (\url{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}).
 Whenever possible, we preferred to start from the raw counts data.
@@
 
-We used the same approach for selecting the best set of methods in the guidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}), developed using the R shiny package.
+We used the same approach for selecting the best set of methods in the guidelines app (\url{http://guidelines.dynverse.org}), developed using the R shiny package.
 This app will also filter the methods, among other things, depending on the predicted running time and memory requirements, the prior information available and the preferred execution environment (using the dynmethods package or standalone).
@@
 
-Further information on research design is available in the Nature Research Reporting Summary, available at \href{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}.
+Further information on research design is available in the Nature Research Reporting Summary, available at \url{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}.
 
@@
 
-The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_himglocalmetric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}).
+The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_himglocalmetric_2015} which was adopted from the R nettools package (\url{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}).
 It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network.
@@
 More specifically, we first calculated the geodesic distances for each cell to all milestones in the trajectory.
-Next, we trained a Random Forest regression model (implemented in the R \textit{ranger} package \cite{wright_rangerfastimplementation_2017}, \href{https://github.com/imbs-hl/ranger}{https://github.com/imbs-hl/ranger}) to predict these distances for each milestone, based on the expression of genes within each cell.
+Next, we trained a Random Forest regression model (implemented in the R \textit{ranger} package \cite{wright_rangerfastimplementation_2017}, \url{https://github.com/imbs-hl/ranger}) to predict these distances for each milestone, based on the expression of genes within each cell.
 We then extracted feature importances using the Mean Decrease in Impurity (importance = 'impurity' parameter of the ranger function), as illustrated in Figure \ref{fig:snote1fig_8}.
@@
 
-We generated a panel of toy datasets (using our \href{https://github.com/dynverse/dyntoy}{\textit{dyntoy}} package, \url{https://github.com/dynverse/dyntoy}) with all possible combinations of:
+We generated a panel of toy datasets (using our \textit{dyntoy} package, \url{https://github.com/dynverse/dyntoy}) with all possible combinations of:
 
@@
 An overview of the conformity of every metric is first given in Table \ref{tab:conformity_overview}.
-The individual rules and metric behaviour are discussed in the Supplementary Material that can be found at \href{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}.
+The individual rules and metric behaviour are discussed in the Supplementary Material that can be found at \url{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}.
 
@@
 We do however want to stress that different use cases will require a different overall score to order the methods.
-Such a context-dependent ranking of all methods is provided through the dynguidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}).
+Such a context-dependent ranking of all methods is provided through the dynguidelines app (\url{http://guidelines.dynverse.org}).
 

### 2019-12-11 scripts/color_pages.R

@@ -1,0 +1,24 @@
+library(tidyverse)
+
+col <- system("gs -o - -sDEVICE=inkcov 'boek_v2.0_20191211.pdf'", intern = TRUE)
+
+pagix <- grep("^Page [0-9]*$", col)
+colix <- grep("CMYK OK$", col)
+
+pages <- col[pagix] %>% gsub("Page ", "", .) %>% as.integer()
+colors <- col[colix] %>% gsub("^ *(.*) *CMYK OK$", "\\1", .) %>% strsplit("  *") %>% map(as.numeric) %>% do.call(rbind, .)
+colnames(colors) <- c("C", "M", "Y", "K")
+
+coldf <- data.frame(
+  page = pages,
+  colors
+)
+
+
+colpag <- coldf %>% filter(C > 0 | M > 0 | Y > 0)
+pages <- colpag$page
+diff(pages)
+
+
+length(pages)
+nrow(coldf) - length(pages)

### 2019-12-12 ch10_discussion/content.tex
update

@@
 The types of computational analyses presented in Section~\ref{sec:comptools} (Clustering, DE, DR, TI, NI) have all become routine methodologies for analysing and interpreting single-cell omics data.
-With projects like the Human Cell Atlas\cite{humancellatlasconsortium_humancellatlas_2018} soon to be churning out millions of single-cell profiles on a regular basis, research in high-throughput computational tools to analyse these data in an unsupervised approach becomes increasingly relevant.
+With projects like the Human Cell Atlas \cite{humancellatlasconsortium_humancellatlas_2018} soon to be churning out millions of single-cell profiles on a regular basis, research in high-throughput computational tools to analyse these data in an unsupervised approach becomes increasingly relevant.
 
@@
 
-Trajectory alignment has been used to compare gene expression kinetics resulting from different biological processes\cite{cacchiarelli_aligningsinglecelldevelopmental_2018}, to compare human, chimpanzee and macaque neuronal development\cite{kanton_organoidsinglecellgenomic_2019}, to find differences in gene regulation in the presence of certain growth factors\cite{mcfaline-figueroa_pooledsinglecellgenetic_2019}, and to compare human and mouse embryogenesis\cite{alpert_alignmentsinglecelltrajectories_2018}.
+Trajectory alignment has been used to compare gene expression kinetics resulting from different biological processes \cite{cacchiarelli_aligningsinglecelldevelopmental_2018}, to compare human, chimpanzee and macaque neuronal development \cite{kanton_organoidsinglecellgenomic_2019}, to find differences in gene regulation in the presence of certain growth factors \cite{mcfaline-figueroa_pooledsinglecellgenetic_2019}, and to compare human and mouse embryogenesis \cite{alpert_alignmentsinglecelltrajectories_2018}.
 

### 2019-12-12 ch1_introduction/content.tex
update

@@
 \section{Hippocratic oath for method developers}
-Before setting out on this perilous journey, I write a Hippocratic oath 
-which I believe all developers of computational tools should swear by 
-before haphazardly creating new tools.
+Before setting out on this perilous journey, I write a Hippocratic oath which I believe all developers of computational tools should swear by before haphazardly developing and publishing new software tools.
 
@@
 			\item I will compare the performance to relevant methods.
-			\item I will report positive achievements of my method, but also be critical of it, and discuss areas in which it performs suboptimally.
+			\item I will report positive achievements of my method, but also be critical of it, and discuss areas in which it performs sub-optimally
 			\item I will write unit tests to ensure that each component in my tool works as intended.
@@
 	\caption{
-		\textbf{A hippocratic oath for computational method developers.} Inspired by Laplante\cite{laplante_firstnoharm_2004} and lists of guidelines for benchmarking\cite{norel_selfassessmenttrapcan_2011,weber_essentialguidelinescomputational_2019}
+		\textbf{A Hippocratic oath for computational method developers.} Inspired by Laplante\cite{laplante_firstnoharm_2004} and lists of guidelines for benchmarking\cite{norel_selfassessmenttrapcan_2011,weber_essentialguidelinescomputational_2019}
 	}

### 2019-12-12 ch3_dynbenchmark/content.tex
update

@@
 Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
-Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}.
+Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}.
 The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.
 TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}.
-Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
+Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.
 
@@
 
-We used the Github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores.
+We used the GitHub issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores.
 About one-third of the authors responded and we improved the wrappers based on their feedback.
-These discussions can be viewed on Github: \url{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}
+These discussions can be viewed on GitHub: \url{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}
 
@@
 If available, this was directly extracted from the reference trajectory; otherwise the geodesic distance from the root milestone was used.
-For synthetic data, the simulation time was uniformily discretized into four timepoints.
+For synthetic data, the simulation time was uniformly discretised into four timepoints.
 	\item \textbf{Continuous time course}: for each cell a time point from which it was sampled.
@@
 The dyngen (Chapter~\ref{chap:dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment.
-At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable.
-We simulated a total of 110 datasets, with 11 different topologies.
+At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendible We simulated a total of 110 datasets, with 11 different topologies.
 
@@
 After simulating base gene expression with simulate\_base\_gene\_exp, we used the sample\_density function to finally simulate expression values of a number of cells (the same as from the reference real dataset), with $\alpha \sim \textit{Lognormal}$ ($\mu = 0.3$ and $\sigma = 1.5$) and $\beta \sim \textit{Lognormal}$ ($\mu = 2$ and $\sigma = 1.5$).
-Each of these parameters were centered around the default values of PROSSTT, but with enough variability to ensure a varied set of datasets.
+Each of these parameters were centred around the default values of PROSSTT, but with enough variability to ensure a varied set of datasets.
 

### 2019-12-12 scripts/color_pages.R

@@ -1,6 +1,6 @@
 library(tidyverse)
 
-col <- system("gs -o - -sDEVICE=inkcov 'boek_v2.0_20191211.pdf'", intern = TRUE)
+col <- system("gs -o - -sDEVICE=inkcov 'boek_v2.1_20191211.pdf'", intern = TRUE)
 
 pagix <- grep("^Page [0-9]*$", col)
 colix <- grep("CMYK OK$", col)
@@ -17,7 +17,7 @@ coldf <- data.frame(
 
 colpag <- coldf %>% filter(C > 0 | M > 0 | Y > 0)
 pages <- colpag$page
-diff(pages)
+paste(pages, collapse = ",")
 
 
 length(pages)

### 2019-12-12 ch5_dyno/content.tex
rename folders

\section{Introduction}

Recent technological advances allow unbiased investigation of cellular
dynamic processes in a high-throughput manner
 \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}.
Trajectory inference (TI) methods aim to give insight into a dynamic
process by inferring a trajectory from omics profiles of cells in which
the dynamic process takes place
 \cite{cannoodt_computationalmethodstrajectory_2016}. In a recent study,
we benchmarked 45 TI methods in terms of their accuracy, scalability,
stability, and
robustness \cite{saelens_comparisonsinglecelltrajectory_2019}. We
construct a set of guidelines to help end-users select an appropriate TI
method for their dataset of interest. However, executing and comparing
multiple methods remains challenging, due to high variability in
input/output data structures, software requirements, and programming
interfaces.

We developed {dyno}, a toolkit to easily infer, visualise and
interpret single-cell trajectories. The user can select the most optimal
set of TI methods based on characteristics of the dataset and user
preferences. More than 50 TI methods can easily be run within a common
interface, and the outputs thereof are converted into a common format.
{dyno} provides downstream analysis such as: visualising a
trajectory in a low-dimensional space or a heatmap, detecting genes
differentially expressed at different stages of the tragectory,
comparing multiple trajectories in a common dimensionality reduction,
and manipulating the trajectory such as adding directionality or
labelling different cell stages.

\section{Results}
This section will present the various components in the {dyno} workflow (Figure \ref{fig:dyno_toolkit}).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.8\linewidth]{fig/toolkit.pdf}
	\caption{
		\textbf{The {dyno} workflow for inferring, visualising and interpreting single-cell trajectories.}
		{dynguidelines} allows you to choose a suitable TI method for your dataset of interest.
		{dynmethods} contains wrappers for `r dynmethods::methods %>% filter(method_source != "control") %>% nrow()` TI methods.
		{dynwrap} provides main functionality for (pre/post-)processing trajectory data structures.
		{dynfeature} calculates gene importance scores based on a trajectory.
		{dynplot} provides visualisation tools for trajectories, including scatterplots and heatmaps.
	}
	\label{fig:dyno_toolkit}
\end{figure}

\subsection{Preparing the dataset}
In order to use your dataset of interest, you will first need to wrap it in a {dynwrap} object. We will use a commonly used dataset of 392 mouse embryonic fibroblasts (MEF) undergoing direct reprogramming to induced neuronal cells \cite{treutlein_dissectingdirectreprogramming_2016}. By wrapping the counts information and possible prior information (e.g. known cell types, a start cell, time course information) into a {dynwrap} object, the dataset can be used as input for the various {dyno} components.

\subsection{Selecting the best methods for a dataset}

We performed a comparative study of 45 trajectory inference
methods \cite{saelens_comparisonsinglecelltrajectory_2019}. We evaluated
each method in terms of four main aspects:

\begin{itemize}
	\tightlist
	\item \textbf{Accuracy}: How similar is the inferred trajectory to the
	``true'' (or ``expected'') trajectory in the data. We used several
	metrics in order to assess similarity in pairwise cellular ordering
	and also the similarity in topology of the trajectories. We used both
	real datasets -- which has the highest biological relevance -- and
	synthetic datasets -- which allow to push methods to their limits more
	easily.
	\item \textbf{Scalability}: How long the method takes to run and how much
	memory it consumes. This mainly depends on the dimensions of the input
	data, i.e.~the number of cells and features.
	\item \textbf{Stability}: How stable the results are when rerunning the
	method with slightly different input data.
	\item \textbf{Usability}: The quality of the corresponding documentation,
	software package, and manuscript. We assessed how easy it is to run
	the method, whether the software adheres to good software development
	practices, and whether the manuscript follows good scientific
	practices. We reviewed `good practices' literature and created a
	`consensus' score-sheet of good practices, and filled in to what extent
	each method adhered to each of the good practices.
\end{itemize}

We found a high diversity in method performance, and that not many
methods perform well across the board. The performance of a method
depended on many factors, mainly the dimensions of the data and the kind
of trajectory present in the data. Based on this, we developed an
interactive Shiny \cite{rstudio_webapplicationframework_2016} app which can be used to explore the results and
select an optimal set of methods for a particular analysis
(Figure \ref{fig:dynguidelines_interface}).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/shiny.pdf}
	\caption{\textbf{A static screenshot of the {dynguidelines} Shiny interface.}}
	\label{fig:dynguidelines_interface}
\end{figure}

\subsection{Inferring trajectories}
There are over 50 TI methods available as part of {dynmethods}, and executing a TI method on a dataset of interest is as simple as running just one command.
Behind the screens, {dyno} will save the dataset, prior
information and parameters as an H5 file and execute a container containing
the desired TI method. This
ensures that the TI method will not have any issues due to software
compatibilities on the host environment. The first time a TI method is
run, it will automatically download the corresponding container from
Docker Hub.

We inferred a trajectory on the dataset of mouse embryonic fibroblasts using the Slingshot \cite{street_slingshotcelllineage_2018} approach. The outputted model contains the main information on the trajectory, namely the milestone network and cell progressions (Table~\ref{tab:trajdata}). These data are hard to interpret manually. Luckily, {dyno} provides many different visualisation functions for interpreting trajectories.

\begin{table}
	\caption{The minimum amount of information to define a trajectory.}\label{tab:trajdata}
	\hfill
	\begin{minipage}[t]{.4\linewidth}
		\footnotesize
		\textbf{Milestone network:}

		\begin{tabular}{llrl}
			\toprule
			from & to & length & directed\\
			\midrule
			5 & 1 & 13.828399 & TRUE\\
			1 & 2 & 4.853443 & TRUE\\
			2 & 3 & 11.325468 & TRUE\\
			2 & 4 & 14.022826 & TRUE\\
			\bottomrule
		\end{tabular}
	\end{minipage}
	\hfill
	\begin{minipage}[t]{.4\linewidth}
		\footnotesize
		\textbf{Cell progressions:}

		\begin{tabular}{lllr}
		\toprule
		cell\_id & from & to & percentage\\
		\midrule
		1\_iN1\_C04 & 2 & 3 & 0.3508990\\
		1\_iN1\_C05 & 2 & 3 & 0.5287099\\
		1\_iN1\_C07 & 1 & 2 & 0.1953806\\
		1\_iN1\_C08 & 2 & 3 & 0.8589863\\
		1\_iN1\_C09 & 1 & 2 & 0.9076217\\
		1\_iN1\_C10 & 1 & 2 & 0.4170644\\
		1\_iN1\_C11 & 2 & 3 & 0.7316277\\
		1\_iN1\_C12 & 1 & 2 & 0.7108862\\
		1\_iN1\_C13 & 1 & 2 & 0.0644059\\
		1\_iN1\_C14 & 1 & 2 & 0.7178277\\
		\multicolumn{4}{c}{\ldots} \\
		\bottomrule
		\end{tabular}
	\end{minipage}
	\hfill\
\end{table}

\subsection{Execution details}

\paragraph{Prior information:} The method will error if it requires some prior information that is not
already provided in the dataset. Take note that the stronger the assumed priors for a given method, the more biased the output will be (Table~\ref{tab:prior}).

\begin{table}
	\caption{Possible prior information accepted or required by TI methods.}\label{tab:prior}
	\footnotesize
	\begin{tabularx}{\linewidth}{lXl}
		\toprule
		Name & Description & Type\\
		\midrule
		Start cell(s) & One or more start cell identifiers & soft\\
	  End cell(s) & One or more end cell identifiers & soft\\
		\# end states & The number of end states & soft\\
		\# start states & The number of start states & soft\\
		\# leaves & The number of leaves & soft\\
		\addlinespace
		Cell clustering & Named character vector linking the cell identifiers to different states/branches & hard\\
		\# states & Number of states/branches, including start, end and intermediary states & soft\\
		State network & Dataframe containing the known network between states/branches. Contains a from and to column & hard\\
		Time course (continuous) & Named numeric vector linking the cell ids to time points & hard\\
		Time course (discrete) & Named numeric vector linking the cell ids to time course points & hard\\
		\addlinespace
		Marker genes & Genes/features known to be important in the dynamic process & soft\\
		\bottomrule
	\end{tabularx}
\end{table}

\paragraph{Reproducibility:} When using this framework for analysing real data, remember to set the seed to ensure reproducibility.

\paragraph{Multiple executions:} Often it is useful to run multiple methods and/or use multiple datasets. While you can easily parallelise this yourself, we provide a helper
function which integrates well with our visualisation functions.

\paragraph{Errors:} Some methods can generate errors which are beyond our control. To know
what and where a method generated an error, remember to turn on the verbosity.

\paragraph{Command-line:} Each method is also executable from command-line using the Docker container. Check out the help documentation by running a TI method container without extra arguments.

\subsection{Visualising trajectories}

The most common way to visualise a trajectory is to plot it on a
dimensionality reduction of the cells (Figure~\ref{fig:dimred}A).
Often (but not always), a TI
method will already output a dimensionality reduction itself, which was
used to construct the trajectory. {dynplot} will use this
dimensionality reduction if available, otherwise it will calculate a
dimensionality reduction under the hood.
You can also supply it with your own dimensionality reduction. In this
case, the trajectory will be \emph{projected} onto this dimensionality
reduction (Figure~\ref{fig:dimred}B).
On this plot, you can colour the cells using various layers of information (Figure~\ref{fig:colorby}): cell ordering, cell grouping, feature expression, and pseudotime.

\begin{figure}[ht!]
  \centering
  A
  \includegraphics[width=.4\linewidth,valign=t]{manuscript_files/figure-latex/plotslingshot-1.pdf}
  B
  \includegraphics[width=.4\linewidth,valign=t]{manuscript_files/figure-latex/plotslingshot_umap-1.pdf}
  \caption{
  	\textbf{A:} Often, the TI method provides its own dimensionality reduction for the trajectory, which will be plotted by default. This is a visualisation of Slingshot \cite{street_slingshotcelllineage_2018} executed on MEF cells from the previous example.
  	\textbf{B:} Alternatively, you can provide your own dimensionality reduction, and the trajectory will be projected to it. In this example the Slingshot trajectory is projected to a UMAP dimensionality reduction \cite{mcinnes_umapuniformmanifold_2018}.
	}
\label{fig:dimred}
\end{figure}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=.6\linewidth]{manuscript_files/figure-latex/slingshot_multi-1.pdf}
  \caption{
		\textbf{Cell ordering:} Cells are coloured by their proximity to the different milestones.
		\textbf{Cell grouping:} A given character vector mapping a cell to a group.
		\textbf{Feature expression:} Colour by the expression levels of a particular gene.
		\textbf{Pseudotime:} The distance to a particular root milestone.
	}
	\label{fig:colorby}
\end{figure}

\paragraph{Various layout and colouring options:} To visualise a trajectory, it is good practice to take into account what the purpose and intended message of the visualisation is (Figure~\ref{fig:cells}).
The cells and trajectory can be positioned to place more emphasis on the topology of the inferred trajectory, or on the cellular heterogeneity between the cells, that the trajectory might or might not have been able to detect. Cells and trajectory can be coloured according to the
topology of the trajectory, according to gene expression, or a custom
input vector of values.

\begin{figure}[ht!]
	\centering
	\includegraphics{manuscript_files/figure-latex/cells-1.pdf}
	\caption{Overview of the different combinations of positioning and
		colouring options, demonstrated on the output of Slingshot \cite{street_slingshotcelllineage_2018}.}
	\label{fig:cells}
\end{figure}

\paragraph{Visualising many genes along a trajectory:}
A one-dimensional visualisation is especially useful if you combine it
with a heatmap (Figure~\ref{fig:heatmap}).

\begin{figure}[ht!]
	\centering
	\includegraphics[width=.6\linewidth]{manuscript_files/figure-latex/heatmap-1.pdf}
	\caption{Selecting relevant features for this heatmap is discussed in a later
		section, By default, features will be selected that best explain the main differences over the
		whole trajectory.}
	\label{fig:heatmap}
\end{figure}

\paragraph{Comparing multiple trajectories:}
Visualising each method with
their own dimensionality reduction can make it hard to interpret to what
extend the methods agree/disagree with each-other (Figure~\ref{fig:compare}A).
Different trajectories become noticeably more comparable when projected
to a common dimensionality reduction (Figure~\ref{fig:compare}B).

\begin{figure}[ht!]
	\centering
	\textbf{A}\includegraphics[width=.6\linewidth,valign=t]{manuscript_files/figure-latex/dimreds-1.pdf}\\\vspace*{1em}
	\textbf{B}\includegraphics[width=.6\linewidth,valign=t]{manuscript_files/figure-latex/dimredscommon-1.pdf}
	\caption{
		\textbf{A:} Comparing outputs of multiple TI methods is tedious when they are each visualised using their own visualisation method of interest.
		\textbf{B:} In contrast, by projecting each output to the same dimensionality reduction, the methods become immediately visually comparable.
	}
	\label{fig:compare}
\end{figure}

\subsection{Manipulating the trajectory}
{dyno} allows manipulating trajectories by simplifying, rooting and annotating them.

\paragraph{Simplifying linear segments:} Intermediate milestones can be removed by simplifying the trajectory (Figure~\ref{fig:annot}A).

\begin{figure}[ht!]
	\centering
	\textbf{A}\includegraphics[width=.7\linewidth,valign=t]{manuscript_files/figure-latex/unnamed-chunk-31-1.pdf} \\
	\textbf{B}\includegraphics[width=.7\linewidth,valign=t]{manuscript_files/figure-latex/rootmanual-1.pdf}\\
	\textbf{C}\includegraphics[width=.35\linewidth,valign=t]{manuscript_files/figure-latex/annotatemanual-1.pdf}
	\caption{
		\textbf{A:} Simplification of the trajectory.
		\textbf{B:} Rooting the trajectory.
		\textbf{C:} Labelling the trajectory.
	}
	\label{fig:annot}
\end{figure}

\paragraph{Rooting:} Most TI methods return undirected trajectories. We provide two ways of
rooting a trajectory, namely by manually specifying the root milestone, or by specifying highly expressed marker genes. After rooting, all other edges will point away from the root (Figure~\ref{fig:annot}B).

\paragraph{Annotating the trajectory:} Similar as with rooting, annotating the trajectory by renaming the
milestones can be done either manually, or with given highly expressed
gene sets (Figure~\ref{fig:annot}C).

\subsection{Differentially expressed genes along the trajectory}

Compared to differential expression between clusters of cells, defining
differential expression on trajectories is not so straightforward. What
constitutes a trajectory differentially expressed gene?

\begin{itemize}
\tightlist
\item A gene that is uniquely expressed in a particular branch?
\item A gene that changes at a branching point?
\item A gene that changes along pseudotime?
\end{itemize}

{dynfeature} allows you to find these different kinds of
differential expression in a trajectory. It first defines a particular
variable that needs to be predicted (for example, whether a cell is
present in a branch or not), and ranks each gene based on their
predictive capability with respect to that variable. This section
reviews the types of feature selection supported by {dynfeature}.

\paragraph{Lineage / transition marker genes:}
We can identify genes that are specifically upregulated or downregulated
at a specific branch (Figure~\ref{fig:fimp_tran}).

\begin{figure}
{\footnotesize
\begin{tabular}{lllr}
\toprule
feature\_id & from & to & importance\\
\midrule
Cav1 & 3 & 2 & 3.289198\\
S100a6 & 3 & 2 & 2.942703\\
1810020D17Rik & 2 & 5 & 2.822247\\
Tagln2 & 3 & 2 & 2.763266\\
Bex2 & 2 & 4 & 2.672312\\
\addlinespace
Anxa3 & 3 & 2 & 2.630013\\
Tm4sf1 & 3 & 2 & 2.547472\\
Vim & 3 & 2 & 2.431169\\
Hmga2 & 3 & 2 & 2.376114\\
Rrm2 & 3 & 2 & 2.346754\\
\bottomrule
\end{tabular}
}
\includegraphics[width=.6\linewidth,valign=m]{manuscript_files/figure-latex/branchheatmap-1.pdf}
\caption{Candidate markers genes for the MEF $\rightarrow$ transition.}
\label{fig:fimp_tran}
\end{figure}

\paragraph{Milestone marker genes:}
We can identify genes that are specifically upregulated or downregulated
at a particular milestone (Figure~\ref{fig:fimp_mil}).

\begin{figure}
	{\footnotesize
\begin{tabular}{llr}
\toprule
milestone\_id & feature\_id & importance\\
\midrule
2 & S100a6 & 2.654416\\
2 & Cav1 & 2.567051\\
2 & Tagln2 & 2.501074\\
2 & Fam198b & 2.334142\\
2 & Timp1 & 2.056474\\
\addlinespace
2 & Anxa3 & 2.009431\\
2 & Tm4sf1 & 1.982274\\
2 & Vim & 1.833045\\
2 & Bex2 & 1.738471\\
2 & Rrm2 & 1.678349\\
\bottomrule
\end{tabular}
}
\includegraphics[width=.6\linewidth,valign=m]{manuscript_files/figure-latex/milestoneimpheatmap-1.pdf}
\caption{Candidate markers genes for the 'induced' milestone.}
\label{fig:fimp_mil}
\end{figure}

\paragraph{Marker genes for the trajectory as a whole:}
We can identify genes that change in function of the ordering of a part
of the trajectory (Figure~\ref{fig:fimp_over}).

\begin{figure}
\hfill
	{\footnotesize
\begin{tabular}{lr}
\toprule
feature\_id & importance\\
\midrule
Tpm2 & 0.4539798\\
Tagln2 & 0.3364078\\
Vim & 0.3319566\\
Plaur & 0.3252199\\
Cav1 & 0.3224918\\
\addlinespace
Hmga2 & 0.3189412\\
Ptrf & 0.3180152\\
Tnnc2 & 0.3092800\\
Acta1 & 0.3087145\\
Myl1 & 0.3014400\\
\bottomrule
\end{tabular}
}\hfill
\includegraphics[width=.6\linewidth,valign=m]{manuscript_files/figure-latex/overallheatmap-1.pdf}
\caption{Candidate markers genes for the overall trajectory.}
\label{fig:fimp_over}
\end{figure}

\section{Discussion}
With {dyno}, we provide a feature-complete toolbox for inferring, visualising and annotating trajectory data. In this work, we demonstrated its usefulness by applying all of its visualisation and manipulation functionality on a particular dataset.

Alternative modalities such as ATAC-Seq or cytometry data are not yet
supported, although it is possible to simply include this data as
expression and counts.
RNA velocity \cite{lamanno_rnavelocitysingle_2018} would be particularly useful, as it would allow
rooting the directory without any further input from the user.

We are also planning to implement additional tools for interpreting the trajectories, such as alignment and differential expression. While the feature importance tools are incredibly useful towards interpreting trajectories, they do not yet provide a statistical ground to find features which are significantly
differentially expressed. In addition, depending on the size of the
dataset and of the predicted trajectory, it might take a long time to
compute.

### 2019-12-12 ch1_introduction/content.tex
update

@@
 \section{Hippocratic oath for method developers}
-Before setting out on this perilous journey, I write a Hippocratic oath which I believe all developers of computational tools should swear by before haphazardly developing and publishing new software tools.
+Before setting out on this perilous journey (i.e.
+the rest of this dissertation), I wrote a Hippocratic oath which I believe all developers of computational tools should swear by before haphazardly developing and publishing new software tools.
 

### 2019-12-12 changes.Rmd
update

---
output: pdf_document
---
# Overall

* Removed 'Paper x' prefix for all the chapters.

* Merged the content on Self-assessment in TI (formerly known as Chapter 8) into the introduction: "1.3 Benchmarking computational tools".

* Fixed many typo's and grammatical errors.

# Chapter 1: introduction

* Added overview figure to section 1.2 of the types of analyses and applications enabled by single-cell omics. For each of the types of analyses, the applications are discussed in more detailed in the respective sections.

* Rewrote research questions and objectives in terms of the three main topics in the thesis: Benchmarking, Trajectory Inference, and Network inference. Added overview diagram of which chapters fit in which group.

# Chapter 2: dyngen

* Added more results to this chapter; added detailed use-cases on how to use dyngen to benchmark RNA velocity (written by Wouter Saelens) and Casewise NI methods (written by me). Quantitative evaluation for either of these types of methods is completely novel. A section on using dyngen to benchmark trajectory alignment tools is ongoing but not included in the thesis.

* Expanded the captions of several figures.

* Reworked methods section on module networks and backbones.

* Expanded explanation on how cells are sampled, included an explanatory figure.

* Added use cases for dyngen. Wouter Saelens developed the use case for RNA velocity.

* Added several sections explaining the formulae in table 2.1.

* Added section on code availability.

# Chapter 3: dynbenchmark

* Removed duplicate section on dyngen's methdology.

* Fixed several layout issues.

* Added standalone captions for supplementary files that were not included in the chapter (see section 3.5)

# Chapter 9: discussion

* Completely rewrote this section. Instead of being a summary of the thesis, the discussion now discusses the impact of the work presented and ongoing developments in computational methods for single-cell omics.

# Backmatter

* Expanded curriculum vitae by listing conferences, oral presentations, poster presentations and master student supervisions.

* Added acknowledgement section.

### 2019-12-15 defence_public/pres_v1.tex
presentation wip

%!TEX TS-program = lualatex
%!TEX encoding = UTF-8 Unicode

%\documentclass[handout]{beamer}
\documentclass{beamer}

\usepackage{xcolor}
\definecolor{myblue}{rgb}{0,0.4,0.7}
\definecolor{myred}{rgb}{0.7,.1,.1}
\definecolor{mygray}{rgb}{0.7,0.7,0.7}
\definecolor{mygreen}{rgb}{0,0.8,0.1}

\useinnertheme{circles}

%\usefonttheme{serif}
%\usepackage{fontspec}
%\setmainfont [Path = ../fonts/,
%	UprightFont = *-300,
%	ItalicFont = *-300-Italic,
%	BoldFont = *-700,
%	BoldItalicFont = *-700-Italic
%]{MuseoSans}

\mode<presentation>{
  \setbeamertemplate{navigation symbols}{}
	\usepackage{lmodern}
	\setbeamercolor{title}{fg=myblue}
	\setbeamercolor{frametitle}{fg=black}
}

\mode<handout>{%
	\usepackage{pgfpages}
	\pgfpagesuselayout{2 on 1}[a4paper] 
	\setbeameroption{show notes}
	\usepackage{lmodern}
	\setbeamercolor{title}{fg=myblue}
	\setbeamercolor{frametitle}{fg=black}
}

\usepackage{graphicx}
\usepackage{transparent}
\usepackage{multimedia}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\usepackage[duration=40,starttime=16:10,lastminutes=5]{pdfpcnotes}

\title{Modelling single-cell dynamics with trajectories and gene regulatory networks}

\author{Robrecht Cannoodt}

\date{16 december 2019}

\begin{document}

{
	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{../cover/vertical4_blurred_small_trans.jpg}}
\begin{frame}[plain]
  \titlepage
  \hfill
  \includegraphics[width=\linewidth]{../fig/logos/mashup}
  \hfill
  \pnote{Goedenamiddag, bedankt voor de talrijke opkomst. In deze thesis voer ik onderzoek uit naar hoe cellen ontwikkelen met behulp van grote hoeveelheid data en computerprogramma's. Maar vooraleer ik exact uitleg wat ik gedaan heb, vertel ik graag eerst een beetje over de biologische insteek tijdens mijn doctoraat.}
\end{frame}
}

\begin{frame}
	\frametitle{Cellen zijn dynamisch}
	\begin{center}
		\only<1>{\includegraphics[width=.5\linewidth]{figures/1_cell/cell_1}}%
		\only<2>{\includegraphics[width=.5\linewidth]{figures/1_cell/cell_2}}%
		\only<3>{\includegraphics[width=.5\linewidth]{figures/1_cell/cell_3}}%
	\end{center}
	\pnote{Het begint bij de cel, die ook wel de basiseenheid van het leven worden genoemd. Cellen bevatten allerlei moleculen die zij nodig hebben om te kunnen reageren op alle soorten veranderingen in hun omgeving. Als een cel een signaal ontvangt, bijvoorbeeld wanneer een naburige cell iets probeert te zeggen, dan kan de cel daar op reageren door bijvoorbeeld nieuwe eiwitten van een bepaald type aan te maken.}
\end{frame}

\begin{frame}
	\frametitle{Hoe maken cellen eiwitten aan?}
	\begin{center}
		\only<1>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_1.pdf}}%
		\only<2>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_2.pdf}}%
		\only<3>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_3.pdf}}%
		\only<4>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_4.pdf}}%
		\only<5>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_5.pdf}}%
		\only<6>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_6.pdf}}%
		\only<6>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_7.pdf}}%
	\end{center}

\end{frame}

\begin{frame}
	\frametitle{Het menselijk lichaam bevat >210 soorten cellen}
	\begin{center}
		\only<1>{\includegraphics[width=.6\linewidth]{figures/3_celtypes/celtypes.jpg}}%
	\end{center}
	\uncover<1>{(placeholder figuur)}

\end{frame}

\begin{frame}
	\frametitle{Cellen kunnen ontwikkelen van één type naar een ander}
	\begin{center}
		\only<1->{\includegraphics[width=.8\linewidth]{figures/3_celtypes/lineagetree.pdf}}%
	\end{center}

\end{frame}

\begin{frame}
	\frametitle{Welke types cellen komen voor in een populatie?}
	\begin{center}
		\only<1>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/scomics_1.pdf}}%
		\only<2>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/scomics_2.pdf}}%
		\only<3>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/scomics_3.pdf}}%
		\only<4>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/scomics_4.pdf}}%
	\end{center}

\end{frame}

\begin{frame}
	\frametitle{2014: doorbraak in schaalbaarheid van scRNA-seq}
	\begin{center}
		\only<1>{\includegraphics[width=\linewidth]{figures/4_singlecellomics/scaling.pdf}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Voorbeeld van een scRNA-seq technologie: Drop-seq}
  \begin{center}
		\movie{\includegraphics[width=\textwidth]{figures/4_singlecellomics/DropSeq_thumbnail.jpg}}{figures/4_singlecellomics/DropSeq.avi}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Moleculen tellen met scRNA-seq}
	\begin{center}
		\only<1>{\includegraphics[width=.8\linewidth]{figures/4_singlecellomics/rnaseq_1.pdf}}%
		\only<2>{\includegraphics[width=.8\linewidth]{figures/4_singlecellomics/rnaseq_2.pdf}}%
		\only<3>{\includegraphics[width=.8\linewidth]{figures/4_singlecellomics/rnaseq_3.pdf}}%
		\only<4>{\includegraphics[width=.8\linewidth]{figures/4_singlecellomics/rnaseq_4.pdf}}%
		\only<5>{\includegraphics[width=\linewidth]{figures/4_singlecellomics/rnaseq_5.pdf}}%
	\end{center}

\end{frame}

\begin{frame}
	\frametitle{Cellen ordenen met SCORPIUS}
	\begin{center}
		\includegraphics[width=.35\linewidth]{figures/4_singlecellomics/rnaseq_plot.pdf}%
		\uncover<2->{%
			\only<1-2>{\includegraphics[width=.35\linewidth]{figures/5_scorpius/dimred.pdf}}%
			\only<3->{\movie{\includegraphics[width=.35\linewidth]{figures/5_scorpius/princurve_it0.png}}{figures/5_scorpius/princurve_loop.avi}}%
			\hspace{.116666\linewidth}
		}\\%
		\uncover<4->{\includegraphics[width=.35\linewidth]{figures/5_scorpius/expr_time.pdf}}%
		\uncover<5->{\includegraphics[width=.46666\linewidth]{figures/5_scorpius/heatmap.pdf}}%
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Niet alle datasets zijn even gemakkelijk}
	\begin{center}
		\only<1>{\includegraphics[width=\linewidth]{figures/6_multifurcation/plots_1.pdf}}%
		\only<2-3>{\includegraphics[width=\linewidth]{figures/6_multifurcation/plots_2.pdf}}%
		\only<4->{\includegraphics[width=\linewidth]{figures/6_multifurcation/plots_3.pdf}}%
	\end{center}
  \uncover<3->{Trajectory inference; >70 methoden}\\
  \uncover<5->{Nood aan een vergelijking van methoden}
\end{frame}

\begin{frame}
	\frametitle{Simulatie van virtuele cellen}
	\begin{center}
		\movie{\includegraphics[width=.6\linewidth]{figures/7_dyngen/insilico_init.png}}{figures/7_dyngen/insilico.avi}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Simulatie van virtuele cellen}
	\begin{center}
		\includegraphics[width=\linewidth]{figures/7_dyngen/example_runs.pdf}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Resultaten benchmark}
\end{frame}

\begin{frame}
	\frametitle{Guidelines}
\end{frame}

\begin{frame}
	\frametitle{Conclusies benchmark paper}
\end{frame}

\begin{frame}
	\frametitle{Toekomstvisie}
	\begin{itemize}
		\item Werk Helena rond GNG (toon filmpje)
		\item Trajectory alignment
		\item Combinatie met spatial omics
		\item Human Cell Atlas (vergelijk met de human genome project)
	\end{itemize}
\end{frame}

\end{document}

### 2019-12-15 defence_public/figures/4_singlecellomics/DropSeq_process.sh

@@ -1,0 +1,9 @@
+#!/bin/sh
+# orig: 640x480
+
+ffmpeg -i DropSeq_original.mp4 -filter:v "crop=260:480:190:0,transpose=2" DropSeq.mp4
+ffmpeg -i DropSeq.mp4 -r 1 -s 480x260 -f image2 foo-%03d.jpg
+ffmpeg -i DropSeq.mp4 -vcodec copy -acodec copy DropSeq.avi
+mv foo-001.jpg DropSeq_thumbnail.jpg
+rm foo*
+

### 2019-12-15 defence_public/scripts/6_multifurcation.R

@@ -1,0 +1,44 @@
+library(tidyverse)
+library(dyno)
+
+set.seed(6)
+dat <- dyntoy::generate_dataset(id = "defence", model = dyntoy::model_multifurcating(num_branchpoints = 2, max_degree = 3), num_cells = 200, num_features = 101, allow_tented_progressions = FALSE, add_velocity = FALSE)
+
+dimred <- dyndimred::dimred_mds(dat$expression) %>% dynutils::scale_uniform()
+plot_dimred(dat, dimred = dimred)
+
+gro <- setNames(rep("NA", length(dat$cell_ids)), dat$cell_ids)
+g1 <-
+  dynplot::plot_dimred(dat, dimred = dimred, plot_trajectory = FALSE, color_cells = "grouping", grouping = gro, size_cells = 1.5, size_milestones = 4) +
+  theme_classic() + labs(x = "Component 1", y = "Component 2") +
+  scale_x_continuous(breaks = NULL, limits = c(-.5, .5)) + scale_y_continuous(breaks = NULL, limits = c(-.5, .5)) +
+  scale_colour_manual(values = c("NA" = "darkgray")) + theme(legend.position = "none")
+
+methods <- list(
+  "Monocle DDRTree" = ti_monocle_ddrtree(),
+  "PAGA" = ti_paga_tree(),
+  "Slingshot" = ti_slingshot(),
+  "CellTrails" = ti_celltrails(),
+  "cellTree" = ti_celltree_maptpx(),
+  "DPT" = ti_dpt()
+)
+
+traj_outs <- map(names(methods), function(method_name) {
+  method <- methods[[method_name]]
+  traj <- infer_trajectory(dat, method)
+  plot <- dynplot::plot_dimred(traj, dimred = dimred, color_cells = "grouping", grouping = gro, size_cells = 1.5, size_milestones = 4) +
+    labs(title = method_name) +
+    scale_colour_manual(values = c("NA" = "lightgray")) + theme(legend.position = "none")
+  lst(traj, plot)
+})
+gs <- patchwork::wrap_plots(map(traj_outs, "plot"), nrow = 2)
+
+
+g2 <-
+  dynplot::plot_dimred(dat, dimred = dimred, color_cells = "grouping", grouping = gro, size_cells = 1.5, size_milestones = 4) + theme_classic() + labs(x = "Component 1", y = "Component 2") +
[…]

### 2019-12-15 defence_public/scripts/7_dyngen.R

@@ -1,0 +1,73 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(3)
+
+backbone <- backbone_bifurcating()
+
+model <-
+  initialise_model(
+    num_cells = 300,
+    num_tfs = 50,
+    num_targets = 200,
+    num_hks = 0,
+    distance_metric = "pearson",
+    backbone = backbone,
+    tf_network_params = tf_network_default(min_tfs_per_module = 1, sample_num_regulators = function() 2),
+    feature_network_params = feature_network_default(target_resampling = 5000),
+    kinetics_params = kinetics_default(),
+    gold_standard_params = gold_standard_default(),
+    simulation_params = simulation_default(
+      census_interval = .1,
+      burn_time = 3,
+      total_time = 10,
+      experiment_params = bind_rows(
+        simulation_type_wild_type(num_simulations = 40),
+        simulation_type_knockdown(num_simulations = 0)
+      )
+    ),
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8
+  )
+
+model <- model %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+model <- model %>%
+  generate_gold_standard()
+
+plot_gold_simulations(model)
+
+model <- model %>% generate_cells()
+
+plot_simulation_expression(model, 1, what = "x")
+plot_gold_mappings(model, do_facet = FALSE) + scale_colour_brewer(palette = "Dark2")
+
+simcounts <- model$simulations$counts
+dr <- dyndimred::dimred_landmark_mds(simcounts, distance_method = "spearman")
+dim(dr)
+
+plot(dr)
+
+drdf <-
+  bind_cols(
+    model$simulations$meta,
+    as.data.frame(dr)
+  )
+
+g <-
+  ggplot(drdf %>% filter(sim_time >= 0)) +
+  geom_path(aes(comp_1, comp_2, group = simulation_i), function(df) df %>% select(-sim_time), colour = "darkgray") +
+  geom_point(aes(comp_1, comp_2), colour = "red") +
+  gganimate::transition_time(sim_time) +
+  theme_classic() +
+  labs(x = "Component 1", y = "Component 2") +
+  scale_x_continuous(breaks = NULL) +
+  scale_y_continuous(breaks = NULL)
[…]

### 2019-12-15 defence_public/scripts/rnaseq.R

@@ -1,0 +1,61 @@
+library(tidyverse)
+
+fig_size <- 3
+
+melt <- function(mat) {
+  mat %>%
+    magrittr::set_colnames(NULL) %>%
+    magrittr::set_rownames(NULL) %>%
+    reshape2::melt(varnames = c("y", "x"), value.name = "value") %>%
+    as_tibble()
+}
+raster <- function(meltdf, legend = FALSE, palette = "RdYlBu") {
+  g <-
+    ggplot(meltdf) +
+    geom_raster(aes(x, y, fill = value)) +
+    scale_fill_distiller(palette = palette, breaks = c(0, 1), labels = c("Min", "Max")) +
+    dynplot::theme_graph() +
+    labs(fill = "Expression") +
+    coord_cartesian(expand = FALSE)
+  if (!legend) {
+    g <- g + theme(legend.position = "none")
+  }
+  g
+}
+
+
+dat <- dyntoy::generate_dataset(id = "defence", model = "multifurcating", num_cells = 100, num_features = 101)
+
+counts <-
+  dat$counts %>%
+  as.matrix()
+colnames(counts) <- gsub("G", "gene ", colnames(counts))
+rownames(counts) <- gsub("C", "cell ", rownames(counts))
+write_tsv(data.frame(counts, check.names = FALSE) %>% rownames_to_column("cell name"), "figures/4_singlecellomics/rnaseq_matrix.tsv")
+
+g <-
+  counts %>%
+  melt() %>%
+  raster()
+
+ggsave("figures/4_singlecellomics/rnaseq_plot.pdf", g, width = 3, height = 3)
+
+
+countsmelt <-
+  reshape2::melt(counts, varnames = c("cell", "gene"), value.name = "count")
+countsmelt_sel <-
+  countsmelt %>%
+  filter(cell == "cell 1", gene %in% paste0("gene ", 1:5))
+g <- ggplot(countsmelt_sel) + geom_bar(aes(gene, count), stat = "identity") + theme_classic() + labs(x = NULL, y = "Aantal RNA transcripten") + coord_flip()
+ggsave("figures/4_singlecellomics/rnaseq_2dplot.pdf", g, width = 3, height = 3)
+
+
+g <-
+  counts %>%
+  melt() %>%
+  filter(y == 1, x %in% 1:5) %>%
+  raster()
+
[…]

### 2019-12-15 defence_public/scripts/scorpius.R

@@ -1,0 +1,64 @@
+library(SCORPIUS)
+library(tidyverse)
+library(gganimate)
+
+set.seed(1)
+
+data <- SCORPIUS::ginhoux
+expr <- data$expression
+sample_info <- data$sample_info
+
+dimred <- SCORPIUS::reduce_dimensionality(expr, ndim = 2)
+
+g2 <-
+  draw_trajectory_plot(dimred, point_size = 1) +
+  scale_x_continuous(breaks = -Inf) +
+  scale_y_continuous(breaks = -Inf)
+ggsave("figures/5_scorpius/dimred.pdf", g2, width = 3, height = 3)
+
+its <- 0:10
+walk(its, function(it) {
+  set.seed(1)
+  traj <- SCORPIUS::infer_trajectory(dimred, maxit = it, k = 0)
+
+  pl <- g2 +
+    geom_path(aes(Comp1, Comp2), as.data.frame(traj$path), colour = "blue") +
+    labs(title = paste0("Iteration ", it))
+  # ggsave(paste0("figures/5_scorpius/princurve_it", it, ".pdf"), pl, width = 3, height = 3)
+  ggsave(paste0("figures/5_scorpius/princurve_it", it, ".png"), pl, width = 3, height = 3)
+})
+
+system("convert figures/5_scorpius/princurve_it*.png -delay 1000 figures/5_scorpius/princurve_loop.gif")
+file.remove("figures/5_scorpius/princurve_loop.avi"); system("ffmpeg -f image2 -r 1 -i figures/5_scorpius/princurve_it%d.png figures/5_scorpius/princurve_loop.avi")
+
+set.seed(1)
+traj <- SCORPIUS::infer_trajectory(dimred, maxit = 10, k = 0)
+
+ix <- sample.int(nrow(expr), 50)
+gimp <- SCORPIUS::gene_importances(expr, traj$time)
+
+expr_df <- data.frame(cell = names(traj$time),time = traj$time, as.matrix(expr[,c("Cd74", "Mpo", "Prtn3", "Siglech")])) %>%
+  gather(gene, value, -time, -cell)
+g6 <-
+  ggplot(expr_df) +
+  geom_point(aes(time, value), size = 1) +
+  geom_smooth(aes(time, value), colour = "blue", fill = NA, size = 1) +
+  facet_wrap(~gene, scales = "free") +
+  theme_classic() +
+  scale_x_continuous(breaks = -Inf) +
+  scale_y_continuous(breaks = -Inf) +
[…]

### 2019-12-15 chX_acknowledgements.tex
update
add space

--- 2019-12-11 ack.tex
@@
 Daan, Mieke \& Lou, thank you for taking matters in your own hands\\and upgrading our friendship status into family.
-Sophie and Marc, thank youfor entertaining Caro\\while Elias, Charlotte and I play Minecraft together. 
+Sophie and Marc, thank you for entertaining Caro\\while Elias, Charlotte and I play Minecraft together. 
 Bart, thank you for being such a good friend\\and listening to me when I needed it the most.
@@
 As part of our family, there will always be a spot for you in our house.\\Let it be a while until one of us has to organise another reception.
-Cedric, no time difference or distance can keep us apart,\\or change our habit to forget the world during our gaming and conversations. 
+Cedric, no time difference or distance can keep us apart,\\or change our habit of forgetting the world around us during our gaming and conversations. 
 놀라운 사람과 최고의 친구가되어 주셔서 감사합니다

### 2019-12-16 defence_public/pres_v1.tex
update

@@
 		\only<6>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_6.pdf}}%
-		\only<6>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_7.pdf}}%
+		\only<7>{\includegraphics[width=\linewidth]{figures/2_centraldogma/central_dogma_7.pdf}}%
 	\end{center}
-
 \end{frame}
@@
 	\end{center}
+\end{frame}
 
+\begin{frame}
+	\frametitle{Overzicht van wetenschappelijke contributies in thesis}
+	\begin{center}
+		\only<1>{\includegraphics[width=.8\linewidth]{figures/overview_1.pdf}}%
+		\only<2>{\includegraphics[width=.8\linewidth]{figures/overview_2.pdf}}%
+	\end{center}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Resultaten benchmark}
+	\frametitle{Grootschalige evaluatie van trajectory inference methoden}
+	\begin{center}
+		\only<1>{\includegraphics[width=.9\linewidth]{figures/8_benchmark/benchmark_1.pdf}}%
+		\only<2>{\includegraphics[width=.9\linewidth]{figures/8_benchmark/benchmark_2.pdf}}%
+		\only<3>{\includegraphics[width=.9\linewidth]{figures/8_benchmark/benchmark_3.pdf}}%
+		\only<4>{\includegraphics[width=.9\linewidth]{figures/8_benchmark/benchmark_4.pdf}}%
+		\only<5->{\includegraphics[width=.9\linewidth]{figures/8_benchmark/benchmark_5.pdf}}%
+	\end{center}
+	\uncover<6->{}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Guidelines}
+	\frametitle{Resultaten van evaluatie}
+	\begin{center}
+		\includegraphics[height=.85\textheight]{figures/8_benchmark/5b_benchmark.pdf}
+	\end{center}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Conclusies benchmark paper}
+	\frametitle{Set van richtlijnen voor gebruikers}
+	\begin{center}
+		\includegraphics[width=\linewidth]{figures/8_benchmark/6_guidelines.pdf}%
+	\end{center}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Eliminatie van sommige methoden}
+	\begin{center}
+		\only<1>{\includegraphics[width=\linewidth]{figures/8_benchmark/aftermath_1.pdf}}%
+		\only<2>{\includegraphics[width=\linewidth]{figures/8_benchmark/aftermath_2.pdf}}%
+	\end{center}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Gevolgen benchmark paper}
+	\begin{center}
+		\only<1>{\includegraphics[width=.65\linewidth]{figures/8_benchmark/caution.pdf}}%
+	\end{center}
 \end{frame}

### 2019-12-16 defence_public/scripts/7_dyngen.R
update

@@ -61,7 +61,7 @@ drdf <-
 g <-
   ggplot(drdf %>% filter(sim_time >= 0)) +
   geom_path(aes(comp_1, comp_2, group = simulation_i), function(df) df %>% select(-sim_time), colour = "darkgray") +
-  geom_point(aes(comp_1, comp_2), colour = "red") +
+  geom_point(aes(comp_1, comp_2), colour = "red", size = 4) +
   gganimate::transition_time(sim_time) +
   theme_classic() +
   labs(x = "Component 1", y = "Component 2") +

### 2019-12-16 chX_acknowledgements.tex
update thesis

@@
 Big hugs for my family who supported me (Vanessa, Marilou, Franco, Alain,\\Luc, Malika, Sien, Silke, Liam, Anne, Patrick, Evelyne, Stefaan, Arne, Laure, 
-Paul, Tania, Dorian, Steven, Joke, Lukas, Lise), and often had to miss me at family parties.
+Paul, Gilbert, Linda, Tania, Dorian, Steven, Joke, Lukas, Lise), and often had to miss me at family parties.
 To mama and papa Jean Paul, thank you for all your love.\\You gave me a carefree youth and a good set of morals,

### 2019-12-16 defence_public/pres_v1.tex
update

@@
 
-%\documentclass[handout]{beamer}
 \documentclass{beamer}
@@
 
-%\usefonttheme{serif}
-%\usepackage{fontspec}
-%\setmainfont [Path = ../fonts/,
-%	UprightFont = *-300,
-%	ItalicFont = *-300-Italic,
-%	BoldFont = *-700,
-%	BoldItalicFont = *-700-Italic
-%]{MuseoSans}
+\usepackage{pdfrender} % for borders
 
@@
   \setbeamertemplate{navigation symbols}{}
-	\usepackage{lmodern}
-	\setbeamercolor{title}{fg=myblue}
-	\setbeamercolor{frametitle}{fg=black}
-}
-
-\mode<handout>{%
-	\usepackage{pgfpages}
-	\pgfpagesuselayout{2 on 1}[a4paper] 
-	\setbeameroption{show notes}
 	\usepackage{lmodern}
@@
 
-\usepackage[duration=40,starttime=16:10,lastminutes=5]{pdfpcnotes}
+\usepackage{pdfpcnotes}
 
@@
 \begin{frame}
-	\frametitle{Cellen zijn dynamisch}
+%	\frametitle{Cellen zijn dynamisch}
 	\begin{center}
@@
 	\end{center}
-	\uncover<1>{(placeholder figuur)}
-
 \end{frame}
@@
 \begin{frame}
+	\frametitle{Veel voorkomende vraag in celbiologie}
+	\begin{center}
+		\only<1>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/vragen_1.pdf}}%
+		\only<2>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/vragen_2.pdf}}%
+		\only<3>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/vragen_3.pdf}}%
+	\end{center}
+\end{frame}
+
+\begin{frame}
 	\frametitle{Welke types cellen komen voor in een populatie?}
@@
 		\only<2>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/scomics_2.pdf}}%
-		\only<3>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/scomics_3.pdf}}%
-		\only<4>{\includegraphics[width=.9\linewidth]{figures/4_singlecellomics/scomics_4.pdf}}%
 	\end{center}
-
 \end{frame}
@@
 	\end{center}
+	\begin{flushright}
+		{\footnotesize Treutlein et al. 2018}
+	\end{flushright}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Overzicht van wetenschappelijke contributies in thesis}
+	\frametitle{Opbouw van de thesis}
 	\begin{center}
@@
 		\only<2>{\includegraphics[width=.8\linewidth]{figures/overview_2.pdf}}%
+	\end{center}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Yvan in 2015:\\\uncover<2->{"hé Robrecht, kan je deze cellen even ordenen?"}}
+	\begin{center}
+		\only<1>{\includegraphics[width=.8\linewidth]{figures/5_scorpius/vraag_1.pdf}}%
+		\only<2>{\includegraphics[width=.8\linewidth]{figures/5_scorpius/vraag_2.pdf}}%
 	\end{center}
@@
 \begin{frame}
-	\frametitle{Niet alle datasets zijn even gemakkelijk}
+	\frametitle{Niet alle datasets zijn even gemakkelijk\\\uncover<2->{$\rightarrow$ Trajectory inference methoden}}
 	\begin{center}
@@
 	\end{center}
-  \uncover<3->{Trajectory inference; >70 methoden}\\
+  \uncover<3->{Zeer pupulair: >70 methoden}\\
   \uncover<5->{Nood aan een vergelijking van methoden}
@@
 \begin{frame}
-	\frametitle{Simulatie van virtuele cellen}
+	\frametitle{dyngen: Simulatie van virtuele cellen}
 	\begin{center}
@@
 \begin{frame}
+	\frametitle{Uitdagingen}
+	\begin{itemize}
+		\item<2-> 45 verschillende tools op meerdere computers (Docker)
+		\item<3-> $\pm$ 100 echte datasets, $\pm$ 100 gesimuleerde datasets
+		\item<4-> Nog geen metrieken
+	\end{itemize}
+\end{frame}
+
+\begin{frame}
 	\frametitle{Resultaten van evaluatie}
@@
 \begin{frame}
-	\frametitle{Gevolgen benchmark paper}
+	\frametitle{Tot nader order: pas meerdere methoden toe en vergelijk}
 	\begin{center}
@@
 	\end{center}
+	\begin{flushright}
+		{\footnotesize Plass et al. 2018}
+	\end{flushright}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Toekomstvisie}
+	\frametitle{Relevantie van dit werk}
+	\begin{columns}
+		\begin{column}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{figures/overview_2.pdf}
+		\end{column}
+		\begin{column}{0.5\textwidth}
+			\uncover<2->{
+				Evaluatie:
 				\begin{itemize}
-		\item Werk Helena rond GNG (toon filmpje)
-		\item Trajectory alignment
-		\item Combinatie met spatial omics
-		\item Human Cell Atlas (vergelijk met de human genome project)
+					\item Wordt gebruikt in onderzoek en industrie 
+					\item Vaak vermeld in artikels en presentaties
 				\end{itemize}
+			}
+			\uncover<3->{
+				Simulator: 
+				\begin{itemize}
+					\item Evaluatie van andere soorten analyses in single-cell data
+				\end{itemize}
+			}	
+			\uncover<4->{
+				SCORPIUS:
+				\begin{itemize}
+					\item Wordt gebruikt om veranderingen in cellen te onderzoeken en visualiseren
+				\end{itemize}
+			}
+		\end{column}
+	\end{columns}
 \end{frame}
 
+\begin{frame}
+	\frametitle{Wat brengt de toekomst?}
+		\begin{columns}
+		\begin{column}{0.3\textwidth}
+			\includegraphics[width=\linewidth]{figures/9_future/hca.png}
+		\end{column}
+		\begin{column}{0.7\textwidth}
+			Human cell atlas
+			\begin{itemize}
+				\item<2-> 50 biljoen cellen in het lichaam, maar hoeveel verschillende soorten cellen?
+				\item<3-> Doelstelling: Maak een Google Maps van cel(types)
+				\item<4-> Begrijpen van ziektes en ontwikkelen van nieuwe medicijnen
+			\end{itemize}
+		\end{column}
+	\end{columns}
+
+\end{frame}
+
+{
+	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{../cover/vertical4_small.jpg}}
+	\begin{frame}
+%		\bfseries \color{white}
+%		\begin{center}
+%			\textpdfrender{
+%				TextRenderingMode=FillStrokeClip,
+%				LineWidth=.2pt,
+%				FillColor=white,
+%				StrokeColor=black,
+%				MiterLimit=1
+%			}{Bedankt voor jullie aandacht}
+%		\end{center}
+	\end{frame}
+}
+
 \end{document}

### 2019-12-16 defence_public/scripts/5_scorpius.R

@@ -1,0 +1,64 @@
+library(SCORPIUS)
+library(tidyverse)
+library(gganimate)
+
+set.seed(1)
+
+data <- SCORPIUS::ginhoux
+expr <- data$expression
+sample_info <- data$sample_info
+
+dimred <- SCORPIUS::reduce_dimensionality(expr, ndim = 2)
+
+g2 <-
+  draw_trajectory_plot(dimred, point_size = 1) +
+  scale_x_continuous(breaks = -Inf) +
+  scale_y_continuous(breaks = -Inf)
+ggsave("figures/5_scorpius/dimred.pdf", g2, width = 3, height = 3)
+
+its <- 0:10
+walk(its, function(it) {
+  set.seed(1)
+  traj <- SCORPIUS::infer_trajectory(dimred, maxit = it, k = 0)
+
+  pl <- g2 +
+    geom_path(aes(Comp1, Comp2), as.data.frame(traj$path), colour = "red") +
+    labs(title = paste0("Iteration ", it))
+  # ggsave(paste0("figures/5_scorpius/princurve_it", it, ".pdf"), pl, width = 3, height = 3)
+  ggsave(paste0("figures/5_scorpius/princurve_it", it, ".png"), pl, width = 3, height = 3)
+})
+
+system("convert figures/5_scorpius/princurve_it*.png -delay 1000 figures/5_scorpius/princurve_loop.gif")
+file.remove("figures/5_scorpius/princurve_loop.avi"); system("ffmpeg -f image2 -r 1 -i figures/5_scorpius/princurve_it%d.png figures/5_scorpius/princurve_loop.avi")
+
+set.seed(1)
+traj <- SCORPIUS::infer_trajectory(dimred, maxit = 10, k = 0)
+
+ix <- sample.int(nrow(expr), 50)
+gimp <- SCORPIUS::gene_importances(expr, traj$time)
+
+expr_df <- data.frame(cell = names(traj$time),time = traj$time, as.matrix(expr[,c("Cd74", "Mpo", "Prtn3", "Siglech")])) %>%
+  gather(gene, value, -time, -cell)
+g6 <-
+  ggplot(expr_df) +
+  geom_point(aes(time, value), size = 1) +
+  geom_smooth(aes(time, value), colour = "red", fill = NA, size = 1) +
+  facet_wrap(~gene, scales = "free") +
+  theme_classic() +
+  scale_x_continuous(breaks = -Inf) +
+  scale_y_continuous(breaks = -Inf) +
[…]

### 2019-12-16 defence_public/scripts/rnaseq.R

@@ -1,61 +1,0 @@
-library(tidyverse)
-
-fig_size <- 3
-
-melt <- function(mat) {
-  mat %>%
-    magrittr::set_colnames(NULL) %>%
-    magrittr::set_rownames(NULL) %>%
-    reshape2::melt(varnames = c("y", "x"), value.name = "value") %>%
-    as_tibble()
-}
-raster <- function(meltdf, legend = FALSE, palette = "RdYlBu") {
-  g <-
-    ggplot(meltdf) +
-    geom_raster(aes(x, y, fill = value)) +
-    scale_fill_distiller(palette = palette, breaks = c(0, 1), labels = c("Min", "Max")) +
-    dynplot::theme_graph() +
-    labs(fill = "Expression") +
-    coord_cartesian(expand = FALSE)
-  if (!legend) {
-    g <- g + theme(legend.position = "none")
-  }
-  g
-}
-
-
-dat <- dyntoy::generate_dataset(id = "defence", model = "multifurcating", num_cells = 100, num_features = 101)
-
-counts <-
-  dat$counts %>%
-  as.matrix()
-colnames(counts) <- gsub("G", "gene ", colnames(counts))
-rownames(counts) <- gsub("C", "cell ", rownames(counts))
-write_tsv(data.frame(counts, check.names = FALSE) %>% rownames_to_column("cell name"), "figures/4_singlecellomics/rnaseq_matrix.tsv")
-
-g <-
-  counts %>%
-  melt() %>%
-  raster()
-
-ggsave("figures/4_singlecellomics/rnaseq_plot.pdf", g, width = 3, height = 3)
-
-
-countsmelt <-
-  reshape2::melt(counts, varnames = c("cell", "gene"), value.name = "count")
-countsmelt_sel <-
-  countsmelt %>%
-  filter(cell == "cell 1", gene %in% paste0("gene ", 1:5))
-g <- ggplot(countsmelt_sel) + geom_bar(aes(gene, count), stat = "identity") + theme_classic() + labs(x = NULL, y = "Aantal RNA transcripten") + coord_flip()
-ggsave("figures/4_singlecellomics/rnaseq_2dplot.pdf", g, width = 3, height = 3)
-
-
-g <-
-  counts %>%
-  melt() %>%
-  filter(y == 1, x %in% 1:5) %>%
-  raster()
-
[…]

### 2019-12-16 defence_public/scripts/scorpius.R

@@ -1,64 +1,0 @@
-library(SCORPIUS)
-library(tidyverse)
-library(gganimate)
-
-set.seed(1)
-
-data <- SCORPIUS::ginhoux
-expr <- data$expression
-sample_info <- data$sample_info
-
-dimred <- SCORPIUS::reduce_dimensionality(expr, ndim = 2)
-
-g2 <-
-  draw_trajectory_plot(dimred, point_size = 1) +
-  scale_x_continuous(breaks = -Inf) +
-  scale_y_continuous(breaks = -Inf)
-ggsave("figures/5_scorpius/dimred.pdf", g2, width = 3, height = 3)
-
-its <- 0:10
-walk(its, function(it) {
-  set.seed(1)
-  traj <- SCORPIUS::infer_trajectory(dimred, maxit = it, k = 0)
-
-  pl <- g2 +
-    geom_path(aes(Comp1, Comp2), as.data.frame(traj$path), colour = "blue") +
-    labs(title = paste0("Iteration ", it))
-  # ggsave(paste0("figures/5_scorpius/princurve_it", it, ".pdf"), pl, width = 3, height = 3)
-  ggsave(paste0("figures/5_scorpius/princurve_it", it, ".png"), pl, width = 3, height = 3)
-})
-
-system("convert figures/5_scorpius/princurve_it*.png -delay 1000 figures/5_scorpius/princurve_loop.gif")
-file.remove("figures/5_scorpius/princurve_loop.avi"); system("ffmpeg -f image2 -r 1 -i figures/5_scorpius/princurve_it%d.png figures/5_scorpius/princurve_loop.avi")
-
-set.seed(1)
-traj <- SCORPIUS::infer_trajectory(dimred, maxit = 10, k = 0)
-
-ix <- sample.int(nrow(expr), 50)
-gimp <- SCORPIUS::gene_importances(expr, traj$time)
-
-expr_df <- data.frame(cell = names(traj$time),time = traj$time, as.matrix(expr[,c("Cd74", "Mpo", "Prtn3", "Siglech")])) %>%
-  gather(gene, value, -time, -cell)
-g6 <-
-  ggplot(expr_df) +
-  geom_point(aes(time, value), size = 1) +
-  geom_smooth(aes(time, value), colour = "blue", fill = NA, size = 1) +
-  facet_wrap(~gene, scales = "free") +
-  theme_classic() +
-  scale_x_continuous(breaks = -Inf) +
-  scale_y_continuous(breaks = -Inf) +
[…]
